### 2020-12-04 manuscript/manuscript.Rmd
and so it begins... :seedling:

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu

abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
output: 
  rticles::arxiv_article:
    keep_tex: true
---

# Introduction

# RL essentials

- terminology, problem definition
- discussion / table of current benchmark algorithms (DQN, PPO, DDPG, A2C, TD3, SAC)
- training, hyperparameter tuning

# Examples

```{r setup, cache = FALSE}
library(tidyverse)
library(reticulate)
```

## Finding a known optimal solution using RL

fishing gym?

```{r cache=TRUE}
## Python dependencies
gym         <- import ("gym")
gym_fishing <- import("gym_fishing")
sb3         <- import ("stable_baselines3")

# train an agent (model) on one of the environments:
env <- gym$make("fishing-v1")
a2c <- sb3$A2C('MlpPolicy', env, verbose=0L) # Must use L for integers!
a2c$learn(total_timesteps=300000L)
sims <- env$simulate(a2c, reps = 500L)
```

```{r}
sims %>%
  group_by(time) %>% 
  summarise(ave_state = mean(state),
            sd_state = sd(state),
            ave_action = mean(action),
            ave_reward = mean(reward),
            .groups = "drop") %>%
  ggplot(aes(time, ave_state)) +  geom_line()

```

```{r}
sims %>% filter(rep < 50) %>% ggplot(aes(time, state, group = as.factor(rep))) + geom_line(alpha=0.1)
```

```{r}
policy <- env$policyfn(a2c, reps = 50L)
policy %>% ggplot(aes(state, action, group=rep)) + geom_point()
```

## Example of solving a problem that is difficult/impossible for conventional methods

two use cases to consider: 
  - model is known but too complex to solve by (approximate) dynamic programming etc, or non-markovian
  - model is unknown

model uncertainty/transfer learning? nonstationary dynamics?

## Example of an open problem without an RL solution

fire gym?

# Open issues

- wild west of AI?
- reproducibility
- AI ethics, transparency

# a vision going forward

- our public/open source environments and agents, leaderboard
- the need for collaborative/competitive development both to make better benchmark environments and better agents

### 2020-12-04 manuscript/manuscript.Rmd
gym_fishing example

@@
 authors:
+  - name: Marcus Lapeyrolerie
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+  - name: Melissa Chapman
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+  - name: Kari Norman
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
   - name: Carl Boettiger
@@
 
-```{r setup, cache = FALSE}
+```{r setup, cache = FALSE, include = FALSE}
 library(tidyverse)
+library(patchwork)
 library(reticulate)
+
+knitr::opts_chunk$set(echo=FALSE)
 ```
@@
 env <- gym$make("fishing-v1")
-a2c <- sb3$A2C('MlpPolicy', env, verbose=0L) # Must use L for integers!
+a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
 a2c$learn(total_timesteps=300000L)
@@
 ```{r}
-sims %>%
-  group_by(time) %>% 
-  summarise(ave_state = mean(state),
-            sd_state = sd(state),
-            ave_action = mean(action),
-            ave_reward = mean(reward),
-            .groups = "drop") %>%
-  ggplot(aes(time, ave_state)) +  geom_line()
-
+fig_fish_sims <- sims %>% 
+  filter(rep < 50) %>%
+  ggplot(aes(time, state, group = as.factor(rep))) + geom_line(alpha=0.1)
+policy <- env$policyfn(a2c, reps = 50L)
+fig_fish_policy <- policy %>% ggplot(aes(state, state - action, group=rep)) + geom_point()
 ```
 
-```{r}
-sims %>% filter(rep < 50) %>% ggplot(aes(time, state, group = as.factor(rep))) + geom_line(alpha=0.1)
-```
-
-```{r}
-policy <- env$policyfn(a2c, reps = 50L)
-policy %>% ggplot(aes(state, action, group=rep)) + geom_point()
+```{r fig1, fig.cap=""}
+fig_fish_sims + fig_fish_policy
 ```

### 2020-12-04 manuscript/manuscript.Rmd
:pen: draft intro material

@@
 # Introduction
+
+Despite rapidly expanding interest in machine learning (ML) and artificial intelligence (AI) applications to ecology, the use of reinforcement learning (RL) has been largely neglected, which have focused primarily on applications of either supervised learning [@examples] or unsupervised learning [@examples].
+In third learning paradigm in AI, reinforcement learning, a _software agent_ learns by repeated experience to select _actions_ to maximize some cumulative reward.
+RL is an incredibly general approach, and the one that in principle looks most like artificial general intelligence.
+RL has been notable for successes such as AlphaZero, a single agent that can beat the best human players in go, chess, and shogi [@alphazero], as well as applications in robotics [@robotics], but also faces major challenges [@henderson2017].
+
+We believe RL approaches hold particular interest for conservation decision-making problems, for two reasons.
+First, ecological systems are complex and difficult to represent accurately using models that remain tractable with classical approaches.
+For example, convenient mathematical assumptions such as stationarity, often required by classical techniques [@Marescot2013], stand in stark odds with the picture of an environment which is constantly changing.
+Second, RL approaches do not require massive amounts of data, in stark contrast with other more widely used applications of ML.
+Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data often remains too sparse in the details that matter.
+And in a changing environment, no amount of historical data may be able to train supervised or unsupervised ML approaches to anticipate a future that has never been observed.
+Stationarity is an even stronger assumption there than in classical approaches.
+However, RL are not trained by historical data alone, but rather, they rely on computer simulations.
+Simulations are our most powerful approach to exploring future scenario, and simulated 'data' represents the largest contribution to the exponential growth of data in climate science [@].
+As we illustrate here, a software agent can be trained to successfully navigate hypothetical scenarios through it's experience in simulation that would be difficult or impossible to solve with classical methods.
+In this manner, RL can compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
+
+If RL has great promise, the approach also imposes great risk.
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but fail unexpectedly under seemingly similar conditions.
+Despite recent progress, Deep RL is a young field and stability and robustness of RL agents remain largely open problems.
+Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as propietary intellectual property.
 

### 2020-12-08 manuscript/manuscript.Rmd
comparison to known solution

@@
 
+np <- import("numpy")
+np.random.seed(24L)
+
 knitr::opts_chunk$set(echo=FALSE)
@@
 
+## initialize the environment
+env <- gym$make("fishing-v1")
+```
+
+```{r eval=!file.exists("cache/a2c.zip")}
 # train an agent (model) on one of the environments:
-env <- gym$make("fishing-v1")
 a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
 a2c$learn(total_timesteps=300000L)
-sims <- env$simulate(a2c, reps = 500L)
+a2c$save("cache/a2c")
 ```
@@
 ```{r}
-fig_fish_sims <- sims %>% 
-  filter(rep < 50) %>%
-  ggplot(aes(time, state, group = as.factor(rep))) + geom_line(alpha=0.1)
+# Simulate management under the trained agent
+a2c <- sb3$A2C$load("cache/a2c")
+sims <- env$simulate(a2c, reps = 500L)
+
 policy <- env$policyfn(a2c, reps = 50L)
-fig_fish_policy <- policy %>% ggplot(aes(state, state - action, group=rep)) + geom_point()
+```
+
+```{r}
+# Simulate under the optimal solution (given the model)
+opt <- gym_fishing$models$escapement(env)
+opt_sims <- env$simulate(opt, reps = 500L)
+opt_policy <- env$policyfn(opt) %>% rename(optimal = action)
+
+sims_df <- bind_rows(sims, opt_sims, .id = "model")
+policy <- left_join(policy, opt_policy)
+```
+
+```{r}
+ymin <- function(x) last(x[(ntile(x, 20)==1)])
+ymax <- function(x) last(x[(ntile(x, 20)==19)])
+
+fig_sims <- 
+sims_df %>% 
+  group_by(time, model) %>% 
+  summarise(ymin = ymin(state),
+            ymax = ymax(state),
+            state = mean(state), .groups = "drop") %>%
+  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
+  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
+```
+
+```{r}
+fig_policy <- 
+  policy %>% ggplot(aes(state, state - action, group=rep)) + 
+  geom_point(alpha=.05) + geom_line(aes(y = state - optimal), col = "blue") + 
+  coord_cartesian(xlim = c(0, 1.5), ylim=c(0,0.9))
 ```
@@
 ```{r fig1, fig.cap=""}
-fig_fish_sims + fig_fish_policy
+fig_sims + fig_policy
 ```

### 2020-12-08 manuscript/manuscript.Rmd
a bit more context

@@
 opt_policy <- env$policyfn(opt) %>% rename(optimal = action)
+```
 
-sims_df <- bind_rows(sims, opt_sims, .id = "model")
+```{r}
+sims_df <- bind_rows(sims, opt_sims, .id = "model") %>%
+  mutate(model = c("A2C", "optimal")[as.integer(model)])
+
 policy <- left_join(policy, opt_policy)
+```
+
+```{r}
+gamma <- 1 #discount
+fig_reward <- sims_df %>% 
+  group_by(rep, model) %>%
+  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
+  group_by(time, model) %>%
+  summarise(mean_reward = mean(cum_reward), 
+            sd = sd(cum_reward), .groups = "drop") %>% 
+  ggplot(aes(time, mean_reward)) + 
+  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
+                  ymax = mean_reward + 2*sd, fill = model),
+              alpha=0.25, show.legend = FALSE) + 
+  geom_line(aes(col = model)) + ylab("mean cumulative reward")
 ```
@@
 ```{r fig1, fig.cap=""}
-fig_sims + fig_policy
+fig_sims / ( fig_policy + fig_reward)
 ```
+Note that the A2C policy learns to shut down fishing entirely below a certain stock size, without any variation.
+However, it keeps the fishery closed even once stock sizes are slightly higher than $B_{MSY}$.
+Once it begins fishing, it slightly under-harvests for stock sizes a little over $B_{MSY}$, but tends to over-harvest for very large stock sizes (rarely ever realized, as fishing pressure is enough to keep it out of this range).
+Note also that the harvest varies across replicates, creating a fuzzy region in the policy function.
+Under such a policy, A2C maintains a stock size on average significantly above $B_{MSY}$, though also sees higher volatility particularly on the low side.
+Note that due to the stochastic nature of the environment, any given replicate A2C may out-perform the optimal policy, though on average the optimal policy provides consistently better yields.
 

### 2020-12-09 manuscript/appendix.Rmd
:pen:

--- 2020-12-08 manuscript/manuscript.Rmd
@@
 ---
-title: Deep Reinforcement Learning for Conservation Decisions
+title: Appendix for: Deep Reinforcement Learning for Conservation Decisions
 authors:
@@
 
-# Introduction
-
-Despite rapidly expanding interest in machine learning (ML) and artificial intelligence (AI) applications to ecology, the use of reinforcement learning (RL) has been largely neglected, which have focused primarily on applications of either supervised learning [@examples] or unsupervised learning [@examples].
-In third learning paradigm in AI, reinforcement learning, a _software agent_ learns by repeated experience to select _actions_ to maximize some cumulative reward.
-RL is an incredibly general approach, and the one that in principle looks most like artificial general intelligence.
-RL has been notable for successes such as AlphaZero, a single agent that can beat the best human players in go, chess, and shogi [@alphazero], as well as applications in robotics [@robotics], but also faces major challenges [@henderson2017].
-
-We believe RL approaches hold particular interest for conservation decision-making problems, for two reasons.
-First, ecological systems are complex and difficult to represent accurately using models that remain tractable with classical approaches.
-For example, convenient mathematical assumptions such as stationarity, often required by classical techniques [@Marescot2013], stand in stark odds with the picture of an environment which is constantly changing.
-Second, RL approaches do not require massive amounts of data, in stark contrast with other more widely used applications of ML.
-Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data often remains too sparse in the details that matter.
-And in a changing environment, no amount of historical data may be able to train supervised or unsupervised ML approaches to anticipate a future that has never been observed.
-Stationarity is an even stronger assumption there than in classical approaches.
-However, RL are not trained by historical data alone, but rather, they rely on computer simulations.
-Simulations are our most powerful approach to exploring future scenario, and simulated 'data' represents the largest contribution to the exponential growth of data in climate science [@].
-As we illustrate here, a software agent can be trained to successfully navigate hypothetical scenarios through it's experience in simulation that would be difficult or impossible to solve with classical methods.
-In this manner, RL can compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
-
-If RL has great promise, the approach also imposes great risk.
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but fail unexpectedly under seemingly similar conditions.
-Despite recent progress, Deep RL is a young field and stability and robustness of RL agents remain largely open problems.
-Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as propietary intellectual property.
-
-# RL essentials
-
-- terminology, problem definition
-- discussion / table of current benchmark algorithms (DQN, PPO, DDPG, A2C, TD3, SAC)
-- training, hyperparameter tuning
-
-# Examples
-
 ```{r setup, cache = FALSE, include = FALSE}
@@
 np <- import("numpy")
-np.random.seed(24L)
+np$random$seed(42L)
 
@@
 
-```{r cache=TRUE}
+```{r }
 ## Python dependencies
@@
 # train an agent (model) on one of the environments:
-a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
-a2c$learn(total_timesteps=300000L)
-a2c$save("cache/a2c")
+## FIXME set tuned hyperparameters
+td3 <- sb3$TD3('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
+td3$learn(total_timesteps=300000L)
+td3$save("cache/td3")
 ```
@@
 # Simulate management under the trained agent
-a2c <- sb3$A2C$load("cache/a2c")
-sims <- env$simulate(a2c, reps = 500L)
-
-policy <- env$policyfn(a2c, reps = 50L)
+td3 <- sb3$A2C$load("cache/td3")
+sims <- env$simulate(td3, reps = 500L)
+policy <- env$policyfn(td3, reps = 50L)
 ```
@@
 sims_df <- bind_rows(sims, opt_sims, .id = "model") %>%
-  mutate(model = c("A2C", "optimal")[as.integer(model)])
+  mutate(model = c("TD3", "optimal")[as.integer(model)])
 

### 2020-12-11 manuscript/appendix.Rmd
hyperparameters

@@
 
-```{r setup, cache = FALSE, include = FALSE}
+```{r knitr, include = FALSE}
+knitr::opts_chunk$set(echo=FALSE)
+```
+
+```{r setup, cache = FALSE}
 library(tidyverse)
@@
 
-knitr::opts_chunk$set(echo=FALSE)
+# force CPU-evaluation for reproducibility
+Sys.setenv("CUDA_VISIBLE_DEVICES"="")
+
 ```
@@
 ## initialize the environment
-env <- gym$make("fishing-v1")
+env <- gym$make("fishing-v1", sigma = 0.1)
 ```
@@
 # train an agent (model) on one of the environments:
-## FIXME set tuned hyperparameters
-td3 <- sb3$TD3('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
+
+#Trial 15 finished with value: 7.6994757652282715 and parameters: 
+hyper = list('gamma'= 0.995, 'lr'= 0.0001355522450968401, 'batch_size'= 128L, 
+         'buffer_size'= 10000L, 'episodic'= FALSE, 'train_freq'= 128L, 
+         'noise_type'= 'normal', 'noise_std'= 0.6656948079225263, 
+         'net_arch'= 'big')
+policy_kwargs = list(net_arch=c(400L, 300L)) # big
+
+#non-episodic:
+hyper['gradient_steps'] = hyper['train_freq']
+hyper['n_episodes_rollout'] = -1
+n_actions = env$action_space$shape[0]
+hyper$action_noise = sb3$common$noise$NormalActionNoise(
+        mean=np$zeros(n_actions),
+        sigma= hyper[['noise_std']] * np$ones(n_actions)
+        )
+
+td3 = sb3$TD3('MlpPolicy', env,  verbose=0L, seed = 42L,
+            gamma = hyper[['gamma']],
+            learning_rate = hyper[['lr']],
+            batch_size = hyper[['batch_size']],            
+            buffer_size = hyper[['buffer_size']],
+            action_noise = hyper[['action_noise']],
+            train_freq = hyper[['train_freq']],
+            gradient_steps = hyper[['train_freq']],
+            n_episodes_rollout = hyper[['n_episodes_rollout']],
+            policy_kwargs=policy_kwargs)
+
 td3$learn(total_timesteps=300000L)
@@
 # Simulate management under the trained agent
-td3 <- sb3$A2C$load("cache/td3")
+#td3 <- sb3$A2C$load("cache/td3") ## fixme, fails to load
 sims <- env$simulate(td3, reps = 500L)
@@
 ```
-Note that the A2C policy learns to shut down fishing entirely below a certain stock size, without any variation.
-However, it keeps the fishery closed even once stock sizes are slightly higher than $B_{MSY}$.
-Once it begins fishing, it slightly under-harvests for stock sizes a little over $B_{MSY}$, but tends to over-harvest for very large stock sizes (rarely ever realized, as fishing pressure is enough to keep it out of this range).
-Note also that the harvest varies across replicates, creating a fuzzy region in the policy function.
-Under such a policy, A2C maintains a stock size on average significantly above $B_{MSY}$, though also sees higher volatility particularly on the low side.
-Note that due to the stochastic nature of the environment, any given replicate A2C may out-perform the optimal policy, though on average the optimal policy provides consistently better yields.
-
-## Example of solving a problem that is difficult/impossible for conventional methods
-
-two use cases to consider: 
-  - model is known but too complex to solve by (approximate) dynamic programming etc, or non-markovian
-  - model is unknown
-
-model uncertainty/transfer learning? nonstationary dynamics?
-
-## Example of an open problem without an RL solution
-
-fire gym?
-
-# Open issues
-
-- wild west of AI?
-- reproducibility
-- AI ethics, transparency
-
-# a vision going forward
-
-- our public/open source environments and agents, leaderboard
-- the need for collaborative/competitive development both to make better benchmark environments and better agents
+```{r}
+```

### 2020-12-12 manuscript/appendix.Rmd
td3 + a2c results & cache

@@
 ---
-title: Appendix for: Deep Reinforcement Learning for Conservation Decisions
+title: "Appendix for: Deep Reinforcement Learning for Conservation Decisions"
 authors:
@@
 ```{r knitr, include = FALSE}
-knitr::opts_chunk$set(echo=FALSE)
+knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
+                      fig.width = 7, fig.height = 4)
+ggplot2::theme_set(ggplot2::theme_bw())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(6)
+txtcolor <- "#586e75"
 ```
 
-```{r setup, cache = FALSE}
+```{r setup, cache = FALSE, message=FALSE}
 library(tidyverse)
@@
 
-# force CPU-evaluation for reproducibility
-Sys.setenv("CUDA_VISIBLE_DEVICES"="")
+# force CPU-evaluation if needing perfect reproducibility
+Sys.setenv("CUDA_VISIBLE_DEVICES"="0")
 
@@
 
-```{r eval=!file.exists("cache/a2c.zip")}
+```{r eval=!file.exists("cache/td3.zip")}
 # train an agent (model) on one of the environments:
@@
 # Simulate management under the trained agent
-#td3 <- sb3$A2C$load("cache/td3") ## fixme, fails to load
-sims <- env$simulate(td3, reps = 500L)
-policy <- env$policyfn(td3, reps = 50L)
+td3 <- sb3$TD3$load("cache/td3")
+td3_sims <- env$simulate(td3, reps = 500L)
+td3_policy <- env$policyfn(td3, reps = 50L)
+```
+
+```{r eval=!file.exists("cache/a2c.zip")}
+# train an agent (model) on one of the environments:
+a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
+a2c$learn(total_timesteps=300000L)
+a2c$save("cache/a2c")
+```
+
+```{r}
+# Simulate management under the trained agent
+a2c <- sb3$A2C$load("cache/a2c")
+a2c_sims <- env$simulate(a2c, reps = 500L)
+a2c_policy <- env$policyfn(a2c, reps = 50L)
 ```
@@
 opt_sims <- env$simulate(opt, reps = 500L)
-opt_policy <- env$policyfn(opt) %>% rename(optimal = action)
+opt_policy <- env$policyfn(opt)
 ```
@@
 ```{r}
-sims_df <- bind_rows(sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "optimal")[as.integer(model)])
+sims_df <- bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
+  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
 
-policy <- left_join(policy, opt_policy)
-```
+policy_df <- bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
+  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
 
-```{r}
 gamma <- 1 #discount
-fig_reward <- sims_df %>% 
+reward_df <- sims_df %>% 
   group_by(rep, model) %>%
@@
   summarise(mean_reward = mean(cum_reward), 
-            sd = sd(cum_reward), .groups = "drop") %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model)) + ylab("mean cumulative reward")
+            sd = sd(cum_reward), .groups = "drop")
+
 ```
@@
 fig_policy <- 
-  policy %>% ggplot(aes(state, state - action, group=rep)) + 
-  geom_point(alpha=.05) + geom_line(aes(y = state - optimal), col = "blue") + 
-  coord_cartesian(xlim = c(0, 1.5), ylim=c(0,0.9))
+  policy_df %>% ggplot(aes(state, state - action, 
+                           group=interaction(rep, model),
+                           col = model)) + 
+  geom_line(show.legend = FALSE) + 
+  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) + 
+  ylab("escapement")
+```
+
+```{r}
+
+fig_reward <- reward_df %>% 
+  ggplot(aes(time, mean_reward)) + 
+  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
+                  ymax = mean_reward + 2*sd, fill = model),
+              alpha=0.25, show.legend = FALSE) + 
+  geom_line(aes(col = model), show.legend = FALSE) + 
+  ylab("reward")
 ```

### 2020-12-12 manuscript/manuscript.Rmd
td3 + a2c results & cache

--- 2020-12-04 manuscript/manuscript.Rmd
@@
 
-```{r setup, cache = FALSE, include = FALSE}
-library(tidyverse)
-library(patchwork)
-library(reticulate)
-
-knitr::opts_chunk$set(echo=FALSE)
-```
-
 ## Finding a known optimal solution using RL
 
-fishing gym?
-
-```{r cache=TRUE}
-## Python dependencies
-gym         <- import ("gym")
-gym_fishing <- import("gym_fishing")
-sb3         <- import ("stable_baselines3")
-
-# train an agent (model) on one of the environments:
-env <- gym$make("fishing-v1")
-a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
-a2c$learn(total_timesteps=300000L)
-sims <- env$simulate(a2c, reps = 500L)
-```
-
-```{r}
-fig_fish_sims <- sims %>% 
-  filter(rep < 50) %>%
-  ggplot(aes(time, state, group = as.factor(rep))) + geom_line(alpha=0.1)
-policy <- env$policyfn(a2c, reps = 50L)
-fig_fish_policy <- policy %>% ggplot(aes(state, state - action, group=rep)) + geom_point()
-```
-
-```{r fig1, fig.cap=""}
-fig_fish_sims + fig_fish_policy
-```
+Note that the A2C policy learns to shut down fishing entirely below a certain stock size, without any variation.
+However, it keeps the fishery closed even once stock sizes are slightly higher than $B_{MSY}$.
+Once it begins fishing, it slightly under-harvests for stock sizes a little over $B_{MSY}$, but tends to over-harvest for very large stock sizes (rarely ever realized, as fishing pressure is enough to keep it out of this range).
+Note also that the harvest varies across replicates, creating a fuzzy region in the policy function.
+Under such a policy, A2C maintains a stock size on average significantly above $B_{MSY}$, though also sees higher volatility particularly on the low side.
+Note that due to the stochastic nature of the environment, any given replicate A2C may out-perform the optimal policy, though on average the optimal policy provides consistently better yields.
 

### 2020-12-12 manuscript/appendix.Rmd
move code to appendix :pen:

@@
 ```{r knitr, include = FALSE}
-knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
+knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
                       fig.width = 7, fig.height = 4)
@@
 ```{r}
+write_csv(sims_df, "figs/sims_df.csv")
+write_csv(policy_df, "figs/policy_df.csv")
+write_csv(reward_df, "figs/reward_df.csv")
+```
+
+```{r}
 ymin <- function(x) last(x[(ntile(x, 20)==1)])

### 2020-12-12 manuscript/manuscript.Rmd
move code to appendix :pen:

@@
 
+```{r knitr, include = FALSE}
+library(tidyverse)
+library(patchwork)
+knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
+                      fig.width = 7, fig.height = 4)
+ggplot2::theme_set(ggplot2::theme_bw())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(6)
+txtcolor <- "#586e75"
+
+```
+
 ## Finding a known optimal solution using RL
+
+```{r}
+sims_df <- read_csv( "figs/sims_df.csv")
+policy_df <- read_csv("figs/policy_df.csv")
+reward_df <- read_csv("figs/reward_df.csv")
+
+ymin <- function(x) last(x[(ntile(x, 20)==1)])
+ymax <- function(x) last(x[(ntile(x, 20)==19)])
+
+fig_sims <- 
+sims_df %>% 
+  group_by(time, model) %>% 
+  summarise(ymin = ymin(state),
+            ymax = ymax(state),
+            state = mean(state), .groups = "drop") %>%
+  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
+  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
+
+fig_policy <- 
+  policy_df %>% ggplot(aes(state, state - action, 
+                           group=interaction(rep, model),
+                           col = model)) + 
+  geom_line(show.legend = FALSE) + 
+  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) + 
+  ylab("escapement")
+
+fig_reward <- reward_df %>% 
+  ggplot(aes(time, mean_reward)) + 
+  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
+                  ymax = mean_reward + 2*sd, fill = model),
+              alpha=0.25, show.legend = FALSE) + 
+  geom_line(aes(col = model), show.legend = FALSE) + 
+  ylab("reward")
+```
+
+```{r fig1, fig.cap=""}
+fig_sims / ( fig_policy + fig_reward)
+```
 

### 2021-01-13 manuscript/appendix.Rmd
WIP

@@
 scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(6)
+pal <- ggthemes::solarized_pal()(8)
 txtcolor <- "#586e75"
@@
 
-## Finding a known optimal solution using RL
+# Finding a known optimal solution using RL
 
-fishing gym?
+## Sustainable Harvest Quotas
 
@@
 fig_policy <- 
-  policy_df %>% ggplot(aes(state, state - action, 
+  policy_df %>% ggplot(aes(state, action, 
                            group=interaction(rep, model),
@@
   geom_line(show.legend = FALSE) + 
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) + 
-  ylab("escapement")
+  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) 
 ```
@@
 ```
+
+## Ecological tipping points
+
 ```{r}
+gym_conservation <- import("gym_conservation")
+env <- gym$make("conservation-v6")
 ```
+
+```{r}
+# Simulate management under the trained agent
+TD3 <- sb3$TD3$load("cache/td3-conservation")
+TD3_sims <- env$simulate(TD3, reps = 100L)
+TD3_policy <- env$policyfn(TD3, reps = 10L)
+```
+
+```{r}
+# Simulate under the steady-state solution (given the model)
+K = 1.5
+alpha = 0.001
+opt <- gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
+opt_sims <- env$simulate(opt, reps = 100L)
+opt_policy <- env$policyfn(opt)
+```
+
+```{r}
+sims_df <- bind_rows(TD3_sims, opt_sims, .id = "model") %>%
+  mutate(model = c("TD3", "steady-state")[as.integer(model)])
+
+policy_df <- bind_rows(TD3_policy, opt_policy, .id = "model") %>%
+  mutate(model = c("TD3", "steady-state")[as.integer(model)])
+
+gamma <- 1 #discount
+reward_df <- sims_df %>% 
+  group_by(rep, model) %>%
+  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
+  group_by(time, model) %>%
+  summarise(mean_reward = mean(cum_reward), 
+            sd = sd(cum_reward), .groups = "drop")
+
+```
+
+```{r}
+write_csv(sims_df, "figs/tipping_sims_df.csv")
+write_csv(policy_df, "figs/tipping_policy_df.csv")
+write_csv(reward_df, "figs/tipping_reward_df.csv")
+```
+
+```{r}
+# ensemble statistics
+ymin <- function(x) last(x[(ntile(x, 20)==1)])
+ymax <- function(x) last(x[(ntile(x, 20)==19)])
+
+fig_sims <- 
+  sims_df %>%
+   group_by(time, model) %>% 
+   summarise(ymin = ymin(state),
+            ymax = ymax(state),
+            state = mean(state), .groups = "drop") %>%
+  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
+   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
+```
+
+```{r}
+# Some individual replicates, for comparison
+sims_df %>%
+  filter(rep < 100, time < 500) %>%
+  ggplot(aes(time, state, col=model, group=interaction(model, rep))) +
+  geom_line(alpha = .1, show.legend = FALSE) + facet_wrap(~model)
+```
+
+```{r}
+# Some individual replicates, for comparison
+is_low <- sims_df %>% filter(model == "TD3") %>% group_by(rep, model) %>% summarize(low = sum(state < .2) > 3) %>% filter(low)
+sims_df %>% inner_join(is_low) %>%
+  ggplot(aes(time, state,  group=interaction(model, rep))) +
+  geom_line(color = pal[3], show.legend = FALSE) + facet_wrap(~rep)
+```
+
+```{r}
+# Some individual replicates, for comparison
+is_low <- sims_df %>% filter(model == "steady-state") %>% group_by(rep, model) %>% summarize(low = sum(state < .2) > 3) %>% filter(low) %>% head(16)
+sims_df %>% inner_join(is_low) %>%
+  ggplot(aes(time, state, group=interaction(model, rep))) +
+  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
+```
+
+```{r}
+fig_policy <- 
+  policy_df %>% ggplot(aes(state, action, 
+                           group=interaction(rep, model),
+                           col = model)) + 
+  geom_line(show.legend = FALSE) 
+```
+
+```{r}
+
+fig_reward <- reward_df %>% 
+  ggplot(aes(time, mean_reward)) + 
+  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
+                  ymax = mean_reward + 2*sd, fill = model),
+              alpha=0.25, show.legend = FALSE) + 
+  geom_line(aes(col = model), show.legend = FALSE) + 
+  ylab("reward")
+```
+
+```{r fig2, fig.cap=""}
+fig_sims / ( fig_policy + fig_reward)
+```
+
+----

### 2021-01-13 manuscript/manuscript.Rmd
WIP

@@
 Despite recent progress, Deep RL is a young field and stability and robustness of RL agents remain largely open problems.
-Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as propietary intellectual property.
+Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.
 
@@
 # Examples
+
+## Sustainable harvest
+
+We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.
+The first example focuses on the well-studied problem of setting harvest quotas in fisheries management [@Schaefer1954; @Clark1974; @Reed1979; @Worm2006; @Costello2018].
+Given an appropriately estimated model, the optimal harvest policy can be determined precisely (e.g.
+by stochastic dynamic programming, SDP, see @Marescot2013), and under most circumstances takes the form of a "constant-escapement" policy [see mathematical proof in @Reed1979].
+This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
+The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
+Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@].
+
+For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 or @Reed1979 (covering the stochastic case).
+In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
+In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
+From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
+In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."  The simplest simulation is merely an implementation of the estimated model itself.
+However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach is in principle agnostic to the details of the simulation.
 
@@
 ```
-
-## Finding a known optimal solution using RL
 

### 2021-01-19 manuscript/manuscript.Rmd
:pen:

@@
 
+<!-- no examples of issues / problems named!-->
+
 Despite rapidly expanding interest in machine learning (ML) and artificial intelligence (AI) applications to ecology, the use of reinforcement learning (RL) has been largely neglected, which have focused primarily on applications of either supervised learning [@examples] or unsupervised learning [@examples].
@@
 First, ecological systems are complex and difficult to represent accurately using models that remain tractable with classical approaches.
-For example, convenient mathematical assumptions such as stationarity, often required by classical techniques [@Marescot2013], stand in stark odds with the picture of an environment which is constantly changing.
+For example, convenient mathematical assumptions such as that a dynamical system is _autonomous_ (parameters do not vary in time) often required by classical techniques [@Marescot2013], stand in stark odds with the picture of an environment which is constantly changing.
 Second, RL approaches do not require massive amounts of data, in stark contrast with other more widely used applications of ML.
@@
 And in a changing environment, no amount of historical data may be able to train supervised or unsupervised ML approaches to anticipate a future that has never been observed.
-Stationarity is an even stronger assumption there than in classical approaches.
 However, RL are not trained by historical data alone, but rather, they rely on computer simulations.
 Simulations are our most powerful approach to exploring future scenario, and simulated 'data' represents the largest contribution to the exponential growth of data in climate science [@].
-As we illustrate here, a software agent can be trained to successfully navigate hypothetical scenarios through it's experience in simulation that would be difficult or impossible to solve with classical methods.
+As we illustrate here, a software agent can be trained to successfully navigate hypothetical scenarios through its experience in simulation that would be difficult or impossible to solve with classical methods.
 In this manner, RL can compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
@@
 If RL has great promise, the approach also imposes great risk.
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but fail unexpectedly under seemingly similar conditions.
-Despite recent progress, Deep RL is a young field and stability and robustness of RL agents remain largely open problems.
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions.
+Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems.
 Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.
+We discuss these issues and potential ways forward.
+
+# Sequential Decision Problems and Reinforcement Learning
+
+Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to  as the reward or utility).
+Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
+In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
+Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
+Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
+Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.
+
+This framework raises two conceptual challenges.
+First, sequential decision-making problem is distinct from the problem of model estimation.
+A sequential decision problem will typically treat the model as given, typically with some uncertainty around the process and possibly the measurmeents and model estimates as well.
+Thus, methods for solving sequential decision-making should not be mistaken as alternatives to methods designed for model estimation, such as regression, Bayesian heirarchical modeling, generalized additive models (GAMs) etc.
+Second, methods for describing and solving sequential decision-making problems are often given different terms by different communities to describe the same thing.
+Sequential decision problems are common to behavioral ecology [@Mangel], conservation [@Marescot2013] natural resource economics [@Clark1990], and some engineering fields [] where they are known as "optimal control" problems, while computer science, neuroscience, and robotics typically refer to them as 'reinfocement learning'.
 
 # RL essentials
+
+**Model-based vs Model-free RL**
+
+**Current algorithms**
 
@@
 
+## Beyond classic methods: Non-stationary models
+
+```{r}
+sims_df <-read_csv("figs/tipping_sims_df.csv")
+policy_df <- read_csv("figs/tipping_policy_df.csv")
+reward_df <- read_csv("figs/tipping_reward_df.csv")
+```
+
+```{r}
+fig_sims <- 
+  sims_df %>%
+   group_by(time, model) %>% 
+   summarise(ymin = ymin(state),
+            ymax = ymax(state),
+            state = mean(state), .groups = "drop") %>%
+  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
+   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
+
+fig_policy <- 
+  policy_df %>% ggplot(aes(state, action, 
+                           group=interaction(rep, model),
+                           col = model)) + 
+  geom_line(show.legend = FALSE)
+
+fig_reward <- reward_df %>% 
+  ggplot(aes(time, mean_reward)) + 
+  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
+                  ymax = mean_reward + 2*sd, fill = model),
+              alpha=0.25, show.legend = FALSE) + 
+  geom_line(aes(col = model), show.legend = FALSE) + 
+  ylab("reward")
+```
+
+```{r fig2, fig.cap=""}
+fig_sims / ( fig_policy + fig_reward)
+```
+
 ## Example of solving a problem that is difficult/impossible for conventional methods

### 2021-01-22 manuscript/manuscript.Rmd
Adding RL section

@@
 bibliography: references.bib
+header-includes:
+  - \usepackage{amsmath}
 output: 
@@
 
-Despite rapidly expanding interest in machine learning (ML) and artificial intelligence (AI) applications to ecology, the use of reinforcement learning (RL) has been largely neglected, which have focused primarily on applications of either supervised learning [@examples] or unsupervised learning [@examples].
-In third learning paradigm in AI, reinforcement learning, a _software agent_ learns by repeated experience to select _actions_ to maximize some cumulative reward.
-RL is an incredibly general approach, and the one that in principle looks most like artificial general intelligence.
-RL has been notable for successes such as AlphaZero, a single agent that can beat the best human players in go, chess, and shogi [@alphazero], as well as applications in robotics [@robotics], but also faces major challenges [@henderson2017].
+Despite a rapidly expanding interest in machine learning (ML) applications to ecology, the use of reinforcement learning (RL) has been largely neglected.
+To date, applications of machine learning to ecology have been predominantly limited to supervised learning [@examples], which attempts to solve problems in classification and regression, and unsupervised learning [@examples], which concentrates on data clustering.
+Reinforcement learning, meanwhile, is the unique subfield of machine learning that focuses on the problem of decision-making.
+The objective of reinforcement learning is to design a _software agent_ that learns by experience to select _actions_ to maximize some cumulative reward.
+RL is an incredibly general approach and has grown rapidly over the past 10 years as a consequence of the simultaneous rise of deep neural networks.
+In this short time, deep reinforcement learning has attained notable succeses such as AlphaZero, a single algorithm that can beat the best human players in go, chess, and shogi [@alphazero], as well as achieving dextrous manipulation in robotics [@robotics].
+However, reinforcement learning faces significant challenges  that limit its application to many real world problems[@henderson2017].
 
-We believe RL approaches hold particular interest for conservation decision-making problems, for two reasons.
+We believe RL approaches hold particular interest for conservation decision-making problems for two reasons.
 First, ecological systems are complex and difficult to represent accurately using models that remain tractable with classical approaches.
-For example, convenient mathematical assumptions such as that a dynamical system is _autonomous_ (parameters do not vary in time) often required by classical techniques [@Marescot2013], stand in stark odds with the picture of an environment which is constantly changing.
-Second, RL approaches do not require massive amounts of data, in stark contrast with other more widely used applications of ML.
+For example, convenient mathematical assumptions such as that a dynamical system is _autonomous_ (parameters do not vary in time) are often required by classical techniques [@Marescot2013]; these assumptions stand in stark odds with the picture of an environment which is constantly changing.
+RL, conversely, has excelled in solving a range of complex decision-making problems that were previously intractable with classical techniques like dynamic programming.
+Second, RL approaches do not require massive amounts of data in stark contrast with other more widely used applications of ML.
 Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data often remains too sparse in the details that matter.
 And in a changing environment, no amount of historical data may be able to train supervised or unsupervised ML approaches to anticipate a future that has never been observed.
-However, RL are not trained by historical data alone, but rather, they rely on computer simulations.
+However, RL algorithms do not need to depend on historical data alone, but rather, they can rely on computer simulations.
 Simulations are our most powerful approach to exploring future scenario, and simulated 'data' represents the largest contribution to the exponential growth of data in climate science [@].
@@
 Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions.
-Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems.
+Despite recent progress, Deep RL is a young field for which stability and robustness of its methods remain largely open problems.
 Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.
@@
 
-Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to  as the reward or utility).
-Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
-In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
+Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must decide what action to take to maximize some objective or cumulative "reward" in response to new information about the "state" of the system the decision-maker seeks to manage.
+Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether to declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
+In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region or how much each region would cost to purchase today are not sequential problems.
 Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
-Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
+Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast, explicitly or implicitly, how each possible action available to them will change that forecast.
 Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.
@@
 First, sequential decision-making problem is distinct from the problem of model estimation.
-A sequential decision problem will typically treat the model as given, typically with some uncertainty around the process and possibly the measurmeents and model estimates as well.
+A sequential decision problem will typically treat the model as given, perhaps with some uncertainty around the process, the measurements and model estimates as well.
 Thus, methods for solving sequential decision-making should not be mistaken as alternatives to methods designed for model estimation, such as regression, Bayesian heirarchical modeling, generalized additive models (GAMs) etc.
@@
 
-**Model-based vs Model-free RL**
+Before providing some examples on how deep reinforcement learning can be applied to ecological and conservation problems, we will provide a brief, general introduction to RL.
+The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
+This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
+The task for the agent is to learn a _policy_, $\pi$, that prescribes what action the agent should perform after observing a certain state.
+The agent interacts with the environment via trial and error in what is called the _training_ period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.
 
-**Current algorithms**
+When the environment is fully observable, i.e.
+when the state that the agent observes fully characterizes the system, the reinforcement learning problem can be formalized as a Markov Decision Process (MDP).
+An MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
+As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
+From this definition, we can provide an agent's trajectory distribution for a given policy as,
+$$
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
+$$
 
-- terminology, problem definition
-- discussion / table of current benchmark algorithms (DQN, PPO, DDPG, A2C, TD3, SAC)
-- training, hyperparameter tuning
+When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
+An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
+This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
+
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected cumulative, discounted reward:
+$$
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
+$$
+
+Note that this formalism extends to an infinite horizon case when $T= \infty$. While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
+
+Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high reward trajectory.
+There are 3 value functions that are commonly used across RL algorithms: the state value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
+The state-value function is defined as the expected return from a given state,
+$$
+  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+$$
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
+$$
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+$$
+And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
+The advantage function can then be succintly defined as
+$$
+  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
+$$
+Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
+Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
+
+**Survey of RL Algorithms**
+
+There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
+RL algorithms can generally be split into 2 groups, model-based and model-free RL.
+In model-based RL, the agent is attempting to learn the state transition operator, $T$.
+Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
+In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
+The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?].
+
+A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
+If an agent uses the same policy to take actions as the policy that the agent attempts to evaulate or improve, then this algorithm is on-policy.
+If an agent uses a different policy to select actions than the policy that the agent evaluates or improves, then this algorithm is off-policy.
+The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [cite SAC paper].
+Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop].
+
+The field of reinforcement learning has proliferated over the past 10 years on account of contributions from deep learning.
+In every reinforcement learning algorithm, the agent is attempting to learn some function, whether it is a policy function and/or a value function.
+Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions which we do not know the form of.
+Additionally, the ability of neural networks to autonomously devise low-dimensional representatoins of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
+The typical process for using neural networks in RL is that the agent is attempting to optimize one or more neural networks using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
+Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable to classic approaches.
+
+In practice, the general process to attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters.
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process.
+For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
+Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
+The hyperparameter tuning process consists of sampling then evaluating a set of hyperparameters, where tuning trials are often ranked by mean cumulative reward.
+It is important to note that hyperparameter tuning can often be a very time intensive process.
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be very time intensive.
+
+(Big table listing all the algorithms)
 
@@
 - reproducibility
+
 - AI ethics, transparency
+
+-model uncertainty part is concerning, you bake in assumptions into environment
+-can separate RL from having gyms implemented -> how do you compare different models, 
+-mention that customization is probably the way to go, baselines only go so far
 

### 2021-01-22 manuscript/manuscript.Rmd
Adding algorithm table

@@
 $$
+
 Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
@@
 $$
+
 And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
@@
 Additionally, the ability of neural networks to autonomously devise low-dimensional representatoins of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
-The typical process for using neural networks in RL is that the agent is attempting to optimize one or more neural networks using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
-Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable to classic approaches.
+The typical process for using neural networks in RL is that the agent is attempting to optimize a policy and/or value network using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
+Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable with classical methods.
 
-In practice, the general process to attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters.
+In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters.
 Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process.
@@
 
-(Big table listing all the algorithms)
+```{=latex}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+Algorithm & Description & Model & Policy & Action Space & State Space \\
+\hline
+MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
+\hline
+MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
+\hline
+A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
+\hline
+A3C  & Asynchronous A2C & Model-free & On-policy & &\\
+\hline
+TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
+\hline
+PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
+\hline
+DQN & Deep Q Networks & Model-free & Off-policy& &\\
+\hline
+DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
+\hline
+TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
+\hline
+SAC & Soft Actor Critic & Model-free & Off-policy & &\\
+\hline
+IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
+\hline
+APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
+\hline
+\end{tabular}
+\end{center}
+```
 

### 2021-01-25 manuscript/manuscript.Rmd
Small edits

@@
 The hyperparameter tuning process consists of sampling then evaluating a set of hyperparameters, where tuning trials are often ranked by mean cumulative reward.
-It is important to note that hyperparameter tuning can often be a very time intensive process.
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be very time intensive.
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
 

### 2021-02-04 manuscript/manuscript.Rmd
Starting on open issues section

@@
 
+Deep reinforcement learning has only become an established subfield of AI over the past decade; and, as with most young fields, there are a range of unsolved problems.
+In practice, the foremost problem with DRL is instability.
+For a range of reasons, including high variance and the exploration vs.
+exploitation trade-off, model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
+For instance, it is not uncommon to observe vastly different behavior when two agents are initialized with different random seeds [cite ibid].
+Since model-based and model-free algorithms typically have a large number of hyperparameters, each of which can contribute to training instability, it is practically impossible to identify the source of instability.
+The hope of the RL practitioner is that after doing an exhaustive hyperparameter search, the practitioner will find a set of hyperparameters that results in a robust, well-performing agent, but there are no guarantees that hyperparameter tuning will lead to this outcome.
+
+In the context of conservation, we anticipate that there will be a host of challenges.
+Interpretability is likely to be the biggest barrier to adoption.
+Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of an RL-trained agent.
+Another challenge is that DRL algorithms are very sample inefficient, and to overcome sample inefficiency for conservation decision-making problems that are often data poor, relying on simulations to train agents will be necessary.
+While there are positives to using simulated data, any assumptions or biases that are included in the model will affect agent behavior.
+
+A problem that is central to AI safety is selecting the right objective. There have been many examples in which explicit reward design goes horribly wrong [cite]. ...
+
 - wild west of AI?
-- reproducibility
+-can separate RL from having gyms implemented -> how do you compare different models, 
+-mention that customization is probably the way to go, baselines only go so far
 
 - AI ethics, transparency
-
--model uncertainty part is concerning, you bake in assumptions into environment
--can separate RL from having gyms implemented -> how do you compare different models, 
--mention that customization is probably the way to go, baselines only go so far
 

### 2021-02-04 manuscript/manuscript.Rmd
More bits to open issues

@@
 
-In the context of conservation, we anticipate that there will be a host of challenges.
+In the context of conservation, we anticipate that there will be a host of specific challenges to using DRL methods.
 Interpretability is likely to be the biggest barrier to adoption.
-Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of an RL-trained agent.
+Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
 Another challenge is that DRL algorithms are very sample inefficient, and to overcome sample inefficiency for conservation decision-making problems that are often data poor, relying on simulations to train agents will be necessary.
@@
 
-A problem that is central to AI safety is selecting the right objective. There have been many examples in which explicit reward design goes horribly wrong [cite]. ...
+A problem that is central to optimization in decision-making is using the right objective.
+There have been many examples in which explicit reward design goes horribly wrong [cite].
+And for ecological systems, that are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
+One recently proposed framework to avoiding bad reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
+Yet, in the context of training an agent to say conserve a species, there are questions that could easily arise, foremost will the agent infer that humans don't care about a certain species so acts in a way that leads to that species' extinction?
+There is a definite need for exploration of the ethics of using DRL for conservation.
+
+Specific to DRL, there
 

### 2021-02-05 manuscript/manuscript.Rmd
Finishing first pass at open issues

@@
 txtcolor <- "#586e75"
-
 ```
@@
 Another challenge is that DRL algorithms are very sample inefficient, and to overcome sample inefficiency for conservation decision-making problems that are often data poor, relying on simulations to train agents will be necessary.
-While there are positives to using simulated data, any assumptions or biases that are included in the model will affect agent behavior.
+While there are positives to using simulated data, any assumptions or biases that are included in the model will determine agent behavior.
 
-A problem that is central to optimization in decision-making is using the right objective.
+An issue that is central to any applied optimization problem is using the right objective.
 There have been many examples in which explicit reward design goes horribly wrong [cite].
-And for ecological systems, that are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
-One recently proposed framework to avoiding bad reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
+And for ecological systems, which are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
+Moreover, specifying a reward function for conservation brings up various ethical concerns -- e.g.
+who gets to specify the reward function and will the values of the greater community be represented in the reward function.
+One recently proposed framework to overcome challenges with explicit reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
 Yet, in the context of training an agent to say conserve a species, there are questions that could easily arise, foremost will the agent infer that humans don't care about a certain species so acts in a way that leads to that species' extinction?
@@
 
-Specific to DRL, there
+Ultimately, we anticipate that getting DRL to work well on fully detailed conservation decision-making problems will be challenging.
+Beyond typically having large state-action spaces for agents to explore, conservation problems tend to have sparse reward structures.
+E.g.
+rarely does an action in conservation lead to immediate rewards; instead, benefits are typically observed after actions have been performed.
+Sparse reward structures are challenging for DRL agents, because of the credit assignment problem [cite].
+To solve sparse reward environments, many researchers have employed customized algorithms to achieve better performance than standard DRL methods [cite Montezuma].
+For conservation, we anticipate that there will likely be a similar pattern, wherein bespoke RL methods will attain better performance than general methods.
 

### 2021-04-07 manuscript/appendix.Rmd
appendix pdf

@@
 # Simulate management under the trained agent
+# See https://github.com/boettiger-lab/conservation-agents/blob/4baaa6c9b994085ca56c42364f7affcc58b0ef3b/conservation/TD3.py
+
 TD3 <- sb3$TD3$load("cache/td3-conservation")

### 2021-04-09 manuscript/appendix.Rmd
update requirements.txt

@@
 ```{r setup, cache = FALSE, message=FALSE}
+# R dependencies
 library(tidyverse)
@@
 
+## Python dependencies loaded via R
+sb3         <- import ("stable_baselines3")
+gym         <- import ("gym")
+gym_fishing <- import("gym_fishing")
+gym_conservation <- import("gym_conservation")
+
+## reproducible settings
 np <- import("numpy")
 np$random$seed(42L)
-
-# force CPU-evaluation if needing perfect reproducibility
-Sys.setenv("CUDA_VISIBLE_DEVICES"="0")
+# Set to "" to force CPU-evaluation if needing perfect reproducibility
+Sys.setenv("CUDA_VISIBLE_DEVICES"="")
 
@@
 ```{r }
-## Python dependencies
-gym         <- import ("gym")
-gym_fishing <- import("gym_fishing")
-sb3         <- import ("stable_baselines3")
-
 ## initialize the environment
@@
 
-```{r eval=!file.exists("cache/td3.zip")}
-# train an agent (model) on one of the environments:
-
-#Trial 15 finished with value: 7.6994757652282715 and parameters: 
-hyper = list('gamma'= 0.995, 'lr'= 0.0001355522450968401, 'batch_size'= 128L, 
-         'buffer_size'= 10000L, 'episodic'= FALSE, 'train_freq'= 128L, 
-         'noise_type'= 'normal', 'noise_std'= 0.6656948079225263, 
-         'net_arch'= 'big')
-policy_kwargs = list(net_arch=c(400L, 300L)) # big
-
-#non-episodic:
-hyper['gradient_steps'] = hyper['train_freq']
-hyper['n_episodes_rollout'] = -1
-n_actions = env$action_space$shape[0]
-hyper$action_noise = sb3$common$noise$NormalActionNoise(
-        mean=np$zeros(n_actions),
-        sigma= hyper[['noise_std']] * np$ones(n_actions)
-        )
-
-td3 = sb3$TD3('MlpPolicy', env,  verbose=0L, seed = 42L,
-            gamma = hyper[['gamma']],
-            learning_rate = hyper[['lr']],
-            batch_size = hyper[['batch_size']],            
-            buffer_size = hyper[['buffer_size']],
-            action_noise = hyper[['action_noise']],
-            train_freq = hyper[['train_freq']],
-            gradient_steps = hyper[['train_freq']],
-            n_episodes_rollout = hyper[['n_episodes_rollout']],
-            policy_kwargs=policy_kwargs)
-
-td3$learn(total_timesteps=300000L)
-td3$save("cache/td3")
-```
-
-```{r}
-# Simulate management under the trained agent
-td3 <- sb3$TD3$load("cache/td3")
-td3_sims <- env$simulate(td3, reps = 500L)
-td3_policy <- env$policyfn(td3, reps = 50L)
-```
+We train the A2C algorithm using default hyper-parameter settings, based on the original publication [@A2C; @SB3].
+While default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achived by _tuning_ the hyperparameter selection.
 
@@
 a2c_policy <- env$policyfn(a2c, reps = 50L)
+```
+
+Hyperparameters are selected through the process of model tuning, see.
+
+```{r}
+# train an agent (model) on one of the environments:
+hyper = list('gamma'= 0.995, 'lr'= 0.0001355522450968401, 'batch_size'= 128L, 
+         'buffer_size'= 10000L, 'episodic'= FALSE, 'train_freq'= 128L, 
+         'noise_type'= 'normal', 'noise_std'= 0.6656948079225263, 
+         'net_arch'= 'big')
+policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
+
+#non-episodic:
+hyper['gradient_steps'] = hyper['train_freq']
+hyper['n_episodes_rollout'] = -1 # deprecated in latest SB3
+n_actions = env$action_space$shape[0]
+hyper$action_noise = sb3$common$noise$NormalActionNoise(
+        mean=np$zeros(n_actions),
+        sigma= hyper[['noise_std']] * np$ones(n_actions)
+        )
+
+td3 = sb3$TD3( 'MlpPolicy', 
+               env,  
+               verbose=0L, 
+               seed = 42L,
+               policy_kwargs=policy_kwargs,
+               learning_rate = hyper[['lr']],
+               gamma = hyper[['gamma']],
+               batch_size = hyper[['batch_size']],            
+               buffer_size = hyper[['buffer_size']],
+               action_noise = hyper[['action_noise']],
+               train_freq = hyper[['train_freq']],
+               gradient_steps = hyper[['train_freq']],
+               n_episodes_rollout = hyper[['n_episodes_rollout']]
+              )
+```
+
+```{r eval=!file.exists("cache/td3.zip")}
+td3$learn(total_timesteps=300000L)
+td3$save("cache/td3")
+```
+
+```{r}
+# Simulate management under the trained agent
+td3 <- sb3$TD3$load("cache/td3")
+td3_sims <- env$simulate(td3, reps = 500L)
+td3_policy <- env$policyfn(td3, reps = 50L)
 ```
@@
 ```{r}
-gym_conservation <- import("gym_conservation")
 env <- gym$make("conservation-v6")
@@
 # Simulate management under the trained agent
-# See https://github.com/boettiger-lab/conservation-agents/blob/4baaa6c9b994085ca56c42364f7affcc58b0ef3b/conservation/TD3.py
+# See https://github.com/boettiger-lab/conservation-agents/blob//conservation/TD3-v6.py
 

### 2021-04-15 manuscript/manuscript.Rmd
:pen:

--- 2021-02-04 manuscript/manuscript.Rmd
@@
 bibliography: references.bib
-header-includes:
-  - \usepackage{amsmath}
 output: 
@@
 
-<!-- no examples of issues / problems named!-->
+Deep Reinforcement Learning (RL), a machine learning technique in which a _software agent_ is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
+Unlike more widely recognized machine learning techniques for either _supervised_ learning (such as image classification, @species_ex, @satellite_ex) or unsupervised learning, RL focuses on solving _dynamic_ problems which require planning ahead, often in uncertain and changing environments.
+Of all machine learning techniques, RL most resembles a general artificial intelligence.
+Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero],
+and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex]. 
+These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.
 
-Despite a rapidly expanding interest in machine learning (ML) applications to ecology, the use of reinforcement learning (RL) has been largely neglected.
-To date, applications of machine learning to ecology have been predominantly limited to supervised learning [@examples], which attempts to solve problems in classification and regression, and unsupervised learning [@examples], which concentrates on data clustering.
-Reinforcement learning, meanwhile, is the unique subfield of machine learning that focuses on the problem of decision-making.
-The objective of reinforcement learning is to design a _software agent_ that learns by experience to select _actions_ to maximize some cumulative reward.
-RL is an incredibly general approach and has grown rapidly over the past 10 years as a consequence of the simultaneous rise of deep neural networks.
-In this short time, deep reinforcement learning has attained notable succeses such as AlphaZero, a single algorithm that can beat the best human players in go, chess, and shogi [@alphazero], as well as achieving dextrous manipulation in robotics [@robotics].
-However, reinforcement learning faces significant challenges  that limit its application to many real world problems[@henderson2017].
+<!-- Millie help on this para? -->
+If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
+The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
+The endangered species act ...
+Ecosystem services ....
+Such simplicity is in no small part due to the challenge associated in determining a 
+Some reference to @TomDietrich on potential for CS-based applciation.
 
-We believe RL approaches hold particular interest for conservation decision-making problems for two reasons.
-First, ecological systems are complex and difficult to represent accurately using models that remain tractable with classical approaches.
-For example, convenient mathematical assumptions such as that a dynamical system is _autonomous_ (parameters do not vary in time) are often required by classical techniques [@Marescot2013]; these assumptions stand in stark odds with the picture of an environment which is constantly changing.
-RL, conversely, has excelled in solving a range of complex decision-making problems that were previously intractable with classical techniques like dynamic programming.
-Second, RL approaches do not require massive amounts of data in stark contrast with other more widely used applications of ML.
-Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data often remains too sparse in the details that matter.
-And in a changing environment, no amount of historical data may be able to train supervised or unsupervised ML approaches to anticipate a future that has never been observed.
-However, RL algorithms do not need to depend on historical data alone, but rather, they can rely on computer simulations.
-Simulations are our most powerful approach to exploring future scenario, and simulated 'data' represents the largest contribution to the exponential growth of data in climate science [@].
-As we illustrate here, a software agent can be trained to successfully navigate hypothetical scenarios through its experience in simulation that would be difficult or impossible to solve with classical methods.
-In this manner, RL can compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
+In contrast to other ML methods, RL does not require massive data sets for training.  
+Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
+Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
+Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
+Methods from _optimal control theory_ have long been used in conservation and resource economics literature [@] to determine optimal policies given such models [@Polasky2011].
+However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
+Deep reinforcement learning offers a way forward. 
+Deep RL agents are trained not with mountains of data, but by interacting with simulators.
+Most recent advances in Deep RL have illustrated algorithms trained by playing arcade game simulators such as Atari BreakOut, or physics-based simulations of robotics movement [@TD3; @SAC; @PPO; @A2C].
+Simulating from complex models is far simpler than solving them.
+The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations.
+Given sufficient computational power, RL can be trained and evaluated across a broad range of possible realistic scenarios well beyond the reach of classical methods.
+In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
 
 If RL has great promise, the approach also imposes great risk.  
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions.
-Despite recent progress, Deep RL is a young field for which stability and robustness of its methods remain largely open problems.
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].  
+Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems. 
+RL approaches are no pancea, and inproper design or testing can easily lead to negative outcomes.
+Technical issues are not the only pitfall of AI applications to conservation problems.
 Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.
-We discuss these issues and potential ways forward.
+
+In this paper, we introduce the fundamentals of deep reinforcement learning and illustrate with simple examples how these approaches can be applied to important conservation decision-making problems, from overharvesting in global fisheries to ecosystem tipping points to wildfire dynamics.
+These examples span a range from problems for which an optimal analytic solution is already known, to problems not previously solved with classical methods, to problems not yet solved by RL. 
+We include a detailed appendix with complete examples in both python and R for training RL algorithms, as well as three stand-alone module "gyms" implementing the each of the conservation problems presented here using a standard benchmark [@gym], which we believe can facilitate future contributions both from researchers in RL searching for interesting and important open problems as well as ecologists who may seek to make those simulations more realistic and precise to specific application areas.
+We hope this paper serves as a port of entry to researchers and students from either field seeking to engage in work on RL applications to conservation. 
+We also hope these examples serve as illustration of both the potential and perils of such applications.
 
@@
 
-Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must decide what action to take to maximize some objective or cumulative "reward" in response to new information about the "state" of the system the decision-maker seeks to manage.
-Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether to declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
-In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region or how much each region would cost to purchase today are not sequential problems.
+<!-- meh, not sure we need this? -->
+
+Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to  as the reward or utility).
+Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
+In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
 Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
-Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast, explicitly or implicitly, how each possible action available to them will change that forecast.
+Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
 Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.
@@
 First, sequential decision-making problem is distinct from the problem of model estimation.
-A sequential decision problem will typically treat the model as given, perhaps with some uncertainty around the process, the measurements and model estimates as well.
+A sequential decision problem will typically treat the model as given, typically with some uncertainty around the process and possibly the measurmeents and model estimates as well.
 Thus, methods for solving sequential decision-making should not be mistaken as alternatives to methods designed for model estimation, such as regression, Bayesian heirarchical modeling, generalized additive models (GAMs) etc.
@@
 
-Before providing some examples on how deep reinforcement learning can be applied to ecological and conservation problems, we will provide a brief, general introduction to RL.
-The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
-This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
-The task for the agent is to learn a _policy_, $\pi$, that prescribes what action the agent should perform after observing a certain state.
-The agent interacts with the environment via trial and error in what is called the _training_ period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.
+**Model-based vs Model-free RL**
 
-When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem can be formalized as a Markov Decision Process (MDP).
-An MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
-From this definition, we can provide an agent's trajectory distribution for a given policy as,
-$$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
-$$
+**Current algorithms**
 
-When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
-An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
-This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
-
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected cumulative, discounted reward:
-$$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
-$$
-
-Note that this formalism extends to an infinite horizon case when $T= \infty$. While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
-
-Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high reward trajectory.
-There are 3 value functions that are commonly used across RL algorithms: the state value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
-The state-value function is defined as the expected return from a given state,
-$$
-  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-$$
-  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succintly defined as
-$$
-  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-$$
-Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
-Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
-
-**Survey of RL Algorithms**
-
-There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
-RL algorithms can generally be split into 2 groups, model-based and model-free RL.
-In model-based RL, the agent is attempting to learn the state transition operator, $T$.
-Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
-In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
-The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?].
-
-A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
-If an agent uses the same policy to take actions as the policy that the agent attempts to evaulate or improve, then this algorithm is on-policy.
-If an agent uses a different policy to select actions than the policy that the agent evaluates or improves, then this algorithm is off-policy.
-The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [cite SAC paper].
-Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop].
-
-The field of reinforcement learning has proliferated over the past 10 years on account of contributions from deep learning.
-In every reinforcement learning algorithm, the agent is attempting to learn some function, whether it is a policy function and/or a value function.
-Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions which we do not know the form of.
-Additionally, the ability of neural networks to autonomously devise low-dimensional representatoins of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
-The typical process for using neural networks in RL is that the agent is attempting to optimize a policy and/or value network using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
-Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable with classical methods.
-
-In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters.
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process.
-For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
-Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
-The hyperparameter tuning process consists of sampling then evaluating a set of hyperparameters, where tuning trials are often ranked by mean cumulative reward.
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
-
-```{=latex}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-Algorithm & Description & Model & Policy & Action Space & State Space \\
-\hline
-MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
-\hline
-MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
-\hline
-A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
-\hline
-A3C  & Asynchronous A2C & Model-free & On-policy & &\\
-\hline
-TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
-\hline
-PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
-\hline
-DQN & Deep Q Networks & Model-free & Off-policy& &\\
-\hline
-DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
-\hline
-TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
-\hline
-SAC & Soft Actor Critic & Model-free & Off-policy & &\\
-\hline
-IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
-\hline
-APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
-\hline
-\end{tabular}
-\end{center}
-```
+- terminology, problem definition
+- discussion / table of current benchmark algorithms (DQN, PPO, DDPG, A2C, TD3, SAC)
+- training, hyperparameter tuning
 
@@
 
-Deep reinforcement learning has only become an established subfield of AI over the past decade; and, as with most young fields, there are a range of unsolved problems.
-In practice, the foremost problem with DRL is instability.
-For a range of reasons, including high variance and the exploration vs.
-exploitation trade-off, model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
-For instance, it is not uncommon to observe vastly different behavior when two agents are initialized with different random seeds [cite ibid].
-Since model-based and model-free algorithms typically have a large number of hyperparameters, each of which can contribute to training instability, it is practically impossible to identify the source of instability.
-The hope of the RL practitioner is that after doing an exhaustive hyperparameter search, the practitioner will find a set of hyperparameters that results in a robust, well-performing agent, but there are no guarantees that hyperparameter tuning will lead to this outcome.
-
-In the context of conservation, we anticipate that there will be a host of challenges.
-Interpretability is likely to be the biggest barrier to adoption.
-Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of an RL-trained agent.
-Another challenge is that DRL algorithms are very sample inefficient, and to overcome sample inefficiency for conservation decision-making problems that are often data poor, relying on simulations to train agents will be necessary.
-While there are positives to using simulated data, any assumptions or biases that are included in the model will affect agent behavior.
-
-A problem that is central to AI safety is selecting the right objective. There have been many examples in which explicit reward design goes horribly wrong [cite]. ...
-
 - wild west of AI?
--can separate RL from having gyms implemented -> how do you compare different models, 
--mention that customization is probably the way to go, baselines only go so far
-
+- reproducibility
 - AI ethics, transparency

### 2021-04-15 manuscript/manuscript.Rmd

--- 2021-02-05 manuscript/manuscript.Rmd
@@
 
-<!-- no examples of issues / problems named!-->
+Deep Reinforcement Learning (RL), a machine learning technique in which a _software agent_ is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
+Unlike more widely recognized machine learning techniques for either _supervised_ learning (such as image classification, @species_ex, @satellite_ex) or unsupervised learning, RL focuses on solving _dynamic_ problems which require planning ahead, often in uncertain and changing environments.
+Of all machine learning techniques, RL most resembles a general artificial intelligence.
+Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero],
+and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex]. 
+These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.
 
-Despite a rapidly expanding interest in machine learning (ML) applications to ecology, the use of reinforcement learning (RL) has been largely neglected.
-To date, applications of machine learning to ecology have been predominantly limited to supervised learning [@examples], which attempts to solve problems in classification and regression, and unsupervised learning [@examples], which concentrates on data clustering.
-Reinforcement learning, meanwhile, is the unique subfield of machine learning that focuses on the problem of decision-making.
-The objective of reinforcement learning is to design a _software agent_ that learns by experience to select _actions_ to maximize some cumulative reward.
-RL is an incredibly general approach and has grown rapidly over the past 10 years as a consequence of the simultaneous rise of deep neural networks.
-In this short time, deep reinforcement learning has attained notable succeses such as AlphaZero, a single algorithm that can beat the best human players in go, chess, and shogi [@alphazero], as well as achieving dextrous manipulation in robotics [@robotics].
-However, reinforcement learning faces significant challenges  that limit its application to many real world problems[@henderson2017].
+<!-- Millie help on this para? -->
+If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
+The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
+The endangered species act ...
+Ecosystem services ....
+Such simplicity is in no small part due to the challenge associated in determining a 
+Some reference to @TomDietrich on potential for CS-based applciation.
 
-We believe RL approaches hold particular interest for conservation decision-making problems for two reasons.
-First, ecological systems are complex and difficult to represent accurately using models that remain tractable with classical approaches.
-For example, convenient mathematical assumptions such as that a dynamical system is _autonomous_ (parameters do not vary in time) are often required by classical techniques [@Marescot2013]; these assumptions stand in stark odds with the picture of an environment which is constantly changing.
-RL, conversely, has excelled in solving a range of complex decision-making problems that were previously intractable with classical techniques like dynamic programming.
-Second, RL approaches do not require massive amounts of data in stark contrast with other more widely used applications of ML.
-Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data often remains too sparse in the details that matter.
-And in a changing environment, no amount of historical data may be able to train supervised or unsupervised ML approaches to anticipate a future that has never been observed.
-However, RL algorithms do not need to depend on historical data alone, but rather, they can rely on computer simulations.
-Simulations are our most powerful approach to exploring future scenario, and simulated 'data' represents the largest contribution to the exponential growth of data in climate science [@].
-As we illustrate here, a software agent can be trained to successfully navigate hypothetical scenarios through its experience in simulation that would be difficult or impossible to solve with classical methods.
-In this manner, RL can compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
+In contrast to other ML methods, RL does not require massive data sets for training.  
+Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
+Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
+Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
+Methods from _optimal control theory_ have long been used in conservation and resource economics literature [@] to determine optimal policies given such models [@Polasky2011].
+However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
+Deep reinforcement learning offers a way forward. 
+Deep RL agents are trained not with mountains of data, but by interacting with simulators.
+Most recent advances in Deep RL have illustrated algorithms trained by playing arcade game simulators such as Atari BreakOut, or physics-based simulations of robotics movement [@TD3; @SAC; @PPO; @A2C].
+Simulating from complex models is far simpler than solving them.
+The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations.
+Given sufficient computational power, RL can be trained and evaluated across a broad range of possible realistic scenarios well beyond the reach of classical methods.
+In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
 
 If RL has great promise, the approach also imposes great risk.  
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions.
-Despite recent progress, Deep RL is a young field for which stability and robustness of its methods remain largely open problems.
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].  
+Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems. 
+RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes.
+Technical issues are not the only pitfall of AI applications to conservation problems.
 Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.
-We discuss these issues and potential ways forward.
+
+In this paper, we introduce the fundamentals of deep reinforcement learning and illustrate with simple examples how these approaches can be applied to important conservation decision-making problems, from over-harvesting in global fisheries to ecosystem tipping points to wildfire dynamics.
+These examples span a range from problems for which an optimal analytic solution is already known, to problems not previously solved with classical methods, to problems not yet solved by RL. 
+We include a detailed appendix with complete examples in both python and R for training RL algorithms, as well as three stand-alone module "gyms" implementing the each of the conservation problems presented here using a standard benchmark [@gym], which we believe can facilitate future contributions both from researchers in RL searching for interesting and important open problems as well as ecologists who may seek to make those simulations more realistic and precise to specific application areas.
+We hope this paper serves as a port of entry to researchers and students from either field seeking to engage in work on RL applications to conservation. 
+We also hope these examples serve as illustration of both the potential and perils of such applications.
 
@@
 
-Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must decide what action to take to maximize some objective or cumulative "reward" in response to new information about the "state" of the system the decision-maker seeks to manage.
-Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether to declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
-In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region or how much each region would cost to purchase today are not sequential problems.
+<!-- meh, not sure we need this? -->
+Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to  as the reward or utility).
+Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
+In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
 Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
-Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast, explicitly or implicitly, how each possible action available to them will change that forecast.
+Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
 Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.

### 2021-04-15 manuscript/appendix.Rmd
:pen:

@@
 
+# Deep Reinforcement Learning Frameworks
+
+At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
+
+# Deep Reinforcement Learning in R
+
+All though all the necessary tooling for RL is implemented in python, the R language is more familiar to most ecologists.
+Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straight forward to use these tools without ever leaving the R interface.
+In this appendix, we detail this "pure R" approach, as well as a "pure python" approach.
+
+In the R based-approach, R functions take responsibility from the user for translating commands into python code before it is executed, an approach commonly referred to as meta-programming.
+This still requires a local python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
+Alternately, users may prefer running the analysis inside a docker container.  
+The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
+This can be particularly useful for users running ML algorithms on remote servers, as configuring R, python, and CUDA environments appropriately can be challenging.
+
+Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
+From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and python modules.
+
+```{r}
+#install.packages("renv")
+renv::restore()
+```
+
+Once the packages have installed, we are ready to load the necessary libraries.
+Note that the `import` function from `reticulate` package acts much like the `library` command in R, though it does not attach the package function to the global namespace.
+To make it more convenient to access those functions, we can assign a shorthand name.
+
 ```{r setup, cache = FALSE, message=FALSE}
@@
 gym_conservation <- import("gym_conservation")
+```
 
+Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration.
+In addition to setting a random seed in our python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.
+
+```{r}
 ## reproducible settings
 np <- import("numpy")
-np$random$seed(42L)
+seed <- 24L # integer
+np$random$seed(seed)
 # Set to "" to force CPU-evaluation if needing perfect reproducibility
@@
 
+We begin by select an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
+
 ```{r }
@@
 
-We train the A2C algorithm using default hyper-parameter settings, based on the original publication [@A2C; @SB3].
-While default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achived by _tuning_ the hyperparameter selection.
+At the heart of each algorithm is a neural network. The type, size and depth of network, choice of activation functions and so forth, are considered "hyper-parameters" of algorithm: 
+we must first make a selection for each of these hyper-parameters to define our (untrained) agent.  
+Other hyper-parameters determine other aspects of the training regime: such as how the agent chooses to explore possible actions, the learning rate, or how future rewards are discounted (gamma).
+Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms,
+for instance, the A2C algorithm originally described in @A2C. 
+Here, we train an agent using the A2C algorithm using default hyperparameter settings.
 
-```{r eval=!file.exists("cache/a2c.zip")}
-# train an agent (model) on one of the environments:
-a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = 42L) # Must use L for integers!
+```{r}
+a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
+```
+
+Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment and the number of training iterations budgeted.
+Therefore, we save the trained agent to disk, and only execute the training method (`learn`) if no previously saved agent is found:
+
+```{r}
+if(!file.exists("cache/a2c.zip")) {
+
+  # Train the agent for a fixed number of timesteps
   a2c$learn(total_timesteps=300000L)
+
+  # Save our trained agent for future use
   a2c$save("cache/a2c")
+}
 ```
+
+Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyperparameter selection.
+This is discussed further below.
+
+We can load a trained agent and use `simulate` and `policyfn` methods provided by the `gym_fishing` environment to simulate the process for a fixed number of replicates.
 
@@
 a2c <- sb3$A2C$load("cache/a2c")
+```
+
+```{r}
+
+## represent the initial state size in the 'rescaled' state space.
+state <- env$get_state( 0.75 )
+
+for(i in 1:100){
+
+  out <- a2c$predict(state)
+  action <- out[[1]]
+  result <- env$step(action)
+  state <- result[[1]]
+
+}
+
+```
+
+```{r}
 a2c_sims <- env$simulate(a2c, reps = 500L)
+```
+
+```{r}
 a2c_policy <- env$policyfn(a2c, reps = 50L)
@@
 
-Hyperparameters are selected through the process of model tuning, see.
+## TD3
+
+We train a second agent based on the TD3 algorithm.  
+This time, we will declare our choices for the hyper-parameters explicitly.
+Note that some of these hyperparameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
+Good choices for hyperparameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance.
+For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
 
@@
 # train an agent (model) on one of the environments:
-hyper = list('gamma'= 0.995, 'lr'= 0.0001355522450968401, 'batch_size'= 128L, 
-         'buffer_size'= 10000L, 'episodic'= FALSE, 'train_freq'= 128L, 
-         'noise_type'= 'normal', 'noise_std'= 0.6656948079225263, 
-         'net_arch'= 'big')
+
 policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
-
-#non-episodic:
-hyper['gradient_steps'] = hyper['train_freq']
-hyper['n_episodes_rollout'] = -1 # deprecated in latest SB3
+# non-episodic action noise:
+noise_std = 0.6656948079225263
 n_actions = env$action_space$shape[0]
-hyper$action_noise = sb3$common$noise$NormalActionNoise(
+action_noise = sb3$common$noise$NormalActionNoise(
         mean=np$zeros(n_actions),
-        sigma= hyper[['noise_std']] * np$ones(n_actions)
+        sigma= noise_std * np$ones(n_actions)
         )
@@
                verbose=0L, 
-               seed = 42L,
-               policy_kwargs=policy_kwargs,
-               learning_rate = hyper[['lr']],
-               gamma = hyper[['gamma']],
-               batch_size = hyper[['batch_size']],            
-               buffer_size = hyper[['buffer_size']],
-               action_noise = hyper[['action_noise']],
-               train_freq = hyper[['train_freq']],
-               gradient_steps = hyper[['train_freq']],
-               n_episodes_rollout = hyper[['n_episodes_rollout']]
+               seed = seed,
+               policy_kwargs = policy_kwargs,
+               learning_rate = 0.0001355522450968401,
+               gamma= 0.995, 
+               batch_size = 128L,
+               buffer_size = 10000L,
+               train_freq = 128L, 
+               n_episodes_rollout = -1, 
+               gradient_steps = 128L,
+               action_noise = action_noise
               )
@@
 
-```{r eval=!file.exists("cache/td3.zip")}
+Once again, after our agent is defined we are ready to train it.
+
+```{r }
+if(!file.exists("cache/td3.zip")){
+
   td3$learn(total_timesteps=300000L)
   td3$save("cache/td3")
+
+}
 ```
@@
 env <- gym$make("conservation-v6")
+```
+
+```{r}
+
+noise_std = 0.4805935357322933
+action_noise = OrnsteinUhlenbeckActionNoise(mean = np.zeros(1),
+                                            sigma = noise_std * np.ones(1))
+
+model = TD3('MlpPolicy', 
+            env, 
+            verbose = 0, 
+            seed = seed,
+            "gamma": 0.995,
+            "learning_rate":  8.315382409902049e-05,
+            "batch_size": 512,
+            "buffer_size": 10000,
+            "train_freq": 1000,
+            "gradient_steps": 1000,
+            "n_episodes_rollout": -1,
+            "action_noise": action_noise,
+            "policy_kwargs": list("net_arch": c(64,64)))
+
+```
+
+```{r}
+model$learn(total_timesteps = 1500000)
+
 ```

### 2021-04-15 manuscript/manuscript.Rmd
:pen:

@@
 
+<!--
+
+Conceptual figure:
+
+Algorithm 
+  hyper-parameters specifying neural network, learning rate, etc.  
+  We refer to a particular selection of algorithm and hyperparameters as an "agent", which is then "trained" to solve one or more environments
+  internal 'parameters', estimated by "training" the agent.
+
+-->
+
 There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
@@
 
+We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.
+
 ## Sustainable harvest
 
-We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.
-The first example focuses on the well-studied problem of setting harvest quotas in fisheries management [@Schaefer1954; @Clark1974; @Reed1979; @Worm2006; @Costello2018].
-Given an appropriately estimated model, the optimal harvest policy can be determined precisely (e.g.
-by stochastic dynamic programming, SDP, see @Marescot2013), and under most circumstances takes the form of a "constant-escapement" policy [see mathematical proof in @Reed1979].
+The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
+Determining fishing quotas is both a critical ecological issue [@Worm2005; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
+Global fisheries represent an essential source of nutrition and natural diversity worldwide with [@Worm2006; @Costello2018]
+Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
+When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].
+
 This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
@@
 
-For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 or @Reed1979 (covering the stochastic case).
+For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
 In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
@@
 From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."  The simplest simulation is merely an implementation of the estimated model itself.
+In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."
+The simplest simulation is merely an implementation of the estimated model itself.
 However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach is in principle agnostic to the details of the simulation.
@@
 
-## Beyond classic methods: Non-stationary models
+## Ecological Tipping Points
 

### 2021-04-15 manuscript/manuscript.Rmd
:pen:

@@
 Determining fishing quotas is both a critical ecological issue [@Worm2005; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
-Global fisheries represent an essential source of nutrition and natural diversity worldwide with [@Worm2006; @Costello2018]
 Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].

### 2021-04-20 manuscript/appendix.Rmd
td3 hyperparameters for cons-v6 (testing :construction:)

@@
 noise_std = 0.4805935357322933
-action_noise = OrnsteinUhlenbeckActionNoise(mean = np.zeros(1),
-                                            sigma = noise_std * np.ones(1))
+action_noise = sb3$common$noise$OrnsteinUhlenbeckActionNoise(mean = np$zeros(1L),
+                                            sigma = noise_std * np$ones(1L))
 
-model = TD3('MlpPolicy', 
+model = sb3$TD3('MlpPolicy', 
             env, 
@@
             seed = seed,
-            "gamma": 0.995,
-            "learning_rate":  8.315382409902049e-05,
-            "batch_size": 512,
-            "buffer_size": 10000,
-            "train_freq": 1000,
-            "gradient_steps": 1000,
-            "n_episodes_rollout": -1,
-            "action_noise": action_noise,
-            "policy_kwargs": list("net_arch": c(64,64)))
+            "gamma"= 0.995,
+            "learning_rate"=  8.315382409902049e-05,
+            "batch_size"= 512L,
+            "buffer_size"= 10000L,
+            "train_freq"= 1000L,
+            "gradient_steps"= 1000L,
+            "n_episodes_rollout"= -1L,
+            "action_noise"= action_noise,
+            "policy_kwargs"= list("net_arch"= c(64L,64L)))
 
@@
 ```{r}
-model$learn(total_timesteps = 1500000)
+if(!file.exists("cache/td3-conservation.zip")){
+
+  model$learn(total_timesteps=1500000L)
+  model$save("cache/td3-conservation")
+
+}
 

### 2021-04-20 manuscript/appendix.Rmd
:pen:

@@
 for instance, the A2C algorithm originally described in @A2C. 
-Here, we train an agent using the A2C algorithm using default hyperparameter settings.
+Here, we train an agent using the A2C algorithm using default hyperparameter settings, utilizing a multi-layer perceptron neural network composed of two 64-neuron layers.
 
@@
 This is discussed further below.
-
-We can load a trained agent and use `simulate` and `policyfn` methods provided by the `gym_fishing` environment to simulate the process for a fixed number of replicates.
+Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.
+Note that a copy of the trained agents are included in the corresponding GitHub repository.
 
@@
 
+We can supply an observation to our trained agent and it will return it's proposed action using the `predict` method. 
+This is all we need to evaluate or employ the agent on a decision-making task.
+Recall that state space and action space in the fishing gym have been re-scaled to a (-1, 1) interval.
+Note that this is equivalent to a choice of appropriate units -- we can re-scale the interval without loss of generality.
+This is often a critical step in the design of an RL environment to facilitate successful training.
+Following the `gym` standard, the core methods such as `predict` and `step` operate on the re-scaled units,
+so it is necessary to first transform the original units into this re-scaled state space.
+For example, if we wish to start a simulation with a stock size of 0.75, we can use the helper
+method `get_state()`, to determine the corresponding value in the re-scaled state space.  
+Unlike `predict` and `step`, `get_state()` is not a standard method of all gym environments -- typically
+a user must first inspect the state space of an environment and choose themselves how to re-scale their
+problem into that state space.
+
 ```{r}
@@
 state <- env$get_state( 0.75 )
+state
+```
 
-for(i in 1:100){
+With an initial state in hand, we are ready to simulate management using our agent.
+The iteration is simple: we use the agent to predict what action we should take given the current state,
+and then we take said action and examine the result to determine the future state. 
+Because these methods return additional information as well, a little extra sub-setting is required in R:
+
+```{r}
+for(i in 1:10){
 
@@
 
+For convenience, `gym_fishing` defines the helper routine `simulate` to perform the above iteration `reps` number of times.
+The `simulate` method returns the state, action, and reward resulting from each time step of each replicate:
+
 ```{r}
@@
 ```
+
+Because the agent typically attempts to retain the fish population near a specific value, simulations with well-trained agents will usually not explore the full range of possible states.
+To get a better idea of how the agent behaves across the full state space, it common to consider a policy function: a plot of the action taken at each possible state.
+Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
+Note that the `policyfn` method is another custom method and not part of the `gym` standard, and the use of a policy function will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTM [@LSTM].
 
@@
 We train a second agent based on the TD3 algorithm. 
+This illustrates some of the salient differences between the different algorithms used in reinfocement learning.
+Bear in mind that the best algorithm for any given environment will depend on the details of the environment, as well
+as the length of training and the selection of hyper-parameters, including the agent's underlying neural networks which provide the 'brains' of the agent's decision-making ability.
 This time, we will declare our choices for the hyper-parameters explicitly.
-Note that some of these hyperparameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
-Good choices for hyperparameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance.
+Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
+Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance.
 For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
@@
 
-Once again, after our agent is defined we are ready to train it.
+Once again, after our agent is defined we are ready to train it:
 
@@
 
+With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:
+
 ```{r}
@@
 
+We compare the performance of these two RL agents trained using the algorithms A2C and TD3 respectively to the known optimal policy for this particular environment.  
+Recall that under the assumptions of the simple model used in the `fishing-v1` environment, we can determine the optimal harvest policy analytically if the model and parameters are known precisely [@Reed1979].
+The optimal strategy is a policy of 'constant escapement', designed to keep the remaining stock size (the population that 'escapes' fishing harvest) at the biomass corresponding to a maximum growth rate, i.e.
+at $B = K/2$ in this model.
+`gym_fishing` defines a collection of non-RL agents in the `models` submodule, including the a human agent that merely asks to enter their desired quota manually. 
+The `escapement` model implements the provably optimal constant escapement rule.
+A third model, `msy`, implements a policy based on "Maximum Sustainable Yield" policy [@Schaefer1954], which is actually more commonly used as a basis for management than constant escapement, despite only being optimal at the steady state under deterministic dynamics.
+
 ```{r}
@@
 ```
+
+Having performed simulations under each method, we gather the results under each management strategy together into a single data frame for simulations (`sims_df`), the policy function (`policy_df`) and the cumulative reward, used to make the plots in the corresponding panels.
+We preserve a copy of each data frames in local `.csv` text files for maximum cross-platform compatibility.
 
@@
 
-```
-
-```{r}
 write_csv(sims_df, "figs/sims_df.csv")
@@
 ```
+
+Standard R methods are used to plot the results summarized across the replicates:
 
@@
 ## Ecological tipping points
+
+Our second example utilizes our `gym_conservation` to provide the necessary environment to simulate an ecosystem approaching a tipping point.
+For simplicity we will train only a TD3-based agent this time.
 
@@
 ```
+
+In general, the optimal solution depends on the ecological dynamics, the benefit of the ecosystem services and the costs associated with a management response.
+Because the tipping point problem is non-autonomous, we cannot solve for the optimal policy even given the model and objective (utility) function using Markov Decision Process methods.
+However, a simple heuristic solution provides a reasonable starting point for comparison.
+
+As discussed in the main text, we have no existing optimal solution to the tipping point problem, and so rely on a common heuristic strategy instead:
+select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in it's current state.
 

### 2021-04-20 manuscript/manuscript.Rmd
:pen:

@@
 From this definition, we can provide an agent's trajectory distribution for a given policy as,
+
 $$
@@
 The state-value function is defined as the expected return from a given state,
+
 $$
@@
 
+```{r knitr, include = FALSE}
+library(tidyverse)
+library(patchwork)
+knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
+                      fig.width = 7, fig.height = 4)
+ggplot2::theme_set(ggplot2::theme_bw())
+
+scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
+pal <- ggthemes::solarized_pal()(6)
+txtcolor <- "#586e75"
+```
+
 ## Sustainable harvest
@@
 The simplest simulation is merely an implementation of the estimated model itself.
-However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach is in principle agnostic to the details of the simulation.
+However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-```{r knitr, include = FALSE}
-library(tidyverse)
-library(patchwork)
-knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4)
-ggplot2::theme_set(ggplot2::theme_bw())
+To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on @TD3.
+A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
+We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 1].
+Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
 
-scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(6)
-txtcolor <- "#586e75"
-```
-
-```{r}
+```{r message=FALSE, include=FALSE}
 sims_df <- read_csv( "figs/sims_df.csv")
@@
 
+Our second example focuses on a case for which we do not already have an existing, provably optimal policy to compare against.
+We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
+This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of @May1977 on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
+On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
+For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $x(t)$.
+Thus when the ecosystem is near the 'high' equilibrium $\hat x_H$, the corresponding ecosystem benefit $b \hat x_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barknosky2014].
+We also enumerate the possible actions which a manager may take in response to environmental degradation. 
+In the absence of any management response, we assume the environment deteriorates at a fixed rate $\alpha$, which can be thought of as the incremental increase in global mean temperature or similar anthropogenic forcing term.
+Management can respond to slow or even reverse this trend by choosing an opposing action $a_t$.
+We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $a_t$ as equal to $c a_t^2$.
+Many alterations of these basic assumptions are also possible: our `gym_conservation` implements a range of different scenarios with user-configurable settings.
+In each case, the manager observes the current state of the system each year and must then select the policy response that year.
+
+Because this problem involves a parameter whose value changes over time (the slowly deteriorating environment), the resulting ecosystem dynamics are not autonomous.
+This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, @Marescot2013), typically used to solve sequential decision-making problems.
+However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham; @Possingham; @morePossingham].
+A common conservation strategy employs a fixed response level rather than a dynamic policy which is toggled up or down each year: for example, declaring certain regions as protected areas in perpetuity.
+An intuitive strategy faced with an ecosystem tipping point would be 'perfect conservation', in which the management response is perfectly calibrated to counter-balance any further decline.
+While the precise rate of such decline may not be known in practice (and will not be known to RL algorithms before-hand either), it is easy to implement such a policy in simulation for comparative purposes.
+We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm:
+
 ```{r}
@@
 
-## Example of solving a problem that is difficult/impossible for conventional methods
+The TD3-trained neural network agent proves far more successful in preventing chance transitions across the tipping point, consistently achieving a higher cumulative ecosystem service value across replicates than the steady-state strategy.
 
-two use cases to consider: 
-  - model is known but too complex to solve by (approximate) dynamic programming etc, or non-markovian
-  - model is unknown
+Examining the replicate management trajectories and corresponding rewards reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
+This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.  
+The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
+Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
+For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
 
-model uncertainty/transfer learning? nonstationary dynamics?
+A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix, fig S1, S2) further highlights the difference between these two approaches.
+Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
+As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards. 
+In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
+In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery (Fig S2).
+Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
+
+Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
+The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
+Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
+
+In particular, the use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
+A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
 

### 2021-04-21 manuscript/appendix.Rmd
:pen:

@@
 np$random$seed(seed)
-# Set to "" to force CPU-evaluation if needing perfect reproducibility
-Sys.setenv("CUDA_VISIBLE_DEVICES"="")
+# Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
+# Sys.setenv("CUDA_VISIBLE_DEVICES"="")
 
@@
 
-```{r fig1, fig.cap=""}
+```{r fig1, fig.cap="", include=FALSE}
 fig_sims / ( fig_policy + fig_reward)
@@
 
-## Ecological tipping points
+# Ecological tipping points
 
 Our second example utilizes our `gym_conservation` to provide the necessary environment to simulate an ecosystem approaching a tipping point.
-For simplicity we will train only a TD3-based agent this time.
+
+## Tipping point model
+
+The tipping point model is based on the consumer-resource model of @May1977, which creates alternative stable states, 
+which we subject to log-normal environmental noise:
+
+\begin{equation}
+\mu_{t} = X_t + X_t r \left(1 - \frac{X_t}{K}) - \frac{a_t X_t^q}{X_t^q + b^q} 
+X_{t+1} \sim \lognormal(\mut_t, \sigma)
+\end{equation}
+
+where we take $r=0.7$, $K=1.2$, $q=3$, $b=0.15$, $a_0 = 0.19$ and $\sigma = 0.2$ 
+Slow change over time in the parameter $a_t$ represents a process of environmental degradation, modeled as a constant increment $a_{t+1} = a_t + \alpha$, where we will take $\alpha = 0.001$.
+This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carptenter2011] to the planet biosphere [@Barnosky2014].
+
+We assume the benefit provided by the ecosystem state is assumed to be directly proportional to the state itself, $b x_t$.  
+We further assume that each year the manager has the option to slow or reverse the environmental degradation by taking action $A_t$, such that under management, the resulting environment in the next time step is given by
+
+\beign{equation}
+a_{t+1} = a_t + \alpha - A_t
+\end{equation}
+
+We assume the cost associated with that action to be proportional to the square of the action, such that large actions are proportionally more costly than small ones.
+Consequently, the utility at time $t$ is given by the sum of costs and benefits:
+
+$$U(X_t, A_t) = b X_t - c A_t ^2$$
+where we will take $b = 1$ and $c=1$.  
+Our implementation in `gym_conservation` allows the user to consider alternative parameter choices, alternative models for the ecological dynamics, and alternate types of actions, such as manipulating the ecosystem state directly rather than manipulating the environmental parameter.
+For some such scenarios the optimal solution is known or can be determined by stochastic dynamic programming, while for others, including the scenario of focus here, the optimal solution is unknown.
+
+For simplicity of illustration, we will train only a TD3-based agent on this scenario.
 
@@
 
-  model$learn(total_timesteps=1500000L)
+  model$learn(total_timesteps=3000000L)
   model$save("cache/td3-conservation")
@@
 
-TD3 <- sb3$TD3$load("cache/td3-conservation")
-TD3_sims <- env$simulate(TD3, reps = 100L)
-TD3_policy <- env$policyfn(TD3, reps = 10L)
+model <- sb3$TD3$load("cache/td3-conservation")
+TD3_sims <- env$simulate(model, reps = 100L)
+TD3_policy <- env$policyfn(model, reps = 10L)
 ```

### 2021-04-21 manuscript/cover-letter.Rmd
:pen:

---
output: pdf_document
---

Dear John Drake,

We are pleased to present a proposal for an article discussing the potential use of deep Reinforcement Learning in theoretical and applied ecological conservation.

Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet's most pressing environmental problems, such as the growing frequency and intensity of wildfire, over-exploited fisheries, declining biodiversity, and zoonotic pandemics. 
Will such approaches really help us tackle a changing planet? 
The reliance of dominant machine learning approaches on existing data can make them a valuable tool for automating well-understood steps in an analysis, but may also make them a poor tool to predict the long-term dynamics of a no-analogue future and the non-representative distribution of available ecological data. 
Supervised Learning methods have proven very successful in tasks such as image classification, with the promise of being able to automate species identification from cameras, acoustic sensors or satellite images. 
Unsupervised learning methods hold out a more ambitious promise of identifying features or making predictions without first requiring a large training set labeled data. 
The least attention has been paid to what may be the most promising member of the ML triumvirate: Reinforcement Learning (RL). 
RL trains a software agent to maximize some objective -- notable examples include navigation in autonomous robots or Google's AlphaZero. 
RL requires agents to make decisions in dynamic and uncertain environments that plan ahead and change the state of the board. Also unlike other approaches, such agents are typically trained, not on mountains of data alone, but on simulations.

We describe how this creates a bridge between the rich, mechanistic or process based models of our field and the power of ML to discover creative solutions to complex decision problems. 
Rather than seeking to displace process-based models of the past century with opaque machines, we illustrate how we can use such mechanistic models to drive realistic simulations which we can then train leading RL algorithms to manage. 
We demonstrate this approach using examples from fisheries management, ecosystem tipping points, and wildfire spread. 
All though our examples are primarily a proof-of-principle, they serve to illuminate both the promise, as well as potential pitfalls 
Those pitfalls are not only technical, but include issues of ethics and power, particularly if the algorithms or data are proprietary. 
We conclude with a discussion of how an open, transparent and reproducible approach can help mitigate some concerns, while also offering a more effective interface between teams of researchers from both ecological and computer sciences.

We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
We further include implementations of three fully featured python modules following current leading standards, which would allow engineers and computer scientists already working in RL to test their agents against both solved and unsolved problems in conservation management.
We believe this piece would provide an effective introduction to the concepts and practices of RL which would allow ecologists to apply, extend, and critically critique such methods.

### 2021-04-21 manuscript/manuscript.Rmd
:pen:

@@
   - \usepackage{amsmath}
+  - \usepackage{amssymb}
+
 output: 
@@
 Such simplicity is in no small part due to the challenge associated in determining a 
-Some reference to @TomDietrich on potential for CS-based applciation.
+Some reference to @TomDietrich on potential for CS-based application.
 
@@
 Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
-Methods from _optimal control theory_ have long been used in conservation and resource economics literature [@] to determine optimal policies given such models [@Polasky2011].
+Methods from _optimal control theory_ have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
 However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
@@
 
-$$
+\[
   p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
-$$
+\]
 
@@
 In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected cumulative, discounted reward:
-$$
+
+\[
   \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
-$$
+\]
 
@@
 
-$$
+\[
   V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
+\]
 
 Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-$$
+\[
   Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
+\]
 
@@
 The advantage function can then be succintly defined as
-$$
+\[
   A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-$$
+\]
 Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
@@
 The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
-Determining fishing quotas is both a critical ecological issue [@Worm2005; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
+Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
 Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
@@
 
-```{r fig1, fig.cap=""}
+```{r fig1, fig.cap="Fisheries management using neural network agents trained with RL algorithms A2C and TD3 compared to optimal management.
+Top panel: mean fish population size over time across 100 replicates.
+Shaded region shows the 95% confidence interval over simulations.
+Lower left: The optimal solution is policy of constant escapement.
+Below the target escapement of 0.5, no harvest occurs, while any stock above that level is immediately harvested back down (red line).
+RL agents both adopt a policies that cease any harvest below this level, while allowing a somewhat higher escapement than optimal just above the optimal escapement level.
+A2C policy is non-deterministic, and overharvest from very high stock sizes.
+Lower Right: Mean cumulative reward over time across simulations.
+TD3 achieves nearly-optimal mean utility.
+Mean reward under A2C is lower, though the confidence interval still overlaps the mean reward of optimal management."}
 fig_sims / ( fig_policy + fig_reward)
@@
 
-Note that the A2C policy learns to shut down fishing entirely below a certain stock size, without any variation.
-However, it keeps the fishery closed even once stock sizes are slightly higher than $B_{MSY}$.
+Note that the A2C policy learns to shut down fishing entirely below a certain stock size, without fail.
+A2C also keeps the fishery closed even once stock sizes are slightly higher than $B_{MSY}$.
 Once it begins fishing, it slightly under-harvests for stock sizes a little over $B_{MSY}$, but tends to over-harvest for very large stock sizes (rarely ever realized, as fishing pressure is enough to keep it out of this range).
@@
 Under such a policy, A2C maintains a stock size on average significantly above $B_{MSY}$, though also sees higher volatility particularly on the low side.
-Note that due to the stochastic nature of the environment, any given replicate A2C may out-perform the optimal policy, though on average the optimal policy provides consistently better yields.
+Note that due to the stochastic nature of the environment, any given replicate A2C may out-perform the optimal policy, though on average the optimal policy provides a higher reward.
+
+These features highlight a common challenge in the design and training of RL algorithms.
+RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
+Because these simulations rarely take stock sizes close to a state of 1.0, the heavy overfishing behavior A2C has in this range has little consequence.
+Likewise it is more difficult for TD3 to learn the appropriate harvest level at these high states that are rarely sampled in the course of training.
+Just as intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
+Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
 
@@
 On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
-For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $x(t)$.
-Thus when the ecosystem is near the 'high' equilibrium $\hat x_H$, the corresponding ecosystem benefit $b \hat x_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barknosky2014].
+For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
+Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2014].
 We also enumerate the possible actions which a manager may take in response to environmental degradation. 
 In the absence of any management response, we assume the environment deteriorates at a fixed rate $\alpha$, which can be thought of as the incremental increase in global mean temperature or similar anthropogenic forcing term.
-Management can respond to slow or even reverse this trend by choosing an opposing action $a_t$.
-We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $a_t$ as equal to $c a_t^2$.
+Management can respond to slow or even reverse this trend by choosing an opposing action $A_t$.
+We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $A_t$ as equal to $c A_t^2$.
 Many alterations of these basic assumptions are also possible: our `gym_conservation` implements a range of different scenarios with user-configurable settings.
 In each case, the manager observes the current state of the system each year and must then select the policy response that year.
+
+```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario.
+The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g.
+global mean temperature, arbitrary units) of 0.19.
+In the absence of conservation action, the environment worsens (e.g.
+rising mean temperature) as the parameter increases.
+This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.
+Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
+bifur_df <- read_csv("figs/bifur.csv", col_types = "dccd")
+bifur_df %>%
+  ggplot(aes(parameter, state, lty=equilibrium, group = group)) + 
+  geom_line()
+
+```
 
@@
 While the precise rate of such decline may not be known in practice (and will not be known to RL algorithms before-hand either), it is easy to implement such a policy in simulation for comparative purposes.
-We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm:
+We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm.
 
@@
 
-```{r}
+```{r fig.cap="Ecosystem dynamics under management using the steady-state rule-of-thumb strategy compared to management using a neural network trained using the TD3 RL algorithm.
+Top panel: mean and 95% confidence interval of ecosystem state over 100 replicate simulations.
+As more replicates cross the tipping point threshold under steady-state strategy, the mean slowly decreases, while the TD3 agent preserves most replicates safely above the tipping point.
+Lower left: the policy function learned using TD3 relative to the policy function under the steady state.
+Lower right: mean rewards under TD3 management evenutally exceed those expected under the steady state strategy as a large initial investment in conservation eventually pays off."}
 fig_sims <- 
@@
 
-Examining the replicate management trajectories and corresponding rewards reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
+Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
 This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.  
@@
 For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
+Recall that the quadratic scaling of cost makes the rapid response of the TD3 algorithm much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
+However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.
 
@@
 Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
-
-In particular, the use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
+The use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
 A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
 
-## Example of an open problem without an RL solution
+# Discussion
 
-fire gym?
+<!--
+
+1. Return to theme of RL vs supervised/unsupervised learning.
+
+-->
+
+<!-- Something about relevance to both the _theory_ and practice of ecological conservation.
+(e.g.
+"strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->
+
+## Open Problems
+
+The examples presented here only scrape the surface of possible RL applicaations to conservation problems.
+The examples we have focused on are intentionally quite simple, though it is worth remembering that these very same simple models have a long history of relevance and application in both research and policy contexts.
+Despite their simplicity, the optimal strategy is not always obvious before hand, however intutive it may appear in retrospect.
+In the case of the ecosystem tipping point scenario, the optimal strategy is unknown, and the approximate solution found by our RL implementation could almost certainly be improved upon.
+
+Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained. 
+In these simple examples in which the simulation implements a single model, training is analgous to classical methods which take the model as given [@Marescot2013].
+But classical approaches can be difficult to generalize when the underlying model is unknown: alternative models may require a completely different approach.
+In contrast, the process of training an RL algorithm on a more complex problem is no different than the simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
+Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @firemodel], disease [@disease], marine ecosystems [@ecopath], or global climate change [@gcm].
+
+# Discussion
 

### 2021-04-21 R/bifurcation-helper.R

@@ -1,0 +1,59 @@
+library(purrr)
+
+states <- seq(0,1.5, length=200)
+p <- list(r=0.7, M=1.2, q=3, b=0.15, a=0.19, alpha=0.001,init_state=0.8)
+may <-  function(y, p){
+  y * p$r * (1 - y / p$M)  - p$a * y ^ p$q / (y ^ p$q + p$b ^ p$q)
+}
+
+## approximate the zeros by minimizing absolute value of function
+f <- function(x, p) abs(may(x,p))
+
+
+# use manual visuals to determine bifurcation value of A and relevant intervals
+# for optimizer search of each equilibrium point
+#tibble(x = states,f = may(x,p), ddx = ddx(x,p)) %>%
+#  ggplot(aes(x,f)) + geom_line() + geom_line(aes(x,ddx), col = 2) + ylim(-.2, .2) + xlim(0,1)
+#P1 <- p
+#P1$a <- 0.165
+#tibble(x = states,f = may(x,P1), abs = f(x,P1)) %>%
+#  ggplot(aes(x,f)) + geom_line() + geom_line(aes(x,abs), col = 2) + ylim(-.05, .07) + xlim(0,1)
+#P2 <- p
+#P2$a <- 0.214
+#tibble(x = states,f = may(x,P2), abs = f(x,P2)) %>%
+#  ggplot(aes(x,f)) + geom_line() + geom_line(aes(x,abs), col = 2) + ylim(-.1, .1)+ xlim(0,1)
+
+
+## Compute bifurcation plot
+## Note that search intervals are tuned to parameter settings
+## Also note we must explicitly handle search outside of bifurcation vale differently
+temp <- seq(.155, .225, length.out = 50)
+bifur <- function(a){
+  p$a <- a
+  ## only valid in range when both stable points exist
+  low <- optimize(f, interval = c(0,0.28), p = p)$minimum
+  threshold <- optimize(f, interval = c(0.22,0.6), p = p)$minimum
+  high <- optimize(f, interval = c(0.4,1), p = p)$minimum
+  if(a > 0.165 & a < 0.214) {
+    data.frame(parameter=a,
+               equilibrium = c("stable", "unstable", "stable"),
+               group = c("low", "threshold", "high"),
+               state = c(low, threshold, high))
+  } else if (a <= 0.165){
+    data.frame(parameter=a,
+               equilibrium = c("stable", "unstable", "stable"),
+               group = c("low", "threshold", "high"),
[…]

### 2021-04-21 manuscript/appendix.Rmd
appendix

@@
 
-abstract: |
-  Enter the text of your abstract here.
 keywords:
@@
 bibliography: references.bib
+
+header-includes:
+  - \usepackage{amsmath}
+  - \usepackage{amssymb}
+
 output: 
@@
 # Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
-# Sys.setenv("CUDA_VISIBLE_DEVICES"="")
-
+Sys.setenv("CUDA_VISIBLE_DEVICES"="")
+set.seed(seed)
 ```
@@
 
+<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
 Standard R methods are used to plot the results summarized across the replicates:
@@
 
-\begin{equation}
-\mu_{t} = X_t + X_t r \left(1 - \frac{X_t}{K}) - \frac{a_t X_t^q}{X_t^q + b^q} 
-X_{t+1} \sim \lognormal(\mut_t, \sigma)
-\end{equation}
+\[
+\mu_{t} = X_t + X_t r \left(1 - \frac{X_t}{K} \right) - \frac{a_t X_t^q}{X_t^q + b^q}
+\]
+\[
+X_{t+1} \sim \text{lognormal}(\mu_t, \sigma)
+\]
 
@@
 
-\beign{equation}
+\[
 a_{t+1} = a_t + \alpha - A_t
-\end{equation}
+\]
 
@@
 
-$$U(X_t, A_t) = b X_t - c A_t ^2$$
-where we will take $b = 1$ and $c=1$.  
+\[U(X_t, A_t) = b X_t - c A_t ^2\]
+
+where we will take \(b = 1\) and \(c=1\).  
 Our implementation in `gym_conservation` allows the user to consider alternative parameter choices, alternative models for the ecological dynamics, and alternate types of actions, such as manipulating the ecosystem state directly rather than manipulating the environmental parameter.
@@
 select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in it's current state.
+This is implemented using the `fixed_action` method provided in our `gym_conservation` module, which also implements other heursitic models, including a human agent which requires interactive input to select the action each year.
 
@@
 ```
+
+We gather together the results under the RL agent and steady-state policy as before,
 
@@
 
+The resulting three data frames contain the necessary data for each of the subplots in figure 3 of the main text.
+
 ```{r}
@@
 ```
+
+<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
 
@@
    geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-```
-
-```{r}
-# Some individual replicates, for comparison
-sims_df %>%
-  filter(rep < 100, time < 500) %>%
-  ggplot(aes(time, state, col=model, group=interaction(model, rep))) +
-  geom_line(alpha = .1, show.legend = FALSE) + facet_wrap(~model)
-```
-
-```{r}
-# Some individual replicates, for comparison
-is_low <- sims_df %>% filter(model == "TD3") %>% group_by(rep, model) %>% summarize(low = sum(state < .2) > 3) %>% filter(low)
-sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state,  group=interaction(model, rep))) +
-  geom_line(color = pal[3], show.legend = FALSE) + facet_wrap(~rep)
-```
-
-```{r}
-# Some individual replicates, for comparison
-is_low <- sims_df %>% filter(model == "steady-state") %>% group_by(rep, model) %>% summarize(low = sum(state < .2) > 3) %>% filter(low) %>% head(16)
-sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state, group=interaction(model, rep))) +
-  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
 ```
@@
 
-```{r fig2, fig.cap=""}
+```{r fig2, include=FALSE}
 fig_sims / ( fig_policy + fig_reward)
@@
 
+Because it is difficult to get a feel for the dynamics of individual replicate simulations from ensemble statistics, we select a few example trajectories to examine directly. 
+Of the 100 replicate simulations, we pick 4 examples that dip below a state value of 0.2 for over 15 consecutive timesteps, indicating a transition into the lower basin of attraction. 
+Comparing the dynamics under the rule of thumb steady-state strategy to that of the RL-trained agent, it is clear that RL agent does a better job at both avoiding tipping points and promoting the recovery of those selected trajectories that cross into the lower attractor.
+
+```{r}
+# Some individual replicates, for comparison
+
+## First 4 of the TD3 reps falling below .2 for more than 15 steps
+is_low <- sims_df %>% 
+  filter(model == "TD3") %>% 
+  group_by(rep, model) %>% 
+  summarize(low = sum(state < .2) > 10) %>% 
+  filter(low) %>% head(6)
+
+## First 4 such cases:
+sims_df %>% inner_join(is_low) %>%
+  ggplot(aes(time, state,  group=interaction(model, rep))) +
+  geom_line(color = pal[3], show.legend = FALSE) + facet_wrap(~rep)
+```
+
+```{r}
+# Some individual replicates, for comparison
+is_low <- sims_df %>% filter(model == "steady-state") %>%
+  group_by(rep, model) %>% summarize(low = sum(state < .2) > 10) %>% 
+  filter(low) %>% head(6)
+sims_df %>% inner_join(is_low) %>%
+  ggplot(aes(time, state, group=interaction(model, rep))) +
+  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
+```
+
 ----

### 2021-04-21 manuscript/cover-letter.Rmd
appendix

@@
 
-We are pleased to present a proposal for an article discussing the potential use of deep Reinforcement Learning in theoretical and applied ecological conservation.
+We are pleased to present a proposal for an article discussing the potential use of deep Reinforcement Learning in theoretical and applied ecological conservation for the Methods section of Ecology Letters.
 

### 2021-04-21 manuscript/appendix.Rmd
rebuild :construction:

@@
 knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4)
+                      fig.width = 7, fig.height = 4, cache = TRUE)
 ggplot2::theme_set(ggplot2::theme_bw())
@@
 
-```{r fig1, fig.cap="", include=FALSE}
+## Manuscript Figure 1
+
+<!-- Not sure if it's helpful to duplicate figures from the paper in the appendix...-->
+
+```{r fig1}
 fig_sims / ( fig_policy + fig_reward)
@@
 This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carptenter2011] to the planet biosphere [@Barnosky2014].
+
+```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario.
+The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g.
+global mean temperature, arbitrary units) of 0.19.
+In the absence of conservation action, the environment worsens (e.g.
+rising mean temperature) as the parameter increases.
+This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.
+Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
+bifur_df <- read_csv("figs/bifur.csv", col_types = "dccd")
+bifur_df %>%
+  ggplot(aes(parameter, state, lty=equilibrium, group = group)) + 
+  geom_line()
+
+```
 
@@
 
-```{r fig2, include=FALSE}
+### Manuscript Figure 3
+
+```{r fig3}
 fig_sims / ( fig_policy + fig_reward)

### 2021-04-23 manuscript/cover-letter.Rmd
First pass at cover letter

@@
 Will such approaches really help us tackle a changing planet? 
-The reliance of dominant machine learning approaches on existing data can make them a valuable tool for automating well-understood steps in an analysis, but may also make them a poor tool to predict the long-term dynamics of a no-analogue future and the non-representative distribution of available ecological data.
-Supervised Learning methods have proven very successful in tasks such as image classification, with the promise of being able to automate species identification from cameras, acoustic sensors or satellite images.
-Unsupervised learning methods hold out a more ambitious promise of identifying features or making predictions without first requiring a large training set labeled data. 
-The least attention has been paid to what may be the most promising member of the ML triumvirate: Reinforcement Learning (RL). 
-RL trains a software agent to maximize some objective -- notable examples include navigation in autonomous robots or Google's AlphaZero. 
-RL requires agents to make decisions in dynamic and uncertain environments that plan ahead and change the state of the board.
-Also unlike other approaches, such agents are typically trained, not on mountains of data alone, but on simulations.
+Machine learning is traditionally divided into three categories: supervised learning, unsupervised learning and reinforcement learning.
+Of these three branches, reinforcement learning has garnered the least amount of attention from ecologists despite it potentially being the most impactful for ecological conservation.
+The goal of RL is to train a software agent to maximize some objective in a dynamic environment -- notable examples include navigation in autonomous robots or Google's AlphaZero. 
+An advantage of RL is that unlike the other branches of ML, RL can rely entirely on simulators to achieve optimal agent performance.
 
-We describe how this creates a bridge between the rich, mechanistic or process based models of our field and the power of ML to discover creative solutions to complex decision problems. 
 Rather than seeking to displace process-based models of the past century with opaque machines, we illustrate how we can use such mechanistic models to drive realistic simulations which we can then train leading RL algorithms to manage.
 We demonstrate this approach using examples from fisheries management, ecosystem tipping points, and wildfire spread. 
-All though our examples are primarily a proof-of-principle, they serve to illuminate both the promise, as well as potential pitfalls 
+Although our examples are primarily a proof-of-principle, they serve to illuminate both the promise as well as potential pitfalls. 
 Those pitfalls are not only technical, but include issues of ethics and power, particularly if the algorithms or data are proprietary. 
@@
 We further include implementations of three fully featured python modules following current leading standards, which would allow engineers and computer scientists already working in RL to test their agents against both solved and unsolved problems in conservation management.
-We believe this piece would provide an effective introduction to the concepts and practices of RL which would allow ecologists to apply, extend, and critically critique such methods.
+We believe this piece would provide an effective introduction to the concepts and practices of RL which would allow ecologists to apply, extend, and critique such methods.

### 2021-04-23 manuscript/cover-letter.Rmd
Small word choice edits
Changing models to RL sentence

@@
 The goal of RL is to train a software agent to maximize some objective in a dynamic environment -- notable examples include navigation in autonomous robots or Google's AlphaZero. 
-An advantage of RL is that unlike the other branches of ML, RL can rely entirely on simulators to achieve optimal agent performance.
+An advantage of RL is that unlike other branches of ML, RL can rely entirely on simulators to achieve optimal performance.
 
-Rather than seeking to displace process-based models of the past century with opaque machines, we illustrate how we can use such mechanistic models to drive realistic simulations which we can then train leading RL algorithms to manage.
+Rather than seeking to displace process-based models of the past century with opaque machines, we illustrate how we can use such mechanistic models to build realistic simulators that the RL agents will learn from.
 We demonstrate this approach using examples from fisheries management, ecosystem tipping points, and wildfire spread. 

### 2021-04-23 manuscript/appendix.Rmd
First broad pass at text in the appendix but will likely revise myself here

@@
 
-All though all the necessary tooling for RL is implemented in python, the R language is more familiar to most ecologists.
-Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straight forward to use these tools without ever leaving the R interface.
-In this appendix, we detail this "pure R" approach, as well as a "pure python" approach.
+Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
+Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
+In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
 
-In the R based-approach, R functions take responsibility from the user for translating commands into python code before it is executed, an approach commonly referred to as meta-programming.
-This still requires a local python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
+In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
+This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
 Alternately, users may prefer running the analysis inside a docker container.  
-The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
-This can be particularly useful for users running ML algorithms on remote servers, as configuring R, python, and CUDA environments appropriately can be challenging.
+The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
+This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.
 
 Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
-From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and python modules.
+From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
 
@@
 Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration.
-In addition to setting a random seed in our python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.
+In addition to setting a random seed in our Python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.
 
@@
 
-We begin by select an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
+We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
 
@@
 
-At the heart of each algorithm is a neural network. The type, size and depth of network, choice of activation functions and so forth, are considered "hyper-parameters" of algorithm: 
-we must first make a selection for each of these hyper-parameters to define our (untrained) agent.  
-Other hyper-parameters determine other aspects of the training regime: such as how the agent chooses to explore possible actions, the learning rate, or how future rewards are discounted (gamma).
+At the heart of each deep RL algorithm is a neural network, called the policy or policy network, which maps states to actions.
+In training, the agent is trying to learn a policy that will maximize its expected cumulative rewards.
+The parameters of a RL algorithm that detail the overall learning process are called hyperparameters while the parameters of a RL algorithm that are learned during training are called parameters.
+For example, the number of layers in a neural network is a hyperparameter but a weight in the neural net is referred to as a parameter.
+Before training an agent, we must first specify the hyperparameters. The neural net's parameters are often randomly initialized.
 Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms,
 for instance, the A2C algorithm originally described in @A2C.
-Here, we train an agent using the A2C algorithm using default hyperparameter settings, utilizing a multi-layer perceptron neural network composed of two 64-neuron layers.
+Here, we train an agent using the A2C algorithm using default hyperparameter settings with a multi-layer perceptron policy network composed of two 64-neuron layers.
 
@@
 
-Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment and the number of training iterations budgeted.
+Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment, the size of the neural networks being used and the number of training iterations budgeted.
 Therefore, we save the trained agent to disk, and only execute the training method (`learn`) if no previously saved agent is found:
@@
 Note that this is equivalent to a choice of appropriate units -- we can re-scale the interval without loss of generality.
-This is often a critical step in the design of an RL environment to facilitate successful training.
+This is often a important step in the design of an RL environment to facilitate successful training.
 Following the `gym` standard, the core methods such as `predict` and `step` operate on the re-scaled units,
@@
 With an initial state in hand, we are ready to simulate management using our agent.
-The iteration is simple: we use the agent to predict what action we should take given the current state,
-and then we take said action and examine the result to determine the future state. 
+The iteration is simple: we use the agent to predict what action we should take given the current state. Then, we take said action and examine the result to determine the future state. 
 Because these methods return additional information as well, a little extra sub-setting is required in R:
@@
 Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
-Note that the `policyfn` method is another custom method and not part of the `gym` standard, and the use of a policy function will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTM [@LSTM].
+Note that the `policyfn` method is another custom method and not part of the `gym` standard.
+The use of a policy function will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTMs [@LSTM].
 
@@
 Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
-Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance.
+Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
 For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
@@
 Of the 100 replicate simulations, we pick 4 examples that dip below a state value of 0.2 for over 15 consecutive timesteps, indicating a transition into the lower basin of attraction. 
-Comparing the dynamics under the rule of thumb steady-state strategy to that of the RL-trained agent, it is clear that RL agent does a better job at both avoiding tipping points and promoting the recovery of those selected trajectories that cross into the lower attractor.
+Comparing the dynamics under the rule of thumb steady-state strategy to that of the RL-trained agent, it is clear that the RL agent does a better job at both avoiding tipping points and promoting the recovery of those selected trajectories that cross into the lower attractor.
 

### 2021-04-26 manuscript/appendix.Rmd
Another pass over the text

@@
 
-At the heart of each deep RL algorithm is a neural network, called the policy or policy network, which maps states to actions.
-In training, the agent is trying to learn a policy that will maximize its expected cumulative rewards.
-The parameters of a RL algorithm that detail the overall learning process are called hyperparameters while the parameters of a RL algorithm that are learned during training are called parameters.
-For example, the number of layers in a neural network is a hyperparameter but a weight in the neural net is referred to as a parameter.
-Before training an agent, we must first specify the hyperparameters. The neural net's parameters are often randomly initialized.
-Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms,
+At the heart of each deep RL algorithm is a neural network or a collection of neural networks.
+The RL agent uses these networks to approximate a policy function and/or a value function.
+The policy function is the agent's mapping of states to actions -- i.e.
+the agent's strategy.
+Value functions attempt to evaluate the goodness of a state or state-action pair towards achieving the RL objective of maximizing cumulative rewards.
+Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyper-parameters_ while the parameters of a RL algorithm that are learned during training are called parameters.
+For example, the number of layers in a neural network is a hyper-parameter but a weight in the neural net is referred to as a parameter.
+Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
+Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms,
 for instance, the A2C algorithm originally described in @A2C.
-Here, we train an agent using the A2C algorithm using default hyperparameter settings with a multi-layer perceptron policy network composed of two 64-neuron layers.
+Here, we train an agent using the A2C algorithm using default hyper-parameter settings with a multi-layer perceptron policy and value network composed of two 64-neuron layers.
 
@@
 
-Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment, the size of the neural networks being used and the number of training iterations budgeted.
+Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment, the neural network architecture and the number of training iterations budgeted.
 Therefore, we save the trained agent to disk, and only execute the training method (`learn`) if no previously saved agent is found:
@@
 
-Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyperparameter selection.
+Note that while default hyper-parameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyper-parameter selection.
 This is discussed further below.
@@
 Note that this is equivalent to a choice of appropriate units -- we can re-scale the interval without loss of generality.
-This is often a important step in the design of an RL environment to facilitate successful training.
+This is often an important step in the design of an RL environment to facilitate successful training.
 Following the `gym` standard, the core methods such as `predict` and `step` operate on the re-scaled units,
@@
 Because the agent typically attempts to retain the fish population near a specific value, simulations with well-trained agents will usually not explore the full range of possible states.
-To get a better idea of how the agent behaves across the full state space, it common to consider a policy function: a plot of the action taken at each possible state.
+To get a better idea of how the agent behaves across the full state space, it is common to examine the policy function: to do this, we'll plot the action taken at every possible state.
 Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
 Note that the `policyfn` method is another custom method and not part of the `gym` standard.
-The use of a policy function will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTMs [@LSTM].
+The use of `policyfn` will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTMs [@LSTM].
 
@@
 We train a second agent based on the TD3 algorithm. 
-This illustrates some of the salient differences between the different algorithms used in reinfocement learning.
-Bear in mind that the best algorithm for any given environment will depend on the details of the environment, as well
-as the length of training and the selection of hyper-parameters, including the agent's underlying neural networks which provide the 'brains' of the agent's decision-making ability.
+This illustrates some of the salient differences between the different algorithms used in reinforcement learning.
+Bear in mind that the best algorithm for any given environment will depend on the details of the environment as well
+as the length of training and the selection of hyper-parameters.
 This time, we will declare our choices for the hyper-parameters explicitly.
 Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
-Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
+Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
 For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].

### 2021-04-27 manuscript/cover-letter.Rmd

@@
 Of these three branches, reinforcement learning has garnered the least amount of attention from ecologists despite it potentially being the most impactful for ecological conservation.
-The goal of RL is to train a software agent to maximize some objective in a dynamic environment -- notable examples include navigation in autonomous robots or Google's AlphaZero. 
+The goal of RL is to train a software agent to maximize some objective in a dynamic environment.
+To date, researchers have used RL to make a number of impressive achievements on decision-making problems that were previously thought to be unsolvable -- one notable example is Deepmind's AlphaGo agent which defeated a Go Grandmaster.
+
 An advantage of RL is that unlike other branches of ML, RL can rely entirely on simulators to achieve optimal performance.
-
-Rather than seeking to displace process-based models of the past century with opaque machines, we illustrate how we can use such mechanistic models to build realistic simulators that the RL agents will learn from.
+Rather than seeking to displace process-based ecological models of the past century with opaque machines, we illustrate how we can use such mechanistic models to build realistic simulators that the RL agents will learn from.
 We demonstrate this approach using examples from fisheries management, ecosystem tipping points, and wildfire spread. 

### 2021-04-28 manuscript/presubmission-letter.Rmd

--- 2021-04-23 manuscript/cover-letter.Rmd
@@
 
-We are pleased to present a proposal for an article discussing the potential use of deep Reinforcement Learning in theoretical and applied ecological conservation for the Methods section of Ecology Letters.
+We would like to propose an article for the Methods section of Ecology Letters which discusses and displays the potential applications of deep Reinforcement Learning in both theoretical and applied ecological conservation and decision making.
 
 Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet's most pressing environmental problems, such as the growing frequency and intensity of wildfire, over-exploited fisheries, declining biodiversity, and zoonotic pandemics.
-Will such approaches really help us tackle a changing planet? 
-Machine learning is traditionally divided into three categories: supervised learning, unsupervised learning and reinforcement learning.
-Of these three branches, reinforcement learning has garnered the least amount of attention from ecologists despite it potentially being the most impactful for ecological conservation.
-The goal of RL is to train a software agent to maximize some objective in a dynamic environment -- notable examples include navigation in autonomous robots or Google's AlphaZero. 
-An advantage of RL is that unlike other branches of ML, RL can rely entirely on simulators to achieve optimal performance.
+But will ML approaches really help us tackle a changing planet?
+Applications of ML in ecology have begun to illustrate the promise of both supervised learning (e.g.
+image recognition) and unsupervised learning (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
+Supervised and unsupervised learning approaches require massive amounts of historical data to learn underlying patterns.  
+Even when such data is eventually available, they may prove reliable for relatively static tasks such as species identification from audio or visual recordings, yet be a poor guide to a future climate we have never observed.
+The task of selecting actions in an uncertain and changing environment to maximize some objective is the explicit focus of RL.
 
-Rather than seeking to displace process-based models of the past century with opaque machines, we illustrate how we can use such mechanistic models to build realistic simulators that the RL agents will learn from.
-We demonstrate this approach using examples from fisheries management, ecosystem tipping points, and wildfire spread. 
+To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and games. 
+_Nature_ recently featured RL agents achieving superhuman performance on classic Atari games, _Science_ described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
+We believe that similar techniques could improve decision-making in managing ecological environments as well. 
+Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
+Detailed, process-based models allow a window into the future scenarios of ecological management.
+Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.
 Although our examples are primarily a proof-of-principle, they serve to illuminate both the promise as well as potential pitfalls.
+
 Those pitfalls are not only technical, but include issues of ethics and power, particularly if the algorithms or data are proprietary. 
 We conclude with a discussion of how an open, transparent and reproducible approach can help mitigate some concerns, while also offering a more effective interface between teams of researchers from both ecological and computer sciences.
-
 We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.

### 2021-04-28 manuscript/appendix.Rmd

@@
 
+<!-- Kari: a brief explaination of the gym$function syntax would probably be helpful - is that a reticulate thing? Just reading later it looks like you reference it as a "method"?-->
+
 ```{r }
@@
 
+<!-- Kari: the role neural networks play in RL is not very well explained in the paper, more detail there would probably make this part more intuitive --> 
 At the heart of each deep RL algorithm is a neural network or a collection of neural networks.

### 2021-04-28 manuscript/manuscript.Rmd

@@
 authors:
+
   - name: Marcus Lapeyrolerie
@@
     email: cboettig@berkeley.edu
-
 abstract: |
@@
   - \usepackage{amssymb}
-
 output: 
@@
     keep_tex: true
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
-Deep Reinforcement Learning (RL), a machine learning technique in which a _software agent_ is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
-Unlike more widely recognized machine learning techniques for either _supervised_ learning (such as image classification, @species_ex, @satellite_ex) or unsupervised learning, RL focuses on solving _dynamic_ problems which require planning ahead, often in uncertain and changing environments.
+Deep Reinforcement Learning (RL), a machine learning technique in which a *software agent* is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
+Unlike more widely recognized machine learning techniques for either *supervised* learning (such as image classification, [@species_ex], [@satellite_ex]) or unsupervised learning, RL focuses on solving *dynamic* problems which require planning ahead, often in uncertain and changing environments.
 Of all machine learning techniques, RL most resembles a general artificial intelligence.
-Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero],
-and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex]. 
+Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero], and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex].
 These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.
@@
 <!-- Millie help on this para? -->
+
 If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
 The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
-The endangered species act ...
-Ecosystem services ....
-Such simplicity is in no small part due to the challenge associated in determining a 
-Some reference to @TomDietrich on potential for CS-based application.
+The endangered species act ... Ecosystem services ....
+Such simplicity is in no small part due to the challenge associated in determining a Some reference to [@TomDietrich] on potential for CS-based application.
 
-In contrast to other ML methods, RL does not require massive data sets for training.  
+In contrast to other ML methods, RL does not require massive data sets for training.\
 Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
@@
 Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
-Methods from _optimal control theory_ have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
+Methods from *optimal control theory* have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
 However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
@@
 
-If RL has great promise, the approach also imposes great risk.  
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].  
-Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems. 
-RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes.
-Technical issues are not the only pitfall of AI applications to conservation problems.
-Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.
+If RL has great promise, the approach also imposes great risk.\
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].\
+Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems
+. RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes
+. Technical issues are not the only pitfall of AI applications to conservation problems
+.
+Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property
+.
 
@@
 
+<!-- Kari: We reference Markov Decision processes as an alternative approach in the tipping points example and just generally elsewhere so it might be helpful to talk about them in the introduction-->
+
+#### Key Terminology
+
+-   model (underlying ecological model)
+
+```{=html}
+<!-- -->
+```
+-   agent
+
+-   trained agent (model)
+
+-   environment
+
+-   predict
+
+-   policy
+
+-   model-free
+
+-   model-based
+
 # Sequential Decision Problems and Reinforcement Learning
@@
 <!-- meh, not sure we need this? -->
+
 Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to as the reward or utility).
@@
 This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
-The task for the agent is to learn a _policy_, $\pi$, that prescribes what action the agent should perform after observing a certain state.
-The agent interacts with the environment via trial and error in what is called the _training_ period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.
+The task for the agent is to learn a *policy*, $\pi$, that prescribes what action the agent should perform after observing a certain state.
+The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.
 
@@
 
-\[
+$$
   p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
-\]
+$$
 
@@
 
-\[
+$$
   \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
-\]
+$$
 
-Note that this formalism extends to an infinite horizon case when $T= \infty$. While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
+Note that this formalism extends to an infinite horizon case when $T= \infty$.
+While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
 
@@
 
-\[
+$$
   V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-\]
+$$
 
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-\[
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, $$
   Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-\]
+$$
 
 And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succintly defined as
-\[
+The advantage function can then be succintly defined as $$
   A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-\]
-Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+$$ Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
 Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
@@
 
+```{=html}
 <!--
@@
 -->
-
+```
 There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
@@
 In model-based RL, the agent is attempting to learn the state transition operator, $T$.
-Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
+Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?]
+.
 In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
-The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?].
+The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
+.
 
@@
 Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
+
+<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
 
@@
 ```
-
 # Examples
@@
 The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
-Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@].
+Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [\@].
 
-For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
+For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1974] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
 In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
@@
 From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."
-The simplest simulation is merely an implementation of the estimated model itself.
+In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment." The simplest simulation is merely an implementation of the estimated model itself.
 However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on @TD3.
+To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on [@TD3] <!-- Kari: are these different algorithms?-->.
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
@@
 We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
-This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of @May1977 on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
+This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
 On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
@@
 Because this problem involves a parameter whose value changes over time (the slowly deteriorating environment), the resulting ecosystem dynamics are not autonomous.
-This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, @Marescot2013), typically used to solve sequential decision-making problems.
+This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, [@Marescot2013]), typically used to solve sequential decision-making problems.
 However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham; @Possingham; @morePossingham].
@@
 Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
-This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.  
+This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.\
 The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
@@
 A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
+<!-- Kari: lots of jargon in this paragraph that needs to be defined-->
 
@@
 -->
-
+```
 <!-- Something about relevance to both the _theory_ and practice of ecological conservation.
@@
 
-The examples presented here only scrape the surface of possible RL applicaations to conservation problems.
+The examples presented here only scrape the surface of possible RL applications to conservation problems.
 The examples we have focused on are intentionally quite simple, though it is worth remembering that these very same simple models have a long history of relevance and application in both research and policy contexts.
-Despite their simplicity, the optimal strategy is not always obvious before hand, however intutive it may appear in retrospect.
+Despite their simplicity, the optimal strategy is not always obvious before hand, however intuitive it may appear in retrospect.
 In the case of the ecosystem tipping point scenario, the optimal strategy is unknown, and the approximate solution found by our RL implementation could almost certainly be improved upon.
@@
 The hope of the RL practitioner is that after doing an exhaustive hyperparameter search, the practitioner will find a set of hyperparameters that results in a robust, well-performing agent, but there are no guarantees that hyperparameter tuning will lead to this outcome.
+<!-- Kari: what does it mean for two algorithms with similar hyperparameters to perform differently? (As happens in our examples) -->
 
@@
 Beyond typically having large state-action spaces for agents to explore, conservation problems tend to have sparse reward structures.
-E.g.
-rarely does an action in conservation lead to immediate rewards; instead, benefits are typically observed after actions have been performed.
+E.g. rarely does an action in conservation lead to immediate rewards; instead, benefits are typically observed after actions have been performed.
 Sparse reward structures are challenging for DRL agents, because of the credit assignment problem [cite].
@@
 -   wild west of AI?
--can separate RL from having gyms implemented -> how do you compare different models, 
--mention that customization is probably the way to go, baselines only go so far
+    -can separate RL from having gyms implemented -\> how do you compare different models, -mention that customization is probably the way to go, baselines only go so far
 

### 2021-04-28 manuscript/appendix.Rmd
data example

--- 2021-04-26 manuscript/appendix.Rmd
@@
 knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4, cache = TRUE)
+                      fig.width = 7, fig.height = 4, cache = FALSE)
 ggplot2::theme_set(ggplot2::theme_bw())
@@
 
+## Applying an RL agent to real data
+
+We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.
+
+Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
+We will use the agent we have just trained using the `gym_fishing` simulator above,
+
+```{r}
+# Simulate management under the trained agent
+env <- gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
+td3 <- sb3$TD3$load("cache/td3")
+
+```
+
+```{r}
+hake <- read_csv("../data/hake.csv")
+x0 <- hake %>%
+ filter(year == min(year)) %>%
+  pull(scaled_biomass)
+Tmax <- 15
+years <-1986:2000
+
+td3_state <- td3_action <- numeric(Tmax)
+td3_state[1] <- x0
+td3_action[1] <- NA
+
+```
+
+Our hindcast considers the scenario of managing catch quotas using TD3, beginning in 1986. 
+We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
+This shows that quotas set by TD3 would have been lower than historical values throughout the late 80s and early 90s, which
+saw a sharp decline in estimated biomass of the Argentine Hake.
+For the same or higher estimated stock sizes from the historical record, we see that TD3 would recommend a lower quota.
+We also see that under TD3 the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
+being set in the now-depleted stock.  
+Whether the TD3 quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
+However, the fact that TD3 would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
+Combined with the observed historical declines, this provides some evidence that the RL agent could have decreased or avoided the over-harvesting in this case.
+
+```{r}
+env$reset()
+## represent the initial state size in the 'rescaled' state space.
+state <- env$get_state( x0 )
+
+for(i in 2:Tmax){
+
+  # RL-recommended harvest action:
+  out <- td3$predict(state)
+  action <- out[[1]]
+
+  # Record state and resulting action
+  td3_state[i] <- env$get_fish_population(state)
+  td3_action[i] <- env$get_quota(action)
+
+  # Implement RL-recommended harvest. 
+  result <- env$step(action)
+  state <- result[[1]]
+
+  #state <- env$get_state(hake$scaled_biomass[i+1])
+}
+
+df <- bind_rows(
+  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
+  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
+
+df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))
+
+```
+
 # Ecological tipping points
@@
 ```{r}
+noise_std = 0.4805935357322933
 
-noise_std = 0.4805935357322933
-action_noise = sb3$common$noise$OrnsteinUhlenbeckActionNoise(mean = np$zeros(1L),
-                                            sigma = noise_std * np$ones(1L))
+OU <- sb3$common$noise$OrnsteinUhlenbeckActionNoise
+action_noise = OU(mean = np$zeros(1L),  sigma = noise_std * np$ones(1L))
 
@@
             verbose = 0, 
-            seed = seed,
+            seed = 42L,
             "gamma"= 0.995,

### 2021-04-28 manuscript/manuscript.Rmd
data example

--- 2021-04-21 manuscript/manuscript.Rmd
@@
 
-In contrast to other ML methods, RL does not require massive data sets for training.  
-Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
+In contrast to other ML methods, RL does not require massive data sets for training, relying instead on model simulations based on the best available ecological data and knowledge of underlying processes.
+Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains sparse and unevenly sampled [CITE].
 Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.

### 2021-04-28 manuscript/presubmission-letter.Rmd
Small edits and a comment

@@
 image recognition) and unsupervised learning (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
-Supervised and unsupervised learning approaches require massive amounts of historical data to learn underlying patterns.  
+Supervised and unsupervised learning approaches often require massive amounts of historical data to learn underlying patterns.
 Even when such data is eventually available, they may prove reliable for relatively static tasks such as species identification from audio or visual recordings, yet be a poor guide to a future climate we have never observed.
@@
 
-To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and games. 
-_Nature_ recently featured RL agents achieving superhuman performance on classic Atari games, _Science_ described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
+To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games. 
+_Nature_ recently featured RL agents achieving superhuman performance on classic Atari games, and _Science_ described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
 We believe that similar techniques could improve decision-making in managing ecological environments as well.
+<!-- Having trouble reconciling lines 21-23, trying to say that these simulators are like the grand challenges and given RL's track record of solving intractable decision-making problems, perhaps RL could shed light here?
+I think having a sentence asserting that RL would use ecological models not replace them would be good somewhere in here.-->
 Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.

### 2021-04-28 manuscript/presubmission-letter.Rmd
Minor clarification edit

@@
 Supervised and unsupervised learning approaches often require massive amounts of historical data to learn underlying patterns.
-Even when such data is eventually available, they may prove reliable for relatively static tasks such as species identification from audio or visual recordings, yet be a poor guide to a future climate we have never observed.
+Even when such data is eventually available, these two methods may prove reliable for relatively static tasks such as species identification from audio or visual recordings, yet be a poor guide to a future climate we have never observed.
 The task of selecting actions in an uncertain and changing environment to maximize some objective is the explicit focus of RL.

### 2021-04-29 python/conservation_TD3.py
Appendix examples as separate python scripts

@@ -1,0 +1,43 @@
+# Ecological Tipping Points using TD3
+
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_conservation
+
+seed = 24
+np.random.seed(seed)
+
+env = gym.make("conservation-v6")
+
+noise_std = 0.4805935357322933
+OU = sb3.common.noise.OrnsteinUhlenbeckActionNoise
+action_noise = OU(mean = 0,  sigma = noise_std)
+
+model = sb3.TD3('MlpPolicy', 
+            env, 
+            verbose = 0, 
+            seed = 42,
+            gamma= 0.995,
+            learning_rate =  8.315382409902049e-05,
+            batch_size = 512,
+            buffer_size = 10000,
+            train_freq = 1000,
+            gradient_steps = 1000,
+            n_episodes_rollout = -1,
+            action_noise = action_noise,
+            policy_kwargs = dict("net_arch"= c(64,64)))
+
+model.learn(total_timesteps=3000000)
+model.save("cache/td3-conservation")
+  
+model = sb3.TD3.load("cache/td3-conservation")
+TD3_sims = env.simulate(model, reps = 100)
+TD3_policy = env.policyfn(model, reps = 10)
+
+# Simulate under the steady-state solution (given the model)
+K = 1.5
+alpha = 0.001
+opt = gym_conservation.models.fixed_action(env, fixed_action = alpha * 100 * 2 * K )
+opt_sims = env.simulate(opt, reps = 100L)
+opt_policy = env.policyfn(opt)

### 2021-04-29 python/conservation_steadystate.py

@@ -1,0 +1,18 @@
+# Using steady-state policy instead of RL:
+
+import numpy as np
+import gym
+import gym_conservation
+from gym_conservation import models
+
+seed = 24
+np.random.seed(seed)
+
+env = gym.make("conservation-v6")
+
+# Simulate under the steady-state solution (given the model)
+K = 1.5
+alpha = 0.001
+steadystate = models.fixed_action(env, fixed_action=alpha * 100 * 2 * K)
+opt_sims = env.simulate(steadystate, reps=100)
+opt_policy = env.policyfn(steadystate)

### 2021-04-29 manuscript/data-example.Rmd

```{r knitr, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4, cache = TRUE)
ggplot2::theme_set(ggplot2::theme_bw())

scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
pal <- ggthemes::solarized_pal()(8)
txtcolor <- "#586e75"
```

```{r setup, cache = FALSE, message=FALSE}
# R dependencies
library(tidyverse)
library(patchwork)
library(reticulate)

## Python dependencies loaded via R
sb3         <- import ("stable_baselines3")
gym         <- import ("gym")
gym_fishing <- import("gym_fishing")
```

```{r}
library(contentid)
ramzip <- resolve("hash://md5/fac27d8b7876df29afaeadcd7f5a3b77")
unzip(ramzip, exdir = "ramlegacy")
load("ramlegacy/RAMLDB v4.491/DB Files With Assessment Data/R Data/DBdata[asmt][v4.491].RData")
```

```{r}
ramlegacy <- 
  timeseries_values_views %>%
  select(stockid, stocklong, year, SSB, TC) %>%
  left_join(stock) %>%
  left_join(area) %>%
  select(stockid, scientificname, 
         commonname, areaname, country, year, 
         SSB, TC) %>%
  left_join(timeseries_units_views %>%
              rename(TC_units = TC, SSB_units = SSB)) %>%
  select(scientificname, commonname, 
         stockid, areaname, country, year, 
         SSB, TC, SSB_units, TC_units)
```

Let's filter out missing data, non-matching units, and obvious reporting errors (catch exceeding total spawning biomass), then we re-scale each series into the 0,1 by appropriate choice of units:

```{r}
stock_ids <- c("PLAICNS", "ARGHAKENARG")

fish <- ramlegacy %>% 
  filter(stockid %in% stock_ids) %>% 
  filter(!is.na(SSB), !is.na(TC)) %>%
  filter(SSB_units == "MT", TC_units=="MT") %>% 
  filter(SSB > TC) %>%
  select(-SSB_units, -TC_units) %>% 
  group_by(stockid) %>%
  mutate(scaled_catch = TC / max(SSB),
         scaled_biomass = SSB / max(SSB))

hake <- fish  %>% filter(commonname=="Argentine hake")
write_csv(hake, "../data/hake.csv")
```

```{r}
# Simulate management under the trained agent
env <- gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
td3 <- sb3$TD3$load("cache/td3")

```

```{r}
hake <- read_csv("../data/hake.csv")
x0 <- hake %>%
 filter(year == min(year)) %>%
  pull(scaled_biomass)
Tmax <- 15
years <-1986:2000

td3_state <- td3_action <- numeric(Tmax)
td3_state[1] <- x0
td3_action[1] <- NA
## represent the initial state size in the 'rescaled' state space.
state <- env$get_state( x0 )
```

```{r}

for(i in 2:Tmax){

  # RL-recommended harvest action:
  out <- td3$predict(state)
  action <- out[[1]]

  # Record state and resulting action
  td3_state[i] <- env$get_fish_population(state)
  td3_action[i] <- env$get_quota(action)

  # Implement RL-recommended harvest:
  result <- env$step(action)
  state <- result[[1]]

}

df <- bind_rows(
  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))

df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))

```

```{r subset}
examples <- df2 %>% 
  ungroup() %>% 
  group_by(commonname)
```

```{r}
## Model does not estimate sigma_m; data is insufficient to do so.
gs_code  <- nimble::nimbleCode({
  r ~ dunif(0, 2)
  K ~ dunif(0, 2)
  sigma ~ dunif(0, 1)

  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K) - min(a[t],x[t])
    x[t+1] ~ dnorm(mu[t], sd = sigma)
  }
})
fit_models <- function(fish, code){
  # fish <- examples %>% filter(stockid == stock_ids[1])

  ## Rescale data
  N <- dim(fish)[1]
  scaled_data <- data.frame(t = 1:N, 
                            y = fish$scaled_biomass, 
                            a = fish$scaled_catch)
  data = data.frame(x = scaled_data$y)
  ## Compile  model
  constants <- list(N = N, a = scaled_data$a)
  inits <- list(r = 0.5, K = 0.5, sigma = 0.02, x0 = scaled_data$y[1])
  model <- nimbleModel(code, constants, data, inits)
  C_model <- compileNimble(model)

  mcmcspec <- configureMCMC(model, thin = 1e2)
  mcmc <- buildMCMC(mcmcspec)
  Cmcmc <- compileNimble(mcmc, project = model)
  Cmcmc$run(1e6)

  samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
  burnin <- 1:(0.05 * dim(samples)[1]) # drop first 5%
  samples <- samples[-burnin,1:(length(inits) - 1)] # drop raised vars, burnin
  #gather(samples) %>% ggplot() + geom_density(aes(value)) + facet_wrap(~key, scale='free')

  ## Return fit
  data.frame(stockid = fish$stockid[1],
             commonname = fish$commonname[1],
             r = mean(samples$r),
             K = mean(samples$K),
             sigma_g = mean(samples$sigma),
             r_sd = sd(samples$r),
             K_sd = sd(samples$K),
             sigma_g_sd = sd(samples$sigma),
             stringsAsFactors = FALSE)

}
```

```{r}
set.seed(123)
fits <- examples %>% do(fit_models(., code=gs_code))
fits 
```

```{r}
pars <- fits %>% ungroup() %>% select(commonname, r, K, sigma_g) 
pars
```

### 2021-04-29 manuscript/manuscript.Rmd

--- 2021-04-28 manuscript/manuscript.Rmd
@@
 
-In contrast to other ML methods, RL does not require massive data sets for training.\
-Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
+In contrast to other ML methods, RL does not require massive data sets for training, relying instead on model simulations based on the best available ecological data and knowledge of underlying processes.
+Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains sparse and unevenly sampled [CITE].
 Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
@@
 
-If RL has great promise, the approach also imposes great risk.\
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].\
+If RL has great promise, the approach also imposes great risk.
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].
 Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems
@@
 
-```{=html}
-<!-- -->
-```
 -   agent
@@
 
-```{=html}
 <!--
@@
 -->
-```
+
 There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.

### 2021-04-29 manuscript/presubmission-letter.Rmd

--- 2021-04-28 manuscript/presubmission-letter.Rmd
@@
 But will ML approaches really help us tackle a changing planet?
-Applications of ML in ecology have begun to illustrate the promise of both supervised learning (e.g.
+Current applications of ML in ecology have made great strides in both supervised learning (e.g.
 image recognition) and unsupervised learning (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
-Supervised and unsupervised learning approaches often require massive amounts of historical data to learn underlying patterns.
-Even when such data is eventually available, they may prove reliable for relatively static tasks such as species identification from audio or visual recordings, yet be a poor guide to a future climate we have never observed.
-The task of selecting actions in an uncertain and changing environment to maximize some objective is the explicit focus of RL.
+Where supervised and unsupervised learning approaches require massive amounts of historical data to learn underlying patterns and focus on relatively static tasks (e.g.
+species identification from audio or visual recordings), RL focuses on the task of selecting actions in an uncertain and changing environment.
+These characteristics make RL particularly well suited for for addressing some of the most challenging problems in ecology.
 
 To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
-_Nature_ recently featured RL agents achieving superhuman performance on classic Atari games, and _Science_ described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
+*Nature* recently featured RL agents achieving superhuman performance on classic Atari games, and *Science* described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
 We believe that similar techniques could improve decision-making in managing ecological environments as well.
 <!-- Having trouble reconciling lines 21-23, trying to say that these simulators are like the grand challenges and given RL's track record of solving intractable decision-making problems, perhaps RL could shed light here?
-I think having a sentence asserting that RL would use ecological models not replace them would be good somewhere in here.-->
-Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
-Detailed, process-based models allow a window into the future scenarios of ecological management.
+I think having a sentence asserting that RL would use ecological models not replace them would be good somewhere in here.--> Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
+<!-- Kari: possible rewrite of the following sentence, not sure if it helps --> While existing process-based models allow a window into the future scenarios of ecological management, RL has the potential to find optimal solutions in those spaces.
 Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.

### 2021-04-29 manuscript/appendix.Rmd
:pen:

@@
 
-scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(8)
-txtcolor <- "#586e75"
+scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
+pal = ggthemes::solarized_pal()(8)
+txtcolor = "#586e75"
 ```
+
+# Overview
+
+Development of any deep reinforcement learning application can be divided into two components: an _environment_ and an _agent_.
+An environment is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
+An agent is typically a computational neural network trained using an appropriate algorithm to suggest an action based on available information about the environment.
+
+Somewhat confusingly, frameworks will often refer to a trained agent as a "model", while also referring to the approaches used to implement and train the agent as "model free" because the agent does not attempt to form an internal model of the environmental process.
+We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.
+
+@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both poblems.
+
+A rapidly growing set of material for learning about deep reinforcement learning is now freely available.
 
@@
 At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
+While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
+Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
+Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3].
+Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
+Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
+However, higher-level interfaces and high-level documentation are still relatively limited.
+OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
+OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
+Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption.
+Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
+The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
+Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
+
+# Building a simulation environment `gym`
+
+[@openai_gym]
 
@@
 ## Python dependencies loaded via R
-sb3         <- import ("stable_baselines3")
-gym         <- import ("gym")
-gym_fishing <- import("gym_fishing")
-gym_conservation <- import("gym_conservation")
+sb3         = import ("stable_baselines3")
+gym         = import ("gym")
+gym_fishing = import("gym_fishing")
+gym_conservation = import("gym_conservation")
 ```
@@
 ## reproducible settings
-np <- import("numpy")
-seed <- 24L # integer
+np = import("numpy")
+seed = 24L # integer
 np$random$seed(seed)
@@
 ## initialize the environment
-env <- gym$make("fishing-v1", sigma = 0.1)
+env = gym$make("fishing-v1", sigma = 0.1)
 ```
@@
 ```{r}
-a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
+a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
 ```
@@
 # Simulate management under the trained agent
-a2c <- sb3$A2C$load("cache/a2c")
+a2c = sb3$A2C$load("cache/a2c")
 ```
@@
 ## represent the initial state size in the 'rescaled' state space.
-state <- env$get_state( 0.75 )
+state = env$get_state( 0.75 )
 state
@@
 
-  out <- a2c$predict(state)
-  action <- out[[1]]
-  result <- env$step(action)
-  state <- result[[1]]
+  out = a2c$predict(state)
+  action = out[[1]]
+  result = env$step(action)
+  state = result[[1]]
 
@@
 ```{r}
-a2c_sims <- env$simulate(a2c, reps = 500L)
+a2c_sims = env$simulate(a2c, reps = 500L)
 ```
@@
 ```{r}
-a2c_policy <- env$policyfn(a2c, reps = 50L)
+a2c_policy = env$policyfn(a2c, reps = 50L)
 ```
@@
 # Simulate management under the trained agent
-td3 <- sb3$TD3$load("cache/td3")
-td3_sims <- env$simulate(td3, reps = 500L)
-td3_policy <- env$policyfn(td3, reps = 50L)
+td3 = sb3$TD3$load("cache/td3")
+td3_sims = env$simulate(td3, reps = 500L)
+td3_policy = env$policyfn(td3, reps = 50L)
 ```
@@
 # Simulate under the optimal solution (given the model)
-opt <- gym_fishing$models$escapement(env)
-opt_sims <- env$simulate(opt, reps = 500L)
-opt_policy <- env$policyfn(opt)
+opt = gym_fishing$models$escapement(env)
+opt_sims = env$simulate(opt, reps = 500L)
+opt_policy = env$policyfn(opt)
 ```
@@
 ```{r}
-sims_df <- bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
+sims_df = bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
   mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
 
-policy_df <- bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
+policy_df = bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
   mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
 
-gamma <- 1 #discount
-reward_df <- sims_df %>% 
+gamma = 1 #discount
+reward_df = sims_df %>% 
   group_by(rep, model) %>%
@@
 ```{r}
-ymin <- function(x) last(x[(ntile(x, 20)==1)])
-ymax <- function(x) last(x[(ntile(x, 20)==19)])
+ymin = function(x) last(x[(ntile(x, 20)==1)])
+ymax = function(x) last(x[(ntile(x, 20)==19)])
 
-fig_sims <- 
+fig_sims = 
 sims_df %>% 
@@
 ```{r}
-fig_policy <- 
+fig_policy = 
   policy_df %>% ggplot(aes(state, action, 
@@
 
-fig_reward <- reward_df %>% 
+fig_reward = reward_df %>% 
   ggplot(aes(time, mean_reward)) + 
@@
 # Simulate management under the trained agent
-env <- gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-td3 <- sb3$TD3$load("cache/td3")
+env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
+td3 = sb3$TD3$load("cache/td3")
 
@@
 ```{r}
-hake <- read_csv("../data/hake.csv")
-x0 <- hake %>%
+hake = read_csv("../data/hake.csv")
+x0 = hake %>%
  filter(year == min(year)) %>%
   pull(scaled_biomass)
-Tmax <- 15
-years <-1986:2000
+Tmax = 15
+years =1986:2000
 
-td3_state <- td3_action <- numeric(Tmax)
-td3_state[1] <- x0
-td3_action[1] <- NA
+td3_state = td3_action = numeric(Tmax)
+td3_state[1] = x0
+td3_action[1] = NA
 
@@
 ## represent the initial state size in the 'rescaled' state space.
-state <- env$get_state( x0 )
+state = env$get_state( x0 )
 
@@
   # RL-recommended harvest action:
-  out <- td3$predict(state)
-  action <- out[[1]]
+  out = td3$predict(state)
+  action = out[[1]]
 
   # Record state and resulting action
-  td3_state[i] <- env$get_fish_population(state)
-  td3_action[i] <- env$get_quota(action)
+  td3_state[i] = env$get_fish_population(state)
+  td3_action[i] = env$get_quota(action)
 
   # Implement RL-recommended harvest. 
-  result <- env$step(action)
-  state <- result[[1]]
+  result = env$step(action)
+  state = result[[1]]
 
-  #state <- env$get_state(hake$scaled_biomass[i+1])
+  #state = env$get_state(hake$scaled_biomass[i+1])
 }
 
-df <- bind_rows(
+df = bind_rows(
   tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
@@
 Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
-bifur_df <- read_csv("figs/bifur.csv", col_types = "dccd")
+bifur_df = read_csv("figs/bifur.csv", col_types = "dccd")
 bifur_df %>%
@@
 ```{r}
-env <- gym$make("conservation-v6")
+env = gym$make("conservation-v6")
 ```
@@
 
-OU <- sb3$common$noise$OrnsteinUhlenbeckActionNoise
+OU = sb3$common$noise$OrnsteinUhlenbeckActionNoise
 action_noise = OU(mean = np$zeros(1L),  sigma = noise_std * np$ones(1L))
@@
 
-model <- sb3$TD3$load("cache/td3-conservation")
-TD3_sims <- env$simulate(model, reps = 100L)
-TD3_policy <- env$policyfn(model, reps = 10L)
+model = sb3$TD3$load("cache/td3-conservation")
+TD3_sims = env$simulate(model, reps = 100L)
+TD3_policy = env$policyfn(model, reps = 10L)
 ```
@@
 alpha = 0.001
-opt <- gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
-opt_sims <- env$simulate(opt, reps = 100L)
-opt_policy <- env$policyfn(opt)
+opt = gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
+opt_sims = env$simulate(opt, reps = 100L)
+opt_policy = env$policyfn(opt)
 ```
@@
 ```{r}
-sims_df <- bind_rows(TD3_sims, opt_sims, .id = "model") %>%
+sims_df = bind_rows(TD3_sims, opt_sims, .id = "model") %>%
   mutate(model = c("TD3", "steady-state")[as.integer(model)])
 
-policy_df <- bind_rows(TD3_policy, opt_policy, .id = "model") %>%
+policy_df = bind_rows(TD3_policy, opt_policy, .id = "model") %>%
   mutate(model = c("TD3", "steady-state")[as.integer(model)])
 
-gamma <- 1 #discount
-reward_df <- sims_df %>% 
+gamma = 1 #discount
+reward_df = sims_df %>% 
   group_by(rep, model) %>%
@@
 # ensemble statistics
-ymin <- function(x) last(x[(ntile(x, 20)==1)])
-ymax <- function(x) last(x[(ntile(x, 20)==19)])
+ymin = function(x) last(x[(ntile(x, 20)==1)])
+ymax = function(x) last(x[(ntile(x, 20)==19)])
 
-fig_sims <- 
+fig_sims = 
   sims_df %>%
@@
 ## First 4 of the TD3 reps falling below .2 for more than 15 steps
-is_low <- sims_df %>% 
+is_low = sims_df %>% 
   filter(model == "TD3") %>% 
@@
 # Some individual replicates, for comparison
-is_low <- sims_df %>% filter(model == "steady-state") %>%
+is_low = sims_df %>% filter(model == "steady-state") %>%
   group_by(rep, model) %>% summarize(low = sum(state < .2) > 10) %>% 

### 2021-04-29 manuscript/appendix.Rmd
WIP

@@
 
-Development of any deep reinforcement learning application can be divided into two components: an _environment_ and an _agent_.
-An environment is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
-An agent is typically a computational neural network trained using an appropriate algorithm to suggest an action based on available information about the environment.
+Development of any deep reinforcement learning application can be divided into two components: an *environment* and an *agent*.
+An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
+An *agent* is typically a computational neural network based on an appropriate RL algorithm that may have been 'trained' on one or more environments.
+
+## Designing and building an environment
+
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
+In principle, there are virtually no restrictions on what comprises a state: it may be continuous or discrete, completely or only partially observed, one or many dimensions.
+For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
+The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
+The reward will in general depend on both the current state and choice of actions.
+An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
+In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
+In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+
+@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
+The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
+This allows the design of software agents which can interact and learn in that environment without knowing anything about its internal details.
+The `gym` python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
+The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see below, 'Deep Reinforcement Learning Frameworks').
+While this library of environments span a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
+We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any Reinforcement Learning framework.
+
+## Designing and training an agent
+
+The training an agent on an environment involves allowing the agent to experience and respond to repeated simulations from the environment, thousands or even millions of times. 
+Over the course of training, weights on the neural network are updated in response to the reward given by the environment, much like reinforcement learning is believed to work in living creatures [@cite].
+A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.
+A typical neural network in RL may have thousands or hundreds of thousands of parameters, giving it the flexibility to learn very complex behaviors.
+The overall architecture of the neural network, as well as other parameters such as those governing how an agent learns, how an agent proposes actions, and how the agent discounts future rewards, and so forth are controlled by *hyperparameters*.
+To initialize a new, untrained agent, we must specify a handful or so of these hyperparameters, which are held fixed over the course of training.
+The particular selection of hyperparameters required depends on which *RL algorithm* our agent will be based.
+The RL algorithm specifies precisely how the agent will learn from its environment:
+Is the algorithm model-based or model-free? 
+Stochastic or deterministic? 
+"Off-policy" or "on-policy"? 
+Does it use a random noise model to propose actions?
+Does it permit a continuous action space or require discrete actions?
+The design of RL algorithms has been and remains an area of intense research and rapid development [Table 1].
+
+The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
+The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
+Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
+Considerations in environmental design, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.
 
@@
 We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.
-
-@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both poblems.
 

### 2021-04-30 manuscript/manuscript.Rmd
WIP
WIP outline

@@
 
-Deep Reinforcement Learning (RL), a machine learning technique in which a *software agent* is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
-Unlike more widely recognized machine learning techniques for either *supervised* learning (such as image classification, [@species_ex], [@satellite_ex]) or unsupervised learning, RL focuses on solving *dynamic* problems which require planning ahead, often in uncertain and changing environments.
-Of all machine learning techniques, RL most resembles a general artificial intelligence.
-Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero], and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex].
-These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.
+<!-- 
+- Global change problems: Important but tough decisions
+- ML: exciting stuff, but only fitting data is driving through the rearview mirror
+- RL is the ML for decisions (in dynamic/uncertain environments) 
+- RL can use existing knowledge and models to improve decisions in complex/uncertain scenarios (read: global change biology)
+- Here be dragons / map paragraph
 
-<!-- Millie help on this para? -->
+-->
 
-If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
-The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
-The endangered species act ... Ecosystem services ....
-Such simplicity is in no small part due to the challenge associated in determining a Some reference to [@TomDietrich] on potential for CS-based application.
+# RL overview
 
-In contrast to other ML methods, RL does not require massive data sets for training, relying instead on model simulations based on the best available ecological data and knowledge of underlying processes.
-Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains sparse and unevenly sampled [CITE].
-Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
-Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
-Methods from *optimal control theory* have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
-However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
-Deep reinforcement learning offers a way forward.
-Deep RL agents are trained not with mountains of data, but by interacting with simulators.
-Most recent advances in Deep RL have illustrated algorithms trained by playing arcade game simulators such as Atari BreakOut, or physics-based simulations of robotics movement [@TD3; @SAC; @PPO; @A2C].
-Simulating from complex models is far simpler than solving them.
-The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations.
-Given sufficient computational power, RL can be trained and evaluated across a broad range of possible realistic scenarios well beyond the reach of classical methods.
-In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
+<!-- Conceptual picture: (A) an "environment", and (B) an "agent"
 
-If RL has great promise, the approach also imposes great risk.
-Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].
-Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems
-. RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes
-. Technical issues are not the only pitfall of AI applications to conservation problems
-.
-Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property
-.
+What do we mean by an "environment"?  
+  - Transition function
+  - "reward"
+  - examples
 
-In this paper, we introduce the fundamentals of deep reinforcement learning and illustrate with simple examples how these approaches can be applied to important conservation decision-making problems, from over-harvesting in global fisheries to ecosystem tipping points to wildfire dynamics.
-These examples span a range from problems for which an optimal analytic solution is already known, to problems not previously solved with classical methods, to problems not yet solved by RL.
-We include a detailed appendix with complete examples in both python and R for training RL algorithms, as well as three stand-alone module "gyms" implementing the each of the conservation problems presented here using a standard benchmark [@gym], which we believe can facilitate future contributions both from researchers in RL searching for interesting and important open problems as well as ecologists who may seek to make those simulations more realistic and precise to specific application areas.
-We hope this paper serves as a port of entry to researchers and students from either field seeking to engage in work on RL applications to conservation.
-We also hope these examples serve as illustration of both the potential and perils of such applications.
+What do we mean by an "agent"?
+  Components of an agent: 
+    - *neural network*, whose *parameters* are estimated by *training* +
+    - an RL algorithm, whose *hyperparameters* (including NN structure) are estimated by *tuning*
+    - High-level overview of an RL algorithm components (model/model-free, on/off policy, stochastic/deterministic, etc).
+
+  conceptual explanation of 'model-free', what it means to have a "policy" without modeling/predicting transitions
+
+-->
 
@@
 
--   model (underlying ecological model)
+-   environment
 
@@
 
--   environment
+-   neural network
 
--   predict
+-   train
+
+-   tune
+
+-   parameters
+
+-   hyper-parameters
+
+-   model (underlying ecological model used to drive step() method in environment)
+
+-   model (SB3 term  for "agent")
+
+-   predict (SB3 term to return agent's action given an observation)
+
+-   predict (ecological model which generates probabilistic forecast of future states)
 
 -   policy
+
+-   reward (utility, value)
+
+-   policy function
+
+-   value function
 
@@
 -   model-based
-
-# Sequential Decision Problems and Reinforcement Learning
-
-<!-- meh, not sure we need this? -->
-
-Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to as the reward or utility).
-Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
-In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
-Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
-Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
-Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.
-
-This framework raises two conceptual challenges.
-First, sequential decision-making problem is distinct from the problem of model estimation.
-A sequential decision problem will typically treat the model as given, typically with some uncertainty around the process and possibly the measurmeents and model estimates as well.
-Thus, methods for solving sequential decision-making should not be mistaken as alternatives to methods designed for model estimation, such as regression, Bayesian heirarchical modeling, generalized additive models (GAMs) etc.
-Second, methods for describing and solving sequential decision-making problems are often given different terms by different communities to describe the same thing.
-Sequential decision problems are common to behavioral ecology [@Mangel], conservation [@Marescot2013] natural resource economics [@Clark1990], and some engineering fields [] where they are known as "optimal control" problems, while computer science, neuroscience, and robotics typically refer to them as 'reinfocement learning'.
 
@@
 RL algorithms can generally be split into 2 groups, model-based and model-free RL.
-In model-based RL, the agent is attempting to learn the state transition operator, $T$.
-Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?]
-.
+In model-based RL, the agent is attempting to learn the state transition operator, $T$. <!-- Sometimes the agent is just assuming this is given, right? -->
+Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
+<!-- For most recent methods focus on an approximate solution, since optimal is computationally impossible.
+In contrast most readers will assume this step is trivial, if the transition matrix is known there's nothing more to do but "look" at the predicted transitions and make a choice.
+-->
 In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
@@
 <!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
+
+<!-- Can we add citations to each please -->
 

### 2021-04-30 manuscript/presubmission-letter.Rmd
WIP outline

--- 2021-04-28 manuscript/presubmission-letter.Rmd
@@
 We believe that similar techniques could improve decision-making in managing ecological environments as well. 
-<!-- Having trouble reconciling lines 21-23, trying to say that these simulators are like the grand challenges and given RL's track record of solving intractable decision-making problems, perhaps RL could shed light here?
-I think having a sentence asserting that RL would use ecological models not replace them would be good somewhere in here.-->
 Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
 Detailed, process-based models allow a window into the future scenarios of ecological management.
+Even simple processed-based ecological models can provide important insight into future scenarios that no amount of historical data would reveal.
+Rather than displacing existing ecological models and knowledge, we illustrate how RL-based approaches can allow us to leverage such understanding to better explore and guide complex decision-making in changing environments.
+Whereas previous decision techniques have historically constrained the complexity and realism of practical modeling efforts, RL may open the door to the design of ever more realistic assumptions. 
 Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.

### 2021-05-01 manuscript/manuscript.Rmd
comments / notes mostly

@@
 What do we mean by an "agent"?
+
+  Model-based vs Model free.  We focus on Model free agents which use Neural networks, a subset of deep RL
+
   Components of an agent: 
@@
 The task for the agent is to learn a *policy*, $\pi$, that prescribes what action the agent should perform after observing a certain state.
-The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.
+The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continually updates its policy in order to maximize its expected cumulative sum of rewards.
 
+<!-- Still can't decide if this statement is too strong; i.e. that at least theoretically / LSTM, RL permits learning processes which are not Markovian in the observed variables -->
 When the environment is fully observable, i.e.
@@
 Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high reward trajectory.
-There are 3 value functions that are commonly used across RL algorithms: the state value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
+There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
 The state-value function is defined as the expected return from a given state,
@@
 And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succintly defined as $$
+The advantage function can then be succinctly defined as
+
+$$
   A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-$$ Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+$$ 
+Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
 Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
@@
 The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
-.
 
 A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
-If an agent uses the same policy to take actions as the policy that the agent attempts to evaulate or improve, then this algorithm is on-policy.
+If an agent uses the same policy to take actions as the policy that the agent attempts to evaluate or improve, then this algorithm is on-policy.
+<!-- Unclear what "to evaluate or improve" means.
+Perhaps an example would help?
+-->
 If an agent uses a different policy to select actions than the policy that the agent evaluates or improves, then this algorithm is off-policy.
 The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [cite SAC paper].
-Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop].
+<!-- Not clear what this means -->
+Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop]. <!-- not clear how "stable" is defined/measured -->
 
@@
 In every reinforcement learning algorithm, the agent is attempting to learn some function, whether it is a policy function and/or a value function.
-Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions which we do not know the form of.
-Additionally, the ability of neural networks to autonomously devise low-dimensional representatoins of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
+Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions whose form is unknown.
+Additionally, the ability of neural networks to autonomously devise low-dimensional representations of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
+<!-- Yes!
+should stress why that matters -->
 The typical process for using neural networks in RL is that the agent is attempting to optimize a policy and/or value network using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
+<!-- Fixme: needs to be clear that NN parameters are being fit by the "training" step.
+No point mentioning back-propagation, evolutionary algorithms, or gradient methdos if we are not going to define them.
+-->
+
 Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable with classical methods.
+<!-- non-tabular undefined -->
 
 In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process.
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
 For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
@@
 
-To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on [@TD3] <!-- Kari: are these different algorithms?-->.
+We will illustrate the potential differences between two RL algorithms used to train the agents, one based on the "Advantage Actor Critic" algorithm, usually referred to as A2C [@A2C], and one based on Twin Delayed Deep Deterministic Policy Gradient, commonly known as TD3 [@TD3] (a generalization of Deep Deterministic Policy Gradient, or DDPG, which extends the DQN algorithm @DQN to a continuous action space using the deterministic policy gradient).
+
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.

### 2021-05-03 manuscript/appendix.Rmd
WIP outline
:pen: :construction: :pen:

--- 2021-04-29 manuscript/appendix.Rmd
@@
 ```
-
-# Overview
-
-Development of any deep reinforcement learning application can be divided into two components: an _environment_ and an _agent_.
-An environment is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
-An agent is typically a computational neural network trained using an appropriate algorithm to suggest an action based on available information about the environment.
-
-Somewhat confusingly, frameworks will often refer to a trained agent as a "model", while also referring to the approaches used to implement and train the agent as "model free" because the agent does not attempt to form an internal model of the environmental process.
-We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.
-
-@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both poblems.
-
-A rapidly growing set of material for learning about deep reinforcement learning is now freely available.
-
-# Deep Reinforcement Learning Frameworks
-
-At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
-While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
-Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
-Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3].
-Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
-Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
-However, higher-level interfaces and high-level documentation are still relatively limited.
-OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
-OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
-Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption.
-Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
-The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
-Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
 

### 2021-05-03 manuscript/manuscript.Rmd
:pen: :construction: :pen:

@@
 
+<!-- Environmental problems, ML -->
+Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
+But will ML approaches really help us tackle a changing planet?
+Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g.
+image recognition) and *unsupervised learning* (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
+Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:
+
+1) RL is explicitly focused on the task of selecting actions in an uncertain and changing environment to maximize some objective,
+2) RL does not require massive amounts of representative sampled historical data,
+3) RL approaches easily integrate with existing ecological models and simulations, which may be our best guide to understanding and predicting future possibilities.
+
+Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
+To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
+RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
+Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
+Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
+Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
+Even simpler models are frequently employed throughout ecology to guide understanding and response.
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
+We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
+We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
+
+In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep Reinforcement Learning techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
+
+The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
+Some familiarity with the design and potential application of RL is important not only to researchers interested in developing these methods to solve new problems, but is also important to any ecologist or manager
+We discuss of how an open, transparent and reproducible approach can help mitigate some concerns, while also offering a more effective interface between teams of researchers from both ecological and computer sciences.
+We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+We further include implementations of three fully featured simulation modules following current leading standards, which would allow engineers and computer scientists already working in RL to test their agents against both solved and unsolved problems in conservation management.
+We believe this piece would provide an effective introduction to the concepts and practices of RL which would allow ecologists to apply, extend, and critique such methods.
+
+Classic approaches require
+
+Reinforcement learning works much like human learning -- the basic techniques of RL have been used to train animals to perform complex tasks long before they were used to train computational neural networks [@behavioral_psych].
+Ecological decision-making has its own rich and rapidly expanding suite of complex simulators:
+SORTIE in forest ecosystems [@sortie], EcoPath with EcoSim in marine systems [@ecopath], GCM models in climate [@gcm], 
+Detailed, process-based models allow a window into the future scenarios of ecological management.
+Even simple ecological models can provide important insight into future scenarios that no amount of historical data would reveal.
+Rather than displacing existing ecological models and knowledge, we illustrate how RL-based approaches can allow us to leverage such understanding to better explore and guide complex decision-making in changing environments.
+Whereas previous decision techniques have historically constrained the complexity and realism of practical modeling efforts, RL may open the door to the design of ever more realistic assumptions.
+Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.
+Although our examples are primarily a proof-of-principle, they serve to illuminate both the promise as well as potential pitfalls.
+
 # RL overview
@@
 -->
+
+Development of any deep reinforcement learning application can be divided into two components: an *environment* and an *agent*.
+An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
+An *agent* is typically a computational neural network based on an appropriate RL algorithm that may have been 'trained' on one or more environments.
+
+## Designing and building an environment
+
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
+In principle, there are virtually no restrictions on what comprises a state: it may be continuous or discrete, completely or only partially observed, one or many dimensions.
+For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
+The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
+The reward will in general depend on both the current state and choice of actions.
+An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
+In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
+In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+
+@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
+The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
+This allows the design of software agents which can interact and learn in that environment without knowing anything about its internal details.
+The `gym` python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
+The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see below, 'Deep Reinforcement Learning Frameworks').
+While this library of environments span a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
+We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any Reinforcement Learning framework.
+
+## Designing and training an agent
+
+The training an agent on an environment involves allowing the agent to experience and respond to repeated simulations from the environment, thousands or even millions of times. 
+Over the course of training, weights on the neural network are updated in response to the reward given by the environment, much like reinforcement learning is believed to work in living creatures [@cite].
+A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.
+A typical neural network in RL may have thousands or hundreds of thousands of parameters, giving it the flexibility to learn very complex behaviors.
+The overall architecture of the neural network, as well as other parameters such as those governing how an agent learns, how an agent proposes actions, and how the agent discounts future rewards, and so forth are controlled by *hyperparameters*.
+To initialize a new, untrained agent, we must specify a handful or so of these hyperparameters, which are held fixed over the course of training.
+The particular selection of hyperparameters required depends on which *RL algorithm* our agent will be based.
+The RL algorithm specifies precisely how the agent will learn from its environment:
+Is the algorithm model-based or model-free? 
+Stochastic or deterministic? 
+"Off-policy" or "on-policy"? 
+Does it use a random noise model to propose actions?
+Does it permit a continuous action space or require discrete actions?
+The design of RL algorithms has been and remains an area of intense research and rapid development [Table 1].
+
+The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
+The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
+Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
+Considerations in environmental design, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.
+
+Somewhat confusingly, frameworks will often refer to a trained agent as a "model", while also referring to the approaches used to implement and train the agent as "model free" because the agent does not attempt to form an internal model of the environmental process.
+We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.
+
+A rapidly growing set of material for learning about deep reinforcement learning is now freely available.
+
+# Deep Reinforcement Learning Frameworks
+
+At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
+While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
+Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
+Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3].
+Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
+Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
+However, higher-level interfaces and high-level documentation are still relatively limited.
+OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
+OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
+Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption.
+Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
+The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
+Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
 

### 2021-05-03 manuscript/appendix.Rmd
:pen:

--- 2021-04-29 manuscript/appendix.Rmd
@@
 
-# Overview
+# Building a simulation environment `gym`
 
-Development of any deep reinforcement learning application can be divided into two components: an _environment_ and an _agent_.
-An environment is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
-An agent is typically a computational neural network trained using an appropriate algorithm to suggest an action based on available information about the environment.
-
-Somewhat confusingly, frameworks will often refer to a trained agent as a "model", while also referring to the approaches used to implement and train the agent as "model free" because the agent does not attempt to form an internal model of the environmental process.
-We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.
-
-@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both poblems.
-
-A rapidly growing set of material for learning about deep reinforcement learning is now freely available.
+[@openai_gym]
 
@@
 Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
-
-# Building a simulation environment `gym`
-
-[@openai_gym]
 
@@
 np$random$seed(seed)
+
 # Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
@@
 ```
+
+<!-- is this too verbose? --> 
+The above code also illustrates a few conventions which may be helpful to bear in mind when using the `reticulate` interface to interact with Python from R: 
+it is often necessary for integer values to be explicitly typed as integers by adding a trailing `L` (corresponding to the primitive C type `long` integer).
+Python is also more strongly object-oriented than many R packages, where "methods" of an "object" are accessed with the list-subset operator `$` in R (equivalent to the use of `.` in Python).
+Lastly, while R can use `<-` or `=` for assignment, Python uses only `=`. For simplicity we will stick with the latter.  
+With few exceptions, the R code shown here can be re-written as python code by dropping the `L` and replacing `$` with `.`, see stand-alone python code in the `python/` subdirectory of the repository.
 
@@
 We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
-
-<!-- Kari: a brief explaination of the gym$function syntax would probably be helpful - is that a reticulate thing? Just reading later it looks like you reference it as a "method"?-->
 

### 2021-05-03 manuscript/manuscript.Rmd
:pen:

@@
 
-In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep Reinforcement Learning techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
+In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
+Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
+Our second example considers ecosystem management under slow environmental degradation: a scenario which is not amenable to classic optimization methods such as dynamic programming.
+In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
+In our third example simulating wildfire spread highlights an open problem which we expect will be considerably harder to find effective solutions.
+These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
+Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
+We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+We further include implementations of three fully featured simulation modules following current standards for RL benchmark problems as Python modules distributed on PyPi and GitHub.
+Each module includes several more challenging variations to the examples described here.
+We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.
 
 The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
-Some familiarity with the design and potential application of RL is important not only to researchers interested in developing these methods to solve new problems, but is also important to any ecologist or manager
-We discuss of how an open, transparent and reproducible approach can help mitigate some concerns, while also offering a more effective interface between teams of researchers from both ecological and computer sciences.
-We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
-We further include implementations of three fully featured simulation modules following current leading standards, which would allow engineers and computer scientists already working in RL to test their agents against both solved and unsolved problems in conservation management.
-We believe this piece would provide an effective introduction to the concepts and practices of RL which would allow ecologists to apply, extend, and critique such methods.
-
-Classic approaches require
-
-Reinforcement learning works much like human learning -- the basic techniques of RL have been used to train animals to perform complex tasks long before they were used to train computational neural networks [@behavioral_psych].
-Ecological decision-making has its own rich and rapidly expanding suite of complex simulators:
-SORTIE in forest ecosystems [@sortie], EcoPath with EcoSim in marine systems [@ecopath], GCM models in climate [@gcm], 
-Detailed, process-based models allow a window into the future scenarios of ecological management.
-Even simple ecological models can provide important insight into future scenarios that no amount of historical data would reveal.
-Rather than displacing existing ecological models and knowledge, we illustrate how RL-based approaches can allow us to leverage such understanding to better explore and guide complex decision-making in changing environments.
-Whereas previous decision techniques have historically constrained the complexity and realism of practical modeling efforts, RL may open the door to the design of ever more realistic assumptions.
-Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.
-Although our examples are primarily a proof-of-principle, they serve to illuminate both the promise as well as potential pitfalls.
+We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
+We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
 
@@
 
-Development of any deep reinforcement learning application can be divided into two components: an *environment* and an *agent*.
+Any application of RL can be divided into two components: an *environment* and an *agent*.
 An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
@@
 The `gym` python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
-The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see below, 'Deep Reinforcement Learning Frameworks').
+The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').
 While this library of environments span a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
-We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any Reinforcement Learning framework.
+We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
 
@@
 
-Somewhat confusingly, frameworks will often refer to a trained agent as a "model", while also referring to the approaches used to implement and train the agent as "model free" because the agent does not attempt to form an internal model of the environmental process.
-We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.
+<!--
+A rapidly growing set of material for learning about deep reinforcement learning is now freely available.... (maybe just describe this in the appendix)
+-->
 
-A rapidly growing set of material for learning about deep reinforcement learning is now freely available.
-
-# Deep Reinforcement Learning Frameworks
-
-At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
-While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
-Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
-Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3].
-Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
-Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
-However, higher-level interfaces and high-level documentation are still relatively limited.
-OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
-OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
-Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption.
-Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
-The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
-Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
-
-<!-- Kari: We reference Markov Decision processes as an alternative approach in the tipping points example and just generally elsewhere so it might be helpful to talk about them in the introduction-->
-
-#### Key Terminology
-
--   environment
-
--   agent
-
--   trained agent (model)
-
--   neural network
-
--   train
-
--   tune
-
--   parameters
-
--   hyper-parameters
-
--   model (underlying ecological model used to drive step() method in environment)
-
--   model (SB3 term  for "agent")
-
--   predict (SB3 term to return agent's action given an observation)
-
--   predict (ecological model which generates probabilistic forecast of future states)
-
--   policy
-
--   reward (utility, value)
-
--   policy function
-
--   value function
-
--   model-free
-
--   model-based
+<!-- 
+Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
+Make a nice table box in latex instead if we're keeping this
+-->
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{table1.pdf}
+\end{figure}
 
@@
 
-Before providing some examples on how deep reinforcement learning can be applied to ecological and conservation problems, we will provide a brief, general introduction to RL.
 The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
@@
 The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
-Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [\@].
+Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
 

### 2021-05-05 manuscript/manuscript.Rmd
Initial edits, thinking about how to frame mdps

--- 2021-05-01 manuscript/manuscript.Rmd
@@
 
-<!-- 
-- Global change problems: Important but tough decisions
-- ML: exciting stuff, but only fitting data is driving through the rearview mirror
-- RL is the ML for decisions (in dynamic/uncertain environments) 
-- RL can use existing knowledge and models to improve decisions in complex/uncertain scenarios (read: global change biology)
-- Here be dragons / map paragraph
+Reinforcement Learning (RL), a machine learning technique in which a *software agent* is trained to select actions to maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
+Unlike the more widely recognized machine learning areas of *supervised* learning and *unsupervised* learning, RL focuses on solving *dynamic* problems which often require planning ahead in uncertain and changing environments.
+Of all machine learning techniques, RL most resembles a general artificial intelligence.
+Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero], and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex].
+These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.
 
--->
+<!-- Millie help on this para? -->
 
-# RL overview
+If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
+The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
+The endangered species act ... Ecosystem services ....
+Such simplicity is in no small part due to the challenge associated in determining a Some reference to [@TomDietrich] on potential for CS-based application.
 
-<!-- Conceptual picture: (A) an "environment", and (B) an "agent"
+In contrast to other ML methods, RL does not require massive data sets for training and can instead rely on simulators.
+Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains sparse and unevenly sampled [CITE].
+Moreover, no amount of data can train supervised or unsupervised ML approaches to predict a future which has no analog in the historical record.<!-- ML: potential counterpoint here of using synthetic data to train SL/USL algorithms-->
+Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
+Methods from *optimal control theory* have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
+However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
+*Deep reinforcement learning*, which applies recent advancements from deep learning to reinforcement learning, has been able to solve previously intractable, high dimensional sequential decision making problems.
+A strength of deep RL methods is that, given sufficient computational power, they can be easily applied to any decision-making problem that can be formulated as a *Markov-decision process* (MDP).
+A Markov-decision process provides a standard ... (working on this)
+The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations .
+In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.
 
-What do we mean by an "environment"?  
-  - Transition function
-  - "reward"
-  - examples
+If RL has great promise, the approach also imposes great risk.
+Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2019].
+Despite recent progress, Deep RL is a young field, and stability and robustness of its methods remain largely open problems [@henderson2019].
 
-What do we mean by an "agent"?
+. RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes
+. Technical issues are not the only pitfall of AI applications to conservation problems
+.
+Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property
+.
 
-  Model-based vs Model free.  We focus on Model free agents which use Neural networks, a subset of deep RL
-
-  Components of an agent: 
-    - *neural network*, whose *parameters* are estimated by *training* +
-    - an RL algorithm, whose *hyperparameters* (including NN structure) are estimated by *tuning*
-    - High-level overview of an RL algorithm components (model/model-free, on/off policy, stochastic/deterministic, etc).
-
-  conceptual explanation of 'model-free', what it means to have a "policy" without modeling/predicting transitions
-
--->
+In this paper, we introduce the fundamentals of deep reinforcement learning and illustrate with simple examples how these approaches can be applied to important conservation decision-making problems, from over-harvesting in global fisheries to ecosystem tipping points to wildfire dynamics.
+These examples span a range from problems for which an optimal analytic solution is already known, to problems not previously solved with classical methods, to problems not yet solved by RL.
+We include a detailed appendix with complete examples in both python and R for training RL algorithms, as well as three stand-alone module "gyms" implementing the each of the conservation problems presented here using a standard benchmark [@gym], which we believe can facilitate future contributions both from researchers in RL searching for interesting and important open problems as well as ecologists who may seek to make those simulations more realistic and precise to specific application areas.
+We hope this paper serves as a port of entry to researchers and students from either field seeking to engage in work on RL applications to conservation.
+We also hope these examples serve as illustration of both the potential and perils of such applications.
 
@@
 
--   environment
+-   model (underlying ecological model)
 
@@
 
--   neural network
+-   environment
 
--   train
-
--   tune
-
--   parameters
-
--   hyper-parameters
-
--   model (underlying ecological model used to drive step() method in environment)
-
--   model (SB3 term  for "agent")
-
--   predict (SB3 term to return agent's action given an observation)
-
--   predict (ecological model which generates probabilistic forecast of future states)
+-   predict
 
 -   policy
-
--   reward (utility, value)
-
--   policy function
-
--   value function
 
@@
 -   model-based
+
+# Sequential Decision Problems and Reinforcement Learning
+
+<!-- meh, not sure we need this? -->
+
+Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to as the reward or utility).
+Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
+In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
+Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
+Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
+Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.
+
+This framework raises two conceptual challenges.
+First, sequential decision-making problem is distinct from the problem of model estimation.
+A sequential decision problem will typically treat the model as given, typically with some uncertainty around the process and possibly the measurmeents and model estimates as well.
+Thus, methods for solving sequential decision-making should not be mistaken as alternatives to methods designed for model estimation, such as regression, Bayesian heirarchical modeling, generalized additive models (GAMs) etc.
+Second, methods for describing and solving sequential decision-making problems are often given different terms by different communities to describe the same thing.
+Sequential decision problems are common to behavioral ecology [@Mangel], conservation [@Marescot2013] natural resource economics [@Clark1990], and some engineering fields [] where they are known as "optimal control" problems, while computer science, neuroscience, and robotics typically refer to them as 'reinfocement learning'.
 
@@
 The task for the agent is to learn a *policy*, $\pi$, that prescribes what action the agent should perform after observing a certain state.
-The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continually updates its policy in order to maximize its expected cumulative sum of rewards.
+The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.
 
-<!-- Still can't decide if this statement is too strong; i.e. that at least theoretically / LSTM, RL permits learning processes which are not Markovian in the observed variables -->
 When the environment is fully observable, i.e.
@@
 Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
-There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high reward trajectory.
+There are 3 value functions that are commonly used across RL algorithms: the state value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
 The state-value function is defined as the expected return from a given state,
@@
 
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, $$
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, 
+$$
   Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
@@
 And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succinctly defined as
-
+The advantage function can then be succintly defined as 
 $$
@@
 Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
-Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
+Learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
 
@@
 RL algorithms can generally be split into 2 groups, model-based and model-free RL.
-In model-based RL, the agent is attempting to learn the state transition operator, $T$. <!-- Sometimes the agent is just assuming this is given, right? -->
-Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
-<!-- For most recent methods focus on an approximate solution, since optimal is computationally impossible.
-In contrast most readers will assume this step is trivial, if the transition matrix is known there's nothing more to do but "look" at the predicted transitions and make a choice.
--->
-In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
+In model-based RL, the agent is attempting to learn the state transition operator, $T$.
+Once the transition model is known, the agent can then use the transition model to plan out what the optimal action is, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?]
+.
+In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both.
 The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
+.
 
 A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
-If an agent uses the same policy to take actions as the policy that the agent attempts to evaluate or improve, then this algorithm is on-policy.
-<!-- Unclear what "to evaluate or improve" means.
-Perhaps an example would help?
--->
+If an agent uses the same policy to take actions as the policy that the agent attempts to evaulate or improve, then this algorithm is on-policy.
 If an agent uses a different policy to select actions than the policy that the agent evaluates or improves, then this algorithm is off-policy.
-The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [cite SAC paper].
-<!-- Not clear what this means -->
-Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop]. <!-- not clear how "stable" is defined/measured -->
+The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [@SAC].
+Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop].
 
@@
 In every reinforcement learning algorithm, the agent is attempting to learn some function, whether it is a policy function and/or a value function.
-Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions whose form is unknown.
+Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in RL since neural networks can be used to approximate policy and/or value functions which we do not know the form of.
 Additionally, the ability of neural networks to autonomously devise low-dimensional representations of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
-<!-- Yes!
-should stress why that matters -->
 The typical process for using neural networks in RL is that the agent is attempting to optimize a policy and/or value network using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
-<!-- Fixme: needs to be clear that NN parameters are being fit by the "training" step.
-No point mentioning back-propagation, evolutionary algorithms, or gradient methdos if we are not going to define them.
--->
-
 Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable with classical methods.
-<!-- non-tabular undefined -->
 
 In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters.
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process.
 For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
@@
 \hline
-MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
+MBPO [@MBPO] & Model-based Policy Optimization & Model-based & On-policy& &\\
 \hline
@@
 \hline
-A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
+A2C  [@A2C] & Advantage Actor Critic & Model-free & On-policy& &\\
 \hline
-A3C  & Asynchronous A2C & Model-free & On-policy & &\\
+A3C [@A3C] & Asynchronous A2C & Model-free & On-policy & &\\
 \hline
-TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
+TRPO [@TRPO] & Trust Region Policy Optimization & Model-free & On-policy & &\\
 \hline
-PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
+PPO [@PPO] & Proximal Policy Optimization & Model-free & On-policy & &\\
 \hline
-DQN & Deep Q Networks & Model-free & Off-policy& &\\
+DQN [@DQN] & Deep Q Networks & Model-free & Off-policy& &\\
 \hline
-DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
+DDPG [@DDPG] & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
 \hline
-TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
+TD3 [@TD3] & Twin Delayed DDPG & Model-free & Off-policy & &\\
 \hline
-SAC & Soft Actor Critic & Model-free & Off-policy & &\\
+SAC [@SAC] & Soft Actor Critic & Model-free & Off-policy & &\\
 \hline
-IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
+IMPALA [@IMPALA] & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
 \hline
-APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
+APE-X [@APE-X] & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
 \hline
@@
 
-We will illustrate the potential differences between two RL algorithms used to train the agents, one based on the "Advantage Actor Critic" algorithm, usually referred to as A2C [@A2C], and one based on Twin Delayed Deep Deterministic Policy Gradient, commonly known as TD3 [@TD3] (a generalization of Deep Deterministic Policy Gradient, or DDPG, which extends the DQN algorithm @DQN to a continuous action space using the deterministic policy gradient).
-
+To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on [@TD3] <!-- Kari: are these different algorithms?-->.
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.

### 2021-05-05 manuscript/manuscript.Rmd
some minor comments on intro

--- 2021-05-03 manuscript/manuscript.Rmd
@@
 Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
+
+<!-- citation suggestion for declining biodiversity  -->
+
+<!-- @article{dirzo2014defaunation, -->
+<!--   title={Defaunation in the Anthropocene}, -->
+<!--   author={Dirzo, Rodolfo and Young, Hillary S and Galetti, Mauro and Ceballos, Gerardo and Isaac, Nick JB and Collen, Ben}, -->
+<!--   journal={science}, -->
+<!--   volume={345}, -->
+<!--   number={6195}, -->
+<!--   pages={401--406}, -->
+<!--   year={2014}, -->
+<!--   publisher={American Association for the Advancement of Science} -->
+<!-- } -->
+
 But will ML approaches really help us tackle a changing planet?
+
+<!-- suggestion: "...tackle the most pressing environmental challenges on a changing planet" -->
+
 Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g.
@@
 
+<!-- really like this list! super clear and helpful! -->
+
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
+
+<!-- another widely used model is ED2 (ecosystem demography model) - I think more common than SORITE? https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2008JG000812 -->
+
 Even simpler models are frequently employed throughout ecology to guide understanding and response.
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+
+<!-- I think this is missing the connection between the line 88 examples and needing more sophisticated methods for decision making below.
+i.e.
+"As we develop more complex process based models (such as Ecosim and ED2), methods with the capacity to effectively integrate high dimensional models into decision making processes will be critical to ensuring their policy relevance" (or something like that, then go on to NN ML)-->
+
 Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
+
+<!-- ^ awesome! -->
+
 Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
+
 Our second example considers ecosystem management under slow environmental degradation: a scenario which is not amenable to classic optimization methods such as dynamic programming.
@@
 In our third example simulating wildfire spread highlights an open problem which we expect will be considerably harder to find effective solutions.
+
+ <!-- I think this is pretty specific (almost like a discussion paragraph, but we haven't read the paper yet.) Could you be more broad about what you set out to do in the paper?
+i.e.: "We start with a simple example from fisheries management that has a known optimal solution, proving that with no prior knowledge of underlying model dynamics RL agents can find a close to optimal solution.
+We go on to show that RL can outperform rule-of-thumb solutions to management scenarios not amenable to classic optimization methods such as dynamic programming.
+Our examples demonstrate that..."  -->
+
 These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
@@
 If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+
+<!-- maybe cite Caleb's paper here? -->
+
 Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
+
+<!-- these last few sentences almost feel like discussion points. I'm torn though, because it it nice to have them in the intro-->
 
@@
 -->
+
+<!-- something to introduce this section would be helpful for transition: "We start with a brief overview of RL problems and solutions:" -->
 

### 2021-05-05 manuscript/manuscript.Rmd
Passed over intro

--- 2021-05-03 manuscript/manuscript.Rmd
@@
 But will ML approaches really help us tackle a changing planet?
-Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g.
-image recognition) and *unsupervised learning* (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
+Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
 Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:
@@
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
-To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
-RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
-Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
-Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
+To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
+Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
+RL agents develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is a common practice.
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
-Even simpler models are frequently employed throughout ecology to guide understanding and response.
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
-Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
-We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
+Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
+<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
+We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
 We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
@@
 An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
-An *agent* is typically a computational neural network based on an appropriate RL algorithm that may have been 'trained' on one or more environments.
+An *agent* is typically a computer algorithm which takes sequential observations from the environment and returns proposed actions.
+RL agents typically use one or more neural networks to represent the relationship between observed states and suitable actions, and 'train' these networks through repeated interactions with the environment to maximize the culumulative reward.
+<!-- Blah blah skip this?-->
+It is important to recognize that these definitions are significantly more general than requirements typically imposed by classical methods.  
+For instance, there is no a priori assumption that the environment is *fully observed*, or that the environment is *Markovian* (future states depend only on the current state), commonly assumed by classical methods [e.g.
+@Marescot2013].
+Just as important is the high degree of interoperability provided by the RL framework.
+As we shall see, an RL algorithm developed to play Atari games can instead be trained on an environment that simulates fish population dynamics without modification.
+(Note that this does not refer to the *trained* agent).
+This not guarantee that such an algorithm will be successful (even after extensive tuning and training in the fish population dynamics environment), but does open the door to more rapid exchange between fields.
+In particular, the adoption of standard interfaces for RL environments [@brockman2016] as well as the implementation of cutting edge algorithms in flexible open source frameworks [@sb3; @tf-agents] means that a researcher can test a range of RL algorithms against a single environment, or a given algorithm across a range of environments with equal ease.
 
@@
 An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
-In principle, there are virtually no restrictions on what comprises a state: it may be continuous or discrete, completely or only partially observed, one or many dimensions.
+In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
+The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
 For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
@@
 In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\linewidth]{fig1.pdf}
+\caption{Reinforcement Learning focuses on the repeated feedback of agent interacting with an \emph{environment}.
+The agent performs an action which influences the environment's future \emph{state}.
+The agent receives an observation of the new state and reward from the environment.}
+\end{figure}
 
@@
 Does it permit a continuous action space or require discrete actions?
-The design of RL algorithms has been and remains an area of intense research and rapid development [Table 1].
+The design of RL algorithms has been and remains an area of intense research and rapid development [Table 2].
+Open source implementations by professional software development teams often follow swiftly after new algorithms are published [@Ssb3; @tf-agents].
+In the following section, we summarize some of the key concepts that make up an RL algorithm, and highlight relevant distinctions between important algorithms today.
 
+<!-- Something about given RL algorithm being able to solve lots of different environments with no pre-specified knowledge. cite examples @AlphaZero and @DQN; but still depends on hyperparameters -->
 The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
@@
 Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
-Considerations in environmental design, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.
+Considerations in how the environment simulation is implemented, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.
 
@@
 
-# RL essentials
+# RL Algorithms
+
+<!--
+
+I think we can lean more heavily both in references to and paralleling the presentation of some of the existing background material.
+
+-->
 
@@
 
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, $$
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
+
+$$
   Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
@@
 <!-- Can we add citations to each please -->
+<!--
 
+Should we mention:
+
+- Temporal Differencing, RIENFORCE
+- SARSA, 
+- Double DQN?  
+Can we add HER (since delayed reward case may be particularly interesting)
+Should we do anything to better reflect the 'family tree' here? (e.g. DQN -> DDPG -> TD3)
+
+-->
 ```{=latex}

### 2021-05-05 manuscript/manuscript.Rmd
Rewriting a transition sentence

@@
 Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
-RL agents develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is a common practice.
+RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.

### 2021-05-05 manuscript/manuscript.Rmd
:pen:

--- 2021-05-05 manuscript/manuscript.Rmd
@@
 But will ML approaches really help us tackle a changing planet?
-Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
+Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g.
+image recognition) and *unsupervised learning* (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
 Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:
@@
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
-To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
-Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
-RL agents develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is a common practice.
+To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
+RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
+Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
+Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
-Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
+Even simpler models are frequently employed throughout ecology to guide understanding and response.
 <!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
-Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
-We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
+We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
 We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.5\linewidth]{fig1.pdf}
-\caption{Reinforcement Learning focuses on the repeated feedback of agent interacting with an \emph{environment}.
-The agent performs an action which influences the environment's future \emph{state}.
-The agent receives an observation of the new state and reward from the environment.}
-\end{figure}
-
 @brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
@@
 
-<!--
-
-I think we can lean more heavily both in references to and paralleling the presentation of some of the existing background material.
-
--->
-
 The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
@@
 
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-
-$$
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, $$
   Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
@@
 
-Should we mention:
-
-- Temporal Differencing, RIENFORCE
-- SARSA, 
-- Double DQN?  
+Should we mention SARSA, Double DQN?  
 Can we add HER (since delayed reward case may be particularly interesting)

### 2021-05-06 manuscript/manuscript.Rmd
minor comments
draft conceptual figure

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.8\linewidth]{fig1.pdf}
+\caption{Reinforcement Learning focuses on the repeated feedback of *agent* interacting with an *environment*.
+The agent performs an action which influences the environment's future *state*.
+The agent recieves an observation of the new state and reward from the environment.}
+\end{figure}
+
 @brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
@@
 # RL Algorithms
+
+<!--
+
+I think we can lean more heavily both in references to and paralleling the presentation of some of the existing background material.
+
+-->
 
@@
 
-Should we mention SARSA, Double DQN?  
+Should we mention:
+
+- Temporal Differencing, RIENFORCE
+- SARSA, 
+- Double DQN?  
 Can we add HER (since delayed reward case may be particularly interesting)

### 2021-05-06 manuscript/manuscript.Rmd
caption

--- 2021-05-05 manuscript/manuscript.Rmd
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.5\linewidth]{fig1.pdf}
+\caption{Reinforcement Learning focuses on the repeated feedback of agent interacting with an \emph{environment}.
+The agent performs an action which influences the environment's future \emph{state}.
+The agent receives an observation of the new state and reward from the environment.}
+\end{figure}
+
 @brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.

### 2021-05-06 manuscript/manuscript.Rmd
most recet changes

@@
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
-To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
-RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
-Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
-Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
+To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
+Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
+RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
-Even simpler models are frequently employed throughout ecology to guide understanding and response.
+Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
 <!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
-Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
-We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
+We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
 We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
@@
 Any application of RL can be divided into two components: an *environment* and an *agent*.
-An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
-An *agent* is typically a computer algorithm which takes sequential observations from the environment and returns proposed actions.
-RL agents typically use one or more neural networks to represent the relationship between observed states and suitable actions, and 'train' these networks through repeated interactions with the environment to maximize the culumulative reward.
-<!-- Blah blah skip this?-->
-It is important to recognize that these definitions are significantly more general than requirements typically imposed by classical methods.  
-For instance, there is no a priori assumption that the environment is *fully observed*, or that the environment is *Markovian* (future states depend only on the current state), commonly assumed by classical methods [e.g.
-@Marescot2013].
-Just as important is the high degree of interoperability provided by the RL framework.
-As we shall see, an RL algorithm developed to play Atari games can instead be trained on an environment that simulates fish population dynamics without modification.
-(Note that this does not refer to the *trained* agent).
-This not guarantee that such an algorithm will be successful (even after extensive tuning and training in the fish population dynamics environment), but does open the door to more rapid exchange between fields.
-In particular, the adoption of standard interfaces for RL environments [@brockman2016] as well as the implementation of cutting edge algorithms in flexible open source frameworks [@sb3; @tf-agents] means that a researcher can test a range of RL algorithms against a single environment, or a given algorithm across a range of environments with equal ease.
+An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics; @offlinesurvey].
+Over the course of an *episode*, an *agent*, which is often a computer program, continuously interacts with the environment. 
+At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
+The goal of the agent is to learn how to maximize its expected cumulative reward.
+The agent learns how to achieve this objective during a period called *training*.
+In training, the agent *explores* the available states and actions. 
+Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+
+To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
+The trade-off between these approaches is that when it is possible to quickly learn a good model or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies.
+Yet, oftentimes learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform.
+Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
+
+Amongst model-free algorithms,
 

### 2021-05-07 manuscript/manuscript.Rmd
:pen:

--- 2021-05-06 manuscript/manuscript.Rmd
@@
 <!-- Environmental problems, ML -->
-Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
+Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
 But will ML approaches really help us tackle a changing planet?
@@
 Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
-Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
+Of course real ecosystems are also far more complex than a game of chess or MoonLander(TM) -- but can the same be said of our efforts at finding the best strategy?
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
 Even simpler models are frequently employed throughout ecology to guide understanding and response.
-<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
@@
 We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.
-
-The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
-We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
-We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
-If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
-Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
 
@@
 Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
-This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.\
+This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.
 The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
@@
 Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
-The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
+The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
 Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
-The use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
-A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
-<!-- Kari: lots of jargon in this paragraph that needs to be defined-->
+The use of a recurrent neural network (RNN) architecture instead of the multi-layer perceptron (MLP) used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
 
@@
 
-```{=html}
 <!-- 
-
-1. Return to theme of RL vs supervised/unsupervised learning.
+- RL:
+  - Focused on decisions
+  - more ambitious than other approaches in structured decision-making or other ML
+- Limited application, mostly from CS side without sufficient ecological grounding (citation to forest fire work, some )
 
 -->
-```
+
 <!-- Something about relevance to both the _theory_ and practice of ecological conservation.
@@
 "strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->
+
+<!-- move to discussion -->
+The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
+We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
+We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
 

### 2021-05-07 manuscript/manuscript.Rmd

--- 2021-05-06 manuscript/manuscript.Rmd
@@
 Any application of RL can be divided into two components: an *environment* and an *agent*.
-An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics; @offlinesurvey].
-Over the course of an *episode*, an *agent*, which is often a computer program, continuously interacts with the environment. 
+An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
+Over the course of an episode, an *agent*, which is often a computer program, continuously interacts with the environment. 
 At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
@@
 The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
-The trade-off between these approaches is that when it is possible to quickly learn a good model or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies.
-Yet, oftentimes learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
 Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
 
-Amongst model-free algorithms,
+Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
+*Policy gradient* algorithms exclusively learn the policy.
+*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
 

### 2021-05-07 manuscript/manuscript.Rmd
minor comments - mostly on intro

@@
 Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
+
+<!-- citation suggestion for declining biodiversity  -->
+
+<!-- @article{dirzo2014defaunation, -->
+<!--   title={Defaunation in the Anthropocene}, -->
+<!--   author={Dirzo, Rodolfo and Young, Hillary S and Galetti, Mauro and Ceballos, Gerardo and Isaac, Nick JB and Collen, Ben}, -->
+<!--   journal={science}, -->
+<!--   volume={345}, -->
+<!--   number={6195}, -->
+<!--   pages={401--406}, -->
+<!--   year={2014}, -->
+<!--   publisher={American Association for the Advancement of Science} -->
+<!-- } -->
+
 But will ML approaches really help us tackle a changing planet?
+
+<!-- suggestion: "...tackle our most pressing environmental challenges" -->
+
 Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
@@
 
+<!-- really like this- super clear and helpful! -->
+
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
+
+<!-- another widely used model is ED2 (ecosystem demography model) - I think more common than SORITE? https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2008JG000812 -->
+
 Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
@@
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+
+<!-- I think there is a missing the connection between the examples and needing more sophisticated methods for decision making below.
+i.e.
+"As we develop more complex process based models (such as Ecosim and ED2), methods that allow us to effectively integrate these high dimensional models into decision making processes will be critical to ensuring their policy relevance." (or something like that, then go on to NN ML)-->
+
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 
+ <!-- I like the summary but could maybe be less specific> "We start with a simple example from fisheries management that has a known optimal solution, proving that with no prior knowledge of underlying model dynamics RL agents can find a close to optimal solution.
+We go on to show that RL can outperform rule-of-thumb solutions to management scenarios not amenable to classic optimization methods such as dynamic programming.
+Our examples demonstrate that..."  -->
+
 The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
@@
 Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
+
+<!-- maybe cite Caleb's paper here? -->
+
+<!-- these last few sentences almost feel like discussion points.
+I'm torn though, because it it nice to have them in the intro.
+Maybe condense into 1 sentence at the end of you "summary" paragraph of the bulk of the paper above.
+for example: "We also explore the potential pitfalls of RL-based solutions to global environmental change problems.
+We show how, and why, these challenges are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary." then add a longer discussion towards the end of the paper-->
 
@@
 
+<!-- something to introduce this section/transition from the intro might be helpful: "We start with a brief overview of RL problems and terminology:" -->
+
 Any application of RL can be divided into two components: an *environment* and an *agent*.
@@
 Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+
+<!-- ^ this is great! I do think keeping a table of these terms would be helpful (and refering to that table in the text here) -->
 
@@
 The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-*Policy gradient* algorithms exclusively learn the policy.
+By contrast, *Policy gradient* algorithms exclusively learn the policy.
 *Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
@@
 An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
-In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
+In contrast to classical approaches, such as dynamic programming, there are few restrictions on what comprises a state or action, at least in principle.
 The state may be continuous or discrete, completely or only partially observed, one or many dimensions.

### 2021-05-07 manuscript/manuscript.Rmd
setup for wildfire example

--- 2021-05-07 manuscript/manuscript.Rmd
@@
 
+## Unsolved environments: Wildfires and other examples
+
+One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
+Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
+Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
+Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+
 # Discussion
+
+In this paper, we have highlighted the great potential that Reinforcement Learning, and in particular, of model-free *deep* RL has for helping confront challenging ecological and environmental management issues in a changing world.
 
@@
 - Limited application, mostly from CS side without sufficient ecological grounding (citation to forest fire work, some )
+
+- Model Free
+
+- Results
+
+- Technical Caveats
+
+- Social Caveats
+
+- Way Forward
+  Open Competition & Collaboration
+  Leaderboards
 

### 2021-05-07 manuscript/manuscript.Rmd
discussion WIP :pen: :construction:

@@
 Even simpler models are frequently employed throughout ecology to guide understanding and response.
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
 Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 
+<!-- decision theory is a fundamental scientific research topic -->
+Ecological challenges facing the planet today are complex and outcomes are as uncertain and highly consequential.
+Even our best models and best research will never provide a crystal ball to the future, only better elucidate possible scenarios.
+Consequently, that research must also confront the challenge of making robust, resilient decisions in a changing world.
+The science of ecological management and quantitative decision-making has a long history [e.g.
+@Schaefer1954; @Walters1974] and remains an active area of research [@Possignham2006; @Fischer2009; @Polasky2011].
+<!-- Something about influence of natural resource economics and preference for analytically tractable optimization -->
+However, the limitations of classical methods such as optimal control frequently constrain these
+
 In this paper, we have highlighted the great potential that Reinforcement Learning, and in particular, of model-free *deep* RL has for helping confront challenging ecological and environmental management issues in a changing world.

### 2021-05-07 manuscript/manuscript.Rmd

@@
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
-To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
-RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
-Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
-Of course real ecosystems are also far more complex than a game of chess or MoonLander(TM) -- but can the same be said of our efforts at finding the best strategy?
+To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
+Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
+RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
-Even simpler models are frequently employed throughout ecology to guide understanding and response.
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
-Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
-We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
+Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
+<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
+Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
+We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
 We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
@@
 Any application of RL can be divided into two components: an *environment* and an *agent*.
-An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
-An *agent* is typically a computer algorithm which takes sequential observations from the environment and returns proposed actions.
-RL agents typically use one or more neural networks to represent the relationship between observed states and suitable actions, and 'train' these networks through repeated interactions with the environment to maximize the culumulative reward.
-<!-- Blah blah skip this?-->
-It is important to recognize that these definitions are significantly more general than requirements typically imposed by classical methods.  
-For instance, there is no a priori assumption that the environment is *fully observed*, or that the environment is *Markovian* (future states depend only on the current state), commonly assumed by classical methods [e.g.
-@Marescot2013].
-Just as important is the high degree of interoperability provided by the RL framework.
-As we shall see, an RL algorithm developed to play Atari games can instead be trained on an environment that simulates fish population dynamics without modification.
-(Note that this does not refer to the *trained* agent).
-This not guarantee that such an algorithm will be successful (even after extensive tuning and training in the fish population dynamics environment), but does open the door to more rapid exchange between fields.
-In particular, the adoption of standard interfaces for RL environments [@brockman2016] as well as the implementation of cutting edge algorithms in flexible open source frameworks [@sb3; @tf-agents] means that a researcher can test a range of RL algorithms against a single environment, or a given algorithm across a range of environments with equal ease.
+An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
+Over the course of an episode, an *agent*, which is often a computer program, continuously interacts with the environment. 
+At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
+The goal of the agent is to learn how to maximize its expected cumulative reward.
+The agent learns how to achieve this objective during a period called *training*.
+In training, the agent *explores* the available states and actions. 
+Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+
+To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
+Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
+
+Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
+*Policy gradient* algorithms exclusively learn the policy.
+*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
 

### 2021-05-07 manuscript/manuscript.Rmd
First pass at a neural net paragraph; need to still revise previous parts

@@
 <!-- Environmental problems, ML -->
-Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
+Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
 But will ML approaches really help us tackle a changing planet?
@@
 <!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.
+
+The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
+We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
+We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
 
@@
 Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
 
@@
 Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
+
+Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
+Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g.
+when the possible states and actions can not be represented in a tractable table [@CITE].
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being universal function approximators [@CITE].
+Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
+While there are other function approximators that can be used in RL, e.g.
+Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
+We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the basic premise is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 
@@
 Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
-This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.
+This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.\
 The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
@@
 Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
-The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
+The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
 Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
-The use of a recurrent neural network (RNN) architecture instead of the multi-layer perceptron (MLP) used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
-
-## Unsolved environments: Wildfires and other examples
-
-One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
-While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
-Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
-Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
-Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+The use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
+A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
+<!-- Kari: lots of jargon in this paragraph that needs to be defined-->
 
@@
 
-<!-- decision theory is a fundamental scientific research topic -->
-Ecological challenges facing the planet today are complex and outcomes are as uncertain and highly consequential.
-Even our best models and best research will never provide a crystal ball to the future, only better elucidate possible scenarios.
-Consequently, that research must also confront the challenge of making robust, resilient decisions in a changing world.
-The science of ecological management and quantitative decision-making has a long history [e.g.
-@Schaefer1954; @Walters1974] and remains an active area of research [@Possignham2006; @Fischer2009; @Polasky2011].
-<!-- Something about influence of natural resource economics and preference for analytically tractable optimization -->
-However, the limitations of classical methods such as optimal control frequently constrain these
+```{=html}
+<!--
 
-In this paper, we have highlighted the great potential that Reinforcement Learning, and in particular, of model-free *deep* RL has for helping confront challenging ecological and environmental management issues in a changing world.
-
-<!-- 
-- RL:
-  - Focused on decisions
-  - more ambitious than other approaches in structured decision-making or other ML
-- Limited application, mostly from CS side without sufficient ecological grounding (citation to forest fire work, some )
-
-- Model Free
-
-- Results
-
-- Technical Caveats
-
-- Social Caveats
-
-- Way Forward
-  Open Competition & Collaboration
-  Leaderboards
+1. Return to theme of RL vs supervised/unsupervised learning.
 
 -->
-
+```
 <!-- Something about relevance to both the _theory_ and practice of ecological conservation.
@@
 "strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->
-
-<!-- move to discussion -->
-The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
-We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
-We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
-If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
-Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
 

### 2021-05-10 manuscript/manuscript.Rmd
Small edits; keeping up to date

@@
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not use or attempt to learn a model of the environment.
+Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 
-Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
+Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
 Generally, model-free agents are either learning a policy function, a value function or both.
@@
 *Policy gradient* algorithms exclusively learn the policy.
-*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+And, lastly, *Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
 Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.

### 2021-05-10 manuscript/manuscript.Rmd
Trying to clarify nn paragraph

@@
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not use or attempt to learn a model of the environment.
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
 Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
@@
 *Policy gradient* algorithms exclusively learn the policy.
-And, lastly, *Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
 There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
@@
 Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the basic premise is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+This same process occurs in value learning and actor-critic algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
 

### 2021-05-10 manuscript/manuscript.Rmd
Few more edits; still need to put citations in

@@
 And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
-Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
 
@@
 Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being universal function approximators [@CITE].
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
 Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
@@
 We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
+This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
 

### 2021-05-10 manuscript/manuscript.Rmd
short paragraph on ethics/power in discussion

--- 2021-05-10 manuscript/manuscript.Rmd
@@
 Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+
+<!-- ^ this is great! I do think keeping a table of these terms would be helpful (and refering to that table in the text here) -->
 
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not use or attempt to learn a model of the environment.
-Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 
-Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
+Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
 Generally, model-free agents are either learning a policy function, a value function or both.
@@
 The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-*Policy gradient* algorithms exclusively learn the policy.
-And, lastly, *Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+By contrast, *Policy gradient* algorithms exclusively learn the policy.
+*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
 Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being universal function approximators [@CITE].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the basic premise is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 
@@
 -->
+
+<!-- maybe consider some combined conceptual figure/table?  But overall I really like the table and think it will be helpful for readers -->
+
 \begin{figure}
@@
 
+<!-- first pass at this -->
+
+Critical engagement in the ethical and power issues presented by decision making algorithms will be imperative to ensuring that applications of RL to ecological challenges are fair and just.
+While widespread discussion of the ethics of decision making algorithms across sectors such as policing, finance, and employment, are a commonplace little has been discussed on the biases and implications of implementing these methods in the environmental context.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Mitigation of the power asymmetries that might arise from algorithmic applications are key [cite caleb's paper?].
+Engagement with experts in governance and ethics to ensure that applications of AI to ecological challenges are applied responsibly and transparently.
+
 # a vision going forward

### 2021-05-10 manuscript/manuscript.Rmd
Starting things up here

--- 2021-05-10 manuscript/manuscript.Rmd
@@
 Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
 
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
-Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 *Policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
 Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being universal function approximators [@CITE].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
 
@@
 
-The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
-This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
-The task for the agent is to learn a *policy*, $\pi$, that prescribes what action the agent should perform after observing a certain state.
-The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continually updates its policy in order to maximize its expected cumulative sum of rewards.
-
 <!-- Still can't decide if this statement is too strong; i.e. that at least theoretically / LSTM, RL permits learning processes which are not Markovian in the observed variables -->
 When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem can be formalized as a Markov Decision Process (MDP).
-An MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
+when the state that the agent observes fully characterizes the system, the reinforcement learning problem is conventionally formalized as a Markov Decision Process (MDP).
+A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
+After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
 As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
-From this definition, we can provide an agent's trajectory distribution for a given policy as,
+From these definitions, we can provide an agent's trajectory distribution for a given policy as,
 

### 2021-05-11 manuscript/manuscript.Rmd
WIP

@@
 <!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 
+The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
+This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
+The task for the agent is to learn a *policy*, $\pi$, that prescribes what action the agent should perform after observing a certain state.
+The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continually updates its policy in order to maximize its expected cumulative sum of rewards.
+
 <!-- Still can't decide if this statement is too strong; i.e. that at least theoretically / LSTM, RL permits learning processes which are not Markovian in the observed variables -->
 When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem is conventionally formalized as a Markov Decision Process (MDP).
-A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
+when the state that the agent observes fully characterizes the system, the reinforcement learning problem can be formalized as a Markov Decision Process (MDP).
+An MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
 As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
-From these definitions, we can provide an agent's trajectory distribution for a given policy as,
+From this definition, we can provide an agent's trajectory distribution for a given policy as,
 
@@
 Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
-The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
+The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
 Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
-The use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
-A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
-<!-- Kari: lots of jargon in this paragraph that needs to be defined-->
+The use of a recurrent neural network (RNN) architecture instead of the multi-layer perceptron (MLP) used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
+
+## Unsolved environments: Wildfires and other examples
+
+One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
+Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
+Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
+Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+
+<!-- Marcus add ~ paragraph discussion of the fire gym, citing motivating examples, basic overview of the state space, action space and reward and a screenshot of the simulation as a figure? -->
 
@@
 
-```{=html}
+<!-- decision theory is a fundamental scientific research topic -->
+Ecological challenges facing the planet today are complex and outcomes are both uncertain and consequential.
+Even our best models and best research will never provide a crystal ball to the future, only better elucidate possible scenarios.
+Consequently, that research must also confront the challenge of making robust, resilient decisions in a changing world.
+The science of ecological management and quantitative decision-making has a long history [e.g.
+@Schaefer1954; @Walters1974] and remains an active area of research [@Possignham2006; @Fischer2009; @Polasky2011].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models, ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
+It is much simpler to accommodate these elements and uncertainties in a suite of simulations than it is to solve for the optimal policy.
+Neural networks used in deep RL have proven remarkably effective in handling such complexity, particularly when leveraging immense computing resources increasingly available through advances in hardware and software [@gpu_computing].
+
+However, successfully applying RL to complex ecological problems is no easy task.
+Deep RL is still a very young field, where despite several landmark successes, potential far outstrips practice.
+
+<!-- Something about influence of natural resource economics and preference for analytically tractable optimization -->
+
 <!-- 
+- RL:
+  - Focused on decisions
+  - more ambitious than other approaches in structured decision-making or other ML
+- Limited application, mostly from CS side without sufficient ecological grounding (citation to forest fire work, some )
 
-1. Return to theme of RL vs supervised/unsupervised learning.
+- Model Free
+
+- Results
+
+- Technical Caveats
+
+- Social Caveats
+
+- Way Forward
+  Open Competition & Collaboration
+  Leaderboards
 
 -->
-```
+
 <!-- Something about relevance to both the _theory_ and practice of ecological conservation.

### 2021-05-11 manuscript/manuscript.Rmd
:pen: :construction:

@@
 The science of ecological management and quantitative decision-making has a long history [e.g.
-@Schaefer1954; @Walters1974] and remains an active area of research [@Possignham2006; @Fischer2009; @Polasky2011].
-However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models, ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
+@Schaefer1954; @Walters1974] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
 It is much simpler to accommodate these elements and uncertainties in a suite of simulations than it is to solve for the optimal policy.
@@
 
+Model-free RL is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
+Rarely do we know the underlying functional forms for ecological processes -- is population recruitment more Beverton-Holt or Ricker?
+Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
+Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
+Model-free RL approaches offer a powerful way to factor this thorny issue.
+The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
+Alternative models can easily be implemented in simulations, allowing an agent to train against both contingencies.
+A successful agent will learn a policy that works well in both scenarios.
+This is analogous to classical approaches for integrating over parameter or model uncertainty [e.g. @Walters1981; @Ludwig1982],
+
 However, successfully applying RL to complex ecological problems is no easy task.
+The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance is another task.
+As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
+Such examples provide important benchmarks against which to calibrate the performance, tuning and training requirements of RL,
+and can easily be extended into more complex problems once RL agents can master the basics.
+Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
+<!-- More precise caveats about reproducibility, convergence, tuning, computational limitations-->
+
 Deep RL is still a very young field, where despite several landmark successes, potential far outstrips practice.
+Recent advances in the field have proven the potential of the approach to solve complex problems [@alphaGo2016; @alphaGoZero; @alphazero; @DQN], but typically leveraging large teams with decades of experience in ML and millions of dollars worth of computing power [@alphaGoZero].
+Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
+Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [@gym], and performance can easily be compared against alternatives on public leaderboards [@gym].
+Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems [@netflix_prize; @imagenet].
+Advances in both computational power and algorithmic efficiency continue to reduce costs at a dramatic pace, significantly faster than Moore's law [@openai_costs].
 
-<!-- Something about influence of natural resource economics and preference for analytically tractable optimization -->
+<!-- Discuss training across mutiple envs, sequentially vs simultaneously -->
 

### 2021-05-12 manuscript/manuscript.Rmd
some references and reworking ethics paragraph (WIP)

--- 2021-05-10 manuscript/manuscript.Rmd
@@
 
+<!-- second pass at this -->
+
+Algorithms are playing increasingly influential decision-making roles in sectors such as policing, finance, and employment, raising concerns about the ethics, biases, and power entrenched in both the algorithms and their outputs [@kalluri2020].
+As algorithms play a greater role in environmental and conservation policy, will similar concerns arise?
+While this question has garnered  recent interest in the literature [@wearn2019; @adams2019; @scoville2021] the use use of decision making algorithms in environmental applications has largely been confined to the context of classification algorithms (i.e.
+to identify wildlife poaching [@adams2019] or illegal fishing behavior [@mcdonald2021]).
+Increasingly, applications of AI to meeting sustainable development goals [@], prioritizing protected area locations [@], and predicting However, good intentions are not sufficient to ensure equitable outcomes.
+Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges are applied responsibly.
+
 # a vision going forward

### 2021-05-12 manuscript/manuscript.Rmd
Getting yesterday's additions up

@@
 Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
-
-<!-- citation suggestion for declining biodiversity  -->
-
-<!-- @article{dirzo2014defaunation, -->
-<!--   title={Defaunation in the Anthropocene}, -->
-<!--   author={Dirzo, Rodolfo and Young, Hillary S and Galetti, Mauro and Ceballos, Gerardo and Isaac, Nick JB and Collen, Ben}, -->
-<!--   journal={science}, -->
-<!--   volume={345}, -->
-<!--   number={6195}, -->
-<!--   pages={401--406}, -->
-<!--   year={2014}, -->
-<!--   publisher={American Association for the Advancement of Science} -->
-<!-- } -->
-
 But will ML approaches really help us tackle a changing planet?
-
-<!-- suggestion: "...tackle our most pressing environmental challenges" -->
-
 Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
@@
 
-<!-- really like this- super clear and helpful! -->
-
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
-
-<!-- another widely used model is ED2 (ecosystem demography model) - I think more common than SORITE? https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2008JG000812 -->
-
 Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
@@
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
-
-<!-- I think there is a missing the connection between the examples and needing more sophisticated methods for decision making below.
-i.e.
-"As we develop more complex process based models (such as Ecosim and ED2), methods that allow us to effectively integrate these high dimensional models into decision making processes will be critical to ensuring their policy relevance." (or something like that, then go on to NN ML)-->
-
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 
- <!-- I like the summary but could maybe be less specific> "We start with a simple example from fisheries management that has a known optimal solution, proving that with no prior knowledge of underlying model dynamics RL agents can find a close to optimal solution.
-We go on to show that RL can outperform rule-of-thumb solutions to management scenarios not amenable to classic optimization methods such as dynamic programming.
-Our examples demonstrate that..."  -->
-
 The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
@@
 Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
-
-<!-- maybe cite Caleb's paper here? -->
-
-<!-- these last few sentences almost feel like discussion points.
-I'm torn though, because it it nice to have them in the intro.
-Maybe condense into 1 sentence at the end of you "summary" paragraph of the bulk of the paper above.
-for example: "We also explore the potential pitfalls of RL-based solutions to global environmental change problems.
-We show how, and why, these challenges are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary." then add a longer discussion towards the end of the paper-->
 
@@
 
-<!-- something to introduce this section/transition from the intro might be helpful: "We start with a brief overview of RL problems and terminology:" -->
-
 Any application of RL can be divided into two components: an *environment* and an *agent*.
@@
 Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
-
-<!-- ^ this is great! I do think keeping a table of these terms would be helpful (and refering to that table in the text here) -->
 
@@
 The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-By contrast, *Policy gradient* algorithms exclusively learn the policy.
+*Policy gradient* algorithms exclusively learn the policy.
 *Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
@@
 An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
-In contrast to classical approaches, such as dynamic programming, there are few restrictions on what comprises a state or action, at least in principle.
+In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
 The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
@@
 -->
-
-<!-- maybe consider some combined conceptual figure/table?  But overall I really like the table and think it will be helpful for readers -->
-
 \begin{figure}
@@
 
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected cumulative, discounted reward:
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
 
 $$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
 $$
@@
 
-Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
+<!-- Explain policy gradients as this is the natural progression
+So sketch out reinforce, ending that its susceptibility to high variance
+
+talk about value-based learning so q values and bellman recursion
+-->
+## Policy Gradients
+The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
+There are both gradient-free and gradient-based policy search methods that have been successful in practice [@CITE; @CITE]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
+In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters. 
+At each gradient step, the following update is performed,
+
+$$
+  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta).
+$$
+
+Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, then $\nabla J(\pi_\theta)$ can be estimated as:
+
+$$
+  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \Big( \sum_{t'=t}^H \gamma^{t'-t} r(s_{i, t'}, a_{i, t'}) \Big).
+$$
+
+See the appendix for how this is derived. 
+The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return. 
+In practice, however, the empirical return tends to have high variance, resulting in a very noisy gradient ascent that impedes learning [@CITE].
+To avoid this issue, most policy gradient algorithms replace the empirical return with a different return estimator that has lower variance [@CITE GAE].
+A sketch of a simple policy gradient algorithm is shown in FIGURE.
+
+## Value Learning
+Another way to approach the RL problem is by estimating a value function, and using the value function to retrieve a policy. 
+Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
 Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
+The state-action-value or Q function finds the expected return provided a state and action,
+
+$$
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg].
+$$
+
+The general motivation is that if we know the Q function, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+In the case that the state-action space can be tractably represented in a table, one could employ an algorithm, like temporal-difference control -- see Sutton Ch.
+6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
+Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; one effective alternative for this case is to use a neural network to approximate the Q-function.
+
 There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
@@
 
-<!-- second pass at this -->
-
-Algorithms are playing increasingly influential decision-making roles in sectors such as policing, finance, and employment, raising concerns about the ethics, biases, and power entrenched in both the algorithms and their outputs [@kalluri2020].
-As algorithms play a greater role in environmental and conservation policy, will similar concerns arise?
-While this question has garnered  recent interest in the literature [@wearn2019; @adams2019; @scoville2021] the use use of decision making algorithms in environmental applications has largely been confined to the context of classification algorithms (i.e.
-to identify wildlife poaching [@adams2019] or illegal fishing behavior [@mcdonald2021]).
-Increasingly, applications of AI to meeting sustainable development goals [@], prioritizing protected area locations [@], and predicting However, good intentions are not sufficient to ensure equitable outcomes.
-Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges are applied responsibly.
-
 # a vision going forward

### 2021-05-12 manuscript/manuscript.Rmd
Almost done with value learning bit

@@
 ## Value Learning
-Another way to approach the RL problem is by estimating a value function, and using the value function to retrieve a policy. 
+Another way to approach the RL problem is by estimating a value function, and then using the value function to retrieve a policy. 
 Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
-The state-action-value or Q function finds the expected return provided a state and action,
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
+For example, the state-action-value or Q function finds the expected return from a state and action pair under a policy, $\pi$,
 
@@
 
-The general motivation is that if we know the Q function, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-In the case that the state-action space can be tractably represented in a table, one could employ an algorithm, like temporal-difference control -- see Sutton Ch.
+The general motivation is that if we know the Q function under the condition that the given policy visits all state-action pairs [@suttonbarto 131; @Qproof], then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+In the case that the state-action space can be tractably represented in a table, one could employ a  algorithm, like temporal-difference control -- see Sutton Ch.
 6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
-Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; one effective alternative for this case is to use a neural network to approximate the Q-function.
+Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; one effective alternative for this case is to use a neural network to approximate the Q-function, a method called deep Q-learning.
+The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
+The Bellman error
 

### 2021-05-12 manuscript/manuscript.Rmd
Draft discussion

--- 2021-05-12 manuscript/manuscript.Rmd
@@
 <!-- Environmental problems, ML -->
-Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
-
-<!-- citation suggestion for declining biodiversity  -->
-
-<!-- @article{dirzo2014defaunation, -->
-<!--   title={Defaunation in the Anthropocene}, -->
-<!--   author={Dirzo, Rodolfo and Young, Hillary S and Galetti, Mauro and Ceballos, Gerardo and Isaac, Nick JB and Collen, Ben}, -->
-<!--   journal={science}, -->
-<!--   volume={345}, -->
-<!--   number={6195}, -->
-<!--   pages={401--406}, -->
-<!--   year={2014}, -->
-<!--   publisher={American Association for the Advancement of Science} -->
-<!-- } -->
-
+Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
 But will ML approaches really help us tackle a changing planet?
-
-<!-- suggestion: "...tackle our most pressing environmental challenges" -->
-
 Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
-
-<!-- another widely used model is ED2 (ecosystem demography model) - I think more common than SORITE? https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2008JG000812 -->
-
 Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
 <!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
-
-<!-- I think there is a missing the connection between the examples and needing more sophisticated methods for decision making below.
-i.e.
-"As we develop more complex process based models (such as Ecosim and ED2), methods that allow us to effectively integrate these high dimensional models into decision making processes will be critical to ensuring their policy relevance." (or something like that, then go on to NN ML)-->
-
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.
-
- <!-- I like the summary but could maybe be less specific> "We start with a simple example from fisheries management that has a known optimal solution, proving that with no prior knowledge of underlying model dynamics RL agents can find a close to optimal solution.
-We go on to show that RL can outperform rule-of-thumb solutions to management scenarios not amenable to classic optimization methods such as dynamic programming.
-Our examples demonstrate that..."  -->
-
-The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
-We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
-We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
-If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
-Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
-
-<!-- maybe cite Caleb's paper here? -->
-
-<!-- these last few sentences almost feel like discussion points.
-I'm torn though, because it it nice to have them in the intro.
-Maybe condense into 1 sentence at the end of you "summary" paragraph of the bulk of the paper above.
-for example: "We also explore the potential pitfalls of RL-based solutions to global environmental change problems.
-We show how, and why, these challenges are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary." then add a longer discussion towards the end of the paper-->
 
@@
 
-```{=html}
-<!--
+<!-- Ecology has complex decisions and we lack necessary tools; RL could be it -->
+Ecological challenges facing the planet today are complex and outcomes are both uncertain and consequential.
+Even our best models and best research will never provide a crystal ball to the future, only better elucidate possible scenarios.
+Consequently, that research must also confront the challenge of making robust, resilient decisions in a changing world.
+The science of ecological management and quantitative decision-making has a long history [e.g.
+@Schaefer1954; @Walters1974] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
+Complexity comes not only from the ecological processes but also the available actions.
+While conservation objectives are often framed in broad, simple terms, the potential action space is often huge: conservation efforts can be implemented in a wide variety of ways, and the most effective policies may vary in both space and time.
+<!-- Millie: can we say something more precise about 'simple terms' or add some citation here?-->
+While such complex dynamics and large state and action spaces are intractable to classical methods, they are straight forward to implement in simulation environments such as those used to train an RL algorithm.
+Neural networks used in deep RL have proven remarkably effective in handling such complexity, particularly when leveraging immense computing resources increasingly available through advances in hardware and software [@gpu_computing].
 
-1. Return to theme of RL vs supervised/unsupervised learning.
+<!-- Model free is great bc we don't know the true models -->
+The rapidly expanding class of Model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
+Rarely do we know the underlying functional forms for ecological processes -- is population recruitment more Beverton-Holt or Ricker?
+Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
+Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
+Model-free RL approaches offer a powerful way to factor this thorny issue.
+The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
+Alternative models can easily be implemented in simulations, allowing an agent to train against both contingencies.
+A successful agent will learn a policy that works well in both scenarios.
+This is analogous to classical approaches for integrating over parameter or model uncertainty [e.g. @Walters1981; @Ludwig1982],
 
--->
-```
-<!-- Something about relevance to both the _theory_ and practice of ecological conservation.
-(e.g.
-"strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->
-
-## Open Problems
-
+<!-- Real problems are harder than these examples -->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 In the case of the ecosystem tipping point scenario, the optimal strategy is unknown, and the approximate solution found by our RL implementation could almost certainly be improved upon.
-
 Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
@@
 
-# Discussion
+<!-- Solving 'real' problems will be hard -->
+However, successfully applying RL to complex ecological problems is no easy task.
+The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance is another task.
+As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
+Such examples provide important benchmarks against which to calibrate the performance, tuning and training requirements of RL,
+and can easily be extended into more complex problems once RL agents can master the basics.
+Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
+<!-- something about several technical caveats: instability, interpretability reward design-->
 
-# Open issues
-
-Deep reinforcement learning has only become an established subfield of AI over the past decade; and, as with most young fields, there are a range of unsolved problems.
-In practice, the foremost problem with DRL is instability.
-For a range of reasons, including high variance and the exploration vs.
-exploitation trade-off, model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
+<!-- reason 1: instability-->
+One leading problem with deep RL is instability.
+For a range of reasons, including high variance [@cite] and the exploration vs.
+exploitation trade-off [@cite], model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
 For instance, it is not uncommon to observe vastly different behavior when two agents are initialized with different random seeds [cite ibid].
@@
 The hope of the RL practitioner is that after doing an exhaustive hyperparameter search, the practitioner will find a set of hyperparameters that results in a robust, well-performing agent, but there are no guarantees that hyperparameter tuning will lead to this outcome.
-<!-- Kari: what does it mean for two algorithms with similar hyperparameters to perform differently? (As happens in our examples) -->
+<!--CB: I think we can say more about how to diagnose this problem if not fix it.
+    Likewise should address the importance of assessing robustness across multiple models/environments, probably at least as important an issue as robustness to nuciance hyperparameters like random seed.
+    -->
 
-In the context of conservation, we anticipate that there will be a host of specific challenges to using DRL methods.
-Interpretability is likely to be the biggest barrier to adoption.
+<!-- reason 2: interpretability --> 
+In the context of conservation, we anticipate that there will be a host of specific challenges to using deep RL methods.
+Interpretability is likely to be the biggest barrier to adoption. <!-- don't need to rank them -->
 Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
@@
 
+<!-- reason 3: reward design -->
 An issue that is central to any applied optimization problem is using the right objective.
@@
 
-Ultimately, we anticipate that getting DRL to work well on fully detailed conservation decision-making problems will be challenging.
+<!-- reason 4/additional reasons, like sparse rewards -->
+Ultimately, we anticipate that getting deep RL to work well on fully detailed conservation decision-making problems will be challenging.
 Beyond typically having large state-action spaces for agents to explore, conservation problems tend to have sparse reward structures.
@@
 
--   wild west of AI?
-    -can separate RL from having gyms implemented -\> how do you compare different models, -mention that customization is probably the way to go, baselines only go so far
+<!-- But never fear, hard problems have been solved before!-->
+Deep RL is still a very young field, where despite several landmark successes, potential far outstrips practice.
+Recent advances in the field have proven the potential of the approach to solve complex problems [@alphaGo2016; @alphaGoZero; @alphazero; @DQN], but typically leveraging large teams with decades of experience in ML and millions of dollars worth of computing power [@alphaGoZero].
+Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
+Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [@gym], and performance can easily be compared against alternatives on public leaderboards [@gym].
+Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].
+Advances in both computational power and algorithmic efficiency continue to reduce costs at a dramatic pace, significantly faster than Moore's law [@openai_costs].
+By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may both be better able to replicate that collaborative, iterative success in cracking hard problems.
+Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.
+If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.
 
--   AI ethics, transparency
+<!-- transition from big promise to perils -- Millie's section on risks here -->
 
-<!-- second pass at this -->
+<!-- 
+The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
+We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
+We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
 
-Algorithms are playing increasingly influential decision-making roles in sectors such as policing, finance, and employment, raising concerns about the ethics, biases, and power entrenched in both the algorithms and their outputs [@kalluri2020].
-As algorithms play a greater role in environmental and conservation policy, will similar concerns arise?
-While this question has garnered  recent interest in the literature [@wearn2019; @adams2019; @scoville2021] the use use of decision making algorithms in environmental applications has largely been confined to the context of classification algorithms (i.e.
-to identify wildlife poaching [@adams2019] or illegal fishing behavior [@mcdonald2021]).
-Increasingly, applications of AI to meeting sustainable development goals [@], prioritizing protected area locations [@], and predicting However, good intentions are not sufficient to ensure equitable outcomes.
-Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges are applied responsibly.
-
-# a vision going forward
-
--   our public/open source environments and agents, leaderboard
--   the need for collaborative/competitive development both to make better benchmark environments and better agents
+-->

### 2021-05-12 manuscript/manuscript.Rmd
comments for remaining sections
Finished first pass at value-learning

--- 2021-05-12 manuscript/manuscript.Rmd
@@
 The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return. 
-In practice, however, the empirical return tends to have high variance, resulting in a very noisy gradient ascent that impedes learning [@CITE].
+In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@CITE].
 To avoid this issue, most policy gradient algorithms replace the empirical return with a different return estimator that has lower variance [@CITE GAE].
@@
 
-The general motivation is that if we know the Q function under the condition that the given policy visits all state-action pairs [@suttonbarto 131; @Qproof], then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+The general motivation is that if we know the Q function under certain conditions -- e.g.
+that the state-action space is discrete and that the given policy visits all state-action pairs repeatedly [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
 In the case that the state-action space can be tractably represented in a table, one could employ a  algorithm, like temporal-difference control -- see Sutton Ch.
@@
 The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
-The Bellman error
+The Bellman error is a measure of how distant a given value function is from that value function which has been re-evaluated with recent environment interaction.
+In the case where the Q-function is approximated by some neural network and thus represented as $Q_\theta$, the Bellman error for one state-action pair would be
+
+$$
+  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\theta(s_{t+1}, a) - Q_\theta(s_t, a_t).
+$$
+
+See the appendix/reference for details on how the Bellman Error is derived from eq. [@CITE]. 
+The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across a range of these samples then perform the following gradient descent,
+
+$$
+  \theta_{t+1} = \theta_t - \alpha \nabla_\theta \, \sum_i \mathcal{E}_i^2,
+$$
+
+where $i$ indexes the sample.
+After convergence to the Q function, the optimal policy could
 

### 2021-05-12 manuscript/manuscript.Rmd
Completing actor-critic paragraph

@@
 
-The general motivation is that if we know the Q function under certain conditions -- e.g.
+The general Q-learning motivation is that if we know the Q function under certain conditions -- e.g.
 that the state-action space is discrete and that the given policy visits all state-action pairs repeatedly [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
@@
 6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
-Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; one effective alternative for this case is to use a neural network to approximate the Q-function, a method called deep Q-learning.
+Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; for this case, one effective alternative is to use a neural network to approximate the Q-function, a method called deep Q-learning.
 The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
 The Bellman error is a measure of how distant a given value function is from that value function which has been re-evaluated with recent environment interaction.
-In the case where the Q-function is approximated by some neural network and thus represented as $Q_\theta$, the Bellman error for one state-action pair would be
+Suppose we attempt to approximate the Q-function with a neural network and thus represent this approximate Q-function as as $Q_\phi$, the Bellman error for one state-action pair would be
 
 $$
-  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\theta(s_{t+1}, a) - Q_\theta(s_t, a_t).
+  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\phi(s_{t+1}, a) - Q_\phi(s_t, a_t).
 $$
@@
 See the appendix/reference for details on how the Bellman Error is derived from eq. [@CITE]. 
-The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across a range of these samples then perform the following gradient descent,
+The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across these samples -- with the samples indexed by $i$ --, and then perform the following gradient descent:
 
 $$
-  \theta_{t+1} = \theta_t - \alpha \nabla_\theta \, \sum_i \mathcal{E}_i^2,
+  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 
 $$
 
-where $i$ indexes the sample.
-After convergence to the Q function, the optimal policy could
+When being evaluated, a deep Q-learning agent will select actions by inputting the current state to the Q network and then identifying the state-action pair with the highest estimated Q value.
+The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITE].
+But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @CITE for more insight on these issues.   
+The primary concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@CITE].
 
-There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
-The state-value function is defined as the expected return from a given state,
+## Actor-Critic
+Actor-critic algorithms integrate the main components from both value-learning and policy gradient based methods.
+The "actor" attempts to learn the policy, and the "critic" attempts to learn a value function.
+While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from vanilla policy gradient algorithms by incorporating the critic in its gradient ascent step.
+Instead of estimating the gradient of the return as in eq ..., the gradient of the return is estimated as
 
 $$
-  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \, \hat{Q}(s_{i,t}, a_{i, t}).
 $$
 
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-
-$$
-  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succinctly defined as
-
-$$
-  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-$$ 
-Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
-Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
+where $\hat{Q}$ is a value function estimated by the critic.
+The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@CITE].
+Consequently, actor-critic algorithms have achieved a large majority of the state of the art results in model-free deep RL [@CITE].
 

### 2021-05-12 manuscript/manuscript.Rmd
Deleting a to-self comment

@@
 
-<!-- Explain policy gradients as this is the natural progression
-So sketch out reinforce, ending that its susceptibility to high variance
-
-talk about value-based learning so q values and bellman recursion
--->
 ## Policy Gradients

### 2021-05-13 manuscript/manuscript.Rmd
Small changes

@@
 In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@CITE].
-To avoid this issue, most policy gradient algorithms replace the empirical return with a different return estimator that has lower variance [@CITE GAE].
-A sketch of a simple policy gradient algorithm is shown in FIGURE.
+To avoid this issue, most policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@CITE GAE].
+A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
 

### 2021-05-13 manuscript/manuscript.Rmd
First pass over the intro env section

@@
 An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
-Over the course of an episode, an *agent*, which is often a computer program, continuously interacts with the environment. 
-At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
+Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
+At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
 The goal of the agent is to learn how to maximize its expected cumulative reward.
@@
 In training, the agent *explores* the available states and actions. 
-Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
 
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
+Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 *Policy gradient* algorithms exclusively learn the policy.
-*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
-Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
+
+Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
+Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g.
+when the possible states and actions can not be represented in a tractable table [@CITE].
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
+Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
+While there are other function approximators that can be used in RL, e.g.
+Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
+We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
 
@@
 
-An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
-In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
-The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
-For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
-The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
-The reward will in general depend on both the current state and choice of actions.
-An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
-In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
+<!-- ML: what are classical approaches below here? -->
+In contrast to classical approaches, there are few restrictions on what comprises a state or action. 
+The state may be continuous or discrete, completely or partially observed, single or multidimensional.
+For instance, we might represent an ecological community of 10 interacting species using an environment with a 10-dimensional continuously valued state space representing the possible population densities of each species.
+The action space can be similarly complex.
+For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at each of these species during a time step.
+The main focus of an RL environment, however, is in the environment's transition dynamics and reward function.
+The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and inputted action.
+This ability allows for RL environments to model a broad range of decision making problems.
+For example, we can set the transitions to be deterministic or stochastic.
+We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that require a long sequence of actions to come across, e.g.
+the end point in a maze.
 In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+Another important consideration in RL is whether to use an episodic or non-episodic environment.
+In episodic environments, there is a finite number of time steps during which the agent interacts with the environment, and for these environments, the designer must specify a stopping condition.
+Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
+For this paper, we only consider episodic environments.
 
@@
 
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected cumulative, discounted reward:
 
 $$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
 $$
@@
 
-## Policy Gradients
-The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
-There are both gradient-free and gradient-based policy search methods that have been successful in practice [@CITE; @CITE]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
-In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters. 
-At each gradient step, the following update is performed,
+Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
+There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
+The state-value function is defined as the expected return from a given state,
 
 $$
-  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta).
+  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
 $$
 
-Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, then $\nabla J(\pi_\theta)$ can be estimated as:
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
 
 $$
-  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \Big( \sum_{t'=t}^H \gamma^{t'-t} r(s_{i, t'}, a_{i, t'}) \Big).
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
 $$
 
-See the appendix for how this is derived. 
-The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return. 
-In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@CITE].
-To avoid this issue, most policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@CITE GAE].
-A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
-
-## Value Learning
-Another way to approach the RL problem is by estimating a value function, and then using the value function to retrieve a policy. 
-Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
-For example, the state-action-value or Q function finds the expected return from a state and action pair under a policy, $\pi$,
+And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
+The advantage function can then be succinctly defined as
 
 $$
-  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg].
+  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
 $$ 
-
-The general Q-learning motivation is that if we know the Q function under certain conditions -- e.g.
-that the state-action space is discrete and that the given policy visits all state-action pairs repeatedly [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-In the case that the state-action space can be tractably represented in a table, one could employ a  algorithm, like temporal-difference control -- see Sutton Ch.
-6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
-Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; for this case, one effective alternative is to use a neural network to approximate the Q-function, a method called deep Q-learning.
-The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
-The Bellman error is a measure of how distant a given value function is from that value function which has been re-evaluated with recent environment interaction.
-Suppose we attempt to approximate the Q-function with a neural network and thus represent this approximate Q-function as as $Q_\phi$, the Bellman error for one state-action pair would be
-
-$$
-  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\phi(s_{t+1}, a) - Q_\phi(s_t, a_t).
-$$
-
-See the appendix/reference for details on how the Bellman Error is derived from eq. [@CITE]. 
-The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across these samples -- with the samples indexed by $i$ --, and then perform the following gradient descent:
-
-$$
-  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 
-$$
-
-When being evaluated, a deep Q-learning agent will select actions by inputting the current state to the Q network and then identifying the state-action pair with the highest estimated Q value.
-The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITE].
-But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @CITE for more insight on these issues.   
-The primary concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@CITE].
-
-## Actor-Critic
-Actor-critic algorithms integrate the main components from both value-learning and policy gradient based methods.
-The "actor" attempts to learn the policy, and the "critic" attempts to learn a value function.
-While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from vanilla policy gradient algorithms by incorporating the critic in its gradient ascent step.
-Instead of estimating the gradient of the return as in eq ..., the gradient of the return is estimated as
-
-$$
-  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \, \hat{Q}(s_{i,t}, a_{i, t}).
-$$
-
-where $\hat{Q}$ is a value function estimated by the critic.
-The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@CITE].
-Consequently, actor-critic algorithms have achieved a large majority of the state of the art results in model-free deep RL [@CITE].
+Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
+Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
 

### 2021-05-13 manuscript/manuscript.Rmd
Some small edits

@@
 The state may be continuous or discrete, completely or partially observed, single or multidimensional.
-For instance, we might represent an ecological community of 10 interacting species using an environment with a 10-dimensional continuously valued state space representing the possible population densities of each species.
-The action space can be similarly complex.
-For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at each of these species during a time step.
+For instance, we might represent an ecological community of 10 interacting species with a 10-dimensional continuously valued state space signifying the population density of each species.
+The action space can be similarly complex, since an action can be continuous or discrete, single or multidimensional.
+For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at the different species over a time step.
 The main focus of an RL environment, however, is in the environment's transition dynamics and reward function.
-The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and inputted action.
-This ability allows for RL environments to model a broad range of decision making problems.
+The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
+This ability allows RL environments to model a broad range of decision making problems.
 For example, we can set the transitions to be deterministic or stochastic.
-We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that require a long sequence of actions to come across, e.g.
+We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that can only be visited after a long sequence of actions, e.g.
 the end point in a maze.
-In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
 Another important consideration in RL is whether to use an episodic or non-episodic environment.
@@
 Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
-For this paper, we only consider episodic environments.
+For this paper, we will only consider episodic environments.
 

### 2021-05-13 manuscript/manuscript.Rmd
First pass at env para complete

@@
 The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
-This allows the design of software agents which can interact and learn in that environment without knowing anything about its internal details.
-The `gym` python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
+This allows the design and application of software agents which can interact and learn in that environment without knowing anything about the environment's internal details.
+The `gym` Python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
 The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').

### 2021-05-13 manuscript/manuscript.Rmd
edits to ethics/power paragraph

--- 2021-05-13 manuscript/manuscript.Rmd
@@
 
+<!-- something to introduce this section/transition from the intro might be helpful: "We start with a brief overview of RL problems and terminology:" -->
+
 Any application of RL can be divided into two components: an *environment* and an *agent*.
 An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
-Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
-At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
+Over the course of an episode, an *agent*, which is often a computer program, continuously interacts with the environment. 
+At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
 The goal of the agent is to learn how to maximize its expected cumulative reward.
@@
 In training, the agent *explores* the available states and actions. 
-Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
+Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
+
+<!-- ^ this is great! I do think keeping a table of these terms would be helpful (and refering to that table in the text here) -->
 
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
-Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-*Policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
+By contrast, *Policy gradient* algorithms exclusively learn the policy.
+*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
 
@@
 
-An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
-<!-- ML: what are classical approaches below here? -->
-In contrast to classical approaches, there are few restrictions on what comprises a state or action. 
-The state may be continuous or discrete, completely or partially observed, single or multidimensional.
-For instance, we might represent an ecological community of 10 interacting species using an environment with a 10-dimensional continuously valued state space representing the possible population densities of each species.
-The action space can be similarly complex.
-For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at each of these species during a time step.
-The main focus of an RL environment, however, is in the environment's transition dynamics and reward function.
-The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and inputted action.
-This ability allows for RL environments to model a broad range of decision making problems.
-For example, we can set the transitions to be deterministic or stochastic.
-We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that require a long sequence of actions to come across, e.g.
-the end point in a maze.
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
+In contrast to classical approaches, such as dynamic programming, there are few restrictions on what comprises a state or action, at least in principle.
+The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
+For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
+The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
+The reward will in general depend on both the current state and choice of actions.
+An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
+In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
 In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
-Another important consideration in RL is whether to use an episodic or non-episodic environment.
-In episodic environments, there is a finite number of time steps during which the agent interacts with the environment, and for these environments, the designer must specify a stopping condition.
-Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
-For this paper, we only consider episodic environments.
 
@@
 
+<!-- second pass at this -->
+
+<!-- from end of Carl's discussion edits: Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.
+-->
+<!-- If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.
+-->
+
+The increasingly influential role of AI in sectors as diverse as policing, finance, and employment has raised concern about the ethics, biases, and power entrenched in both decision-making algorithms and their outputs [@kalluri2020].
+As algorithms play a greater role in environmental and conservation policy, will similar issues arise?
+While this question has garnered  recent interest in the academic literature [@wearn2019; @adams2019; @scoville2021], the use of decision making algorithms in environmental contexts is still relatively isolated and in it's early stages.
+Examples of AI applications, and corresponding concerns, have largely been confined to the context of enforcing environmental laws (i.e.
+algorithms which predict wildlife poaching events [@adams2019] or identify illegal fishing behavior [@mcdonald2021; @swartz2021]).
+However, more diverse applications of AI algorithms - such as applications which contribute to meeting sustainable development goals [@vinuesa2020] or designing  protected area networks - raise additional concerns of entrenching power asymmetries between those reliant on natural resources and those designing algorithms and their objectives.
+These issues of power become particularly acute when private industry holds the keys to propriety algorithms or environmental data, and generally operate without the same community engagement obligations of public agencies.
+In these cases, good intentions might not be sufficient to ensure equitable outcomes. 
+Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges promotes equity in both the procedural process of developing objective functions and the distribution of benefits that result from decisions.
+
 # a vision going forward

### 2021-05-13 manuscript/manuscript.Rmd
Getting into intro agent para

--- 2021-05-13 manuscript/manuscript.Rmd
@@
 
+<!-- Write paragraph on classic approaches then compare to deep RL very broadly-->
+
 Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
@@
 Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g.
+Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
 when the possible states and actions can not be represented in a tractable table [@CITE].
@@
 ## Designing and training an agent
+<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
 
-The training an agent on an environment involves allowing the agent to experience and respond to repeated simulations from the environment, thousands or even millions of times. 
-Over the course of training, weights on the neural network are updated in response to the reward given by the environment, much like reinforcement learning is believed to work in living creatures [@cite].
+Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 
+Over the course of training, the agent will engage in a trial and error process whereby the agent continually updates weights on a collection of neural networks in response to the environment's reward signal, much like how reinforcement learning is believed to work in living creatures [@cite].
+Fundamental to this trial and error process is a balancing of explorative and exploitative behavior.
+The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
+Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
+As the exploration-exploitation is currently unsolved, there are a range of approaches that RL algorithms employ.
+
+Model-free algorithms can explore the environment in either an off-policy or on-policy manner. 
+The distinction here comes from what policy the agent is using during training.
+For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algoirthms, the agent uses the same policy in training as the agent will use during evaluation.
+On top of this distinction, agents can learn either a deterministic or stochastic policy.
+
 A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.

### 2021-05-13 manuscript/manuscript.Rmd
Small comments

@@
 <!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
-
+<!-- Ecologist job to build environment; CS job to build agent 
+      Ok so talk about this in very general terms
+      selecting an agent, why you might want to use different ones
+-->
 Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 

### 2021-05-13 manuscript/manuscript.Rmd

@@
 
-Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
-There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
-The state-value function is defined as the expected return from a given state,
+## Policy Gradients
+The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
+There are both gradient-free and gradient-based policy search methods that have been successful in practice [@CITE; @CITE]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
+In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters. 
+At each gradient step, the following update is performed,
 
 $$
-  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta).
 $$
 
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
+Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, then $\nabla J(\pi_\theta)$ can be estimated as:
 
 $$
-  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \Big( \sum_{t'=t}^H \gamma^{t'-t} r(s_{i, t'}, a_{i, t'}) \Big).
 $$
 
-And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succinctly defined as
+See the appendix for how this is derived. 
+The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return. 
+In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@CITE].
+To avoid this issue, most policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@CITE GAE].
+A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
+
+## Value Learning
+Another way to approach the RL problem is by estimating a value function, and then using the value function to retrieve a policy. 
+Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
+For example, the state-action-value or Q function finds the expected return from a state and action pair under a policy, $\pi$,
 
 $$
-  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg].
 $$
-Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
-Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
+
+The general Q-learning motivation is that if we know the Q function under certain conditions -- e.g.
+that the state-action space is discrete and that the given policy visits all state-action pairs repeatedly [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+In the case that the state-action space can be tractably represented in a table, one could employ a  algorithm, like temporal-difference control -- see Sutton Ch.
+6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
+Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; for this case, one effective alternative is to use a neural network to approximate the Q-function, a method called deep Q-learning.
+The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
+The Bellman error is a measure of how distant a given value function is from that value function which has been re-evaluated with recent environment interaction.
+Suppose we attempt to approximate the Q-function with a neural network and thus represent this approximate Q-function as as $Q_\phi$, the Bellman error for one state-action pair would be
+
+$$
+  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\phi(s_{t+1}, a) - Q_\phi(s_t, a_t).
+$$
+
+See the appendix/reference for details on how the Bellman Error is derived from eq. [@CITE]. 
+The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across these samples -- with the samples indexed by $i$ --, and then perform the following gradient descent:
+
+$$
+  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 
+$$
+
+When being evaluated, a deep Q-learning agent will select actions by inputting the current state to the Q network and then identifying the state-action pair with the highest estimated Q value.
+The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITE].
+But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @CITE for more insight on these issues.   
+The primary concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@CITE].
+
+## Actor-Critic
+Actor-critic algorithms integrate the main components from both value-learning and policy gradient based methods.
+The "actor" attempts to learn the policy, and the "critic" attempts to learn a value function.
+While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from vanilla policy gradient algorithms by incorporating the critic in its gradient ascent step.
+Instead of estimating the gradient of the return as in eq ..., the gradient of the return is estimated as
+
+$$
+  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \, \hat{Q}(s_{i,t}, a_{i, t}).
+$$
+
+where $\hat{Q}$ is a value function estimated by the critic.
+The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@CITE].
+Consequently, actor-critic algorithms have achieved a large majority of the state of the art results in model-free deep RL [@CITE].
 

### 2021-05-14 manuscript/manuscript.Rmd
Cutting and moving

--- 2021-05-13 manuscript/manuscript.Rmd
@@
 
-Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
-Generally, model-free agents are either learning a policy function, a value function or both.
-For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
-If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
-The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-*Policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
-
 ## Designing and building an environment
@@
 
-## Designing and training an agent
+## Training and tuning an agent
 <!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
-
+<!-- Ecologist job to build environment; CS job to build agent 
+      Ok so talk about this in very general terms
+      selecting an agent, why you might want to use different ones
+-->
 Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 
@@
 
-A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.
-A typical neural network in RL may have thousands or hundreds of thousands of parameters, giving it the flexibility to learn very complex behaviors.
-The overall architecture of the neural network, as well as other parameters such as those governing how an agent learns, how an agent proposes actions, and how the agent discounts future rewards, and so forth are controlled by *hyperparameters*.
-To initialize a new, untrained agent, we must specify a handful or so of these hyperparameters, which are held fixed over the course of training.
-The particular selection of hyperparameters required depends on which *RL algorithm* our agent will be based.
-The RL algorithm specifies precisely how the agent will learn from its environment:
-Is the algorithm model-based or model-free? 
-Stochastic or deterministic? 
-"Off-policy" or "on-policy"? 
-Does it use a random noise model to propose actions?
-Does it permit a continuous action space or require discrete actions?
-The design of RL algorithms has been and remains an area of intense research and rapid development [Table 2].
-Open source implementations by professional software development teams often follow swiftly after new algorithms are published [@Ssb3; @tf-agents].
-In the following section, we summarize some of the key concepts that make up an RL algorithm, and highlight relevant distinctions between important algorithms today.
-
-<!-- Something about given RL algorithm being able to solve lots of different environments with no pre-specified knowledge. cite examples @AlphaZero and @DQN; but still depends on hyperparameters -->
-The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
+In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
+For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
+Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
 The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
-Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
-Considerations in how the environment simulation is implemented, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
 
@@
 While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
-
-Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
-There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
-The state-value function is defined as the expected return from a given state,
-
-$$
-  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-
-$$
-  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succinctly defined as
-
-$$
-  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-$$ 
-Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
-Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
 
@@
 There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
-RL algorithms can generally be split into 2 groups, model-based and model-free RL.
-In model-based RL, the agent is attempting to learn the state transition operator, $T$. <!-- Sometimes the agent is just assuming this is given, right? -->
-Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
-<!-- For most recent methods focus on an approximate solution, since optimal is computationally impossible.
-In contrast most readers will assume this step is trivial, if the transition matrix is known there's nothing more to do but "look" at the predicted transitions and make a choice.
--->
-In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
-The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
 

### 2021-05-14 manuscript/appendix.Rmd
Generally should have things in the right places

@@
 
+# RL
+
+<!-- Not tightened up yet to make sense in the appendix, just migrating more nuts and bolts RL stuff
+     from manuscript here for sake of word count
+-->
+
+## Policy Gradients
+The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
+There are both gradient-free and gradient-based policy search methods that have been successful in practice [@CITE; @CITE]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
+In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters. 
+At each gradient step, the following update is performed,
+
+$$
+  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta).
+$$
+
+Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, then $\nabla J(\pi_\theta)$ can be estimated as:
+
+$$
+  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \Big( \sum_{t'=t}^H \gamma^{t'-t} r(s_{i, t'}, a_{i, t'}) \Big).
+$$
+
+See the appendix for how this is derived. 
+The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return. 
+In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@CITE].
+To avoid this issue, most policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@CITE GAE].
+A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
+
+## Value Learning
+Another way to approach the RL problem is by estimating a value function, and then using the value function to retrieve a policy. 
+Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
+For example, the state-action-value or Q function finds the expected return from a state and action pair under a policy, $\pi$,
+
+$$
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg].
+$$
+
+The general Q-learning motivation is that if we know the Q function under certain conditions -- e.g.
+that the state-action space is discrete and that the given policy visits all state-action pairs repeatedly [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+In the case that the state-action space can be tractably represented in a table, one could employ a  algorithm, like temporal-difference control -- see Sutton Ch.
+6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
+Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; for this case, one effective alternative is to use a neural network to approximate the Q-function, a method called deep Q-learning.
+The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
+The Bellman error is a measure of how distant a given value function is from that value function which has been re-evaluated with recent environment interaction.
+Suppose we attempt to approximate the Q-function with a neural network and thus represent this approximate Q-function as as $Q_\phi$, the Bellman error for one state-action pair would be
+
+$$
+  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\phi(s_{t+1}, a) - Q_\phi(s_t, a_t).
+$$
+
+See the appendix/reference for details on how the Bellman Error is derived from eq. [@CITE]. 
+The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across these samples -- with the samples indexed by $i$ --, and then perform the following gradient descent:
+
+$$
+  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 
+$$
+
+When being evaluated, a deep Q-learning agent will select actions by inputting the current state to the Q network and then identifying the state-action pair with the highest estimated Q value.
+The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITE].
+But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @CITE for more insight on these issues.   
+The primary concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@CITE].
+
+## Actor-Critic
+Actor-critic algorithms integrate the main components from both value-learning and policy gradient based methods.
+The "actor" attempts to learn the policy, and the "critic" attempts to learn a value function.
+While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from vanilla policy gradient algorithms by incorporating the critic in its gradient ascent step.
+Instead of estimating the gradient of the return as in eq ..., the gradient of the return is estimated as
+
+$$
+  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \, \hat{Q}(s_{i,t}, a_{i, t}).
+$$
+
+where $\hat{Q}$ is a value function estimated by the critic.
+The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@CITE].
+Consequently, actor-critic algorithms have achieved a large majority of the state of the art results in model-free deep RL [@CITE].
+
+## Value Functions
+
+There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
+The state-value function is defined as the expected return from a given state,
+
+$$
+  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+$$
+
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
+
+$$
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+$$
+
+And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
+The advantage function can then be succinctly defined as
+
+$$
+  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
+$$ 
+Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
+Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
 ----

### 2021-05-14 manuscript/manuscript.Rmd
Generally should have things in the right places

@@
 
-To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
-Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
-Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
+When the environment is fully observable, i.e.
+when the state that the agent observes fully characterizes the system, the reinforcement learning problem is typically formalized as a Markov Decision Process (MDP).
+A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
+After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
+As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
+From these definitions, we can provide an agent's trajectory distribution for a given policy as,
 
-<!-- Write paragraph on classic approaches then compare to deep RL very broadly-->
+$$
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) \, T(s_{t+1} \mid s_t, a_t)
+$$
+
+When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
+An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
+This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
+
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
+
+$$
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
+$$
+
+Note that this formalism extends to an infinite horizon case when $T= \infty$.
+While there are RL methods to deal with infinite horizon cases, we will only finite horizon or episodic MDP and POMDPs.
 
@@
 In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
-Another important consideration in RL is whether to use an episodic or non-episodic environment.
-In episodic environments, there is a finite number of time steps during which the agent interacts with the environment, and for these environments, the designer must specify a stopping condition.
-Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
-For this paper, we will only consider episodic environments.
 
@@
 
-## Training and tuning an agent
-<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
-<!-- Ecologist job to build environment; CS job to build agent 
-      Ok so talk about this in very general terms
-      selecting an agent, why you might want to use different ones
--->
-Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 
-Over the course of training, the agent will engage in a trial and error process whereby the agent continually updates weights on a collection of neural networks in response to the environment's reward signal, much like how reinforcement learning is believed to work in living creatures [@cite].
-Fundamental to this trial and error process is a balancing of explorative and exploitative behavior.
-The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
-Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
-As the exploration-exploitation is currently unsolved, there are a range of approaches that RL algorithms employ.
-
-Model-free algorithms can explore the environment in either an off-policy or on-policy manner. 
-The distinction here comes from what policy the agent is using during training.
-For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algoirthms, the agent uses the same policy in training as the agent will use during evaluation.
-On top of this distinction, agents can learn either a deterministic or stochastic policy.
-
-In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
-For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
-Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
-The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
-
-<!--
-A rapidly growing set of material for learning about deep reinforcement learning is now freely available.... (maybe just describe this in the appendix)
--->
-
-<!-- 
-Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
-Make a nice table box in latex instead if we're keeping this
--->
-\begin{figure}
-\centering
-\includegraphics[width=0.9\linewidth]{table1.pdf}
-\end{figure}
-
 # RL Algorithms
-
-<!--
-
-I think we can lean more heavily both in references to and paralleling the presentation of some of the existing background material.
-
--->
-
-<!-- Still can't decide if this statement is too strong; i.e. that at least theoretically / LSTM, RL permits learning processes which are not Markovian in the observed variables -->
-When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem is conventionally formalized as a Markov Decision Process (MDP).
-A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
-From these definitions, we can provide an agent's trajectory distribution for a given policy as,
-
-$$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
-$$
-
-When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
-An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
-This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
-
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
-
-$$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
-$$
-
-Note that this formalism extends to an infinite horizon case when $T= \infty$.
-While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
 
@@
 
+To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
+Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents throughout the rest of the paper.
+
+<!-- classic algorithms transition to neural networks>
+
+<!-- These 2 paragraphs might make it back to the manuscript but going to keep them here for now -->
+Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
+*Policy gradient* algorithms exclusively learn the policy.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
+
+Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
+Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
+when the possible states and actions can not be represented in a tractable table [@CITE].
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
+Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
+While there are other function approximators that can be used in RL, e.g.
+Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
+We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
+
 A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
@@
 Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop]. <!-- not clear how "stable" is defined/measured -->
-
-The field of reinforcement learning has proliferated over the past 10 years on account of contributions from deep learning.
-In every reinforcement learning algorithm, the agent is attempting to learn some function, whether it is a policy function and/or a value function.
-Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions whose form is unknown.
-Additionally, the ability of neural networks to autonomously devise low-dimensional representations of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
-<!-- Yes!
-should stress why that matters -->
-The typical process for using neural networks in RL is that the agent is attempting to optimize a policy and/or value network using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
-<!-- Fixme: needs to be clear that NN parameters are being fit by the "training" step.
-No point mentioning back-propagation, evolutionary algorithms, or gradient methdos if we are not going to define them.
--->
-
-Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable with classical methods.
-<!-- non-tabular undefined -->
-
-In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
-For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
-Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
-The hyperparameter tuning process consists of sampling then evaluating a set of hyperparameters, where tuning trials are often ranked by mean cumulative reward.
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
 
@@
 ```
+
+## Training and tuning an agent
+<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
+<!-- Ecologist job to build environment; CS job to build agent 
+      Ok so talk about this in very general terms
+      selecting an agent, why you might want to use different ones
+-->
+Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 
+Over the course of training, the agent will engage in a trial and error process whereby the agent continually updates weights on a collection of neural networks in response to the environment's reward signal, much like how reinforcement learning is believed to work in living creatures [@cite].
+Fundamental to this trial and error process is a balancing of explorative and exploitative behavior.
+The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
+Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
+As the exploration-exploitation is currently unsolved, there are a range of approaches that RL algorithms employ.
+
+Model-free algorithms can explore the environment in either an off-policy or on-policy manner. 
+The distinction here comes from what policy the agent is using during training.
+For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algoirthms, the agent uses the same policy in training as the agent will use during evaluation.
+On top of this distinction, agents can learn either a deterministic or stochastic policy.
+
+In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
+For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
+Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
+The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
+
+<!--
+A rapidly growing set of material for learning about deep reinforcement learning is now freely available.... (maybe just describe this in the appendix)
+-->
+
+<!-- 
+Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
+Make a nice table box in latex instead if we're keeping this
+-->
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{table1.pdf}
+\end{figure}
+
 # Examples

### 2021-05-14 manuscript/manuscript.Rmd
drop bit on conservation policy actions being too simple

--- 2021-05-14 manuscript/manuscript.Rmd
@@
 An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
-Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
-At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
+Over the course of an episode, an *agent*, which is often a computer program, continuously interacts with the environment. 
+At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
 The goal of the agent is to learn how to maximize its expected cumulative reward.
@@
 In training, the agent *explores* the available states and actions. 
-Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
+Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.
 
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
-Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 
-<!-- Write paragraph on classic approaches then compare to deep RL very broadly-->
+Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
+*Policy gradient* algorithms exclusively learn the policy.
+*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
+Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.
 
@@
 
-An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
-<!-- ML: what are classical approaches below here? -->
-In contrast to classical approaches, there are few restrictions on what comprises a state or action. 
-The state may be continuous or discrete, completely or partially observed, single or multidimensional.
-For instance, we might represent an ecological community of 10 interacting species with a 10-dimensional continuously valued state space signifying the population density of each species.
-The action space can be similarly complex, since an action can be continuous or discrete, single or multidimensional.
-For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at the different species over a time step.
-The main focus of an RL environment, however, is in the environment's transition dynamics and reward function.
-The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
-This ability allows RL environments to model a broad range of decision making problems.
-For example, we can set the transitions to be deterministic or stochastic.
-We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that can only be visited after a long sequence of actions, e.g.
-the end point in a maze.
-In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
-Another important consideration in RL is whether to use an episodic or non-episodic environment.
-In episodic environments, there is a finite number of time steps during which the agent interacts with the environment, and for these environments, the designer must specify a stopping condition.
-Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
-For this paper, we will only consider episodic environments.
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
+In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
+The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
+For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
+The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
+The reward will in general depend on both the current state and choice of actions.
+An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
+In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
+In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
 
@@
 The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
-This allows the design and application of software agents which can interact and learn in that environment without knowing anything about the environment's internal details.
-The `gym` Python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
+This allows the design of software agents which can interact and learn in that environment without knowing anything about its internal details.
+The `gym` python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
 The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').
@@
 
-## Training and tuning an agent
-<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
-<!-- Ecologist job to build environment; CS job to build agent 
-      Ok so talk about this in very general terms
-      selecting an agent, why you might want to use different ones
--->
-Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 
-Over the course of training, the agent will engage in a trial and error process whereby the agent continually updates weights on a collection of neural networks in response to the environment's reward signal, much like how reinforcement learning is believed to work in living creatures [@cite].
-Fundamental to this trial and error process is a balancing of explorative and exploitative behavior.
-The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
-Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
-As the exploration-exploitation is currently unsolved, there are a range of approaches that RL algorithms employ.
+## Designing and training an agent
 
-Model-free algorithms can explore the environment in either an off-policy or on-policy manner. 
-The distinction here comes from what policy the agent is using during training.
-For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algoirthms, the agent uses the same policy in training as the agent will use during evaluation.
-On top of this distinction, agents can learn either a deterministic or stochastic policy.
+The training an agent on an environment involves allowing the agent to experience and respond to repeated simulations from the environment, thousands or even millions of times. 
+Over the course of training, weights on the neural network are updated in response to the reward given by the environment, much like reinforcement learning is believed to work in living creatures [@cite].
+A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.
+A typical neural network in RL may have thousands or hundreds of thousands of parameters, giving it the flexibility to learn very complex behaviors.
+The overall architecture of the neural network, as well as other parameters such as those governing how an agent learns, how an agent proposes actions, and how the agent discounts future rewards, and so forth are controlled by *hyperparameters*.
+To initialize a new, untrained agent, we must specify a handful or so of these hyperparameters, which are held fixed over the course of training.
+The particular selection of hyperparameters required depends on which *RL algorithm* our agent will be based.
+The RL algorithm specifies precisely how the agent will learn from its environment:
+Is the algorithm model-based or model-free? 
+Stochastic or deterministic? 
+"Off-policy" or "on-policy"? 
+Does it use a random noise model to propose actions?
+Does it permit a continuous action space or require discrete actions?
+The design of RL algorithms has been and remains an area of intense research and rapid development [Table 2].
+Open source implementations by professional software development teams often follow swiftly after new algorithms are published [@Ssb3; @tf-agents].
+In the following section, we summarize some of the key concepts that make up an RL algorithm, and highlight relevant distinctions between important algorithms today.
 
-In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
-For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
-Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
+<!-- Something about given RL algorithm being able to solve lots of different environments with no pre-specified knowledge. cite examples @AlphaZero and @DQN; but still depends on hyperparameters -->
+The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
 The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
+Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
+Considerations in how the environment simulation is implemented, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.
 
@@
 While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
+
+Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
+Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
+There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
+The state-value function is defined as the expected return from a given state,
+
+$$
+  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+$$
+
+Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
+
+$$
+  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
+$$
+
+And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
+The advantage function can then be succinctly defined as
+
+$$
+  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
+$$ 
+Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
+Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
+Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
 
@@
 There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
+RL algorithms can generally be split into 2 groups, model-based and model-free RL.
+In model-based RL, the agent is attempting to learn the state transition operator, $T$. <!-- Sometimes the agent is just assuming this is given, right? -->
+Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
+<!-- For most recent methods focus on an approximate solution, since optimal is computationally impossible.
+In contrast most readers will assume this step is trivial, if the transition matrix is known there's nothing more to do but "look" at the predicted transitions and make a choice.
+-->
+In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
+The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
 
@@
 
-```{=html}
-<!--
+<!-- Ecology has complex decisions and we lack necessary tools; RL could be it -->
+Ecological challenges facing the planet today are complex and outcomes are both uncertain and consequential.
+Even our best models and best research will never provide a crystal ball to the future, only better elucidate possible scenarios.
+Consequently, that research must also confront the challenge of making robust, resilient decisions in a changing world.
+The science of ecological management and quantitative decision-making has a long history [e.g.
+@Schaefer1954; @Walters1974] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
+Complexity comes not only from the ecological processes but also the available actions.
+Conservation efforts can be implemented in a wide variety of ways, and the most effective policies may vary in both space and time.
+While such complex dynamics and large state and action spaces are intractable to classical methods, they are straight forward to implement in simulation environments such as those used to train an RL algorithm.
+Neural networks used in deep RL have proven remarkably effective in handling such complexity, particularly when leveraging immense computing resources increasingly available through advances in hardware and software [@gpu_computing].
 
-1. Return to theme of RL vs supervised/unsupervised learning.
+<!-- Model free is great bc we don't know the true models -->
+The rapidly expanding class of Model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
+Rarely do we know the underlying functional forms for ecological processes -- is population recruitment more Beverton-Holt or Ricker?
+Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
+Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
+Model-free RL approaches offer a powerful way to factor this thorny issue.
+The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
+Alternative models can easily be implemented in simulations, allowing an agent to train against both contingencies.
+A successful agent will learn a policy that works well in both scenarios.
+This is analogous to classical approaches for integrating over parameter or model uncertainty [e.g. @Walters1981; @Ludwig1982],
 
--->
-```
-<!-- Something about relevance to both the _theory_ and practice of ecological conservation.
-(e.g.
-"strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->
-
-## Open Problems
-
+<!-- Real problems are harder than these examples -->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 
--   wild west of AI?
-    -can separate RL from having gyms implemented -\> how do you compare different models, -mention that customization is probably the way to go, baselines only go so far
+<!-- But never fear, hard problems have been solved before!-->
+Deep RL is still a very young field, where despite several landmark successes, potential far outstrips practice.
+Recent advances in the field have proven the potential of the approach to solve complex problems [@alphaGo2016; @alphaGoZero; @alphazero; @DQN], but typically leveraging large teams with decades of experience in ML and millions of dollars worth of computing power [@alphaGoZero].
+Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
+Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [@gym], and performance can easily be compared against alternatives on public leaderboards [@gym].
+Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].
+Advances in both computational power and algorithmic efficiency continue to reduce costs at a dramatic pace, significantly faster than Moore's law [@openai_costs].
+By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may both be better able to replicate that collaborative, iterative success in cracking hard problems.
+Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.
+If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.
 
--   AI ethics, transparency
+<!-- transition from big promise to perils -- Millie's section on risks here -->
 
-# a vision going forward
+<!-- Something about the way forward, point to appendix resources etc -->
 
--   our public/open source environments and agents, leaderboard
--   the need for collaborative/competitive development both to make better benchmark environments and better agents
+<!-- Discussion really doesn't discuss/draw conclusions from the fisheries and tipping points examples! Do we need to or maybe not since it's not about the science-->
+
+<!-- 
+The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
+We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
+We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
+
+-->

### 2021-05-14 manuscript/manuscript.Rmd
No major edits, mostly notes to self

--- 2021-05-14 manuscript/manuscript.Rmd
@@
 To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
@@
 
-<!-- classic algorithms transition to neural networks>
+<!-- 
+- Say that classically TD learning and dynamic program were common approaches to solving RL; but they fail with large state-action spaces
+- over past 10 years, neural network based rl algorithms have succeeded on large state-action spaces
+
+- on-policy vs off-policy, exploration vs exploitation
+
+other algorithm tidbits:
+- stochastic vs. determinstic 
+- all deep rl algorithms are formulated to work on different action/state spaces
+-->
 
@@
 If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
-The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-*Policy gradient* algorithms exclusively learn the policy.
+The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
+In contrast, *policy gradient* algorithms exclusively learn the policy.
 And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
 There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
 

### 2021-05-14 manuscript/manuscript.Rmd
Coming around on RL algorithms section

@@
 
-# RL Algorithms
+# Deep RL Algorithms
 
@@
 Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
-Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents throughout the rest of the paper.
+Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
 
-<!-- 
-- Say that classically TD learning and dynamic program were common approaches to solving RL; but they fail with large state-action spaces
-- over past 10 years, neural network based rl algorithms have succeeded on large state-action spaces
-
-- on-policy vs off-policy, exploration vs exploitation
-
-other algorithm tidbits:
-- stochastic vs. determinstic 
-- all deep rl algorithms are formulated to work on different action/state spaces
--->
-
-<!-- These 2 paragraphs might make it back to the manuscript but going to keep them here for now -->
 Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
@@
 And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
 Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
@@
 Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the agent's neural network that it is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
 
-A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
-If an agent uses the same policy to take actions as the policy that the agent attempts to evaluate or improve, then this algorithm is on-policy.
-<!-- Unclear what "to evaluate or improve" means.
-Perhaps an example would help?
--->
-If an agent uses a different policy to select actions than the policy that the agent evaluates or improves, then this algorithm is off-policy.
-The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [cite SAC paper].
-<!-- Not clear what this means -->
-Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop]. <!-- not clear how "stable" is defined/measured -->
+Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit former high rewarding sequences.
+The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
+Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
+The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
+This *exploration-exploitation dilemma* is an open research problem [@CITE], and current deep RL algorithms take a range of approaches towards exploration -- in the appendix, we discuss some of the more specific exploration algorithms.
+Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
+The distinction between on and off-policy exploration comes from what policy the agent is using during training.
+For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
+There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the appendix.
 
-<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
+On top of all the already mentioned possible distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
+One important point of divergence is whether the agent learns a deterministic or stochastic policy.
+Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous action space.
+In table **, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
+In the appendix, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
 
@@
 Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
-Over the course of training, the agent will engage in a trial and error process whereby the agent continually updates weights on a collection of neural networks in response to the environment's reward signal, much like how reinforcement learning is believed to work in living creatures [@cite].
-Fundamental to this trial and error process is a balancing of explorative and exploitative behavior.
-The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
-Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
-As the exploration-exploitation is currently unsolved, there are a range of approaches that RL algorithms employ.
-
-Model-free algorithms can explore the environment in either an off-policy or on-policy manner. 
-The distinction here comes from what policy the agent is using during training.
-For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algoirthms, the agent uses the same policy in training as the agent will use during evaluation.
-On top of this distinction, agents can learn either a deterministic or stochastic policy.
 

### 2021-05-14 manuscript/manuscript.Rmd
reworking 'issues' paragraphs (WIP)

--- 2021-05-14 manuscript/manuscript.Rmd
@@
 authors:
-
   - name: Marcus Lapeyrolerie
@@
 \begin{figure}
-\centering
-\includegraphics[width=0.5\linewidth]{fig1.pdf}
-\caption{Reinforcement Learning focuses on the repeated feedback of agent interacting with an \emph{environment}.
-The agent performs an action which influences the environment's future \emph{state}.
-The agent receives an observation of the new state and reward from the environment.}
+\includegraphics[width=\linewidth]{fig1.pdf}
+\caption{Deep Reinforcement Learning:  An artificial intelligence \emph{agent} uses a \emph{neural network} to select an \emph{action} in response to an \emph{observation} of the \emph{environment}, receiving a \emph{reward} (or punishment) from the environment as a result.
+The agent is \emph{trained} to maximize the reward through the experience of trying different actions over repeated \emph{epsiodes} of interacting with the environment.
+Over the course of training, the agent adjusts parameter weights on the neural network as it learns which actions lead to the best cumulative rewards.
+A researcher may \emph{tune} an agent by considering iterative trials that train different agents with different \emph{hyperparameters}, such as different neural network architectures.
+An agent trained with the most successful combination of hyperparameters can then be tested in an evaluation environment.}
 \end{figure}
@@
 <!-- Solving 'real' problems will be hard -->
-However, successfully applying RL to complex ecological problems is no easy task.
+Successfully applying RL to complex ecological problems is no easy task.
 The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance is another task.
@@
 Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
-<!-- something about several technical caveats: instability, interpretability reward design-->
+In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: instability, interpretability, and reward design.
 
 <!-- reason 1: instability-->
-One leading problem with deep RL is instability.
+One leading challenge for deep RL today is the problem of instability.
+RL agents trained on narrowly defined environments may perform poorly even when the environment is altered in seemingly trivial way, such as a re-scaling of parameters.
+To avoid this, researchers should seek to both train and evaluate agents against a range of environments representing different parameters or processes dynamics.
+Stable performance can be challenging even when using the same environment.
 For a range of reasons, including high variance [@cite] and the exploration vs.
 exploitation trade-off [@cite], model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
-For instance, it is not uncommon to observe vastly different behavior when two agents are initialized with different random seeds [cite ibid].
-Since model-based and model-free algorithms typically have a large number of hyperparameters, each of which can contribute to training instability, it is practically impossible to identify the source of instability.
-The hope of the RL practitioner is that after doing an exhaustive hyperparameter search, the practitioner will find a set of hyperparameters that results in a robust, well-performing agent, but there are no guarantees that hyperparameter tuning will lead to this outcome.
-<!--CB: I think we can say more about how to diagnose this problem if not fix it.
-    Likewise should address the importance of assessing robustness across multiple models/environments, probably at least as important an issue as robustness to nuciance hyperparameters like random seed.
+For instance, it is not uncommon to observe vastly different behavior when an agent is initialized with different random seeds [cite ibid].
+Such unstable behavior can often be avoided by alternate selection of hyperparameters, so researchers should consider stability as well as performance during hyperparameter selection.
+The choice of RL algorithm, design of the reward function, and duration of training can also influence stability.
+
+<!--
+Under some conditions, RL agents can prove adept at _transfer learning_, in which an agent trained in one environment performs well under a different environment.
 -->
+<!-- if we're really recommending 'testing' for stability, maybe we need to demo that in the appendix-->
 
 <!-- reason 2: interpretability -->
-In the context of conservation, we anticipate that there will be a host of specific challenges to using deep RL methods.
-Interpretability is likely to be the biggest barrier to adoption. <!-- don't need to rank them -->
+Interpretability is a common challenge ot the adoption of most ML methods [@cite]. 
 Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
-Another challenge is that DRL algorithms are very sample inefficient, and to overcome sample inefficiency for conservation decision-making problems that are often data poor, relying on simulations to train agents will be necessary.
-While there are positives to using simulated data, any assumptions or biases that are included in the model will determine agent behavior.
+As our examples have illustrated, it may be possible to gain intuition on an agent's behavior by visualizing the action selected against each possible state, though this becomes more difficult in large problems.
+Another challenge is that Deep RL algorithms are very sample inefficient, all though progress is being made to improve this [@cite].  
+-->
 
@@
 An issue that is central to any applied optimization problem is using the right objective.
-There have been many examples in which explicit reward design goes horribly wrong [cite].
+There have been many examples in which explicit reward design goes very wrong [cite].
 And for ecological systems, which are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
@@
 who gets to specify the reward function and will the values of the greater community be represented in the reward function.
+RL researchers often introduce modified reward functions during training that help an agent perform better on the original objective [@cite].
+This approach is particularly common when rewards are sparse, such as in a game of Chess where the only objective that counts is whether or not you win, but during training, an RL algorithm may reward the agent for intermediate steps such as capturing a piece.
+
+<!--
 One recently proposed framework to overcome challenges with explicit reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
-Yet, in the context of training an agent to say conserve a species, there are questions that could easily arise, foremost will the agent infer that humans don't care about a certain species so acts in a way that leads to that species' extinction?
 There is a definite need for exploration of the ethics of using DRL for conservation.
-
+-->
 <!-- reason 4/additional reasons, like sparse rewards -->

### 2021-05-14 manuscript/manuscript.Rmd
Finishing training and tuning

--- 2021-05-14 manuscript/manuscript.Rmd
@@
 
+## Training and tuning an agent
+<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
+<!-- Ecologist job to build environment; CS job to build agent 
+      Ok so talk about this in very general terms
+      selecting an agent, why you might want to use different ones
+-->
+Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
+The training period involves a trial and error process whereby the agent will perform an action and then receive feedback from the environment in the form of a reward. 
+This process will repeat until the environment transitions to a terminal state, after which the environment will reset to some starting condition.
+According to the algorithm that the agent follows, the agent will explore different state-action trajectories and use its experience to improve its policy.
+The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, the simulation complexity of the environment, model complexity of the agent and more.
+Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@cite], so it is recommended to provide a generous training budget for these agents.
+
+Hyperparameters refer to the parameters in a deep, RL algorithm that are not directly learned during the training process.
+For example, a weight in an agent's neural network is not a hyperparameter, but the number of hidden layers in this network would be a hyperparameter.
+Since the optimal hyperparameters cannot be found analytically [@cite], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
+Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with this set, then examining the resulting performance.
+This procedure can be automated through standard multi-dimensional optimization approaches. 
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
+Hyperparameter tuning, however, is a necessary step in evaluating the performance of an algorithm and should be a standard component of any RL workflow.
+
+<!-- 
+Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
+Make a nice table box in latex instead if we're keeping this
+-->
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{table1.pdf}
+\end{figure}
+
 # Deep RL Algorithms
@@
 ```
-
-## Training and tuning an agent
-<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
-<!-- Ecologist job to build environment; CS job to build agent 
-      Ok so talk about this in very general terms
-      selecting an agent, why you might want to use different ones
--->
-Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
-
-In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
-Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
-For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
-Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
-The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
-
-<!--
-A rapidly growing set of material for learning about deep reinforcement learning is now freely available.... (maybe just describe this in the appendix)
--->
-
-<!-- 
-Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
-Make a nice table box in latex instead if we're keeping this
--->
-\begin{figure}
-\centering
-\includegraphics[width=0.9\linewidth]{table1.pdf}
-\end{figure}
 

### 2021-05-15 manuscript/manuscript.Rmd
Adding bits to convey practicality

@@
 ## Training and tuning an agent
-<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
-<!-- Ecologist job to build environment; CS job to build agent 
-      Ok so talk about this in very general terms
-      selecting an agent, why you might want to use different ones
--->
+
 Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
@@
 This process will repeat until the environment transitions to a terminal state, after which the environment will reset to some starting condition.
-According to the algorithm that the agent follows, the agent will explore different state-action trajectories and use its experience to improve its policy.
+According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its policy.
 The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, the simulation complexity of the environment, model complexity of the agent and more.
 Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@cite], so it is recommended to provide a generous training budget for these agents.
+As practitioners, an advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing much about the environment's transition dynamics or all the inner workings of the algorithm.
 

### 2021-05-15 manuscript/manuscript.Rmd
Getting rid of a problematic sentence

@@
 Hyperparameters refer to the parameters in a deep, RL algorithm that are not directly learned during the training process.
-For example, a weight in an agent's neural network is not a hyperparameter, but the number of hidden layers in this network would be a hyperparameter.
 Since the optimal hyperparameters cannot be found analytically [@cite], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
@@
 Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
-Hyperparameter tuning, however, is a necessary step in evaluating the performance of an algorithm and should be a standard component of any RL workflow.
+Hyperparameter tuning, however, is a necessary step in evaluating the performance of an algorithm and should be a standard component of any deep RL workflow.
 

### 2021-05-15 manuscript/manuscript.Rmd
Small edits

@@
 A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
+After the agent observes the state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
 As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
@@
 For example, we can set the transitions to be deterministic or stochastic.
-We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that can only be visited after a long sequence of actions, e.g.
-the end point in a maze.
-In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+We can also specify the reward function to be *sparse*, whereby a positive reward can only be received after a long sequence of actions, e.g. the end point in a maze.
+In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward as we illustrate in examples here.
 
@@
 
-@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
+@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research; to address these issues, @brockman2016 proposed the creation of the OpenAI `gym` framework.
 The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
@@
 The training period involves a trial and error process whereby the agent will perform an action and then receive feedback from the environment in the form of a reward. 
-This process will repeat until the environment transitions to a terminal state, after which the environment will reset to some starting condition.
+This process will repeat until the environment transitions to a terminal state, after which the environment will reset to an initial state.
 According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its policy.
-The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, the simulation complexity of the environment, model complexity of the agent and more.
+The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more [@cite].
 Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@cite], so it is recommended to provide a generous training budget for these agents.
-As practitioners, an advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing much about the environment's transition dynamics or all the inner workings of the algorithm.
+An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing much about the environment's transition dynamics or all the inner workings of the algorithm.
 
 Hyperparameters refer to the parameters in a deep, RL algorithm that are not directly learned during the training process.
-Since the optimal hyperparameters cannot be found analytically [@cite], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
-Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with this set, then examining the resulting performance.
+Since the optimal hyperparameters vary across environments and can only be found through exhaustive search [@cite], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
+Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with these hyperparameters, then examining the resulting performance.
 This procedure can be automated through standard multi-dimensional optimization approaches. 
@@
 In contrast, *policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
+And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not get into in depth -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
 Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
@@
 
-Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit former high rewarding sequences.
-The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
-Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
+Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit past high rewarding sequences.
+Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen. 
+Yet, the agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
 The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@CITE], and current deep RL algorithms take a range of approaches towards exploration -- in the appendix, we discuss some of the more specific exploration algorithms.
+This *exploration-exploitation dilemma* is an open research problem [@CITE], and current deep RL algorithms take a range of approaches towards this dilemma -- in the appendix, we discuss some of the more specific exploration algorithms.
 Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
-The distinction between on and off-policy exploration comes from what policy the agent is using during training.
-For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
+The distinction between on and off-policy exploration comes from what policy the agent is using to select actions during training.
+For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
 There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the appendix.
 
-On top of all the already mentioned possible distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
+On top of the aforementioned distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
 One important point of divergence is whether the agent learns a deterministic or stochastic policy.
-Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous action space.
+Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
 In table **, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.

### 2021-05-17 manuscript/manuscript.Rmd
Adding references

--- 2021-05-15 manuscript/manuscript.Rmd
@@
 Any application of RL can be divided into two components: an *environment* and an *agent*.
-An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
+An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment.
 Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
@@
 
-When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem is typically formalized as a Markov Decision Process (MDP).
-A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
-From these definitions, we can provide an agent's trajectory distribution for a given policy as,
+To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
+Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
+Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
 
-$$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) \, T(s_{t+1} \mid s_t, a_t)
-$$
-
-When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
-An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
-This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
-
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
-
-$$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
-$$
-
-Note that this formalism extends to an infinite horizon case when $T= \infty$.
-While there are RL methods to deal with infinite horizon cases, we will only finite horizon or episodic MDP and POMDPs.
+<!-- Write paragraph on classic approaches then compare to deep RL very broadly-->
 
@@
 In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
+Another important consideration in RL is whether to use an episodic or non-episodic environment.
+In episodic environments, there is a finite number of time steps during which the agent interacts with the environment, and for these environments, the designer must specify a stopping condition.
+Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
+For this paper, we will only consider episodic environments.
 
@@
 ## Training and tuning an agent
+<!-- Maybe rename this as Training and Tuning an Agent -> how to practically get an agent up and running -->
+<!-- Ecologist job to build environment; CS job to build agent 
+      Ok so talk about this in very general terms
+      selecting an agent, why you might want to use different ones
+-->
+Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps. 
+Over the course of training, the agent will engage in a trial and error process whereby the agent continually updates weights on a collection of neural networks in response to the environment's reward signal, much like how reinforcement learning is believed to work in living creatures [@rl_brain].
+Fundamental to this trial and error process is a balancing of explorative and exploitative behavior.
+The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
+Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
+As the exploration-exploitation is currently unsolved, there are a range of approaches that RL algorithms employ.
 
-Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
-The training period involves a trial and error process whereby the agent will perform an action and then receive feedback from the environment in the form of a reward. 
-This process will repeat until the environment transitions to a terminal state, after which the environment will reset to some starting condition.
-According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its policy.
-The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, the simulation complexity of the environment, model complexity of the agent and more.
-Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@cite], so it is recommended to provide a generous training budget for these agents.
-As practitioners, an advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing much about the environment's transition dynamics or all the inner workings of the algorithm.
+Model-free algorithms can explore the environment in either an off-policy or on-policy manner. 
+The distinction here comes from what policy the agent is using during training.
+For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algoirthms, the agent uses the same policy in training as the agent will use during evaluation.
+On top of this distinction, agents can learn either a deterministic or stochastic policy.
 
-Hyperparameters refer to the parameters in a deep, RL algorithm that are not directly learned during the training process.
-Since the optimal hyperparameters cannot be found analytically [@cite], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
-Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with this set, then examining the resulting performance.
-This procedure can be automated through standard multi-dimensional optimization approaches. 
+In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
+For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
+Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
+The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
 Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
-Hyperparameter tuning, however, is a necessary step in evaluating the performance of an algorithm and should be a standard component of any deep RL workflow.
+
+<!--
+A rapidly growing set of material for learning about deep reinforcement learning is now freely available.... (maybe just describe this in the appendix)
+-->
 
@@
 
-# Deep RL Algorithms
+# RL Algorithms
+
+<!--
+
+I think we can lean more heavily both in references to and paralleling the presentation of some of the existing background material.
+
+-->
+
+<!-- Still can't decide if this statement is too strong; i.e. that at least theoretically / LSTM, RL permits learning processes which are not Markovian in the observed variables -->
+When the environment is fully observable, i.e.
+when the state that the agent observes fully characterizes the system, the reinforcement learning problem is conventionally formalized as a Markov Decision Process (MDP).
+A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
+After the agent observes a state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
+As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
+From these definitions, we can provide an agent's trajectory distribution for a given policy as,
+
+$$
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
+$$
+
+When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
+An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
+This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
+
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
+
+$$
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
+$$
+
+Note that this formalism extends to an infinite horizon case when $T= \infty$.
+While there are RL methods to deal with infinite horizon cases, we will only consider episodic MDP and POMDPs.
 
@@
 
-To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
-Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
+A further distinction to be made amongst RL algorithms is in what policy the agent uses to select actions.
+If an agent uses the same policy to take actions as the policy that the agent attempts to evaluate or improve, then this algorithm is on-policy.
+<!-- Unclear what "to evaluate or improve" means.
+Perhaps an example would help?
+-->
+If an agent uses a different policy to select actions than the policy that the agent evaluates or improves, then this algorithm is off-policy.
+The benefit of off-policy learning is that the agent does not need to collect a sample every time the agent updates its behavior; instead, the agent can learn from past experience, which increases sample efficiency [cite SAC paper].
+<!-- Not clear what this means -->
+Yet, on-policy algorithms tend to be more stable in practice [cite Q-prop]. <!-- not clear how "stable" is defined/measured -->
 
-Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
-Generally, model-free agents are either learning a policy function, a value function or both.
-For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
-If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
-The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
-In contrast, *policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
+The field of reinforcement learning has proliferated over the past 10 years on account of contributions from deep learning.
+In every reinforcement learning algorithm, the agent is attempting to learn some function, whether it is a policy function and/or a value function.
+Neural networks, which have the property of being universal function approximators [cite], are thus a powerful tool in reinforcement learning since neural networks can be used to approximate policy and/or value functions whose form is unknown.
+Additionally, the ability of neural networks to autonomously devise low-dimensional representations of data is very beneficial when state observations are high dimension, for instance in the case of image-based observations.
+<!-- Yes!
+should stress why that matters -->
+The typical process for using neural networks in RL is that the agent is attempting to optimize a policy and/or value network using the backpropagation algorithm -- an exception is made here for evolutionary algorithms which did not employ gradient-based methods.
+<!-- Fixme: needs to be clear that NN parameters are being fit by the "training" step.
+No point mentioning back-propagation, evolutionary algorithms, or gradient methdos if we are not going to define them.
+-->
 
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the agent's neural network that it is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
+Deep RL has excelled in solving sequential decision-making problems that have non-tabular state-action spaces, problems that are typically intractable with classical methods.
+<!-- non-tabular undefined -->
 
-Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit former high rewarding sequences.
-The agent will occasionally need to propose exploratory actions that may result in higher rewards than previously seen. 
-Yet, the agent will also need to occasionally exploit the best perceived sequence of actions which may require marginal improvement to reach optimality.
-The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@CITE], and current deep RL algorithms take a range of approaches towards exploration -- in the appendix, we discuss some of the more specific exploration algorithms.
-Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
-The distinction between on and off-policy exploration comes from what policy the agent is using during training.
-For off-policy algorithms, the agent is using a policy that can be different than the policy the agent will use during evaluation to select actions; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
-There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the appendix.
+In practice, the general process for attaining a good performing agent is to tune model hyperparameters then train an agent using the best hyperparameters. 
+Hyperparameters refer to the parameters in a RL algorithm that are not directly learned during the training process. <!-- better to define before using in opening sentence.  -->
+For example, a weight in the neural network is a model parameter, but the number of hidden layers in the neural network is a hyperparameter.
+Since the optimal hyperparameters cannot be found analytically, tuning should be done to find a good-performing set of hyperparameters.
+The hyperparameter tuning process consists of sampling then evaluating a set of hyperparameters, where tuning trials are often ranked by mean cumulative reward.
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
 
-On top of all the already mentioned possible distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
-One important point of divergence is whether the agent learns a deterministic or stochastic policy.
-Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous action space.
-In table **, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
-In the appendix, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
+<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
 
@@
 \hline
-MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
+MBPO & Model-based Policy Optimization [@mbpo] & Model-based & On-policy & &\\
 \hline
-MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
+MCTS & Monte Carlo Tree Search [@suttonbarto] & Model-based & Either & & \\
 \hline
-A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
+A2C  & Advantage Actor Critic [@a2c] & Model-free & On-policy& &\\
 \hline
-A3C  & Asynchronous A2C & Model-free & On-policy & &\\
+A3C  & Asynchronous A2C & Model-free [@a3c] & On-policy & &\\
 \hline
-TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
+TRPO  & Trust Region Policy Optimization [@trpo] & Model-free & On-policy & &\\
 \hline
-PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
+PPO  & Proximal Policy Optimization [@ppo] & Model-free & On-policy & &\\
 \hline
-DQN & Deep Q Networks & Model-free & Off-policy& &\\
+DQN & Deep Q Networks [@dqn] & Model-free & Off-policy& &\\
 \hline
-DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
+DDPG & Deep Deterministic Policy Gradient [@ddpg] & Model-free & Off-policy & &\\
 \hline
-TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
+TD3 & Twin Delayed DDPG [@td3] & Model-free & Off-policy & &\\
 \hline
-SAC & Soft Actor Critic & Model-free & Off-policy & &\\
+SAC & Soft Actor Critic [@sac] & Model-free & Off-policy & &\\
 \hline
-IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
+IMPALA & Importance Weighted Actor Learner [@impala] & Model-free & Off-policy & &\\
 \hline
-APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
+APE-X & Distributed Prioritized Experience Replay [@apex] & Model-free & Off-policy & & \\
 \hline

### 2021-05-17 manuscript/manuscript.Rmd
Small updates
Overhauling with writing from alternate

--- 2021-05-15 manuscript/manuscript.Rmd
@@
 
-When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem is typically formalized as a Markov Decision Process (MDP).
-A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-After the agent observes the state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
-From these definitions, we can provide an agent's trajectory distribution for a given policy as,
-
-$$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) \, T(s_{t+1} \mid s_t, a_t)
-$$
-
-When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
-An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
-This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
-
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
-
-$$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
-$$
-
-Note that this formalism extends to an infinite horizon case when $T= \infty$.
-While there are RL methods to deal with infinite horizon cases, we will only finite horizon or episodic MDP and POMDPs.
-
 ## Designing and building an environment
@@
 \end{figure}
+
+# RL Background
+
+When the environment is fully observable, i.e.
+when the state that the agent observes fully characterizes the system, the reinforcement learning problem is typically formalized as a Markov Decision Process (MDP).
+A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
+After the agent observes the state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
+As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
+From these definitions, we can provide an agent's trajectory distribution for a given policy as,
+
+$$
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) \, T(s_{t+1} \mid s_t, a_t)
+$$
+
+When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
+An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
+This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
+
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
+
+$$
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
+$$
+
+Note that this formalism extends to an infinite horizon case when $T= \infty$.
+While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic MDP and POMDPs.
 
@@
 
+<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
+
 <!-- Can we add citations to each please -->
@@
 \hline
-MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
+MBPO & Model-based Policy Optimization [@mbpo] & Model-based & On-policy & Continuous & Continuous\\
 \hline
-MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
+MCTS & Monte Carlo Tree Search [@suttonbarto] & Model-based & Either & & \\
 \hline
-A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
+A2C  & Advantage Actor Critic [@a2c] & Model-free & On-policy& &\\
 \hline
-A3C  & Asynchronous A2C & Model-free & On-policy & &\\
+A3C  & Asynchronous A2C & Model-free [@a3c] & On-policy & &\\
 \hline
-TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
+TRPO  & Trust Region Policy Optimization [@trpo] & Model-free & On-policy & &\\
 \hline
-PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
+PPO  & Proximal Policy Optimization [@ppo] & Model-free & On-policy & &\\
 \hline
-DQN & Deep Q Networks & Model-free & Off-policy& &\\
+DQN & Deep Q Networks [@dqn] & Model-free & Off-policy& &\\
 \hline
-DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
+DDPG & Deep Deterministic Policy Gradient [@ddpg] & Model-free & Off-policy & &\\
 \hline
-TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
+TD3 & Twin Delayed DDPG [@td3] & Model-free & Off-policy & &\\
 \hline
-SAC & Soft Actor Critic & Model-free & Off-policy & &\\
+SAC & Soft Actor Critic [@sac] & Model-free & Off-policy & &\\
 \hline
-IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
+IMPALA & Importance Weighted Actor Learner [@impala] & Model-free & Off-policy & &\\
 \hline
-APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
+APE-X & Distributed Prioritized Experience Replay [@apex] & Model-free & Off-policy & & \\
 \hline

### 2021-05-17 manuscript/manuscript.Rmd
Adding citations

@@
 According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its policy.
-The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more [@cite].
-Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@cite], so it is recommended to provide a generous training budget for these agents.
+The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
+Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
 An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing much about the environment's transition dynamics or all the inner workings of the algorithm.
@@
 Hyperparameters refer to the parameters in a deep, RL algorithm that are not directly learned during the training process.
-Since the optimal hyperparameters vary across environments and can only be found through exhaustive search [@cite], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
+Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
 Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with these hyperparameters, then examining the resulting performance.
@@
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
 Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
@@
 And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not get into in depth -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
+There is a handful of advantages and disadvantages for each of these approaches that we will not get into in depth -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@suttonbarto].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@suttonbarto; @qbias].
 
 Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
+Classic RL algorithms, like policy iteration, value iteration and TD-learning, fail when the state-action space is large, e.g.
+when the possible states and actions can not be represented in a tractable table [@suttonbarto].
 Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
 Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
 While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the agent's neural network that it is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
+We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the agent's neural network that it is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
@@
 The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@CITE], and current deep RL algorithms take a range of approaches towards this dilemma -- in the appendix, we discuss some of the more specific exploration algorithms.
+This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014], and current deep RL algorithms take a range of approaches towards this dilemma -- in the appendix, we discuss some of the more specific exploration algorithms.
 Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.

### 2021-05-18 manuscript/manuscript.Rmd
Some revisions to the formal section

@@
 
-Training a deep, model-free agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
+Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
 The training period involves a trial and error process whereby the agent will perform an action and then receive feedback from the environment in the form of a reward. 
@@
 
-# RL Background
+# Formal RL Concepts
 
-When the environment is fully observable, i.e.
-when the state that the agent observes fully characterizes the system, the reinforcement learning problem is typically formalized as a Markov Decision Process (MDP).
-A MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$.
-After the agent observes the state at time t, $s_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps states to a probability distribution over actions, $\pi: \mathcal{S} \rightarrow p(\mathcal{A} \mid \mathcal{S})$.
+The reinforcement learning problem is typically formalized as a partially observable Markov decision process (POMDP).
+A POMDP is a 7-tuple that consists of the following: 
+- $\mathcal{S}$: a set of states called the state space
+- $\mathcal{A}$: a set of actions called the action space
+- $\Omega$ : a set of observations called the  observation space
+- $E(o_{t}| s_{t})$: an emission distribution, which accounts for an agent's observation being different from environment's state
+- $T(s_{t+1}|s_t, a_t)$: a state transition operator which describes the dynamics of the system
+- $r(s_t, a_t)$: a reward function
+- $d_0(s_0)$: an initial state distribution
+- $\gamma \in (0,1]$: a discount factor
+
+After the agent makes an observation at time t, $o_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
+This action will cause the environment to transition to a new state according to its state transition operator. 
+The agent will then receive a new observation and the process will repeat until a terminal state is reached.
 As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
@@
 $$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) \, T(s_{t+1} \mid s_t, a_t)
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t)
 $$
 
-When the agent cannot fully observe the state of the system, reinforcement learning can then be described by a Partially Observable Markov Decision Process (POMDP), wherein there is an additional distribution, $E(o_{t}| s_{t})$, which accounts for the observation being different from system's actual state.
 An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
-This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we will discuss.
+This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we discuss in the next section.
 
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:
+In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t|o_t)$, that maximizes the expected return, $J(\pi)$:
 
@@
 \hline
-Algorithm & Description & Model & Policy & Action Space & State Space \\
+Algorithm & Description & Model & Evaluation & Policy & Constraints  \\
 \hline
-MBPO & Model-based Policy Optimization [@mbpo] & Model-based & On-policy & Continuous & Continuous\\
+MBPO & Model-based Policy Optimization [@mbpo] & Model-based & On-policy & Deterministic & \\
 \hline

### 2021-05-18 manuscript/manuscript.Rmd
Adding to the table

@@
 
-The reinforcement learning problem is typically formalized as a partially observable Markov decision process (POMDP).
+The reinforcement learning problem is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
 A POMDP is a 7-tuple that consists of the following: 
@@
 Note that this formalism extends to an infinite horizon case when $T= \infty$.
-While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic MDP and POMDPs.
+While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
 
@@
 Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
-The distinction between on and off-policy exploration comes from what policy the agent is using to select actions during training.
+The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
 For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
@@
 \hline
-Algorithm & Description & Model & Evaluation & Policy & Constraints  \\
+Algorithm & Description & Model & Method & Evaluation  & Constraints  \\
 \hline
-MBPO & Model-based Policy Optimization [@mbpo] & Model-based & On-policy & Deterministic & \\
+MBPO & Model-based Policy Optimization [@mbpo] & Model-based & Planning & On-policy & \\
 \hline
-MCTS & Monte Carlo Tree Search [@suttonbarto] & Model-based & Either & & \\
+MCTS* & Monte Carlo Tree Search [@suttonbarto] & Model-based & Planning & Either & Not suited for continuous action and state spaces \\
 \hline
-A2C  & Advantage Actor Critic [@a2c] & Model-free & On-policy& &\\
+A2C  & Advantage Actor Critic [@a2c] & Model-free & Actor Critic  & On-policy & \\
 \hline
-A3C  & Asynchronous A2C & Model-free [@a3c] & On-policy & &\\
+A3C  & Asynchronous A2C [@a3c] & Model-free & Actor Critic & On-policy & \\
 \hline
-TRPO  & Trust Region Policy Optimization [@trpo] & Model-free & On-policy & &\\
+ACKTR  &  Actor Critic using Kronecker-Factored Trust Region [@acktr] & Model-free & Actor Critic & On-policy & \\
 \hline
-PPO  & Proximal Policy Optimization [@ppo] & Model-free & On-policy & &\\
+ACER  &  Actor Critic with Experience Replay [@wang_sample_2017] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces \\
 \hline
-DQN & Deep Q Networks [@dqn] & Model-free & Off-policy& &\\
+TRPO  & Trust Region Policy Optimization [@trpo] & Model-free & Actor Critic & On-policy & \\
 \hline
-DDPG & Deep Deterministic Policy Gradient [@ddpg] & Model-free & Off-policy & &\\
+PPO  & Proximal Policy Optimization [@ppo] & Model-free & Actor Critic & On-policy & \\
 \hline
-TD3 & Twin Delayed DDPG [@td3] & Model-free & Off-policy & &\\
+DQN & Deep Q Networks [@dqn] & Model-free & Value-based & Off-policy & Not suited for continuous action spaces; Suffers from overestimation bias\\
 \hline
-SAC & Soft Actor Critic [@sac] & Model-free & Off-policy & &\\
+Double DQN & Deep Q Networks [@van_hasselt_deep_2015] & Model-free & Value-based & Off-policy & Not suited for continuous action spaces \\
 \hline
-IMPALA & Importance Weighted Actor Learner [@impala] & Model-free & Off-policy & &\\
+DDPG & Deep Deterministic Policy Gradient [@ddpg] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces\\
 \hline
-APE-X & Distributed Prioritized Experience Replay [@apex] & Model-free & Off-policy & & \\
+TD3 & Twin Delayed DDPG [@td3] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces \\
+\hline
+SAC & Soft Actor Critic [@sac] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces\\
+\hline
+IMPALA & Importance Weighted Actor Learner [@impala] & Model-free & Actor Critic & Off-policy & \\
 \hline
 \end{tabular}
+\caption{A survey of popular deep RL algorithms.
+With the plethora of deep RL algorithms, this table aims to show some of the most common ones.
+Model-free algorithms are in the majority in this table as the model-free approach has garnered more attention in recent years.
+Model-based approaches like MCTS, which is the basis for AlphaZero, are still able to achieve state of the art performances on certain tasks, so they should not be overlooked.
+Despite there being a range of different algorithms in this table, many of them are closely related to one another.
+E.g.
+Double DQN makes a small adjustment to DQN; PPO is an evolution of TRPO; TD3 is an evolution of DDPG; and, A2C is marginally different from A3C.
+These small differences, however, can result in large change in agent performance.
+All deep RL algorithms have their respective weak points, and it is impossible to know how algorithms will perform on a task until they are tuned and evaluated.
+As a deep rl practitioner, the best thing to do then is to find a range of algorithms that can work on your problem's state and action space, tune the algorithms, and see how they perform.}
 \end{center}

### 2021-05-18 manuscript/manuscript.Rmd
Small changes to formal section

@@
 
-The reinforcement learning problem is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
+In this brief section, we will introduce the concept of the RL environment as a POMDP; we will then use POMDP definitions to ultimately define the RL objective.
+Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is crucial to understanding how RL algorithms work.
+<!-- ML: this seems like an obvious sentence to me but am stuck on it atm so revisions welcomed-->
+
+The reinforcement learning environment is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
 A POMDP is a 7-tuple that consists of the following: 
@@
 
-After the agent makes an observation at time t, $o_t$, the agent selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
-This action will cause the environment to transition to a new state according to its state transition operator. 
-The agent will then receive a new observation and the process will repeat until a terminal state is reached.
+The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
+At each time step, the agent makes an observation of the environment at time t, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
+The action will cause the environment to transition to a new state according to the state transition operator. 
+The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
+This process will repeat until a terminal state is reached.
 As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
@@
 
-An important property of POMDPs and MDPs is that they are memoryless; the future, $s_{t+1}$, does not depend on the past but only on the state and action at the present time.
-This Markov property, which is often an unrealistic assumption, is held in most of the RL algorithms that we discuss in the next section.
-
-In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t|o_t)$, that maximizes the expected return, $J(\pi)$:
+The goal of reinforcement learning is for the agent to find an optimal policy distribution, $\pi^*(a_t|o_t)$, that maximizes the expected return, $J(\pi)$:
 
@@
 While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
+In the appendix, we will go into further detail on the general approaches that model-free deep RL algorithms take towards optimizing the RL objective.
 

### 2021-05-19 manuscript/appendix.Rmd

@@
 
+<!-- These 2 paragraphs might make it back to the manuscript but going to keep them here for now -->
+Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
+*Policy gradient* algorithms exclusively learn the policy.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
+There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
+
+Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
+Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
+when the possible states and actions can not be represented in a tractable table [@CITE].
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
+Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
+While there are other function approximators that can be used in RL, e.g.
+Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
+We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
+
 ## Policy Gradients

### 2021-05-19 manuscript/manuscript.Rmd

@@
 
-In this brief section, we will introduce the concept of the RL environment as a POMDP; we will then use POMDP definitions to ultimately define the RL objective.
+In this brief section, we will introduce the concept of the RL environment as a POMDP; we will then use POMDP definitions to define the RL objective.
 Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is crucial to understanding how RL algorithms work.

### 2021-05-19 manuscript/manuscript.Rmd

@@
 \begin{figure}
-\centering
-\includegraphics[width=0.5\linewidth]{fig1.pdf}
-\caption{Reinforcement Learning focuses on the repeated feedback of agent interacting with an \emph{environment}.
-The agent performs an action which influences the environment's future \emph{state}.
-The agent receives an observation of the new state and reward from the environment.}
+\includegraphics[width=\linewidth]{fig1.pdf}
+\caption{Deep Reinforcement Learning:  An artificial intelligence \emph{agent} uses a \emph{neural network} to select an \emph{action} in response to an \emph{observation} of the \emph{environment}, receiving a \emph{reward} (or punishment) from the environment as a result.
+The agent is \emph{trained} to maximize the reward through the experience of trying different actions over repeated \emph{epsiodes} of interacting with the environment.
+Over the course of training, the agent adjusts parameter weights on the neural network as it learns which actions lead to the best cumulative rewards.
+A researcher may \emph{tune} an agent by considering iterative trials that train different agents with different \emph{hyperparameters}, such as different neural network architectures.
+An agent trained with the most successful combination of hyperparameters can then be tested in an evaluation environment.}
 \end{figure}
 
-@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research; to address these issues, @brockman2016 proposed the creation of the OpenAI `gym` framework.
+@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
 The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
@@
 
-# Discussion
+<!-- Solving 'real' problems will be hard -->
+Successfully applying RL to complex ecological problems is no easy task.
+The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance is another task.
+As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
+Such examples provide important benchmarks against which to calibrate the performance, tuning and training requirements of RL,
+and can easily be extended into more complex problems once RL agents can master the basics.
+Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
+In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: instability, interpretability, and reward design.
 
-# Open issues
+<!-- reason 1: instability-->
+One leading challenge for deep RL today is the problem of instability.
+RL agents trained on narrowly defined environments may perform poorly even when the environment is altered in seemingly trivial way, such as a re-scaling of parameters.
+To avoid this, researchers should seek to both train and evaluate agents against a range of environments representing different parameters or processes dynamics.
+Stable performance can be challenging even when using the same environment.
+For a range of reasons, including high variance [@cite] and the exploration vs.
+exploitation trade-off [@cite], model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
+For instance, it is not uncommon to observe vastly different behavior when an agent is initialized with different random seeds [cite ibid].
+Such unstable behavior can often be avoided by alternate selection of hyperparameters, so researchers should consider stability as well as performance during hyperparameter selection.
+The choice of RL algorithm, design of the reward function, and duration of training can also influence stability.
 
-Deep reinforcement learning has only become an established subfield of AI over the past decade; and, as with most young fields, there are a range of unsolved problems.
-In practice, the foremost problem with DRL is instability.
-For a range of reasons, including high variance and the exploration vs.
-exploitation trade-off, model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
-For instance, it is not uncommon to observe vastly different behavior when two agents are initialized with different random seeds [cite ibid].
-Since model-based and model-free algorithms typically have a large number of hyperparameters, each of which can contribute to training instability, it is practically impossible to identify the source of instability.
-The hope of the RL practitioner is that after doing an exhaustive hyperparameter search, the practitioner will find a set of hyperparameters that results in a robust, well-performing agent, but there are no guarantees that hyperparameter tuning will lead to this outcome.
-<!-- Kari: what does it mean for two algorithms with similar hyperparameters to perform differently? (As happens in our examples) -->
+<!--
+Under some conditions, RL agents can prove adept at _transfer learning_, in which an agent trained in one environment performs well under a different environment.
+-->
+<!-- if we're really recommending 'testing' for stability, maybe we need to demo that in the appendix-->
 
-In the context of conservation, we anticipate that there will be a host of specific challenges to using DRL methods.
-Interpretability is likely to be the biggest barrier to adoption.
+<!-- reason 2: interpretability -->
+Interpretability is a common challenge ot the adoption of most ML methods [@cite]. 
 Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
-Another challenge is that DRL algorithms are very sample inefficient, and to overcome sample inefficiency for conservation decision-making problems that are often data poor, relying on simulations to train agents will be necessary.
-While there are positives to using simulated data, any assumptions or biases that are included in the model will determine agent behavior.
+As our examples have illustrated, it may be possible to gain intuition on an agent's behavior by visualizing the action selected against each possible state, though this becomes more difficult in large problems.
+Another challenge is that Deep RL algorithms are very sample inefficient, all though progress is being made to improve this [@cite].  
+-->
 
+<!-- reason 3: reward design -->
 An issue that is central to any applied optimization problem is using the right objective.
-There have been many examples in which explicit reward design goes horribly wrong [cite].
+There have been many examples in which explicit reward design goes very wrong [cite].
 And for ecological systems, which are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
@@
 who gets to specify the reward function and will the values of the greater community be represented in the reward function.
+RL researchers often introduce modified reward functions during training that help an agent perform better on the original objective [@cite].
+This approach is particularly common when rewards are sparse, such as in a game of Chess where the only objective that counts is whether or not you win, but during training, an RL algorithm may reward the agent for intermediate steps such as capturing a piece.
+
+<!--
 One recently proposed framework to overcome challenges with explicit reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
-Yet, in the context of training an agent to say conserve a species, there are questions that could easily arise, foremost will the agent infer that humans don't care about a certain species so acts in a way that leads to that species' extinction?
 There is a definite need for exploration of the ethics of using DRL for conservation.
-
-Ultimately, we anticipate that getting DRL to work well on fully detailed conservation decision-making problems will be challenging.
+-->
+<!-- reason 4/additional reasons, like sparse rewards -->
+Ultimately, we anticipate that getting deep RL to work well on fully detailed conservation decision-making problems will be challenging.
 Beyond typically having large state-action spaces for agents to explore, conservation problems tend to have sparse reward structures.

### 2021-05-19 manuscript/manuscript.Rmd
manual merge millie-edits thanks @milliechapman

@@
 
+<!-- from end of Carl's discussion edits: Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.
+-->
+<!-- If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.
+-->
+
+The increasingly influential role of AI in sectors as diverse as policing, finance, and employment has raised concern about the ethics, biases, and power entrenched in both decision-making algorithms and their outputs [@kalluri2020].
+As algorithms play a greater role in environmental and conservation policy, will similar issues arise?
+While this question has garnered  recent interest in the academic literature [@wearn2019; @adams2019; @scoville2021], the use of decision making algorithms in environmental contexts is still relatively isolated and in it's early stages.
+Examples of AI applications, and corresponding concerns, have largely been confined to the context of enforcing environmental laws (i.e.
+algorithms which predict wildlife poaching events [@adams2019] or identify illegal fishing behavior [@mcdonald2021; @swartz2021]).
+However, more diverse applications of AI algorithms - such as applications which contribute to meeting sustainable development goals [@vinuesa2020] or designing  protected area networks - raise additional concerns of entrenching power asymmetries between those reliant on natural resources and those designing algorithms and their objectives.
+These issues of power become particularly acute when private industry holds the keys to propriety algorithms or environmental data, and generally operate without the same community engagement obligations of public agencies.
+In these cases, good intentions might not be sufficient to ensure equitable outcomes. 
+Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges promotes equity in both the procedural process of developing objective functions and the distribution of benefits that result from decisions.
+
 <!-- Something about the way forward, point to appendix resources etc -->

### 2021-05-19 manuscript/manuscript.Rmd
fix knit

@@
 
-<!-- 
-Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
-Make a nice table box in latex instead if we're keeping this
--->
-\begin{figure}
-\centering
-\includegraphics[width=0.9\linewidth]{table1.pdf}
-\end{figure}
-
 # Formal RL Concepts
@@
 -->
-```{=latex}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-Algorithm & Description & Model & Method & Evaluation  & Constraints  \\
-\hline
-MBPO & Model-based Policy Optimization [@mbpo] & Model-based & Planning & On-policy & \\
-\hline
-MCTS* & Monte Carlo Tree Search [@suttonbarto] & Model-based & Planning & Either & Not suited for continuous action and state spaces \\
-\hline
-A2C  & Advantage Actor Critic [@a2c] & Model-free & Actor Critic  & On-policy & \\
-\hline
-A3C  & Asynchronous A2C [@a3c] & Model-free & Actor Critic & On-policy & \\
-\hline
-ACKTR  &  Actor Critic using Kronecker-Factored Trust Region [@acktr] & Model-free & Actor Critic & On-policy & \\
-\hline
-ACER  &  Actor Critic with Experience Replay [@wang_sample_2017] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces \\
-\hline
-TRPO  & Trust Region Policy Optimization [@trpo] & Model-free & Actor Critic & On-policy & \\
-\hline
-PPO  & Proximal Policy Optimization [@ppo] & Model-free & Actor Critic & On-policy & \\
-\hline
-DQN & Deep Q Networks [@dqn] & Model-free & Value-based & Off-policy & Not suited for continuous action spaces; Suffers from overestimation bias\\
-\hline
-Double DQN & Deep Q Networks [@van_hasselt_deep_2015] & Model-free & Value-based & Off-policy & Not suited for continuous action spaces \\
-\hline
-DDPG & Deep Deterministic Policy Gradient [@ddpg] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces\\
-\hline
-TD3 & Twin Delayed DDPG [@td3] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces \\
-\hline
-SAC & Soft Actor Critic [@sac] & Model-free & Actor Critic & Off-policy & Not suited for discrete action spaces\\
-\hline
-IMPALA & Importance Weighted Actor Learner [@impala] & Model-free & Actor Critic & Off-policy & \\
-\hline
-\end{tabular}
-\caption{A survey of popular deep RL algorithms.
-With the plethora of deep RL algorithms, this table aims to show some of the most common ones.
-Model-free algorithms are in the majority in this table as the model-free approach has garnered more attention in recent years.
-Model-based approaches like MCTS, which is the basis for AlphaZero, are still able to achieve state of the art performances on certain tasks, so they should not be overlooked.
-Despite there being a range of different algorithms in this table, many of them are closely related to one another.
-E.g.
-Double DQN makes a small adjustment to DQN; PPO is an evolution of TRPO; TD3 is an evolution of DDPG; and, A2C is marginally different from A3C.
-These small differences, however, can result in large change in agent performance.
-All deep RL algorithms have their respective weak points, and it is impossible to know how algorithms will perform on a task until they are tuned and evaluated.
-As a deep rl practitioner, the best thing to do then is to find a range of algorithms that can work on your problem's state and action space, tune the algorithms, and see how they perform.}
-\end{center}
-```
+
 # Examples
@@
 
-```{r fig1, fig.cap="Fisheries management using neural network agents trained with RL algorithms A2C and TD3 compared to optimal management.
+```{r fig_fish, fig.cap="Fisheries management using neural network agents trained with RL algorithms A2C and TD3 compared to optimal management.
 Top panel: mean fish population size over time across 100 replicates.
@@
 
-```{r fig2, fig.cap=""}
+```{r fig_conservation, fig.cap=""}
 fig_sims / ( fig_policy + fig_reward)

### 2021-05-19 manuscript/table-terms.Rmd
fix knit

---
output: pdf_document
header-includes:
  - \pagestyle{empty}
  - \geometry{margin=0in}
---

Term        | Definition 
------------|------------------------
environment | A mathematical model, computer simulation, or real world in which the agent interacts.  The environment is described by a current *state* which is observed by the agent and influenced by the agent's *actions*.  The agent also experiences a reward (cost or benefit) from interacting with the environment, depending on the states and the actions taken.
agent | A software algorithm which generates proposed *actions* in response to observations of the current *state* of the *environment*. 
neural network |
training | Process by which an agent improves its performance through repeated episodes of interaction with the environment.  Training effectively allows estimation of parameters describing weights of the agent's neural networks.  
tuning | Process by which a researcher selects the hyper-parameters governing the behavior of agent training.
policy function | A function which maps observations of the current *state* to the *actions* selected by an *agent*.  
reward | The cost or benefit an agent receives for any interaction with the environment. Note that an agent seeks to maximize the discounted cumulative reward over an episode, and that a reward may be given after every action or potentially only at some end condition (e.g. winning or losing a chess match).  Researchers can often improve agent learning by appropriate reward design during training. 
value function | 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$ (expected value of being in a given state), the state-action value function, $Q^\pi(s_t, a_t)$ the expected value of a given state conditional on a certain action, often called the Q-value, and the advantage function, $A^\pi(s_t, a_t)$, the Q-value relative to some baseline.
state | The state of the world (environment) with which the agent interacts. The state can be discrete or continuous and in an arbitrary number of dimensions. An agent may observe the current state completely, or only partially. 
action | The space of possible actions the agent can take.  Actions can be discrete or continuous and in an arbitrary number of dimensions. Both the state and action space are part of the environment definition -- the agent may not necessarily know what the valid range of possible actions is.
hyper-parameters | parameters controlling behavior of the RL algorithm that are fixed over the course of training, e.g. the learning rate, discount rate, or size and type of neural network(s) used by the agent.
model | We use the term "model" to describe a mathematical model of the ecological process (environment). An RL algorithm is considered "model-based" if the agent seeks to 'learn' this underlying model, and "model-free" if it does not. Confusingly, some frameworks (stable-baselines) also refer to the agent as the "model" (even when the algorithm being used is model-free).

### 2021-05-20 appendices/appendix-A-definitions.Rmd
divide up appendices

--- 2021-05-19 manuscript/appendix.Rmd
@@
 ---
-title: "Appendix for: Deep Reinforcement Learning for Conservation Decisions"
+title: "Appendix A: Deep Reinforcement Learning"
 authors:
@@
 ---
-
-```{r knitr, include = FALSE}
-knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4, cache = FALSE)
-ggplot2::theme_set(ggplot2::theme_bw())
-
-scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
-pal = ggthemes::solarized_pal()(8)
-txtcolor = "#586e75"
-```
-
-# Building a simulation environment `gym`
-
-[@openai_gym]
-
-# Deep Reinforcement Learning Frameworks
-
-At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
-While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
-Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
-Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3].
-Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
-Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
-However, higher-level interfaces and high-level documentation are still relatively limited.
-OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
-OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
-Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption.
-Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
-The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
-Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
-
-# Deep Reinforcement Learning in R
-
-Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
-Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
-In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
-
-In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
-This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
-Alternately, users may prefer running the analysis inside a docker container.  
-The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
-This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.
-
-Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
-From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
-
-```{r}
-#install.packages("renv")
-renv::restore()
-```
-
-Once the packages have installed, we are ready to load the necessary libraries.
-Note that the `import` function from `reticulate` package acts much like the `library` command in R, though it does not attach the package function to the global namespace.
-To make it more convenient to access those functions, we can assign a shorthand name.
-
-```{r setup, cache = FALSE, message=FALSE}
-# R dependencies
-library(tidyverse)
-library(patchwork)
-library(reticulate)
-
-## Python dependencies loaded via R
-sb3         = import ("stable_baselines3")
-gym         = import ("gym")
-gym_fishing = import("gym_fishing")
-gym_conservation = import("gym_conservation")
-```
-
-Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration.
-In addition to setting a random seed in our Python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.
-
-```{r}
-## reproducible settings
-np = import("numpy")
-seed = 24L # integer
-np$random$seed(seed)
-
-# Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
-Sys.setenv("CUDA_VISIBLE_DEVICES"="")
-set.seed(seed)
-```
-
-<!-- is this too verbose? --> 
-The above code also illustrates a few conventions which may be helpful to bear in mind when using the `reticulate` interface to interact with Python from R: 
-it is often necessary for integer values to be explicitly typed as integers by adding a trailing `L` (corresponding to the primitive C type `long` integer).
-Python is also more strongly object-oriented than many R packages, where "methods" of an "object" are accessed with the list-subset operator `$` in R (equivalent to the use of `.` in Python).
-Lastly, while R can use `<-` or `=` for assignment, Python uses only `=`. For simplicity we will stick with the latter.  
-With few exceptions, the R code shown here can be re-written as python code by dropping the `L` and replacing `$` with `.`, see stand-alone python code in the `python/` subdirectory of the repository.
-
-# Finding a known optimal solution using RL
-
-## Sustainable Harvest Quotas
-
-We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
-
-```{r }
-## initialize the environment
-env = gym$make("fishing-v1", sigma = 0.1)
-```
-
-<!-- Kari: the role neural networks play in RL is not very well explained in the paper, more detail there would probably make this part more intuitive --> 
-At the heart of each deep RL algorithm is a neural network or a collection of neural networks.
-The RL agent uses these networks to approximate a policy function and/or a value function.
-The policy function is the agent's mapping of states to actions -- i.e.
-the agent's strategy.
-Value functions attempt to evaluate the goodness of a state or state-action pair towards achieving the RL objective of maximizing cumulative rewards.
-Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyper-parameters_ while the parameters of a RL algorithm that are learned during training are called parameters.
-For example, the number of layers in a neural network is a hyper-parameter but a weight in the neural net is referred to as a parameter.
-Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
-Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms,
-for instance, the A2C algorithm originally described in @A2C.
-Here, we train an agent using the A2C algorithm using default hyper-parameter settings with a multi-layer perceptron policy and value network composed of two 64-neuron layers.
-
-```{r}
-a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
-```
-
-Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment, the neural network architecture and the number of training iterations budgeted.
-Therefore, we save the trained agent to disk, and only execute the training method (`learn`) if no previously saved agent is found:
-
-```{r}
-if(!file.exists("cache/a2c.zip")) {
-
-  # Train the agent for a fixed number of timesteps
-  a2c$learn(total_timesteps=300000L)
-
-  # Save our trained agent for future use
-  a2c$save("cache/a2c")
-}
-```
-
-Note that while default hyper-parameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyper-parameter selection.
-This is discussed further below.
-Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.
-Note that a copy of the trained agents are included in the corresponding GitHub repository.
-
-```{r}
-# Simulate management under the trained agent
-a2c = sb3$A2C$load("cache/a2c")
-```
-
-We can supply an observation to our trained agent and it will return it's proposed action using the `predict` method. 
-This is all we need to evaluate or employ the agent on a decision-making task.
-Recall that state space and action space in the fishing gym have been re-scaled to a (-1, 1) interval.
-Note that this is equivalent to a choice of appropriate units -- we can re-scale the interval without loss of generality.
-This is often an important step in the design of an RL environment to facilitate successful training.
-Following the `gym` standard, the core methods such as `predict` and `step` operate on the re-scaled units,
-so it is necessary to first transform the original units into this re-scaled state space.
-For example, if we wish to start a simulation with a stock size of 0.75, we can use the helper
-method `get_state()`, to determine the corresponding value in the re-scaled state space.  
-Unlike `predict` and `step`, `get_state()` is not a standard method of all gym environments -- typically
-a user must first inspect the state space of an environment and choose themselves how to re-scale their
-problem into that state space.
-
-```{r}
-
-## represent the initial state size in the 'rescaled' state space.
-state = env$get_state( 0.75 )
-state
-```
-
-With an initial state in hand, we are ready to simulate management using our agent.
-The iteration is simple: we use the agent to predict what action we should take given the current state. Then, we take said action and examine the result to determine the future state. 
-Because these methods return additional information as well, a little extra sub-setting is required in R:
-
-```{r}
-for(i in 1:10){
-
-  out = a2c$predict(state)
-  action = out[[1]]
-  result = env$step(action)
-  state = result[[1]]
-
-}
-
-```
-
-For convenience, `gym_fishing` defines the helper routine `simulate` to perform the above iteration `reps` number of times.
-The `simulate` method returns the state, action, and reward resulting from each time step of each replicate:
-
-```{r}
-a2c_sims = env$simulate(a2c, reps = 500L)
-```
-
-Because the agent typically attempts to retain the fish population near a specific value, simulations with well-trained agents will usually not explore the full range of possible states.
-To get a better idea of how the agent behaves across the full state space, it is common to examine the policy function: to do this, we'll plot the action taken at every possible state.
-Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
-Note that the `policyfn` method is another custom method and not part of the `gym` standard.
-The use of `policyfn` will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTMs [@LSTM].
-
-```{r}
-a2c_policy = env$policyfn(a2c, reps = 50L)
-```
-
-## TD3
-
-We train a second agent based on the TD3 algorithm. 
-This illustrates some of the salient differences between the different algorithms used in reinforcement learning.
-Bear in mind that the best algorithm for any given environment will depend on the details of the environment as well
-as the length of training and the selection of hyper-parameters.
-This time, we will declare our choices for the hyper-parameters explicitly.
-Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
-Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
-For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
-
-```{r}
-# train an agent (model) on one of the environments:
-
-policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
-# non-episodic action noise:
-noise_std = 0.6656948079225263
-n_actions = env$action_space$shape[0]
-action_noise = sb3$common$noise$NormalActionNoise(
-        mean=np$zeros(n_actions),
-        sigma= noise_std * np$ones(n_actions)
-        )
-
-td3 = sb3$TD3( 'MlpPolicy', 
-               env,  
-               verbose=0L, 
-               seed = seed,
-               policy_kwargs = policy_kwargs,
-               learning_rate = 0.0001355522450968401,
-               gamma= 0.995, 
-               batch_size = 128L,
-               buffer_size = 10000L,
-               train_freq = 128L, 
-               n_episodes_rollout = -1, 
-               gradient_steps = 128L,
-               action_noise = action_noise
-              )
-```
-
-Once again, after our agent is defined we are ready to train it:
-
-```{r }
-if(!file.exists("cache/td3.zip")){
-
-  td3$learn(total_timesteps=300000L)
-  td3$save("cache/td3")
-
-}
-```
-
-With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:
-
-```{r}
-# Simulate management under the trained agent
-td3 = sb3$TD3$load("cache/td3")
-td3_sims = env$simulate(td3, reps = 500L)
-td3_policy = env$policyfn(td3, reps = 50L)
-```
-
-We compare the performance of these two RL agents trained using the algorithms A2C and TD3 respectively to the known optimal policy for this particular environment.  
-Recall that under the assumptions of the simple model used in the `fishing-v1` environment, we can determine the optimal harvest policy analytically if the model and parameters are known precisely [@Reed1979].
-The optimal strategy is a policy of 'constant escapement', designed to keep the remaining stock size (the population that 'escapes' fishing harvest) at the biomass corresponding to a maximum growth rate, i.e.
-at $B = K/2$ in this model.
-`gym_fishing` defines a collection of non-RL agents in the `models` submodule, including the a human agent that merely asks to enter their desired quota manually. 
-The `escapement` model implements the provably optimal constant escapement rule.
-A third model, `msy`, implements a policy based on "Maximum Sustainable Yield" policy [@Schaefer1954], which is actually more commonly used as a basis for management than constant escapement, despite only being optimal at the steady state under deterministic dynamics.
-
-```{r}
-# Simulate under the optimal solution (given the model)
-opt = gym_fishing$models$escapement(env)
-opt_sims = env$simulate(opt, reps = 500L)
-opt_policy = env$policyfn(opt)
-```
-
-Having performed simulations under each method, we gather the results under each management strategy together into a single data frame for simulations (`sims_df`), the policy function (`policy_df`) and the cumulative reward, used to make the plots in the corresponding panels.
-We preserve a copy of each data frames in local `.csv` text files for maximum cross-platform compatibility.
-
-```{r}
-sims_df = bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
-
-policy_df = bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
-  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
-
-gamma = 1 #discount
-reward_df = sims_df %>% 
-  group_by(rep, model) %>%
-  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
-  group_by(time, model) %>%
-  summarise(mean_reward = mean(cum_reward), 
-            sd = sd(cum_reward), .groups = "drop")
-
-write_csv(sims_df, "figs/sims_df.csv")
-write_csv(policy_df, "figs/policy_df.csv")
-write_csv(reward_df, "figs/reward_df.csv")
-```
-
-<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
-Standard R methods are used to plot the results summarized across the replicates:
-
-```{r}
-ymin = function(x) last(x[(ntile(x, 20)==1)])
-ymax = function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims = 
-sims_df %>% 
-  group_by(time, model) %>% 
-  summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-```
-
-```{r}
-fig_policy = 
-  policy_df %>% ggplot(aes(state, action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) + 
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) 
-```
-
-```{r}
-
-fig_reward = reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
-```
-
-## Manuscript Figure 1
-
-<!-- Not sure if it's helpful to duplicate figures from the paper in the appendix...-->
-
-```{r fig1}
-fig_sims / ( fig_policy + fig_reward)
-```
-
-## Applying an RL agent to real data
-
-We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.
-
-Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
-We will use the agent we have just trained using the `gym_fishing` simulator above,
-
-```{r}
-# Simulate management under the trained agent
-env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-td3 = sb3$TD3$load("cache/td3")
-
-```
-
-```{r}
-hake = read_csv("../data/hake.csv")
-x0 = hake %>%
- filter(year == min(year)) %>%
-  pull(scaled_biomass)
-Tmax = 15
-years =1986:2000
-
-td3_state = td3_action = numeric(Tmax)
-td3_state[1] = x0
-td3_action[1] = NA
-
-```
-
-Our hindcast considers the scenario of managing catch quotas using TD3, beginning in 1986. 
-We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
-This shows that quotas set by TD3 would have been lower than historical values throughout the late 80s and early 90s, which
-saw a sharp decline in estimated biomass of the Argentine Hake.
-For the same or higher estimated stock sizes from the historical record, we see that TD3 would recommend a lower quota.
-We also see that under TD3 the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
-being set in the now-depleted stock.  
-Whether the TD3 quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
-However, the fact that TD3 would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
-Combined with the observed historical declines, this provides some evidence that the RL agent could have decreased or avoided the over-harvesting in this case.
-
-```{r}
-env$reset()
-## represent the initial state size in the 'rescaled' state space.
-state = env$get_state( x0 )
-
-for(i in 2:Tmax){
-
-  # RL-recommended harvest action:
-  out = td3$predict(state)
-  action = out[[1]]
-
-  # Record state and resulting action
-  td3_state[i] = env$get_fish_population(state)
-  td3_action[i] = env$get_quota(action)
-
-  # Implement RL-recommended harvest. 
-  result = env$step(action)
-  state = result[[1]]
-
-  #state = env$get_state(hake$scaled_biomass[i+1])
-}
-
-df = bind_rows(
-  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
-  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
-
-df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))
-
-```
-
-# Ecological tipping points
-
-Our second example utilizes our `gym_conservation` to provide the necessary environment to simulate an ecosystem approaching a tipping point.
-
-## Tipping point model
-
-The tipping point model is based on the consumer-resource model of @May1977, which creates alternative stable states, 
-which we subject to log-normal environmental noise:
-
-\[
-\mu_{t} = X_t + X_t r \left(1 - \frac{X_t}{K} \right) - \frac{a_t X_t^q}{X_t^q + b^q}
-\]
-\[
-X_{t+1} \sim \text{lognormal}(\mu_t, \sigma)
-\]
-
-where we take $r=0.7$, $K=1.2$, $q=3$, $b=0.15$, $a_0 = 0.19$ and $\sigma = 0.2$ 
-Slow change over time in the parameter $a_t$ represents a process of environmental degradation, modeled as a constant increment $a_{t+1} = a_t + \alpha$, where we will take $\alpha = 0.001$.
-This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carptenter2011] to the planet biosphere [@Barnosky2014].
-
-```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario.
-The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g.
-global mean temperature, arbitrary units) of 0.19.
-In the absence of conservation action, the environment worsens (e.g.
-rising mean temperature) as the parameter increases.
-This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.
-Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
-bifur_df = read_csv("figs/bifur.csv", col_types = "dccd")
-bifur_df %>%
-  ggplot(aes(parameter, state, lty=equilibrium, group = group)) + 
-  geom_line()
-
-```
-
-We assume the benefit provided by the ecosystem state is assumed to be directly proportional to the state itself, $b x_t$.  
-We further assume that each year the manager has the option to slow or reverse the environmental degradation by taking action $A_t$, such that under management, the resulting environment in the next time step is given by
-
-\[
-a_{t+1} = a_t + \alpha - A_t
-\]
-
-We assume the cost associated with that action to be proportional to the square of the action, such that large actions are proportionally more costly than small ones.
-Consequently, the utility at time $t$ is given by the sum of costs and benefits:
-
-\[U(X_t, A_t) = b X_t - c A_t ^2\]
-
-where we will take \(b = 1\) and \(c=1\).  
-Our implementation in `gym_conservation` allows the user to consider alternative parameter choices, alternative models for the ecological dynamics, and alternate types of actions, such as manipulating the ecosystem state directly rather than manipulating the environmental parameter.
-For some such scenarios the optimal solution is known or can be determined by stochastic dynamic programming, while for others, including the scenario of focus here, the optimal solution is unknown.
-
-For simplicity of illustration, we will train only a TD3-based agent on this scenario.
-
-```{r}
-env = gym$make("conservation-v6")
-```
-
-```{r}
-noise_std = 0.4805935357322933
-
-OU = sb3$common$noise$OrnsteinUhlenbeckActionNoise
-action_noise = OU(mean = np$zeros(1L),  sigma = noise_std * np$ones(1L))
-
-model = sb3$TD3('MlpPolicy', 
-            env, 
-            verbose = 0, 
-            seed = 42L,
-            "gamma"= 0.995,
-            "learning_rate"=  8.315382409902049e-05,
-            "batch_size"= 512L,
-            "buffer_size"= 10000L,
-            "train_freq"= 1000L,
-            "gradient_steps"= 1000L,
-            "n_episodes_rollout"= -1L,
-            "action_noise"= action_noise,
-            "policy_kwargs"= list("net_arch"= c(64L,64L)))
-
-```
-
-```{r}
-if(!file.exists("cache/td3-conservation.zip")){
-
-  model$learn(total_timesteps=3000000L)
-  model$save("cache/td3-conservation")
-
-}
-
-```
-
-```{r}
-# Simulate management under the trained agent
-# See https://github.com/boettiger-lab/conservation-agents/blob//conservation/TD3-v6.py
-
-model = sb3$TD3$load("cache/td3-conservation")
-TD3_sims = env$simulate(model, reps = 100L)
-TD3_policy = env$policyfn(model, reps = 10L)
-```
-
-In general, the optimal solution depends on the ecological dynamics, the benefit of the ecosystem services and the costs associated with a management response.
-Because the tipping point problem is non-autonomous, we cannot solve for the optimal policy even given the model and objective (utility) function using Markov Decision Process methods.
-However, a simple heuristic solution provides a reasonable starting point for comparison.
-
-As discussed in the main text, we have no existing optimal solution to the tipping point problem, and so rely on a common heuristic strategy instead:
-select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in it's current state.
-This is implemented using the `fixed_action` method provided in our `gym_conservation` module, which also implements other heursitic models, including a human agent which requires interactive input to select the action each year.
-
-```{r}
-# Simulate under the steady-state solution (given the model)
-K = 1.5
-alpha = 0.001
-opt = gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
-opt_sims = env$simulate(opt, reps = 100L)
-opt_policy = env$policyfn(opt)
-```
-
-We gather together the results under the RL agent and steady-state policy as before,
-
-```{r}
-sims_df = bind_rows(TD3_sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "steady-state")[as.integer(model)])
-
-policy_df = bind_rows(TD3_policy, opt_policy, .id = "model") %>%
-  mutate(model = c("TD3", "steady-state")[as.integer(model)])
-
-gamma = 1 #discount
-reward_df = sims_df %>% 
-  group_by(rep, model) %>%
-  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
-  group_by(time, model) %>%
-  summarise(mean_reward = mean(cum_reward), 
-            sd = sd(cum_reward), .groups = "drop")
-
-```
-
-The resulting three data frames contain the necessary data for each of the subplots in figure 3 of the main text.
-
-```{r}
-write_csv(sims_df, "figs/tipping_sims_df.csv")
-write_csv(policy_df, "figs/tipping_policy_df.csv")
-write_csv(reward_df, "figs/tipping_reward_df.csv")
-```
-
-<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
-
-```{r}
-# ensemble statistics
-ymin = function(x) last(x[(ntile(x, 20)==1)])
-ymax = function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims = 
-  sims_df %>%
-   group_by(time, model) %>% 
-   summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-```
-
-```{r}
-fig_policy = 
-  policy_df %>% ggplot(aes(state, action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) 
-```
-
-```{r}
-
-fig_reward = reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
-```
-
-### Manuscript Figure 3
-
-```{r fig3}
-fig_sims / ( fig_policy + fig_reward)
-```
-
-Because it is difficult to get a feel for the dynamics of individual replicate simulations from ensemble statistics, we select a few example trajectories to examine directly. 
-Of the 100 replicate simulations, we pick 4 examples that dip below a state value of 0.2 for over 15 consecutive timesteps, indicating a transition into the lower basin of attraction. 
-Comparing the dynamics under the rule of thumb steady-state strategy to that of the RL-trained agent, it is clear that the RL agent does a better job at both avoiding tipping points and promoting the recovery of those selected trajectories that cross into the lower attractor.
-
-```{r}
-# Some individual replicates, for comparison
-
-## First 4 of the TD3 reps falling below .2 for more than 15 steps
-is_low = sims_df %>% 
-  filter(model == "TD3") %>% 
-  group_by(rep, model) %>% 
-  summarize(low = sum(state < .2) > 10) %>% 
-  filter(low) %>% head(6)
-
-## First 4 such cases:
-sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state,  group=interaction(model, rep))) +
-  geom_line(color = pal[3], show.legend = FALSE) + facet_wrap(~rep)
-```
-
-```{r}
-# Some individual replicates, for comparison
-is_low = sims_df %>% filter(model == "steady-state") %>%
-  group_by(rep, model) %>% summarize(low = sum(state < .2) > 10) %>% 
-  filter(low) %>% head(6)
-sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state, group=interaction(model, rep))) +
-  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
-```
-
-# RL
 

### 2021-05-20 appendices/appendix-B-examples.Rmd
divide up appendices

--- 2021-05-03 manuscript/appendix.Rmd
@@
 ---
-title: "Appendix for: Deep Reinforcement Learning for Conservation Decisions"
+title: "Appendix B: Examples of Deep RL in Ecological Decision Problems"
 authors:
@@
 ```
-
-----

### 2021-05-20 manuscript/manuscript.Rmd
First pass; edited some sentences for grammar or spelling

@@
 
+<!-- ML: need to change some of the language here to be consistent with POMDP set up later-->
+
 Any application of RL can be divided into two components: an *environment* and an *agent*.
@@
 For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at the different species over a time step.
-The main focus of an RL environment, however, is in the environment's transition dynamics and reward function.
+The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
 The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
@@
 Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
-An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing much about the environment's transition dynamics or all the inner workings of the algorithm.
+An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing anything about the environment's transition dynamics or the inner workings of the algorithm.
 
-Hyperparameters refer to the parameters in a deep, RL algorithm that are not directly learned during the training process.
+Hyperparameters refer to the parameters in a deep, RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
 Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
@@
 A POMDP is a 7-tuple that consists of the following:
+
 - $\mathcal{S}$: a set of states called the state space
+
 - $\mathcal{A}$: a set of actions called the action space
+
 - $\Omega$ : a set of observations called the observation space
+
 - $E(o_{t}| s_{t})$: an emission distribution, which accounts for an agent's observation being different from environment's state
+
 - $T(s_{t+1}|s_t, a_t)$: a state transition operator which describes the dynamics of the system
+
 - $r(s_t, a_t)$: a reward function
+
 - $d_0(s_0)$: an initial state distribution
+
 - $\gamma \in (0,1]$: a discount factor
@@
 In contrast, *policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function is used to inform the agent on the goodness of a selected action.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
 There is a handful of advantages and disadvantages for each of these approaches that we will not get into in depth -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@suttonbarto].
 Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@suttonbarto; @qbias].
+<!-- ML: I think we can keep the bias-variance discussion in the appendix and cut these 2 sentences-->
 
@@
 Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
+Whenever there is a need for an agent to approximate some function, say a policy function or value function, neural networks can be used in this capacity.
 While there are other function approximators that can be used in RL, e.g.
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the agent's neural network that it is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the neural network that the agent is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
@@
 Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
-Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
+Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
 When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].
@@
 However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+<!-- ML: I think this sentence should be tightened up re RL, like one could use a model-based algorithm which uses an inputted model under the RL umbrella-->
 
@@
 
+<!-- ML: I'm have some uneasiness about this figure. Is A2C tuned? I think we'd probably want to do tuning and training with a set seed so that everything is reproducible?  -->
 ```{r message=FALSE, include=FALSE}
@@
 Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
-This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.
+This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady state strategy.
 The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
@@
 The use of a recurrent neural network (RNN) architecture instead of the multi-layer perceptron (MLP) used here would open the door for the policy to reflect the non-autonomous dynamics more directly. 
+<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote-->
 
@@
 The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
+<!-- ML: I feel like the remaining part of this dont full drive home the benefit of model-free rl, like its able to learn good performing policies even in the case when a model is very difficult or impossible to learn.
+-->
 Alternative models can easily be implemented in simulations, allowing an agent to train against both contingencies.
@@
 <!-- Real problems are harder than these examples -->
+<!-- ML: I feel like this paragraph could be condensed with the one above -->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
-In these simple examples in which the simulation implements a single model, training is analgous to classical methods which take the model as given [@Marescot2013].
+In these simple examples in which the simulation implements a single model, training is analogous to classical methods which take the model as given [@Marescot2013].
 But classical approaches can be difficult to generalize when the underlying model is unknown: alternative models may require a completely different approach.
@@
 <!-- reason 1: instability-->
+<!-- ML: I think this needs to be tightened up a bit-->
 One leading challenge for deep RL today is the problem of instability.
@@
 <!-- reason 2: interpretability -->
-Interpretability is a common challenge ot the adoption of most ML methods [@cite]. 
+<!-- ML: Re our OU talk, might be worth saying that this is a barrier to adoption?
+If we want to ultimately use these RL agents to make important decisions, it would be hard to swallow us blindly using them without having any sense to what is going on.
+-->
+Interpretability is a common challenge at the adoption of most ML methods [@cite]. 
 Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
@@
 Another challenge is that Deep RL algorithms are very sample inefficient, all though progress is being made to improve this [@cite].
+<!-- ML: I think we can cut this sentence, mentionned it in passing earlier-->
 -->
@@
 <!-- reason 3: reward design -->
+<!-- ML: I'll take a pass at this paragraph in a new branch. I think it would be useful to put terms on stuff in here, like reward shaping and reward hacking which we are talking around. -->
 An issue that is central to any applied optimization problem is using the right objective.
@@
 <!-- reason 4/additional reasons, like sparse rewards -->
+<!-- ML: I don't know if this paragraph is adding that much. Since we talk about sparsity in reward design, i think we can add a sentence or 2 there that conveys the message here.-->
 Ultimately, we anticipate that getting deep RL to work well on fully detailed conservation decision-making problems will be challenging.
@@
 
+<!-- ML: I think talking about the examples in a very passing manner is fine as the point is primarily to show that RL can work.
+To conclude, I feel like we could portray the vision of open ai gym environments building a CS-ecology community.-->
+
 <!-- 

### 2021-05-20 appendices/appendix-B-examples.Rmd
update paths

@@
 
-write_csv(sims_df, "figs/sims_df.csv")
-write_csv(policy_df, "figs/policy_df.csv")
-write_csv(reward_df, "figs/reward_df.csv")
+write_csv(sims_df, "../manuscript/figs/sims_df.csv")
+write_csv(policy_df, "../manuscript/figs/policy_df.csv")
+write_csv(reward_df, "../manuscript/figs/reward_df.csv")
 ```
@@
 Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
-bifur_df = read_csv("figs/bifur.csv", col_types = "dccd")
+bifur_df = read_csv("../manuscript/figs/bifur.csv", col_types = "dccd")
 bifur_df %>%
@@
 ```{r}
-write_csv(sims_df, "figs/tipping_sims_df.csv")
-write_csv(policy_df, "figs/tipping_policy_df.csv")
-write_csv(reward_df, "figs/tipping_reward_df.csv")
+write_csv(sims_df, "../manuscript/figs/tipping_sims_df.csv")
+write_csv(policy_df, "../manuscript/figs/tipping_policy_df.csv")
+write_csv(reward_df, "../manuscript/figs/tipping_reward_df.csv")
 ```

### 2021-05-21 appendices/appendix-C-environments.Rmd

---
title: "Appendix C: Building Environments for Deep Reinforcement Learning"
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Melissa Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Kari Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu

bibliography: ../manuscript/references.bib

header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}

output: rticles::arxiv_article
---

# Defining an Environment

In Deep Reinforcement Learning (RL), and _environment_ is a mathematical model, computer simulation, or real world situation with which the agent interacts.
Defining an _environment_  requires 5 components:

1. Describe the states defining the environment
2. Describe the possible actions available to the agent
3. Update it's internal _state_ in response to the _action_ of the _agent_
4. Provide the agent with an _observation_ of the new state 
5. Provide the agent with a _reward_ based on the its state and the agent's action.

Mathematically, these actions can be expressed as functions: a _step function_ $f$ providing the new state, $s_{t+1}$ given the current state $s_t$ and the agent's most recent action $a_t$:

\[ s_{t+1} = f(s_t, a_t), \]

a observation function (sometimes called the 'emission' function, $E$) giving the observed state as a function of the current state and possibly also the action (i.e. different actions could potentially influence the accuracy of the observation):

\[ o_{t+1} = E(s_t, a_t) \]

and a reward $r_t$ given by a utility function $U$ of the current state and action,

\[ r_t = U(s_t, a_t) \]

This corresponds to the formal definition of a Partially Observed Markov Decision Process (POMDP), see main text and Appendix A.

# A Minimal Example based on Fishing Quotas

For example, consider the fishing quota problem.

1. The state space is defined as the current total biomass of the fish stock, that is, any non-negative real number.  For simplicity, we might restrict our representation of the state space to a sufficiently large value unlikely to be reached, such as twice the carrying capacity, $s_t \in [0, 2K]$.  Note that in a more general problem, the state space could be a vector of real values, e.g. giving the biomass of different age classes and/or different interacting species.

2. For simplicity, we assume the harvest quota can reach any possible state value, $a_t \in [0, 2K]$.

3. The step function is given by the Gordon-Schaefer model,

\[ s_{t+1} = s_t + r s_t \left( 1 - \frac{s_t}{K} \right) - a_t, \]

(In `gym_fishing-v1` an optional stochastic growth term is also included).

4. observations are made without error, $o_t = s_t$,

5. and the utility is given by the amount harvested (i.e. the harvest quota, so long as it does not exceed the total stock):

\[ r_t = \min(a_t, x_t) \]

# A Minimal Example in Python

In order to interface with existing frameworks for implementing deep RL, it is convenient to express these mathematical functions as computational functions.  By following the specification defined in @brockman2016, we can create a python module, often referred to as a "gym", that expresses the five elements in standardized form.

\pagebreak

A minimal `gym` object inherits from the `gym.Env` class, and provides a `step()` method and `reset()` method:

```python
import gym
import numpy as np
from gym import spaces

class FishingEnv(gym.Env):

    def __init__(self, r = 0.3, K = 1, s0 = 0.75, Tmax=100):

        # parameters
        self.K = K
        self.r = r
        self.s0 = s0
        self.Tmax = Tmax

        # Initial conditions
        self.state = self.s0
        self.year = 1

        # State and action space
        self.observation_space = spaces.Box(0., 2.*self.K)
        self.action_space = spaces.Box(0., 2.*self.K)

    def step(self, action):

        # Step function
        self.state = self.state + self.r * self.state (1 - self.state / self.K) - action

        # Utility function
        reward = min(quota, self.state)

        # Observation function
        observed = self.state

        # End conditions
        self.year = self.year + 1
        done = bool(self.year > self.Tmax)

        # Return observation, reward, end condition and optional info 
        return observed, reward, done, {}

    def reset(self):
        self.state = self.s0
        self.year = 0
        return self.state

```

In practice, an environment may include additional logic for rendering information about it's current states as output.
Effective environments may often work in a transformed variable space, for example, `gym_fishing` maps the space $[0,2K]$ into the space $[-1, 1]$, which significantly improves the performance. 
Other features, such as 32 bit precision numeric values for state and action spaces, can improve the performance of the environment on GPU-based computation.

Function inheritance provides a natural way to express many variations of a gym, such as alternate step functions, utility functions, or even alternate action and state spaces.
Standard python module structure facilitates installing and loading a gym. 
Unit testing and continuous integration provide a reliable mechanism to ensure the gym is working as expected.
The `gym_fishing` repository, https://github.com/boettiger-lab/gym_fishing illustrates all of these elements.

# References

### 2021-05-24 manuscript/manuscript.Rmd
just adding a few citations

@@
 <!-- Environmental problems, ML -->
-Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
+Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@Worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
 But will ML approaches really help us tackle a changing planet?
@@
 The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
-Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
+Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
 Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
@@
 The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
-Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
+Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2016] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
 
-For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1974] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
+For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
 In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
@@
 We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
-This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
+This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012] lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
 On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
 For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
-Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2014].
+Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].
 We also enumerate the possible actions which a manager may take in response to environmental degradation.
@@
 The science of ecological management and quantitative decision-making has a long history [e.g.
-@Schaefer1954; @Walters1974] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
-However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
+@Schaefer1954; @Walters1978] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@Hastings2012].
 Complexity comes not only from the ecological processes but also the available actions.
@@
 In contrast, the process of training an RL algorithm on a more complex problem is no different than the simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
-Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @firemodel], disease [@disease], marine ecosystems [@ecopath], or global climate change [@gcm].
+Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
 
@@
 Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
-Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [@gym], and performance can easily be compared against alternatives on public leaderboards [@gym].
+Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [brockman2016].
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].

### 2021-05-24 manuscript/manuscript.Rmd
a few minor comments and tweaks

@@
 RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
-Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
+Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
 Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
-<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
@@
 <!-- Conceptual picture: (A) an "environment", and (B) an "agent"
-
-What do we mean by an "environment"?  
-  - Transition function
-  - "reward"
-  - examples
-
-What do we mean by an "agent"?
-
-  Model-based vs Model free.  We focus on Model free agents which use Neural networks, a subset of deep RL
-
   Components of an agent: 
@@
     - High-level overview of an RL algorithm components (model/model-free, on/off policy, stochastic/deterministic, etc).
-
   conceptual explanation of 'model-free', what it means to have a "policy" without modeling/predicting transitions
@@
 -->
-
-<!-- ML: need to change some of the language here to be consistent with POMDP set up later-->
 
@@
 An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
-<!-- ML: what are classical approaches below here? -->
-In contrast to classical approaches, there are few restrictions on what comprises a state or action. 
-The state may be continuous or discrete, completely or partially observed, single or multidimensional.
-For instance, we might represent an ecological community of 10 interacting species with a 10-dimensional continuously valued state space signifying the population density of each species.
-The action space can be similarly complex, since an action can be continuous or discrete, single or multidimensional.
+In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
+States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
 For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at the different species over a time step.
@@
 
-**Survey of RL Algorithms**
-
 <!--
@@
 Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
-In table **, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
+In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
 In the appendix, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
-
-<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
-
-<!-- Can we add citations to each please -->
-<!--
-
-Should we mention:
-
-- Temporal Differencing, RIENFORCE
-- SARSA, 
-- Double DQN?  
-Can we add HER (since delayed reward case may be particularly interesting)
-Should we do anything to better reflect the 'family tree' here? (e.g. DQN -> DDPG -> TD3)
-
--->
 
@@
 
-We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.
+We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
 
@@
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
-We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 1].
+We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
 Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
@@
 
-A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix, fig S1, S2) further highlights the difference between these two approaches.
+A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix B, fig S1, S2) further highlights the difference between these two approaches.
 Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
@@
 The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
-Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
-The use of a recurrent neural network (RNN) architecture instead of the multi-layer perceptron (MLP) used here would open the door for the policy to reflect the non-autonomous dynamics more directly. 
-<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote-->
+Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
+<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote.  CB: agree, revised above-->
 
@@
 The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
-<!-- ML: I feel like the remaining part of this dont full drive home the benefit of model-free rl, like its able to learn good performing policies even in the case when a model is very difficult or impossible to learn.
--->
-Alternative models can easily be implemented in simulations, allowing an agent to train against both contingencies.
-A successful agent will learn a policy that works well in both scenarios.
-This is analogous to classical approaches for integrating over parameter or model uncertainty [e.g. @Walters1981; @Ludwig1982],
+Model-free approaches have proven successful at learning effective policies even when the underlying model may be difficult or impossible to learn [@alphago], as long as simulations of possible mechanisms are available.
 
@@
 <!-- ML: I feel like this paragraph could be condensed with the one above -->
+<!-- CB: I think "model free is great" and "real problems are harder than these examples" are two distinct messages though and at least one would get lost in mashing them together.
+Perhaps the message doesn't come across in the current version.
+Does look like this paragraph is similar in overall message to the next one though, maybe could be combined with that?
+-->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
+<!-- Marcus is working on tightening up these next few paragraphs, precisely which issues actually get highlighted TBD-->
 In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: instability, interpretability, and reward design.

### 2021-05-25 manuscript/manuscript.Rmd
fix the minor issues flagged in #17
will save larger edit of discussion paragraphs on 'rl is hard' for separate PR

@@
 States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
-For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at the different species over a time step.
 The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
@@
 The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
-Model-free approaches have proven successful at learning effective policies even when the underlying model may be difficult or impossible to learn [@alphago], as long as simulations of possible mechanisms are available.
+Model-free approaches have proven successful at learning effective policies even when the underlying model may be difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
 

### 2021-05-25 python/conservation_A2C.py
mapping hyperparameters over

@@ -1,0 +1,33 @@
+# Fishing using A2C
+
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_conservation
+from torch import nn as nn
+from stable_baselines3.common.env_util import make_vec_env
+
+seed = 24
+# Create a fishing environment
+env = make_vec_env("conservation-v6", n_envs = 4, seed = seed)
+
+# Create an A2C agent
+a2c = sb3.A2C("MlpPolicy",
+              env, 
+              seed=seed,
+              gamma= 0.9999,
+              normalize_advantage= True,
+              max_grad_norm= 0.6,
+              use_rms_prop= False,
+              gae_lambda= 0.8,
+              n_steps= 128,
+              learning_rate= 0.021243170810461984,
+              ent_coef= 1.7498876702865585e-07,
+              vf_coef= 0.12495864179484306,
+              policy_kwargs= dict(ortho_init=False))
+
+# Train the agent for a fixed number of timesteps
+a2c.learn(total_timesteps=300000)
+
+# Save our trained agent for future use
+a2c.save("cache/conservation_a2c")

### 2021-05-25 python/conservation_PPO.py

@@ -1,0 +1,31 @@
+from stable_baselines3 import PPO
+from stable_baselines3.common.env_util import make_vec_env
+from torch import nn as nn
+import gym_conservation
+
+seed = 24
+env = make_vec_env("conservation-v6", n_envs = 4, seed = seed)
+
+agent = PPO('MlpPolicy', 
+            env, 
+            seed = seed,
+            batch_size= 64,
+            n_steps= 256,
+            gamma= 0.9,
+            learning_rate= 0.0008330851234475645,
+            ent_coef= 7.290496088342335e-08,
+            clip_range= 0.3,
+            n_epochs= 1,
+            gae_lambda= 0.95,
+            max_grad_norm= 0.5,
+            vf_coef= 0.7638921186547223,
+            policy_kwargs= dict(activation_fn=nn.ReLU,
+                                net_arch=[dict(pi=[256, 256], vf=[256, 256])]
+                                )
+          )
+agent.learn(total_timesteps = 3e5)
+
+# Save our trained agent for future use
+agent.save("cache/conservation_ppo")
+
+

### 2021-05-25 python/fishing_A2C.py

@@ -4,34 +4,34 @@ import numpy as np
 import stable_baselines3 as sb3
 import gym
 import gym_fishing
+from torch import nn as nn
+from stable_baselines3.common.env_util import make_vec_env
+
 
 seed = 24
-np.random.seed(seed)
-
 # Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.1)
+env = make_vec_env("fishing-v1", sigma=0.1, n_envs = 4, seed = seed)
 
 # Create an A2C agent
-a2c = sb3.A2C("MlpPolicy", env, seed=seed)
+a2c = sb3.A2C("MlpPolicy",
+              env, 
+              seed=seed,
+              gamma= 0.995,
+              normalize_advantage= False,
+              max_grad_norm= 0.7,
+              use_rms_prop= False,
+              gae_lambda= 0.92,
+              n_steps= 8,
+              learning_rate= 0.0006987399679768267,
+              ent_coef= 3.585469128347999e-05,
+              vf_coef= 0.48712388559858977,
+              policy_kwargs= dict(activation_fn=nn.ReLU,
+                                 ortho_init=True,
+                                 net_arch=[dict(pi=[256, 256], vf=[256, 256])]
+                                 ))
 
 # Train the agent for a fixed number of timesteps
 a2c.learn(total_timesteps=300000)
 
 # Save our trained agent for future use
-a2c.save("cache/a2c")
-
-# Load the saved agent
-a2c = sb3.A2C.load("cache/a2c")
-
-
-# Example simulation of the agent
-state = env.get_state(0.75)
-for i in range(1, 10):
-    action, null = a2c.predict(state)
-    state, reward, done, info = env.step(action)
-
-## Or more consisely:
-a2c_sims = env.simulate(a2c, reps=500)
-
-# Estimate the 'Policy function'
-a2c_policy = env.policyfn(a2c, reps=50)
+a2c.save("cache/fishing_a2c")

### 2021-05-25 python/fishing_PPO.py

@@ -1,0 +1,29 @@
+from stable_baselines3 import PPO
+from stable_baselines3.common.env_util import make_vec_env
+import gym_fishing
+
+seed = 24
+env = make_vec_env("fishing-v1", n_envs = 4, seed = seed)
+total_timesteps = 3e5
+
+agent = PPO('MlpPolicy', 
+          env, 
+          seed = seed,
+          batch_size= 8,
+          n_steps= 64,
+          gamma= 0.95,
+          learning_rate= 0.009063123943359215,
+          ent_coef= 0.002465145165808194,
+          clip_range= 0.4,
+          n_epochs= 5,
+          gae_lambda= 0.98,
+          max_grad_norm= 0.8,
+          vf_coef= 0.8153389910080954
+          )
+
+agent.learn(total_timesteps = 3e5)
+
+# Save our trained agent for future use
+agent.save("cache/fishing_ppo")
+
+

### 2021-05-26 appendices/testing_figures.Rmd
testing

--- 2021-05-20 appendices/appendix-B-examples.Rmd
@@
   - these are optional and can be removed
-bibliography: references.bib
+bibliography: ../manuscript/references.bib
 
@@
 ```{r knitr, include = FALSE}
-knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
+knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
                       fig.width = 7, fig.height = 4, cache = FALSE)
@@
 scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
-pal = ggthemes::solarized_pal()(8)
+pal = ggthemes::solarized_pal()(6)
 txtcolor = "#586e75"
 ```
-
-# Building a simulation environment `gym`
-
-[@openai_gym]
-
-# Deep Reinforcement Learning Frameworks
-
-At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
-While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
-Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
-Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3].
-Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
-Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
-However, higher-level interfaces and high-level documentation are still relatively limited.
-OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
-OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
-Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption.
-Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
-The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
-Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.
-
-# Deep Reinforcement Learning in R
-
-Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
-Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
-In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
-
-In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
-This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
-Alternately, users may prefer running the analysis inside a docker container.  
-The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
-This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.
-
-Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
-From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
-
-```{r}
-#install.packages("renv")
-renv::restore()
-```
-
-Once the packages have installed, we are ready to load the necessary libraries.
-Note that the `import` function from `reticulate` package acts much like the `library` command in R, though it does not attach the package function to the global namespace.
-To make it more convenient to access those functions, we can assign a shorthand name.
 
@@
 gym_conservation = import("gym_conservation")
-```
 
-Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration.
-In addition to setting a random seed in our Python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.
-
-```{r}
 ## reproducible settings
@@
 
-<!-- is this too verbose? --> 
-The above code also illustrates a few conventions which may be helpful to bear in mind when using the `reticulate` interface to interact with Python from R: 
-it is often necessary for integer values to be explicitly typed as integers by adding a trailing `L` (corresponding to the primitive C type `long` integer).
-Python is also more strongly object-oriented than many R packages, where "methods" of an "object" are accessed with the list-subset operator `$` in R (equivalent to the use of `.` in Python).
-Lastly, while R can use `<-` or `=` for assignment, Python uses only `=`. For simplicity we will stick with the latter.  
-With few exceptions, the R code shown here can be re-written as python code by dropping the `L` and replacing `$` with `.`, see stand-alone python code in the `python/` subdirectory of the repository.
-
-# Finding a known optimal solution using RL
-
-## Sustainable Harvest Quotas
-
-We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
-
 ```{r }
@@
 
-<!-- Kari: the role neural networks play in RL is not very well explained in the paper, more detail there would probably make this part more intuitive --> 
-At the heart of each deep RL algorithm is a neural network or a collection of neural networks.
-The RL agent uses these networks to approximate a policy function and/or a value function.
-The policy function is the agent's mapping of states to actions -- i.e.
-the agent's strategy.
-Value functions attempt to evaluate the goodness of a state or state-action pair towards achieving the RL objective of maximizing cumulative rewards.
-Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyper-parameters_ while the parameters of a RL algorithm that are learned during training are called parameters.
-For example, the number of layers in a neural network is a hyper-parameter but a weight in the neural net is referred to as a parameter.
-Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
-Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms,
-for instance, the A2C algorithm originally described in @A2C.
-Here, we train an agent using the A2C algorithm using default hyper-parameter settings with a multi-layer perceptron policy and value network composed of two 64-neuron layers.
+```{r}
+a2c = sb3$A2C$load("../python/cache/fishing_a2c")
+a2c_sims = env$simulate(a2c, reps = 100L)
+a2c_policy = env$policyfn(a2c, reps = 5L)
+```
 
 ```{r}
-a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
+ppo = sb3$A2C$load("../python/cache/fishing_ppo")
+ppo_sims = env$simulate(ppo, reps = 100L)
+ppo_policy = env$policyfn(ppo, reps = 5L)
 ```
-
-Training is the main computationally intensive process, which can take anywhere from a few minutes to many days, depending on the complexity of the environment, the neural network architecture and the number of training iterations budgeted.
-Therefore, we save the trained agent to disk, and only execute the training method (`learn`) if no previously saved agent is found:
-
-```{r}
-if(!file.exists("cache/a2c.zip")) {
-
-  # Train the agent for a fixed number of timesteps
-  a2c$learn(total_timesteps=300000L)
-
-  # Save our trained agent for future use
-  a2c$save("cache/a2c")
-}
-```
-
-Note that while default hyper-parameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyper-parameter selection.
-This is discussed further below.
-Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.
-Note that a copy of the trained agents are included in the corresponding GitHub repository.
-
-```{r}
-# Simulate management under the trained agent
-a2c = sb3$A2C$load("cache/a2c")
-```
-
-We can supply an observation to our trained agent and it will return it's proposed action using the `predict` method. 
-This is all we need to evaluate or employ the agent on a decision-making task.
-Recall that state space and action space in the fishing gym have been re-scaled to a (-1, 1) interval.
-Note that this is equivalent to a choice of appropriate units -- we can re-scale the interval without loss of generality.
-This is often an important step in the design of an RL environment to facilitate successful training.
-Following the `gym` standard, the core methods such as `predict` and `step` operate on the re-scaled units,
-so it is necessary to first transform the original units into this re-scaled state space.
-For example, if we wish to start a simulation with a stock size of 0.75, we can use the helper
-method `get_state()`, to determine the corresponding value in the re-scaled state space.  
-Unlike `predict` and `step`, `get_state()` is not a standard method of all gym environments -- typically
-a user must first inspect the state space of an environment and choose themselves how to re-scale their
-problem into that state space.
-
-```{r}
-
-## represent the initial state size in the 'rescaled' state space.
-state = env$get_state( 0.75 )
-state
-```
-
-With an initial state in hand, we are ready to simulate management using our agent.
-The iteration is simple: we use the agent to predict what action we should take given the current state. Then, we take said action and examine the result to determine the future state. 
-Because these methods return additional information as well, a little extra sub-setting is required in R:
-
-```{r}
-for(i in 1:10){
-
-  out = a2c$predict(state)
-  action = out[[1]]
-  result = env$step(action)
-  state = result[[1]]
-
-}
-
-```
-
-For convenience, `gym_fishing` defines the helper routine `simulate` to perform the above iteration `reps` number of times.
-The `simulate` method returns the state, action, and reward resulting from each time step of each replicate:
-
-```{r}
-a2c_sims = env$simulate(a2c, reps = 500L)
-```
-
-Because the agent typically attempts to retain the fish population near a specific value, simulations with well-trained agents will usually not explore the full range of possible states.
-To get a better idea of how the agent behaves across the full state space, it is common to examine the policy function: to do this, we'll plot the action taken at every possible state.
-Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
-Note that the `policyfn` method is another custom method and not part of the `gym` standard.
-The use of `policyfn` will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTMs [@LSTM].
-
-```{r}
-a2c_policy = env$policyfn(a2c, reps = 50L)
-```
-
-## TD3
-
-We train a second agent based on the TD3 algorithm. 
-This illustrates some of the salient differences between the different algorithms used in reinforcement learning.
-Bear in mind that the best algorithm for any given environment will depend on the details of the environment as well
-as the length of training and the selection of hyper-parameters.
-This time, we will declare our choices for the hyper-parameters explicitly.
-Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
-Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
-For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
-
-```{r}
-# train an agent (model) on one of the environments:
-
-policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
-# non-episodic action noise:
-noise_std = 0.6656948079225263
-n_actions = env$action_space$shape[0]
-action_noise = sb3$common$noise$NormalActionNoise(
-        mean=np$zeros(n_actions),
-        sigma= noise_std * np$ones(n_actions)
-        )
-
-td3 = sb3$TD3( 'MlpPolicy', 
-               env,  
-               verbose=0L, 
-               seed = seed,
-               policy_kwargs = policy_kwargs,
-               learning_rate = 0.0001355522450968401,
-               gamma= 0.995, 
-               batch_size = 128L,
-               buffer_size = 10000L,
-               train_freq = 128L, 
-               n_episodes_rollout = -1, 
-               gradient_steps = 128L,
-               action_noise = action_noise
-              )
-```
-
-Once again, after our agent is defined we are ready to train it:
-
-```{r }
-if(!file.exists("cache/td3.zip")){
-
-  td3$learn(total_timesteps=300000L)
-  td3$save("cache/td3")
-
-}
-```
-
-With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:
 
@@
 td3_sims = env$simulate(td3, reps = 500L)
-td3_policy = env$policyfn(td3, reps = 50L)
+td3_policy = env$policyfn(td3, reps = 5L)
 ```
-
-We compare the performance of these two RL agents trained using the algorithms A2C and TD3 respectively to the known optimal policy for this particular environment.  
-Recall that under the assumptions of the simple model used in the `fishing-v1` environment, we can determine the optimal harvest policy analytically if the model and parameters are known precisely [@Reed1979].
-The optimal strategy is a policy of 'constant escapement', designed to keep the remaining stock size (the population that 'escapes' fishing harvest) at the biomass corresponding to a maximum growth rate, i.e.
-at $B = K/2$ in this model.
-`gym_fishing` defines a collection of non-RL agents in the `models` submodule, including the a human agent that merely asks to enter their desired quota manually. 
-The `escapement` model implements the provably optimal constant escapement rule.
-A third model, `msy`, implements a policy based on "Maximum Sustainable Yield" policy [@Schaefer1954], which is actually more commonly used as a basis for management than constant escapement, despite only being optimal at the steady state under deterministic dynamics.
 
@@
 opt = gym_fishing$models$escapement(env)
-opt_sims = env$simulate(opt, reps = 500L)
+opt_sims = env$simulate(opt, reps = 100L)
 opt_policy = env$policyfn(opt)
@@
 
-Having performed simulations under each method, we gather the results under each management strategy together into a single data frame for simulations (`sims_df`), the policy function (`policy_df`) and the cumulative reward, used to make the plots in the corresponding panels.
-We preserve a copy of each data frames in local `.csv` text files for maximum cross-platform compatibility.
-
 ```{r}
-sims_df = bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
+sims_df = bind_rows(td3_sims, a2c_sims, ppo_sims, opt_sims, .id = "model") %>%
+  mutate(model = c("TD3", "A2C", "PPO", "optimal")[as.integer(model)])
 
 policy_df = bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
-  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
+  mutate(model = c("TD3", "A2C", "PPO", "optimal")[as.integer(model)])
 
@@
 
-write_csv(sims_df, "figs/sims_df.csv")
-write_csv(policy_df, "figs/policy_df.csv")
-write_csv(reward_df, "figs/reward_df.csv")
 ```
-
-<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
-Standard R methods are used to plot the results summarized across the replicates:
 
@@
 
-## Manuscript Figure 1
-
-<!-- Not sure if it's helpful to duplicate figures from the paper in the appendix...-->
-
 ```{r fig1}
@@
 
-## Applying an RL agent to real data
-
-We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.
-
-Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
-We will use the agent we have just trained using the `gym_fishing` simulator above,
-
-```{r}
-# Simulate management under the trained agent
-env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-td3 = sb3$TD3$load("cache/td3")
-
-```
-
-```{r}
-hake = read_csv("../data/hake.csv")
-x0 = hake %>%
- filter(year == min(year)) %>%
-  pull(scaled_biomass)
-Tmax = 15
-years =1986:2000
-
-td3_state = td3_action = numeric(Tmax)
-td3_state[1] = x0
-td3_action[1] = NA
-
-```
-
-Our hindcast considers the scenario of managing catch quotas using TD3, beginning in 1986. 
-We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
-This shows that quotas set by TD3 would have been lower than historical values throughout the late 80s and early 90s, which
-saw a sharp decline in estimated biomass of the Argentine Hake.
-For the same or higher estimated stock sizes from the historical record, we see that TD3 would recommend a lower quota.
-We also see that under TD3 the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
-being set in the now-depleted stock.  
-Whether the TD3 quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
-However, the fact that TD3 would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
-Combined with the observed historical declines, this provides some evidence that the RL agent could have decreased or avoided the over-harvesting in this case.
-
-```{r}
-env$reset()
-## represent the initial state size in the 'rescaled' state space.
-state = env$get_state( x0 )
-
-for(i in 2:Tmax){
-
-  # RL-recommended harvest action:
-  out = td3$predict(state)
-  action = out[[1]]
-
-  # Record state and resulting action
-  td3_state[i] = env$get_fish_population(state)
-  td3_action[i] = env$get_quota(action)
-
-  # Implement RL-recommended harvest. 
-  result = env$step(action)
-  state = result[[1]]
-
-  #state = env$get_state(hake$scaled_biomass[i+1])
-}
-
-df = bind_rows(
-  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
-  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
-
-df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))
-
-```
-
-# Ecological tipping points
-
-Our second example utilizes our `gym_conservation` to provide the necessary environment to simulate an ecosystem approaching a tipping point.
-
-## Tipping point model
-
-The tipping point model is based on the consumer-resource model of @May1977, which creates alternative stable states, 
-which we subject to log-normal environmental noise:
-
-\[
-\mu_{t} = X_t + X_t r \left(1 - \frac{X_t}{K} \right) - \frac{a_t X_t^q}{X_t^q + b^q}
-\]
-\[
-X_{t+1} \sim \text{lognormal}(\mu_t, \sigma)
-\]
-
-where we take $r=0.7$, $K=1.2$, $q=3$, $b=0.15$, $a_0 = 0.19$ and $\sigma = 0.2$ 
-Slow change over time in the parameter $a_t$ represents a process of environmental degradation, modeled as a constant increment $a_{t+1} = a_t + \alpha$, where we will take $\alpha = 0.001$.
-This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carptenter2011] to the planet biosphere [@Barnosky2014].
-
-```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario.
-The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g.
-global mean temperature, arbitrary units) of 0.19.
-In the absence of conservation action, the environment worsens (e.g.
-rising mean temperature) as the parameter increases.
-This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.
-Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
-bifur_df = read_csv("figs/bifur.csv", col_types = "dccd")
-bifur_df %>%
-  ggplot(aes(parameter, state, lty=equilibrium, group = group)) + 
-  geom_line()
-
-```
-
-We assume the benefit provided by the ecosystem state is assumed to be directly proportional to the state itself, $b x_t$.  
-We further assume that each year the manager has the option to slow or reverse the environmental degradation by taking action $A_t$, such that under management, the resulting environment in the next time step is given by
-
-\[
-a_{t+1} = a_t + \alpha - A_t
-\]
-
-We assume the cost associated with that action to be proportional to the square of the action, such that large actions are proportionally more costly than small ones.
-Consequently, the utility at time $t$ is given by the sum of costs and benefits:
-
-\[U(X_t, A_t) = b X_t - c A_t ^2\]
-
-where we will take \(b = 1\) and \(c=1\).  
-Our implementation in `gym_conservation` allows the user to consider alternative parameter choices, alternative models for the ecological dynamics, and alternate types of actions, such as manipulating the ecosystem state directly rather than manipulating the environmental parameter.
-For some such scenarios the optimal solution is known or can be determined by stochastic dynamic programming, while for others, including the scenario of focus here, the optimal solution is unknown.
-
-For simplicity of illustration, we will train only a TD3-based agent on this scenario.
+# Conservation
 
@@
 ```{r}
-noise_std = 0.4805935357322933
-
-OU = sb3$common$noise$OrnsteinUhlenbeckActionNoise
-action_noise = OU(mean = np$zeros(1L),  sigma = noise_std * np$ones(1L))
-
-model = sb3$TD3('MlpPolicy', 
-            env, 
-            verbose = 0, 
-            seed = 42L,
-            "gamma"= 0.995,
-            "learning_rate"=  8.315382409902049e-05,
-            "batch_size"= 512L,
-            "buffer_size"= 10000L,
-            "train_freq"= 1000L,
-            "gradient_steps"= 1000L,
-            "n_episodes_rollout"= -1L,
-            "action_noise"= action_noise,
-            "policy_kwargs"= list("net_arch"= c(64L,64L)))
-
+model = sb3$TD3$load("cache/td3-conservation")
+TD3_sims = env$simulate(model, reps = 100L)
+TD3_policy = env$policyfn(model, reps = 5L)
 ```
@@
 ```{r}
-if(!file.exists("cache/td3-conservation.zip")){
-
-  model$learn(total_timesteps=3000000L)
-  model$save("cache/td3-conservation")
-
-}
-
+a2c = sb3$A2C$load("../python/cache/conservation_a2c")
+a2c_sims = env$simulate(a2c, reps = 100L)
+a2c_policy = env$policyfn(a2c, reps = 5L)
 ```
@@
 ```{r}
-# Simulate management under the trained agent
-# See https://github.com/boettiger-lab/conservation-agents/blob//conservation/TD3-v6.py
-
-model = sb3$TD3$load("cache/td3-conservation")
-TD3_sims = env$simulate(model, reps = 100L)
-TD3_policy = env$policyfn(model, reps = 10L)
+ppo = sb3$A2C$load("../python/cache/conservation_ppo")
+ppo_sims = env$simulate(ppo, reps = 100L)
+ppo_policy = env$policyfn(ppo, reps = 5L)
 ```
-
-In general, the optimal solution depends on the ecological dynamics, the benefit of the ecosystem services and the costs associated with a management response.
-Because the tipping point problem is non-autonomous, we cannot solve for the optimal policy even given the model and objective (utility) function using Markov Decision Process methods.
-However, a simple heuristic solution provides a reasonable starting point for comparison.
-
-As discussed in the main text, we have no existing optimal solution to the tipping point problem, and so rely on a common heuristic strategy instead:
-select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in it's current state.
-This is implemented using the `fixed_action` method provided in our `gym_conservation` module, which also implements other heursitic models, including a human agent which requires interactive input to select the action each year.
 
@@
 ```{r}
-sims_df = bind_rows(TD3_sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "steady-state")[as.integer(model)])
+sims_df = bind_rows(TD3_sims, a2c_sims, ppo_sims, opt_sims, .id = "model") %>%
+  mutate(model = c("TD3", "A2C", "PPO", "steady-state")[as.integer(model)])
 
-policy_df = bind_rows(TD3_policy, opt_policy, .id = "model") %>%
-  mutate(model = c("TD3", "steady-state")[as.integer(model)])
+policy_df = bind_rows(TD3_policy, a2c_policy, ppo_policy, opt_policy, .id = "model") %>%
+  mutate(model = c("TD3", "A2C", "PPO", "steady-state")[as.integer(model)])
 
@@
 
-The resulting three data frames contain the necessary data for each of the subplots in figure 3 of the main text.
-
 ```{r}
-write_csv(sims_df, "figs/tipping_sims_df.csv")
-write_csv(policy_df, "figs/tipping_policy_df.csv")
-write_csv(reward_df, "figs/tipping_reward_df.csv")
-```
-
-<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
-
-```{r}
-# ensemble statistics
-ymin = function(x) last(x[(ntile(x, 20)==1)])
-ymax = function(x) last(x[(ntile(x, 20)==19)])
-
 fig_sims = 
@@
    geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-```
 
-```{r}
 fig_policy = 
@@
   geom_line(show.legend = FALSE)
-```
-
-```{r}
 
@@
   ylab("reward")
-```
 
-### Manuscript Figure 3
-
-```{r fig3}
 fig_sims / ( fig_policy + fig_reward)
 ```
-
-Because it is difficult to get a feel for the dynamics of individual replicate simulations from ensemble statistics, we select a few example trajectories to examine directly. 
-Of the 100 replicate simulations, we pick 4 examples that dip below a state value of 0.2 for over 15 consecutive timesteps, indicating a transition into the lower basin of attraction. 
-Comparing the dynamics under the rule of thumb steady-state strategy to that of the RL-trained agent, it is clear that the RL agent does a better job at both avoiding tipping points and promoting the recovery of those selected trajectories that cross into the lower attractor.
-
-```{r}
-# Some individual replicates, for comparison
-
-## First 4 of the TD3 reps falling below .2 for more than 15 steps
-is_low = sims_df %>% 
-  filter(model == "TD3") %>% 
-  group_by(rep, model) %>% 
-  summarize(low = sum(state < .2) > 10) %>% 
-  filter(low) %>% head(6)
-
-## First 4 such cases:
-sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state,  group=interaction(model, rep))) +
-  geom_line(color = pal[3], show.legend = FALSE) + facet_wrap(~rep)
-```
-
-```{r}
-# Some individual replicates, for comparison
-is_low = sims_df %>% filter(model == "steady-state") %>%
-  group_by(rep, model) %>% summarize(low = sum(state < .2) > 10) %>% 
-  filter(low) %>% head(6)
-sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state, group=interaction(model, rep))) +
-  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
-```

### 2021-05-26 python/fishing_A2C.py

@@ -10,7 +10,7 @@ from stable_baselines3.common.env_util import make_vec_env
 
 seed = 24
 # Create a fishing environment
-env = make_vec_env("fishing-v1", sigma=0.1, n_envs = 4, seed = seed)
+env = make_vec_env("fishing-v1", n_envs = 4, seed = seed, env_kwargs = {"sigma": 0.1})
 
 # Create an A2C agent
 a2c = sb3.A2C("MlpPolicy",

### 2021-05-26 python/fishing_PPO.py

@@ -3,7 +3,7 @@ from stable_baselines3.common.env_util import make_vec_env
 import gym_fishing
 
 seed = 24
-env = make_vec_env("fishing-v1", n_envs = 4, seed = seed)
+env = make_vec_env("fishing-v1", n_envs = 4, seed = seed, env_kwargs = {"sigma": 0.1})
 total_timesteps = 3e5
 
 agent = PPO('MlpPolicy', 

### 2021-05-26 python/conservation_A2C.py
train with tuned hyperparameters

@@ -31,3 +31,12 @@ a2c.learn(total_timesteps=300000)
 
 # Save our trained agent for future use
 a2c.save("cache/conservation_a2c")
+
+
+# Evaluation
+env = gym.make("conservation-v6")
+sims = env.simulate(agent, reps = 100)
+policy = env.policyfn(agent, reps = 5)
+
+env.plot(sims, "conservation_a2c.png")
+env.plot_policy(policy, "conservation_a2c_policy.png")

### 2021-05-26 python/conservation_PPO.py

@@ -28,4 +28,12 @@ agent.learn(total_timesteps = 3e5)
 # Save our trained agent for future use
 agent.save("cache/conservation_ppo")
 
+# Evaluation
+env = gym.make("conservation-v6")
+sims = env.simulate(agent, reps = 100)
+policy = env.policyfn(agent, reps = 5)
+
+env.plot(sims, "conservation_ppo.png")
+env.plot_policy(policy, "conservation_ppo_policy.png")
+
 

### 2021-05-26 python/fishing_A2C.py

@@ -35,3 +35,11 @@ a2c.learn(total_timesteps=300000)
 
 # Save our trained agent for future use
 a2c.save("cache/fishing_a2c")
+
+# Evaluation
+env = gym.make("fishing-v1", sigma= 0.1)
+sims = env.simulate(agent, reps = 100)
+policy = env.policyfn(agent, reps = 5)
+
+env.plot(sims, "fishing_a2c.png")
+env.plot_policy(policy, "fishing_a2c_policy.png")

### 2021-05-26 python/fishing_PPO.py

@@ -27,3 +27,10 @@ agent.learn(total_timesteps = 3e5)
 agent.save("cache/fishing_ppo")
 
 
+# Evaluation 
+env = gym.make("fishing-v1", sigma= 0.1)
+sims = env.simulate(agent, reps = 100)
+policy = env.policyfn(agent, reps = 5)
+
+env.plot(sims, "fishing_ppo.png")
+env.plot_policy(policy, "fishing_ppo_policy.png")

### 2021-05-26 python/conservation_A2C.py
generate example output for reference

@@ -12,7 +12,7 @@ seed = 24
 env = make_vec_env("conservation-v6", n_envs = 4, seed = seed)
 
 # Create an A2C agent
-a2c = sb3.A2C("MlpPolicy",
+agent = sb3.A2C("MlpPolicy",
               env, 
               seed=seed,
               gamma= 0.9999,
@@ -27,10 +27,10 @@ a2c = sb3.A2C("MlpPolicy",
               policy_kwargs= dict(ortho_init=False))
 
 # Train the agent for a fixed number of timesteps
-a2c.learn(total_timesteps=300000)
+agent.learn(total_timesteps=300000)
 
 # Save our trained agent for future use
-a2c.save("cache/conservation_a2c")
+agent.save("cache/conservation_a2c")
 
 
 # Evaluation

### 2021-05-26 python/conservation_PPO.py

@@ -2,6 +2,7 @@ from stable_baselines3 import PPO
 from stable_baselines3.common.env_util import make_vec_env
 from torch import nn as nn
 import gym_conservation
+import gym
 
 seed = 24
 env = make_vec_env("conservation-v6", n_envs = 4, seed = seed)

### 2021-05-26 python/fishing_A2C.py

@@ -13,7 +13,7 @@ seed = 24
 env = make_vec_env("fishing-v1", n_envs = 4, seed = seed, env_kwargs = {"sigma": 0.1})
 
 # Create an A2C agent
-a2c = sb3.A2C("MlpPolicy",
+agent = sb3.A2C("MlpPolicy",
               env, 
               seed=seed,
               gamma= 0.995,
@@ -31,10 +31,10 @@ a2c = sb3.A2C("MlpPolicy",
                                  ))
 
 # Train the agent for a fixed number of timesteps
-a2c.learn(total_timesteps=300000)
+agent.learn(total_timesteps=300000)
 
 # Save our trained agent for future use
-a2c.save("cache/fishing_a2c")
+agent.save("cache/fishing_a2c")
 
 # Evaluation
 env = gym.make("fishing-v1", sigma= 0.1)

### 2021-05-26 python/fishing_PPO.py

@@ -1,6 +1,7 @@
 from stable_baselines3 import PPO
 from stable_baselines3.common.env_util import make_vec_env
 import gym_fishing
+import gym
 
 seed = 24
 env = make_vec_env("fishing-v1", n_envs = 4, seed = seed, env_kwargs = {"sigma": 0.1})

### 2021-05-26 manuscript/manuscript.Rmd
Just keeping up to date

@@
 <!-- Marcus is working on tightening up these next few paragraphs, precisely which issues actually get highlighted TBD-->
-In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: instability, interpretability, and reward design.
+In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: generalizability, interpretability, and reward design.
 
 <!-- reason 1: instability-->
-<!-- ML: I think this needs to be tightened up a bit-->
-One leading challenge for deep RL today is the problem of instability.
-RL agents trained on narrowly defined environments may perform poorly even when the environment is altered in seemingly trivial way, such as a re-scaling of parameters.
-To avoid this, researchers should seek to both train and evaluate agents against a range of environments representing different parameters or processes dynamics.
-Stable performance can be challenging even when using the same environment.
-For a range of reasons, including high variance [@cite] and the exploration vs.
-exploitation trade-off [@cite], model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
-For instance, it is not uncommon to observe vastly different behavior when an agent is initialized with different random seeds [cite ibid].
-Such unstable behavior can often be avoided by alternate selection of hyperparameters, so researchers should consider stability as well as performance during hyperparameter selection.
-The choice of RL algorithm, design of the reward function, and duration of training can also influence stability.
+<!-- ML: pivoting this to generalizability but still rewrote past paragraph which might fit in the appendix somewhere-->
+Currently, deep RL agents struggle to generalize their skills [@generalization].
+This inability can be observed in training: if you train an agent on an environment that demands learning a wide range of tasks, the agent will likely fail to converge to an optimal policy [@CITE].
+And, during evaluation, if you give an agent any novel task, including tasks that are closely related to the examples that the agent saw in training, the agent will likely fail to generalize its policy to this new situation [@CITE].
+These failures are all due to the brittleness of deep neural networks. 
+Across deep learning, neural networks tend to perform exceptionally well on narrowly specified problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@CITE].
+
+<!-- 
+A significant current challenge in deep RL is the problem of instability.
+For a range of reasons, including high variance of returns and the exploration vs.
+exploitation dilemma, deep RL algorithms tend to have unstable performance on environments with large state-action spaces [@drlthatmatters].
+For example, @drlthatmatters show that initializing the same agent with different random seeds can lead to significant differences in agent performance.
+Such unstable behavior can be avoided by finding a robust set of hyperparameters or using algorithms that are designed to have greater reliability [@qprop], thus researchers should consider stability as well as performance during hyperparameter tuning and algorithm selection.
+-->
 
@@
 -->
-Interpretability is a common challenge at the adoption of most ML methods [@cite]. 
+Interpretability is an issue that may hinder the adoption of deep RL methods [@cite].
+
 Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.

### 2021-05-26 python/fishing_TD3_untuned.py
cleaning up

@@ -1,0 +1,29 @@
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.1)
+
+td3 = sb3.TD3(
+    "MlpPolicy",
+    env,
+    seed=seed)
+
+td3.learn(total_timesteps=300000)
+td3.save("cache/td3_untuned")
+
+
+# Simulate management under the trained agent
+td3 = sb3.TD3.load("cache/td3_untuned")
+td3_sims = env.simulate(td3, reps=500)
+td3_policy = env.policyfn(td3, reps=50)
+
+
+env.plot(sims, "fishing_td3_untuned.png")
+env.plot_policy(policy, "fishing_td3_untuned.png")
+

### 2021-05-26 manuscript/manuscript.Rmd
Marginal slow improvement here. Still trying to frame the reward paragraph

@@
 
-<!-- reason 1: instability-->
+<!-- reason 1: generalizability-->
 <!-- ML: pivoting this to generalizability but still rewrote past paragraph which might fit in the appendix somewhere-->
 Currently, deep RL agents struggle to generalize their skills [@generalization].
-This inability can be observed in training: if you train an agent on an environment that demands learning a wide range of tasks, the agent will likely fail to converge to an optimal policy [@CITE].
-And, during evaluation, if you give an agent any novel task, including tasks that are closely related to the examples that the agent saw in training, the agent will likely fail to generalize its policy to this new situation [@CITE].
+This inability can affect performance during training and at evaluation. 
+For instance, training an agent on an environment that demands learning an assortment of skills is typically much more difficult than training an agent to learn one specific skill [@CITE].
+And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent will likely fail to generalize its learned behavior to this new situation [@CITE].
 These failures are all due to the brittleness of deep neural networks. 
 Across deep learning, neural networks tend to perform exceptionally well on narrowly specified problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@CITE].
+The ability to adapt to new situations is critical for any decision-making problem in the real world, so the current lack of this ability in deep RL may limit deep RL's effectiveness to conservation issues.
 
@@
 <!-- if we're really recommending 'testing' for stability, maybe we need to demo that in the appendix-->
-
 <!-- reason 2: interpretability -->
@@
 -->
-Interpretability is an issue that may hinder the adoption of deep RL methods [@cite].
-
-Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
-As our examples have illustrated, it may be possible to gain intuition on an agent's behavior by visualizing the action selected against each possible state, though this becomes more difficult in large problems.
-Another challenge is that Deep RL algorithms are very sample inefficient, all though progress is being made to improve this [@cite].
-<!-- ML: I think we can cut this sentence, mentionned it in passing earlier-->
--->
+Compounded with their difficulties to generalize, deep RL agents also suffer from an inability to be fully interpreted [@cite].
+Deep neural networks are commonly referred to as black box models, since it is not currently possible to explain why a deep neural network produces a specific output from a given input.
+In the context of deep RL, this means that we cannot explain why an agents makes a decision besides that the agent is attempting to maximize cumulative rewards.
+The lack of explainability dovetails dangerously with generalizability concerns. 
+Because we are not able to interpret deep RL agents, we are not able to anticipate how an agent will behave when in an unfamiliar situation.
+Despite limited progress to make neural networks more explainable, interpretability will likely be a concern that limits adoption.
 
@@
 <!-- ML: I'll take a pass at this paragraph in a new branch. I think it would be useful to put terms on stuff in here, like reward shaping and reward hacking which we are talking around. -->
-An issue that is central to any applied optimization problem is using the right objective.
-There have been many examples in which explicit reward design goes very wrong [cite].
-And for ecological systems, which are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
+A fundamental issue for all applied optimization problems is using the right objective.
+There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@CITE].
+For ecological systems, which are typically very complex and difficult to fully understand, it seems that bad reward design is quite likely.
+To avoid these undesirable outcomes, RL researchers often *shape* the reward functions so that the agent can more easily learn a good-performing policy [@cite].
+This approach is particularly common when rewards are *sparse*, such as in a game of Chess where the objective is to win and there are no true immediate rewards.
+In this case, one might shape the reward function so that the agent is awarded for capturing the opponent's pieces, but this could easily lead to a policy where the agent focuses on capturing pieces instead of winning the game.
+Since benefits in conservation typically come long after their corresponding actions, sparse reward.
 Moreover, specifying a reward function for conservation brings up various ethical concerns -- e.g.
 who gets to specify the reward function and will the values of the greater community be represented in the reward function.
-RL researchers often introduce modified reward functions during training that help an agent perform better on the original objective [@cite].
-This approach is particularly common when rewards are sparse, such as in a game of Chess where the only objective that counts is whether or not you win, but during training, an RL algorithm may reward the agent for intermediate steps such as capturing a piece.
 

### 2021-05-27 appendices/appendix-B-examples.Rmd
appendix B: tuned vs untuned

@@
 
-keywords:
-  - blah
-  - blee
-  - bloo
-  - these are optional and can be removed
-bibliography: references.bib
+bibliography: ../manuscript/references.bib
 
@@
 
-# Building a simulation environment `gym`
-
-[@openai_gym]
+In this appendix, we illustrate how to apply a deep reinfocement learning framework to train previously published RL algorithms on the novel decision environments as illustrated in the main text.
+Appendix C discusses the construction of such environments in code.
 
@@
 Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
+Note that the RMarkdown source-code for this file can be found in the `appendices` directory of the project repository.
 From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
 
-```{r}
+```{r message=FALSE}
 #install.packages("renv")
@@
 gym_conservation = import("gym_conservation")
+
+# 
+source("../R/plotting-helpers.R")
 ```
@@
 
-<!-- Kari: the role neural networks play in RL is not very well explained in the paper, more detail there would probably make this part more intuitive --> 
+## An optimal solution
+
+Recall that under the assumptions of the simple model used in the `fishing-v1` environment, we can determine the optimal harvest policy analytically if the model and parameters are known precisely [@Reed1979].
+The optimal strategy is a policy of 'constant escapement', designed to keep the remaining stock size (the population that 'escapes' fishing harvest) at the biomass corresponding to a maximum growth rate, i.e.
+at $B_{MSY} = K/2$ in this model.
+`gym_fishing` defines a collection of non-RL agents in the `models` submodule, including the a human agent that merely asks to enter their desired quota manually. 
+The `escapement` model implements the provably optimal constant escapement rule.
+A third model, `msy`, implements a policy based on "Maximum Sustainable Yield" policy [@Schaefer1954], which is actually more commonly used as a basis for management than constant escapement, despite only being optimal at the steady state under deterministic dynamics.
+
+```{r}
+# Simulate under the optimal solution (given the model)
+opt = gym_fishing$models$escapement(env)
+```
+
+Note that the `escapement` function makes very specific assumptions about the environment that it is given.
+In particular, it assumes we can compute $B_{MSY}$ directly from the internal model of stock recruitment contained in the environment.
+In contrast, the RL methods make no such assumption of being able to access that internal model directly.
+
+```{r}
+opt_sims = env$simulate(opt, reps = 100L)
+opt_policy = env$policyfn(opt)
+```
+
+We can plot the resulting data.frames using standard `ggplot2` methods shown in the `R/plotting-helpers.R`, or in python using either the `plot` methods defined in `gym_fishing` or standard plotting libraries.
+
+```{r}
+sims_df <- opt_sims %>% mutate(model = "1_optimal")
+policy_df <- opt_policy  %>% mutate(model = "1_optimal")
+
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
+```
+
+## RL-based solutions
+
+We will compare this optimal solution with results from the several different RL algorithms.
+Unlike the optimal solution, the RL methods never know the underlying growth function used by the simulation (or even if such a function actually exists).
+These methods merely seek to learn a strategy for setting harvest quotas which maximizes the cumulative reward the receive from the environment.
 At the heart of each deep RL algorithm is a neural network or a collection of neural networks.
@@
 ```{r}
-a2c_sims = env$simulate(a2c, reps = 500L)
+a2c_sims = env$simulate(a2c, reps = 100L)
 ```
@@
 ```{r}
-a2c_policy = env$policyfn(a2c, reps = 50L)
+a2c_policy = env$policyfn(a2c, reps = 5L)
+```
+
+```{r}
+sims_df <- a2c_sims %>% mutate(model = "A2C") %>% bind_rows(sims_df)
+policy_df <- a2c_policy  %>% mutate(model = "A2C")  %>% bind_rows(policy_df)
+
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
 ```
@@
 We train a second agent based on the TD3 algorithm.
-This illustrates some of the salient differences between the different algorithms used in reinforcement learning.
-Bear in mind that the best algorithm for any given environment will depend on the details of the environment as well
-as the length of training and the selection of hyper-parameters.
-This time, we will declare our choices for the hyper-parameters explicitly.
-Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
+Recall that the Twin Delayed Deep Deterministic Policy Gradient, commonly known as TD3 [@TD3], is a generalization of Deep Deterministic Policy Gradient, or DDPG, which extends the DQN algorithm @DQN to a continuous action space using the deterministic policy gradient.
+DDPG and DQN are also available in the stable baselines framework [@sb3].
+One of the most salient differences here is that A2C uses an Actor-Critic strategy that combines policy gradient and value iteration approaches, while TD3 is a policy gradient approach.
+
+We will start with the default hyperparameters, which correspond to those used in @TD3, where the algorithm was first introduced.
+
+```{r}
+td3_untuned = sb3$TD3('MlpPolicy', env, seed = seed)
+```
+
+Once again, after our agent is defined we are ready to train it:
+
+```{r }
+if(!file.exists("../python/cache/td3_untuned.zip")){
+
+  td3_untuned$learn(total_timesteps=300000L)
+  td3_untuned$save("../python/cache/td3_untuned")
+
+}
+```
+
+With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:
+
+```{r}
+# Simulate management under the trained agent
+td3_untuned = sb3$TD3$load("../python/cache/td3_untuned")
+td3_sims_untuned = env$simulate(td3_untuned, reps = 100L)
+td3_policy_untuned = env$policyfn(td3_untuned, reps = 5L)
+```
+
+```{r}
+sims_df <- td3_sims_untuned %>% mutate(model = "TD3_untuned") %>% bind_rows(sims_df)
+policy_df <- td3_policy_untuned  %>% mutate(model = "TD3_untuned")  %>% bind_rows(policy_df)
+
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
+```
+The untuned TD3 policy tends to overfish relative to the optimal solution, but is less variable than the A2C solution.
+While it is possible that with much more training, this TD3 algorithm could further improve, subsequent learning may be difficult.  
+Note that the reward under TD3 is nevertheless often within the margin of error of the optimal policy, which may limit future learning since further policy exploration is more likely to reduce rewards.
+Instead, we may be able to improve performance of TD3 by choosing different hyperparameters, which may allow it to imagine more possible policies or learn more effectively.
+
+## Tuned TD3
+
+Some of these hyper-parameters of the TD3 algorithm, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm. 
+Others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
 Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
-For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
+For simplicity, we show only training with specified hyper-parameters here, found by various tuning experimentation (see <https://github.com/boettiger-lab/conservation-agents>) and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
 
@@
                train_freq = 128L, 
-               n_episodes_rollout = -1, 
                gradient_steps = 128L,
@@
 
-Once again, after our agent is defined we are ready to train it:
+We train this new TD3 agent:
 
@@
 
-With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:
+With our trained agent, we can now simulate replicate scenarios or scan across states to determine the implicit policy function:
 
@@
 td3 = sb3$TD3$load("cache/td3")
-td3_sims = env$simulate(td3, reps = 500L)
-td3_policy = env$policyfn(td3, reps = 50L)
+td3_sims = env$simulate(td3, reps = 100L)
+td3_policy = env$policyfn(td3, reps = 5L)
 ```
 
-We compare the performance of these two RL agents trained using the algorithms A2C and TD3 respectively to the known optimal policy for this particular environment.  
-Recall that under the assumptions of the simple model used in the `fishing-v1` environment, we can determine the optimal harvest policy analytically if the model and parameters are known precisely [@Reed1979].
-The optimal strategy is a policy of 'constant escapement', designed to keep the remaining stock size (the population that 'escapes' fishing harvest) at the biomass corresponding to a maximum growth rate, i.e.
-at $B = K/2$ in this model.
-`gym_fishing` defines a collection of non-RL agents in the `models` submodule, including the a human agent that merely asks to enter their desired quota manually. 
-The `escapement` model implements the provably optimal constant escapement rule.
-A third model, `msy`, implements a policy based on "Maximum Sustainable Yield" policy [@Schaefer1954], which is actually more commonly used as a basis for management than constant escapement, despite only being optimal at the steady state under deterministic dynamics.
+```{r}
+sims_df <- td3_sims %>% mutate(model = "TD3") %>% bind_rows(sims_df)
+policy_df <- td3_policy  %>% mutate(model = "TD3")  %>% bind_rows(policy_df)
 
-```{r}
-# Simulate under the optimal solution (given the model)
-opt = gym_fishing$models$escapement(env)
-opt_sims = env$simulate(opt, reps = 500L)
-opt_policy = env$policyfn(opt)
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
 ```
 
-Having performed simulations under each method, we gather the results under each management strategy together into a single data frame for simulations (`sims_df`), the policy function (`policy_df`) and the cumulative reward, used to make the plots in the corresponding panels.
+The tuned TD3 algorithm, though trained for an identical 300,000 timesteps as the untined one, settles on a policy much closer to the optimal policy.
+The tuned TD3 agent tends to underfish on average, but only slightly, and avoids the occassional sharp drops in stock size seen when A2C occassionally sets too high a havest quota.
+The tuned solution is still not optimal, but it is very close.
+Moreover, the RL algorithm has been able to learn this nearly-optimal policy with no prior knowledge about the shape or nature of the population growth function used in the environment.
+
 We preserve a copy of each data frames in local `.csv` text files for maximum cross-platform compatibility.
@@
 ```{r}
-sims_df = bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
-
-policy_df = bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
-  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])
-
-gamma = 1 #discount
-reward_df = sims_df %>% 
-  group_by(rep, model) %>%
-  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
-  group_by(time, model) %>%
-  summarise(mean_reward = mean(cum_reward), 
-            sd = sd(cum_reward), .groups = "drop")
-
 write_csv(sims_df, "../manuscript/figs/sims_df.csv")
 write_csv(policy_df, "../manuscript/figs/policy_df.csv")
-write_csv(reward_df, "../manuscript/figs/reward_df.csv")
 ```
@@
 <!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
-Standard R methods are used to plot the results summarized across the replicates:
-
-```{r}
-ymin = function(x) last(x[(ntile(x, 20)==1)])
-ymax = function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims = 
-sims_df %>% 
-  group_by(time, model) %>% 
-  summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-```
-
-```{r}
-fig_policy = 
-  policy_df %>% ggplot(aes(state, action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) + 
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) 
-```
-
-```{r}
-
-fig_reward = reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
-```
-
-## Manuscript Figure 1
-
-<!-- Not sure if it's helpful to duplicate figures from the paper in the appendix...-->
-
-```{r fig1}
-fig_sims / ( fig_policy + fig_reward)
-```
 
@@
             "gradient_steps"= 1000L,
-            "n_episodes_rollout"= -1L,
             "action_noise"= action_noise,
@@
 
-gamma = 1 #discount
-reward_df = sims_df %>% 
-  group_by(rep, model) %>%
-  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
-  group_by(time, model) %>%
-  summarise(mean_reward = mean(cum_reward), 
-            sd = sd(cum_reward), .groups = "drop")
-
 ```
@@
 write_csv(policy_df, "../manuscript/figs/tipping_policy_df.csv")
-write_csv(reward_df, "../manuscript/figs/tipping_reward_df.csv")
-```
-
-<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
-
-```{r}
-# ensemble statistics
-ymin = function(x) last(x[(ntile(x, 20)==1)])
-ymax = function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims = 
-  sims_df %>%
-   group_by(time, model) %>% 
-   summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-```
-
-```{r}
-fig_policy = 
-  policy_df %>% ggplot(aes(state, action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) 
-```
-
-```{r}
-
-fig_reward = reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
 ```
@@
 ```{r fig3}
-fig_sims / ( fig_policy + fig_reward)
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_policy(sims_df))
 ```

### 2021-05-27 manuscript/manuscript.Rmd
Getting closer on rewards paragraph

@@
 There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@CITE].
-For ecological systems, which are typically very complex and difficult to fully understand, it seems that bad reward design is quite likely.
-To avoid these undesirable outcomes, RL researchers often *shape* the reward functions so that the agent can more easily learn a good-performing policy [@cite].
-This approach is particularly common when rewards are *sparse*, such as in a game of Chess where the objective is to win and there are no true immediate rewards.
-In this case, one might shape the reward function so that the agent is awarded for capturing the opponent's pieces, but this could easily lead to a policy where the agent focuses on capturing pieces instead of winning the game.
-Since benefits in conservation typically come long after their corresponding actions, sparse reward.
+Thus, for ecological systems, which are typically difficult to control and fully understand, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
+To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn a certain behavior[@cite].
+This approach is particularly common when rewards are *sparse*, which occurs when there are few state-action pairs that have strong feedback signals.
+Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still produce unintended agent behavior [@CITE]. 
+Given the nature of many conservation problems, where benefits rarely come immediately and thus the reward structures are sparse, RL agents will likely have difficulty learning anything, or, in the case that researchers shape the rewards to aid learning, the agents may learn policies that are undesirable.
 Moreover, specifying a reward function for conservation brings up various ethical concerns -- e.g.
 who gets to specify the reward function and will the values of the greater community be represented in the reward function.
-
-<!--
-One recently proposed framework to overcome challenges with explicit reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
-There is a definite need for exploration of the ethics of using DRL for conservation.
--->
-<!-- reason 4/additional reasons, like sparse rewards -->
-<!-- ML: I don't know if this paragraph is adding that much. Since we talk about sparsity in reward design, i think we can add a sentence or 2 there that conveys the message here.-->
-Ultimately, we anticipate that getting deep RL to work well on fully detailed conservation decision-making problems will be challenging.
-Beyond typically having large state-action spaces for agents to explore, conservation problems tend to have sparse reward structures.
-E.g. rarely does an action in conservation lead to immediate rewards; instead, benefits are typically observed after actions have been performed.
-Sparse reward structures are challenging for DRL agents, because of the credit assignment problem [cite].
-To solve sparse reward environments, many researchers have employed customized algorithms to achieve better performance than standard DRL methods [cite Montezuma].
-For conservation, we anticipate that there will likely be a similar pattern, wherein bespoke RL methods will attain better performance than general methods.
 

### 2021-05-27 manuscript/manuscript.Rmd
Getting closer

@@
 Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
-<!-- Marcus is working on tightening up these next few paragraphs, precisely which issues actually get highlighted TBD-->
 In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: generalizability, interpretability, and reward design.
@@
 For instance, training an agent on an environment that demands learning an assortment of skills is typically much more difficult than training an agent to learn one specific skill [@CITE].
-And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent will likely fail to generalize its learned behavior to this new situation [@CITE].
+And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent can easily fail to generalize its learned behavior to this new situation [@CITE].
 These failures are all due to the brittleness of deep neural networks. 
@@
 -->
-Compounded with their difficulties to generalize, deep RL agents also suffer from an inability to be fully interpreted [@cite].
-Deep neural networks are commonly referred to as black box models, since it is not currently possible to explain why a deep neural network produces a specific output from a given input.
-In the context of deep RL, this means that we cannot explain why an agents makes a decision besides that the agent is attempting to maximize cumulative rewards.
-The lack of explainability dovetails dangerously with generalizability concerns. 
-Because we are not able to interpret deep RL agents, we are not able to anticipate how an agent will behave when in an unfamiliar situation.
-Despite limited progress to make neural networks more explainable, interpretability will likely be a concern that limits adoption.
+Deep RL agents also suffer from an inability to be interpreted [@cite].
+Deep neural networks are commonly referred to as black box models, since it is not possible to fully explain why a deep neural network produces a specific output from a given input [@cite].
+For deep RL, this means that we cannot provide a meaningful explanation for why any agent selects an action.
+Compounded with the previously mentioned generalizability issues, the lack of explainability introduces a lot of safety concerns when attempting to apply these methods to the real world.
+When going from a simulator to the real world, the real world may introduce novel situations that a simulator-trained agent is not familiar with, and since we can't anticipate how an agent will respond to new situations, there is an inherent risk in letting RL agents make decisions in the real world.
 
@@
 There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@CITE].
-Thus, for ecological systems, which are typically difficult to control and fully understand, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
-To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn a certain behavior[@cite].
+Thus, for ecological systems, which are typically difficult to control to begin with, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
+To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn certain behaviors[@cite].
 This approach is particularly common when rewards are *sparse*, which occurs when there are few state-action pairs that have strong feedback signals.
-Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still produce unintended agent behavior [@CITE]. 
+Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still backfire and produce unintended agent behavior [@CITE].
 Given the nature of many conservation problems, where benefits rarely come immediately and thus the reward structures are sparse, RL agents will likely have difficulty learning anything, or, in the case that researchers shape the rewards to aid learning, the agents may learn policies that are undesirable.
-Moreover, specifying a reward function for conservation brings up various ethical concerns -- e.g.
-who gets to specify the reward function and will the values of the greater community be represented in the reward function.
+Moreover, specifying a reward function for conservation brings up various ethical concerns.
+e.g., Who gets to specify the reward function, and will the values of the greater community be represented in the reward function?
 

### 2021-05-27 manuscript/manuscript.Rmd
Passing over discussion again

@@
 Successfully applying RL to complex ecological problems is no easy task.
-The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance is another task.
+The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance can often be more challenging than expected.
 As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
@@
 These failures are all due to the brittleness of deep neural networks. 
-Across deep learning, neural networks tend to perform exceptionally well on narrowly specified problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@CITE].
-The ability to adapt to new situations is critical for any decision-making problem in the real world, so the current lack of this ability in deep RL may limit deep RL's effectiveness to conservation issues.
-
-<!-- 
-A significant current challenge in deep RL is the problem of instability.
-For a range of reasons, including high variance of returns and the exploration vs.
-exploitation dilemma, deep RL algorithms tend to have unstable performance on environments with large state-action spaces [@drlthatmatters].
-For example, @drlthatmatters show that initializing the same agent with different random seeds can lead to significant differences in agent performance.
-Such unstable behavior can be avoided by finding a robust set of hyperparameters or using algorithms that are designed to have greater reliability [@qprop], thus researchers should consider stability as well as performance during hyperparameter tuning and algorithm selection.
--->
+Across deep learning, neural networks tend to perform exceptionally well on narrowly defined problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@CITE].
+The ability to adapt to new situations is critical for any decision-making problem in the real world, so the current lack of this ability in deep RL may limit deep RL's effectiveness to conservation problems.
 
@@
 <!-- reason 2: interpretability -->
-<!-- ML: Re our OU talk, might be worth saying that this is a barrier to adoption?
-If we want to ultimately use these RL agents to make important decisions, it would be hard to swallow us blindly using them without having any sense to what is going on.
--->
-Deep RL agents also suffer from an inability to be interpreted [@cite].
+Deep RL agents are also very difficult to interpret [@cite].
 Deep neural networks are commonly referred to as black box models, since it is not possible to fully explain why a deep neural network produces a specific output from a given input [@cite].
@@
 <!-- reason 3: reward design -->
-<!-- ML: I'll take a pass at this paragraph in a new branch. I think it would be useful to put terms on stuff in here, like reward shaping and reward hacking which we are talking around. -->
-A fundamental issue for all applied optimization problems is using the right objective.
+And lastly, a fundamental issue for all applied optimization problems is using the right objective.
 There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@CITE].
-Thus, for ecological systems, which are typically difficult to control to begin with, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
+Thus, for ecological systems, which are typically quite difficult to control, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
 To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn certain behaviors[@cite].
 This approach is particularly common when rewards are *sparse*, which occurs when there are few state-action pairs that have strong feedback signals.
-Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still backfire and produce unintended agent behavior [@CITE].
+Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still backfire by producing unintended agent behavior [@CITE].
 Given the nature of many conservation problems, where benefits rarely come immediately and thus the reward structures are sparse, RL agents will likely have difficulty learning anything, or, in the case that researchers shape the rewards to aid learning, the agents may learn policies that are undesirable.
 Moreover, specifying a reward function for conservation brings up various ethical concerns.
-e.g., Who gets to specify the reward function, and will the values of the greater community be represented in the reward function?
+E.g., who gets to specify the reward function, and will the values of the greater community be represented in the reward function?
 
@@
 Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
-Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [brockman2016].
+Leading developers of new deep RL algorithms benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [brockman2016].
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].

### 2021-05-27 manuscript/manuscript.Rmd
Adding citations

@@
 <!-- ML: pivoting this to generalizability but still rewrote past paragraph which might fit in the appendix somewhere-->
-Currently, deep RL agents struggle to generalize their skills [@generalization].
+Currently, deep RL agents struggle to generalize their skills [@pmlr-v97-cobbe19a].
 This inability can affect performance during training and at evaluation. 
-For instance, training an agent on an environment that demands learning an assortment of skills is typically much more difficult than training an agent to learn one specific skill [@CITE].
-And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent can easily fail to generalize its learned behavior to this new situation [@CITE].
+For instance, training an agent on an environment that demands learning an assortment of skills is typically much more difficult than training an agent to learn one specific skill [@teh_distral_2017].
+And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent can easily fail to generalize its learned behavior to this new situation [@pmlr-v97-cobbe19a].
 These failures are all due to the brittleness of deep neural networks. 
-Across deep learning, neural networks tend to perform exceptionally well on narrowly defined problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@CITE].
+Across deep learning, neural networks tend to perform exceptionally well on narrowly defined problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@cogswell_reducing_2016].
 The ability to adapt to new situations is critical for any decision-making problem in the real world, so the current lack of this ability in deep RL may limit deep RL's effectiveness to conservation problems.
@@
 <!-- reason 2: interpretability -->
-Deep RL agents are also very difficult to interpret [@cite].
-Deep neural networks are commonly referred to as black box models, since it is not possible to fully explain why a deep neural network produces a specific output from a given input [@cite].
+Deep RL agents are also difficult to interpret [@heuillet_explainability_2021].
+Deep neural networks are commonly referred to as black box models, since it is challenging to fully explain why a deep neural network produces a specific output from a given input [@castelvecchi_can_2016].
 For deep RL, this means that we cannot provide a meaningful explanation for why any agent selects an action.
@@
 And lastly, a fundamental issue for all applied optimization problems is using the right objective.
-There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@CITE].
+There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@hadfield-menell_inverse_2020].
 Thus, for ecological systems, which are typically quite difficult to control, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
-To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn certain behaviors[@cite].
+To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn certain behaviors [@reward_shaping].
 This approach is particularly common when rewards are *sparse*, which occurs when there are few state-action pairs that have strong feedback signals.
-Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still backfire by producing unintended agent behavior [@CITE].
+Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing, but reward shaping can still backfire by producing unintended agent behavior [@hadfield-menell_inverse_2020].
 Given the nature of many conservation problems, where benefits rarely come immediately and thus the reward structures are sparse, RL agents will likely have difficulty learning anything, or, in the case that researchers shape the rewards to aid learning, the agents may learn policies that are undesirable.

### 2021-05-27 manuscript/manuscript.Rmd
small changes

@@
 Deep neural networks are commonly referred to as black box models, since it is challenging to fully explain why a deep neural network produces a specific output from a given input [@castelvecchi_can_2016].
-For deep RL, this means that we cannot provide a meaningful explanation for why any agent selects an action.
+For deep RL, this means that we can rarely provide a meaningful explanation for why any agent selects an action.
 Compounded with the previously mentioned generalizability issues, the lack of explainability introduces a lot of safety concerns when attempting to apply these methods to the real world.
-When going from a simulator to the real world, the real world may introduce novel situations that a simulator-trained agent is not familiar with, and since we can't anticipate how an agent will respond to new situations, there is an inherent risk in letting RL agents make decisions in the real world.
+When going from a simulator to the real world, the real world may introduce novel situations that a simulator-trained agent is not familiar with, and since we can not anticipate how an agent will respond to new situations, there is an inherent risk in letting RL agents make decisions in the real world.
 

### 2021-05-27 appendices/appendix-B-examples.Rmd
WIP

@@
 # Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
-Sys.setenv("CUDA_VISIBLE_DEVICES"="")
+#Sys.setenv("CUDA_VISIBLE_DEVICES"="")
 set.seed(seed)
@@
 ```{r}
-sims_df <- opt_sims %>% mutate(model = "1_optimal")
-policy_df <- opt_policy  %>% mutate(model = "1_optimal")
+algos <- c("Optimal", "A2C", "TD3", "TD3_tuned")
+sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
+policy_df <- opt_policy  %>% mutate(model = factor("Optimal", algos))
 
@@
 ```{r}
-if(!file.exists("cache/a2c.zip")) {
+if(!file.exists("../python/cache/a2c.zip")) {
 
@@
   # Save our trained agent for future use
-  a2c$save("cache/a2c")
+  a2c$save("../python/cache/a2c")
 }
@@
 # Simulate management under the trained agent
-a2c = sb3$A2C$load("cache/a2c")
+a2c = sb3$A2C$load("../python/cache/a2c")
 ```
@@
 ```{r}
-sims_df <- a2c_sims %>% mutate(model = "A2C") %>% bind_rows(sims_df)
-policy_df <- a2c_policy  %>% mutate(model = "A2C")  %>% bind_rows(policy_df)
+sims_df <- a2c_sims %>% mutate(model = factor("A2C", algos)) %>% bind_rows(sims_df)
+policy_df <- a2c_policy  %>% mutate(model = factor("A2C", algos))  %>% bind_rows(policy_df)
 
@@
 ```{r}
-sims_df <- td3_sims_untuned %>% mutate(model = "TD3_untuned") %>% bind_rows(sims_df)
-policy_df <- td3_policy_untuned  %>% mutate(model = "TD3_untuned")  %>% bind_rows(policy_df)
+sims_df <- td3_sims_untuned %>% mutate(model = factor("TD3", algos)) %>% bind_rows(sims_df)
+policy_df <- td3_policy_untuned  %>% mutate(model = factor("TD3", algos))  %>% bind_rows(policy_df)
 
@@
 ```{r }
-if(!file.exists("cache/td3.zip")){
+if(!file.exists("../python/cache/td3_tuned.zip")){
 
   td3$learn(total_timesteps=300000L)
-  td3$save("cache/td3")
+  td3$save("../python/cache/td3_tuned")
 
@@
 # Simulate management under the trained agent
-td3 = sb3$TD3$load("cache/td3")
+td3 = sb3$TD3$load("../python/cache/td3_tuned")
 td3_sims = env$simulate(td3, reps = 100L)
@@
 ```{r}
-sims_df <- td3_sims %>% mutate(model = "TD3") %>% bind_rows(sims_df)
-policy_df <- td3_policy  %>% mutate(model = "TD3")  %>% bind_rows(policy_df)
+sims_df <- td3_sims %>% mutate(model = factor("TD3_tuned", algos)) %>% bind_rows(sims_df)
+policy_df <- td3_policy  %>% mutate(model = factor("TD3_tuned", algos))  %>% bind_rows(policy_df)
 
@@
 ```{r}
+# Python users see python/conservation_TD3.py
 env = gym$make("conservation-v6")
-```
 
-```{r}
 noise_std = 0.4805935357322933
-
 OU = sb3$common$noise$OrnsteinUhlenbeckActionNoise
@@
 
+As before, we train the agent unless a cached version of the trained agent is already available:
+
 ```{r}
-if(!file.exists("cache/td3-conservation.zip")){
+if(!file.exists("../python/cache/td3-conservation.zip")){
 
   model$learn(total_timesteps=3000000L)
-  model$save("cache/td3-conservation")
+  model$save("../python/cache/td3-conservation")
 
@@
 
-model = sb3$TD3$load("cache/td3-conservation")
+model = sb3$TD3$load("../python/cache/td3-conservation")
 TD3_sims = env$simulate(model, reps = 100L)
@@
 ```
+
+```{r}
+```

### 2021-05-27 manuscript/manuscript.Rmd
WIP

--- 2021-05-27 manuscript/manuscript.Rmd
@@
 <!-- Environmental problems, ML -->
-Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@Worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
+Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
 But will ML approaches really help us tackle a changing planet?
@@
 RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
-Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
+Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.
 Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
+<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
@@
 <!-- Conceptual picture: (A) an "environment", and (B) an "agent"
+
+What do we mean by an "environment"?  
+  - Transition function
+  - "reward"
+  - examples
+
+What do we mean by an "agent"?
+
+  Model-based vs Model free.  We focus on Model free agents which use Neural networks, a subset of deep RL
+
   Components of an agent: 
@@
     - High-level overview of an RL algorithm components (model/model-free, on/off policy, stochastic/deterministic, etc).
+
   conceptual explanation of 'model-free', what it means to have a "policy" without modeling/predicting transitions
@@
 -->
+
+<!-- ML: need to change some of the language here to be consistent with POMDP set up later-->
 
@@
 An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
-In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
-States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
+<!-- ML: what are classical approaches below here? -->
+In contrast to classical approaches, there are few restrictions on what comprises a state or action. 
+The state may be continuous or discrete, completely or partially observed, single or multidimensional.
+For instance, we might represent an ecological community of 10 interacting species with a 10-dimensional continuously valued state space signifying the population density of each species.
+The action space can be similarly complex, since an action can be continuous or discrete, single or multidimensional.
+For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at the different species over a time step.
 The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
@@
 
+**Survey of RL Algorithms**
+
 <!--
@@
 Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
-In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
+In table **, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
 In the appendix, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
+
+<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->
+
+<!-- Can we add citations to each please -->
+<!--
+
+Should we mention:
+
+- Temporal Differencing, RIENFORCE
+- SARSA, 
+- Double DQN?  
+Can we add HER (since delayed reward case may be particularly interesting)
+Should we do anything to better reflect the 'family tree' here? (e.g. DQN -> DDPG -> TD3)
+
+-->
 
@@
 
-We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
+We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and challenges there-in.
 
@@
 txtcolor <- "#586e75"
+
+source("../R/plotting-helpers.R")
 ```
@@
 The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
-Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
+Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2018], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
 Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
@@
 The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
-Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2016] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
+Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
 
-For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
+For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1974] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
 In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
@@
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
-We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
+We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 1].
 Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
 
-<!-- ML: I'm have some uneasiness about this figure. Is A2C tuned? I think we'd probably want to do tuning and training with a set seed so that everything is reproducible?  -->
 ```{r message=FALSE, include=FALSE}
-sims_df <- read_csv( "figs/sims_df.csv")
-policy_df <- read_csv("figs/policy_df.csv")
-reward_df <- read_csv("figs/reward_df.csv")
+sims_df <- read_csv( "figs/sims_df.csv") %>% filter(model %in% c("Optimal", "TD3_tuned"))
+policy_df <- read_csv("figs/policy_df.csv") %>% filter(model %in% c("Optimal", "TD3_tuned"))
 
-ymin <- function(x) last(x[(ntile(x, 20)==1)])
-ymax <- function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims <- 
-sims_df %>% 
-  group_by(time, model) %>% 
-  summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-
-fig_policy <- 
-  policy_df %>% ggplot(aes(state, state - action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) + 
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) + 
-  ylab("escapement")
-
-fig_reward <- reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
 ```
@@
 Mean reward under A2C is lower, though the confidence interval still overlaps the mean reward of optimal management."}
-fig_sims / ( fig_policy + fig_reward)
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
 ```
@@
 We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
-This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012] lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
+This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
 On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
 For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
-Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].
+Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2014].
 We also enumerate the possible actions which a manager may take in response to environmental degradation.
@@
 
-A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix B, fig S1, S2) further highlights the difference between these two approaches.
+A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix, fig S1, S2) further highlights the difference between these two approaches.
 Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
@@
 The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
-Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
-<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote.  CB: agree, revised above-->
+Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
+The use of a recurrent neural network (RNN) architecture instead of the multi-layer perceptron (MLP) used here would open the door for the policy to reflect the non-autonomous dynamics more directly. 
+<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote-->
 
@@
 The science of ecological management and quantitative decision-making has a long history [e.g.
-@Schaefer1954; @Walters1978] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
-However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@Hastings2012].
+@Schaefer1954; @Walters1974] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@HastingsEncylopedia].
 Complexity comes not only from the ecological processes but also the available actions.
@@
 The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
-Model-free approaches have proven successful at learning effective policies even when the underlying model may be difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
+<!-- ML: I feel like the remaining part of this dont full drive home the benefit of model-free rl, like its able to learn good performing policies even in the case when a model is very difficult or impossible to learn.
+-->
+Alternative models can easily be implemented in simulations, allowing an agent to train against both contingencies.
+A successful agent will learn a policy that works well in both scenarios.
+This is analogous to classical approaches for integrating over parameter or model uncertainty [e.g. @Walters1981; @Ludwig1982],
 
@@
 <!-- ML: I feel like this paragraph could be condensed with the one above -->
-<!-- CB: I think "model free is great" and "real problems are harder than these examples" are two distinct messages though and at least one would get lost in mashing them together.
-Perhaps the message doesn't come across in the current version.
-Does look like this paragraph is similar in overall message to the next one though, maybe could be combined with that?
--->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 In contrast, the process of training an RL algorithm on a more complex problem is no different than the simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
-Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
+Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @firemodel], disease [@disease], marine ecosystems [@ecopath], or global climate change [@gcm].
 
@@
 Successfully applying RL to complex ecological problems is no easy task.
-The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance can often be more challenging than expected.
+The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance is another task.
 As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
@@
 Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
-In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: generalizability, interpretability, and reward design.
+In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: instability, interpretability, and reward design.
 
-<!-- reason 1: generalizability-->
-<!-- ML: pivoting this to generalizability but still rewrote past paragraph which might fit in the appendix somewhere-->
-Currently, deep RL agents struggle to generalize their skills [@generalization].
-This inability can affect performance during training and at evaluation. 
-For instance, training an agent on an environment that demands learning an assortment of skills is typically much more difficult than training an agent to learn one specific skill [@CITE].
-And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent can easily fail to generalize its learned behavior to this new situation [@CITE].
-These failures are all due to the brittleness of deep neural networks. 
-Across deep learning, neural networks tend to perform exceptionally well on narrowly defined problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@CITE].
-The ability to adapt to new situations is critical for any decision-making problem in the real world, so the current lack of this ability in deep RL may limit deep RL's effectiveness to conservation problems.
+<!-- reason 1: instability-->
+<!-- ML: I think this needs to be tightened up a bit-->
+One leading challenge for deep RL today is the problem of instability.
+RL agents trained on narrowly defined environments may perform poorly even when the environment is altered in seemingly trivial way, such as a re-scaling of parameters.
+To avoid this, researchers should seek to both train and evaluate agents against a range of environments representing different parameters or processes dynamics.
+Stable performance can be challenging even when using the same environment.
+For a range of reasons, including high variance [@cite] and the exploration vs.
+exploitation trade-off [@cite], model-free and model-based algorithms tend to have unstable performance on environments with non-tabular state-action spaces [cite RL that matters].
+For instance, it is not uncommon to observe vastly different behavior when an agent is initialized with different random seeds [cite ibid].
+Such unstable behavior can often be avoided by alternate selection of hyperparameters, so researchers should consider stability as well as performance during hyperparameter selection.
+The choice of RL algorithm, design of the reward function, and duration of training can also influence stability.
 
@@
 <!-- if we're really recommending 'testing' for stability, maybe we need to demo that in the appendix-->
+
 <!-- reason 2: interpretability -->
-Deep RL agents are also very difficult to interpret [@cite].
-Deep neural networks are commonly referred to as black box models, since it is not possible to fully explain why a deep neural network produces a specific output from a given input [@cite].
-For deep RL, this means that we cannot provide a meaningful explanation for why any agent selects an action.
-Compounded with the previously mentioned generalizability issues, the lack of explainability introduces a lot of safety concerns when attempting to apply these methods to the real world.
-When going from a simulator to the real world, the real world may introduce novel situations that a simulator-trained agent is not familiar with, and since we can't anticipate how an agent will respond to new situations, there is an inherent risk in letting RL agents make decisions in the real world.
+<!-- ML: Re our OU talk, might be worth saying that this is a barrier to adoption?
+If we want to ultimately use these RL agents to make important decisions, it would be hard to swallow us blindly using them without having any sense to what is going on.
+-->
+Interpretability is a common challenge at the adoption of most ML methods [@cite]. 
+Although there have been some attempts to increase the transparency of RL agents that use neural network as function approximators [cite], it is still very difficult to explain the behavior of RL-trained agents.
+As our examples have illustrated, it may be possible to gain intuition on an agent's behavior by visualizing the action selected against each possible state, though this becomes more difficult in large problems.
+Another challenge is that Deep RL algorithms are very sample inefficient, all though progress is being made to improve this [@cite].
+<!-- ML: I think we can cut this sentence, mentionned it in passing earlier-->
+-->
 
 <!-- reason 3: reward design -->
-And lastly, a fundamental issue for all applied optimization problems is using the right objective.
-There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@CITE].
-Thus, for ecological systems, which are typically quite difficult to control, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
-To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn certain behaviors[@cite].
-This approach is particularly common when rewards are *sparse*, which occurs when there are few state-action pairs that have strong feedback signals.
-Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing [@CITE], but reward shaping can still backfire by producing unintended agent behavior [@CITE].
-Given the nature of many conservation problems, where benefits rarely come immediately and thus the reward structures are sparse, RL agents will likely have difficulty learning anything, or, in the case that researchers shape the rewards to aid learning, the agents may learn policies that are undesirable.
-Moreover, specifying a reward function for conservation brings up various ethical concerns.
-E.g., who gets to specify the reward function, and will the values of the greater community be represented in the reward function?
+<!-- ML: I'll take a pass at this paragraph in a new branch. I think it would be useful to put terms on stuff in here, like reward shaping and reward hacking which we are talking around. -->
+An issue that is central to any applied optimization problem is using the right objective.
+There have been many examples in which explicit reward design goes very wrong [cite].
+And for ecological systems, which are often very complex and impossible to fully observe, it seems that bad reward design is quite likely.
+Moreover, specifying a reward function for conservation brings up various ethical concerns -- e.g.
+who gets to specify the reward function and will the values of the greater community be represented in the reward function.
+RL researchers often introduce modified reward functions during training that help an agent perform better on the original objective [@cite].
+This approach is particularly common when rewards are sparse, such as in a game of Chess where the only objective that counts is whether or not you win, but during training, an RL algorithm may reward the agent for intermediate steps such as capturing a piece.
+
+<!--
+One recently proposed framework to overcome challenges with explicit reward design is called AI alignment, wherein the agent observes humans, infers the preferences of humans and then acts as to satisfy human preferences.
+There is a definite need for exploration of the ethics of using DRL for conservation.
+-->
+<!-- reason 4/additional reasons, like sparse rewards -->
+<!-- ML: I don't know if this paragraph is adding that much. Since we talk about sparsity in reward design, i think we can add a sentence or 2 there that conveys the message here.-->
+Ultimately, we anticipate that getting deep RL to work well on fully detailed conservation decision-making problems will be challenging.
+Beyond typically having large state-action spaces for agents to explore, conservation problems tend to have sparse reward structures.
+E.g. rarely does an action in conservation lead to immediate rewards; instead, benefits are typically observed after actions have been performed.
+Sparse reward structures are challenging for DRL agents, because of the credit assignment problem [cite].
+To solve sparse reward environments, many researchers have employed customized algorithms to achieve better performance than standard DRL methods [cite Montezuma].
+For conservation, we anticipate that there will likely be a similar pattern, wherein bespoke RL methods will attain better performance than general methods.
 
@@
 Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
-Leading developers of new deep RL algorithms benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [brockman2016].
+Leading developers of new algorithms for deep RL benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [@gym], and performance can easily be compared against alternatives on public leaderboards [@gym].
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].

### 2021-05-27 python/conservation_TD3.py

@@ -12,32 +12,25 @@ env = gym.make("conservation-v6")
 
 noise_std = 0.4805935357322933
 OU = sb3.common.noise.OrnsteinUhlenbeckActionNoise
-action_noise = OU(mean = 0,  sigma = noise_std)
-
-model = sb3.TD3('MlpPolicy', 
-            env, 
-            verbose = 0, 
-            seed = 42,
-            gamma= 0.995,
-            learning_rate =  8.315382409902049e-05,
-            batch_size = 512,
-            buffer_size = 10000,
-            train_freq = 1000,
-            gradient_steps = 1000,
-            n_episodes_rollout = -1,
-            action_noise = action_noise,
-            policy_kwargs = dict("net_arch"= c(64,64)))
+action_noise = OU(mean=np.zeros(1), sigma=noise_std * np.ones(1))
+model = sb3.TD3(
+    "MlpPolicy",
+    env,
+    verbose=0,
+    seed=42,
+    gamma=0.995,
+    learning_rate=8.315382409902049e-05,
+    batch_size=512,
+    buffer_size=10000,
+    train_freq=1000,
+    gradient_steps=1000,
+    action_noise=action_noise,
+    policy_kwargs={"net_arch": [64, 64]},
+)
 
 model.learn(total_timesteps=3000000)
 model.save("cache/td3-conservation")
-  
-model = sb3.TD3.load("cache/td3-conservation")
-TD3_sims = env.simulate(model, reps = 100)
-TD3_policy = env.policyfn(model, reps = 10)
 
-# Simulate under the steady-state solution (given the model)
-K = 1.5
-alpha = 0.001
-opt = gym_conservation.models.fixed_action(env, fixed_action = alpha * 100 * 2 * K )
-opt_sims = env.simulate(opt, reps = 100L)
-opt_policy = env.policyfn(opt)
+model = sb3.TD3.load("cache/td3-conservation")
+TD3_sims = env.simulate(model, reps=100)
+TD3_policy = env.policyfn(model, reps=10)

### 2021-05-27 python/fishing_A2C_untuned.py

@@ -1,4 +1,4 @@
-# Fishing using A2C
+# Fishing using (untuned) A2C
 
 import numpy as np
 import stable_baselines3 as sb3
@@ -24,13 +24,13 @@ a2c.save("cache/a2c")
 a2c = sb3.A2C.load("cache/a2c")
 
 
-# Example simulation of the agent
+# Example manual simulation of the agent
 state = env.get_state(0.75)
 for i in range(1, 10):
     action, null = a2c.predict(state)
     state, reward, done, info = env.step(action)
 
-## Or more consisely:
+## Or more consisely using built-in method:
 a2c_sims = env.simulate(a2c, reps=500)
 
 # Estimate the 'Policy function'

### 2021-05-27 python/fishing_TD3_tuned.py

@@ -1,3 +1,5 @@
+# Fishing using tuned TD3
+
 import numpy as np
 import stable_baselines3 as sb3
 import gym
@@ -9,9 +11,8 @@ np.random.seed(seed)
 # Create a fishing environment
 env = gym.make("fishing-v1", sigma=0.1)
 
-
+# Define the agent
 policy_kwargs = dict(net_arch=[400, 300])  # "big"
-# non-episodic action noise:
 noise_std = 0.6656948079225263
 action_noise = sb3.common.noise.NormalActionNoise(mean=0, sigma=noise_std)
 
@@ -25,16 +26,21 @@ td3 = sb3.TD3(
     batch_size=128,
     buffer_size=10000,
     train_freq=128,
-    n_episodes_rollout=-1,
     gradient_steps=128,
     action_noise=action_noise,
 )
 
+# Train the agent
 td3.learn(total_timesteps=300000)
-td3.save("cache/td3")
+td3.save("cache/td3_tuned")
 
 
 # Simulate management under the trained agent
-td3 = sb3.TD3.load("cache/td3")
-td3_sims = env.simulate(td3, reps=500)
-td3_policy = env.policyfn(td3, reps=50)
+td3 = sb3.TD3.load("cache/td3_tuned")
+td3_sims = env.simulate(td3, reps=100)
+td3_policy = env.policyfn(td3, reps=5)
+
+# Example of built-in plotting methods
+env.plot(td3_sims, "fishing_td3_tuned.png")
+env.plot_policy(td3_policy, "fishing_td3_tuned_policy.png")
+

### 2021-05-27 python/fishing_TD3_untuned.py

@@ -1,3 +1,5 @@
+# Fishing using untuned TD3
+
 import numpy as np
 import stable_baselines3 as sb3
 import gym
@@ -8,22 +10,18 @@ np.random.seed(seed)
 
 # Create a fishing environment
 env = gym.make("fishing-v1", sigma=0.1)
+# Create the agent
+td3 = sb3.TD3("MlpPolicy", env, seed=seed)
 
-td3 = sb3.TD3(
-    "MlpPolicy",
-    env,
-    seed=seed)
-
+# Train the agent
 td3.learn(total_timesteps=300000)
 td3.save("cache/td3_untuned")
 
-
 # Simulate management under the trained agent
 td3 = sb3.TD3.load("cache/td3_untuned")
-td3_sims = env.simulate(td3, reps=500)
-td3_policy = env.policyfn(td3, reps=50)
+td3_sims = env.simulate(td3, reps=100)
+td3_policy = env.policyfn(td3, reps=5)
 
-
-env.plot(sims, "fishing_td3_untuned.png")
-env.plot_policy(policy, "fishing_td3_untuned.png")
-
+# Example of built-in plotting methods
+env.plot(td3_sims, "fishing_td3_untuned.png")
+env.plot_policy(td3_policy, "fishing_td3_untuned_policy.png")

### 2021-05-28 appendices/appendix-A-definitions.Rmd
Trying to get this up to par

@@
 keywords:
-  - blah
-  - blee
-  - bloo
-  - these are optional and can be removed
-bibliography: references.bib
+  - Decision Theory
+  - Conservation
+  - Artificial Intelligence
+  - Machine Learning
+  - Tipping Points
+  - Fisheries
+bibliography: ../manuscript/references.bib
 
@@
 
-<!-- Not tightened up yet to make sense in the appendix, just migrating more nuts and bolts RL stuff
-     from manuscript here for sake of word count
--->
-
-<!-- These 2 paragraphs might make it back to the manuscript but going to keep them here for now -->
-Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
-Generally, model-free agents are either learning a policy function, a value function or both.
-For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
-If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
-The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
-*Policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and TD-learning [@suttonbarto], fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@CITE].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
-Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
-We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.
+In this appendix, we will go into further detail on how deep RL algorithms work.
+The aim here is not to give full treatment of all deep RL methods -- @suttonbarto is an excellent resource for theoretical underpinnings while the white papers behind the algorithms go into greater detail on modern approaches --; instead, this section is meant to serve as an abbreviated background view on common, model-free deep RL techniques.
+We hope to convey the sense of what the different algorithm classes try to achieve and will point to specific resources for readers looking for a more complete presentation.
+We will discuss policy gradients, value-based and actor-critic critic methods which were briefly introduced in the manuscript. 
+These sections presume some familiarity with gradient descent optimization -- see @ruder_overview_2017 for a thorough treatment.
+Throughout these sections, we'll consider the environment to be an MDP which makes for a cleaner presentation here and is also consistent with the referenced sources.
 
@@
 The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
-There are both gradient-free and gradient-based policy search methods that have been successful in practice [@CITE; @CITE]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
+There are both gradient-free and gradient-based policy search methods that have been successful in practice [@salimans_evolution_2017]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
 In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters.
@@
 
-Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, then $\nabla J(\pi_\theta)$ can be estimated as:
+Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, we can employ the policy gradient theorem -- see @suttonbarto for the proof of the policy gradient theorem -- to estimate $\nabla J(\pi_\theta)$ as:
 
@@
 
-See the appendix for how this is derived. 
 The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return.
-In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@CITE].
-To avoid this issue, most policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@CITE GAE].
+In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@suttonbarto].
+To avoid this issue, many policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@suttonbarto].
+
+Putting the 
 A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
 
-## Value Learning
+## Value-based Methods 
 Another way to approach the RL problem is by estimating a value function, and then using the value function to retrieve a policy. 
@@
 Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
-For example, the state-action-value or Q function finds the expected return from a state and action pair under a policy, $\pi$,
+For example, the state-action-value or Q function finds the expected return from a state-action pair under a policy, $\pi$,
 
@@
 
-The general Q-learning motivation is that if we know the Q function under certain conditions -- e.g.
-that the state-action space is discrete and that the given policy visits all state-action pairs repeatedly [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-In the case that the state-action space can be tractably represented in a table, one could employ a  algorithm, like temporal-difference control -- see Sutton Ch.
-6 for details -- which approximates the Q value for every state-action pair [@suttonbarto].
-Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; for this case, one effective alternative is to use a neural network to approximate the Q-function, a method called deep Q-learning.
-The general process for deep Q-learning differs from policy gradient methods, since instead of trying to maximize the return directly, the objective is to minimize the Bellman error.
-The Bellman error is a measure of how distant a given value function is from that value function which has been re-evaluated with recent environment interaction.
-Suppose we attempt to approximate the Q-function with a neural network and thus represent this approximate Q-function as as $Q_\phi$, the Bellman error for one state-action pair would be
+There are more value functions that can used in practice [@gaeschulman], but since Q-functions are commonly encountered, we will focus on Q-learning.
+From the above equation, we can define the Q function recursively,
+
+$$
+  Q^\pi(s_t, a_t) = r (s_t, a_t) + \gamma  \, \mathbb{E}_{s_{t+1} \sim T(s_{t+1} | s_t, a_t), \, a_{t+1} \sim \pi(a_{t+1}| s_{t+1})}\big[ Q^\pi(s_{t+1}, a_{t+1})\big]
+$$
+
+The motivation behind Q-learning is that if we know the Q function under certain conditions -- e.g.
+that the state-action space is discrete and that all state-action pairs are visited ad infinitum [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q (s_t, a_t)$.
+Using this optimal policy, we can write an equation for the optimal Q-function from the previous definitions,
+
+$$
+  Q^*(s_t, a_t) = r (s_t, a_t) + \gamma  \, \mathbb{E}_{s_{t+1} \sim T(s_{t+1} | s_t, a_t)}\big[ \underset{a_{t+1}}{\text{max}} \,\,Q^*(s_{t+1}, a_{t+1})\big]
+$$
+
+The objective for deep Q-learning is to find an approximate Q-function that satisfies the above equation, where we represent our approximate Q-function, $Q_\phi$, with a neural network that has the parameters of $\phi$.
+Subtracting the left hand and right hand side of the recursive Q-function definition, we can define the temporal difference error, $\mathcal{E}$, for a state-action pair as
 
@@
 
-See the appendix/reference for details on how the Bellman Error is derived from eq. [@CITE]. 
-The general scheme for approximate value-learning with deep neural networks is to sample trajectories from the environment, evaluate the Bellman Error across these samples -- with the samples indexed by $i$ --, and then perform the following gradient descent:
+The goal is to find a $Q_\phi$ that sets the TD error to zero across all state-action pairs. 
+The general process to achieve this objective is to sample trajectories from the environment, evaluate the TD Error across different state-action-reward transitions -- with samples indexed by $i$ --, and then perform the following gradient descent on the parameters of the Q-function neural network:
 
@@
 
-When being evaluated, a deep Q-learning agent will select actions by inputting the current state to the Q network and then identifying the state-action pair with the highest estimated Q value.
-The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITE].
-But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @CITE for more insight on these issues.   
-The primary concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@CITE].
+After the Q-function neural network or Q-network has been fitted, the agent will select actions by inputting the current state into the Q network and then identifying the state-action pair with the highest estimated Q value.
+The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITEmnihhumancontrol?].
+But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @mnihhumancontrol and @ddqn for more insight on these issues.   
+A notable concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@ddqn]. 
+In review, a sketch of simple deep Q learning algorithm is shown in ALGORITHM.
 
@@
 The "actor" attempts to learn the policy, and the "critic" attempts to learn a value function.
-While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from vanilla policy gradient algorithms by incorporating the critic in its gradient ascent step.
+While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from policy gradient algorithms by incorporating the critic in its gradient ascent step.
 Instead of estimating the gradient of the return as in eq ..., the gradient of the return is estimated as
@@
 where $\hat{Q}$ is a value function estimated by the critic.
-The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@CITE].
-Consequently, actor-critic algorithms have achieved a large majority of the state of the art results in model-free deep RL [@CITE].
+The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@suttonbarto].
+Consequently, actor-critic algorithms have achieved many of the state of the art results in model-free deep RL [@SAC; @TD3].  
+A simple sketch of an actor critic algorithm is shown in ALGORITHM.
 
-## Value Functions
+## Exploration
 
-There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
-The state-value function is defined as the expected return from a given state,
+Exploration is a central component of any deep RL algorithm, and there are a variety of exploration algorithms that an agent can employ.
+As mentioned in the manuscript, the exploration vs. exploitation dilemma is an open research problem.
 
-$$
-  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
-
-$$
-  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
-$$
-
-And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
-The advantage function can then be succinctly defined as
-
-$$
-  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
-$$ 
-Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
-Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
-Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].
 ----

### 2021-05-29 appendices/appendix-A-definitions.Rmd
Adding exploration paragraph

@@
 where $\hat{Q}$ is a value function estimated by the critic.
-The benefit of actor-critic algorithms is that they balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@suttonbarto].
+The benefit of actor-critic algorithms is that they can balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@suttonbarto].
 Consequently, actor-critic algorithms have achieved many of the state of the art results in model-free deep RL [@SAC; @TD3].  
@@
 Exploration is a central component of any deep RL algorithm, and there are a variety of exploration algorithms that an agent can employ.
-As mentioned in the manuscript, the exploration vs. exploitation dilemma is an open research problem.
+In the sections above, we procedurally mentioned that the agent must interact with the environment and then use its experiences to maximize rewards.
+Yet, how the agent explores is very important because if an agent fails to explore the state-action space well, then the agent will ultimately learn a bad policy.
+In this section, we will go over some common exploration algorithms and try to provide some practical insight.
+
+### Epsilon-Greedy Algorithm
+
+The Epsilon-Greedy algorithm takes a very simple approach towards balancing exploration and exploitation. 
+For some $\epsilon \in (0, 1)$ and typically close to 0, an Epsilon-Greedy agent exploits the best available action with probability $1-\epsilon$ and explores a random action with probability $\epsilon$.
+While simplistic, this algorithm has been effective on a range of problems notably @mnihhumanlevelcontrol.
+
+### Boltzmann Exploration
+
+For algorithms that learn a value function, agents can select actions by sampling from a Boltzmann or Softmax distribution that is constructed over the value function.
+So if the agent is doing Q-learning, the probability for the agent to select the action, $a \in \mathcal A$, would be:
+
+$$
+P(a) = \frac{e^{\frac{Q(s_t, a)}{\tau}}}{\sum_{a'} e^{\frac{Q(s_t, a')}{\tau}}}
+$$
+where $\tau \in (0, \infty)$ is called the temperature parameter and controls how much the agent will weight high Q scores over low Q scores.
+The advantage of using Boltzmann exploration is that since the probability of selecting an action is proportional to how good the agent thinks that action is, the agent can explore the space more efficiently than with Epsilon-Greedy exploration.
+This is because an Epsilon-Greedy agent will explore all actions equally including low reward yielding actions; in contrast, with Boltzmann exploration, an agent will spend less time on these unpromising actions.
+
+### Noise-Based Exploration
+
+Injecting noise into an agent's action, observation or parameter space is another effective exploration method.
+With noise addition, there are a lot design choices that can be made.
+For instance, noise can be sampled from different distributions [@parameter_space], scaled adaptively based on the policy[@parameter_space], or the noise can be sampled according to the state that the agent observes [@gsde].
+There are a variety of reasons why one would want to use these variations of noise injection, which are well discussed in @parameter_space and @gsde.
+
+### Entropy-based Exploration
+
+An increasingly common way to achieve exploration is by adding a function of the entropy term, $H(\pi(a|s))$, to the loss function of the policy network.
+While this introduces some bias as the agent no longer exclusively maximizes cumulative rewards, entropy-based exploration has been effective in practice at improving exploration, resulting in better agent performance.
+@understandingtheimpactofentropy gives a very complete discussion on the benefits of using entropy.
+
+There is a multitude of exploration algorithms that we have not touched on, so it is important to say that the algorithms mentionned above are generally suited for dense reward or "easy-exploration" problems.
+For "hard-exploration" problems -- for instance, environments that have very few state-action pairs with significant rewards, like a maze, are seen as hard-exploration problems --, more intricate and bespoke exploration algorithms tend to outperform.
+We will not go over these algorithms for the sake of brevity but a good example of an algorithm for hard-exploration problems is @go-explore.
 

### 2021-05-29 appendices/appendix-A-definitions.Rmd
Small edits

@@
 
-There is a multitude of exploration algorithms that we have not touched on, so it is important to say that the algorithms mentionned above are generally suited for dense reward or "easy-exploration" problems.
+There is a multitude of exploration algorithms that we have not touched on, so it is important to say that the algorithms mentioned above are generally suited for dense reward or "easy-exploration" problems.
 For "hard-exploration" problems -- for instance, environments that have very few state-action pairs with significant rewards, like a maze, are seen as hard-exploration problems --, more intricate and bespoke exploration algorithms tend to outperform.
-We will not go over these algorithms for the sake of brevity but a good example of an algorithm for hard-exploration problems is @go-explore.
+We will not go over these more advanced exploration algorithms but a good example is seen in @go-explore.
 

### 2021-06-01 appendices/appendix-A-definitions.Rmd
Have this done roughly but needs to be edited

@@
 The aim here is not to give full treatment of all deep RL methods -- @suttonbarto is an excellent resource for theoretical underpinnings while the white papers behind the algorithms go into greater detail on modern approaches --; instead, this section is meant to serve as an abbreviated background view on common, model-free deep RL techniques.
-We hope to convey the sense of what the different algorithm classes try to achieve and will point to specific resources for readers looking for a more complete presentation.
-We will discuss policy gradients, value-based and actor-critic critic methods which were briefly introduced in the manuscript. 
+We hope to introduce some of the important design choices that are made for model-free deep RL algorithms and will point to specific resources that give a more complete presentation.
 These sections presume some familiarity with gradient descent optimization -- see @ruder_overview_2017 for a thorough treatment.
@@
 
-Putting the 
 A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
@@
 After the Q-function neural network or Q-network has been fitted, the agent will select actions by inputting the current state into the Q network and then identifying the state-action pair with the highest estimated Q value.
-The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@CITEmnihhumancontrol?].
-But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @mnihhumancontrol and @ddqn for more insight on these issues.   
+The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@DQN].
+But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @DQN and @ddqn for more insight on these issues.   
 A notable concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@ddqn]. 
@@
 Exploration is a central component of any deep RL algorithm, and there are a variety of exploration algorithms that an agent can employ.
-In the sections above, we procedurally mentioned that the agent must interact with the environment and then use its experiences to maximize rewards.
-Yet, how the agent explores is very important because if an agent fails to explore the state-action space well, then the agent will ultimately learn a bad policy.
+In the sections above, we procedurally mentioned that the agent must select actions during training.
+Yet, how the agent selects these actions is very important because if an agent fails to explore the state-action space well, the agent will likely learn a bad policy.
 In this section, we will go over some common exploration algorithms and try to provide some practical insight.
@@
 For some $\epsilon \in (0, 1)$ and typically close to 0, an Epsilon-Greedy agent exploits the best available action with probability $1-\epsilon$ and explores a random action with probability $\epsilon$.
-While simplistic, this algorithm has been effective on a range of problems notably @mnihhumanlevelcontrol.
+While simplistic, this algorithm has been effective on a range of problems notably @DQN.
 
@@
 $$
-P(a) = \frac{e^{\frac{Q(s_t, a)}{\tau}}}{\sum_{a'} e^{\frac{Q(s_t, a')}{\tau}}}
+P(a) = \frac{e^{\frac{Q(s_t, a)}{\tau}}}{\underset{a'}{\sum} e^{\frac{Q(s_t, a')}{\tau}}}
 $$
@@
 
+# Replay Buffers vs. Parallel Workers
+
+Another important algorithm design choice involves how an RL agent collects and samples past experiences.
+The main issue that arises from sampling past experiences is correlation of samples [@DQN].
+In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated.
+Learning from correlated samples is undesirable since this can result in agents getting stuck in local optima, which will impede learning.
+There are two commonly used solutions to this problem of correlated samples: Replay Buffers and Parallel Workers.
+If the agent is off-policy, the agent can employ a replay buffer.
+With a replay buffer, the agent will save state-action-reward transitions in the replay buffer, and during learning updates, the agent will sample transitions from the replay buffer.
+Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer will likely produce transitions that cover a broader region in state-action space than if the agent were to sample from one trajectory.
+On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- note that in a replay buffer, it is possible to have samples from different policies.
+The solution to de-correlate samples for on-policy agents is to use parallel workers, whereby multiple workers explore the environment according to the current policy. 
+Due to stochasticity that can come from exploration and the environment, the workers will go on different trajectories in state-action space.
+The on-policy agent will collect samples across these workers in a learning step, and then repeat the process of collecting and sampling transitions with an updated policy.
+Once again, the idea behind using parallel workers is that sampling over multiple trajectories will de-correlate samples.
+In practice, using a large replay buffer, on the order of millions of transitions, or a modest number of parallel, around 4-16 for simple RL problems, is very important to achieve good agent performance for off-policy and on-policy algorithms respectively.
+
+# Next Steps
+
+Reinforcement learning is a huge, constantly evolving field, and we've only touched on a small sliver of RL in our study.
+In this section, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with current state of the art algorithms.
+If you are interested in learning more about model-free deep RL, then reading papers like @SAC, @DQN and @PPO, are good places to start.
+We have also not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @deep_neuro for more here.
+A benefit of model-free deep RL over these other approaches is that there has been more of a general focus on model-free deep RL, so there are widely available packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL; but the focus on model-free algorithm could easily change in coming years.
+Deep RL has only become established in the last decade, and with its amount of unanswered research questions, deep RL may evolve significantly in the near future.
+
+# References
 ----

### 2021-06-01 appendices/appendix-A-definitions.Rmd
Adding citations

@@
 In this appendix, we will go into further detail on how deep RL algorithms work.
-The aim here is not to give full treatment of all deep RL methods -- @suttonbarto is an excellent resource for theoretical underpinnings while the white papers behind the algorithms go into greater detail on modern approaches --; instead, this section is meant to serve as an abbreviated background view on common, model-free deep RL techniques.
+The aim here is not to give full treatment of all deep RL methods -- @suttonbarto is an excellent resource for theoretical underpinnings while the white papers behind the algorithms go into greater detail on modern approaches --; instead, this section is meant to serve as an abbreviated background view on model-free deep RL.
 We hope to introduce some of the important design choices that are made for model-free deep RL algorithms and will point to specific resources that give a more complete presentation.
@@
 Throughout these sections, we'll consider the environment to be an MDP which makes for a cleaner presentation here and is also consistent with the referenced sources.
+
+# Gradient-based Model-free RL Algorithms
+
+Policy gradients, value-based and actor-critic methods are the general classes of algorithms for gradient-based model-free RL.
+In this section, we will go over these algorithm classes to convey a broad sense of what they are attempting to achieve and the trade-offs between them.
 
@@
 
-There are more value functions that can used in practice [@gaeschulman], but since Q-functions are commonly encountered, we will focus on Q-learning.
+There are more value functions that can used in practice [@schulman_high-dimensional_2018], but since Q-functions are commonly encountered, we will focus on Q-learning.
 From the above equation, we can define the Q function recursively,
@@
 The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@DQN].
-But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @DQN and @ddqn for more insight on these issues.   
-A notable concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@ddqn]. 
+But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @DQN and @van_hasselt_deep_2015 for more insight on these issues.   
+A notable concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@van_hasselt_deep_2015].
 In review, a sketch of simple deep Q learning algorithm is shown in ALGORITHM.
@@
 The benefit of actor-critic algorithms is that they can balance the issues with bias and variance that are common with value-learning and policy gradient methods respectively [@suttonbarto].
-Consequently, actor-critic algorithms have achieved many of the state of the art results in model-free deep RL [@SAC; @TD3].  
+Consequently, actor-critic algorithms have achieved many of the state of the art results in model-free deep RL [@sac; @TD3].  
 A simple sketch of an actor critic algorithm is shown in ALGORITHM.
 
-## Exploration
+# Exploration
 
@@
 
-### Epsilon-Greedy Algorithm
+## Epsilon-Greedy Algorithm
 
@@
 
-### Boltzmann Exploration
+## Boltzmann Exploration
 
@@
 
-### Noise-Based Exploration
+## Noise-Based Exploration
 
@@
 With noise addition, there are a lot design choices that can be made.
-For instance, noise can be sampled from different distributions [@parameter_space], scaled adaptively based on the policy[@parameter_space], or the noise can be sampled according to the state that the agent observes [@gsde].
-There are a variety of reasons why one would want to use these variations of noise injection, which are well discussed in @parameter_space and @gsde.
+For instance, noise can be sampled from different distributions [@plappert_parameter_2018], scaled adaptively based on the policy[@plappert_parameter_2018], or the noise can be sampled according to the state that the agent observes [@gsde].
+There are a variety of reasons why one would want to use these variations of noise injection, which are well discussed in @plappert_parameter_2018 and @gsde.
 
-### Entropy-based Exploration
+## Entropy-based Exploration
 
@@
 While this introduces some bias as the agent no longer exclusively maximizes cumulative rewards, entropy-based exploration has been effective in practice at improving exploration, resulting in better agent performance.
-@understandingtheimpactofentropy gives a very complete discussion on the benefits of using entropy.
+@ahmed_understanding_2019 gives a very complete discussion on the benefits of using entropy.
 
@@
 In this section, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with current state of the art algorithms.
-If you are interested in learning more about model-free deep RL, then reading papers like @SAC, @DQN and @PPO, are good places to start.
-We have also not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @deep_neuro for more here.
+If you are interested in learning more about model-free deep RL, then reading papers like @sac, @DQN and @ppo, are good places to start.
+We have also not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @salimans_evolution_2017 for more here.
 A benefit of model-free deep RL over these other approaches is that there has been more of a general focus on model-free deep RL, so there are widely available packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL; but the focus on model-free algorithm could easily change in coming years.

### 2021-06-02 appendices/appendix-A-definitions.Rmd
Adding algorithm blocks

@@
     keep_tex: true
+    extra_dependencies:
+      algorithm: null
+      algpseudocode: noend
+
 ---
@@
 
-A sketch of a simple policy gradient algorithm is shown in ALGORITHM.
+A sketch of a simple policy gradient algorithm is shown in Algorithm 1.
+
+\begin{algorithm}
+\label{alg:policy_gradient}
+\caption{Vanilla On-policy Policy Gradient Algorithm}
+\begin{algorithmic}[1]
+    \State Input initial policy parameters $\theta_0$
+    \For{$k = 0, 1, 2, \dots$}
+        \State Generate a trajectory $\{s_i, a_i, r_i\}$ by following an exploration algorithm with the policy $\pi_k$ (e.g. $\epsilon$-greedy)
+        \For{$t = 0, 1, \dots, T-1$}
+            \State $G \gets \sum_{j=t}^{T} \gamma^{j-t} r_j$
+            \State $\theta_{k+1} \gets \theta_k + \alpha \gamma^t G \, \nabla \log \pi_{\theta_k}(a_t | s_t)$
+        \EndFor
+    \EndFor
+\end{algorithmic}
+\end{algorithm}
 
@@
 A notable concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@van_hasselt_deep_2015].
-In review, a sketch of simple deep Q learning algorithm is shown in ALGORITHM.
+In review, a sketch of simple deep Q learning algorithm is shown in Algorithm 2.
+
+\begin{algorithm}
+\caption{Vanilla On-policy Q-learning Algorithm}
+\begin{algorithmic}[1]
+    \State Input initial Q-network parameters $\phi_0$
+    \For{$k = 0, 1, 2, \dots$}
+        \For{$t = 0, 1, \dots, T-1$}
+            \State Choose the action $a_t$ using $Q_{\phi_k}$ according to an exploration algorithm (e.g. $\epsilon$-greedy)
+            \State Take action $a_t$ and observe $r_{t}, s_{t+1}$
+            \State $ \mathcal{E} \gets r_t + \gamma \, \underset{a}{\text{max}} \, Q_{\phi_{k}}(s_{t+1}, a) - Q_{\phi_{k}}(s_t, a_t)$
+            \State  $\phi_{k+1} \gets \phi_{k} - \alpha \nabla_{\phi_k} \, \mathcal{E}^2$
+        \EndFor
+    \EndFor
+\end{algorithmic}
+\end{algorithm}
 
@@
 Consequently, actor-critic algorithms have achieved many of the state of the art results in model-free deep RL [@sac; @TD3].  
-A simple sketch of an actor critic algorithm is shown in ALGORITHM.
+A simple sketch of an actor critic algorithm is shown in Algorithm 3.
+
+\begin{algorithm}
+\caption{Vanilla On-policy Actor-Critic Algorithm}
+\begin{algorithmic}[1]
+    \State Input initial policy and Q-network parameters $\theta_0, \phi_0$
+    \For{$k = 0, 1, 2, \dots$}
+        \State Generate a trajectory $\{s_i, a_i, r_i\}$ by following an exploration algorithm with the policy $\pi_k$ (e.g. $\epsilon$-greedy)
+        \For{$t = 0, 1, \dots, T-1$}
+            \State $ \mathcal{E} \gets r_t + \gamma \, \underset{a}{\text{max}} \, Q_{\phi_{k}}(s_{t+1}, a) - Q_{\phi_{k}}(s_t, a_t)$
+            \State $\phi_{k+1} \gets \phi_{k} - \alpha \nabla_{\phi_k} \, \mathcal{E}^2$
+            \State $\theta_{k+1} \gets \theta_k + \alpha \, \nabla \log \pi_{\theta_k}(a_t | s_t) Q_{\phi_k}(s_t, a_t)$
+        \EndFor
+    \EndFor
+\end{algorithmic}
+\end{algorithm}
 
@@
 Another important algorithm design choice involves how an RL agent collects and samples past experiences.
-The main issue that arises from sampling past experiences is correlation of samples [@DQN].
+A significant issue that arises from sampling past experiences is correlation of samples [@DQN, @schaul_prioritized_2016].
 In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated.
-Learning from correlated samples is undesirable since this can result in agents getting stuck in local optima, which will impede learning.
+Learning from correlated samples is undesirable since this can result in agents getting stuck in local optima, which will impede learning, but there are other reasons why learning from a single trajectory causes problems -- see @schaul_prioritized_2016 and @DQN for more insight here.
 There are two commonly used solutions to this problem of correlated samples: Replay Buffers and Parallel Workers.
-If the agent is off-policy, the agent can employ a replay buffer.
+If the agent is off-policy, the agent can employ a replay buffer -- see @DQN for a now classic example for a replay buffer-based agent.
 With a replay buffer, the agent will save state-action-reward transitions in the replay buffer, and during learning updates, the agent will sample transitions from the replay buffer.
 Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer will likely produce transitions that cover a broader region in state-action space than if the agent were to sample from one trajectory.
-On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- note that in a replay buffer, it is possible to have samples from different policies.
+On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- replay buffers are generally large enough to hold transitions from old policies.
 The solution to de-correlate samples for on-policy agents is to use parallel workers, whereby multiple workers explore the environment according to the current policy.
@@
 The on-policy agent will collect samples across these workers in a learning step, and then repeat the process of collecting and sampling transitions with an updated policy.
-Once again, the idea behind using parallel workers is that sampling over multiple trajectories will de-correlate samples.
+The idea behind using parallel workers is that sampling over multiple trajectories will de-correlate samples -- see @a3c on more details here.
 In practice, using a large replay buffer, on the order of millions of transitions, or a modest number of parallel, around 4-16 for simple RL problems, is very important to achieve good agent performance for off-policy and on-policy algorithms respectively.
@@
 # References
+
 ----

### 2021-06-02 manuscript/manuscript.Rmd

--- 2021-05-27 manuscript/manuscript.Rmd
@@
 Successfully applying RL to complex ecological problems is no easy task.
-The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem, but training an RL agent to reach an acceptable level of performance can often be more challenging than expected.
+The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem.
+Yet, even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters].
 As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
@@
 and can easily be extended into more complex problems once RL agents can master the basics.
-Finding acceptable policies even in relatively simple problems may prove beyond the reach of existing algorithms and available computing power.
-In seeking either to adopt or assess a deep RL-based approach to an ecological issue, we highlight three key issues that are important to bear in mind: generalizability, interpretability, and reward design.
+In the case that someone trains an agent to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
 
-<!-- reason 1: generalizability-->
-<!-- ML: pivoting this to generalizability but still rewrote past paragraph which might fit in the appendix somewhere-->
-Currently, deep RL agents struggle to generalize their skills [@pmlr-v97-cobbe19a].
-This inability can affect performance during training and at evaluation. 
-For instance, training an agent on an environment that demands learning an assortment of skills is typically much more difficult than training an agent to learn one specific skill [@teh_distral_2017].
-And, at evaluation, if you give an agent a novel task, including tasks that are closely related to the examples that the agent saw in training, the agent can easily fail to generalize its learned behavior to this new situation [@pmlr-v97-cobbe19a].
-These failures are all due to the brittleness of deep neural networks. 
-Across deep learning, neural networks tend to perform exceptionally well on narrowly defined problems, yet neural networks show great fragility when given new situations or problem statements that require flexibility [@cogswell_reducing_2016].
-The ability to adapt to new situations is critical for any decision-making problem in the real world, so the current lack of this ability in deep RL may limit deep RL's effectiveness to conservation problems.
+- Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the deep RL agent will generalize its past experience to new situations?
 
-<!--
-Under some conditions, RL agents can prove adept at _transfer learning_, in which an agent trained in one environment performs well under a different environment.
--->
-<!-- if we're really recommending 'testing' for stability, maybe we need to demo that in the appendix-->
-<!-- reason 2: interpretability -->
-Deep RL agents are also difficult to interpret [@heuillet_explainability_2021].
-Deep neural networks are commonly referred to as black box models, since it is challenging to fully explain why a deep neural network produces a specific output from a given input [@castelvecchi_can_2016].
-For deep RL, this means that we cannot provide a meaningful explanation for why any agent selects an action.
-Compounded with the previously mentioned generalizability issues, the lack of explainability introduces a lot of safety concerns when attempting to apply these methods to the real world.
-When going from a simulator to the real world, the real world may introduce novel situations that a simulator-trained agent is not familiar with, and since we can't anticipate how an agent will respond to new situations, there is an inherent risk in letting RL agents make decisions in the real world.
-
-<!-- reason 3: reward design -->
-And lastly, a fundamental issue for all applied optimization problems is using the right objective.
-There have been many examples in which explicit reward design produces unanticipated agent behavior that is highly undesirable [@hadfield-menell_inverse_2020].
-Thus, for ecological systems, which are typically quite difficult to control, specifying a reward function could easily produce behavior that opposes the intents of the RL practitioner.
-To guide agents to learn desirable policies, RL researchers often *shape* the reward function so that the agent will more readily learn certain behaviors [@reward_shaping].
-This approach is particularly common when rewards are *sparse*, which occurs when there are few state-action pairs that have strong feedback signals.
-Frequently, when researchers do not shape the rewards on sparse reward problems, the RL agent learns nothing, but reward shaping can still backfire by producing unintended agent behavior [@hadfield-menell_inverse_2020].
-Given the nature of many conservation problems, where benefits rarely come immediately and thus the reward structures are sparse, RL agents will likely have difficulty learning anything, or, in the case that researchers shape the rewards to aid learning, the agents may learn policies that are undesirable.
-Moreover, specifying a reward function for conservation brings up various ethical concerns.
-E.g., who gets to specify the reward function, and will the values of the greater community be represented in the reward function?
+- Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?
 

### 2021-06-02 manuscript/manuscript.Rmd
Starting up a comments, small edit branch

@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
-Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
+Even simpler simulators are frequently employed throughout ecology to guide understanding and response. <!-- ML: I think this sentence could be dropped -->
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
@@
 In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
+<!-- ML: If we are dropping wildfire example, we'll need to cut wildfire mentions here-->
 Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
@@
 
-An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of the environment's current *state* and an associated *reward* as output.
 In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
@@
 
+<!-- ML: I have some qualms about this figure; I think it can do a bit more to display what is going on.
+Like perhaps can put dashed arrows within a trial to show that there is continuity between episodes but not continuity across trials.
+We could also put a dashed box around the trials to this effect as well.
+In these dashed boxes, we could put smaller hyperparameter boxes to convey that different trials have distinct hyperparameters.
+I am also not a fan of the object used for neural networks.
+-->
 \begin{figure}
@@
 The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').
-While this library of environments span a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
+While this library of environments spans a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
 We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
+In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
 
@@
 # Formal RL Concepts
+<!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
 
-In this brief section, we will introduce the concept of the RL environment as a POMDP; we will then use POMDP definitions to define the RL objective.
-Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is crucial to understanding how RL algorithms work.
+In this brief section, we will introduce the concept of the RL environment as partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.
+Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is critical to understanding how RL algorithms work.
 <!-- ML: this seems like an obvious sentence to me but am stuck on it atm so revisions welcomed-->
 
-The reinforcement learning environment is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
+The reinforcement learning environment is typically formalized as a discrete-time POMDP.
 A POMDP is a 7-tuple that consists of the following:
@@
 And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
-There is a handful of advantages and disadvantages for each of these approaches that we will not get into in depth -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@suttonbarto].
-Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@suttonbarto; @qbias].
-<!-- ML: I think we can keep the bias-variance discussion in the appendix and cut these 2 sentences-->
+We go into greater depth on these different classes of RL algorithms in Appendix A.
 
@@
 when the possible states and actions can not be represented in a tractable table [@suttonbarto].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments [@DQN, @alphaGo2016].
 Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
@@
 The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014], and current deep RL algorithms take a range of approaches towards this dilemma -- in the appendix, we discuss some of the more specific exploration algorithms.
-Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
+This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014], and current deep RL algorithms take a range of approaches towards this dilemma -- in Appendix A, we discuss some of the more specific exploration algorithms.
+Yet, overall, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
 The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
-For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
-There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the appendix.
+For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent updates the same policy in training as the agent will use during evaluation.
+There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the Appendix A.
 
@@
 Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
-In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
-In the appendix, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
+In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics. <!-- ML: Have we dropped the table? -->
+In Appendices A and B, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
 
@@
 When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].
+<!-- ML: Will probably want to de-emphasize this given the momentum to sigma of 0 -->
 
@@
 
-<!-- ML: I'm have some uneasiness about this figure. Is A2C tuned? I think we'd probably want to do tuning and training with a set seed so that everything is reproducible?  -->
 ```{r message=FALSE, include=FALSE}

### 2021-06-03 manuscript/manuscript.Rmd
Few more comments

@@
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
-From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
+From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
 In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment." The simplest simulation is merely an implementation of the estimated model itself.
@@
 
-The TD3-trained neural network agent proves far more successful in preventing chance transitions across the tipping point, consistently achieving a higher cumulative ecosystem service value across replicates than the steady-state strategy.
+The TD3-trained agent proves far more successful in preventing chance transitions across the tipping point, consistently achieving a higher cumulative ecosystem service value across replicates than the steady-state strategy.
 
@@
 For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
-Recall that the quadratic scaling of cost makes the rapid response of the TD3 algorithm much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
+Recall that the quadratic scaling of cost makes the rapid response of the TD3 agent much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
 However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.
@@
 The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
+<!-- ML: the parenthetical here I understand but I feel a little ehh on the word choice.
+I also don't know how to rephrase it -->
 Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
-<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote.  CB: agree, revised above-->
 
@@
 Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
+<!-- ML: A counterpoint here is that a model-based agent may learn a correct model that generalizes outside of the examples it has seen-->
 In these simple examples in which the simulation implements a single model, training is analogous to classical methods which take the model as given [@Marescot2013].
@@
 and can easily be extended into more complex problems once RL agents can master the basics.
-In the case that someone trains an agent to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
+In the case that an agent has been trained to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
 

### 2021-06-03 manuscript/manuscript.Rmd
Cutting out wildfire mentions in intro

@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
-Even simpler simulators are frequently employed throughout ecology to guide understanding and response. <!-- ML: I think this sentence could be dropped -->
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
@@
 
-In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-<!-- ML: If we are dropping wildfire example, we'll need to cut wildfire mentions here-->
+In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
 Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
@@
 In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
-In our third example simulating wildfire spread highlights an open problem which we expect will be considerably harder to find effective solutions.
 These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
@@
 
+<!-- ML: Need to rework this to talk about other environments that we are working on basically point to conservation gym website -->
 One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.

### 2021-06-03 python/fishing_SAC.py
try SAC

@@ -1,0 +1,50 @@
+# Fishing using tuned SB3
+
+import numpy as np
+import os
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+from stable_baselines3.common.monitor import Monitor
+from stable_baselines3.common.evaluation import evaluate_policy
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+# Create an agent
+hyper = {
+    "gamma": 0.95,
+    "learning_rate": 0.000115,
+    "batch_size": 128,
+    "buffer_size": 100000,
+    "learning_starts": 0,
+    "train_freq": 128,
+    "tau": 0.01,
+    "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
+}
+
+agent = sb3.SAC("MlpPolicy", env, seed=seed, **hyper)
+
+# Train the agent
+if not os.path.exists("cache/sac_tuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/sac_tuned")
+
+
+agent = sb3.SAC.load("cache/sac_tuned")
+
+# Evaluate the trained agent
+eval_env = Monitor(gym.make("fishing-v1"))
+mean_reward, std_reward = evaluate_policy(agent, eval_env, n_eval_episodes=100)
+print("mean reward:", mean_reward, "std:", std_reward)
+
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=5)
+
+# Plot results
+env.plot(agent_sims, "results/fishing_sac_tuned.png")
+env.plot_policy(agent_policy, "results/fishing_sac_tuned_policy.png")

### 2021-06-03 python/fishing_SAC_untuned.py

@@ -1,0 +1,33 @@
+# Fishing using tuned TD3
+
+import os
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+# Create an agent
+agent = sb3.SAC("MlpPolicy", env, seed=seed)
+
+# Train the agent
+if not os.path.exists("cache/sac_untuned.zip"):
+  agent.learn(total_timesteps=300000)
+  agent.save("cache/sac_untuned")
+
+# Evaluate the trained agent
+agent = sb3.SAC.load("cache/sac_untuned")
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=5)
+
+# Plot results
+env.plot(agent_sims, "results/fishing_sac_untuned.png")
+env.plot_policy(agent_policy, "results/fishing_sac_untuned_policy.png")
+
+
+

### 2021-06-03 python/tuning/utils/record_training.py

@@ -1,159 +1,0 @@
-import argparse
-import os
-import re
-import shutil
-import subprocess
-from copy import deepcopy
-
-from utils.utils import ALGOS, get_latest_run_id
-
-if __name__ == "__main__":  # noqa: C901
-    parser = argparse.ArgumentParser()
-    parser.add_argument("--env", help="environment ID", type=str, default="CartPole-v1")
-    parser.add_argument("-f", "--folder", help="Log folder", type=str, default="rl-trained-agents")
-    parser.add_argument("-o", "--output-folder", help="Output folder", type=str)
-    parser.add_argument("--algo", help="RL Algorithm", default="ppo", type=str, required=False, choices=list(ALGOS.keys()))
-    parser.add_argument("-n", "--n-timesteps", help="number of timesteps", default=1000, type=int)
-    parser.add_argument("--n-envs", help="number of environments", default=1, type=int)
-    parser.add_argument("--deterministic", action="store_true", default=False, help="Use deterministic actions")
-    parser.add_argument("-g", "--gif", action="store_true", default=False, help="Convert final video to gif")
-    parser.add_argument("--seed", help="Random generator seed", type=int, default=0)
-    parser.add_argument("--exp-id", help="Experiment ID (default: 0: latest, -1: no exp folder)", default=0, type=int)
-    args = parser.parse_args()
-
-    env_id = args.env
-    algo = args.algo
-    folder = args.folder
-    n_timesteps = args.n_timesteps
-    n_envs = args.n_envs
-    video_folder = args.output_folder
-    seed = args.seed
-    deterministic = args.deterministic
-    convert_to_gif = args.gif
-
-    if args.exp_id == 0:
-        args.exp_id = get_latest_run_id(os.path.join(folder, algo), env_id)
-        print(f"Loading latest experiment, id={args.exp_id}")
-    # Sanity checks
-    if args.exp_id > 0:
-        log_path = os.path.join(folder, algo, f"{env_id}_{args.exp_id}")
-    else:
-        log_path = os.path.join(folder, algo)
-
-    assert os.path.isdir(log_path), f"The {log_path} folder was not found"
-
-    if video_folder is None:
-        video_folder = os.path.abspath(os.path.join(log_path, "videos"))
-    shutil.rmtree(video_folder, ignore_errors=True)
-    os.makedirs(video_folder, exist_ok=True)
-
-    # record a video of every model
[…]

### 2021-06-03 python/tuning/utils/record_video.py

@@ -1,111 +1,0 @@
-import argparse
-import os
-
-from stable_baselines3.common.vec_env import VecVideoRecorder
-
-from utils.exp_manager import ExperimentManager
-from utils.utils import ALGOS, create_test_env, get_latest_run_id, get_saved_hyperparams
-
-if __name__ == "__main__":
-    parser = argparse.ArgumentParser()
-    parser.add_argument("--env", help="environment ID", type=str, default="CartPole-v1")
-    parser.add_argument("-f", "--folder", help="Log folder", type=str, default="rl-trained-agents")
-    parser.add_argument("-o", "--output-folder", help="Output folder", type=str)
-    parser.add_argument("--algo", help="RL Algorithm", default="ppo", type=str, required=False, choices=list(ALGOS.keys()))
-    parser.add_argument("-n", "--n-timesteps", help="number of timesteps", default=1000, type=int)
-    parser.add_argument("--n-envs", help="number of environments", default=1, type=int)
-    parser.add_argument("--deterministic", action="store_true", default=False, help="Use deterministic actions")
-    parser.add_argument("--seed", help="Random generator seed", type=int, default=0)
-    parser.add_argument(
-        "--no-render", action="store_true", default=False, help="Do not render the environment (useful for tests)"
-    )
-    parser.add_argument("--exp-id", help="Experiment ID (default: 0: latest, -1: no exp folder)", default=0, type=int)
-    parser.add_argument(
-        "--load-best", action="store_true", default=False, help="Load best model instead of last model if available"
-    )
-    parser.add_argument(
-        "--load-checkpoint",
-        type=int,
-        help="Load checkpoint instead of last model if available, you must pass the number of timesteps corresponding to it",
-    )
-    args = parser.parse_args()
-
-    env_id = args.env
-    algo = args.algo
-    folder = args.folder
-    video_folder = args.output_folder
-    seed = args.seed
-    deterministic = args.deterministic
-    video_length = args.n_timesteps
-    n_envs = args.n_envs
-    load_best = args.load_best
-    load_checkpoint = args.load_checkpoint
-
-    if args.exp_id == 0:
-        args.exp_id = get_latest_run_id(os.path.join(folder, algo), env_id)
-        print(f"Loading latest experiment, id={args.exp_id}")
-    # Sanity checks
-    if args.exp_id > 0:
[…]

### 2021-06-03 manuscript/manuscript.Rmd
Starting to work on an additional environment paragraph

--- 2021-06-02 manuscript/manuscript.Rmd
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
-Even simpler simulators are frequently employed throughout ecology to guide understanding and response. <!-- ML: I think this sentence could be dropped -->
+Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
@@
 In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-<!-- ML: If we are dropping wildfire example, we'll need to cut wildfire mentions here-->
 Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
@@
 
-An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of the environment's current *state* and an associated *reward* as output.
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
 In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
@@
 
-<!-- ML: I have some qualms about this figure; I think it can do a bit more to display what is going on.
-Like perhaps can put dashed arrows within a trial to show that there is continuity between episodes but not continuity across trials.
-We could also put a dashed box around the trials to this effect as well.
-In these dashed boxes, we could put smaller hyperparameter boxes to convey that different trials have distinct hyperparameters.
-I am also not a fan of the object used for neural networks.
--->
 \begin{figure}
@@
 The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').
-While this library of environments spans a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
+While this library of environments span a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
 We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
-In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
 
@@
 # Formal RL Concepts
-<!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
 
-In this brief section, we will introduce the concept of the RL environment as partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.
-Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is critical to understanding how RL algorithms work.
+In this brief section, we will introduce the concept of the RL environment as a POMDP; we will then use POMDP definitions to define the RL objective.
+Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is crucial to understanding how RL algorithms work.
 <!-- ML: this seems like an obvious sentence to me but am stuck on it atm so revisions welcomed-->
 
-The reinforcement learning environment is typically formalized as a discrete-time POMDP.
+The reinforcement learning environment is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
 A POMDP is a 7-tuple that consists of the following:
@@
 And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
-We go into greater depth on these different classes of RL algorithms in Appendix A.
+There is a handful of advantages and disadvantages for each of these approaches that we will not get into in depth -- for more, see the appendix; yet, the main point of divergence comes from the bias-variance trade-off [@suttonbarto].
+Generally, policy gradient algorithms tend to struggle with high variance, value-based methods tends to be biased while actor-critic methods balance these two concerns [@suttonbarto; @qbias].
+<!-- ML: I think we can keep the bias-variance discussion in the appendix and cut these 2 sentences-->
 
@@
 when the possible states and actions can not be represented in a tractable table [@suttonbarto].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments [@DQN, @alphaGo2016].
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
 Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
@@
 The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014], and current deep RL algorithms take a range of approaches towards this dilemma -- in Appendix A, we discuss some of the more specific exploration algorithms.
-Yet, overall, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
+This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014], and current deep RL algorithms take a range of approaches towards this dilemma -- in the appendix, we discuss some of the more specific exploration algorithms.
+Yet, generally, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
 The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
-For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent updates the same policy in training as the agent will use during evaluation.
-There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the Appendix A.
+For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent uses the same policy in training as the agent will use during evaluation.
+There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the appendix.
 
@@
 Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
-In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics. <!-- ML: Have we dropped the table? -->
-In Appendices A and B, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
+In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics.
+In the appendix, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
 
@@
 When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].
-<!-- ML: Will probably want to de-emphasize this given the momentum to sigma of 0 -->
 
@@
 
-One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
-While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
-Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
-Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
-Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+One feature that most distinguishes deep RL from other approaches to decision theory (such as optimal control) is its flexibility and scope.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, deep RL is scoped more ambitiously.
+Large, complex problems are still challenging to RL, and existing RL approaches can easily struggle to solve problems that can already be solved much more efficiently by other means.
+Nevertheless, deep RL approaches have proven successful on problems that classical approaches could not solve despite immense computational resources -- e.g. deep RL  [@DQN, @alphago].
+Our previous examples have focused on relatively simple problems that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+Yet, we also have developed
 

### 2021-06-04 manuscript/manuscript.Rmd
Updating comment for figure 1

--- 2021-06-03 manuscript/manuscript.Rmd
@@
 
-<!-- ML: I have some qualms about this figure; I think it can do a bit more to display what is going on.
-Like perhaps can put dashed arrows within a trial to show that there is continuity between episodes but not continuity across trials.
-We could also put a dashed box around the trials to this effect as well.
-In these dashed boxes, we could put smaller hyperparameter boxes to convey that different trials have distinct hyperparameters.
-I am also not a fan of the object used for neural networks.
--->
 \begin{figure}
 \includegraphics[width=\linewidth]{fig1.pdf}
-\caption{Deep Reinforcement Learning:  An artificial intelligence \emph{agent} uses a \emph{neural network} to select an \emph{action} in response to an \emph{observation} of the \emph{environment}, receiving a \emph{reward} (or punishment) from the environment as a result.
-The agent is \emph{trained} to maximize the reward through the experience of trying different actions over repeated \emph{epsiodes} of interacting with the environment.
-Over the course of training, the agent adjusts parameter weights on the neural network as it learns which actions lead to the best cumulative rewards.
-A researcher may \emph{tune} an agent by considering iterative trials that train different agents with different \emph{hyperparameters}, such as different neural network architectures.
-An agent trained with the most successful combination of hyperparameters can then be tested in an evaluation environment.}
+\caption{Deep Reinforcement Learning:  A deep RL \emph{agent} uses a \emph{neural network} to select an \emph{action} in response to an \emph{observation} of the \emph{environment}, and receives a \emph{reward} from the environment as a result.
+During \emph{training}, the agent tries to maximize its cumulative reward by interacting with the environment and learning from experience.
+In the RL loop, the agent performs an action, and the environment returns a reward as well as an observation of the environment's state.
+The agent-environment loop continues until the environment reaches a terminal state, after which the environment will reset, causing a new \emph{episode} to begin.
+Across training episodes, the agent will continually update the \emph{parameters} in its neural network, so that the agent will select better actions.
+Before training starts, the researcher must input a set of \emph{hyperparameters} to the agent; hyperparameters direct the learning process and thus affect the outcome of training.
+A researcher finds the best set of hyperparameters during \emph{tuning}.
+Hyperparameter tuning consists of iterative \emph{trials}, in which the agent is trained with different sets of hyperparameters.
+At the end of a trial, the agent is evaluated to see which agent attains the highest cumulative reward.
+An agent is \emph{evaluated} by recording the cumulative reward over one episode, or the mean reward over multiple episodes.
+Within evaluation, the agent does not update its neural network; instead, the agent only uses the neural network to select actions.}
 \end{figure}

### 2021-06-04 appendices/appendix-B-examples.Rmd
reworking example to PPO

--- 2021-05-20 appendices/appendix-B-examples.Rmd
@@
 ---
+
+# Overview
+
+In this appendix, we provide examples with annotated code illustrating how to use recent deep RL algorithms
+to train RL agents to make decisions in two distinct ecological management scenarios: setting a harvest quota in a fishery, and managing conservation effort in the face of ecological tipping points.
+The approach illustrated here should be sufficient to reproduce the results presented in the main text,
+and should serve as an effective starting point for researchers seeking to apply deep RL to these simulation environments, 
+to other simulation environments we have built and continue to develop, or to their own environments.
+(Appendix C details the process of defining one's own `gym` environment from an ecological model). 
+This overview is neither a complete introduction to deep RL or a comprehensive analysis of RL solutions to these environments.
 
@@
 
-# Building a simulation environment `gym`
+# Deep Reinforcement Learning in R
 
-[@openai_gym]
+Although all the necessary tooling for deep RL is most extensively implemented in Python, the R language is more familiar to most ecologists.
+Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
+In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
+
+In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
+This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
+Alternately, users may prefer running the analysis inside a docker container.  
+The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
+This can be particularly useful for users running ML algorithms on remote servers.
+
+By using a cache of pre-trained agents distributed with this appendix, it should be possible to efficiently reproduce the results shown here on any recent laptop with the versions of software indicated.
+Reproducing the training of RL agents is more computationally intensive, and a more powerful multi-core CPU or CUDA-enabled GPU is recommended.
+Note that hardware differences are also a source of instability in the reproduciblity of RL training (see main text and Appendix A).
 
@@
 
-At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
+Rather than implement deep RL algorithms from scratch, we illustrate the use of one of the leading software frameworks for research and application of deep RL methods.
+At the time of writing, several major frameworks exist which provide reference implementations of the leading deep RL algorithms (see table 1 and Appendix A).
 While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
@@
 
-# Deep Reinforcement Learning in R
-
-Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
-Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
-In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
-
-In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
-This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
-Alternately, users may prefer running the analysis inside a docker container.  
-The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
-This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.
+# Getting started
 
@@
 From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
+(Python users can install the dependencies listed in the repository's `requirements.txt` and run the corresponding python scripts found in the `python/` subidrectory instead).
 
@@
 
-We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
+We begin by selecting an environment from `gym_fishing`.
+While we use default parameter values for this example, note that is possible to pass alternative parameterizations using additional optional arguments.
 
@@
 ## initialize the environment
-env = gym$make("fishing-v1", sigma = 0.1)
+env = gym$make("fishing-v1")
 ```
@@
 Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
-Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms,
-for instance, the A2C algorithm originally described in @A2C.
-Here, we train an agent using the A2C algorithm using default hyper-parameter settings with a multi-layer perceptron policy and value network composed of two 64-neuron layers.
+Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
+For example, here we will try the Proximal Policy Optimization (PPO), originally described in @PPO.
+Here, we train an agent using the PPO algorithm using default hyper-parameter settings with a multi-layer perceptron ('MlpPolicy') policy network and value network composed of two 64-neuron layers.
 
 ```{r}
-a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
+ppo = sb3$PPO('MlpPolicy', env, seed = seed) 
 ```
@@
 ```{r}
-if(!file.exists("cache/a2c.zip")) {
+if(!file.exists("../python/cache/ppo.zip")) {
 
   # Train the agent for a fixed number of timesteps
-  a2c$learn(total_timesteps=300000L)
+  ppo$learn(total_timesteps=300000L)
 
   # Save our trained agent for future use
-  a2c$save("cache/a2c")
+  ppo$save("cache/ppo")
 }
@@
 # Simulate management under the trained agent
-a2c = sb3$A2C$load("cache/a2c")
+ppo = sb3$PPO$load("cache/ppo")
 ```
@@
 
-  out = a2c$predict(state)
+  out = ppo$predict(state, deterministic=TRUE)
   action = out[[1]]
@@
 ```
+
+## Improving performance by tuning:
 

### 2021-06-04 python/fishing_SAC.py

@@ -1,50 +1,0 @@
-# Fishing using tuned SB3
-
-import numpy as np
-import os
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-from stable_baselines3.common.monitor import Monitor
-from stable_baselines3.common.evaluation import evaluate_policy
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-
-# Create an agent
-hyper = {
-    "gamma": 0.95,
-    "learning_rate": 0.000115,
-    "batch_size": 128,
-    "buffer_size": 100000,
-    "learning_starts": 0,
-    "train_freq": 128,
-    "tau": 0.01,
-    "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
-}
-
-agent = sb3.SAC("MlpPolicy", env, seed=seed, **hyper)
-
-# Train the agent
-if not os.path.exists("cache/sac_tuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/sac_tuned")
-
-
-agent = sb3.SAC.load("cache/sac_tuned")
-
-# Evaluate the trained agent
-eval_env = Monitor(gym.make("fishing-v1"))
-mean_reward, std_reward = evaluate_policy(agent, eval_env, n_eval_episodes=100)
-print("mean reward:", mean_reward, "std:", std_reward)
-
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=5)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_sac_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_sac_tuned_policy.png")

### 2021-06-04 R/plotting-helpers.R
plotting helpers

@@ -1,0 +1,37 @@
+# Standard R methods are used to plot the results summarized across the replicates:
+
+library(tidyverse)
+
+ymin = function(x) last(x[(ntile(x, 20)==1)])
+ymax = function(x) last(x[(ntile(x, 20)==19)])
+
+plot_sims <- function(sims_df)
+  sims_df %>%
+  group_by(time, model) %>%
+  summarise(ymin = ymin(state),
+            ymax = ymax(state),
+            state = mean(state), .groups = "drop") %>%
+  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) +
+  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
+
+plot_policy <- function(policy_df)
+  policy_df %>% ggplot(aes(state, action,
+                           group=interaction(rep, model),
+                           col = model)) +
+  geom_line(show.legend = FALSE) +
+  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9))
+
+plot_reward <- function(sims_df, gamma = 1)
+  sims_df %>%
+  group_by(rep, model) %>%
+  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
+  group_by(time, model) %>%
+  summarise(mean_reward = mean(cum_reward),
+            sd = sd(cum_reward), .groups = "drop") %>%
+  ggplot(aes(time, mean_reward)) +
+  geom_ribbon(aes(ymin = mean_reward - 2*sd,
+                  ymax = mean_reward + 2*sd, fill = model),
+              alpha=0.25, show.legend = FALSE) +
+  geom_line(aes(col = model), show.legend = FALSE) +
+  ylab("reward")
+

### 2021-06-04 appendices/appendix-B-examples.Rmd

--- 2021-05-27 appendices/appendix-B-examples.Rmd
@@
 
+# Overview
+
+In this appendix, we provide examples with annotated code illustrating how to use recent deep RL algorithms
+to train RL agents to make decisions in two distinct ecological management scenarios: setting a harvest quota in a fishery, and managing conservation effort in the face of ecological tipping points.
+The approach illustrated here should be sufficient to reproduce the results presented in the main text,
+and should serve as an effective starting point for researchers seeking to apply deep RL to these simulation environments, 
+to other simulation environments we have built and continue to develop, or to their own environments.
+(Appendix C details the process of defining one's own `gym` environment from an ecological model). 
+This overview is neither a complete introduction to deep RL or a comprehensive analysis of RL solutions to these environments.
+
 ```{r knitr, include = FALSE}
@@
 
+# Deep Reinforcement Learning in R
+
+Although all the necessary tooling for deep RL is most extensively implemented in Python, the R language is more familiar to most ecologists.
+Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
+In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
+
+In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
+This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
+Alternately, users may prefer running the analysis inside a docker container.  
+The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
+This can be particularly useful for users running ML algorithms on remote servers.
+
+By using a cache of pre-trained agents distributed with this appendix, it should be possible to efficiently reproduce the results shown here on any recent laptop with the versions of software indicated.
+Reproducing the training of RL agents is more computationally intensive, and a more powerful multi-core CPU or CUDA-enabled GPU is recommended.
+Note that hardware differences are also a source of instability in the reproduciblity of RL training (see main text and Appendix A).
+
 In this appendix, we illustrate how to apply a deep reinfocement learning framework to train previously published RL algorithms on the novel decision environments as illustrated in the main text.
@@
 
-# Deep Reinforcement Learning in R
-
-Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
-Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
-In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
-
-In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
-This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
-Alternately, users may prefer running the analysis inside a docker container.  
-The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
-This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.
+# Getting started
 
 Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
-Note that the RMarkdown source-code for this file can be found in the `appendices` directory of the project repository.
+
 From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
@@
 #install.packages("renv")
-renv::restore()
+#renv::restore()
 ```
 
+Python users can install the dependencies listed in the repository's `requirements.txt` and run the corresponding python scripts found in the `python/` sub-directory instead. 
+Meanwhile, the RMarkdown source-code for this file can be found in the `appendices` directory of the project repository.
 Once the packages have installed, we are ready to load the necessary libraries.
@@
 ## Python dependencies loaded via R
-sb3         = import ("stable_baselines3")
-gym         = import ("gym")
-gym_fishing = import("gym_fishing")
-gym_conservation = import("gym_conservation")
+sb3      = import("stable_baselines3")
+torch    = import("torch")
+gym      = import("gym")
+gym_fish = import("gym_fishing")
+gym_cons = import("gym_conservation")
 
-# 
 source("../R/plotting-helpers.R")
@@
 ```{r}
-algos <- c("Optimal", "A2C", "TD3", "TD3_tuned")
+algos <- c("Optimal", "PPO", "PPO_tuned")
 sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
@@
 ```{r}
-if(!file.exists("../python/cache/a2c.zip")) {
+if(!file.exists("../python/cache/ppo.zip")) {
 
   # Train the agent for a fixed number of timesteps
-  a2c$learn(total_timesteps=300000L)
+  ppo$learn(total_timesteps=300000L)
 
   # Save our trained agent for future use
-  a2c$save("../python/cache/a2c")
+  ppo$save("../python/cache/ppo")
+
 }
@@
 This is discussed further below.
-Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.
+Having saved our trained agent to disk (`cache/ppo.zip`), we can then re-load this agent for evaluation or to continue training.
 Note that a copy of the trained agents are included in the corresponding GitHub repository.
@@
 # Simulate management under the trained agent
-a2c = sb3$A2C$load("../python/cache/a2c")
+ppo = sb3$PPO$load("../python/cache/ppo")
 ```
@@
 ```{r}
-
 ## represent the initial state size in the 'rescaled' state space.
@@
 ```{r}
-a2c_sims = env$simulate(a2c, reps = 100L)
+ppo_sims = env$simulate(ppo, reps = 100L)
 ```
@@
 ```{r}
-a2c_policy = env$policyfn(a2c, reps = 5L)
+ppo_policy = env$policyfn(ppo, reps = 5L)
 ```
@@
 ```{r}
-sims_df <- a2c_sims %>% mutate(model = factor("A2C", algos)) %>% bind_rows(sims_df)
-policy_df <- a2c_policy  %>% mutate(model = factor("A2C", algos))  %>% bind_rows(policy_df)
+# stack result simulations with those above and plot:
+sims_df <- ppo_sims %>% 
+  mutate(model = factor("PPO", algos)) %>% 
+  bind_rows(sims_df)
+policy_df <- ppo_policy  %>% 
+  mutate(model = factor("PPO", algos))  %>% 
+  bind_rows(policy_df)
 
@@
 
-## TD3
+Note that the un-tuned version of PPO has found a viable policy which slowly increases its total reward over the course of the simulation,
+but nevertheless is substantially sub-optimal in its performance.
+Under the management of this agent, stock sizes average much higher than $B_{MSY}$ level of the optimal solution.
+The stock sizes also fluctuate as the agent constantly adjusts the harvest quota in response to the previous level.
 
-We train a second agent based on the TD3 algorithm.
-Recall that the Twin Delayed Deep Deterministic Policy Gradient, commonly known as TD3 [@TD3], is a generalization of Deep Deterministic Policy Gradient, or DDPG, which extends the DQN algorithm @DQN to a continuous action space using the deterministic policy gradient.
-DDPG and DQN are also available in the stable baselines framework [@sb3].
-One of the most salient differences here is that A2C uses an Actor-Critic strategy that combines policy gradient and value iteration approaches, while TD3 is a policy gradient approach.
+## Improving performance by tuning:
 
-We will start with the default hyperparameters, which correspond to those used in @TD3, where the algorithm was first introduced.
+Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
+For simplicity, we show only training with specified hyper-parameters here, found by various tuning experimentation (see <https://github.com/boettiger-lab/conservation-agents>) or the standard reference examples presented in `stable-baselines3-zoo` [@zoo].
 
 ```{r}
-td3_untuned = sb3$TD3('MlpPolicy', env, seed = seed)
+env = sb3$common$env_util$make_vec_env("fishing-v1", n_envs=4L, seed=seed)
+
+# Create a fishing environment
+# Create an agent
+ppo_tuned = sb3$PPO("MlpPolicy",
+                    "env" = env, 
+                    "seed" = seed,
+                    "batch_size" = 64L,
+                    "n_steps" = 32L,
+                    "gamma" = 0.9999,
+                    "learning_rate" = 0.00038,
+                    "ent_coef" = 0.0,
+                    "clip_range" = 0.3,
+                    "n_epochs" = 10L,
+                    "gae_lambda" = 0.92,
+                    "max_grad_norm" = 0.7,
+                    "vf_coef" = 0.486355,
+                    "policy_kwargs" = list(
+                      "net_arch" = c(64L, 64L),
+                      "activation_fn" = torch$nn$ReLU
+                     )
+                   )
+
 ```
 
-Once again, after our agent is defined we are ready to train it:
+We train this new PPO-based agent as before:
 
 ```{r }
-if(!file.exists("../python/cache/td3_untuned.zip")){
+if(!file.exists("../python/cache/ppo_tuned.zip")){
 
-  td3_untuned$learn(total_timesteps=300000L)
-  td3_untuned$save("../python/cache/td3_untuned")
-
-}
-```
-
-With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:
-
-```{r}
-# Simulate management under the trained agent
-td3_untuned = sb3$TD3$load("../python/cache/td3_untuned")
-td3_sims_untuned = env$simulate(td3_untuned, reps = 100L)
-td3_policy_untuned = env$policyfn(td3_untuned, reps = 5L)
-```
-
-```{r}
-sims_df <- td3_sims_untuned %>% mutate(model = factor("TD3", algos)) %>% bind_rows(sims_df)
-policy_df <- td3_policy_untuned  %>% mutate(model = factor("TD3", algos))  %>% bind_rows(policy_df)
-
-plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
-```
-The untuned TD3 policy tends to overfish relative to the optimal solution, but is less variable than the A2C solution.
-While it is possible that with much more training, this TD3 algorithm could further improve, subsequent learning may be difficult.  
-Note that the reward under TD3 is nevertheless often within the margin of error of the optimal policy, which may limit future learning since further policy exploration is more likely to reduce rewards.
-Instead, we may be able to improve performance of TD3 by choosing different hyperparameters, which may allow it to imagine more possible policies or learn more effectively.
-
-## Tuned TD3
-
-Some of these hyper-parameters of the TD3 algorithm, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm. 
-Others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
-Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
-For simplicity, we show only training with specified hyper-parameters here, found by various tuning experimentation (see <https://github.com/boettiger-lab/conservation-agents>) and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].
-
-```{r}
-# train an agent (model) on one of the environments:
-
-policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
-# non-episodic action noise:
-noise_std = 0.6656948079225263
-n_actions = env$action_space$shape[0]
-action_noise = sb3$common$noise$NormalActionNoise(
-        mean=np$zeros(n_actions),
-        sigma= noise_std * np$ones(n_actions)
-        )
-
-td3 = sb3$TD3( 'MlpPolicy', 
-               env,  
-               verbose=0L, 
-               seed = seed,
-               policy_kwargs = policy_kwargs,
-               learning_rate = 0.0001355522450968401,
-               gamma= 0.995, 
-               batch_size = 128L,
-               buffer_size = 10000L,
-               train_freq = 128L, 
-               gradient_steps = 128L,
-               action_noise = action_noise
-              )
-```
-
-We train this new TD3 agent:
-
-```{r }
-if(!file.exists("../python/cache/td3_tuned.zip")){
-
-  td3$learn(total_timesteps=300000L)
-  td3$save("../python/cache/td3_tuned")
+  ppo_tuned$learn(total_timesteps=300000L)
+  ppo_tuned$save("../python/cache/ppo_tuned")
 
@@
 # Simulate management under the trained agent
-td3 = sb3$TD3$load("../python/cache/td3_tuned")
-td3_sims = env$simulate(td3, reps = 100L)
-td3_policy = env$policyfn(td3, reps = 5L)
+ppo_tuned = sb3$PPO$load("../python/cache/ppo_tuned")
+ppo_tuned_sims = env$simulate(ppo_tuned, reps = 100L)
+ppo_tuned_policy = env$policyfn(ppo_tuned, reps = 5L)
 ```
@@
 ```{r}
-sims_df <- td3_sims %>% mutate(model = factor("TD3_tuned", algos)) %>% bind_rows(sims_df)
-policy_df <- td3_policy  %>% mutate(model = factor("TD3_tuned", algos))  %>% bind_rows(policy_df)
+# stack result simulations with those above and plot:
+sims_df <- ppo_tuned_sims %>% 
+  mutate(model = factor("PPO_tuned", algos)) %>% 
+  bind_rows(sims_df)
+policy_df <- ppo_tuned_policy  %>% 
+  mutate(model = factor("PPO_tuned", algos))  %>% 
+  bind_rows(policy_df)
 
@@
 
-The tuned TD3 algorithm, though trained for an identical 300,000 timesteps as the untined one, settles on a policy much closer to the optimal policy.
-The tuned TD3 agent tends to underfish on average, but only slightly, and avoids the occassional sharp drops in stock size seen when A2C occassionally sets too high a havest quota.
 The tuned solution is still not optimal, but it is very close.
@@
 env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-td3 = sb3$TD3$load("cache/td3")
+agent = sb3$PPO$load("../python/cache/ppo_tuned")
 
@@
 
-td3_state = td3_action = numeric(Tmax)
-td3_state[1] = x0
-td3_action[1] = NA
+state = action = numeric(Tmax)
+state[1] = x0
+action[1] = NA
 
@@
   # RL-recommended harvest action:
-  out = td3$predict(state)
+  out = agent$predict(state, determinstic=TRUE)
   action = out[[1]]
@@
   # Record state and resulting action
-  td3_state[i] = env$get_fish_population(state)
-  td3_action[i] = env$get_quota(action)
+  state[i] = env$get_fish_population(state)
+  action[i] = env$get_quota(action)
 
@@
 df = bind_rows(
-  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
+  tibble(year = years, state = state, action = action, model = "(tuned) PPO"),
   tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))

### 2021-06-04 appendices/appendix-B-examples.Rmd
instability...

@@
 Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
-For simplicity, we show only training with specified hyper-parameters here, found by various tuning experimentation (see <https://github.com/boettiger-lab/conservation-agents>) or the standard reference examples presented in `stable-baselines3-zoo` [@zoo].
+In practice, tuning may attempt to set only a subset of parameters through an optimal search algorithm, while others are chosen manually.
+Because each iteration in tuning must retrain the algorithm over thousands or millions of episodes, tuning can easily become very computationally intensive.
+For simplicity, here we illustrate how to train PPO using custom hyperparameters.
+The PPO algorithm supports training on a vectorized environment, in which the algorithm learns by interacting with `n` independent copies of the environment simultaneously.
+The choice of `n` can itself be considered another hyperparameter.
+Additional choices for many of the hyperparameters used by the PPO model are specified as additional optional arguments.
+In this case, the parameter choices shown were already derived through tuning (see <https://github.com/boettiger-lab/conservation-agents>, or the standard reference examples presented in `stable-baselines3-zoo` [@zoo]).
+More generally, these choices could represent any alternative trial of possible hyperparameters.
 
 ```{r}
-env = sb3$common$env_util$make_vec_env("fishing-v1", n_envs=4L, seed=seed)
+vec_env = sb3$common$env_util$make_vec_env("fishing-v1", n_envs=4L, seed=seed)
 
@@
 ppo_tuned = sb3$PPO("MlpPolicy",
-                    "env" = env, 
+                    "env" = vec_env, 
                     "seed" = seed,
@@
                     "gamma" = 0.9999,
-                    "learning_rate" = 0.00038,
-                    "ent_coef" = 0.0,
+                    "learning_rate" = 0.00037996229760279524,
+                    "ent_coef" = 2.392750198613048e-08,
                     "clip_range" = 0.3,
@@
                     "max_grad_norm" = 0.7,
-                    "vf_coef" = 0.486355,
+                    "vf_coef" = 0.4863551514846155,
                     "policy_kwargs" = list(
@@
 
-We train this new PPO-based agent as before:
+We train this new PPO-based agent for 300,000 steps, as before:
 

### 2021-06-04 python/fishing_PPO_tuned.py

@@ -16,18 +16,19 @@ hyper = {
     "batch_size": 64,
     "n_steps": 32,
     "gamma": 0.9999,
-    "learning_rate": 0.00038,
-    "ent_coef": 0.0,
+    "learning_rate": 0.00037996229760279524,
+    "ent_coef": 2.392750198613048e-08,
     "clip_range": 0.3,
     "n_epochs": 10,
     "gae_lambda": 0.92,
     "max_grad_norm": 0.7,
-    "vf_coef": 0.486355,
+    "vf_coef": 0.4863551514846155,
      "policy_kwargs": {
         "net_arch": [64, 64],
         "activation_fn": nn.ReLU,
     }
 }
+
 # Create an agent
 agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
 # Train the agent

### 2021-06-04 appendices/appendix-B-examples.Rmd

--- 2021-06-04 appendices/appendix-B-examples.Rmd
@@
 
-# Overview
-
-In this appendix, we provide examples with annotated code illustrating how to use recent deep RL algorithms
-to train RL agents to make decisions in two distinct ecological management scenarios: setting a harvest quota in a fishery, and managing conservation effort in the face of ecological tipping points.
-The approach illustrated here should be sufficient to reproduce the results presented in the main text,
-and should serve as an effective starting point for researchers seeking to apply deep RL to these simulation environments, 
-to other simulation environments we have built and continue to develop, or to their own environments.
-(Appendix C details the process of defining one's own `gym` environment from an ecological model). 
-This overview is neither a complete introduction to deep RL or a comprehensive analysis of RL solutions to these environments.
-
 ```{r knitr, include = FALSE}
@@
 
-# Deep Reinforcement Learning in R
+# Building a simulation environment `gym`
 
-Although all the necessary tooling for deep RL is most extensively implemented in Python, the R language is more familiar to most ecologists.
-Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
-In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
-
-In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
-This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
-Alternately, users may prefer running the analysis inside a docker container.  
-The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
-This can be particularly useful for users running ML algorithms on remote servers.
-
-By using a cache of pre-trained agents distributed with this appendix, it should be possible to efficiently reproduce the results shown here on any recent laptop with the versions of software indicated.
-Reproducing the training of RL agents is more computationally intensive, and a more powerful multi-core CPU or CUDA-enabled GPU is recommended.
-Note that hardware differences are also a source of instability in the reproduciblity of RL training (see main text and Appendix A).
+[@openai_gym]
 
@@
 
-Rather than implement deep RL algorithms from scratch, we illustrate the use of one of the leading software frameworks for research and application of deep RL methods.
-At the time of writing, several major frameworks exist which provide reference implementations of the leading deep RL algorithms (see table 1 and Appendix A).
+At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
 While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
@@
 
-# Getting started
+# Deep Reinforcement Learning in R
+
+Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
+Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
+In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.
+
+In the R-based approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
+This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
+Alternately, users may prefer running the analysis inside a docker container.  
+The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
+This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.
 
@@
 From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.
-(Python users can install the dependencies listed in the repository's `requirements.txt` and run the corresponding python scripts found in the `python/` subidrectory instead).
 
@@
 
-We begin by selecting an environment from `gym_fishing`.
-While we use default parameter values for this example, note that is possible to pass alternative parameterizations using additional optional arguments.
+We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.
 
@@
 ## initialize the environment
-env = gym$make("fishing-v1")
+env = gym$make("fishing-v1", sigma = 0.1)
 ```
@@
 Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
-Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
-For example, here we will try the Proximal Policy Optimization (PPO), originally described in @PPO.
-Here, we train an agent using the PPO algorithm using default hyper-parameter settings with a multi-layer perceptron ('MlpPolicy') policy network and value network composed of two 64-neuron layers.
+Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms,
+for instance, the A2C algorithm originally described in @A2C.
+Here, we train an agent using the A2C algorithm using default hyper-parameter settings with a multi-layer perceptron policy and value network composed of two 64-neuron layers.
 
 ```{r}
-ppo = sb3$PPO('MlpPolicy', env, seed = seed) 
+a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
 ```
@@
 ```{r}
-if(!file.exists("../python/cache/ppo.zip")) {
+if(!file.exists("cache/a2c.zip")) {
 
   # Train the agent for a fixed number of timesteps
-  ppo$learn(total_timesteps=300000L)
+  a2c$learn(total_timesteps=300000L)
 
   # Save our trained agent for future use
-  ppo$save("cache/ppo")
+  a2c$save("cache/a2c")
 }
@@
 # Simulate management under the trained agent
-ppo = sb3$PPO$load("cache/ppo")
+a2c = sb3$A2C$load("cache/a2c")
 ```
@@
 Because these methods return additional information as well, a little extra sub-setting is required in R:
-
+<!-- ML: Given recent excursions, we'll probably want to put a deterministic=True flag in here and following sections; also potentially problematic here if the agent crashes the population and restarts.
+-->
 ```{r}
@@
 
-  out = ppo$predict(state, deterministic=TRUE)
+  out = a2c$predict(state)
   action = out[[1]]
@@
 To get a better idea of how the agent behaves across the full state space, it is common to examine the policy function: to do this, we'll plot the action taken at every possible state.
-Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
+Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior. <!-- ML: This might be interesting -->
 Note that the `policyfn` method is another custom method and not part of the `gym` standard.
@@
 ```
-
-## Improving performance by tuning:
 

### 2021-06-04 appendices/appendix-C-environments.Rmd

@@
 
-In Deep Reinforcement Learning (RL), and _environment_ is a mathematical model, computer simulation, or real world situation with which the agent interacts.
+In Deep Reinforcement Learning (DRL), an _environment_ is a mathematical model, computer simulation, or real world situation with which the agent interacts.
 Defining an _environment_  requires 5 components:
@@
 2. Describe the possible actions available to the agent
-3. Update it's internal _state_ in response to the _action_ of the _agent_
+3. Update the environment's internal _state_ in response to the _action_ of the _agent_
 4. Provide the agent with an _observation_ of the new state 
-5. Provide the agent with a _reward_ based on the its state and the agent's action.
+5. Provide the agent with a _reward_ based on the environment's state and the agent's action.
 
@@
 
-This corresponds to the formal definition of a Partially Observed Markov Decision Process (POMDP), see main text and Appendix A.
+This corresponds to the formal definition of a Partially Observed Markov Decision Process (POMDP), see main text.
 

### 2021-06-04 manuscript/manuscript.Rmd
Small edit to the comment on fig1

@@
 Hyperparameter tuning consists of iterative \emph{trials}, in which the agent is trained with different sets of hyperparameters.
-At the end of a trial, the agent is evaluated to see which agent attains the highest cumulative reward.
+At the end of a trial, the agent is evaluated to see which set of hyperparameters results in the highest cumulative reward.
 An agent is \emph{evaluated} by recording the cumulative reward over one episode, or the mean reward over multiple episodes.
@@
 
-<!-- ML: Need to rework this to talk about other environments that we are working on basically point to conservation gym website -->
-One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
-While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
-Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
-Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
-Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+One feature that most distinguishes deep RL from other approaches to decision theory (such as optimal control) is its flexibility and scope.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, deep RL is scoped more ambitiously.
+Large, complex problems are still challenging to RL, and existing RL approaches can easily struggle to solve problems that can already be solved much more efficiently by other means.
+Nevertheless, deep RL approaches have proven successful on problems that classical approaches could not solve despite immense computational resources -- e.g. deep RL  [@DQN, @alphago].
+Our previous examples have focused on relatively simple problems that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+Yet, we also have developed
 

### 2021-06-04 python/fishing_PPO_tuned.py
conditional logic

@@ -2,7 +2,7 @@ import numpy as np
 import stable_baselines3 as sb3
 from stable_baselines3.common.env_util import make_vec_env
 from torch import nn as nn
-
+import os
 import gym
 import gym_fishing
 
@@ -32,8 +32,9 @@ hyper = {
 # Create an agent
 agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
 # Train the agent
-agent.learn(total_timesteps=300000)
-agent.save("cache/PPO_tuned")
+if not os.path.exists("cache/PPO_tuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/PPO_tuned")
 
 # Evaluate the trained agent
 env = gym.make("fishing-v1")

### 2021-06-04 appendices/appendix-B-examples.Rmd
fix import

--- 2021-06-04 appendices/appendix-B-examples.Rmd
@@
 gym      = import("gym")
-gym_fish = import("gym_fishing")
-gym_cons = import("gym_conservation")
+gym_fishing = import("gym_fishing")
+gym_conservation = import("gym_conservation")
 
@@
 
-Our hindcast considers the scenario of managing catch quotas using TD3, beginning in 1986. 
+Our hindcast considers the scenario of managing catch quotas using the RL agent, beginning in 1986. 
 We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
-This shows that quotas set by TD3 would have been lower than historical values throughout the late 80s and early 90s, which
+This shows that quotas set by the RL agent would have been lower than historical values throughout the late 80s and early 90s, which
 saw a sharp decline in estimated biomass of the Argentine Hake.
-For the same or higher estimated stock sizes from the historical record, we see that TD3 would recommend a lower quota.
-We also see that under TD3 the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
+For the same or higher estimated stock sizes from the historical record, we see that the RL agent would recommend a lower quota.
+We also see that under the RL agent the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
 being set in the now-depleted stock.  
-Whether the TD3 quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
+Whether the RL agent quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
 However, the fact that TD3 would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.

### 2021-06-04 manuscript/manuscript.Rmd
First pass

@@
 
-One feature that most distinguishes deep RL from other approaches to decision theory (such as optimal control) is its flexibility and scope.
-While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, deep RL is scoped more ambitiously.
-Large, complex problems are still challenging to RL, and existing RL approaches can easily struggle to solve problems that can already be solved much more efficiently by other means.
-Nevertheless, deep RL approaches have proven successful on problems that classical approaches could not solve despite immense computational resources -- e.g. deep RL  [@DQN, @alphago].
-Our previous examples have focused on relatively simple problems that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-Yet, we also have developed
+The feature that most distinguishes deep RL from other approaches to decision theory (such as optimal control) is its scope.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, deep RL can work on a much broader range of problems.
+Deep RL algorithms have proven successful on a range of problems that classical approaches could not solve despite immense computational resources -- e.g.
+deep RL can solve control problems with image-based observations that would be intractable with classical methods [@DQN]--; yet, getting deep RL to work in practice can be unexpectedly challenging and computationally intensive.
+Our previous examples have thus focused on relatively simple problems that can be addressed by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+We have developed more conservation-based RL environments that range in difficulty to solve; an active list of these environments can be found at <https://boettiger-lab.github.io/conservation-gym/>.
 

### 2021-06-04 manuscript/manuscript.Rmd
Condensing

@@
 Our previous examples have thus focused on relatively simple problems that can be addressed by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-We have developed more conservation-based RL environments that range in difficulty to solve; an active list of these environments can be found at <https://boettiger-lab.github.io/conservation-gym/>.
-
-<!-- Marcus add ~ paragraph discussion of the fire gym, citing motivating examples, basic overview of the state space, action space and reward and a screenshot of the simulation as a figure? -->
+We have developed more conservation-based RL environments, some of which pose much more challenging control problems; an active list of these environments can be found at <https://boettiger-lab.github.io/conservation-gym/>.
 

### 2021-06-04 manuscript/manuscript.Rmd
Hopefully this catches all the changes
Changing intro sentence on additional envs para

@@
 
-## Unsolved environments: Wildfires and other examples
+## Additional environments
 
-The feature that most distinguishes deep RL from other approaches to decision theory (such as optimal control) is its scope.
-While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, deep RL can work on a much broader range of problems.
 Deep RL algorithms have proven successful on a range of problems that classical approaches could not solve despite immense computational resources -- e.g.
@@
 Our previous examples have thus focused on relatively simple problems that can be addressed by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-We have developed more conservation-based RL environments, some of which pose much more challenging control problems; an active list of these environments can be found at <https://boettiger-lab.github.io/conservation-gym/>.
+We have developed more OpenAI `gym`'s, which deal with a range of ecological control problems and pose a wider spectrum of difficulty; an active list of these gyms can be found at <https://boettiger-lab.github.io/conservation-gym/>.
+Examples of these additional gyms include a wildfire gym and an epidemic gym.
+Further, with the fishing and conservation environments we've detailed in this study, we've defined supplementary fishing and conservation environments that implement a variety of models that present more challenging control problems.
 

### 2021-06-04 manuscript/manuscript.Rmd
Small word choice change

@@
 Examples of these additional gyms include a wildfire gym and an epidemic gym.
-Further, with the fishing and conservation environments we've detailed in this study, we've defined supplementary fishing and conservation environments that implement a variety of models that present more challenging control problems.
+Further, with the fishing and conservation environments we've detailed in this study, we've defined supplementary fishing and conservation environments that implement a variety of models that present more difficult control problems.
 

### 2021-06-04 manuscript/manuscript.Rmd
Small grammar change

@@
 Our previous examples have thus focused on relatively simple problems that can be addressed by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-We have developed more OpenAI `gym`'s, which deal with a range of ecological control problems and pose a wider spectrum of difficulty; an active list of these gyms can be found at <https://boettiger-lab.github.io/conservation-gym/>.
+We have developed more OpenAI `gym`s, which deal with a range of ecological control problems and pose a wider spectrum of difficulty; an active list of these gyms can be found at <https://boettiger-lab.github.io/conservation-gym/>.
 Examples of these additional gyms include a wildfire gym and an epidemic gym.

### 2021-06-04 appendices/appendix-B-examples.Rmd
typo

@@
                     "vf_coef" = 0.4863551514846155,
+                    "use_sde" = TRUE,
                     "policy_kwargs" = list(
-                      "net_arch" = c(64L, 64L),
+                      "net_arch" = list(list(pi=c(64L, 64L), vf=c(64L, 64L))),
                       "activation_fn" = torch$nn$ReLU
@@
   # RL-recommended harvest action:
-  out = agent$predict(state, determinstic=TRUE)
+  out = agent$predict(state, deterministic = TRUE)
   action = out[[1]]

### 2021-06-04 manuscript/manuscript.Rmd
Almost good on add. envs
Revising the add. envs. section

@@
 
-## Sustainable harvest
+## Sustainable harvest \label{sec:fishery}
 
@@
 
-## Ecological Tipping Points
+## Ecological Tipping Points \label{sec:tipping}
 
@@
 
-## Additional environments
+## Additional Environments
 
-Deep RL algorithms have proven successful on a range of problems that classical approaches could not solve despite immense computational resources -- e.g.
-deep RL can solve control problems with image-based observations that would be intractable with classical methods [@DQN]--; yet, getting deep RL to work in practice can be unexpectedly challenging and computationally intensive.
-Our previous examples have thus focused on relatively simple problems that can be addressed by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-We have developed more OpenAI `gym`s, which deal with a range of ecological control problems and pose a wider spectrum of difficulty; an active list of these gyms can be found at <https://boettiger-lab.github.io/conservation-gym/>.
-Examples of these additional gyms include a wildfire gym and an epidemic gym.
-Further, with the fishing and conservation environments we've detailed in this study, we've defined supplementary fishing and conservation environments that implement a variety of models that present more difficult control problems.
+Ecology holds many open problems for deep RL.  
+It is straight-forward to extend the simple environments presented here to reflect greater biological complexity, or to consider a host of more realistic decision scenarios.
+We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
+Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
 

### 2021-06-04 appendices/appendix-B-examples.Rmd
sp
variants...

--- 2021-06-04 appendices/appendix-B-examples.Rmd
@@
 ## initialize the environment
-env = gym$make("fishing-v1")
+env = gym$make("fishing-v1", sigma=0.1)
 ```
@@
 ```{r}
-algos <- c("Optimal", "PPO", "PPO_tuned")
+# add a column indicating the model
+algos <- c("Optimal", "A2C")
 sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
@@
 
+# plot results as separate panels
 plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
@@
 Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
-For example, here we will try the Proximal Policy Optimization (PPO), originally described in @PPO.
-Here, we train an agent using the PPO algorithm using default hyper-parameter settings with a multi-layer perceptron ('MlpPolicy') policy network and value network composed of two 64-neuron layers.
+For example, here we will try the A2C actor-critic algorithm, an extension of the A3C algorithm originally described in @A2C.
+We train an agent using the A2C algorithm using default hyper-parameter settings (which correspond to those used in @A2C),
+using a multi-layer perceptron ('MlpPolicy') using the default architecture for the policy network and value network composed of two 64-neuron layers.
 
 ```{r}
-ppo = sb3$PPO('MlpPolicy', env, seed = seed) 
+a2c = sb3$A2C('MlpPolicy', env, seed = seed) 
 ```
@@
 ```{r}
-if(!file.exists("../python/cache/ppo.zip")) {
+if(!file.exists("../python/cache/a2c.zip")) {
 
   # Train the agent for a fixed number of timesteps
-  ppo$learn(total_timesteps=300000L)
+  a2c$learn(total_timesteps=300000L)
 
   # Save our trained agent for future use
-  ppo$save("../python/cache/ppo")
+  a2c$save("../python/cache/a2c")
 
@@
 This is discussed further below.
-Having saved our trained agent to disk (`cache/ppo.zip`), we can then re-load this agent for evaluation or to continue training.
+Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.
 Note that a copy of the trained agents are included in the corresponding GitHub repository.
@@
 # Simulate management under the trained agent
-ppo = sb3$PPO$load("../python/cache/ppo")
+a2c = sb3$A2C$load("../python/cache/a2c")
 ```
@@
 With an initial state in hand, we are ready to simulate management using our agent.
-The iteration is simple: we use the agent to predict what action we should take given the current state. Then, we take said action and examine the result to determine the future state. 
+The iteration is simple: we use the agent to predict what action we should take given the current state. 
+Then, we take said action and examine the result to determine the future state. 
 Because these methods return additional information as well, a little extra sub-setting is required in R:
@@
 
-  out = ppo$predict(state, deterministic=TRUE)
+  out = a2c$predict(state, deterministic=TRUE)
   action = out[[1]]
@@
 ```{r}
-ppo_sims = env$simulate(ppo, reps = 100L)
+a2c_sims = env$simulate(a2c, reps = 100L)
 ```
@@
 ```{r}
-ppo_policy = env$policyfn(ppo, reps = 5L)
+a2c_policy = env$policyfn(a2c, reps = 5L)
 ```
@@
 # stack result simulations with those above and plot:
-sims_df <- ppo_sims %>% 
-  mutate(model = factor("PPO", algos)) %>% 
+sims_df <- a2c_sims %>% 
+  mutate(model = factor("A2C", algos)) %>% 
   bind_rows(sims_df)
-policy_df <- ppo_policy  %>% 
-  mutate(model = factor("PPO", algos))  %>% 
+policy_df <- a2c_policy  %>% 
+  mutate(model = factor("A2C", algos))  %>% 
   bind_rows(policy_df)
@@
 
-Note that the un-tuned version of PPO has found a viable policy which slowly increases its total reward over the course of the simulation,
-but nevertheless is substantially sub-optimal in its performance.
-Under the management of this agent, stock sizes average much higher than $B_{MSY}$ level of the optimal solution.
-The stock sizes also fluctuate as the agent constantly adjusts the harvest quota in response to the previous level.
+After an initial over-harvest (top panel), A2C achieves nearly optimal performance as measured by reward over time, (bottom right), or compared to the steady-state management solution.
+The policy function (bottom left) explains this behavior.
+The RL agent has a nearly optimal policy for stock sizes between zero and just above the $B_{MSY}$ target.
+At higher stocks, the agent has a tendency to over-harvest.
 
-## Improving performance by tuning:
+In this simple example, A2C is able to achieve nearly-optimal performance (as defined by the reward conditions) without additional tuning.
+It is worth noting that this is not true of all RL algorithms.
+For example, the PPO algorithm [@A2C] is a somewhat more computationally intensive alternative that combines strategies from A2C and another algorithm, TRPO [@PPO].
+Nevertheless, using default hyper-parameters, PPO does considerably less well (not shown).
+For harder problems, it is often necessary to consider a range of algorithms, and tune the hyper-parameters extensively before training will be able to discover an acceptable solution.
+We illustrate this briefly in our second example gym.
 
-Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance.
-In practice, tuning may attempt to set only a subset of parameters through an optimal search algorithm, while others are chosen manually.
-Because each iteration in tuning must retrain the algorithm over thousands or millions of episodes, tuning can easily become very computationally intensive.
-For simplicity, here we illustrate how to train PPO using custom hyperparameters.
-The PPO algorithm supports training on a vectorized environment, in which the algorithm learns by interacting with `n` independent copies of the environment simultaneously.
-The choice of `n` can itself be considered another hyperparameter.
-Additional choices for many of the hyperparameters used by the PPO model are specified as additional optional arguments.
-In this case, the parameter choices shown were already derived through tuning (see <https://github.com/boettiger-lab/conservation-agents>, or the standard reference examples presented in `stable-baselines3-zoo` [@zoo]).
-More generally, these choices could represent any alternative trial of possible hyperparameters.
-
-```{r}
-vec_env = sb3$common$env_util$make_vec_env("fishing-v1", n_envs=4L, seed=seed)
-
-# Create a fishing environment
-# Create an agent
-ppo_tuned = sb3$PPO("MlpPolicy",
-                    "env" = vec_env, 
-                    "seed" = seed,
-                    "batch_size" = 64L,
-                    "n_steps" = 32L,
-                    "gamma" = 0.9999,
-                    "learning_rate" = 0.00037996229760279524,
-                    "ent_coef" = 2.392750198613048e-08,
-                    "clip_range" = 0.3,
-                    "n_epochs" = 10L,
-                    "gae_lambda" = 0.92,
-                    "max_grad_norm" = 0.7,
-                    "vf_coef" = 0.4863551514846155,
-                    "policy_kwargs" = list(
-                      "net_arch" = c(64L, 64L),
-                      "activation_fn" = torch$nn$ReLU
-                     )
-                   )
-
-```
-
-We train this new PPO-based agent for 300,000 steps, as before:
-
-```{r }
-if(!file.exists("../python/cache/ppo_tuned.zip")){
-
-  ppo_tuned$learn(total_timesteps=300000L)
-  ppo_tuned$save("../python/cache/ppo_tuned")
-
-}
-```
-
-With our trained agent, we can now simulate replicate scenarios or scan across states to determine the implicit policy function:
-
-```{r}
-# Simulate management under the trained agent
-ppo_tuned = sb3$PPO$load("../python/cache/ppo_tuned")
-ppo_tuned_sims = env$simulate(ppo_tuned, reps = 100L)
-ppo_tuned_policy = env$policyfn(ppo_tuned, reps = 5L)
-```
-
-```{r}
-# stack result simulations with those above and plot:
-sims_df <- ppo_tuned_sims %>% 
-  mutate(model = factor("PPO_tuned", algos)) %>% 
-  bind_rows(sims_df)
-policy_df <- ppo_tuned_policy  %>% 
-  mutate(model = factor("PPO_tuned", algos))  %>% 
-  bind_rows(policy_df)
-
-plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
-```
-
-The tuned solution is still not optimal, but it is very close.
-Moreover, the RL algorithm has been able to learn this nearly-optimal policy with no prior knowledge about the shape or nature of the population growth function used in the environment.
-
-We preserve a copy of each data frames in local `.csv` text files for maximum cross-platform compatibility.
-
-```{r}
+```{r include=FALSE}
+# Store data.frames for manuscript plotting
 write_csv(sims_df, "../manuscript/figs/sims_df.csv")
@@
 ```
-
-<!-- Remaining is just plotting code, repeated in the main .Rmd.  Should we omit it from the appendix? -->
 
@@
 # Simulate management under the trained agent
+# parameters based on statistical best model estimate, see Memarzadeh & Boettiger 2019
+# or Appendix D
 env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-agent = sb3$PPO$load("../python/cache/ppo_tuned")
+agent = sb3$A2C$load("../python/cache/a2c_untuned")
 
@@
 years =1986:2000
-
-state = action = numeric(Tmax)
-state[1] = x0
-action[1] = NA
-
+agent_state = agent_action = numeric(Tmax)
+agent_state[1] = x0
+agent_action[1] = NA
 ```
 
-Our hindcast considers the scenario of managing catch quotas using the RL agent, beginning in 1986. 
+Our hindcast considers the scenario of managing catch quotas using TD3, beginning in 1986. 
 We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
-This shows that quotas set by the RL agent would have been lower than historical values throughout the late 80s and early 90s, which
+This shows that quotas set by TD3 would have been lower than historical values throughout the late 80s and early 90s, which
 saw a sharp decline in estimated biomass of the Argentine Hake.
-For the same or higher estimated stock sizes from the historical record, we see that the RL agent would recommend a lower quota.
-We also see that under the RL agent the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
+For the same or higher estimated stock sizes from the historical record, we see that TD3 would recommend a lower quota.
+We also see that under TD3 the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
 being set in the now-depleted stock.  
-Whether the RL agent quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
+Whether the TD3 quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
 However, the fact that TD3 would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
@@
 state = env$get_state( x0 )
-
 for(i in 2:Tmax){
@@
   # RL-recommended harvest action:
-  out = agent$predict(state, determinstic=TRUE)
+  out = agent$predict(state)
   action = out[[1]]
@@
   # Record state and resulting action
-  state[i] = env$get_fish_population(state)
-  action[i] = env$get_quota(action)
+  agent_state[i] = env$get_fish_population(state)
+  agent_action[i] = env$get_quota(action)
 
@@
 }
-
 df = bind_rows(
-  tibble(year = years, state = state, action = action, model = "(tuned) PPO"),
+  tibble(year = years, state = agent_state, action = agent_action, model = "RL"),
   tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
-
 df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))
-
 ```

### 2021-06-04 python/fishing/determinstic/fishing_A2C_tuned.py

@@ -1,0 +1,54 @@
+# Fishing using (untuned) A2C
+
+import numpy as np
+from torch import nn as nn
+import stable_baselines3 as sb3
+import os
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+hyper = {
+    "gamma": 0.98,
+    "normalize_advantage": True,
+    "max_grad_norm": 0.5,
+    "use_rms_prop": False,
+    "gae_lambda": 0.92,
+    "n_steps": 16,
+    "learning_rate": 0.0014125669914905178,
+    "ent_coef": 0.000415417442958291,
+    "vf_coef": 0.7417182197443206,
+    "policy_kwargs": {
+        "ortho_init": False,
+        "net_arch": [64, 64],
+        "activation_fn": nn.ReLU,
+    }
+}
+
+# Create an A2C agent
+agent = sb3.A2C("MlpPolicy", env, seed=seed, **hyper)
+
+# Train
+if not os.path.exists("cache/a2c_tuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/a2c_tuned")
+
+# Load the saved agent
+agent = sb3.A2C.load("cache/a2c_tuned")
+
+# Evaluate
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
+print("mean score:", mean)
+
+# Plot results
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=1)
+env.plot(agent_sims, "results/fishing_a2c_tuned.png")
+env.plot_policy(agent_policy, "results/fishing_a2c_tuned_policy.png")
+
+

### 2021-06-04 python/fishing/determinstic/fishing_A2C_untuned.py

@@ -1,0 +1,33 @@
+# Fishing using (untuned) A2C
+
+import os
+import gym
+import gym_fishing
+import stable_baselines3 as sb3
+
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.0)
+
+# Create an A2C agent
+agent = sb3.A2C("MlpPolicy", env, seed=24)
+
+# Train
+if not os.path.exists("cache/a2c_untuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/a2c_untuned")
+
+# Load the saved agent
+agent = sb3.A2C.load("cache/a2c_untuned")
+
+# Evaluate
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
+print("mean score:", mean)
+
+# Plot results
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=1)
+env.plot(agent_sims, "results/fishing_a2c_untuned.png")
+env.plot_policy(agent_policy, "results/fishing_a2c_untuned_policy.png")
+
+

### 2021-06-04 python/fishing/determinstic/fishing_DDPG_untuned.py

@@ -1,0 +1,27 @@
+# Fishing using untuned ddpg
+
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.)
+# Create the agent
+ddpg = sb3.DDPG("MlpPolicy", env, seed=seed)
+
+# Train the agent
+ddpg.learn(total_timesteps=300000)
+ddpg.save("cache/ddpg_untuned")
+
+# Simulate management under the trained agent
+ddpg = sb3.DDPG.load("cache/ddpg_untuned")
+ddpg_sims = env.simulate(ddpg, reps=100)
+ddpg_policy = env.policyfn(ddpg, reps=5)
+
+# Example of built-in plotting methods
+env.plot(ddpg_sims, "results/fishing_ddpg_untuned.png")
+env.plot_policy(ddpg_policy, "results/fishing_ddpg_untuned_policy.png")

### 2021-06-04 python/fishing/determinstic/fishing_PPO_tuned.py

@@ -1,0 +1,45 @@
+import numpy as np
+import stable_baselines3 as sb3
+from stable_baselines3.common.env_util import make_vec_env
+from torch import nn as nn
+
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+env = make_vec_env("fishing-v1", n_envs=8, seed=seed, env_kwargs={"sigma": 0.})
+
+# Create a fishing environment
+hyper = {
+    "batch_size": 16,
+    "n_steps": 8,
+    "gamma": 0.99,
+    "learning_rate": 0.0012512477171887717,
+    "ent_coef": 6.908006009288375e-08,
+    "clip_range": 0.3,
+    "n_epochs": 1,
+    "gae_lambda": 0.9,
+    "max_grad_norm": 0.7,
+    "vf_coef": 0.7688840852349988,
+     "policy_kwargs": {
+        "net_arch": [256, 256],
+        "activation_fn": nn.ReLU,
+    }
+}
+# Create an agent
+agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
+# Train the agent
+agent.learn(total_timesteps=300000)
+agent.save("cache/PPO_tuned")
+
+# Evaluate the trained agent
+env = gym.make("fishing-v1", sigma=0.1)
+
+agent = sb3.PPO.load("cache/PPO_tuned")
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=5)
+
+# Plot results
+env.plot(agent_sims, "results/fishing_PPO_tuned.png")
+env.plot_policy(agent_policy, "results/fishing_PPO_tuned_policy.png")

### 2021-06-04 python/fishing/determinstic/fishing_SAC_tuned.py

@@ -1,0 +1,45 @@
+import os
+import gym
+import gym_fishing
+import stable_baselines3 as sb3
+
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+# Create an agent
+hyper = {
+    "gamma": 0.95,
+    "learning_rate": 0.000115,
+    "batch_size": 128,
+    "buffer_size": 100000,
+    "learning_starts": 0,
+    "train_freq": 128,
+    "tau": 0.01,
+    "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
+}
+
+agent = sb3.SAC("MlpPolicy", env, seed=24, **hyper)
+
+# Train the agent
+if not os.path.exists("cache/sac_tuned"):
+  agent.learn(total_timesteps=300000)
+  agent.save("cache/sac_tuned")
+
+
+agent = sb3.SAC.load("cache/sac_tuned")
+
+# Evaluate the trained agent
+eval_env = sb3.common.monitor.Monitor(gym.make("fishing-v1"))
+mean_reward, std_reward = evaluate_policy(agent, eval_env, n_eval_episodes=100)
+print("mean reward:", mean_reward, "std:", std_reward, "tuned_value:", 7.755079)
+
+
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=1)
+
+print("mean reward from sims:", np.sum(agent_sims["reward"]) / 100)
+
+# Plot results
+env.plot(agent_sims, "results/fishing_sac_tuned.png")
+env.plot_policy(agent_policy, "results/fishing_sac_tuned_policy.png")

### 2021-06-04 python/fishing/determinstic/fishing_SAC_untuned.py

@@ -1,0 +1,31 @@
+import os
+import gym
+import gym_fishing
+import stable_baselines3 as sb3
+
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+# Create an agent
+agent = sb3.SAC("MlpPolicy", env, seed=24)
+
+# Train the agent
+if not os.path.exists("cache/sac_untuned.zip"):
+  agent.learn(total_timesteps=300000)
+  agent.save("cache/sac_untuned")
+
+agent = sb3.SAC.load("cache/sac_untuned")
+
+# Evaluate
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, sb3.common.monitor.Monitor(env))
+print("mean score:", mean)
+
+# Plot results
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=2)
+env.plot(agent_sims, "results/fishing_sac_untuned.png")
+env.plot_policy(agent_policy, "results/fishing_sac_untuned_policy.png")
+
+
+

### 2021-06-04 python/fishing/determinstic/fishing_TD3_tuned.py

@@ -1,0 +1,43 @@
+# Fishing using tuned TD3
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+# Define the agent
+policy_kwargs = dict(net_arch=[400, 300])  # "big"
+noise_std = 0.6656948079225263
+action_noise = sb3.common.noise.NormalActionNoise(mean=0, sigma=noise_std)
+
+td3 = sb3.TD3(
+    "MlpPolicy",
+    env,
+    seed=seed,
+    policy_kwargs=policy_kwargs,
+    learning_rate=0.0001355522450968401,
+    gamma=0.995,
+    batch_size=128,
+    buffer_size=10000,
+    train_freq=(128, "step"),
+    gradient_steps=128,
+    action_noise=action_noise,
+)
+
+# Train the agent
+td3.learn(total_timesteps=300000)
+td3.save("cache/td3_tuned")
+
+
+# Simulate management under the trained agent
+td3 = sb3.TD3.load("cache/td3_tuned")
+td3_sims = env.simulate(td3, reps=100)
+td3_policy = env.policyfn(td3, reps=5)
+
+# Example of built-in plotting methods
+env.plot(td3_sims, "results/fishing_td3_tuned.png")
+env.plot_policy(td3_policy, "results/fishing_td3_tuned_policy.png")
+

### 2021-06-04 python/fishing/determinstic/fishing_TD3_untuned.py

@@ -1,0 +1,25 @@
+# Fishing using untuned TD3
+
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+# Create the agent
+td3 = sb3.TD3("MlpPolicy", env, seed=seed)
+
+# Train the agent
+td3.learn(total_timesteps=300000)
+td3.save("cache/td3_untuned")
+
+# Simulate management under the trained agent
+td3 = sb3.TD3.load("cache/td3_untuned")
+td3_sims = env.simulate(td3, reps=100)
+td3_policy = env.policyfn(td3, reps=5)
+
+# Example of built-in plotting methods
+env.plot(td3_sims, "results/fishing_td3_untuned.png")
+env.plot_policy(td3_policy, "results/fishing_td3_untuned_policy.png")

### 2021-06-04 python/fishing/stochastic/fishing_A2C_tuned.py

@@ -1,0 +1,54 @@
+# Fishing using (untuned) A2C
+
+import numpy as np
+from torch import nn as nn
+import stable_baselines3 as sb3
+import os
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.1)
+
+hyper = {
+    "gamma": 0.98,
+    "normalize_advantage": True,
+    "max_grad_norm": 0.5,
+    "use_rms_prop": False,
+    "gae_lambda": 0.92,
+    "n_steps": 16,
+    "learning_rate": 0.0014125669914905178,
+    "ent_coef": 0.000415417442958291,
+    "vf_coef": 0.7417182197443206,
+    "policy_kwargs": {
+        "ortho_init": False,
+        "net_arch": [64, 64],
+        "activation_fn": nn.ReLU,
+    }
+}
+
+# Create an A2C agent
+agent = sb3.A2C("MlpPolicy", env, seed=seed, **hyper)
+
+# Train
+if not os.path.exists("cache/a2c_tuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/a2c_tuned")
+
+# Load the saved agent
+agent = sb3.A2C.load("cache/a2c_tuned")
+
+# Evaluate
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
+print("mean score:", mean)
+
+# Plot results
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=1)
+env.plot(agent_sims, "results/fishing_a2c_tuned.png")
+env.plot_policy(agent_policy, "results/fishing_a2c_tuned_policy.png")
+
+

### 2021-06-04 python/fishing/stochastic/fishing_PPO_tuned.py

@@ -2,42 +2,39 @@ import numpy as np
 import stable_baselines3 as sb3
 from stable_baselines3.common.env_util import make_vec_env
 from torch import nn as nn
-import os
+
 import gym
 import gym_fishing
 
 seed = 24
 np.random.seed(seed)
-env = make_vec_env("fishing-v1", n_envs=4, seed=seed)
-
+env = make_vec_env("fishing-v1", n_envs=8, seed=seed, env_kwargs={"sigma": 0.1})
 
 # Create a fishing environment
 hyper = {
-    "batch_size": 64,
-    "n_steps": 32,
-    "gamma": 0.9999,
-    "learning_rate": 0.00037996229760279524,
-    "ent_coef": 2.392750198613048e-08,
+    "batch_size": 16,
+    "n_steps": 8,
+    "gamma": 0.99,
+    "learning_rate": 0.0012512477171887717,
+    "ent_coef": 6.908006009288375e-08,
     "clip_range": 0.3,
-    "n_epochs": 10,
-    "gae_lambda": 0.92,
+    "n_epochs": 1,
+    "gae_lambda": 0.9,
     "max_grad_norm": 0.7,
-    "vf_coef": 0.4863551514846155,
+    "vf_coef": 0.7688840852349988,
      "policy_kwargs": {
-        "net_arch": [64, 64],
+        "net_arch": [256, 256],
         "activation_fn": nn.ReLU,
     }
 }
-
 # Create an agent
 agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
 # Train the agent
-if not os.path.exists("cache/PPO_tuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/PPO_tuned")
+agent.learn(total_timesteps=300000)
+agent.save("cache/PPO_tuned")
 
 # Evaluate the trained agent
-env = gym.make("fishing-v1")
+env = gym.make("fishing-v1", sigma=0.1)
 
 agent = sb3.PPO.load("cache/PPO_tuned")
 agent_sims = env.simulate(agent, reps=100)

### 2021-06-04 python/fishing/stochastic/fishing_SAC_tuned.py

@@ -1,0 +1,46 @@
+# Fishing using tuned SB3
+import os
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.1)
+
+# Create an agent
+hyper = {
+    "gamma": 0.95,
+    "learning_rate": 0.000115,
+    "batch_size": 128,
+    "buffer_size": 100000,
+    "learning_starts": 0,
+    "train_freq": 128,
+    "tau": 0.01,
+    "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
+}
+
+agent = sb3.SAC("MlpPolicy", env, seed=seed, **hyper)
+
+# Train the agent
+if not os.path.exists("cache/sac_tuned,zip"):
+  agent.learn(total_timesteps=300000)
+  agent.save("cache/sac_tuned")
+
+
+agent = sb3.SAC.load("cache/sac_tuned")
+
+# Evaluate the trained agent
+eval_env = sb3.common.monitor.Monitor(gym.make("fishing-v1"))
+mean_reward, std_reward = evaluate_policy(agent, eval_env, n_eval_episodes=100)
+print("mean reward:", mean_reward, "std:", std_reward, "tuned_value:", 7.755079)
+
+
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=5)
+
+print("mean reward from sims:", np.sum(agent_sims["reward"]) / 100)
+
+# Plot results
+env.plot(agent_sims, "results/fishing_sac_tuned.png")
+env.plot_policy(agent_policy, "results/fishing_sac_tuned_policy.png")

### 2021-06-04 python/fishing/stochastic/fishing_SAC_untuned.py

@@ -1,0 +1,36 @@
+# Fishing using tuned TD3
+
+import stable_baselines3 as sb3
+import os
+import gym
+import gym_fishing
+
+
+# Create a fishing environment
+env = gym.make("fishing-v1")
+
+# Create an agent
+agent = sb3.SAC("MlpPolicy", env, seed=24)
+
+# Train the agent
+if not os.path.exists("cache/sac_untuned.zip"):
+  agent.learn(total_timesteps=300000)
+  agent.save("cache/sac_untuned")
+
+
+
+agent = sb3.SAC.load("cache/sac_untuned")
+
+# Evaluate
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
+print("mean score:", mean)
+
+
+# Plot results
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=2)
+env.plot(agent_sims, "results/fishing_sac_untuned.png")
+env.plot_policy(agent_policy, "results/fishing_sac_untuned_policy.png")
+
+
+

### 2021-06-04 python/fishing/stochastic/fishing_TD3_tuned.py

@@ -1,0 +1,46 @@
+# Fishing using tuned TD3
+
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.1)
+
+# Define the agent
+policy_kwargs = dict(net_arch=[400, 300])  # "big"
+noise_std = 0.6656948079225263
+action_noise = sb3.common.noise.NormalActionNoise(mean=0, sigma=noise_std)
+
+td3 = sb3.TD3(
+    "MlpPolicy",
+    env,
+    seed=seed,
+    policy_kwargs=policy_kwargs,
+    learning_rate=0.0001355522450968401,
+    gamma=0.995,
+    batch_size=128,
+    buffer_size=10000,
+    train_freq=(128, "step"),
+    gradient_steps=128,
+    action_noise=action_noise,
+)
+
+# Train the agent
+td3.learn(total_timesteps=300000)
+td3.save("cache/td3_tuned")
+
+
+# Simulate management under the trained agent
+td3 = sb3.TD3.load("cache/td3_tuned")
+td3_sims = env.simulate(td3, reps=100)
+td3_policy = env.policyfn(td3, reps=5)
+
+# Example of built-in plotting methods
+env.plot(td3_sims, "results/fishing_td3_tuned.png")
+env.plot_policy(td3_policy, "results/fishing_td3_tuned_policy.png")
+

### 2021-06-04 python/fishing/stochastic/fishing_TD3_untuned.py

@@ -1,0 +1,27 @@
+# Fishing using untuned TD3
+
+import numpy as np
+import stable_baselines3 as sb3
+import gym
+import gym_fishing
+
+seed = 24
+np.random.seed(seed)
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.)
+# Create the agent
+td3 = sb3.TD3("MlpPolicy", env, seed=seed)
+
+# Train the agent
+td3.learn(total_timesteps=300000)
+td3.save("cache/td3_untuned")
+
+# Simulate management under the trained agent
+td3 = sb3.TD3.load("cache/td3_untuned")
+td3_sims = env.simulate(td3, reps=100)
+td3_policy = env.policyfn(td3, reps=5)
+
+# Example of built-in plotting methods
+env.plot(td3_sims, "results/fishing_td3_untuned.png")
+env.plot_policy(td3_policy, "results/fishing_td3_untuned_policy.png")

### 2021-06-04 python/fishing_A2C_untuned.py

@@ -1,0 +1,34 @@
+# Fishing using (untuned) A2C
+
+import os
+import gym
+import gym_fishing
+import stable_baselines3 as sb3
+
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.1)
+
+# Create an A2C agent
+agent = sb3.A2C("MlpPolicy", env, seed=24)
+
+# Train
+if not os.path.exists("cache/a2c_untuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/a2c_untuned")
+
+# Load the saved agent
+agent = sb3.A2C.load("cache/a2c_untuned")
+
+# Evaluate
+eval_env = sb3.common.monitor.Monitor(env)
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, eval_env)
+print("mean score:", mean)
+
+# Plot results
+agent_sims = env.simulate(agent, reps=100)
+agent_policy = env.policyfn(agent, reps=1)
+env.plot(agent_sims, "results/fishing_a2c_untuned.png")
+env.plot_policy(agent_policy, "results/fishing_a2c_untuned_policy.png")
+
+

### 2021-06-05 appendices/appendix-B-examples.Rmd

@@
 
-In this appendix, we illustrate how to apply a deep reinfocement learning framework to train previously published RL algorithms on the novel decision environments as illustrated in the main text.
+In this appendix, we illustrate how to apply a deep reinforcement learning framework to train previously published RL algorithms on the novel decision environments as illustrated in the main text.
 Appendix C discusses the construction of such environments in code.

### 2021-06-05 python/fishing/determinstic/fishing_PPO_tuned.py
cirrus-based comparisons across algos on fishing_v1
stochastic and deterministic environments, tuned (on det environment only) and untuned agents.

@@ -1,14 +1,12 @@
-import numpy as np
 import stable_baselines3 as sb3
 from stable_baselines3.common.env_util import make_vec_env
 from torch import nn as nn
-
+import os
 import gym
 import gym_fishing
 
 seed = 24
-np.random.seed(seed)
-env = make_vec_env("fishing-v1", n_envs=8, seed=seed, env_kwargs={"sigma": 0.})
+env = make_vec_env("fishing-v1", n_envs=4, seed=seed, env_kwargs={"sigma": 0.})
 
 # Create a fishing environment
 hyper = {
@@ -36,7 +34,7 @@ if not os.path.exists("cache/ppo_tuned.zip"):
     agent.save("cache/ppo_tuned")
 
 # Evaluate the trained agent
-env = gym.make("fishing-v1", sigma=0.1)
+env = gym.make("fishing-v1", sigma=0.)
 
 agent = sb3.PPO.load("cache/ppo_tuned")
 agent_sims = env.simulate(agent, reps=100)

### 2021-06-05 python/fishing/stochastic/fishing_PPO_tuned.py

@@ -6,9 +6,7 @@ from torch import nn as nn
 import gym
 import gym_fishing
 
-seed = 24
-np.random.seed(seed)
-env = make_vec_env("fishing-v1", n_envs=8, seed=seed, env_kwargs={"sigma": 0.1})
+env = make_vec_env("fishing-v1", n_envs=4, seed=24, env_kwargs={"sigma": 0.1})
 
 # Create a fishing environment
 hyper = {
@@ -28,7 +26,7 @@ hyper = {
     }
 }
 # Create an agent
-agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
+agent = sb3.PPO("MlpPolicy", env, seed=24, **hyper)
 # Train the agent
 agent.learn(total_timesteps=300000)
 agent.save("cache/PPO_tuned")

### 2021-06-05 python/fishing/stochastic/fishing_SAC_tuned.py

@@ -20,7 +20,7 @@ hyper = {
     "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
 }
 
-agent = sb3.SAC("MlpPolicy", env, seed=seed, **hyper)
+agent = sb3.SAC("MlpPolicy", env, seed=24, **hyper)
 
 # Train the agent
 if not os.path.exists("cache/sac_tuned.zip"):

### 2021-06-05 appendices/appendix-B-examples.Rmd
sac-based example

--- 2021-06-04 appendices/appendix-B-examples.Rmd
@@
 # add a column indicating the model
-algos <- c("Optimal", "A2C")
+algos <- c("Optimal", "SAC")
 sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
@@
 Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
-For example, here we will try the A2C actor-critic algorithm, an extension of the A3C algorithm originally described in @A2C.
-We train an agent using the A2C algorithm using default hyper-parameter settings (which correspond to those used in @A2C),
-using a multi-layer perceptron ('MlpPolicy') using the default architecture for the policy network and value network composed of two 64-neuron layers.
+For example, here we will try the Soft Actor-Critic algorithm, SAC, training with the default hyperparameters originally posed in @SAC.
+We use SAC to train an agent that has a multi-layer perceptron ('MlpPolicy') neural network, with the default architecture of two 64-neuron layers.
 
 ```{r}
-a2c = sb3$A2C('MlpPolicy', env, seed = seed) 
+env = gym$make("fishing-v1", sigma=0.1)
+agent = sb3$SAC('MlpPolicy', env, seed = seed) 
 ```
@@
 ```{r}
-if(!file.exists("../python/cache/a2c.zip")) {
+if(!file.exists("../python/cache/sac.zip")) {
 
   # Train the agent for a fixed number of timesteps
-  a2c$learn(total_timesteps=300000L)
+  agent$learn(total_timesteps=300000L)
 
   # Save our trained agent for future use
-  a2c$save("../python/cache/a2c")
+  agent$save("../python/cache/sac")
 
@@
 This is discussed further below.
-Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.
+Having saved our trained agent to disk (`cache/sac.zip`), we can then re-load this agent for evaluation or to continue training.
 Note that a copy of the trained agents are included in the corresponding GitHub repository.
@@
 ```{r}
-# Simulate management under the trained agent
-a2c = sb3$A2C$load("../python/cache/a2c")
+# load the trained agent from cache
+agent = sb3$SAC$load("../python/cache/sac")
 ```
@@
 
-  out = a2c$predict(state, deterministic=TRUE)
+  out = agent$predict(state, deterministic=TRUE)
   action = out[[1]]
@@
 ```{r}
-a2c_sims = env$simulate(a2c, reps = 100L)
+agent_sims = env$simulate(agent, reps = 100L)
 ```
@@
 ```{r}
-a2c_policy = env$policyfn(a2c, reps = 5L)
+agent_policy = env$policyfn(agent, reps = 5L)
 ```
@@
 # stack result simulations with those above and plot:
-sims_df <- a2c_sims %>% 
-  mutate(model = factor("A2C", algos)) %>% 
+sims_df <- agent_sims %>% 
+  mutate(model = factor("SAC", algos)) %>% 
   bind_rows(sims_df)
-policy_df <- a2c_policy  %>% 
-  mutate(model = factor("A2C", algos))  %>% 
+policy_df <- agent_policy  %>% 
+  mutate(model = factor("SAC", algos))  %>% 
   bind_rows(policy_df)
@@
 
-After an initial over-harvest (top panel), A2C achieves nearly optimal performance as measured by reward over time, (bottom right), or compared to the steady-state management solution.
+After an initial over-harvest (top panel), SAC achieves nearly optimal performance as measured by reward over time, (bottom right), or compared to the steady-state management solution.
 The policy function (bottom left) explains this behavior.
@@
 
-In this simple example, A2C is able to achieve nearly-optimal performance (as defined by the reward conditions) without additional tuning.
-It is worth noting that this is not true of all RL algorithms.
-For example, the PPO algorithm [@A2C] is a somewhat more computationally intensive alternative that combines strategies from A2C and another algorithm, TRPO [@PPO].
-Nevertheless, using default hyper-parameters, PPO does considerably less well (not shown).
-For harder problems, it is often necessary to consider a range of algorithms, and tune the hyper-parameters extensively before training will be able to discover an acceptable solution.
-We illustrate this briefly in our second example gym.
+In this simple example, SAC is able to achieve nearly-optimal performance (as defined by the reward conditions) without tuning, simply by using hyperparameters originally chosen for different RL tasks.
+As noted in appendix C, careful design of the environment, such as rescaling of state and action space, can also improve the ability of a Deep RL algorithm to find an acceptable policy without altering the substance of the challenge.
+Even so, our agent still had extensive time to train on this specific environment, a total of 300,000 timesteps or 3000 training episodes.
+Sufficient tuning of the hyperparameters to this environment may be able result in better performance than we have seen here.
+For more difficult problems, it may be impossible to find an acceptable policy without first tuning candidate approaches to select the best hyperparameters for the environment.
 
@@
 Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
+Here, we load a local copy of the stock assessment for the Argentine Hake from 1986 to 2000.
+
+```{r}
+hake = read_csv("../data/hake.csv")
+```
+
 We will use the agent we have just trained using the `gym_fishing` simulator above,
@@
 # or Appendix D
-env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-agent = sb3$A2C$load("../python/cache/a2c_untuned")
+agent = sb3$SAC$load("../python/cache/sac")
 
@@
 ```{r}
-hake = read_csv("../data/hake.csv")
 x0 = hake %>%
@@
 
-Our hindcast considers the scenario of managing catch quotas using TD3, beginning in 1986. 
+Our hindcast considers the scenario of managing catch quotas using the RL agent, beginning in 1986. 
 We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
-This shows that quotas set by TD3 would have been lower than historical values throughout the late 80s and early 90s, which
+This shows that quotas set by the RL agent would have been lower than historical values throughout the late 80s and early 90s, which
 saw a sharp decline in estimated biomass of the Argentine Hake.
-For the same or higher estimated stock sizes from the historical record, we see that TD3 would recommend a lower quota.
-We also see that under TD3 the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
+For the same or higher estimated stock sizes from the historical record, we see that the RL agent would recommend a lower quota.
+We also see that under the RL agent the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
 being set in the now-depleted stock.  
-Whether the TD3 quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
-However, the fact that TD3 would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
+Whether the the RL agent quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
+However, the fact that the RL agent would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
 Combined with the observed historical declines, this provides some evidence that the RL agent could have decreased or avoided the over-harvesting in this case.
@@
 ```{r}
-env$reset()
-## represent the initial state size in the 'rescaled' state space.
+# Env is used only for helper methods to translate between original units and transformed space
+env = gym$make("fishing-v1")
+## Now we can represent the initial stock size in the 'rescaled' state space.
 state = env$get_state( x0 )
+
 for(i in 2:Tmax){
 
-  # RL-recommended harvest action:
+  # The RL agent recommends a Quota based on current state
   out = agent$predict(state)
@@
 
-  # Record state and resulting action
+  # We record state and proposed quota (in the original units)
   agent_state[i] = env$get_fish_population(state)
@@
 
-  # Implement RL-recommended harvest. 
-  result = env$step(action)
-  state = result[[1]]
-
-  #state = env$get_state(hake$scaled_biomass[i+1])
+  # Use historical data to determine the next state
+  state = env$get_state(hake$scaled_biomass[i+1])
 }

### 2021-06-05 appendices/algorithm_table.Rmd
Still having issues with pdf output

---
output: pdf_document
header-includes:
  - \pagestyle{empty}
  - \geometry{margin=0in}
bibliography: ../manuscript/references.bib
suppress-bibliography: True
---
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
| Algorithm | Description                                   | Model Use   | Evaluation | Policy        | Method            |
+===========+===============================================+=============+============+===============+===================+
|    MBPO   | Model-based Policy Optimization [@mbpo]       | Model-based | On-policy  | Deterministic | Policy Gradient   |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    MCTS   | Monte Carlo Tree Search [@suttonbarto]        | Model-based | Either     | Either        | Any               |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    A2C    | Advantage Actor Critic [@A2C]                 | Model-free  | On-policy  | Stochastic    | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    A3C    | Asynchronous A2C       [@a3c]                 | Model-free  | On-policy  | Stochastic    | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    TRPO   | Trust Region Policy Optimization [@trpo]      | Model-free  | On-policy  | Stochastic    | Policy gradient   |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    PPO    | Proximal Policy Optimization [@ppo]           | Model-free  | On-policy  | Stochastic    | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    DQN    | Deep Q Networks [@DQN]                        |  Model-free | Off-policy | Deterministic | Value-based       |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    DDPG   | Deep Deterministic Policy Gradient [@ddpg]    | Model-free  | Off-policy | Deterministic | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    TD3    | Twin Delayed DDPG [@TD3]                      |  Model-free | Off-policy | Deterministic | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    SAC    |  Soft Actor Critic [@sac]                     | Model-free  | Off-policy | Stochastic    | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+
|    IMPALA |Importance Weighted Actor Learner [@impala]    | Model-free  | Off-policy | Stochastic    | Actor-critic      |
+-----------+-----------------------------------------------+-------------+------------+---------------+-------------------+

### 2021-06-05 appendices/appendix-A-definitions.Rmd
Still having issues with pdf output

@@
 Reinforcement learning is a huge, constantly evolving field, and we've only touched on a small sliver of RL in our study.
-In this section, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with current state of the art algorithms.
-If you are interested in learning more about model-free deep RL, then reading papers like @sac, @DQN and @ppo, are good places to start.
+In this appendix, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with current state of the art algorithms.
+In table ..., we present some commonly used deep RL algorithms.
+Readers that are interested in learning more about the current state of deep RL should read the papers referenced in table ...; @DQN, @sac and @PPO are good places to start. 
 We have also not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @salimans_evolution_2017 for more here.
-A benefit of model-free deep RL over these other approaches is that there has been more of a general focus on model-free deep RL, so there are widely available packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL; but the focus on model-free algorithm could easily change in coming years.
-Deep RL has only become established in the last decade, and with its amount of unanswered research questions, deep RL may evolve significantly in the near future.
+A benefit of model-free deep RL over these other approaches is that there has been more of a general focus on model-free deep RL, so there are widely available packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
+It is important to keep in mind though that deep RL has only become established in the last decade, and with the large amount of open research questions, deep RL may evolve significantly in the near future.
 
+\begin{figure}
+\includegraphics[width=\linewidth]{algorithm_table.pdf}
+\caption{Common Deep RL Algorithms. }
+\end{figure}
 # References
-
-----

### 2021-06-05 appendices/algorithm_table_tex.Rmd
Getting a table in tex up

---
output: 
  pdf_document:
    fig_width: 7
    fig_height: 6.5
    fig_crop: False

header-includes:
  - \RequirePackage{colortbl}
  - \RequirePackage{xcolor}
  - \usepackage{xcolor}[tbl]
  - \pagestyle{empty}
  - \geometry{margin=0in}

bibliography: ../manuscript/references.bib

---
\definecolor{lightgray}{gray}{0.9}
\begin{center}
\begin{tabular}{c|c|c|c|c}
\hline
Algorithm & Description                                     & Model       & Policy     & Method \\
\hline
\rowcolor{lightgray}
MBPO      & Model-based Policy Optimization \cite{mbpo}    & Model-based & On-policy  & Policy Gradient \\
MCTS      & Monte Carlo Tree Search \cite{suttonbarto}     & Model-based & Either     & Any\\
\rowcolor{lightgray}
A2C       & Advantage Actor Critic  \cite{A2C}             & Model-free  & On-policy  & Actor-critic\\
A3C       & Asynchronous A2C  \cite{a3c}                   & Model-free  & On-policy  & Actor-critic\\
\rowcolor{lightgray}
TRPO      & Trust Region Policy Optimization  \cite{trpo}  & Model-free  & On-policy  & Policy Gradient\\
PPO       & Proximal Policy Optimization  \cite{ppo}       & Model-free  & On-policy  & Actor-critic\\
\rowcolor{lightgray}
DQN       & Deep Q Networks  \cite{DQN}                    & Model-free  & Off-policy & Value-Based\\
DDPG      & Deep Deterministic Policy Gradient  \cite{ddpg}& Model-free  & Off-policy & Actor-critic\\
\rowcolor{lightgray}
TD3       & Twin Delayed DDPG  \cite{TD3}                  & Model-free  & Off-policy & Actor-critic\\
SAC       & Soft Actor Critic  \cite{sac}                  & Model-free  & Off-policy & Actor-critic\\
\rowcolor{lightgray}
IMPALA    & Importance Weighted Actor Learner \cite{impala}& Model-free  & Off-policy & Actor-critic\\
\end{tabular}
\end{center}

### 2021-06-05 appendices/appendix-A-definitions.Rmd
Adding the latex table to the document

@@
   - \usepackage{amssymb}
+  - \RequirePackage{colortbl}
+  - \RequirePackage{xcolor}
+  - \usepackage{xcolor}[tbl]
 
@@
 
-\begin{figure}
-\includegraphics[width=\linewidth]{algorithm_table.pdf}
-\caption{Common Deep RL Algorithms. }
-\end{figure}
+\begin{tabular}{c|c|c|c|c}
+\hline
+Algorithm & Description                                     & Model       & Policy     & Method \\
+\hline
+\rowcolor{lightgray}
+MBPO      & Model-based Policy Optimization \cite{mbpo}    & Model-based & On-policy  & Policy Gradient \\
+MCTS      & Monte Carlo Tree Search \cite{suttonbarto}     & Model-based & Either     & Any\\
+\rowcolor{lightgray}
+A2C       & Advantage Actor Critic  \cite{A2C}             & Model-free  & On-policy  & Actor-critic\\
+A3C       & Asynchronous A2C  \cite{a3c}                   & Model-free  & On-policy  & Actor-critic\\
+\rowcolor{lightgray}
+TRPO      & Trust Region Policy Optimization  \cite{trpo}  & Model-free  & On-policy  & Policy Gradient\\
+PPO       & Proximal Policy Optimization  \cite{ppo}       & Model-free  & On-policy  & Actor-critic\\
+\rowcolor{lightgray}
+DQN       & Deep Q Networks  \cite{DQN}                    & Model-free  & Off-policy & Value-Based\\
+DDPG      & Deep Deterministic Policy Gradient  \cite{ddpg}& Model-free  & Off-policy & Actor-critic\\
+\rowcolor{lightgray}
+TD3       & Twin Delayed DDPG  \cite{TD3}                  & Model-free  & Off-policy & Actor-critic\\
+SAC       & Soft Actor Critic  \cite{sac}                  & Model-free  & Off-policy & Actor-critic\\
+\rowcolor{lightgray}
+IMPALA    & Importance Weighted Actor Learner \cite{impala}& Model-free  & Off-policy & Actor-critic\\
+\end{tabular}
+
+<!-- \begin{figure} -->
+<!-- \includegraphics[width=\linewidth]{algorithm_table.pdf} -->
+<!-- \caption{Common Deep RL Algorithms. } -->
+<!-- \end{figure} -->
 # References

### 2021-06-06 appendices/appendix-A-definitions.Rmd
Getting the table to look okay

--- 2021-06-05 appendices/appendix-A-definitions.Rmd
@@
 
+nocite: |
+    @A2C, @trpo, @ddpg, @TD3, @impala
+
 header-includes:
@@
 In this appendix, we will go into further detail on how deep RL algorithms work.
-The aim here is not to give full treatment of all deep RL methods -- @suttonbarto is an excellent resource for theoretical underpinnings while the white papers behind the algorithms go into greater detail on modern approaches --; instead, this section is meant to serve as an abbreviated background view on model-free deep RL.
-We hope to introduce some of the important design choices that are made for model-free deep RL algorithms and will point to specific resources that give a more complete presentation.
+The aim here is not to give full treatment of all deep RL methods; instead, this section is meant to serve as an abbreviated background view on model-free deep RL.
+We will introduce some of the important design choices that are made for model-free deep RL algorithms and will point to specific resources that give a more complete presentation.
 These sections presume some familiarity with gradient descent optimization -- see @ruder_overview_2017 for a thorough treatment.
-Throughout these sections, we'll consider the environment to be an MDP which makes for a cleaner presentation here and is also consistent with the referenced sources.
+Throughout these sections, we'll consider the environment to be an MDP which makes for a cleaner presentation and is also consistent with the referenced sources.
 
@@
 Policy gradients, value-based and actor-critic methods are the general classes of algorithms for gradient-based model-free RL.
-In this section, we will go over these algorithm classes to convey a broad sense of what they are attempting to achieve and the trade-offs between them.
+In this section, we will go over these algorithm classes to convey a sense of what they are attempting to achieve and the trade-offs between them.
 
 ## Policy Gradients
-The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
-There are both gradient-free and gradient-based policy search methods that have been successful in practice [@salimans_evolution_2017]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered.
+The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates its policy to maximize rewards.
+In deep RL, the agent's policy, $\pi$, is parameterized by a neural network.
+The parameters of the policy neural network are commonly denoted by $\theta$.
 In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters.
@@
 The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return.
-In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impedes learning [@suttonbarto].
+In practice, however, the empirical return tends to have high variance, resulting in very noisy gradients that impede learning [@suttonbarto].
 To avoid this issue, many policy gradient algorithms replace the empirical return with a return estimator that has lower variance [@suttonbarto].
@@
 Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
-Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
+Given that the objective of RL is to maximize the cumulative reward, "goodness" refers to how high of a cumulative reward the agent can expect to receive.
 For example, the state-action-value or Q function finds the expected return from a state-action pair under a policy, $\pi$,
@@
 There are more value functions that can used in practice [@schulman_high-dimensional_2018], but since Q-functions are commonly encountered, we will focus on Q-learning.
-From the above equation, we can define the Q function recursively,
+From the above equation, we can also define the Q function recursively,
 
@@
 that the state-action space is discrete and that all state-action pairs are visited ad infinitum [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q (s_t, a_t)$.
-Using this optimal policy, we can write an equation for the optimal Q-function from the previous definitions,
+Using this optimal policy, we can write an equation for the optimal Q-function from the previous definition,
 
@@
 
-After the Q-function neural network or Q-network has been fitted, the agent will select actions by inputting the current state into the Q network and then identifying the state-action pair with the highest estimated Q value.
-The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions over unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@DQN].
+After the Q-function neural network or Q-network has been fitted, the agent will select actions by inputting the current state into the Q network and then identifying the state-action pair with the highest approximated Q value.
+The benefit of using neural networks to approximate Q-functions is that neural networks are able to generalize Q functions to unseen state-action pairs, whereas classic methods like TD-learning are not able to generalize [@DQN].
 But there a host of problems with getting neural-network-based value-learning agents to work in practice; see @DQN and @van_hasselt_deep_2015 for more insight on these issues.   
@@
 In the sections above, we procedurally mentioned that the agent must select actions during training.
-Yet, how the agent selects these actions is very important because if an agent fails to explore the state-action space well, the agent will likely learn a bad policy.
-In this section, we will go over some common exploration algorithms and try to provide some practical insight.
+Yet, how the agent selects these actions is very important because if an agent explores the state-action space poorly, the agent will learn a sub-optimal policy.
+As mentioned in the manuscript, the dilemma in training is that it is difficult for the agent to know when to balance exploration of new actions and exploitation of past behavior.
+For RL problems with small state-action spaces, e.g.
+space-action spaces that can be represented in a table, simple exploration algorithms will converge to the optimal policy, but with large or infinite state-action spaces, there are no convergence guarantees.
+In the sections below, we will present some of the more commonly encountered exploration algorithms and then point to some more advanced methods.
 
@@
 
+## More Advanced Exploration Methods
+
 There is a multitude of exploration algorithms that we have not touched on, so it is important to say that the algorithms mentioned above are generally suited for dense reward or "easy-exploration" problems.
 For "hard-exploration" problems -- for instance, environments that have very few state-action pairs with significant rewards, like a maze, are seen as hard-exploration problems --, more intricate and bespoke exploration algorithms tend to outperform.
-We will not go over these more advanced exploration algorithms but a good example is seen in @go-explore.
+The issue for "hard-exploration" problems is that the agent needs to explore the state-action space efficiently, and simple exploration algorithms like epsilon-greedy will waste compute by revisiting bad actions.
+Two common approaches to hard exploration problems are memory-based and curiosity-driven techniques.
+For the sake of brevity, we will not go into further detail on these approaches but will provide some promising recent work.
+Look at @go-explore and @nevergiveup for recent applications of memory-based exploration.
+See @largescalecuriosity and @bellemare2016 for recent studies of curiosity-driven methods.
 
@@
 With a replay buffer, the agent will save state-action-reward transitions in the replay buffer, and during learning updates, the agent will sample transitions from the replay buffer.
-Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer will likely produce transitions that cover a broader region in state-action space than if the agent were to sample from one trajectory.
-On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- replay buffers are generally large enough to hold transitions from old policies.
+Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer will result in decorrelated samples.
+On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- note replay buffers are generally large enough to hold transitions from old policies.
 The solution to de-correlate samples for on-policy agents is to use parallel workers, whereby multiple workers explore the environment according to the current policy.
 Due to stochasticity that can come from exploration and the environment, the workers will go on different trajectories in state-action space.
-The on-policy agent will collect samples across these workers in a learning step, and then repeat the process of collecting and sampling transitions with an updated policy.
-The idea behind using parallel workers is that sampling over multiple trajectories will de-correlate samples -- see @a3c on more details here.
+The agent will collect samples across these workers in a learning step, perform its learning step and then repeat the process of collecting and sampling transitions with an updated policy.
+The idea behind using parallel workers is that sampling over multiple trajectories de-correlate samples -- see @a3c on more details here.
 In practice, using a large replay buffer, on the order of millions of transitions, or a modest number of parallel, around 4-16 for simple RL problems, is very important to achieve good agent performance for off-policy and on-policy algorithms respectively.
@@
 In table ..., we present some commonly used deep RL algorithms.
-Readers that are interested in learning more about the current state of deep RL should read the papers referenced in table ...; @DQN, @sac and @PPO are good places to start. 
-We have also not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @salimans_evolution_2017 for more here.
-A benefit of model-free deep RL over these other approaches is that there has been more of a general focus on model-free deep RL, so there are widely available packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
+Readers that are interested in learning more about the current state of deep RL should read the papers referenced in table ...; @DQN, @sac and @ppo are good places to start. 
+Yet, we have not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @salimans_evolution_2017 for more here.
+A benefit of model-free deep RL over these other approaches is that there has been more of a general focus recently on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
 It is important to keep in mind though that deep RL has only become established in the last decade, and with the large amount of open research questions, deep RL may evolve significantly in the near future.
 
-\begin{figure}
-\includegraphics[width=\linewidth]{algorithm_table.pdf}
-\caption{Common Deep RL Algorithms. }
-\end{figure}
+\resizebox{\textwidth}{!}{\begin{tabular}{l l l l l}
+\hline
+Algorithm & Description                                                       & Model       & Policy     & Method \\
+\hline
+MBPO      & Model-based Policy Optimization (Janner et al. 2019)              & Model-based & On-policy  & Policy Gradient \\
+MCTS      & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based & Either     & Any\\
+A2C       & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  & On-policy  & Actor-critic\\
+A3C       & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  & On-policy  & Actor-critic\\
+TRPO      & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  & On-policy  & Policy Gradient\\
+PPO       & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  & On-policy  & Actor-critic\\
+DQN       & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  & Off-policy & Value-Based\\
+DDPG      & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  & Off-policy & Actor-critic\\
+TD3       & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  & Off-policy & Actor-critic\\
+SAC       & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  & Off-policy & Actor-critic\\
+IMPALA    & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  & Off-policy & Actor-critic\\
+\end{tabular}}
+
 # References

### 2021-06-06 appendices/appendix-A-definitions.Rmd
Adding a footnote

@@
 The parameters of the policy neural network are commonly denoted by $\theta$.
-In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters.
+In policy gradient methods, the RL agent performs gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters.
 At each gradient step, the following update is performed,
@@
 
-A sketch of a simple policy gradient algorithm is shown in Algorithm 1.
+A sketch of a simple policy gradient algorithm is shown in Algorithm 1^[The algorithm sketches showed throughout this study are not practical for a variety of reasons including high bias estimators, high variance estimators and correlated samples.
+The algorithm examples are instead intended to convey how the general steps of the different algorithm classes are integrated by the agents.].
 

### 2021-06-06 appendices/appendix-A-definitions.Rmd
Tidying up

@@
 $$
-  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta).
+  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta) .
 $$
@@
 A sketch of a simple policy gradient algorithm is shown in Algorithm 1^[The algorithm sketches showed throughout this study are not practical for a variety of reasons including high bias estimators, high variance estimators and correlated samples.
-The algorithm examples are instead intended to convey how the general steps of the different algorithm classes are integrated by the agents.].
+The algorithm examples are instead intended to convey how the general steps of the different algorithm classes come together.].
 
@@
 \label{alg:policy_gradient}
-\caption{Vanilla On-policy Policy Gradient Algorithm}
+\caption{Generic On-policy Policy Gradient Algorithm}
 \begin{algorithmic}[1]
@@
 $$
-  Q^\pi(s_t, a_t) = r (s_t, a_t) + \gamma  \, \mathbb{E}_{s_{t+1} \sim T(s_{t+1} | s_t, a_t), \, a_{t+1} \sim \pi(a_{t+1}| s_{t+1})}\big[ Q^\pi(s_{t+1}, a_{t+1})\big]
+  Q^\pi(s_t, a_t) = r (s_t, a_t) + \gamma  \, \mathbb{E}_{s_{t+1} \sim T(s_{t+1} | s_t, a_t), \, a_{t+1} \sim \pi(a_{t+1}| s_{t+1})}\big[ Q^\pi(s_{t+1}, a_{t+1})\big].
 $$
@@
 that the state-action space is discrete and that all state-action pairs are visited ad infinitum [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q (s_t, a_t)$.
-Using this optimal policy, we can write an equation for the optimal Q-function from the previous definition,
+Using this optimal policy, we can write an equation for the optimal Q-function from the recursive definition,
 
 $$
-  Q^*(s_t, a_t) = r (s_t, a_t) + \gamma  \, \mathbb{E}_{s_{t+1} \sim T(s_{t+1} | s_t, a_t)}\big[ \underset{a_{t+1}}{\text{max}} \,\,Q^*(s_{t+1}, a_{t+1})\big]
+  Q^*(s_t, a_t) = r (s_t, a_t) + \gamma  \, \mathbb{E}_{s_{t+1} \sim T(s_{t+1} | s_t, a_t)}\big[ \underset{a_{t+1}}{\text{max}} \,\,Q^*(s_{t+1}, a_{t+1})\big].
 $$
@@
 $$
-  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 
+  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 .
 $$
@@
 \begin{algorithm}
-\caption{Vanilla On-policy Q-learning Algorithm}
+\caption{Generic On-policy Q-learning Algorithm}
 \begin{algorithmic}[1]
@@
 \begin{algorithm}
-\caption{Vanilla On-policy Actor-Critic Algorithm}
+\caption{Generic On-policy Actor-Critic Algorithm}
 \begin{algorithmic}[1]
@@
 The advantage of using Boltzmann exploration is that since the probability of selecting an action is proportional to how good the agent thinks that action is, the agent can explore the space more efficiently than with Epsilon-Greedy exploration.
-This is because an Epsilon-Greedy agent will explore all actions equally including low reward yielding actions; in contrast, with Boltzmann exploration, an agent will spend less time on these unpromising actions.
+An Epsilon-Greedy agent will explore all actions equally including low reward yielding actions; in contrast, with Boltzmann exploration, an agent will spend less time on these unpromising actions.
 
@@
 
-Injecting noise into an agent's action, observation or parameter space is another effective exploration method.
-With noise addition, there are a lot design choices that can be made.
+Injecting noise into an agent's action, observation or parameter space is another simple yet effective exploration method.
+With noise addition, there are a number of design choices that can be made.
 For instance, noise can be sampled from different distributions [@plappert_parameter_2018], scaled adaptively based on the policy[@plappert_parameter_2018], or the noise can be sampled according to the state that the agent observes [@gsde].
-There are a variety of reasons why one would want to use these variations of noise injection, which are well discussed in @plappert_parameter_2018 and @gsde.
+There are a variety of reasons why one would want to use each of these variations; these reasons are well discussed in @plappert_parameter_2018 and @gsde.
 
@@
 An increasingly common way to achieve exploration is by adding a function of the entropy term, $H(\pi(a|s))$, to the loss function of the policy network.
-While this introduces some bias as the agent no longer exclusively maximizes cumulative rewards, entropy-based exploration has been effective in practice at improving exploration, resulting in better agent performance.
-@ahmed_understanding_2019 gives a very complete discussion on the benefits of using entropy.
+By adding this term, the agent has an incentive to learn a policy that balances reward-seeking and explorative behavior. 
+While this introduces some bias as the agent no longer exclusively maximizes cumulative rewards, agents that use entropy-based exploration have achieved state of the art results [@sac].
+For more, @ahmed_understanding_2019 gives a thorough discussion on the benefits of using entropy.
 
@@
 
-There is a multitude of exploration algorithms that we have not touched on, so it is important to say that the algorithms mentioned above are generally suited for dense reward or "easy-exploration" problems.
+There is a multitude of exploration algorithms that we have not touched on. 
+It is important to note that the algorithms mentioned above are generally suited for dense reward or "easy-exploration" problems.
 For "hard-exploration" problems -- for instance, environments that have very few state-action pairs with significant rewards, like a maze, are seen as hard-exploration problems --, more intricate and bespoke exploration algorithms tend to outperform.
-The issue for "hard-exploration" problems is that the agent needs to explore the state-action space efficiently, and simple exploration algorithms like epsilon-greedy will waste compute by revisiting bad actions.
+The issue for "hard-exploration" problems is that the agent needs to explore the state-action space efficiently, and simple exploration algorithms will waste compute by revisiting bad actions.
 Two common approaches to hard exploration problems are memory-based and curiosity-driven techniques.
@@
 
-Another important algorithm design choice involves how an RL agent collects and samples past experiences.
+Another critical algorithm design choice involves how an RL agent collects and samples past experiences.
 A significant issue that arises from sampling past experiences is correlation of samples [@DQN, @schaul_prioritized_2016].
@@
 Reinforcement learning is a huge, constantly evolving field, and we've only touched on a small sliver of RL in our study.
-In this appendix, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with current state of the art algorithms.
-In table ..., we present some commonly used deep RL algorithms.
-Readers that are interested in learning more about the current state of deep RL should read the papers referenced in table ...; @DQN, @sac and @ppo are good places to start. 
-Yet, we have not given any space to discuss model-based or gradient-free techniques which also warrant attention -- see @mbpo and @salimans_evolution_2017 for more here.
-A benefit of model-free deep RL over these other approaches is that there has been more of a general focus recently on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
-It is important to keep in mind though that deep RL has only become established in the last decade, and with the large amount of open research questions, deep RL may evolve significantly in the near future.
+In this appendix, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with state of the art model-free agents.
+In Table 1, we collect some of the more commonly used deep RL algorithms.
+Readers who are interested in learning more about the current state of deep RL should read the literature referenced in Table 1; @DQN, @sac and @ppo are good places to start. 
+It is important to note that while we have focused on gradient-based, model-free methods, there has also been promising recent development with model-based and gradient-free techniques.
+See @mbpo for a recent example of a model-based deep RL algorithm and @salimans_evolution_2017 as a recent example of a gradient-free approach.
+A benefit of model-free deep RL over these other methods is that there has been more of a general focus on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
+However, it is important to keep in mind though that deep RL has only become established in the last decade, and with the large amount of open research questions, deep RL may evolve significantly in the near future.
 
+\begin{table}
 \resizebox{\textwidth}{!}{\begin{tabular}{l l l l l}
@@
 \end{tabular}}
+\caption{Survey of common deep RL algorithms. }
+\end{table}
 

### 2021-06-06 appendices/appendix-A-definitions.Rmd
Fixing table positioning

@@
 There are two commonly used solutions to this problem of correlated samples: Replay Buffers and Parallel Workers.
-If the agent is off-policy, the agent can employ a replay buffer -- see @DQN for a now classic example for a replay buffer-based agent.
+If the agent is off-policy, the agent can employ a replay buffer -- see @DQN for an example of a replay buffer-based agent.
 With a replay buffer, the agent will save state-action-reward transitions in the replay buffer, and during learning updates, the agent will sample transitions from the replay buffer.
@@
 The agent will collect samples across these workers in a learning step, perform its learning step and then repeat the process of collecting and sampling transitions with an updated policy.
-The idea behind using parallel workers is that sampling over multiple trajectories de-correlate samples -- see @a3c on more details here.
+The idea behind using parallel workers is that sampling over multiple trajectories de-correlates samples -- see @a3c on more details here.
 In practice, using a large replay buffer, on the order of millions of transitions, or a modest number of parallel, around 4-16 for simple RL problems, is very important to achieve good agent performance for off-policy and on-policy algorithms respectively.
@@
 
-\begin{table}
+\begin{table}[H]
 \resizebox{\textwidth}{!}{\begin{tabular}{l l l l l}

### 2021-06-07 manuscript/manuscript.Rmd
Cutting and editing a few sentences

@@
 An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
-Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
+The *agent*, which is often a computer program, continuously interacts with the environment. 
 At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
@@
 Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
-The training period involves a trial and error process whereby the agent will perform an action and then receive feedback from the environment in the form of a reward. 
-This process will repeat until the environment transitions to a terminal state, after which the environment will reset to an initial state.
-According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its policy.
+The training period involves a trial and error process in which the agent inputs actions to the environment and uses the reward signal to learn an optimal policy. 
+According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its decisions.
 The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
@@
 
-In this brief section, we will introduce the concept of the RL environment as partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.
-Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is critical to understanding how RL algorithms work.
-<!-- ML: this seems like an obvious sentence to me but am stuck on it atm so revisions welcomed-->
+In this brief section, we will introduce the concept of the RL environment as a partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.
 
@@
 The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
-This process will repeat until a terminal state is reached.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
+This process will repeat until a terminal state is reached, after which the environment can reset to an initial state.
+As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the length of the state-action sequence.
 From these definitions, we can provide an agent's trajectory distribution for a given policy as,
@@
 
-Note that this formalism extends to an infinite horizon case when $T= \infty$.
+Note that this formalism extends to an infinite horizon case when $H= \infty$.
 While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
@@
 Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
-Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.
+Since our examples use model-free agents, we will focus on model-free algorithms in this section.
 
-Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
+Among model-free methods, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
 Generally, model-free agents are either learning a policy function, a value function or both.
@@
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the neural network that the agent is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
@@
 Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
-In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics. <!-- ML: Have we dropped the table? -->
 In Appendices A and B, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
@@
 
-## Sustainable harvest \label{sec:fishery}
+## Sustainable harvest
 
@@
 
-## Ecological Tipping Points \label{sec:tipping}
+## Ecological Tipping Points
 
@@
 
-## Additional Environments
+## Unsolved environments: Wildfires and other examples
 
-Ecology holds many open problems for deep RL.  
-It is straight-forward to extend the simple environments presented here to reflect greater biological complexity, or to consider a host of more realistic decision scenarios.
-We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
-Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
+<!-- ML: Need to rework this to talk about other environments that we are working on basically point to conservation gym website -->
+One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
+Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
+Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
+Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+
+<!-- Marcus add ~ paragraph discussion of the fire gym, citing motivating examples, basic overview of the state space, action space and reward and a screenshot of the simulation as a figure? -->
 

### 2021-06-07 appendices/appendix-B-examples.Rmd
td3-fishing overfish slightly

@@
 # add a column indicating the model
-algos <- c("Optimal", "SAC")
+algos <- c("Optimal", "TD3", "TD3-tuned")
 sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
@@
 Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
-For example, here we will try the Soft Actor-Critic algorithm, SAC, training with the default hyperparameters originally posed in @SAC.
-We use SAC to train an agent that has a multi-layer perceptron ('MlpPolicy') neural network, with the default architecture of two 64-neuron layers.
+<!-- For example, here we will try the Actor-Critic algorithm A2C, training with the default hyperparameters originally posed in @A2C.-->
+We use this algorithm to train an agent that has a multi-layer perceptron ('MlpPolicy') neural network, with the default architecture of two 64-neuron layers.
 
@@
 env = gym$make("fishing-v1", sigma=0.1)
-agent = sb3$SAC('MlpPolicy', env, seed = seed) 
+agent = sb3$TD3('MlpPolicy', env, seed = seed) 
 ```
@@
 ```{r}
-if(!file.exists("../python/cache/sac.zip")) {
+if(!file.exists("../python/cache/td3_untuned.zip")) {
 
@@
   # Save our trained agent for future use
-  agent$save("../python/cache/sac")
+  agent$save("../python/cache/td3_untuned")
 
@@
 This is discussed further below.
-Having saved our trained agent to disk (`cache/sac.zip`), we can then re-load this agent for evaluation or to continue training.
+Having saved our trained agent to disk, we can then re-load this agent for evaluation or to continue training.
 Note that a copy of the trained agents are included in the corresponding GitHub repository.
@@
 # load the trained agent from cache
-agent = sb3$SAC$load("../python/cache/sac")
+agent = sb3$TD3$load("../python/cache/td3_untuned")
 ```
@@
 sims_df <- agent_sims %>% 
-  mutate(model = factor("SAC", algos)) %>% 
+  mutate(model = factor("RL", algos)) %>% 
   bind_rows(sims_df)
 policy_df <- agent_policy  %>% 
-  mutate(model = factor("SAC", algos))  %>% 
+  mutate(model = factor("RL", algos))  %>% 
   bind_rows(policy_df)
@@
 
-After an initial over-harvest (top panel), SAC achieves nearly optimal performance as measured by reward over time, (bottom right), or compared to the steady-state management solution.
-The policy function (bottom left) explains this behavior.
-The RL agent has a nearly optimal policy for stock sizes between zero and just above the $B_{MSY}$ target.
-At higher stocks, the agent has a tendency to over-harvest.
+## After Tuning
 
-In this simple example, SAC is able to achieve nearly-optimal performance (as defined by the reward conditions) without tuning, simply by using hyperparameters originally chosen for different RL tasks.
+```{r}
+hyper$action_noise = sb3$common$noise$NormalActionNoise(
+        mean=np$zeros(1L),
+        sigma= 0.6656948079225263* np$ones(1L))
+td3 = sb3$TD3('MlpPolicy',
+              env,  
+              seed = 42L,
+              gamma = 0.995,
+              learning_rate = 0.0001355522450968401,
+              batch_size = 128L,
+              buffer_size = 10000L,
+              train_freq = 128L,
+              gradient_steps = 128L,
+              policy_kwargs = list(net_arch=c(400L, 300L)))
+```
+
+```{r}
+if(!file.exists("../python/cache/td3_tuned.zip")){
+  td3$learn(total_timesteps=300000L)
+  td3$save("../python/cache/td3_tuned.zip")
+}
+```
+
+```{r}
+td3_sims = env$simulate(td3, reps = 100L)
+td3_policy = env$policyfn(td3, reps = 5L)
+
+# stack result simulations with those above and plot:
+sims_df <- td3_sims %>% 
+  mutate(model = "TD3_tuned") %>% 
+  bind_rows(sims_df)
+policy_df <- td3_policy  %>% 
+  mutate(model = "TD3_tuned")  %>% 
+  bind_rows(policy_df)
+
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
+```
+
+In this simple example, the RL agent is able to achieve nearly-optimal performance (as defined by the reward conditions) without tuning, simply by using hyperparameters originally chosen for different RL tasks.
 As noted in appendix C, careful design of the environment, such as rescaling of state and action space, can also improve the ability of a Deep RL algorithm to find an acceptable policy without altering the substance of the challenge.
@@
 We will use the agent we have just trained using the `gym_fishing` simulator above,
-
-```{r}
-# Simulate management under the trained agent
-# parameters based on statistical best model estimate, see Memarzadeh & Boettiger 2019
-# or Appendix D
-agent = sb3$SAC$load("../python/cache/sac")
-
-```
 
@@
 For some such scenarios the optimal solution is known or can be determined by stochastic dynamic programming, while for others, including the scenario of focus here, the optimal solution is unknown.
-
-For simplicity of illustration, we will train only a TD3-based agent on this scenario.
 

### 2021-06-07 appendices/appendix-B-examples.Rmd
tidy up

@@
 txtcolor = "#586e75"
+
+colors <- set_names(c(pal[c(1,3,5,8)], txtcolor),
+                    c("Optimal", "TD3", "TD3_tuned", "RL", "historical"))
+scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
+scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
 ```
@@
 ```{r}
+
 # add a column indicating the model
-algos <- c("Optimal", "TD3", "TD3-tuned")
-sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
-policy_df <- opt_policy  %>% mutate(model = factor("Optimal", algos))
+sims_df <- opt_sims %>% mutate(model = "Optimal")
+policy_df <- opt_policy  %>% mutate(model = "Optimal")
 
@@
 sims_df <- agent_sims %>% 
-  mutate(model = factor("RL", algos)) %>% 
+  mutate(model = "TD3") %>% 
   bind_rows(sims_df)
 policy_df <- agent_policy  %>% 
-  mutate(model = factor("RL", algos))  %>% 
+  mutate(model = "TD3")  %>% 
   bind_rows(policy_df)
@@
 
-## After Tuning
+As we can see in the figure above, without hyper-parameter tuning, our agent has found only a sub-optimal policy.
+The RL agent trained under an un-tuned TD3 algorithm here tends to over-harvest the fishery, resulting in an initially higher reward that grows steadily but more slowly than the optimal strategy.
+As discussed in the main text, training of Deep RL algorithms is subject to several sources of instability that can make it difficult or impossible to reproduce the same results from training with the same hyper-parameters on different hardware or hardware drivers, even when controlling for random seed [@instability_2019; @drlthatmatters].
+Consequently, users may not recover identical behavior when training under identical parameters.
+However, by using the cached neural network parameter weights of the trained agent, users should be able to reproduce the same agent behavior.
+
+## Tuning hyperparameters
+
+By tuning the hyper-parameters that control the training regime used by the TD3 algorithm, we are able to further improve upon on the performance of the RL agent.
+Tuning may be performed manually by adjusting hyperparameters based on mechanistic reasoning (e.g.
+larger action noise may improve exploration over exploitation, a larger neural network may better approximate a more complex policy).
+This search can also be automated using standard non-linear optimization techniques, adjusting hyperparameters to maximize the reward attained by the trained network.
+Using such an approach (see <https://github.com/boettiger-lab/conservation-agents>), we discover the following hyperparameters achieve better performance (see @TD3 or @SB3 documentation for formal definitions of each hyperparameter) :
 
 ```{r}
-hyper$action_noise = sb3$common$noise$NormalActionNoise(
-        mean=np$zeros(1L),
-        sigma= 0.6656948079225263* np$ones(1L))
 td3 = sb3$TD3('MlpPolicy',
@@
               gradient_steps = 128L,
+              action_noise = sb3$common$noise$NormalActionNoise(
+                mean=np$zeros(1L),
+                sigma= 0.6656948079225263 * np$ones(1L)),
               policy_kwargs = list(net_arch=c(400L, 300L)))
 ```
+
+We train this new RL agent using it's tuned hyperparameters, and saving the resulting trained agent as before:
 
@@
 
+Once our agent has been trained, we can again evaluate its performance across replicate simulations:
+
 ```{r}
+td3 <- sb3$TD3$load("../python/cache/td3_tuned.zip")
+
 td3_sims = env$simulate(td3, reps = 100L)
-td3_policy = env$policyfn(td3, reps = 5L)
+td3_policy = env$policyfn(td3)
+```
 
+```{r}
 # stack result simulations with those above and plot:
@@
 
+colors <- set_names(c(pal[c(1,3,8,2)], txtcolor),
+                    c("Optimal", "TD3", "TD3_tuned", "RL", "historical"))
+scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
+scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
 plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
@@
 
-In this simple example, the RL agent is able to achieve nearly-optimal performance (as defined by the reward conditions) without tuning, simply by using hyperparameters originally chosen for different RL tasks.
-As noted in appendix C, careful design of the environment, such as rescaling of state and action space, can also improve the ability of a Deep RL algorithm to find an acceptable policy without altering the substance of the challenge.
+In this simple example, the RL tuned agent is able to achieve nearly-optimal performance (as defined by the reward conditions).
+Note that the largest deviations from the optimal policy occur at large stock sizes, at or above carrying capacity of the model.
+This pattern is not surprising, since these large stock values are least likely to observed over the simulated episodes.
+For the same reason, that deviation contributes very little to reward. 
+Only by comparing average stock sizes directly can we see that the RL agent slightly under-harvests on average relative to the optimal policy.
+As noted in appendix C, careful design of the environment, such as re-scaling of state and action space, can also improve the ability of a Deep RL algorithm to find an acceptable policy without altering the substance of the challenge.
 Even so, our agent still had extensive time to train on this specific environment, a total of 300,000 timesteps or 3000 training episodes.
-Sufficient tuning of the hyperparameters to this environment may be able result in better performance than we have seen here.
-For more difficult problems, it may be impossible to find an acceptable policy without first tuning candidate approaches to select the best hyperparameters for the environment.
 
@@
 We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.  
-
 Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
 Here, we load a local copy of the stock assessment for the Argentine Hake from 1986 to 2000.
+(Annotated code for data cleaning can be found in Appendix D).
 
-```{r}
+```{r message=FALSE}
 hake = read_csv("../data/hake.csv")
@@
 
-We will use the agent we have just trained using the `gym_fishing` simulator above,
+For each year in the data, we compare the historical harvest to the quota that would have been recommended by the TD3-based Deep RL agent we have just trained.
+To utilize our RL agent to set a quota for the stock, we only have to map the observed biomass estimates and historical harvests into the re-scaled state space of the environment. 
+Recall that the `gym_fishing` environment includes helper function `get_state()` to turn observed biomass to the state-space units used in the gym.
 
 ```{r}
-x0 = hake %>%
- filter(year == min(year)) %>%
-  pull(scaled_biomass)
-Tmax = 15
-years =1986:2000
+# Initial states
+x0 = hake[[1, "scaled_biomass"]]
 agent_state = agent_action = numeric(Tmax)
@@
 agent_action[1] = NA
-```
 
-Our hindcast considers the scenario of managing catch quotas using the RL agent, beginning in 1986. 
-We compare both the anticipated harvest and resulting stock biomass to the historically recorded harvest and biomass.
-This shows that quotas set by the RL agent would have been lower than historical values throughout the late 80s and early 90s, which
-saw a sharp decline in estimated biomass of the Argentine Hake.
-For the same or higher estimated stock sizes from the historical record, we see that the RL agent would recommend a lower quota.
-We also see that under the RL agent the stock recovers sufficiently to predict higher quotas by the second half of the 90s than were
-being set in the now-depleted stock.  
-Whether the the RL agent quotas would have truly been sufficient to permit recovery in stock sizes shown in the simulation clearly depends on the accuracy of the underlying simulation.
-However, the fact that the RL agent would set a lower quota than the historical management given the same or even higher estimate of stock size is independent of the Hake simulation.
-Combined with the observed historical declines, this provides some evidence that the RL agent could have decreased or avoided the over-harvesting in this case.
-
-```{r}
-# Env is used only for helper methods to translate between original units and transformed space
-env = gym$make("fishing-v1")
-## Now we can represent the initial stock size in the 'rescaled' state space.
-state = env$get_state( x0 )
-
-for(i in 2:Tmax){
+# `env` is used only to translate between original units and transformed space
+# Represent the initial stock size in the 'rescaled' state space:
+state = env$get_state(x0)
+for(i in 1:15){
 
   # The RL agent recommends a Quota based on current state
-  out = agent$predict(state)
+  out = td3$predict(state)
   action = out[[1]]
@@
 }
+```
+
+Plot the results:
+
+```{r}
+years =1986:2000
 df = bind_rows(
-  tibble(year = years, state = agent_state, action = agent_action, model = "RL"),
-  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
-df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))
+  tibble(year = years, harvest = agent_action, model = "TD3_tuned"),
+  tibble(year = years, harvest = hake$scaled_catch, model = "historical")
+)
+stock <-  tibble(year = years, biomass = hake$scaled_biomass, model = "historical")
+
+top_panel <- ggplot(stock, aes(year, biomass, col=model)) + geom_line(lty=2) + 
+  theme(axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank())
+bottom_panel <- ggplot(df, aes(year, harvest, col=model)) + 
+  geom_line() + geom_point()
+
+top_panel / bottom_panel
 ```
+
+The Argentine hake biomass exhibited a steep decline between 1986 and 2000 as shown by the stock assessment estimates.
+Over that time, historical harvests showed some decline in step with the declining biomass.
+By contrast, for each of the historically observed biomass estimates the RL agent would have proposed much steeper cuts to the biomass quota which may have avoided the continued loss of biomass and eventual decline in harvest as well.
+This result should not be over-interpreted. 
+Without being able to replay history we have no way of truly knowing if the RL agent would have managed this stock better.
+While this example shows that it is in principle very straight forward to utilize a trained RL agent to recommend real harvest quotas, we must bear in mind that the agent is only as good as the environments on which it was trained.
+In practice, the RL agent would have to first be trained and evaluated on much more realistic environments than only the `fishing-v1` gym illustrated here.
 
@@
 
-### Manuscript Figure 3
-
 ```{r fig3}

### 2021-06-07 manuscript/manuscript.Rmd
tidy up

--- 2021-06-04 manuscript/manuscript.Rmd
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
+Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
 Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
@@
 
-In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
+In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
 Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
@@
 In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
+In our third example simulating wildfire spread highlights an open problem which we expect will be considerably harder to find effective solutions.
 These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
@@
 \includegraphics[width=\linewidth]{fig1.pdf}
-\caption{Deep Reinforcement Learning:  A deep RL \emph{agent} uses a \emph{neural network} to select an \emph{action} in response to an \emph{observation} of the \emph{environment}, and receives a \emph{reward} from the environment as a result.
-During \emph{training}, the agent tries to maximize its cumulative reward by interacting with the environment and learning from experience.
-In the RL loop, the agent performs an action, and the environment returns a reward as well as an observation of the environment's state.
-The agent-environment loop continues until the environment reaches a terminal state, after which the environment will reset, causing a new \emph{episode} to begin.
-Across training episodes, the agent will continually update the \emph{parameters} in its neural network, so that the agent will select better actions.
-Before training starts, the researcher must input a set of \emph{hyperparameters} to the agent; hyperparameters direct the learning process and thus affect the outcome of training.
-A researcher finds the best set of hyperparameters during \emph{tuning}.
-Hyperparameter tuning consists of iterative \emph{trials}, in which the agent is trained with different sets of hyperparameters.
-At the end of a trial, the agent is evaluated to see which set of hyperparameters results in the highest cumulative reward.
-An agent is \emph{evaluated} by recording the cumulative reward over one episode, or the mean reward over multiple episodes.
-Within evaluation, the agent does not update its neural network; instead, the agent only uses the neural network to select actions.}
+\caption{Deep Reinforcement Learning:  An artificial intelligence \emph{agent} uses a \emph{neural network} to select an \emph{action} in response to an \emph{observation} of the \emph{environment}, receiving a \emph{reward} (or punishment) from the environment as a result.
+The agent is \emph{trained} to maximize the reward through the experience of trying different actions over repeated \emph{epsiodes} of interacting with the environment.
+Over the course of training, the agent adjusts parameter weights on the neural network as it learns which actions lead to the best cumulative rewards.
+A researcher may \emph{tune} an agent by considering iterative trials that train different agents with different \emph{hyperparameters}, such as different neural network architectures.
+An agent trained with the most successful combination of hyperparameters can then be tested in an evaluation environment.}
 \end{figure}
@@
 txtcolor <- "#586e75"
+
+source("../R/plotting-helpers.R")
 ```
 
-## Sustainable harvest \label{sec:fishery}
+## Sustainable harvest
 
@@
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
-From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
+From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
 In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment." The simplest simulation is merely an implementation of the estimated model itself.
@@
 ```{r message=FALSE, include=FALSE}
-sims_df <- read_csv( "figs/sims_df.csv")
-policy_df <- read_csv("figs/policy_df.csv")
-reward_df <- read_csv("figs/reward_df.csv")
+sims_df <- read_csv( "figs/sims_df.csv") %>% filter(model %in% c("Optimal", "TD3_tuned"))
+policy_df <- read_csv("figs/policy_df.csv") %>% filter(model %in% c("Optimal", "TD3_tuned"))
 
-ymin <- function(x) last(x[(ntile(x, 20)==1)])
-ymax <- function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims <- 
-sims_df %>% 
-  group_by(time, model) %>% 
-  summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-
-fig_policy <- 
-  policy_df %>% ggplot(aes(state, state - action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) + 
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) + 
-  ylab("escapement")
-
-fig_reward <- reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
 ```
@@
 
-The TD3-trained agent proves far more successful in preventing chance transitions across the tipping point, consistently achieving a higher cumulative ecosystem service value across replicates than the steady-state strategy.
+The TD3-trained neural network agent proves far more successful in preventing chance transitions across the tipping point, consistently achieving a higher cumulative ecosystem service value across replicates than the steady-state strategy.
 
@@
 For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
-Recall that the quadratic scaling of cost makes the rapid response of the TD3 agent much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
+Recall that the quadratic scaling of cost makes the rapid response of the TD3 algorithm much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
 However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.
@@
 The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
-<!-- ML: the parenthetical here I understand but I feel a little ehh on the word choice.
-I also don't know how to rephrase it -->
 Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
+<!-- ML: I think it might be confusing to introduce RNNs, maybe just say different architectures, maybe make a footnote.  CB: agree, revised above-->
 
-## Additional Environments
+## Unsolved environments: Wildfires and other examples
 
-Ecology holds many open problems for deep RL.  
-It is straight-forward to extend the simple environments presented here to reflect greater biological complexity, or to consider a host of more realistic decision scenarios.
-We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
-Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
+One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
+While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
+Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
+Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
+Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
+To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
+
+<!-- Marcus add ~ paragraph discussion of the fire gym, citing motivating examples, basic overview of the state space, action space and reward and a screenshot of the simulation as a figure? -->
 
@@
 Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
-<!-- ML: A counterpoint here is that a model-based agent may learn a correct model that generalizes outside of the examples it has seen-->
 In these simple examples in which the simulation implements a single model, training is analogous to classical methods which take the model as given [@Marescot2013].
@@
 and can easily be extended into more complex problems once RL agents can master the basics.
-In the case that an agent has been trained to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
+In the case that someone trains an agent to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
 

### 2021-06-07 R/plotting-helpers.R

@@ -34,4 +34,3 @@ plot_reward <- function(sims_df, gamma = 1)
               alpha=0.25, show.legend = FALSE) +
   geom_line(aes(col = model), show.legend = FALSE) +
   ylab("reward")
-

### 2021-06-07 python/fishing/determinstic/fishing_A2C_tuned.py

@@ -1,54 +1,0 @@
-# Fishing using (untuned) A2C
-
-import numpy as np
-from torch import nn as nn
-import stable_baselines3 as sb3
-import os
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-
-hyper = {
-    "gamma": 0.98,
-    "normalize_advantage": True,
-    "max_grad_norm": 0.5,
-    "use_rms_prop": False,
-    "gae_lambda": 0.92,
-    "n_steps": 16,
-    "learning_rate": 0.0014125669914905178,
-    "ent_coef": 0.000415417442958291,
-    "vf_coef": 0.7417182197443206,
-    "policy_kwargs": {
-        "ortho_init": False,
-        "net_arch": [64, 64], #use separate networks?
-        "activation_fn": nn.ReLU,
-    }
-}
-
-# Create an A2C agent
-agent = sb3.A2C("MlpPolicy", env, seed=seed, **hyper)
-
-# Train
-if not os.path.exists("cache/a2c_tuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/a2c_tuned")
-
-# Load the saved agent
-agent = sb3.A2C.load("cache/a2c_tuned")
-
-# Evaluate
-mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
-print("mean score:", mean)
-
-# Plot results
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=1)
-env.plot(agent_sims, "results/fishing_a2c_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_a2c_tuned_policy.png")
-
-

### 2021-06-07 python/fishing/determinstic/fishing_A2C_untuned.py

@@ -1,33 +1,0 @@
-# Fishing using (untuned) A2C
-
-import os
-import gym
-import gym_fishing
-import stable_baselines3 as sb3
-
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.0)
-
-# Create an A2C agent
-agent = sb3.A2C("MlpPolicy", env, seed=24)
-
-# Train
-if not os.path.exists("cache/a2c_untuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/a2c_untuned")
-
-# Load the saved agent
-agent = sb3.A2C.load("cache/a2c_untuned")
-
-# Evaluate
-mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
-print("mean score:", mean)
-
-# Plot results
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=1)
-env.plot(agent_sims, "results/fishing_a2c_untuned.png")
-env.plot_policy(agent_policy, "results/fishing_a2c_untuned_policy.png")
-
-

### 2021-06-07 python/fishing/determinstic/fishing_DDPG_untuned.py

@@ -1,27 +1,0 @@
-# Fishing using untuned ddpg
-
-import numpy as np
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.)
-# Create the agent
-ddpg = sb3.DDPG("MlpPolicy", env, seed=seed)
-
-# Train the agent
-ddpg.learn(total_timesteps=300000)
-ddpg.save("cache/ddpg_untuned")
-
-# Simulate management under the trained agent
-ddpg = sb3.DDPG.load("cache/ddpg_untuned")
-ddpg_sims = env.simulate(ddpg, reps=100)
-ddpg_policy = env.policyfn(ddpg, reps=5)
-
-# Example of built-in plotting methods
-env.plot(ddpg_sims, "results/fishing_ddpg_untuned.png")
-env.plot_policy(ddpg_policy, "results/fishing_ddpg_untuned_policy.png")

### 2021-06-07 python/fishing/determinstic/fishing_PPO_tuned.py

@@ -1,47 +1,0 @@
-import numpy as np
-import stable_baselines3 as sb3
-from stable_baselines3.common.env_util import make_vec_env
-from torch import nn as nn
-
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-env = make_vec_env("fishing-v1", n_envs=8, seed=seed, env_kwargs={"sigma": 0.})
-
-# Create a fishing environment
-hyper = {
-    "batch_size": 16,
-    "n_steps": 8,
-    "gamma": 0.99,
-    "learning_rate": 0.0012512477171887717,
-    "ent_coef": 6.908006009288375e-08,
-    "clip_range": 0.3,
-    "n_epochs": 1,
-    "gae_lambda": 0.9,
-    "max_grad_norm": 0.7,
-    "vf_coef": 0.7688840852349988,
-     "policy_kwargs": {
-        "net_arch": [256, 256], # should be separate networks(?)
-        "activation_fn": nn.ReLU,
-    }
-}
-# Create an agent
-agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
-# Train the agent
-
-if not os.path.exists("cache/ppo_tuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/ppo_tuned")
-
-# Evaluate the trained agent
-env = gym.make("fishing-v1", sigma=0.1)
-
-agent = sb3.PPO.load("cache/ppo_tuned")
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=5)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_ppo_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_ppo_tuned_policy.png")

### 2021-06-07 python/fishing/determinstic/fishing_PPO_untuned.py

@@ -1,27 +1,0 @@
-import numpy as np
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.0)
-
-# Create an agent
-agent = sb3.PPO("MlpPolicy", env,seed=seed)
-
-# Train the agent
-agent.learn(total_timesteps=300000)
-agent.save("cache/ppo_untuned")
-
-# Evaluate the trained agent
-agent = sb3.PPO.load("cache/ppo_untuned")
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=5)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_ppo_untuned.png")
-env.plot_policy(agent_policy, "results/fishing_ppo_untuned_policy.png")
-

### 2021-06-07 python/fishing/determinstic/fishing_SAC_tuned.py

@@ -1,45 +1,0 @@
-import os
-import gym
-import gym_fishing
-import stable_baselines3 as sb3
-
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-
-# Create an agent
-hyper = {
-    "gamma": 0.95,
-    "learning_rate": 0.000115,
-    "batch_size": 128,
-    "buffer_size": 100000,
-    "learning_starts": 0,
-    "train_freq": 128,
-    "tau": 0.01,
-    "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
-}
-
-agent = sb3.SAC("MlpPolicy", env, seed=24, **hyper)
-
-# Train the agent
-if not os.path.exists("cache/sac_tuned"):
-  agent.learn(total_timesteps=300000)
-  agent.save("cache/sac_tuned")
-
-
-agent = sb3.SAC.load("cache/sac_tuned")
-
-# Evaluate the trained agent
-eval_env = sb3.common.monitor.Monitor(gym.make("fishing-v1"))
-mean_reward, std_reward = evaluate_policy(agent, eval_env, n_eval_episodes=100)
-print("mean reward:", mean_reward, "std:", std_reward, "tuned_value:", 7.755079)
-
-
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=1)
-
-print("mean reward from sims:", np.sum(agent_sims["reward"]) / 100)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_sac_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_sac_tuned_policy.png")

### 2021-06-07 python/fishing/determinstic/fishing_SAC_untuned.py

@@ -1,31 +1,0 @@
-import os
-import gym
-import gym_fishing
-import stable_baselines3 as sb3
-
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-
-# Create an agent
-agent = sb3.SAC("MlpPolicy", env, seed=24)
-
-# Train the agent
-if not os.path.exists("cache/sac_untuned.zip"):
-  agent.learn(total_timesteps=300000)
-  agent.save("cache/sac_untuned")
-
-agent = sb3.SAC.load("cache/sac_untuned")
-
-# Evaluate
-mean, sd = sb3.common.evaluation.evaluate_policy(agent, sb3.common.monitor.Monitor(env))
-print("mean score:", mean)
-
-# Plot results
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=2)
-env.plot(agent_sims, "results/fishing_sac_untuned.png")
-env.plot_policy(agent_policy, "results/fishing_sac_untuned_policy.png")
-
-
-

### 2021-06-07 python/fishing/determinstic/fishing_TD3_tuned.py

@@ -1,43 +1,0 @@
-# Fishing using tuned TD3
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-
-# Define the agent
-policy_kwargs = dict(net_arch=[400, 300])  # "big"
-noise_std = 0.6656948079225263
-action_noise = sb3.common.noise.NormalActionNoise(mean=0, sigma=noise_std)
-
-td3 = sb3.TD3(
-    "MlpPolicy",
-    env,
-    seed=seed,
-    policy_kwargs=policy_kwargs,
-    learning_rate=0.0001355522450968401,
-    gamma=0.995,
-    batch_size=128,
-    buffer_size=10000,
-    train_freq=(128, "step"),
-    gradient_steps=128,
-    action_noise=action_noise,
-)
-
-# Train the agent
-td3.learn(total_timesteps=300000)
-td3.save("cache/td3_tuned")
-
-
-# Simulate management under the trained agent
-td3 = sb3.TD3.load("cache/td3_tuned")
-td3_sims = env.simulate(td3, reps=100)
-td3_policy = env.policyfn(td3, reps=5)
-
-# Example of built-in plotting methods
-env.plot(td3_sims, "results/fishing_td3_tuned.png")
-env.plot_policy(td3_policy, "results/fishing_td3_tuned_policy.png")
-

### 2021-06-07 python/fishing/determinstic/fishing_TD3_untuned.py

@@ -1,25 +1,0 @@
-# Fishing using untuned TD3
-
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-# Create the agent
-td3 = sb3.TD3("MlpPolicy", env, seed=seed)
-
-# Train the agent
-td3.learn(total_timesteps=300000)
-td3.save("cache/td3_untuned")
-
-# Simulate management under the trained agent
-td3 = sb3.TD3.load("cache/td3_untuned")
-td3_sims = env.simulate(td3, reps=100)
-td3_policy = env.policyfn(td3, reps=5)
-
-# Example of built-in plotting methods
-env.plot(td3_sims, "results/fishing_td3_untuned.png")
-env.plot_policy(td3_policy, "results/fishing_td3_untuned_policy.png")

### 2021-06-07 python/fishing/stochastic/fishing_A2C_tuned.py

@@ -1,54 +1,0 @@
-# Fishing using (untuned) A2C
-
-import numpy as np
-from torch import nn as nn
-import stable_baselines3 as sb3
-import os
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.1)
-
-hyper = {
-    "gamma": 0.98,
-    "normalize_advantage": True,
-    "max_grad_norm": 0.5,
-    "use_rms_prop": False,
-    "gae_lambda": 0.92,
-    "n_steps": 16,
-    "learning_rate": 0.0014125669914905178,
-    "ent_coef": 0.000415417442958291,
-    "vf_coef": 0.7417182197443206,
-    "policy_kwargs": {
-        "ortho_init": False,
-        "net_arch": [64, 64],
-        "activation_fn": nn.ReLU,
-    }
-}
-
-# Create an A2C agent
-agent = sb3.A2C("MlpPolicy", env, seed=seed, **hyper)
-
-# Train
-if not os.path.exists("cache/a2c_tuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/a2c_tuned")
-
-# Load the saved agent
-agent = sb3.A2C.load("cache/a2c_tuned")
-
-# Evaluate
-mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
-print("mean score:", mean)
-
-# Plot results
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=1)
-env.plot(agent_sims, "results/fishing_a2c_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_a2c_tuned_policy.png")
-
-

### 2021-06-07 python/fishing/stochastic/fishing_PPO_tuned.py

@@ -1,45 +1,0 @@
-import numpy as np
-import stable_baselines3 as sb3
-from stable_baselines3.common.env_util import make_vec_env
-from torch import nn as nn
-
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-env = make_vec_env("fishing-v1", n_envs=8, seed=seed, env_kwargs={"sigma": 0.1})
-
-# Create a fishing environment
-hyper = {
-    "batch_size": 16,
-    "n_steps": 8,
-    "gamma": 0.99,
-    "learning_rate": 0.0012512477171887717,
-    "ent_coef": 6.908006009288375e-08,
-    "clip_range": 0.3,
-    "n_epochs": 1,
-    "gae_lambda": 0.9,
-    "max_grad_norm": 0.7,
-    "vf_coef": 0.7688840852349988,
-     "policy_kwargs": {
-        "net_arch": [256, 256],
-        "activation_fn": nn.ReLU,
-    }
-}
-# Create an agent
-agent = sb3.PPO("MlpPolicy", env, seed=seed, **hyper)
-# Train the agent
-agent.learn(total_timesteps=300000)
-agent.save("cache/PPO_tuned")
-
-# Evaluate the trained agent
-env = gym.make("fishing-v1", sigma=0.1)
-
-agent = sb3.PPO.load("cache/PPO_tuned")
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=5)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_PPO_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_PPO_tuned_policy.png")

### 2021-06-07 python/fishing/stochastic/fishing_PPO_untuned.py

@@ -1,27 +1,0 @@
-import numpy as np
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.0)
-
-# Create an agent
-agent = sb3.PPO("MlpPolicy", env,seed=seed)
-
-# Train the agent
-agent.learn(total_timesteps=300000)
-agent.save("cache/PPO_untuned")
-
-# Evaluate the trained agent
-agent = sb3.PPO.load("cache/PPO_untuned")
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=5)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_PPO_untuned.png")
-env.plot_policy(agent_policy, "results/fishing_PPO_untuned_policy.png")
-

### 2021-06-07 python/fishing/stochastic/fishing_SAC_tuned.py

@@ -1,46 +1,0 @@
-# Fishing using tuned SB3
-import os
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.1)
-
-# Create an agent
-hyper = {
-    "gamma": 0.95,
-    "learning_rate": 0.000115,
-    "batch_size": 128,
-    "buffer_size": 100000,
-    "learning_starts": 0,
-    "train_freq": 128,
-    "tau": 0.01,
-    "policy_kwargs": {"log_std_init": 0.680754, "net_arch": [400, 300]},
-}
-
-agent = sb3.SAC("MlpPolicy", env, seed=seed, **hyper)
-
-# Train the agent
-if not os.path.exists("cache/sac_tuned,zip"):
-  agent.learn(total_timesteps=300000)
-  agent.save("cache/sac_tuned")
-
-
-agent = sb3.SAC.load("cache/sac_tuned")
-
-# Evaluate the trained agent
-eval_env = sb3.common.monitor.Monitor(gym.make("fishing-v1"))
-mean_reward, std_reward = evaluate_policy(agent, eval_env, n_eval_episodes=100)
-print("mean reward:", mean_reward, "std:", std_reward, "tuned_value:", 7.755079)
-
-
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=5)
-
-print("mean reward from sims:", np.sum(agent_sims["reward"]) / 100)
-
-# Plot results
-env.plot(agent_sims, "results/fishing_sac_tuned.png")
-env.plot_policy(agent_policy, "results/fishing_sac_tuned_policy.png")

### 2021-06-07 python/fishing/stochastic/fishing_SAC_untuned.py

@@ -1,36 +1,0 @@
-# Fishing using tuned TD3
-
-import stable_baselines3 as sb3
-import os
-import gym
-import gym_fishing
-
-
-# Create a fishing environment
-env = gym.make("fishing-v1")
-
-# Create an agent
-agent = sb3.SAC("MlpPolicy", env, seed=24)
-
-# Train the agent
-if not os.path.exists("cache/sac_untuned.zip"):
-  agent.learn(total_timesteps=300000)
-  agent.save("cache/sac_untuned")
-
-
-
-agent = sb3.SAC.load("cache/sac_untuned")
-
-# Evaluate
-mean, sd = sb3.common.evaluation.evaluate_policy(agent, env)
-print("mean score:", mean)
-
-
-# Plot results
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=2)
-env.plot(agent_sims, "results/fishing_sac_untuned.png")
-env.plot_policy(agent_policy, "results/fishing_sac_untuned_policy.png")
-
-
-

### 2021-06-07 python/fishing/stochastic/fishing_TD3_tuned.py

@@ -1,46 +1,0 @@
-# Fishing using tuned TD3
-
-import numpy as np
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.1)
-
-# Define the agent
-policy_kwargs = dict(net_arch=[400, 300])  # "big"
-noise_std = 0.6656948079225263
-action_noise = sb3.common.noise.NormalActionNoise(mean=0, sigma=noise_std)
-
-td3 = sb3.TD3(
-    "MlpPolicy",
-    env,
-    seed=seed,
-    policy_kwargs=policy_kwargs,
-    learning_rate=0.0001355522450968401,
-    gamma=0.995,
-    batch_size=128,
-    buffer_size=10000,
-    train_freq=(128, "step"),
-    gradient_steps=128,
-    action_noise=action_noise,
-)
-
-# Train the agent
-td3.learn(total_timesteps=300000)
-td3.save("cache/td3_tuned")
-
-
-# Simulate management under the trained agent
-td3 = sb3.TD3.load("cache/td3_tuned")
-td3_sims = env.simulate(td3, reps=100)
-td3_policy = env.policyfn(td3, reps=5)
-
-# Example of built-in plotting methods
-env.plot(td3_sims, "results/fishing_td3_tuned.png")
-env.plot_policy(td3_policy, "results/fishing_td3_tuned_policy.png")
-

### 2021-06-07 python/fishing/stochastic/fishing_TD3_untuned.py

@@ -1,27 +1,0 @@
-# Fishing using untuned TD3
-
-import numpy as np
-import stable_baselines3 as sb3
-import gym
-import gym_fishing
-
-seed = 24
-np.random.seed(seed)
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.)
-# Create the agent
-td3 = sb3.TD3("MlpPolicy", env, seed=seed)
-
-# Train the agent
-td3.learn(total_timesteps=300000)
-td3.save("cache/td3_untuned")
-
-# Simulate management under the trained agent
-td3 = sb3.TD3.load("cache/td3_untuned")
-td3_sims = env.simulate(td3, reps=100)
-td3_policy = env.policyfn(td3, reps=5)
-
-# Example of built-in plotting methods
-env.plot(td3_sims, "results/fishing_td3_untuned.png")
-env.plot_policy(td3_policy, "results/fishing_td3_untuned_policy.png")

### 2021-06-07 python/fishing/stochastic/fishing_optimal.py

@@ -1,14 +1,0 @@
-# Using provably optimal 'constant escapement' policy instead of RL:
-
-import numpy as np
-import gym
-import gym_fishing
-from gym_fishing import models
-
-seed = 24
-np.random.seed(seed)
-
-env = gym.make("fishing-v1", sigma=0.1)
-optimal_policy = models.escapement(env)
-opt_sims = env.simulate(optimal_policy, reps=500)
-opt_policy = env.policyfn(optimal_policy)

### 2021-06-07 python/fishing_A2C_untuned.py

@@ -1,34 +1,0 @@
-# Fishing using (untuned) A2C
-
-import os
-import gym
-import gym_fishing
-import stable_baselines3 as sb3
-
-
-# Create a fishing environment
-env = gym.make("fishing-v1", sigma=0.1)
-
-# Create an A2C agent
-agent = sb3.A2C("MlpPolicy", env, seed=24)
-
-# Train
-if not os.path.exists("cache/a2c_untuned.zip"):
-    agent.learn(total_timesteps=300000)
-    agent.save("cache/a2c_untuned")
-
-# Load the saved agent
-agent = sb3.A2C.load("cache/a2c_untuned")
-
-# Evaluate
-eval_env = sb3.common.monitor.Monitor(env)
-mean, sd = sb3.common.evaluation.evaluate_policy(agent, eval_env)
-print("mean score:", mean)
-
-# Plot results
-agent_sims = env.simulate(agent, reps=100)
-agent_policy = env.policyfn(agent, reps=1)
-env.plot(agent_sims, "results/fishing_a2c_untuned.png")
-env.plot_policy(agent_policy, "results/fishing_a2c_untuned_policy.png")
-
-

### 2021-06-07 python/fishing_TD3_untuned.py

@@ -1,0 +1,39 @@
+import numpy as np
+import stable_baselines3 as sb3
+import os
+import gym
+import gym_fishing
+
+seed = 24
+
+# Create a fishing environment
+env = gym.make("fishing-v1", sigma=0.1)
+
+# Initialize the agent to train using TD3 algorithm
+agent = sb3.TD3("MlpPolicy", env, seed=24)
+
+# Train the agent
+if not os.path.exists("cache/td3_untuned.zip"):
+    agent.learn(total_timesteps=300000)
+    agent.save("cache/td3_untuned")
+
+# Load the saved agent
+agent = sb3.TD3.load("cache/td3_untuned")
+
+# Evaluate agent score using built-in methods:
+eval_env = sb3.common.monitor.Monitor(env)
+mean, sd = sb3.common.evaluation.evaluate_policy(agent, eval_env)
+print("mean score:", mean)
+
+# Simulate management under the trained agent (using deterministic policy)
+for i in range(0,10):
+  state, action = agent.predict(state, deterministic=True)
+  state = env.step(action)
+
+# Or gym_fishing provides additional helper methods for simulation & plotting
+td3_sims = env.simulate(agent, reps=100)
+td3_policy = env.policyfn(agent, reps=1)
+
+# Example of built-in plotting methods
+env.plot(td3_sims, "results/fishing_td3_tuned.png")
+env.plot_policy(td3_policy, "results/fishing_td3_tuned_policy.png")

### 2021-06-07 appendices/explore.Rmd

---
output: pdf_document
---

```{r setup, cache = FALSE, message=FALSE, include=FALSE}
# R dependencies
library(tidyverse)
library(patchwork)
library(reticulate)

## Python dependencies loaded via R
sb3      = import("stable_baselines3")
torch    = import("torch")
gym      = import("gym")
gym_fishing = import("gym_fishing")
gym_conservation = import("gym_conservation")

source("../R/plotting-helpers.R")
ggplot2::theme_set(ggplot2::theme_bw())
scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()

knitr::opts_chunk$set(echo=FALSE)
```

```{r}

sims_all <- function(cache, env) {

  names(cache) <- gsub("_\\w*tuned\\.zip", "", basename(cache))
  algos <- stringr::str_to_upper(names(cache))

  names(algos) <- algos
  sims_df <- map_dfr(algos, function(algo){
    i <- which(algo==algos)
    agent <- sb3[[algo]]$load(cache[[i]])
    env$simulate(agent, 100L)
  }, .id = "model")

  opt <- gym_fishing$models$escapement(env)
  opt_sims <- env$simulate(opt, reps = 100L)%>% mutate(model="optimal")
  bind_rows(opt_sims, sims_df) 
}

policyfns <- function(cache, env) {

  names(cache) <- gsub("_\\w*tuned\\.zip", "", basename(cache))
  algos <- stringr::str_to_upper(names(cache))
  names(algos) <- algos
  policy_df <- map_dfr(algos, function(algo){
    i <- which(algo==algos)
    agent <- sb3[[algo]]$load(cache[[i]])
    env$policyfn(agent, 1L)
  }, .id = "model")

  opt <- gym_fishing$models$escapement(env)
  opt_policy <- env$policyfn(opt)%>% mutate(model="optimal")
  bind_rows(opt_policy, policy_df) 
}

#  policy_df = env$policyfn(agent) %>% mutate(model="RL") %>% bind_rows(opt_policy)
```

# Deterministic Environment, Untuned agents

```{r}
env = gym$make("fishing-v1", sigma = 0)
untuned_det <- fs::dir_ls("../python/fishing/determinstic/cache/", regexp = ".*_untuned\\.zip")

df <- sims_all(untuned_det, env)
pf <- policyfns(untuned_det, env)
plot_sims(df) / (plot_policy(pf) + plot_reward(df))
```

# Stochastic Environment, Untuned Agents

```{r}
env = gym$make("fishing-v1", sigma = 0.1)
cache <- fs::dir_ls("../python/fishing/stochastic/cache/", regexp = ".*_untuned\\.zip")

df <- sims_all(cache, env)
pf <- policyfns(cache, env)
plot_sims(df) / (plot_policy(pf) + plot_reward(df))
```

# Deterministic Environment, Tuned Agents

```{r}
env = gym$make("fishing-v1", sigma = 0)
tuned_det <- fs::dir_ls("../python/fishing/determinstic/cache/", regexp = ".*_tuned\\.zip")

df <- sims_all(tuned_det, env)
pf <- policyfns(tuned_det, env)
plot_sims(df) / (plot_policy(pf) + plot_reward(df))
```

# Stochastic Environment, Tuned Agents

```{r}
env = gym$make("fishing-v1", sigma = 0.1)
tuned_stoch <- fs::dir_ls("../python/fishing/stochastic/cache/", regexp = ".*_tuned\\.zip")

df <- sims_all(tuned_stoch, env)
pf <- policyfns(tuned_stoch, env)
plot_sims(df) / (plot_policy(pf) + plot_reward(df))
```

### 2021-06-07 manuscript/manuscript.Rmd
Some edits and comments in the discussion here

--- 2021-06-07 manuscript/manuscript.Rmd
@@
 
-## Unsolved environments: Wildfires and other examples
+## Additional Environments
 
-<!-- ML: Need to rework this to talk about other environments that we are working on basically point to conservation gym website -->
-One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
-While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
-Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
-Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
-Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
-To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.
-
-<!-- Marcus add ~ paragraph discussion of the fire gym, citing motivating examples, basic overview of the state space, action space and reward and a screenshot of the simulation as a figure? -->
+Ecology holds many open problems for deep RL.  
+It is straight-forward to extend the simple environments presented here to reflect greater biological complexity, or to consider a host of more realistic decision scenarios.
+We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
+Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
 
@@
 <!-- Real problems are harder than these examples -->
-<!-- ML: I feel like this paragraph could be condensed with the one above -->
-<!-- CB: I think "model free is great" and "real problems are harder than these examples" are two distinct messages though and at least one would get lost in mashing them together.
-Perhaps the message doesn't come across in the current version.
-Does look like this paragraph is similar in overall message to the next one though, maybe could be combined with that?
--->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
-<!-- ML: A counterpoint here is that a model-based agent may learn a correct model that generalizes outside of the examples it has seen-->
+<!-- ML: I think we could delete this sentence and not lose anything.
+Another issue here is that the sentence implies that RL is not able to optimize the tipping point example becasue of not seeing the right state-action pairs ehh.
+-->
 In these simple examples in which the simulation implements a single model, training is analogous to classical methods which take the model as given [@Marescot2013].
 But classical approaches can be difficult to generalize when the underlying model is unknown: alternative models may require a completely different approach.
-In contrast, the process of training an RL algorithm on a more complex problem is no different than the simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
+In contrast, the process of training an RL algorithm on a more complex problem is no different than training on a simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
 Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
+<!-- ML: The second half of this paragraph doesn't quite drive home what I think is the intended point which is that RL can do well with model uncertainty and complexity.
+-->
 
@@
 As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
-Such examples provide important benchmarks against which to calibrate the performance, tuning and training requirements of RL,
-and can easily be extended into more complex problems once RL agents can master the basics.
-In the case that an agent has been trained to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
+Such examples provide important benchmarks to calibrate the performance, tuning and training requirements of RL.
+Once RL agents have mastered the basics, these examples can be easily extended into more complex problems by changing the environment.
+Yet, even in the case that an agent performs well on a realistic problem, there will still be a range of open questions in using deep RL to inform decision-making:
 
@@
 Advances in both computational power and algorithmic efficiency continue to reduce costs at a dramatic pace, significantly faster than Moore's law [@openai_costs].
-By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may both be better able to replicate that collaborative, iterative success in cracking hard problems.
+<!-- ML: I think we can let go of this sentence and not lose the point; I think we touch on it a bit in the last sentence of the first paragraph here also -->
+By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may be better able to replicate that collaborative, iterative success in cracking hard problems.
 Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.

### 2021-06-07 appendices/appendix-B-examples.Rmd
updated examples

@@
 
-output: 
-  rticles::arxiv_article:
-    keep_tex: true
+output: rticles::arxiv_article
 ---
@@
 This overview is neither a complete introduction to deep RL or a comprehensive analysis of RL solutions to these environments.
-
-```{r knitr, include = FALSE}
-knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4, cache = FALSE)
-ggplot2::theme_set(ggplot2::theme_bw())
-
-scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
-pal = ggthemes::solarized_pal()(8)
-txtcolor = "#586e75"
-
-colors <- set_names(c(pal[c(1,3,5,8)], txtcolor),
-                    c("Optimal", "TD3", "TD3_tuned", "RL", "historical"))
-scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
-scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
-```
 
@@
 source("../R/plotting-helpers.R")
+```
+
+```{r knitr, include = FALSE}
+# This chunk controls plot aesthetics only, and can be omitted with no material change
+knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
+                      fig.width = 7, fig.height = 4, cache = FALSE)
+ggplot2::theme_set(ggplot2::theme_bw())
+
+scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
+pal = ggthemes::solarized_pal()(8)
+txtcolor = "#586e75"
+
+colors <- set_names(c(pal[c(1,3,8,2)], txtcolor, pal[6]),
+                    c("Optimal", "TD3", "TD3_tuned", "RL", "historical", "steady-state"))
+scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
+scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
 ```
@@
 Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
-<!-- For example, here we will try the Actor-Critic algorithm A2C, training with the default hyperparameters originally posed in @A2C.-->
-We use this algorithm to train an agent that has a multi-layer perceptron ('MlpPolicy') neural network, with the default architecture of two 64-neuron layers.
+
+For example, here we will use the Twin Delayed DDPG method (TD3) described in @TD3.
+This algorithm is a successor of Deep Deterministic Policy Gradient (DDPG) method [@DDPG], itself a policy-gradient-modification that extends the original Deep Q-Network (DQN) algorithm [@DQN] to continuous action spaces.
+We use this algorithm to train an agent that has a multi-layer perceptron (`MlpPolicy`) neural network, with the default architecture of two 64-neuron layers.
 
@@
 
-colors <- set_names(c(pal[c(1,3,8,2)], txtcolor),
-                    c("Optimal", "TD3", "TD3_tuned", "RL", "historical"))
-scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
-scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
 plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
@@
 x0 = hake[[1, "scaled_biomass"]]
-agent_state = agent_action = numeric(Tmax)
+agent_state = agent_action = numeric(15)
 agent_state[1] = x0
@@
 ```{r}
-# Simulate management under the trained agent
-# See https://github.com/boettiger-lab/conservation-agents/blob//conservation/TD3-v6.py
-
 model = sb3$TD3$load("../python/cache/td3-conservation")
-TD3_sims = env$simulate(model, reps = 100L)
-TD3_policy = env$policyfn(model, reps = 10L)
+TD3_sims = env$simulate(model, reps = 100L) %>% mutate(model = "TD3_tuned")
+TD3_policy = env$policyfn(model, reps = 1L)  %>% mutate(model = "TD3_tuned")
 ```
@@
 opt = gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
-opt_sims = env$simulate(opt, reps = 100L)
-opt_policy = env$policyfn(opt)
+opt_sims = env$simulate(opt, reps = 100L) %>% mutate(model="steady-state")
+opt_policy = env$policyfn(opt) %>% mutate(model="steady-state")
 ```
@@
 ```{r}
-sims_df = bind_rows(TD3_sims, opt_sims, .id = "model") %>%
-  mutate(model = c("TD3", "steady-state")[as.integer(model)])
-
-policy_df = bind_rows(TD3_policy, opt_policy, .id = "model") %>%
-  mutate(model = c("TD3", "steady-state")[as.integer(model)])
+sims_df = bind_rows(TD3_sims, opt_sims)
+policy_df = bind_rows(TD3_policy, opt_policy)
 
@@
 ```{r fig3}
-plot_sims(sims_df) / ( plot_policy(policy_df) + plot_policy(sims_df))
+plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
 ```
+
+## Rescues: individual simulation trajectories
 
@@
 is_low = sims_df %>% 
-  filter(model == "TD3") %>% 
+  filter(model == "TD3_tuned") %>% 
   group_by(rep, model) %>% 
@@
 sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state,  group=interaction(model, rep))) +
-  geom_line(color = pal[3], show.legend = FALSE) + facet_wrap(~rep)
+  ggplot(aes(time, state,  col=model, group=interaction(model, rep))) +
+  geom_line(show.legend = FALSE) + facet_wrap(~rep)
 ```
@@
 sims_df %>% inner_join(is_low) %>%
-  ggplot(aes(time, state, group=interaction(model, rep))) +
-  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
+  ggplot(aes(time, state, col=model, group=interaction(model, rep))) +
+  geom_line(show.legend = FALSE) + facet_wrap(~rep)
 ```
@@
 ```
+
+\pagebreak
+
+# References

### 2021-06-07 appendices/appendix-D-data.Rmd
updated examples

--- 2021-04-29 manuscript/data-example.Rmd
@@
+---
+title: "Appendix D: Accessing Empirical Data for Management Comparison"
+authors:
+  - name: Marcus Lapeyrolerie
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+  - name: Melissa Chapman
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+  - name: Kari Norman
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+  - name: Carl Boettiger
+    department: Department of Environmental Science, Policy, and Management
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+    email: cboettig@berkeley.edu
+
+bibliography: ../manuscript/references.bib
+
+header-includes:
+  - \usepackage{amsmath}
+  - \usepackage{amssymb}
+
+output: rticles::arxiv_article
+---
+
+This appendix provides reproducible code used to access and clean data from the R.A. Myers Legacy Stock Assessment Database used in for empirical comparison in the hindcasts shown in the main text.
+
 ```{r knitr, include = FALSE}
@@
 
+Access data from the RAM Legacy Stock Assessment Database. (2020). Version v4.491 [Data set]. Zenodo. http://doi.org/10.5281/zenodo.3676088.
+Note that DOIs resolve only to HTML landing pages.  
+For programmatic access to the data, we resolve content by it's content hash instead:
+
 ```{r}
@@
 ```
+
+Relevant data is stored across several tables, following relational database practices.  
+Before analysis we join these tables by shared primary key:
 
@@
 
-Let's filter out missing data, non-matching units, and obvious reporting errors (catch exceeding total spawning biomass), then we re-scale each series into the 0,1 by appropriate choice of units:
+Let's filter out missing data, non-matching units, and obvious reporting errors (catch exceeding total spawning biomass), then we re-scale each series into the 0,1 by appropriate choice of units.
+We store the processed data locally for future use.
 
 ```{r}
-stock_ids <- c("PLAICNS", "ARGHAKENARG")
+stock_ids <- c("ARGHAKENARG")
 
@@
 ```
-
-```{r}
-# Simulate management under the trained agent
-env <- gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
-td3 <- sb3$TD3$load("cache/td3")
-
-```
-
-```{r}
-hake <- read_csv("../data/hake.csv")
-x0 <- hake %>%
- filter(year == min(year)) %>%
-  pull(scaled_biomass)
-Tmax <- 15
-years <-1986:2000
-
-td3_state <- td3_action <- numeric(Tmax)
-td3_state[1] <- x0
-td3_action[1] <- NA
-## represent the initial state size in the 'rescaled' state space.
-state <- env$get_state( x0 )
-```
-
-```{r}
-
-for(i in 2:Tmax){
-
-  # RL-recommended harvest action:
-  out <- td3$predict(state)
-  action <- out[[1]]
-
-  # Record state and resulting action
-  td3_state[i] <- env$get_fish_population(state)
-  td3_action[i] <- env$get_quota(action)
-
-  # Implement RL-recommended harvest:
-  result <- env$step(action)
-  state <- result[[1]]
-
-}
-
-df <- bind_rows(
-  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
-  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
-
-df %>% ggplot(aes(year, state, col=model)) + geom_line() + geom_point(aes(year, action, col=model))
-
-```
-
-```{r subset}
-examples <- df2 %>% 
-  ungroup() %>% 
-  group_by(commonname)
-```
-
-```{r}
-## Model does not estimate sigma_m; data is insufficient to do so.
-gs_code  <- nimble::nimbleCode({
-  r ~ dunif(0, 2)
-  K ~ dunif(0, 2)
-  sigma ~ dunif(0, 1)
-
-  x[1] <- x0
-  for(t in 1:(N-1)){
-    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K) - min(a[t],x[t])
-    x[t+1] ~ dnorm(mu[t], sd = sigma)
-  }
-})
-fit_models <- function(fish, code){
-  # fish <- examples %>% filter(stockid == stock_ids[1])
-
-  ## Rescale data
-  N <- dim(fish)[1]
-  scaled_data <- data.frame(t = 1:N, 
-                            y = fish$scaled_biomass, 
-                            a = fish$scaled_catch)
-  data = data.frame(x = scaled_data$y)
-  ## Compile  model
-  constants <- list(N = N, a = scaled_data$a)
-  inits <- list(r = 0.5, K = 0.5, sigma = 0.02, x0 = scaled_data$y[1])
-  model <- nimbleModel(code, constants, data, inits)
-  C_model <- compileNimble(model)
-
-  mcmcspec <- configureMCMC(model, thin = 1e2)
-  mcmc <- buildMCMC(mcmcspec)
-  Cmcmc <- compileNimble(mcmc, project = model)
-  Cmcmc$run(1e6)
-
-  samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
-  burnin <- 1:(0.05 * dim(samples)[1]) # drop first 5%
-  samples <- samples[-burnin,1:(length(inits) - 1)] # drop raised vars, burnin
-  #gather(samples) %>% ggplot() + geom_density(aes(value)) + facet_wrap(~key, scale='free')
-
-  ## Return fit
-  data.frame(stockid = fish$stockid[1],
-             commonname = fish$commonname[1],
-             r = mean(samples$r),
-             K = mean(samples$K),
-             sigma_g = mean(samples$sigma),
-             r_sd = sd(samples$r),
-             K_sd = sd(samples$K),
-             sigma_g_sd = sd(samples$sigma),
-             stringsAsFactors = FALSE)
-
-}
-```
-
-```{r}
-set.seed(123)
-fits <- examples %>% do(fit_models(., code=gs_code))
-fits 
-```
-
-```{r}
-pars <- fits %>% ungroup() %>% select(commonname, r, K, sigma_g) 
-pars
-```

### 2021-06-07 R/plotting-helpers.R

@@ -19,7 +19,7 @@ plot_policy <- function(policy_df)
                            group=interaction(rep, model),
                            col = model)) +
   geom_line(show.legend = FALSE) +
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9))
+  coord_cartesian(xlim = c(0, 1.2))
 
 plot_reward <- function(sims_df, gamma = 1)
   sims_df %>%

### 2021-06-07 appendices/appendix-B-examples.Rmd
updating manuscript plotting code

@@
 colors <- set_names(c(pal[c(1,3,8,2)], txtcolor, pal[6]),
-                    c("Optimal", "TD3", "TD3_tuned", "RL", "historical", "steady-state"))
+                    c("Optimal", "TD3_untuned", "TD3", "RL", "historical", "steady-state"))
 scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
@@
 sims_df <- agent_sims %>% 
-  mutate(model = "TD3") %>% 
+  mutate(model = "TD3_untuned") %>% 
   bind_rows(sims_df)
 policy_df <- agent_policy  %>% 
-  mutate(model = "TD3")  %>% 
+  mutate(model = "TD3_untuned")  %>% 
   bind_rows(policy_df)
@@
 sims_df <- td3_sims %>% 
-  mutate(model = "TD3_tuned") %>% 
+  mutate(model = "TD3") %>% 
   bind_rows(sims_df)
 policy_df <- td3_policy  %>% 
-  mutate(model = "TD3_tuned")  %>% 
+  mutate(model = "TD3")  %>% 
   bind_rows(policy_df)
@@
 years =1986:2000
-df = bind_rows(
-  tibble(year = years, harvest = agent_action, model = "TD3_tuned"),
+harvest = bind_rows(
+  tibble(year = years, harvest = agent_action, model = "TD3"),
   tibble(year = years, harvest = hake$scaled_catch, model = "historical")
@@
 stock <-  tibble(year = years, biomass = hake$scaled_biomass, model = "historical")
+
+write_csv(harvest, "../manuscript/figs/historical_harvest.csv")
+write_csv(stock, "../manuscript/figs/historical_stock.csv")
 
@@
         axis.ticks.x=element_blank())
-bottom_panel <- ggplot(df, aes(year, harvest, col=model)) + 
+bottom_panel <- ggplot(harvest, aes(year, harvest, col=model)) + 
   geom_line() + geom_point()
@@
 model = sb3$TD3$load("../python/cache/td3-conservation")
-TD3_sims = env$simulate(model, reps = 100L) %>% mutate(model = "TD3_tuned")
-TD3_policy = env$policyfn(model, reps = 1L)  %>% mutate(model = "TD3_tuned")
+TD3_sims = env$simulate(model, reps = 100L) %>% mutate(model = "TD3")
+TD3_policy = env$policyfn(model, reps = 1L)  %>% mutate(model = "TD3")
 ```
@@
 is_low = sims_df %>% 
-  filter(model == "TD3_tuned") %>% 
+  filter(model == "TD3") %>% 
   group_by(rep, model) %>% 

### 2021-06-07 manuscript/manuscript.Rmd
updating manuscript plotting code

@@
 library(patchwork)
-knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4)
+# This chunk controls plot aesthetics only, and can be omitted with no material change
+knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
+                      fig.width = 7, fig.height = 4, cache = FALSE)
 ggplot2::theme_set(ggplot2::theme_bw())
 
-scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(6)
-txtcolor <- "#586e75"
+scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
+pal = ggthemes::solarized_pal()(8)
+txtcolor = "#586e75"
+
+colors <- set_names(c(pal[c(1,3,8,2)], txtcolor, pal[6]),
+                    c("Optimal", "TD3_untuned", "TD3", "RL", "historical", "steady-state"))
+scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
+scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
+source("../R/plotting-helpers.R")
 ```
@@
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment." The simplest simulation is merely an implementation of the estimated model itself.
-However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
-<!-- ML: I think this sentence should be tightened up re RL, like one could use a model-based algorithm which uses an inputted model under the RL umbrella-->
+However, unlike an analytic or SDP solution that is derived directly from the estimated model, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-We will illustrate the potential differences between two RL algorithms used to train the agents, one based on the "Advantage Actor Critic" algorithm, usually referred to as A2C [@A2C], and one based on Twin Delayed Deep Deterministic Policy Gradient, commonly known as TD3 [@TD3] (a generalization of Deep Deterministic Policy Gradient, or DDPG, which extends the DQN algorithm @DQN to a continuous action space using the deterministic policy gradient).
-
+Here, we apply a version of the deep RL algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
@@
 ```{r message=FALSE, include=FALSE}
-sims_df <- read_csv( "figs/sims_df.csv")
-policy_df <- read_csv("figs/policy_df.csv")
-reward_df <- read_csv("figs/reward_df.csv")
+sims_df <- read_csv( "figs/sims_df.csv") %>% filter(model != "TD3_untuned")
+policy_df <- read_csv("figs/policy_df.csv") %>% filter(model != "TD3_untuned")
 
-ymin <- function(x) last(x[(ntile(x, 20)==1)])
-ymax <- function(x) last(x[(ntile(x, 20)==19)])
-
-fig_sims <- 
-sims_df %>% 
-  group_by(time, model) %>% 
-  summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
-
-fig_policy <- 
-  policy_df %>% ggplot(aes(state, state - action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE) + 
-  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) + 
-  ylab("escapement")
-
-fig_reward <- reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
 ```
 
-```{r fig_fish, fig.cap="Fisheries management using neural network agents trained with RL algorithms A2C and TD3 compared to optimal management.
+```{r fig_fish, fig.cap="Fisheries management using neural network agents trained with RL algorithm TD3 compared to optimal management.
 Top panel: mean fish population size over time across 100 replicates.
@@
 RL agents both adopt a policies that cease any harvest below this level, while allowing a somewhat higher escapement than optimal just above the optimal escapement level.
-A2C policy is non-deterministic, and overharvest from very high stock sizes.
-Lower Right: Mean cumulative reward over time across simulations.
-TD3 achieves nearly-optimal mean utility.
-Mean reward under A2C is lower, though the confidence interval still overlaps the mean reward of optimal management."}
-fig_sims / ( fig_policy + fig_reward)
+TD3 achieves nearly-optimal mean utility."}
+
+plot_sims(sims_df) / ( plot_policy(policy_df) + ylim(c(0,0.8)) + plot_reward(sims_df))
+
 ```
-
-Note that the A2C policy learns to shut down fishing entirely below a certain stock size, without fail.
-A2C also keeps the fishery closed even once stock sizes are slightly higher than $B_{MSY}$.
-Once it begins fishing, it slightly under-harvests for stock sizes a little over $B_{MSY}$, but tends to over-harvest for very large stock sizes (rarely ever realized, as fishing pressure is enough to keep it out of this range).
-Note also that the harvest varies across replicates, creating a fuzzy region in the policy function.
-Under such a policy, A2C maintains a stock size on average significantly above $B_{MSY}$, though also sees higher volatility particularly on the low side.
-Note that due to the stochastic nature of the environment, any given replicate A2C may out-perform the optimal policy, though on average the optimal policy provides a higher reward.
 
@@
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely take stock sizes close to a state of 1.0, the heavy overfishing behavior A2C has in this range has little consequence.
-Likewise it is more difficult for TD3 to learn the appropriate harvest level at these high states that are rarely sampled in the course of training.
-Just as intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
+Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at lower stock sizes.
+As intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
 Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
+
+<!-- 
+Add text about empirical example
+-->
+
+```{r message=FALSE}
+
+harvest <- read_csv( "../manuscript/figs/historical_harvest.csv")
+stock <- read_csv("../manuscript/figs/historical_stock.csv")
+
+top_panel <- ggplot(stock, aes(year, biomass, col=model)) +
+  geom_line(lty=2, show.legend = FALSE) + 
+  theme(axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()) + ggtitle("Argentine Hake")
+bottom_panel <- ggplot(harvest, aes(year, harvest, col=model)) + 
+  geom_line() + geom_point()
+
+top_panel / bottom_panel
+
+```
 
@@
 ```{r}
-sims_df <-read_csv("figs/tipping_sims_df.csv")
-policy_df <- read_csv("figs/tipping_policy_df.csv")
-reward_df <- read_csv("figs/tipping_reward_df.csv")
+tip_sims_df <-read_csv("figs/tipping_sims_df.csv")
+tip_policy_df <- read_csv("figs/tipping_policy_df.csv")
 ```
@@
 Lower right: mean rewards under TD3 management evenutally exceed those expected under the steady state strategy as a large initial investment in conservation eventually pays off."}
-fig_sims <- 
-  sims_df %>%
-   group_by(time, model) %>% 
-   summarise(ymin = ymin(state),
-            ymax = ymax(state),
-            state = mean(state), .groups = "drop") %>%
-  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
-   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
 
-fig_policy <- 
-  policy_df %>% ggplot(aes(state, action, 
-                           group=interaction(rep, model),
-                           col = model)) + 
-  geom_line(show.legend = FALSE)
-
-fig_reward <- reward_df %>% 
-  ggplot(aes(time, mean_reward)) + 
-  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
-                  ymax = mean_reward + 2*sd, fill = model),
-              alpha=0.25, show.legend = FALSE) + 
-  geom_line(aes(col = model), show.legend = FALSE) + 
-  ylab("reward")
-```
-
-```{r fig_conservation, fig.cap=""}
-fig_sims / ( fig_policy + fig_reward)
+plot_sims(tip_sims_df) / ( plot_policy(tip_policy_df) + plot_reward(tip_sims_df))
 ```

### 2021-06-08 manuscript/manuscript.Rmd
updated
fix gh-actions:
gh-actions cannot use cuda-based image: no space left on device

@@
 
-<!-- Something about the way forward, point to appendix resources etc -->
+\pagebreak
 
-<!-- Discussion really doesn't discuss/draw conclusions from the fisheries and tipping points examples! Do we need to or maybe not since it's not about the science-->
-
-<!-- ML: I think talking about the examples in a very passing manner is fine as the point is primarily to show that RL can work.
-To conclude, I feel like we could portray the vision of open ai gym environments building a CS-ecology community.-->
-
-<!-- 
-The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
-We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
-We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
-If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
-Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
-
--->
+# References

### 2021-06-08 manuscript/manuscript.Rmd
examples text and a few tweaks

@@
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy function and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
@@
 Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen. 
-Yet, the agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
-The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014], and current deep RL algorithms take a range of approaches towards this dilemma -- in Appendix A, we discuss some of the more specific exploration algorithms.
-Yet, overall, the manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
+The agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
+It is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
+This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014]. 
+Current deep RL algorithms take a range of approaches towards this (see  Appendix A).
+The manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
 The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
 For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent updates the same policy in training as the agent will use during evaluation.
-There are trade-offs to consider in using an off-policy or on-policy learner, which we will further discuss in the Appendix A.
-
-On top of the aforementioned distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
-One important point of divergence is whether the agent learns a deterministic or stochastic policy.
-Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
-In Appendices A and B, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.
+Other points of divergence include whether the agent learns a deterministic or stochastic policy, and
+whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
 
@@
 
-We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
+We provide two examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
 
@@
 Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
+
 Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
@@
 For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
-In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
+The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-However, unlike an analytic or SDP solution that is derived directly from the estimated model, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+In constrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-Here, we apply a version of the deep RL algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
+We illustrate the deep RL approach using the algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
+TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
 We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
-Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
+Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
 
@@
 
+The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
+Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
+Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
+This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
 These features highlight a common challenge in the design and training of RL algorithms.
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at lower stock sizes.
-As intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
+Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
 Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
 
-<!-- 
-Add text about empirical example
--->
+How would an RL agent be applied to empirical data?
+In principle, this is straight forward.
+Once our agent has been trained on appropriate simulation environments, it is able to propose a quota given an observation of the stock after appropriate transformation of variables.
+To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2000 [@ram, see Appendix D].
+Hake stocks showed a marked decline throughout this period, while harvests deceased only in proportion [Fig 3].
+In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
+While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.
 
@@
 
+This approach is not as different from conventional strategies at it may seem.
+In a conventional approach, ecological models are first estimated from empirical data, (stock assessments in the fisheries case).
+Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (the policy function) against model-based simulations of stock dynamics.
+This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.
+Instead of researchers evaluating a handful of control rules, the RL agent takes on the task of proposing and evaluating the millions of possible control rules represented by its neural network.
+
 ## Ecological Tipping Points \label{sec:tipping}
@@
 Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].
+
 We also enumerate the possible actions which a manager may take in response to environmental degradation.
@@
 
-A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix B, fig S1, S2) further highlights the difference between these two approaches.
+A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see appendix B] further highlights the difference between these two approaches.
 Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
@@
 In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
-In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery (Fig S2).
+In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
 Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
-
-Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
-The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
-<!-- ML: the parenthetical here I understand but I feel a little ehh on the word choice.
-I also don't know how to rephrase it -->
+The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
+Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
 Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
+Nevertheless, such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.
 

### 2021-06-08 manuscript/manuscript.Rmd
echo=FALSE

@@
 output: 
+  word_document: default
   rticles::arxiv_article:
@@
 # This chunk controls plot aesthetics only, and can be omitted with no material change
-knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
+knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
                       fig.width = 7, fig.height = 4, cache = FALSE)

### 2021-06-08 appendices/appendix-B-examples.Rmd
updated
data example -> appendix D

@@
 #install.packages("renv")
-#renv::restore()
+renv::restore()
 ```
@@
 
-## Applying an RL agent to real data
-
-We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.  
-Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
-Here, we load a local copy of the stock assessment for the Argentine Hake from 1986 to 2000.
-(Annotated code for data cleaning can be found in Appendix D).
-
-```{r message=FALSE}
-hake = read_csv("../data/hake.csv")
-```
-
-For each year in the data, we compare the historical harvest to the quota that would have been recommended by the TD3-based Deep RL agent we have just trained.
-To utilize our RL agent to set a quota for the stock, we only have to map the observed biomass estimates and historical harvests into the re-scaled state space of the environment. 
-Recall that the `gym_fishing` environment includes helper function `get_state()` to turn observed biomass to the state-space units used in the gym.
-
-```{r}
-# Initial states
-x0 = hake[[1, "scaled_biomass"]]
-agent_state = agent_action = numeric(15)
-agent_state[1] = x0
-agent_action[1] = NA
-
-# `env` is used only to translate between original units and transformed space
-# Represent the initial stock size in the 'rescaled' state space:
-state = env$get_state(x0)
-for(i in 1:15){
-
-  # The RL agent recommends a Quota based on current state
-  out = td3$predict(state)
-  action = out[[1]]
-
-  # We record state and proposed quota (in the original units)
-  agent_state[i] = env$get_fish_population(state)
-  agent_action[i] = env$get_quota(action)
-
-  # Use historical data to determine the next state
-  state = env$get_state(hake$scaled_biomass[i+1])
-}
-```
-
-Plot the results:
-
-```{r}
-years =1986:2000
-harvest = bind_rows(
-  tibble(year = years, harvest = agent_action, model = "TD3"),
-  tibble(year = years, harvest = hake$scaled_catch, model = "historical")
-)
-stock <-  tibble(year = years, biomass = hake$scaled_biomass, model = "historical")
-
-write_csv(harvest, "../manuscript/figs/historical_harvest.csv")
-write_csv(stock, "../manuscript/figs/historical_stock.csv")
-
-top_panel <- ggplot(stock, aes(year, biomass, col=model)) + geom_line(lty=2) + 
-  theme(axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
-bottom_panel <- ggplot(harvest, aes(year, harvest, col=model)) + 
-  geom_line() + geom_point()
-
-top_panel / bottom_panel
-```
-
-The Argentine hake biomass exhibited a steep decline between 1986 and 2000 as shown by the stock assessment estimates.
-Over that time, historical harvests showed some decline in step with the declining biomass.
-By contrast, for each of the historically observed biomass estimates the RL agent would have proposed much steeper cuts to the biomass quota which may have avoided the continued loss of biomass and eventual decline in harvest as well.
-This result should not be over-interpreted. 
-Without being able to replay history we have no way of truly knowing if the RL agent would have managed this stock better.
-While this example shows that it is in principle very straight forward to utilize a trained RL agent to recommend real harvest quotas, we must bear in mind that the agent is only as good as the environments on which it was trained.
-In practice, the RL agent would have to first be trained and evaluated on much more realistic environments than only the `fishing-v1` gym illustrated here.
-
 # Ecological tipping points

### 2021-06-08 manuscript/manuscript.Rmd
data example -> appendix D

@@
 Once our agent has been trained on appropriate simulation environments, it is able to propose a quota given an observation of the stock after appropriate transformation of variables.
-To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2000 [@ram, see Appendix D].
+To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
 Hake stocks showed a marked decline throughout this period, while harvests deceased only in proportion [Fig 3].

### 2021-06-08 manuscript/manuscript.Rmd
add pdfs

@@
 output: 
-  word_document: default
   rticles::arxiv_article:
     keep_tex: true
+  word_document: default
 editor_options: 

### 2021-06-08 appendices/appendix-A-definitions.Rmd
add pdfs
Moving exploration paragraph to the appendix

@@
 
-Exploration is a central component of any deep RL algorithm, and there are a variety of exploration algorithms that an agent can employ.
+Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit past high rewarding sequences.
 In the sections above, we procedurally mentioned that the agent must select actions during training.
 Yet, how the agent selects these actions is very important because if an agent explores the state-action space poorly, the agent will learn a sub-optimal policy.
-As mentioned in the manuscript, the dilemma in training is that it is difficult for the agent to know when to balance exploration of new actions and exploitation of past behavior.
+Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen.
+Yet, the agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
+The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior.
+This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014].
 For RL problems with small state-action spaces, e.g.
 space-action spaces that can be represented in a table, simple exploration algorithms will converge to the optimal policy, but with large or infinite state-action spaces, there are no convergence guarantees.
+
+Current deep RL algorithms take a range of approaches towards this dilemma.
 In the sections below, we will present some of the more commonly encountered exploration algorithms and then point to some more advanced methods.
@@
 
-# Replay Buffers vs. Parallel Workers
+# On-Policy vs. Off-policy Evaluation
 
-Another critical algorithm design choice involves how an RL agent collects and samples past experiences.
-A significant issue that arises from sampling past experiences is correlation of samples [@DQN, @schaul_prioritized_2016].
-In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated.
+Before deciding how to explore the environment, RL algorithms must have a policy to use during training.
+RL agents can either select actions from the same policy that they intend to use during evaluation or they can sample actions from a policy that is different.
+The former is called on-policy learning; the agent learns to select actions with the same policy it will use in testing.
+The latter is called off-policy learning; the agent uses a policy during training that is potentially different than the policy that the agent will evaluate.
+
+A significant issue that arises when sampling actions and then using this experience to maximize the RL objective is correlation [@DQN, @schaul_prioritized_2016].
+In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated[@DQN].
 Learning from correlated samples is undesirable since this can result in agents getting stuck in local optima, which will impede learning, but there are other reasons why learning from a single trajectory causes problems -- see @schaul_prioritized_2016 and @DQN for more insight here.
-There are two commonly used solutions to this problem of correlated samples: Replay Buffers and Parallel Workers.
+
+On-policy and off-policy agents have different solutions to the correlated samples problem.
 If the agent is off-policy, the agent can employ a replay buffer -- see @DQN for an example of a replay buffer-based agent.
-With a replay buffer, the agent will save state-action-reward transitions in the replay buffer, and during learning updates, the agent will sample transitions from the replay buffer.
-Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer will result in decorrelated samples.
-On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- note replay buffers are generally large enough to hold transitions from old policies.
+With a replay buffer, the agent saves state-action-reward transitions in the replay buffer, and during learning updates, the agent samples transitions from the buffer.
+Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer results in decorrelated samples.
+On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- note replay buffers are generally large enough to hold transitions from multiple policies.
 The solution to de-correlate samples for on-policy agents is to use parallel workers, whereby multiple workers explore the environment according to the current policy.
-Due to stochasticity that can come from exploration and the environment, the workers will go on different trajectories in state-action space.
+Due to stochasticity that can come from exploration and the environment, the workers go on different trajectories in state-action space.
 The agent will collect samples across these workers in a learning step, perform its learning step and then repeat the process of collecting and sampling transitions with an updated policy.

### 2021-06-08 manuscript/manuscript.Rmd
Moving exploration paragraph to the appendix

@@
     keep_tex: true
-  word_document: default
 editor_options: 
@@
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy function and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
 
-Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit past high rewarding sequences.
-Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen. 
-The agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
-It is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014]. 
-Current deep RL algorithms take a range of approaches towards this (see  Appendix A).
-The manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
-The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
-For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent updates the same policy in training as the agent will use during evaluation.
-Other points of divergence include whether the agent learns a deterministic or stochastic policy, and
-whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
+On top of the aforementioned distinctions of deep RL algorithms, there are more algorithms design choices that are made. 
+A very significant consideration is deciding how the agent will explore the environment.
+Another is whether the algorithm can work on a discrete, continuous or both a discrete and continuous state and action space.
+In Appendix A, we provide more detail on the important components of RL algorithms.
 
@@
 
-We provide two examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
+We provide three examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
 
@@
 # This chunk controls plot aesthetics only, and can be omitted with no material change
-knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE,
+knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
                       fig.width = 7, fig.height = 4, cache = FALSE)
@@
 Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
-
 Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
@@
 For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
-The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
+In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In constrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+However, unlike an analytic or SDP solution that is derived directly from the estimated model, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-We illustrate the deep RL approach using the algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
-TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
+Here, we apply a version of the deep RL algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
 A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
 We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
-Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
+Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
 
@@
 
-The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
-Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
-Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
-This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
 These features highlight a common challenge in the design and training of RL algorithms.
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
+Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at lower stock sizes.
+As intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
 Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
 
-How would an RL agent be applied to empirical data?
-In principle, this is straight forward.
-Once our agent has been trained on appropriate simulation environments, it is able to propose a quota given an observation of the stock after appropriate transformation of variables.
-To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
-Hake stocks showed a marked decline throughout this period, while harvests deceased only in proportion [Fig 3].
-In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
-While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.
+<!-- 
+Add text about empirical example
+-->
 
@@
 
-This approach is not as different from conventional strategies at it may seem.
-In a conventional approach, ecological models are first estimated from empirical data, (stock assessments in the fisheries case).
-Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (the policy function) against model-based simulations of stock dynamics.
-This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.
-Instead of researchers evaluating a handful of control rules, the RL agent takes on the task of proposing and evaluating the millions of possible control rules represented by its neural network.
-
 ## Ecological Tipping Points \label{sec:tipping}
@@
 Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].
-
 We also enumerate the possible actions which a manager may take in response to environmental degradation.
@@
 
-A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see appendix B] further highlights the difference between these two approaches.
+A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix B, fig S1, S2) further highlights the difference between these two approaches.
 Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
@@
 In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
-In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
+In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery (Fig S2).
 Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
-The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
-Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
+
+Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
+The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral-like features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
+<!-- ML: the parenthetical here I understand but I feel a little ehh on the word choice.
+I also don't know how to rephrase it -->
 Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
-Nevertheless, such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.
 
@@
 
+<!-- Something about the way forward, point to appendix resources etc -->
+
+<!-- Discussion really doesn't discuss/draw conclusions from the fisheries and tipping points examples! Do we need to or maybe not since it's not about the science-->
+
+<!-- ML: I think talking about the examples in a very passing manner is fine as the point is primarily to show that RL can work.
+To conclude, I feel like we could portray the vision of open ai gym environments building a CS-ecology community.-->
+
+<!-- 
+The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
+We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
+We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
+If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
+Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
+
+-->
+
 \pagebreak

### 2021-06-08 manuscript/manuscript.Rmd

@@
 
-<!-- Something about the way forward, point to appendix resources etc -->
-
-<!-- Discussion really doesn't discuss/draw conclusions from the fisheries and tipping points examples! Do we need to or maybe not since it's not about the science-->
-
-<!-- ML: I think talking about the examples in a very passing manner is fine as the point is primarily to show that RL can work.
-To conclude, I feel like we could portray the vision of open ai gym environments building a CS-ecology community.-->
-
-<!-- 
-The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
-We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
-We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
-If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
-Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.
-
--->
+\pagebreak
 

### 2021-06-08 manuscript/manuscript.Rmd
Added some spaces accidentally
Deleting empty lines
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

--- 2021-06-08 manuscript/manuscript.Rmd
@@
 We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy function and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
 
@@
 TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
-A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
+A step-by-step walk-through for training agents on this environment is provided in the Appendix.
 We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].

### 2021-06-08 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
-Hake stocks showed a marked decline throughout this period, while harvests deceased only in proportion [Fig 3].
+Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
 In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
@@
 This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.
-Instead of researchers evaluating a handful of control rules, the RL agent takes on the task of proposing and evaluating the millions of possible control rules represented by its neural network.
+Instead of researchers evaluating a handful of control rules, the RL agent proposes and evaluates a plethora of possible control rules autonomously.
 

### 2021-06-08 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 In a conventional approach, ecological models are first estimated from empirical data, (stock assessments in the fisheries case).
-Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (the policy function) against model-based simulations of stock dynamics.
+Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (policies) against model-based simulations of stock dynamics.
 This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.

### 2021-06-08 manuscript/manuscript.Rmd

@@
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy function and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
 
-Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit past high rewarding sequences.
-Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen. 
-The agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
-It is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014]. 
-Current deep RL algorithms take a range of approaches towards this (see  Appendix A).
-The manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
-The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
-For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent updates the same policy in training as the agent will use during evaluation.
-Other points of divergence include whether the agent learns a deterministic or stochastic policy, and
-whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
+On top of the aforementioned distinctions of deep RL algorithms, there are more algorithms design choices that are made. 
+A very significant consideration is deciding how the agent will explore the environment.
+Another is whether the algorithm can work on a discrete, continuous or both a discrete and continuous state and action space.
+In Appendix A, we provide more detail on the important components of RL algorithms.
 
@@
 For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
-The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
+In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In constrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+However, unlike an analytic or SDP solution that is derived directly from the estimated model, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-We illustrate the deep RL approach using the algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
-TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
-A step-by-step walk-through for training agents on this environment is provided in the Appendix.
+Here, we apply a version of the deep RL algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
+A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
 We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
-Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
+Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
 
@@
 
-The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
-Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
-Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
-This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
 These features highlight a common challenge in the design and training of RL algorithms.
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
+Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at lower stock sizes.
+As intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
 Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
 
-How would an RL agent be applied to empirical data?
-In principle, this is straight forward.
-Once our agent has been trained on appropriate simulation environments, it is able to propose a quota given an observation of the stock after appropriate transformation of variables.
-To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
-Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
-In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
-While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.
+<!-- 
+Add text about empirical example
+-->
 
@@
 
-This approach is not as different from conventional strategies at it may seem.
-In a conventional approach, ecological models are first estimated from empirical data, (stock assessments in the fisheries case).
-Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (policies) against model-based simulations of stock dynamics.
-This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.
-Instead of researchers evaluating a handful of control rules, the RL agent proposes and evaluates a plethora of possible control rules autonomously.
-
 ## Ecological Tipping Points \label{sec:tipping}

### 2021-06-09 manuscript/manuscript.Rmd
Fixing all the references that I could easily find

@@
 But will ML approaches really help us tackle a changing planet?
-Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
+Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph_neural_2020] and *unsupervised learning* [@valletta_applications_2017], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
 Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:
@@
 Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
-To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
+To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar]
 Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
@@
 Any application of RL can be divided into two components: an *environment* and an *agent*.
-An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
+An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@ha_learning_2020].
 The *agent*, which is often a computer program, continuously interacts with the environment. 
@@
 Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2016] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
+<!-- ML: This seems repetitive with second sentence in first paragraph above so going to leave this citation unfulfilled-->
 
@@
 This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, [@Marescot2013]), typically used to solve sequential decision-making problems.
-However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham; @Possingham; @morePossingham].
+However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham2004; @Possingham2006; @Possingham2009].
 A common conservation strategy employs a fixed response level rather than a dynamic policy which is toggled up or down each year: for example, declaring certain regions as protected areas in perpetuity.
@@
 Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
+<!-- ML: Having difficulty finding what the reference here should be.
+-->
 Model-free RL approaches offer a powerful way to factor this thorny issue.
@@
 <!-- Real problems are harder than these examples -->
+<!-- ML: I feel like this paragraph could be condensed with the one above -->
+<!-- CB: I think "model free is great" and "real problems are harder than these examples" are two distinct messages though and at least one would get lost in mashing them together.
+Perhaps the message doesn't come across in the current version.
+Does look like this paragraph is similar in overall message to the next one though, maybe could be combined with that?
+-->
 The examples presented here only scrape the surface of possible RL applications to conservation problems.
@@
 Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
-<!-- ML: I think we could delete this sentence and not lose anything.
-Another issue here is that the sentence implies that RL is not able to optimize the tipping point example becasue of not seeing the right state-action pairs ehh.
--->
+<!-- ML: A counterpoint here is that a model-based agent may learn a correct model that generalizes outside of the examples it has seen-->
 In these simple examples in which the simulation implements a single model, training is analogous to classical methods which take the model as given [@Marescot2013].
 But classical approaches can be difficult to generalize when the underlying model is unknown: alternative models may require a completely different approach.
-In contrast, the process of training an RL algorithm on a more complex problem is no different than training on a simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
+In contrast, the process of training an RL algorithm on a more complex problem is no different than the simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
 Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
-<!-- ML: The second half of this paragraph doesn't quite drive home what I think is the intended point which is that RL can do well with model uncertainty and complexity.
--->
 
@@
 As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
-Such examples provide important benchmarks to calibrate the performance, tuning and training requirements of RL.
-Once RL agents have mastered the basics, these examples can be easily extended into more complex problems by changing the environment.
-Yet, even in the case that an agent performs well on a realistic problem, there will still be a range of open questions in using deep RL to inform decision-making:
+Such examples provide important benchmarks against which to calibrate the performance, tuning and training requirements of RL,
+and can easily be extended into more complex problems once RL agents can master the basics.
+In the case that an agent has been trained to perform well on a realistic environment, there will still be a range of open questions in using this agent to inform decision-making:
 
@@
 Recent advances in the field have proven the potential of the approach to solve complex problems [@alphaGo2016; @alphaGoZero; @alphazero; @DQN], but typically leveraging large teams with decades of experience in ML and millions of dollars worth of computing power [@alphaGoZero].
-Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
-Leading developers of new deep RL algorithms benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [brockman2016].
+Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@popova_deep_2018; @zhou_optimizing_2017].
+Leading developers of new deep RL algorithms benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [@brockman2016].
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].
 Advances in both computational power and algorithmic efficiency continue to reduce costs at a dramatic pace, significantly faster than Moore's law [@openai_costs].
-<!-- ML: I think we can let go of this sentence and not lose the point; I think we touch on it a bit in the last sentence of the first paragraph here also -->
-By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may be better able to replicate that collaborative, iterative success in cracking hard problems.
+By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may both be better able to replicate that collaborative, iterative success in cracking hard problems.
 Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.

### 2021-06-09 appendices/appendix-B-examples.Rmd
Adding references for appendix B

--- 2021-06-07 appendices/appendix-B-examples.Rmd
@@
 For example, here we will use the Twin Delayed DDPG method (TD3) described in @TD3.
-This algorithm is a successor of Deep Deterministic Policy Gradient (DDPG) method [@DDPG], itself a policy-gradient-modification that extends the original Deep Q-Network (DQN) algorithm [@DQN] to continuous action spaces.
+This algorithm is a successor of Deep Deterministic Policy Gradient (DDPG) method [@ddpg], itself a policy-gradient-modification that extends the original Deep Q-Network (DQN) algorithm [@DQN] to continuous action spaces.
 We use this algorithm to train an agent that has a multi-layer perceptron (`MlpPolicy`) neural network, with the default architecture of two 64-neuron layers.
@@
 This search can also be automated using standard non-linear optimization techniques, adjusting hyperparameters to maximize the reward attained by the trained network.
-Using such an approach (see <https://github.com/boettiger-lab/conservation-agents>), we discover the following hyperparameters achieve better performance (see @TD3 or @SB3 documentation for formal definitions of each hyperparameter) :
+Using such an approach (see <https://github.com/boettiger-lab/conservation-agents>), we discover the following hyperparameters achieve better performance (see @TD3 or @sb3 documentation for formal definitions of each hyperparameter) :
 
@@
 This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carptenter2011] to the planet biosphere [@Barnosky2014].
+<!-- ML: Not sure what citation to use here for Barnosky or Carptenter in these years-->
 

### 2021-06-09 manuscript/manuscript.Rmd

--- 2021-06-08 manuscript/manuscript.Rmd
@@
 
-Over the course of training, RL agents will engage in trial and error learning, whereby it will be necessary for the agent to explore new sequences of actions as well as exploit past high rewarding sequences.
-Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen. 
-The agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
-It is very difficult for the agent to know when it should engage in explorative or exploitative behavior. 
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014]. 
-Current deep RL algorithms take a range of approaches towards this (see  Appendix A).
-The manner in which RL algorithms explore depend on whether the algorithms are *on-policy* or *off-policy*.
-The distinction between on and off-policy evaluation comes from what policy the agent is using to select actions during training.
-For off-policy algorithms, the agent uses a policy during training that can be different from the policy that the agent will use during evaluation; while in on-policy algorithms, the agent updates the same policy in training as the agent will use during evaluation.
-Other points of divergence include whether the agent learns a deterministic or stochastic policy, and
-whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
+On top of these distinctions of deep RL algorithms, there are other algorithms design choices to be made. 
+A significant consideration is deciding how the agent will explore the environment.
+Another is whether the algorithm can work on a discrete, or continuous state and action space.
+In Appendix A, we provide more detail on the important components of RL algorithms.
 

### 2021-06-09 appendices/appendix-D-data.Rmd

@@
 
-```{r knitr, include = FALSE}
-knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
-                      fig.width = 7, fig.height = 4, cache = TRUE)
-ggplot2::theme_set(ggplot2::theme_bw())
-
-scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
-scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
-pal <- ggthemes::solarized_pal()(8)
-txtcolor <- "#586e75"
-```
-
 ```{r setup, cache = FALSE, message=FALSE}
@@
 ## Python dependencies loaded via R
-sb3         <- import ("stable_baselines3")
-gym         <- import ("gym")
-gym_fishing <- import("gym_fishing")
+sb3      = import("stable_baselines3")
+torch    = import("torch")
+gym      = import("gym")
+gym_fishing = import("gym_fishing")
+gym_conservation = import("gym_conservation")
+
+source("../R/plotting-helpers.R")
+```
+
+```{r knitr, include = FALSE}
+# This chunk controls plot aesthetics only, and can be omitted with no material change
+knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
+                      fig.width = 7, fig.height = 4, cache = FALSE)
+ggplot2::theme_set(ggplot2::theme_bw())
+
+scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
+scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
+pal = ggthemes::solarized_pal()(8)
+txtcolor = "#586e75"
+
+colors <- set_names(c(pal[c(1,3,8,2)], txtcolor, pal[6]),
+                    c("Optimal", "TD3_untuned", "TD3", "RL", "historical", "steady-state"))
+scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
+scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
 ```
@@
 hake <- fish  %>% filter(commonname=="Argentine hake")
+
+hake %>% ggplot(aes(year, scaled_biomass)) + geom_point() + geom_line(aes(year, scaled_catch))
 write_csv(hake, "../data/hake.csv")
 ```
+
+## Applying an RL agent to real data
+
+We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.  
+Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
+Here, we load a local copy of the stock assessment for the Argentine Hake from 1986 to 2000.
+(Annotated code for data cleaning can be found in Appendix D).
+
+```{r message=FALSE}
+hake = read_csv("../data/hake.csv")
+td3 = sb3$TD3$load("../python/cache/td3_tuned.zip") 
+env = gym$make("fishing-v1")
+```
+
+For each year in the data, we compare the historical harvest to the quota that would have been recommended by the TD3-based Deep RL agent we have just trained.
+To utilize our RL agent to set a quota for the stock, we only have to map the observed biomass estimates and historical harvests into the re-scaled state space of the environment. 
+Recall that the `gym_fishing` environment includes helper function `get_state()` to turn observed biomass to the state-space units used in the gym.
+
+```{r}
+# Initial states
+x0 = hake[[1, "scaled_biomass"]]
+agent_state = agent_action = numeric(15)
+agent_state[1] = x0
+agent_action[1] = NA
+N <- dim(hake)[1]
+# `env` is used only to translate between original units and transformed space
+# Represent the initial stock size in the 'rescaled' state space:
+state = env$get_state(x0)
+for(i in 1:N){
+
+  # The RL agent recommends a Quota based on current state
+  out = td3$predict(state)
+  action = out[[1]]
+
+  # We record state and proposed quota (in the original units)
+  agent_state[i] = env$get_fish_population(state)
+  agent_action[i] = env$get_quota(action)
+
+  # Use historical data to determine the next state
+  state = env$get_state(hake$scaled_biomass[i+1])
+}
+```
+
+Plot the results:
+
+```{r}
+harvest = bind_rows(
+  tibble(year = hake$year, harvest = agent_action, model = "TD3"),
+  tibble(year = hake$year, harvest = hake$scaled_catch, model = "historical")
+)
+stock <-  tibble(year =hake$year, biomass = hake$scaled_biomass, model = "historical")
+
+write_csv(harvest, "../manuscript/figs/historical_harvest.csv")
+write_csv(stock, "../manuscript/figs/historical_stock.csv")
+
+top_panel <- ggplot(stock, aes(year, biomass, col=model)) + geom_line(lty=2) + 
+  theme(axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank())
+bottom_panel <- ggplot(harvest, aes(year, harvest, col=model)) + 
+  geom_line() + geom_point()
+
+top_panel / bottom_panel
+```
+
+The Argentine hake biomass exhibited a steep decline between 1986 and 2014 as shown by the stock assessment estimates.
+Over that time, historical harvests showed some decline in step with the declining biomass.
+By contrast, for each of the historically observed biomass estimates the RL agent would have proposed much steeper cuts to the biomass quota which may have avoided the continued loss of biomass and eventual decline in harvest as well.
+This result should not be over-interpreted. 
+Without being able to replay history we have no way of truly knowing if the RL agent would have managed this stock better.
+While this example shows that it is in principle very straight forward to utilize a trained RL agent to recommend real harvest quotas, we must bear in mind that the agent is only as good as the environments on which it was trained.
+In practice, the RL agent would have to first be trained and evaluated on much more realistic environments than only the `fishing-v1` gym illustrated here.

### 2021-06-09 manuscript/manuscript.Rmd

@@
 In principle, this is straight forward.
-Once our agent has been trained on appropriate simulation environments, it is able to propose a quota given an observation of the stock after appropriate transformation of variables.
+After we train an agent on a simulation environment that approximates the fishery of interest, we can query the policy of the agent to find a quota for the observed stock.    
 To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].

### 2021-06-09 manuscript/manuscript.Rmd
Cutting and moving RL stuff to appendix A.

@@
 @brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
-The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
-This allows the design and application of software agents which can interact and learn in that environment without knowing anything about the environment's internal details.
-The `gym` Python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
-The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').
-While this library of environments spans a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
-We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
+We use the `gym` framework to turn existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
 In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
@@
 Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
-The training period involves a trial and error process in which the agent inputs actions to the environment and uses the reward signal to learn an optimal policy. 
-According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its decisions.
 The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
-Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
-An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing anything about the environment's transition dynamics or the inner workings of the algorithm.
+Yet, overall, it has been well established that deep RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
 
-Hyperparameters refer to the parameters in a deep, RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
-Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
-Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with these hyperparameters, then examining the resulting performance.
-This procedure can be automated through standard multi-dimensional optimization approaches. 
-Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
-Hyperparameter tuning, however, is a necessary step in evaluating the performance of an algorithm and should be a standard component of any deep RL workflow.
+Hyperparameters refer to the parameters in a deep RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
+Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning, which can be automated through standard multi-dimensional optimization approaches.
+We further discuss and show the benefits of hyperparameter tuning in Appendix B.
 
@@
 
-Note that this formalism extends to an infinite horizon case when $H= \infty$.
-While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
-In the appendix, we will go into further detail on the general approaches that model-free deep RL algorithms take towards optimizing the RL objective.
+In Appendix B, we will go into greater detail on the general approaches that deep RL algorithms take towards optimizing the RL objective.
 
-# Deep RL Algorithms
+# Deep RL Overview <!-- ML: Might need to change the title to this section. It's not much of an overview imo. -->
 
@@
 
-There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
-
-To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
-Since our examples use model-free agents, we will focus on model-free algorithms in this section.
-
-Among model-free methods, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
-Generally, model-free agents are either learning a policy function, a value function or both.
-For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
-If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
-The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
-In contrast, *policy gradient* algorithms exclusively learn the policy.
-And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
-We go into greater depth on these different classes of RL algorithms in Appendix A.
-
 Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
@@
 Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
-Whenever there is a need for an agent to approximate some function, say a policy function or value function, neural networks can be used in this capacity.
+Whenever there is a need for an agent to approximate some function, say a policy, transition or value function^[Value functions estimate the expected return from a state or state-action pair.
+If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
+We talk more about value-based learning methods in Appendix A], neural networks can be used in this capacity.
 While there are other function approximators that can be used in RL, e.g.
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy function and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy, transition and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 
-On top of these distinctions of deep RL algorithms, there are other algorithms design choices to be made. 
+Beyond how neural networks are integrated, there are many other algorithm design choices to be made. 
 A significant consideration is deciding how the agent will explore the environment.

### 2021-06-09 appendices/appendix-A-definitions.Rmd
Ooops forgot to add appendix A

@@
   rticles::arxiv_article:
-    keep_tex: true
     extra_dependencies:
@@
 
-Policy gradients, value-based and actor-critic methods are the general classes of algorithms for gradient-based model-free RL.
-In this section, we will go over these algorithm classes to convey a sense of what they are attempting to achieve and the trade-offs between them.
+To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
+Since our examples use model-free agents, we will focus on model-free algorithms in this section.
+
+Among model-free methods, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
+In contrast, *policy gradient* algorithms exclusively learn the policy.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
+We will go into further detail on these different approaches below.
 

### 2021-06-09 manuscript/manuscript.Rmd
Deleting some sentences in the intro

@@
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
-We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
-We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
 
 In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
-Our second example considers ecosystem management under slow environmental degradation: a scenario which is not amenable to classic optimization methods such as dynamic programming.
-In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
-These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
+We present multiple examples which demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
 Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
 We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
-We further include implementations of three fully featured simulation modules following current standards for RL benchmark problems as Python modules distributed on PyPi and GitHub.
-Each module includes several more challenging variations to the examples described here.
-We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.
 

### 2021-06-09 appendices/appendix-A-definitions.Rmd
Keeping model-free bit in RL section; Also midway through discussion edits here

@@
 # Gradient-based Model-free RL Algorithms
-
-To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
-Since our examples use model-free agents, we will focus on model-free algorithms in this section.
 

### 2021-06-09 manuscript/manuscript.Rmd
Keeping model-free bit in RL section; Also midway through discussion edits here

@@
 
-# Formal RL Concepts
+# Reinforcement Learning
 <!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
@@
 
-In Appendix B, we will go into greater detail on the general approaches that deep RL algorithms take towards optimizing the RL objective.
-
-# Deep RL Overview <!-- ML: Might need to change the title to this section. It's not much of an overview imo. -->
-
-<!--
-
-Conceptual figure:
-
-Algorithm 
-  hyper-parameters specifying neural network, learning rate, etc.  
-  We refer to a particular selection of algorithm and hyperparameters as an "agent", which is then "trained" to solve one or more environments
-  internal 'parameters', estimated by "training" the agent.
-
--->
+To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
 
 Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Classic RL algorithms, like policy iteration, value iteration and TD-learning, fail when the state-action space is large, e.g.
-when the possible states and actions can not be represented in a tractable table [@suttonbarto].
-Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments [@DQN, @alphaGo2016].
-Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
 Whenever there is a need for an agent to approximate some function, say a policy, transition or value function^[Value functions estimate the expected return from a state or state-action pair.
 If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
-We talk more about value-based learning methods in Appendix A], neural networks can be used in this capacity.
+We talk more about value-based learning methods in Appendix A], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
 While there are other function approximators that can be used in RL, e.g.
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy, transition and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 
-Beyond how neural networks are integrated, there are many other algorithm design choices to be made. 
-A significant consideration is deciding how the agent will explore the environment.
-Another is whether the algorithm can work on a discrete, or continuous state and action space.
 In Appendix A, we provide more detail on the important components of RL algorithms.
@@
 Complexity comes not only from the ecological processes but also the available actions.
-Conservation efforts can be implemented in a wide variety of ways, and the most effective policies may vary in both space and time.
 While such complex dynamics and large state and action spaces are intractable to classical methods, they are straight forward to implement in simulation environments such as those used to train an RL algorithm.

### 2021-06-09 manuscript/manuscript.Rmd
Mostly changes to the discussion

@@
 output: 
+  word_document: default
   rticles::arxiv_article:
     keep_tex: true
-  word_document: default
 editor_options: 
@@
 
-In this brief section, we will introduce the concept of the RL environment as a partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.
-
-The reinforcement learning environment is typically formalized as a discrete-time POMDP.
+The reinforcement learning environment is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
 A POMDP is a 7-tuple that consists of the following:
@@
 If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
-We talk more about value-based learning methods in Appendix A], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+We talk more about value-based learning methods in Appendix A.], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
 While there are other function approximators that can be used in RL, e.g.
@@
 
-In Appendix A, we provide more detail on the important components of RL algorithms.
+In this paper, we will focus on model-free RL approaches that employ neural networks.
+We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
 
@@
 
-We illustrate the deep RL approach using the algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
+We illustrate the deep RL approach using the model-free algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
 TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
@@
 Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
-Model-free RL approaches offer a powerful way to factor this thorny issue.
-The model-free RL agent needs to make no such a priori hypothesis about a particular functional form.
-Model-free approaches have proven successful at learning effective policies even when the underlying model may be difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
+Model-free RL approaches offer a powerful solution for this thorny issue.
+Model-free algorithms have proven successful at learning effective policies even when the underlying model is difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
 
@@
 In the case of the ecosystem tipping point scenario, the optimal strategy is unknown, and the approximate solution found by our RL implementation could almost certainly be improved upon.
-Even at it's best, an RL-trained agent can only be as good as the scenarios on which it has been trained.
-<!-- ML: I think we could delete this sentence and not lose anything.
-Another issue here is that the sentence implies that RL is not able to optimize the tipping point example becasue of not seeing the right state-action pairs ehh.
--->
 In these simple examples in which the simulation implements a single model, training is analogous to classical methods which take the model as given [@Marescot2013].
-But classical approaches can be difficult to generalize when the underlying model is unknown: alternative models may require a completely different approach.
+But classical approaches can be difficult to generalize when the underlying model is unknown.
 In contrast, the process of training an RL algorithm on a more complex problem is no different than training on a simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
-Many of our most realistic ecological models exist only as numerical simulations of lower-level dynamics, such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
-<!-- ML: The second half of this paragraph doesn't quite drive home what I think is the intended point which is that RL can do well with model uncertainty and complexity.
--->
+This flexibility of RL, for instance, could allow us to use RL agents on our most realistic ecological models such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
 
@@
 Successfully applying RL to complex ecological problems is no easy task.
-The flexibility of the approach makes it relatively straightforward to include realistic biology and environmental science in framing an RL problem.
-Yet, even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters].
-As the examples we have presented illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
+Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters].
+As Section 5.1 and 5.2 illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
 Such examples provide important benchmarks to calibrate the performance, tuning and training requirements of RL.
-Once RL agents have mastered the basics, these examples can be easily extended into more complex problems by changing the environment.
+Once RL agents have mastered the basics, the examples can be easily extended into more complex problems by changing the environment.
 Yet, even in the case that an agent performs well on a realistic problem, there will still be a range of open questions in using deep RL to inform decision-making:
@@
 
+- Will using software-based agents create power assymetries between those developing the agents and those that rely on the natural resources at stake?
+
 <!-- But never fear, hard problems have been solved before!-->
@@
 Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
-Leading developers of new deep RL algorithms benchmark the performance of their algorithms against such problems in part because simulations of games and robotic movement are readily available through open source frameworks designed to interface with RL algorithms [brockman2016], and performance can easily be compared against alternatives on public leaderboards [brockman2016].
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].
-Advances in both computational power and algorithmic efficiency continue to reduce costs at a dramatic pace, significantly faster than Moore's law [@openai_costs].
-<!-- ML: I think we can let go of this sentence and not lose the point; I think we touch on it a bit in the last sentence of the first paragraph here also -->
 By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may be better able to replicate that collaborative, iterative success in cracking hard problems.
-Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.
-If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.
 
@@
 
-The increasingly influential role of AI in sectors as diverse as policing, finance, and employment has raised concern about the ethics, biases, and power entrenched in both decision-making algorithms and their outputs [@kalluri2020].
-As algorithms play a greater role in environmental and conservation policy, will similar issues arise?
-While this question has garnered  recent interest in the academic literature [@wearn2019; @adams2019; @scoville2021], the use of decision making algorithms in environmental contexts is still relatively isolated and in it's early stages.
-Examples of AI applications, and corresponding concerns, have largely been confined to the context of enforcing environmental laws (i.e.
+<!-- The increasingly influential role of AI in sectors as diverse as policing, finance, and employment has raised concern about the ethics, biases, and power entrenched in both decision-making algorithms and their outputs [@kalluri2020].
+-->
+<!-- As algorithms play a greater role in environmental and conservation policy, will similar issues arise? -->
+<!-- While this question has garnered  recent interest in the academic literature [@wearn2019; @adams2019; @scoville2021], the use of decision making algorithms in environmental contexts is still relatively isolated and in it's early stages.
+-->
+<!-- Examples of AI applications, and corresponding concerns, have largely been confined to the context of enforcing environmental laws (i.e.
 algorithms which predict wildlife poaching events [@adams2019] or identify illegal fishing behavior [@mcdonald2021; @swartz2021]).
-However, more diverse applications of AI algorithms - such as applications which contribute to meeting sustainable development goals [@vinuesa2020] or designing  protected area networks - raise additional concerns of entrenching power asymmetries between those reliant on natural resources and those designing algorithms and their objectives.
-These issues of power become particularly acute when private industry holds the keys to propriety algorithms or environmental data, and generally operate without the same community engagement obligations of public agencies.
-In these cases, good intentions might not be sufficient to ensure equitable outcomes. 
-Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges promotes equity in both the procedural process of developing objective functions and the distribution of benefits that result from decisions.
+-->
+<!-- However, more diverse applications of AI algorithms - such as applications which contribute to meeting sustainable development goals [@vinuesa2020] or designing  protected area networks - raise additional concerns of entrenching power asymmetries between those reliant on natural resources and those designing algorithms and their objectives.
+-->
+<!-- These issues of power become particularly acute when private industry holds the keys to propriety algorithms or environmental data, and generally operate without the same community engagement obligations of public agencies.
+-->
+<!-- In these cases, good intentions might not be sufficient to ensure equitable outcomes. -->
+<!-- Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges promotes equity in both the procedural process of developing objective functions and the distribution of benefits that result from decisions.
+-->
 

### 2021-06-09 manuscript/manuscript.Rmd
Fixing one ref

--- 2021-06-09 manuscript/manuscript.Rmd
@@
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
+We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
+We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.
 
 In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-We present multiple examples which demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
+Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
+Our second example considers ecosystem management under slow environmental degradation: a scenario which is not amenable to classic optimization methods such as dynamic programming.
+In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
+These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
 Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
 We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+We further include implementations of three fully featured simulation modules following current standards for RL benchmark problems as Python modules distributed on PyPi and GitHub.
+Each module includes several more challenging variations to the examples described here.
+We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.
 
@@
 @brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
-We use the `gym` framework to turn existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
+The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
+This allows the design and application of software agents which can interact and learn in that environment without knowing anything about the environment's internal details.
+The `gym` Python module created by OpenAI also provides over 100 pre-built environments that serve as benchmark problems against which new RL algorithms can be tested.
+The `gym` framework is now recognized by all major software frameworks for writing RL algorithms (see Appendix, 'Deep RL Frameworks').
+While this library of environments spans a range of difficulties from classic control problems that can be solved in closed form to complex three-dimensional robotics simulators, they include no examples of ecological or biological problems.
+We use the `gym` framework to turn several existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
 In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
@@
 Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
+The training period involves a trial and error process in which the agent inputs actions to the environment and uses the reward signal to learn an optimal policy. 
+According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its decisions.
 The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
-Yet, overall, it has been well established that deep RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
+Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
+An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing anything about the environment's transition dynamics or the inner workings of the algorithm.
 
-Hyperparameters refer to the parameters in a deep RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
-Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning, which can be automated through standard multi-dimensional optimization approaches.
-We further discuss and show the benefits of hyperparameter tuning in Appendix B.
+Hyperparameters refer to the parameters in a deep, RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
+Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning.
+Hyperparameter tuning consists of the following general steps in repetition: proposing a new set of hyperparameters, training an agent with these hyperparameters, then examining the resulting performance.
+This procedure can be automated through standard multi-dimensional optimization approaches. 
+Since one round of training can be time intensive, hyperparameter tuning, which consists of usually numerous rounds of training, can be particularly time intensive.
+Hyperparameter tuning, however, is a necessary step in evaluating the performance of an algorithm and should be a standard component of any deep RL workflow.
 
-# Reinforcement Learning
+# Formal RL Concepts
 <!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
@@
 
-To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
+Note that this formalism extends to an infinite horizon case when $H= \infty$.
+While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
+In the appendix, we will go into further detail on the general approaches that model-free deep RL algorithms take towards optimizing the RL objective.
+
+# Deep RL Algorithms
+
+<!--
+
+Conceptual figure:
+
+Algorithm 
+  hyper-parameters specifying neural network, learning rate, etc.  
+  We refer to a particular selection of algorithm and hyperparameters as an "agent", which is then "trained" to solve one or more environments
+  internal 'parameters', estimated by "training" the agent.
+
+-->
+
+There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
+
+To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
 The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
@@
 Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
+Since our examples use model-free agents, we will focus on model-free algorithms in this section.
+
+Among model-free methods, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
+Generally, model-free agents are either learning a policy function, a value function or both.
+For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
+If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
+The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
+In contrast, *policy gradient* algorithms exclusively learn the policy.
+And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
+We go into greater depth on these different classes of RL algorithms in Appendix A.
 
 Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Whenever there is a need for an agent to approximate some function, say a policy, transition or value function^[Value functions estimate the expected return from a state or state-action pair.
-If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
-We talk more about value-based learning methods in Appendix A], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+Classic RL algorithms, like policy iteration, value iteration and TD-learning, fail when the state-action space is large, e.g.
+when the possible states and actions can not be represented in a tractable table [@suttonbarto].
+Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments [@DQN, @alphaGo2016].
+Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
+Whenever there is a need for an agent to approximate some function, say a policy function or value function, neural networks can be used in this capacity.
 While there are other function approximators that can be used in RL, e.g.
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
+We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
+This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
 
+On top of the aforementioned distinctions of deep RL algorithms, there are more algorithms design choices that are made. 
+A very significant consideration is deciding how the agent will explore the environment.
+Another is whether the algorithm can work on a discrete, continuous or both a discrete and continuous state and action space.
 In Appendix A, we provide more detail on the important components of RL algorithms.
@@
 Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2016] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
+<!-- ML: This seems repetitive with second sentence in first paragraph above so going to leave this citation unfulfilled-->
 
 For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
-The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
+In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In constrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+However, unlike an analytic or SDP solution that is derived directly from the estimated model, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-We illustrate the deep RL approach using the algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
-TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
-A step-by-step walk-through for training agents on this environment is provided in the Appendix.
+Here, we apply a version of the deep RL algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
+A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
 We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
-Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
+Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
 
@@
 
-The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
-Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
-Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
-This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
 These features highlight a common challenge in the design and training of RL algorithms.
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
+Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at lower stock sizes.
+As intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
 Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
 
-How would an RL agent be applied to empirical data?
-In principle, this is straight forward.
-After we train an agent on a simulation environment that approximates the fishery of interest, we can query the policy of the agent to find a quota for the observed stock.    
-To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
-Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
-In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
-While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.
+<!-- 
+Add text about empirical example
+-->
 
@@
 Complexity comes not only from the ecological processes but also the available actions.
+Conservation efforts can be implemented in a wide variety of ways, and the most effective policies may vary in both space and time.
 While such complex dynamics and large state and action spaces are intractable to classical methods, they are straight forward to implement in simulation environments such as those used to train an RL algorithm.
@@
 Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
-Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
+Frequently available ecological data is insufficient to distinguish between possible alternative models [@knape_are_2012], or the correct model may be non-identifiable with any amount of data.
 Model-free RL approaches offer a powerful way to factor this thorny issue.

### 2021-06-09 manuscript/manuscript.Rmd

@@
     keep_tex: true
+  word_document: default
 editor_options: 
@@
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
-This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively
+We will not discuss in detail how neural networks work [see @lecun_deep_2015], but the general process for how neural networks are used in RL is that at each learning step, the agent, will adjust the parameters in its neural network with which it approximates the policy function and/or value function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future.
 
-On top of the aforementioned distinctions of deep RL algorithms, there are more algorithms design choices that are made. 
-A very significant consideration is deciding how the agent will explore the environment.
-Another is whether the algorithm can work on a discrete, continuous or both a discrete and continuous state and action space.
+On top of these distinctions of deep RL algorithms, there are other algorithms design choices to be made. 
+A significant consideration is deciding how the agent will explore the environment.
+Another is whether the algorithm can work on a discrete, or continuous state and action space.
 In Appendix A, we provide more detail on the important components of RL algorithms.
@@
 For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
-In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
+The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
 In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-However, unlike an analytic or SDP solution that is derived directly from the estimated model, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+In constrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
-Here, we apply a version of the deep RL algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
-A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
+We illustrate the deep RL approach using the algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
+TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
+A step-by-step walk-through for training agents on this environment is provided in the Appendix.
 We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
-Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.
+Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
 
@@
 
+The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
+Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
+Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
+This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
 These features highlight a common challenge in the design and training of RL algorithms.
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at lower stock sizes.
-As intuition would suggest, it is possible to devise alternative training schemes which may do a better job sampling the simulation dynamics at high stock levels.
+Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
 Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
 
-<!-- 
-Add text about empirical example
--->
+How would an RL agent be applied to empirical data?
+In principle, this is straight forward.
+After we train an agent on a simulation environment that approximates the fishery of interest, we can query the policy of the agent to find a quota for the observed stock.    
+To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
+Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
+In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
+While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.
 
@@
 
+This approach is not as different from conventional strategies at it may seem.
+In a conventional approach, ecological models are first estimated from empirical data, (stock assessments in the fisheries case).
+Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (policies) against model-based simulations of stock dynamics.
+This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.
+Instead of researchers evaluating a handful of control rules, the RL agent proposes and evaluates a plethora of possible control rules autonomously.
+
 ## Ecological Tipping Points \label{sec:tipping}

### 2021-06-09 manuscript/manuscript.Rmd

--- 2021-06-09 manuscript/manuscript.Rmd
@@
 Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
-Frequently available ecological data is insufficient to distinguish between possible alternative models [@devalpine], or the correct model may be non-identifiable with any amount of data.
+Frequently available ecological data is insufficient to distinguish between possible alternative models [@knape_are_2012], or the correct model may be non-identifiable with any amount of data.
 Model-free RL approaches offer a powerful solution for this thorny issue.
@@
 Recent advances in the field have proven the potential of the approach to solve complex problems [@alphaGo2016; @alphaGoZero; @alphazero; @DQN], but typically leveraging large teams with decades of experience in ML and millions of dollars worth of computing power [@alphaGoZero].
-Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@protein_folding].
+Successes have so far been concentrated in applications to games and robotics, not scientific and policy domains, though this is beginning to change [@popova_deep_2018; @zhou_optimizing_2017].
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].

### 2021-06-10 manuscript/manuscript.Rmd

@@
 We illustrate the deep RL approach using the model-free algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
-TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @DDPG], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
+TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @ddpg], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
 A step-by-step walk-through for training agents on this environment is provided in the Appendix.
@@
 After we train an agent on a simulation environment that approximates the fishery of interest, we can query the policy of the agent to find a quota for the observed stock.    
-To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ram, see Appendix D].
+To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ramlegacy, see Appendix D].
 Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
@@
 Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (policies) against model-based simulations of stock dynamics.
-This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @Punt2016] is already very closely analogous to the RL process.
+This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @punt_management_2016] is already very closely analogous to the RL process.
 Instead of researchers evaluating a handful of control rules, the RL agent proposes and evaluates a plethora of possible control rules autonomously.
@@
 Once RL agents have mastered the basics, the examples can be easily extended into more complex problems by changing the environment.
-Yet, even in the case that an agent performs well on a realistic problem, there will still be a range of open questions in using deep RL to inform decision-making:
-
-- Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the deep RL agent will generalize its past experience to new situations?
-
-- Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?
-
-- Will using software-based agents create power assymetries between those developing the agents and those that rely on the natural resources at stake?
+Yet, even in the case that an agent performs well on a realistic problem, there will be a range of open questions in using deep RL to inform decision-making. 
+Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the deep RL agent will generalize its past experience to new situations?
+Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?
+A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021, @chapman_promoting_2021].
 

### 2021-06-10 manuscript/manuscript.Rmd
manuscript edits up to examples

@@
 <!-- Environmental problems, ML -->
-Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@Worm2006], declining biodiversity [@biodiversity], and zoonotic pandemics [@covid].
-But will ML approaches really help us tackle a changing planet?
-Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph_neural_2020] and *unsupervised learning* [@valletta_applications_2017], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
-Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:
+Advances in both available data and computing power are opening the door for machine learning (ML) to play a greater role in addressing some of our planet’s most pressing environmental problems. 
+But will ML approaches really help us tackle our most pressing environmental problems?
+From the growing frequency and intensity of wildfire [@wildfire], to over-exploited fisheries [@Worm2006] and declining biodiversity [@biodiversity], to emergent zoonotic pandemics [@covid], the diversity and scope of environmental problems are unprecedented.
+Applications of ML in ecology have to-date illustrated the promise of two methods: *supervised learning* [@joseph_neural_2020] and *unsupervised learning* [@valletta_applications_2017].
+However, the fields of ecology and conservation science have so far overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
+Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing issues of global ecological change:
 
@@
 
-Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in these fields.
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar]
 Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
-RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
+RL agents typically develop strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
@@
 In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-We present multiple examples which demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
+We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
 Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
@@
 
-Any application of RL can be divided into two components: an *environment* and an *agent*.
-An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@ha_learning_2020].
+All applications of RL can be divided into two components: an *environment* and an *agent*.
+The *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@ha_learning_2020].
 The *agent*, which is often a computer program, continuously interacts with the environment. 
-At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
-The goal of the agent is to learn how to maximize its expected cumulative reward.
+At each time step, the agent observes the current *state* of the environment then performs an *action*.
+As a result of this action, the environment transitions to a new state and transmits a numerical *reward* signal to the agent.
+The goal of the agent is to learn how to maximize its expected cumulative reward in the environment.
 The agent learns how to achieve this objective during a period called *training*.
@@
 @brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
+<!-- this reads a little abrupt - might be worth transitioning a bit to these challenges -->
+
 We use the `gym` framework to turn existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
@@
 
-Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
+Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands to millions of time steps.
 The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
@@
 <!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
+
+<!-- Mc: I agree-- could you intregrate this into the other parts? Like when you introduce states and actions rather than having a separate section? -->
 
@@
 The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
+
+<!-- yeah this is pretty repetative... I dont think you necessarily need to avoid some of the formal notation in the first subsection-->
+
 At each time step, the agent makes an observation of the environment at time t, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
@@
 We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
+
+<!-- I really like the last few paragraphs here!
+I do feel like what is above model-free vs model based paragrph is repetative and could just be introduced with the concepts in the previous section.
+Then you could make this whole subsection about the distrinction between model free and model based?
+-->
 

### 2021-06-10 manuscript/manuscript.Rmd
a couple more edits!

@@
 The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
-Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and also a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
+Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
 
@@
 <!-- ML: This seems repetitive with second sentence in first paragraph above so going to leave this citation unfulfilled-->
+<!-- ^MC:+1 to this sentence seeming repetative -->
 
@@
 From this model, the optimal harvest policy -- that is, the function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
-In constrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
+In contrast, a model-free deep RL algorithm makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.
 
@@
 
-This approach is not as different from conventional strategies at it may seem.
+This approach is not as different from conventional strategies as it may seem.
 In a conventional approach, ecological models are first estimated from empirical data, (stock assessments in the fisheries case).
 Quotas can then be set based directly on these model estimates, or by comparing alternative candidate "harvest control rules" (policies) against model-based simulations of stock dynamics.
-This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @punt_management_2016] is already very closely analogous to the RL process.
+This latter approach, known in fisheries as Management Strategy Evaluation [MSE; @punt_management_2016] is already closely analogous to the RL process.
 Instead of researchers evaluating a handful of control rules, the RL agent proposes and evaluates a plethora of possible control rules autonomously.
@@
 In the absence of any management response, we assume the environment deteriorates at a fixed rate $\alpha$, which can be thought of as the incremental increase in global mean temperature or similar anthropogenic forcing term.
-Management can respond to slow or even reverse this trend by choosing an opposing action $A_t$.
+Management can slow or even reverse this trend by choosing an opposing action $A_t$.
 We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $A_t$ as equal to $c A_t^2$.
@@
 Iterative improvements to well-posed public challenges have proven immensely effective in the computer science community in tackling difficult problems, which allow many teams with diverse expertise not only to compete but to learn from each other [@netflix_prize; @imagenet].
-By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may be better able to replicate that collaborative, iterative success in cracking hard problems.
+By working to develop similarly well-posed challenges as clear benchmarks, ecology and environmental science researchers may be able to replicate that collaborative, iterative success in cracking hard problems.
 

### 2021-06-10 manuscript/presubmission-letter.Rmd
Edits presubmission letter

--- 2021-04-29 manuscript/presubmission-letter.Rmd
@@
 
-To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
-*Nature* recently featured RL agents achieving superhuman performance on classic Atari games, and *Science* described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
+To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and games.
+*Nature* recently featured RL agents achieving superhuman performance on classic Atari games, *Science* described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
 We believe that similar techniques could improve decision-making in managing ecological environments as well.
-<!-- Having trouble reconciling lines 21-23, trying to say that these simulators are like the grand challenges and given RL's track record of solving intractable decision-making problems, perhaps RL could shed light here?
-I think having a sentence asserting that RL would use ecological models not replace them would be good somewhere in here.--> Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
-<!-- Kari: possible rewrite of the following sentence, not sure if it helps --> While existing process-based models allow a window into the future scenarios of ecological management, RL has the potential to find optimal solutions in those spaces.
+Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
+Detailed, process-based models allow a window into the future scenarios of ecological management.
 Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.

### 2021-06-10 manuscript/manuscript.Rmd
Minor comments

@@
 Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands to millions of time steps.
-The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
+The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute <!-- Kari: is this supposed to be "dedicated compute time"?
+Not sure what dedicated compute is -->, complexity of the environment, complexity of the agent and more.
 Yet, overall, it has been well established that deep RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
@@
 Rarely do we know the underlying functional forms for ecological processes -- is population recruitment more Beverton-Holt or Ricker?
+<!-- I don't know what those things are - might not be the place to introduce new concepts even if they're off hand?
+Or am I the only one that doesn't recognize them?
+-->
 Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.

### 2021-06-10 manuscript/manuscript.Rmd
Shifting around; also had some overlooked mistakes in the formal section

--- 2021-06-10 manuscript/manuscript.Rmd
@@
 All applications of RL can be divided into two components: an *environment* and an *agent*.
-The *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@ha_learning_2020].
-The *agent*, which is often a computer program, continuously interacts with the environment. 
-At each time step, the agent observes the current *state* of the environment then performs an *action*.
-As a result of this action, the environment transitions to a new state and transmits a numerical *reward* signal to the agent.
-The goal of the agent is to learn how to maximize its expected cumulative reward in the environment.
+
+The *environment* is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
+A POMDP is a tuple that consists of the following:
+
+- $\mathcal{S}$: a set of states called the state space
+
+- $\mathcal{A}$: a set of actions called the action space
+
+- $\Omega$ : a set of observations called the observation space
+
+- $E(o_{t}| s_{t})$: an emission distribution, which accounts for an agent's observation being different from environment's state
+
+- $T(s_{t+1}|s_t, a_t)$: a state transition operator which describes the dynamics of the system
+
+- $r(s_t, a_t)$: a reward function
+
+- $d_0(s_0)$: an initial state distribution
+
+- $\gamma \in (0,1]$: a discount factor
+
+The *agent* interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
+At each time step, the agent makes an observation of the environment, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
+The action will cause the environment to transition to a new state according to the state transition operator. 
+The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
+This process will repeat until a terminal state is reached, after which the environment can reset to an initial state^[While there are RL methods for infinite horizon problems, i.e.
+when $H=\infty$, we will only present methods for episodic or finite horizon POMDPs.].
+As the agent interacts with the environment, the agent will create a trajectory given as $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$.
+From these definitions, we can provide an agent's trajectory distribution for a given policy as,
+
+$$
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^{H-1} \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t) .
+$$
+
+The goal of reinforcement learning is for the agent to find an optimal policy distribution, $\pi^*(a_t|o_t)$, that maximizes the expected return, $J(\pi)$:
+
+$$
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi) .
+$$
+
 The agent learns how to achieve this objective during a period called *training*.
-In training, the agent *explores* the available states and actions. 
-Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
+In training, the agent *explores* the available actions. 
+Once the agent comes across a highly rewarding sequence of observations and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
+
+## Deep RL Algorithms
+
+To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
+
+Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
+Whenever there is a need for an agent to approximate some function, say a policy, transition or value function^[Value functions estimate the expected return from a state or state-action pair.
+If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
+We talk more about value-based learning methods in Appendix A.], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+While there are other function approximators that can be used in RL, e.g.
+Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
+
+In this paper, we will focus on model-free RL approaches that employ neural networks.
+We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
 
@@
 
-Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands to millions of time steps.
+Training a deep RL agent on an environment involves allowing the agent to interact with the environment for typically thousands to millions of time steps.
 The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
@@
 We further discuss and show the benefits of hyperparameter tuning in Appendix B.
-
-# Reinforcement Learning
-<!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
-
-<!-- Mc: I agree-- could you intregrate this into the other parts? Like when you introduce states and actions rather than having a separate section? -->
-
-The reinforcement learning environment is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
-A POMDP is a 7-tuple that consists of the following:
-
-- $\mathcal{S}$: a set of states called the state space
-
-- $\mathcal{A}$: a set of actions called the action space
-
-- $\Omega$ : a set of observations called the observation space
-
-- $E(o_{t}| s_{t})$: an emission distribution, which accounts for an agent's observation being different from environment's state
-
-- $T(s_{t+1}|s_t, a_t)$: a state transition operator which describes the dynamics of the system
-
-- $r(s_t, a_t)$: a reward function
-
-- $d_0(s_0)$: an initial state distribution
-
-- $\gamma \in (0,1]$: a discount factor
-
-The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
-
-<!-- yeah this is pretty repetative... I dont think you necessarily need to avoid some of the formal notation in the first subsection-->
-
-At each time step, the agent makes an observation of the environment at time t, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
-The action will cause the environment to transition to a new state according to the state transition operator. 
-The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
-This process will repeat until a terminal state is reached, after which the environment can reset to an initial state.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the length of the state-action sequence.
-From these definitions, we can provide an agent's trajectory distribution for a given policy as,
-
-$$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t)
-$$
-
-The goal of reinforcement learning is for the agent to find an optimal policy distribution, $\pi^*(a_t|o_t)$, that maximizes the expected return, $J(\pi)$:
-
-$$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
-$$
-
-To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Whenever there is a need for an agent to approximate some function, say a policy, transition or value function^[Value functions estimate the expected return from a state or state-action pair.
-If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
-We talk more about value-based learning methods in Appendix A.], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-
-In this paper, we will focus on model-free RL approaches that employ neural networks.
-We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
-
-<!-- I really like the last few paragraphs here!
-I do feel like what is above model-free vs model based paragrph is repetative and could just be introduced with the concepts in the previous section.
-Then you could make this whole subsection about the distrinction between model free and model based?
--->
 

### 2021-06-10 manuscript/manuscript.Rmd
Keeping same structure but cutting a RL loop disc

--- 2021-06-10 manuscript/manuscript.Rmd
@@
 The agent learns how to achieve this objective during a period called *training*.
-In training, the agent *explores* the available states and actions. 
-Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
+In training, the agent *explores* the available actions. 
+Once the agent comes across a highly rewarding sequence of observations and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
 Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
@@
 The reinforcement learning environment is typically formalized as a discrete-time partially observable Markov decision process (POMDP).
-A POMDP is a 7-tuple that consists of the following:
+A POMDP is a tuple that consists of the following:
 
@@
 The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
-
-<!-- yeah this is pretty repetative... I dont think you necessarily need to avoid some of the formal notation in the first subsection-->
-
-At each time step, the agent makes an observation of the environment at time t, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
-The action will cause the environment to transition to a new state according to the state transition operator. 
-The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
-This process will repeat until a terminal state is reached, after which the environment can reset to an initial state.
-As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the length of the state-action sequence.
+As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$, the agent will create a trajectory given as $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$^[While there are RL methods for infinite horizon problems, i.e.
+when $H=\infty$, we will only present methods for episodic or finite horizon POMDPs.].
 From these definitions, we can provide an agent's trajectory distribution for a given policy as,
@@
 $$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t)
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^{H-1} \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t)
 $$

### 2021-06-10 manuscript/manuscript.Rmd
Keeping up to date
Moving algorithm section above
Small edit

@@
 
+## RL Algorithm Overview
+
+To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
+
+Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
+Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+While there are other function approximators that can be used in RL, e.g.
+Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
+
+In this paper, we will focus on model-free RL approaches that employ neural networks.
+We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
+
 ## Designing and building an environment
@@
 
-# Reinforcement Learning
+## RL Formalism 
 <!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
@@
 $$
-  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^{H-1} \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t)
+  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^{H-1} \pi(a_t | o_t) \, E(o_{t}| s_{t}) \, T(s_{t+1} | s_t, a_t).
 $$
@@
 $$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi).
 $$
-
-To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Whenever there is a need for an agent to approximate some function, say a policy, transition or value function^[Value functions estimate the expected return from a state or state-action pair.
-If you know the value function under certain conditions, you can use the value function to retrieve the optimal policy by selecting the action that maximizes the value function.
-We talk more about value-based learning methods in Appendix A.], neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-
-In this paper, we will focus on model-free RL approaches that employ neural networks.
-We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
 

### 2021-06-10 manuscript/manuscript.Rmd
Putting a footnote in the body

@@
 The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
-As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$, the agent will create a trajectory given as $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$^[While there are RL methods for infinite horizon problems, i.e.
-when $H=\infty$, we will only present methods for episodic or finite horizon POMDPs.].
+As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$, the agent will create a trajectory given as $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$.
 From these definitions, we can provide an agent's trajectory distribution for a given policy as,
@@
 $$
+
+While there are RL methods for infinite horizon problems, i.e. when $H=\infty$, we will only handle episodic or finite horizon POMDPs.
 

### 2021-06-10 manuscript/manuscript.Rmd
Put in some edits following Kari's suggs

@@
 
-## RL Algorithm Overview
-
-To optimize the RL objective, algorithms either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
-Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
-
-Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
-Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
-While there are other function approximators that can be used in RL, e.g.
-Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-
-In this paper, we will focus on model-free RL approaches that employ neural networks.
-We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
-
-## Designing and building an environment
+## RL Environments
 
@@
 The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
-The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
+The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the currentstate and action.
 This ability allows RL environments to model a broad range of decision making problems.
@@
 
-## Training and tuning an agent
+## Deep RL Agents
 
-Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands to millions of time steps.
-The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
+To optimize the RL objective, agents either take a *model-free* or *model-based* approach. 
+The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
+
+Neural networks become useful in RL when the environment has a large observation-action space^[Conventionally, an observation-action space is considered to be large when it is non-tabular, i.e.
+cannot be represented in a computationally tractable table.], which happens frequently with realistic decision-making problems.
+Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+Although there being other function approximators that can be used in RL, e.g.
+Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
+In this paper, we will focus on model-free RL approaches that employ neural networks.
+We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
+
+Training a deep RL agent involves allowing the agent to interact with the environment for potentially thousands to millions of time steps.
+The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the complexity of the environment, complexity of the agent and more.
 Yet, overall, it has been well established that deep RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
 
-Hyperparameters refer to the parameters in a deep RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
-Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning, which can be automated through standard multi-dimensional optimization approaches.
+*Hyperparameters* refer to the parameters in a deep RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
+Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning which uses standard multi-dimensional optimization methods.
 We further discuss and show the benefits of hyperparameter tuning in Appendix B.
@@
 ## RL Formalism
-<!-- ML: As I read this section, I have some concerns on this coming off as overly repetitive with the parts above. -->
-
-<!-- Mc: I agree-- could you intregrate this into the other parts? Like when you introduce states and actions rather than having a separate section? -->
 
@@
 
-While there are RL methods for infinite horizon problems, i.e. when $H=\infty$, we will only handle episodic or finite horizon POMDPs.
-
-<!-- I really like the last few paragraphs here!
-I do feel like what is above model-free vs model based paragrph is repetative and could just be introduced with the concepts in the previous section.
-Then you could make this whole subsection about the distrinction between model free and model based?
--->
+Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
 
@@
 
-Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straight forward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
+Given a population growth model that predicts the total biomass of a fish stock in the following year as a function of the current biomass, it is straightforward to determine what biomass corresponds to the maximum growth rate of the stock, or $B_{\textrm{MSY}}$, the biomass at Maximum Sustainable Yield (MSY) [@Schaefer1954].
 When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].
@@
 How would an RL agent be applied to empirical data?
-In principle, this is straight forward.
+In principle, this is straightforward.
 After we train an agent on a simulation environment that approximates the fishery of interest, we can query the policy of the agent to find a quota for the observed stock.    
@@
 Ecology holds many open problems for deep RL.  
-It is straight-forward to extend the simple environments presented here to reflect greater biological complexity, or to consider a host of more realistic decision scenarios.
+To extend the examples presented here to reflect greater biological complexity or more realistic decision scenarios, the transition, emission and/or reward functions of the environment can be modified.
 We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
@@
 Complexity comes not only from the ecological processes but also the available actions.
-While such complex dynamics and large state and action spaces are intractable to classical methods, they are straight forward to implement in simulation environments such as those used to train an RL algorithm.
-Neural networks used in deep RL have proven remarkably effective in handling such complexity, particularly when leveraging immense computing resources increasingly available through advances in hardware and software [@gpu_computing].
+Deep RL agents have proven remarkably effective in handling such complexity, particularly when leveraging immense computing resources increasingly available through advances in hardware and software [@gpu_computing].
 

### 2021-06-10 manuscript/manuscript.Rmd
Keeping up to date

@@
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-In this paper, we will focus on model-free RL approaches that employ neural networks.
+In this paper, we will focus on model-free RL approaches that employ neural networks. <!-- ML: put in sentence to introduce table here-->
 We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
@@
 
-## RL Formalism
+## RL Objective
 
@@
 Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
+<!-- Put in a sentence to point to the appendix -->
 

### 2021-06-10 manuscript/manuscript.Rmd
Fishing para edits

@@
 The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
-The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the currentstate and action.
+The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
 This ability allows RL environments to model a broad range of decision making problems.
@@
 The first example focuses on the important but well-studied problem of setting harvest quotas in fisheries management.
+This provides a natural benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
 Determining fishing quotas is both a critical ecological issue [@Worm2006; @Worm2009; @Costello2016], and a textbook example that has long informed the management of renewable resources within fisheries and beyond [@Clark1990].
@@
 When the population growth rate is stochastic, the problem is slightly harder to solve, as the harvest quota must constantly adjust to the ups and downs of stochastic growth, but it is still possible to show the optimal strategy merely seeks to maintain the stock at $B_{\textrm{MSY}}$, adjusted for any discounting of future yields [@Reed1979].
-
-This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
-The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
-Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2016] as well as a common test case to understand ecological dynamics and conservation management more broadly [@something].
-<!-- ML: This seems repetitive with second sentence in first paragraph above so going to leave this citation unfulfilled-->
-<!-- ^MC:+1 to this sentence seeming repetative -->
 

### 2021-06-11 manuscript/manuscript.Rmd

@@
 
-@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both problems.
-<!-- this reads a little abrupt - might be worth transitioning a bit to these challenges -->
-
-We use the `gym` framework to turn existing ecological models into valid environmental simulators that can similarly be used in any RL framework.
+The OpenAI `gym` software framework was created to address the lack of standardization of RL environments and the need for better benchmark environments to advance RL research [@brockman2016].
+The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
+This allows the design and application of software agents which can interact and learn in that environment without knowing anything about the environment's internal details.
+Using the `gym` framework, we turn existing ecological models into valid environmental simulators that can be used in any RL framework.
 In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
@@
 Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
-Although there being other function approximators that can be used in RL, e.g.
+Although there are other function approximators that can be used in RL, e.g.
 Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
-In this paper, we will focus on model-free RL approaches that employ neural networks. <!-- ML: put in sentence to introduce table here-->
-We provide more conceptual and practical background on model-free deep RL in Appendix A and B respectively.
+There is a multitude of deep RL algorithms since there are many design choices that can be made in constructing a deep RL agent -- see Appendix A for more detail on these engineering decisions. 
+In Table 1, we present some of the more common deep RL algorithms which serve as good reference points for the current state of deep RL.
+
+\begin{table}
+\begin{tabular}{l l l}
+\hline
+Abbreviation & Algorithm Name                                                    & Model \\
+\hline
+MCTS         & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based \\
+MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based \\
+DQN          & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  \\
+A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  \\
+A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  \\
+TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  \\
+PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  \\
+DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  \\
+TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  \\
+SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  \\
+IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  \\
+\hline
+\end{tabular}
+\caption{Survey of common deep RL algorithms.}
+\end{table}
 
@@
 Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
-<!-- Put in a sentence to point to the appendix -->
+In appendix A, we will further discuss how deep RL algorithms attempt to optimize the RL objective.
 
@@
 We illustrate the deep RL approach using the model-free algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
-TD3 is an extension of Deep Deterministic Policy Gradient [DDPG, @ddpg], which itself extended the original Deep Q-Network [DQN, @DQN] algorithm to continuous action space.
 A step-by-step walk-through for training agents on this environment is provided in the Appendix.
@@
 The rapidly expanding class of Model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
-Rarely do we know the underlying functional forms for ecological processes -- is population recruitment more Beverton-Holt or Ricker?
+Rarely do we know the underlying functional forms for ecological processes.
 Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.

### 2021-06-11 manuscript/manuscript.Rmd
finally, an abstract

@@
 abstract: |
-  Enter the text of your abstract here.
+  Can machine learning really help us make better decisions about a changing planet?
+  In this paper, we illustrate and discuss the potential of a promising 
+  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
+  most challenging conservation decision problems. RL is uniquely
+  well suited to conservation and global change challenges for three reasons: (1)
+  rather than passively modeling observed data, RL is explicitly designed around 
+  the actions of an agent who _interacts_ with an environment which is dynamic and uncertain,
+  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
+  utilize rather than replace existing models, simulations, and the knowledge they contain.
+  We provide a conceptual and technical introduction to RL and its relevance to ecological
+  and conservation challenges, including examples of both a previously solved problem in setting
+  fisheries quotas and a previously unsolved problem in managing ecological tipping points.
+  Four appendices with annotated code provide a tangible introduction to researchers 
+  looking to adopt, evaluate, or extend these approaches.
+
 keywords:
-  - blah
-  - blee
-  - bloo
-  - these are optional and can be removed
+  - Reinforcement Learning
+  - Conservation
+  - Machine Learning
+  - Artificial Intelligence
+  - Tipping points
 bibliography: references.bib

### 2021-06-11 manuscript/manuscript.Rmd
now with figure caption

@@
 
-```{r message=FALSE}
+```{r message=FALSE, fig.cap="Setting fisheries harvest quotas using Deep RL.
+Argentine Hake fish stocks show a marked decline between 1986 and 2014 (upper panel).
+Historical harvests (lower panel) declined only slowly in response to consistently falling stocks, suggesting overfishing.
+In contrast, RL-based quotas would have been set considerably lower than observed harvests in each year of the data.
+As decline persists, the RL-based management would have closed the fishery to future harvest until the stock recovered."}
 
@@
 Our second example focuses on a case for which we do not already have an existing, provably optimal policy to compare against.
-We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
+We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 4].
 This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012] lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
@@
 
-Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
+Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
 This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady state strategy.
@@
 
-A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see appendix B] further highlights the difference between these two approaches.
+A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see Appendix B] further highlights the difference between these two approaches.
 Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
@@
 
-<!-- transition from big promise to perils -- Millie's section on risks here -->
-
-<!-- from end of Carl's discussion edits: Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet.
--->
-<!-- If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.
--->
-
-<!-- The increasingly influential role of AI in sectors as diverse as policing, finance, and employment has raised concern about the ethics, biases, and power entrenched in both decision-making algorithms and their outputs [@kalluri2020].
--->
-<!-- As algorithms play a greater role in environmental and conservation policy, will similar issues arise? -->
-<!-- While this question has garnered  recent interest in the academic literature [@wearn2019; @adams2019; @scoville2021], the use of decision making algorithms in environmental contexts is still relatively isolated and in it's early stages.
--->
-<!-- Examples of AI applications, and corresponding concerns, have largely been confined to the context of enforcing environmental laws (i.e.
-algorithms which predict wildlife poaching events [@adams2019] or identify illegal fishing behavior [@mcdonald2021; @swartz2021]).
--->
-<!-- However, more diverse applications of AI algorithms - such as applications which contribute to meeting sustainable development goals [@vinuesa2020] or designing  protected area networks - raise additional concerns of entrenching power asymmetries between those reliant on natural resources and those designing algorithms and their objectives.
--->
-<!-- These issues of power become particularly acute when private industry holds the keys to propriety algorithms or environmental data, and generally operate without the same community engagement obligations of public agencies.
--->
-<!-- In these cases, good intentions might not be sufficient to ensure equitable outcomes. -->
-<!-- Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges promotes equity in both the procedural process of developing objective functions and the distribution of benefits that result from decisions.
--->
-
 \pagebreak

### 2021-06-11 manuscript/manuscript.Rmd
Looking good to me; changed a few paragraphs to help with flow

--- 2021-06-11 manuscript/manuscript.Rmd
@@
 bibliography: references.bib
+nocite: |
+    @A2C, @trpo, @ddpg, @TD3, @impala
 header-includes:
@@
 Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in these fields.
-To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar]
+To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
 Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
@@
 As a result of this action, the environment transitions to a new state and transmits a numerical *reward* signal to the agent.
-The goal of the agent is to learn how to maximize its expected cumulative reward in the environment.
+The goal of the agent is to learn how to maximize its expected cumulative reward.
 The agent learns how to achieve this objective during a period called *training*.
@@
 Once the agent comes across a highly rewarding sequence of observations and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
-Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.
+Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given observation.
 
@@
 During \emph{training}, the agent tries to maximize its cumulative reward by interacting with the environment and learning from experience.
-In the RL loop, the agent performs an action, and the environment returns a reward as well as an observation of the environment's state.
+In the RL loop, the agent performs an action, then the environment returns a reward and an observation of the environment's state.
 The agent-environment loop continues until the environment reaches a terminal state, after which the environment will reset, causing a new \emph{episode} to begin.
@@
 An agent is \emph{evaluated} by recording the cumulative reward over one episode, or the mean reward over multiple episodes.
-Within evaluation, the agent does not update its neural network; instead, the agent only uses the neural network to select actions.}
+Within evaluation, the agent does not update its neural network; instead, the agent only uses a trained neural network to select actions.}
 \end{figure}
@@
 The `gym` framework defines a standard interface and methods by which a developer can describe an arbitrary environment in a computer program.
-This allows the design and application of software agents which can interact and learn in that environment without knowing anything about the environment's internal details.
-Using the `gym` framework, we turn existing ecological models into valid environmental simulators that can be used in any RL framework.
+This interface allows for the application of software agents that can interact and learn in that environment without knowing anything about the environment's internal details.
+Using the `gym` framework, we turn existing ecological models into valid environmental simulators that can be used with any RL agents.
 In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
@@
 The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
-A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
+A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo; @suttonbarto].
 Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
@@
 MCTS         & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based \\
+DYNA         & Dyna-Q (Sutton and Barto 2018)                                    & Model-based \\
 MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based \\
@@
 Training a deep RL agent involves allowing the agent to interact with the environment for potentially thousands to millions of time steps.
-The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the complexity of the environment, complexity of the agent and more.
+During training, the deep RL agent continually updates its neural network parameters so that it will converge to an optimal policy. 
+The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the complexity of the environment, the complexity of the agent, and more.
 Yet, overall, it has been well established that deep RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
 
-*Hyperparameters* refer to the parameters in a deep RL algorithm that control the learning process, e.g. the step size to use for gradient descent steps.
+The deep RL agent controls the learning process with parameters called *hyperparameters*.
+Examples of hyperparameters include the step size used for gradient ascent and the interval to interact with the environment before updating the policy.
+In contrast, a weight or bias in an agent's neural network is simply called a *parameter*.
+Parameters are learned by the agent, but the hyperparameters must be specified by the RL practitioner.
 Since the optimal hyperparameters vary across environments and can not be predetermined [@drlthatmatters], it is necessary to find a good-performing set of hyperparameters in a process called hyperparameter tuning which uses standard multi-dimensional optimization methods.
@@
 The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
-As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$, the agent will create a trajectory given as $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$.
+As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$, the agent will create a trajectory, $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$.
 From these definitions, we can provide an agent's trajectory distribution for a given policy as,
@@
 Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
-In appendix A, we will further discuss how deep RL algorithms attempt to optimize the RL objective.
+In appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.
 
@@
 A step-by-step walk-through for training agents on this environment is provided in the Appendix.
-We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 2].
+We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig 2].
 Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
@@
 Lower left: The optimal solution is policy of constant escapement.
-Below the target escapement of 0.5, no harvest occurs, while any stock above that level is immediately harvested back down (red line).
-RL agents both adopt a policies that cease any harvest below this level, while allowing a somewhat higher escapement than optimal just above the optimal escapement level.
-TD3 achieves nearly-optimal mean utility."}
+Below the target escapement of 0.5, no harvest occurs, while any stock above that level is immediately harvested back down.
+The TD3 agent adopts a policy that ceases any harvest below this level, while allowing a somewhat higher escapement than optimal.
+TD3 achieves a nearly-optimal mean utility."}
 
@@
 
-Our second example focuses on a case for which we do not already have an existing, provably optimal policy to compare against.
+Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
 We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
-This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012] lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
+This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
 On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
@@
 Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
+<!-- ML: Nit picky; it looks like it almost goes negative but not quite on average in the plot -->
 This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady state strategy.
@@
 <!-- Model free is great bc we don't know the true models -->
-The rapidly expanding class of Model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
+The rapidly expanding class of model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
 Rarely do we know the underlying functional forms for ecological processes.
@@
 In contrast, the process of training an RL algorithm on a more complex problem is no different than training on a simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
-This flexibility of RL, for instance, could allow us to use RL agents on our most realistic ecological models such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
+This flexibility of RL could allow us to attain better decision-making insight for our most realistic ecological models such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
 
@@
 Successfully applying RL to complex ecological problems is no easy task.
-Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters].
+Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters; @berger-tal_exploration-exploitation_2014].
 As Section 5.1 and 5.2 illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.

### 2021-06-11 manuscript/manuscript.Rmd
edits suggested by millie

--- 2021-06-11 manuscript/manuscript.Rmd
@@
     location: Berkeley, California
-  - name: Melissa Chapman
+  - name: Melissa S. Chapman
     department: Department of Environmental Science, Policy, and Management
@@
     location: Berkeley, California
-  - name: Kari Norman
+  - name: Kari E. A. Norman
     department: Department of Environmental Science, Policy, and Management
@@
 abstract: |
-  Can machine learning really help us make better decisions about a changing planet?
+  Can machine learning help us make better decisions about a changing planet?
   In this paper, we illustrate and discuss the potential of a promising 
@@
   rather than passively modeling observed data, RL is explicitly designed around 
-  the actions of an agent who _interacts_ with an environment which is dynamic and uncertain,
+  the actions of an agent who _interacts_ with an environment that is dynamic and uncertain,
   (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
@@
   We provide a conceptual and technical introduction to RL and its relevance to ecological
-  and conservation challenges, including examples of both a previously solved problem in setting
-  fisheries quotas and a previously unsolved problem in managing ecological tipping points.
+  and conservation challenges, including examples of a problem in setting
+  fisheries quotas and in managing ecological tipping points.
   Four appendices with annotated code provide a tangible introduction to researchers 

### 2021-06-11 appendices/appendix-A-definitions.Rmd
Few transitions, putting in algorithm table
Various edits mainly for consistency

@@
 In this appendix, we will go into further detail on how deep RL algorithms work.
-The aim here is not to give full treatment of all deep RL methods; instead, this section is meant to serve as an abbreviated background view on model-free deep RL.
+The aim here is not to give full treatment of all deep RL methods; instead, this section is meant to serve as an abbreviated background view on model-free deep RL^[We focus on model-free methods because the agents we use in the manuscript and Appendix B are all model-free.
+Further, model-based algorithms generalize from the methods we discuss here except model-based agents attempt to learn or use the environment's state-transition operator, which is typically used for planning or supplementary sample generation [@suttonbarto].].
 We will introduce some of the important design choices that are made for model-free deep RL algorithms and will point to specific resources that give a more complete presentation.
 These sections presume some familiarity with gradient descent optimization -- see @ruder_overview_2017 for a thorough treatment.
-Throughout these sections, we'll consider the environment to be an MDP which makes for a cleaner presentation and is also consistent with the referenced sources.
+Throughout these sections, we'll consider the environment to be an MDP which makes for a cleaner presentation and is also consistent with the referenced literature.
+This treatment can be extended to the POMDP case by including observations and the emission function.
 
@@
 Among model-free methods, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
-Generally, model-free agents are either learning a policy function, a value function or both.
+Generally, model-free agents either learn a policy function, a value function or both.
 For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
 If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
-The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
+The class of algorithms that exclusively tries to learn a value function are called *value-based* methods.
 In contrast, *policy gradient* algorithms exclusively learn the policy.
@@
 In policy gradient methods, the RL agent performs gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters.
-At each gradient step, the following update is performed,
+At each gradient step, which we'll index using $k$, the following update is performed,
 
 $$
-  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta) .
+  \theta_{k+1} = \theta_k + \alpha \nabla_{\theta_k} J(\pi_{\theta_k}) .
 $$
@@
 A sketch of a simple policy gradient algorithm is shown in Algorithm 1^[The algorithm sketches showed throughout this study are not practical for a variety of reasons including high bias estimators, high variance estimators and correlated samples.
-The algorithm examples are instead intended to convey how the general steps of the different algorithm classes come together.].
+The algorithm examples are instead intended to convey how the general steps of the different algorithm classes come together.] ^[In this algorithm, we mention exploration algorithms which we discuss later in Section 2.].
 
@@
     \For{$k = 0, 1, 2, \dots$}
-        \State Generate a trajectory $\{s_i, a_i, r_i\}$ by following an exploration algorithm with the policy $\pi_k$ (e.g. $\epsilon$-greedy)
+        \State Generate a trajectory $\{s_i, a_i, r_i\}$ by following an exploration algorithm with the policy $\pi_{\theta_k}$ (e.g. $\epsilon$-greedy)
         \For{$t = 0, 1, \dots, T-1$}
             \State $G \gets \sum_{j=t}^{T} \gamma^{j-t} r_j$
-            \State $\theta_{k+1} \gets \theta_k + \alpha \gamma^t G \, \nabla \log \pi_{\theta_k}(a_t | s_t)$
+            \State $\theta_{k+1} \gets \theta_k + \alpha \gamma^t G \, \nabla_{\theta_k} \log \pi_{\theta_k}(a_t | s_t)$
         \EndFor
@@
 The motivation behind Q-learning is that if we know the Q function under certain conditions -- e.g.
-that the state-action space is discrete and that all state-action pairs are visited ad infinitum [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q (s_t, a_t)$.
+that the state-action space is discrete and that all state-action pairs are visited ad infinitum^[These conditions are of course rarely true, but approximate Q-learning has still been able to achieve notable successes [@DQN].] [@suttonbarto 131; @Qproof] --, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q (s_t, a_t)$.
 Using this optimal policy, we can write an equation for the optimal Q-function from the recursive definition,
@@
 The objective for deep Q-learning is to find an approximate Q-function that satisfies the above equation, where we represent our approximate Q-function, $Q_\phi$, with a neural network that has the parameters of $\phi$.
-Subtracting the left hand and right hand side of the recursive Q-function definition, we can define the temporal difference error, $\mathcal{E}$, for a state-action pair as
+Subtracting the left hand and right hand side of the recursive Q-function definition, we can define the temporal difference (TD) error, $\mathcal{E}$, for a state-action pair as
 
 $$
-  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_\phi(s_{t+1}, a) - Q_\phi(s_t, a_t).
+  \mathcal{E} = r(s_t, a_t) + \gamma \, \underset{a}{\text{max}} \, Q_{\phi}(s_{t+1}, a) - Q_{\phi}(s_t, a_t).
 $$
@@
 $$
-  \phi_{t+1} = \phi_t - \alpha \nabla_\phi \, \sum_i \mathcal{E}_i^2 .
+  \phi_{k+1} = \phi_k - \alpha \nabla_{\phi_k} \, \sum_i \mathcal{E}_i^2 .
 $$
@@
 A notable concern with these methods is that they tend to suffer from bias which is introduced by fitting a value network to a target that incorporates an estimated value function [@van_hasselt_deep_2015].
-In review, a sketch of simple deep Q learning algorithm is shown in Algorithm 2.
+In review, a sketch of a simple deep Q learning algorithm is shown in Algorithm 2.
 
@@
         \For{$t = 0, 1, \dots, T-1$}
-            \State Choose the action $a_t$ using $Q_{\phi_k}$ according to an exploration algorithm (e.g. $\epsilon$-greedy)
+            \State Select an action $a_t$ using $Q_{\phi_k}$ according to an exploration algorithm (e.g. $\epsilon$-greedy)
             \State Take action $a_t$ and observe $r_{t}, s_{t+1}$
@@
 While the general algorithm for the critic is exactly the same as mentioned in the value-learning section, the actor differs from policy gradient algorithms by incorporating the critic in its gradient ascent step.
-Instead of estimating the gradient of the return as in eq ..., the gradient of the return is estimated as
+The gradient of the return is estimated as
 
@@
 Consequently, actor-critic algorithms have achieved many of the state of the art results in model-free deep RL [@sac; @TD3].  
-A simple sketch of an actor critic algorithm is shown in Algorithm 3.
+A simple sketch of an actor-critic algorithm is shown in Algorithm 3.
 
@@
             \State $\phi_{k+1} \gets \phi_{k} - \alpha \nabla_{\phi_k} \, \mathcal{E}^2$
-            \State $\theta_{k+1} \gets \theta_k + \alpha \, \nabla \log \pi_{\theta_k}(a_t | s_t) Q_{\phi_k}(s_t, a_t)$
+            \State $\theta_{k+1} \gets \theta_k + \alpha \, \nabla_{\theta_k} \log \pi_{\theta_k}(a_t | s_t) Q_{\phi_k}(s_t, a_t)$
         \EndFor
@@
 Yet, how the agent selects these actions is very important because if an agent explores the state-action space poorly, the agent will learn a sub-optimal policy.
-Occasionally, the agent will need to propose exploratory actions that may result in higher rewards than previously seen.
-Yet, the agent may also need to focus on past sequences of actions which may require marginal improvement to reach optimality.
 The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior.
@@
 For RL problems with small state-action spaces, e.g.
-space-action spaces that can be represented in a table, simple exploration algorithms will converge to the optimal policy, but with large or infinite state-action spaces, there are no convergence guarantees.
+space-action spaces that can be represented in a table, simple exploration algorithms will converge to the optimal policy, but with large or infinite state-action spaces, there are no convergence guarantees [@berger-tal_exploration-exploitation_2014].
 
@@
 The Epsilon-Greedy algorithm takes a very simple approach towards balancing exploration and exploitation. 
-For some $\epsilon \in (0, 1)$ and typically close to 0, an Epsilon-Greedy agent exploits the best available action with probability $1-\epsilon$ and explores a random action with probability $\epsilon$.
+For some $\epsilon \in (0, 1)$, an Epsilon-Greedy agent exploits the best available action with probability $1-\epsilon$ and explores a random action with probability $\epsilon$.
 While simplistic, this algorithm has been effective on a range of problems notably @DQN.
@@
 For algorithms that learn a value function, agents can select actions by sampling from a Boltzmann or Softmax distribution that is constructed over the value function.
-So if the agent is doing Q-learning, the probability for the agent to select the action, $a \in \mathcal A$, would be:
+So if the agent is doing Q-learning, the probability for the agent to select the action, $a$, would be:
 
@@
 The advantage of using Boltzmann exploration is that since the probability of selecting an action is proportional to how good the agent thinks that action is, the agent can explore the space more efficiently than with Epsilon-Greedy exploration.
-An Epsilon-Greedy agent will explore all actions equally including low reward yielding actions; in contrast, with Boltzmann exploration, an agent will spend less time on these unpromising actions.
 
@@
 With noise addition, there are a number of design choices that can be made.
-For instance, noise can be sampled from different distributions [@plappert_parameter_2018], scaled adaptively based on the policy[@plappert_parameter_2018], or the noise can be sampled according to the state that the agent observes [@gsde].
+For instance, noise can be sampled from different distributions [@plappert_parameter_2018], scaled adaptively based on the policy[@plappert_parameter_2018], or sampled according to the state that the agent observes [@gsde].
 There are a variety of reasons why one would want to use each of these variations; these reasons are well discussed in @plappert_parameter_2018 and @gsde.
@@
 By adding this term, the agent has an incentive to learn a policy that balances reward-seeking and explorative behavior. 
-While this introduces some bias as the agent no longer exclusively maximizes cumulative rewards, agents that use entropy-based exploration have achieved state of the art results [@sac].
+While this introduces some bias as the agent no longer exclusively maximizes cumulative rewards, entropy-based exploration has achieved state of the art results [@sac].
 For more, @ahmed_understanding_2019 gives a thorough discussion on the benefits of using entropy.
@@
 It is important to note that the algorithms mentioned above are generally suited for dense reward or "easy-exploration" problems.
-For "hard-exploration" problems -- for instance, environments that have very few state-action pairs with significant rewards, like a maze, are seen as hard-exploration problems --, more intricate and bespoke exploration algorithms tend to outperform.
-The issue for "hard-exploration" problems is that the agent needs to explore the state-action space efficiently, and simple exploration algorithms will waste compute by revisiting bad actions.
+"Hard-exploration" problems, like a maze environment which has very few state-action pairs with significant rewards, often require more intricate and bespoke exploration algorithms.
+The issue for "hard-exploration" problems is that the agent needs to explore the state-action space efficiently, and simple exploration algorithms will waste computational time by revisiting bad actions.
 Two common approaches to hard exploration problems are memory-based and curiosity-driven techniques.
@@
 RL agents can either select actions from the same policy that they intend to use during evaluation or they can sample actions from a policy that is different.
-The former is called on-policy learning; the agent learns to select actions with the same policy it will use in testing.
-The latter is called off-policy learning; the agent uses a policy during training that is potentially different than the policy that the agent will evaluate.
+The former is called *on-policy* learning: the on-policy agent learns to select actions with the same policy it will use in testing.
+The latter is called *off-policy* learning: the off-policy agent uses a policy during training that is potentially different than the policy that the agent will evaluate.
 
-A significant issue that arises when sampling actions and then using this experience to maximize the RL objective is correlation [@DQN, @schaul_prioritized_2016].
-In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated[@DQN].
+A significant issue that arises when agents learn from their experience is correlation of samples [@DQN, @schaul_prioritized_2016].
+In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated [@DQN].
 Learning from correlated samples is undesirable since this can result in agents getting stuck in local optima, which will impede learning, but there are other reasons why learning from a single trajectory causes problems -- see @schaul_prioritized_2016 and @DQN for more insight here.
@@
 Since the replay buffer can be arbitrarily large and thus store transitions from numerous trajectories, sampling from the buffer results in decorrelated samples.
-On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- note replay buffers are generally large enough to hold transitions from multiple policies.
+On-policy agents, however, cannot use a replay buffer since these agents are not allowed to use samples from a policy different than the policy they are trying to improve -- n.b.
+replay buffers are generally large enough to hold transitions from multiple policies.
 The solution to de-correlate samples for on-policy agents is to use parallel workers, whereby multiple workers explore the environment according to the current policy.
@@
 \hline
-Algorithm & Description                                                       & Model       & Policy     & Method \\
+Abbreviation & Algorithm Name                                                   & Model       & Policy     & Method \\
 \hline

### 2021-06-11 manuscript/manuscript.Rmd

@@
   well suited to conservation and global change challenges for three reasons: (1)
-  rather than passively modeling observed data, RL is explicitly designed around 
-  the actions of an agent who _interacts_ with an environment that is dynamic and uncertain,
+  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
   (2) RL approaches do not require massive amounts of data, (3) RL approaches would 

### 2021-06-11 appendices/appendix-B-examples.Rmd
Going over appendix B

@@
 and should serve as an effective starting point for researchers seeking to apply deep RL to these simulation environments, 
-to other simulation environments we have built and continue to develop, or to their own environments.
+to other simulation environments we continue to develop, or to their own environments.
 (Appendix C details the process of defining one's own `gym` environment from an ecological model). 
@@
 Reproducing the training of RL agents is more computationally intensive, and a more powerful multi-core CPU or CUDA-enabled GPU is recommended.
-Note that hardware differences are also a source of instability in the reproduciblity of RL training (see main text and Appendix A).
+Note that hardware differences are also a source of instability in the reproduciblity of RL training [@instability_2019].
 
@@
 Rather than implement deep RL algorithms from scratch, we illustrate the use of one of the leading software frameworks for research and application of deep RL methods.
-At the time of writing, several major frameworks exist which provide reference implementations of the leading deep RL algorithms (see table 1 and Appendix A).
+At the time of writing, several major frameworks exist which provide reference implementations of the leading deep RL algorithms (see Table 1 and Appendix A).
 While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
@@
 
-Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration. 
+Numerical reproducibility can be challenging in machine learning problems, particularly when using GPU-based acceleration [@instability_2019]. 
 In addition to setting a random seed in our Python environment, we can optionally disable GPU use to improve reproducibility by setting the `CUDA_VISIBLE_DEVICES` to a blank value.
@@
 Unlike the optimal solution, the RL methods never know the underlying growth function used by the simulation (or even if such a function actually exists).
-These methods merely seek to learn a strategy for setting harvest quotas which maximizes the cumulative reward the receive from the environment.
+These methods merely seek to learn a strategy for setting harvest quotas which maximizes the cumulative reward they receive from the environment.
 At the heart of each deep RL algorithm is a neural network or a collection of neural networks. 
-The RL agent uses these networks to approximate a policy function and/or a value function.
-The policy function is the agent's mapping of states to actions -- i.e.
-the agent's strategy.
-Value functions attempt to evaluate the goodness of a state or state-action pair towards achieving the RL objective of maximizing cumulative rewards.
-Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyper-parameters_ while the parameters of a RL algorithm that are learned during training are called parameters.
-For example, the number of layers in a neural network is a hyper-parameter but a weight in the neural net is referred to as a parameter.
-Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
-Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
+Model-free RL agents use these networks to approximate a policy function and/or a value function -- see Appendix A for more detail here. 
+Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyperparameters_ while the parameters of a RL algorithm that are learned during training are called parameters.
+For example, the number of layers in a neural network is a hyperparameter, but a weight in the neural net is referred to as a parameter.
+Before training an agent, we must first specify the hyperparameters; the parameters, meanwhile, are typically randomly initialized.
+Note that each algorithm may have different hyperparameters, so it is helpful to look at the algorithm's documentation to know what these hyperparameters are. 
+`stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms.
 
@@
 This algorithm is a successor of Deep Deterministic Policy Gradient (DDPG) method [@ddpg], itself a policy-gradient-modification that extends the original Deep Q-Network (DQN) algorithm [@DQN] to continuous action spaces.
-We use this algorithm to train an agent that has a multi-layer perceptron (`MlpPolicy`) neural network, with the default architecture of two 64-neuron layers.
+We use this algorithm to train an agent that employs multi-layer perceptron (`MlpPolicy`) neural networks, with the default architecture of two 64-neuron layers.
 
@@
 
-Note that while default hyper-parameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyper-parameter selection.
+Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss below), better performance can almost always be achieved by _tuning_ the hyperparameter selection.
+<!-- ML: eek I think we dropped normalization talk -->
 This is discussed further below.
@@
 
-We can supply an observation to our trained agent and it will return it's proposed action using the `predict` method. 
+We can supply an observation to our trained agent and it will return its proposed action using the `predict` method. 
 This is all we need to evaluate or employ the agent on a decision-making task.
@@
 
-As we can see in the figure above, without hyper-parameter tuning, our agent has found only a sub-optimal policy.
+As we can see in the figure above, without hyperparameter tuning, our agent has found only a sub-optimal policy.
 The RL agent trained under an un-tuned TD3 algorithm here tends to over-harvest the fishery, resulting in an initially higher reward that grows steadily but more slowly than the optimal strategy.
-As discussed in the main text, training of Deep RL algorithms is subject to several sources of instability that can make it difficult or impossible to reproduce the same results from training with the same hyper-parameters on different hardware or hardware drivers, even when controlling for random seed [@instability_2019; @drlthatmatters].
+As discussed in the main text, training of Deep RL algorithms is subject to several sources of instability that can make it difficult or impossible to reproduce the same results from training with the same hyperparameters on different hardware or hardware drivers, even when controlling for random seed [@instability_2019; @drlthatmatters].
 Consequently, users may not recover identical behavior when training under identical parameters.
@@
 
-By tuning the hyper-parameters that control the training regime used by the TD3 algorithm, we are able to further improve upon on the performance of the RL agent.
+By tuning the hyperparameters that control the training regime used by the TD3 algorithm, we are able to further improve upon on the performance of the RL agent.
 Tuning may be performed manually by adjusting hyperparameters based on mechanistic reasoning (e.g.
@@
 
-We train this new RL agent using it's tuned hyperparameters, and saving the resulting trained agent as before:
+We train this new RL agent using its tuned hyperparameters, and saving the resulting trained agent as before:
 
@@
 Note that the largest deviations from the optimal policy occur at large stock sizes, at or above carrying capacity of the model.
-This pattern is not surprising, since these large stock values are least likely to observed over the simulated episodes.
-For the same reason, that deviation contributes very little to reward. 
+This pattern is not surprising, since these large stock values are least likely to be observed over the simulated episodes.
+For the same reason, that deviation contributes very little to the observed reward. 
 Only by comparing average stock sizes directly can we see that the RL agent slightly under-harvests on average relative to the optimal policy.
 As noted in appendix C, careful design of the environment, such as re-scaling of state and action space, can also improve the ability of a Deep RL algorithm to find an acceptable policy without altering the substance of the challenge.
-Even so, our agent still had extensive time to train on this specific environment, a total of 300,000 timesteps or 3000 training episodes.
+Even so, our agent still had extensive time to train on this specific environment, a total of 300,000 timesteps.
 
@@
 As discussed in the main text, we have no existing optimal solution to the tipping point problem, and so rely on a common heuristic strategy instead:
-select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in it's current state.
-This is implemented using the `fixed_action` method provided in our `gym_conservation` module, which also implements other heursitic models, including a human agent which requires interactive input to select the action each year.
+select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in its current state.
+This is implemented using the `fixed_action` method provided in our `gym_conservation` module, which also implements other heuristic models, including a human agent which requires interactive input to select the action each year.
 

### 2021-06-11 appendices/appendix-C-environments.Rmd
Small edits on appendix C

@@
 
-a observation function (sometimes called the 'emission' function, $E$) giving the observed state as a function of the current state and possibly also the action (i.e.
+an observation function (sometimes called the 'emission' function, $E$) giving the observed state as a function of the current state and possibly also the action (i.e.
 different actions could potentially influence the accuracy of the observation):
@@
 In practice, an environment may include additional logic for rendering information about it's current states as output.
-Effective environments may often work in a transformed variable space, for example, `gym_fishing` maps the space $[0,2K]$ into the space $[-1, 1]$, which significantly improves the performance. 
-Other features, such as 32 bit precision numeric values for state and action spaces, can improve the performance of the environment on GPU-based computation.
+Effective environments may often work in a transformed variable space, for example, `gym_fishing` maps the space $[0,2K]$ into the space $[-1, 1]$, which significantly improves agent performance. 
+Other features, such as 32 bit precision numeric values for state and action spaces, can also improve agent performance with GPU-based computation.
 

### 2021-06-11 manuscript/manuscript.Rmd

--- 2021-06-11 manuscript/manuscript.Rmd
@@
 abstract: |
-  Enter the text of your abstract here.
+  Can machine learning help us make better decisions about a changing planet?
+  In this paper, we illustrate and discuss the potential of a promising 
+  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
+  most challenging conservation decision problems. RL is uniquely
+  well suited to conservation and global change challenges for three reasons: (1)
+  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
+  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
+  utilize rather than replace existing models, simulations, and the knowledge they contain.
+  We provide a conceptual and technical introduction to RL and its relevance to ecological
+  and conservation challenges, including examples of a problem in setting
+  fisheries quotas and in managing ecological tipping points.
+  Four appendices with annotated code provide a tangible introduction to researchers 
+  looking to adopt, evaluate, or extend these approaches.
+
 keywords:
-  - blah
-  - blee
-  - bloo
-  - these are optional and can be removed
+  - Reinforcement Learning
+  - Conservation
+  - Machine Learning
+  - Artificial Intelligence
+  - Tipping points
 bibliography: references.bib
@@
 Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
-We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
+We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 4].
 This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
@@
 
-Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
+Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
 <!-- ML: Nit picky; it looks like it almost goes negative but not quite on average in the plot -->

### 2021-06-11 appendices/appendix-B-examples.Rmd
Fixing some edits
Fixing 2 misspecified refs

@@
 Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss below), better performance can almost always be achieved by _tuning_ the hyperparameter selection.
-<!-- ML: eek I think we dropped normalization talk -->
 This is discussed further below.
@@
 Slow change over time in the parameter $a_t$ represents a process of environmental degradation, modeled as a constant increment $a_{t+1} = a_t + \alpha$, where we will take $\alpha = 0.001$.
-This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carptenter2011] to the planet biosphere [@Barnosky2014].
-<!-- ML: Not sure what citation to use here for Barnosky or Carptenter in these years-->
+This model supports the dynamics of a fold bifurcation, widely used to model critical transitions in both theory and empirical manipulation in systems from microbes [@Dai2012] to lakes [@Carpenter2011] to the planet biosphere [@Barnosky2012].
 

### 2021-06-11 manuscript/manuscript.Rmd
Fixing the conservation agent desc and usage edits re steady-state

@@
 Lower left: the policy function learned using TD3 relative to the policy function under the steady state.
-Lower right: mean rewards under TD3 management evenutally exceed those expected under the steady state strategy as a large initial investment in conservation eventually pays off."}
+Lower right: mean rewards under TD3 management evenutally exceed those expected under the steady-state strategy as a large initial investment in conservation eventually pays off."}
 
@@
 
-Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
-<!-- ML: Nit picky; it looks like it almost goes negative but not quite on average in the plot -->
-This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady state strategy.
+Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean stead-state reward initially before exceeding it in the long run.
+This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady-state strategy.
 The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.

### 2021-06-11 appendices/appendix-D-data.Rmd

@@
 We use historical data from Argentine Hake to illustrate how an RL agent might be applied in practice.  
-Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy].
+Historical biomass and catch data for Argentine Hake can be found in the R.A. Myers Legacy Stock Assessment Database [@ramlegacy; @ricard_examining_2012].
 Here, we load a local copy of the stock assessment for the Argentine Hake from 1986 to 2000.
@@
 
-For each year in the data, we compare the historical harvest to the quota that would have been recommended by the TD3-based Deep RL agent we have just trained.
+For each year in the data, we compare the historical harvest to the quota that would have been recommended by the TD3-based deep RL agent we trained in Appendix B.
 To utilize our RL agent to set a quota for the stock, we only have to map the observed biomass estimates and historical harvests into the re-scaled state space of the environment. 
@@
 This result should not be over-interpreted. 
-Without being able to replay history we have no way of truly knowing if the RL agent would have managed this stock better.
-While this example shows that it is in principle very straight forward to utilize a trained RL agent to recommend real harvest quotas, we must bear in mind that the agent is only as good as the environments on which it was trained.
+Without being able to replay history, we have no way of truly knowing if the RL agent would have managed this stock better.
+While this example shows that it is in principle very straightforward to utilize a trained RL agent to recommend real harvest quotas, we must bear in mind that the agent is only as good as the environments on which it was trained.
 In practice, the RL agent would have to first be trained and evaluated on much more realistic environments than only the `fishing-v1` gym illustrated here.

### 2021-06-11 manuscript/manuscript.Rmd

@@
 
-Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean stead-state reward initially before exceeding it in the long run.
+Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
 This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady-state strategy.

### 2021-06-13 manuscript/manuscript.Rmd
build all pdfs

@@
     location: Berkeley, California
+    email: mlapeyro@berkeley.edu
   - name: Melissa S. Chapman
@@
     location: Berkeley, California
+    email: mchapman@berkeley.edu
   - name: Kari E. A. Norman
@@
     location: Berkeley, California
+    email: kari.norman@berkeley.edu
   - name: Carl Boettiger
@@
     location: Berkeley, California
-    email: cboettig@berkeley.edu
+    email: cboettig@berkeley.edu (corresponding author)
 abstract: |
@@
 
+a short running title: **Deep Reinforcement Learning**
+
+the type of article: **Method**
+
+the number of words: **4,814**
+
+the number of references: **62**
+
+the number of figures, tables, and text boxes: **6** (five figures, 1 table)
+
+\pagebreak
+
 # Introduction
@@
 Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
-We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+We include an extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
 
@@
 Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
-In appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.
+In Appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.
 
@@
 We provide two examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
+Annotated code for these examples may be found in Appendix B and at <https://github.com/boettiger-lab/rl-intro>.
 

### 2021-06-13 manuscript/manuscript.Rmd
journal submission formatting

@@
   - \usepackage{amssymb}
+  - \usepackage{lineno}
+  - \usepackage{endfloat}
+  - \linenumbers
+
 output: 
@@
 ---
+
+Statement of Authorship:  ML and CB developed the code and wrote the manuscript with support from KEAN and MSC.
+
+Data Accessibility Statement: No new empirical data were used in this analysis.
+Processed forms of public data used in this analysis, simulated data, and code are provided at <https://github.com/boettiger-lab/rl-intro>, archived on Zenodo Data Archive under series-DOI: <https://doi.org/10.5281/zenodo.4940139>.
 

### 2021-06-13 manuscript/manuscript.Rmd
arxiv format

@@
   - \usepackage{amssymb}
-  - \usepackage{lineno}
-  - \usepackage{endfloat}
-  - \linenumbers
+#  - \usepackage{lineno}
+#  - \usepackage{endfloat}
+#  - \linenumbers
 
@@
 ---
+
+<!-- Journal Submission Required Statements
 
@@
 \pagebreak
+
+-->
 

### 2021-06-13 appendices/appendix-A-definitions.Rmd
Most significant edit to the RL tables; diettrich citation added also

@@
 In this appendix, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with state of the art model-free agents.
-In Table 1, we collect some of the more commonly used deep RL algorithms.
+In Table 1, we collect some of the more commonly used model-free deep RL algorithms.
 Readers who are interested in learning more about the current state of deep RL should read the literature referenced in Table 1; @DQN, @sac and @ppo are good places to start. 
 It is important to note that while we have focused on gradient-based, model-free methods, there has also been promising recent development with model-based and gradient-free techniques.
-See @mbpo for a recent example of a model-based deep RL algorithm and @salimans_evolution_2017 as a recent example of a gradient-free approach.
+See @mbpo and @weber_imagination-augmented_2018 for a recent examples of model-based deep RL algorithms, and @salimans_evolution_2017 as a recent example of a gradient-free approach.
 A benefit of model-free deep RL over these other methods is that there has been more of a general focus on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
@@
 \hline
-MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based & On-policy  & Policy Gradient \\
-MCTS         & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based & Either     & Any\\
 A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  & On-policy  & Actor-critic\\
@@
 \end{tabular}}
-\caption{Survey of common deep RL algorithms. }
+\caption{Survey of common model-free deep RL algorithms. }
 \end{table}

### 2021-06-13 manuscript/manuscript.Rmd
Most significant edit to the RL tables; diettrich citation added also

--- 2021-06-13 manuscript/manuscript.Rmd
@@
     location: Berkeley, California
-    email: mlapeyro@berkeley.edu
   - name: Melissa S. Chapman
@@
     location: Berkeley, California
-    email: mchapman@berkeley.edu
   - name: Kari E. A. Norman
@@
     location: Berkeley, California
-    email: kari.norman@berkeley.edu
   - name: Carl Boettiger
@@
     location: Berkeley, California
-    email: cboettig@berkeley.edu (corresponding author)
+    email: cboettig@berkeley.edu
 abstract: |
@@
 nocite: |
-    @A2C, @trpo, @ddpg, @TD3, @impala
+    @A2C, @trpo, @ddpg, @TD3, @impala, @lenz_deepmpc_2015, @weber_imagination-augmented_2018
 header-includes:
@@
 ---
-
-a short running title: **Deep Reinforcement Learning**
-
-the type of article: **Method**
-
-the number of words: **4,814**
-
-the number of references: **62**
-
-the number of figures, tables, and text boxes: **6** (five figures, 1 table)
-
-\pagebreak
 
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013; @ferrer-mestres_k-n-momdps_2021].
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
@@
 Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
-We include an extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
 
@@
 States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
-The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
+The main focus of building an RL environment, however, is on the environment's transition dynamics and reward function.
 The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
@@
 \hline
-MCTS         & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based \\
-DYNA         & Dyna-Q (Sutton and Barto 2018)                                    & Model-based \\
+D-MPC        & Deep Model-Predictive Control (Lenz, Knepper, and Saxena 2015)    & Model-based \\
+I2A          & Imagination-Augmented Agents (Weber et al. 2018)                  & Model-based \\
 MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based \\
@@
 $$
-  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi).
+  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H-1} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi).
 $$
@@
 Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
-In Appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.
+In appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.
 
@@
 We provide two examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
-Annotated code for these examples may be found in Appendix B and at <https://github.com/boettiger-lab/rl-intro>.
 
@@
 Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
+Such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.
 Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
-Nevertheless, such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.
 
@@
 We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
-Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
+Some environments in this library include a wildfire `gym` that poses the problem of wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
 

### 2021-06-14 appendices/appendix-A-definitions.Rmd

@@
         \For{$t = 0, 1, \dots, T-1$}
-            \State $G \gets \sum_{j=t}^{T} \gamma^{j-t} r_j$
+            \State $G \gets \sum_{j=t}^{T-1} \gamma^{j-t} r_j$
             \State $\theta_{k+1} \gets \theta_k + \alpha \gamma^t G \, \nabla_{\theta_k} \log \pi_{\theta_k}(a_t | s_t)$
@@
 The fundamental issue is that it is very difficult for the agent to know when it should engage in explorative or exploitative behavior.
-This *exploration-exploitation dilemma* is an open research problem [@berger-tal_exploration-exploitation_2014].
+This *exploration-exploitation dilemma* is an open research question [@berger-tal_exploration-exploitation_2014].
 For RL problems with small state-action spaces, e.g.
@@
 For some $\epsilon \in (0, 1)$, an Epsilon-Greedy agent exploits the best available action with probability $1-\epsilon$ and explores a random action with probability $\epsilon$.
-While simplistic, this algorithm has been effective on a range of problems notably @DQN.
+While simplistic, this algorithm has been effective on a range of problems [@DQN].
 
@@
 
-A significant issue that arises when agents learn from their experience is correlation of samples [@DQN, @schaul_prioritized_2016].
+A significant issue that arises when agents learn from their experience is correlation of samples [@DQN; @schaul_prioritized_2016].
 In the simple case that we have an RL agent that interacts with the environment for a few time steps and then uses this trajectory for its learning update, these samples, since they come from the same trajectory, will be strongly correlated [@DQN].
@@
 The idea behind using parallel workers is that sampling over multiple trajectories de-correlates samples -- see @a3c on more details here.
-In practice, using a large replay buffer, on the order of millions of transitions, or a modest number of parallel, around 4-16 for simple RL problems, is very important to achieve good agent performance for off-policy and on-policy algorithms respectively.
+In practice, using a large replay buffer, on the order of millions of transitions, or a modest number of parallel, around 4-16 for simple RL problems, is very important to achieve good agent performance.
 
@@
 It is important to note that while we have focused on gradient-based, model-free methods, there has also been promising recent development with model-based and gradient-free techniques.
-See @mbpo and @weber_imagination-augmented_2018 for a recent examples of model-based deep RL algorithms, and @salimans_evolution_2017 as a recent example of a gradient-free approach.
+See @mbpo and @weber_imagination-augmented_2018 for recent examples of model-based deep RL algorithms, and @salimans_evolution_2017 as a recent example of a gradient-free approach.
 A benefit of model-free deep RL over these other methods is that there has been more of a general focus on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
@@
 IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  & Off-policy & Actor-critic\\
+\hline
 \end{tabular}}

### 2021-06-14 appendices/appendix-A-definitions.Rmd

@@
 \hline
-Abbreviation & Algorithm Name                                                   & Model       & Policy     & Method \\
+Abbreviation & Algorithm Name                                                     & Policy     & Method \\
 \hline
-A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  & On-policy  & Actor-critic\\
-A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  & On-policy  & Actor-critic\\
-TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  & On-policy  & Policy Gradient\\
-PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  & On-policy  & Actor-critic\\
-DQN          & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  & Off-policy & Value-Based\\
-DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  & Off-policy & Actor-critic\\
-TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  & Off-policy & Actor-critic\\
-SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  & Off-policy & Actor-critic\\
-IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  & Off-policy & Actor-critic\\
-\hline
+A2C          & Advantage Actor Critic  (Mnih et al. 2016)                         & On-policy  & Actor-critic\\
+A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                        & On-policy  & Actor-critic\\
+TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)   & On-policy  & Policy Gradient\\
+PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)      & On-policy  & Actor-critic\\
+DQN          & Deep Q Networks   (Mnih et al. 2015)                               & Off-policy & Value-Based\\
+DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)       & Off-policy & Actor-critic\\
+TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)                & Off-policy & Actor-critic\\
+SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                         & Off-policy & Actor-critic\\
+IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)           & Off-policy & Actor-critic\\
+\hlin
 \end{tabular}}

### 2021-08-10 manuscript/manuscript.Rmd
Replacing MPC in table

@@
 nocite: |
-    @A2C, @trpo, @ddpg, @TD3, @impala, @lenz_deepmpc_2015, @weber_imagination-augmented_2018
+    @A2C, @trpo, @ddpg, @TD3, @impala, @hafner_learning_2019, @weber_imagination-augmented_2018
 header-includes:
@@
 \hline
-D-MPC        & Deep Model-Predictive Control (Lenz, Knepper, and Saxena 2015)    & Model-based \\
+PlaNet       & Deep Planning Network (Hafner et al. 2019)                        & Model-based \\
 I2A          & Imagination-Augmented Agents (Weber et al. 2018)                  & Model-based \\

### 2021-08-11 manuscript/manuscript.Rmd
Added some behavioral eco. references but not too happy with phrasing here

@@
 Applications of ML in ecology have to-date illustrated the promise of two methods: *supervised learning* [@joseph_neural_2020] and *unsupervised learning* [@valletta_applications_2017].
-However, the fields of ecology and conservation science have so far overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
+However, the fields of ecology and conservation science have largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
 Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing issues of global ecological change:
@@
 
-Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in these fields.
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, RL has not been widely applied to these fields except in behavioral ecology [@yamada_evolution_2020, @wang_reinforcement_2020].
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].

### 2021-08-11 appendices/appendix-B-examples.Rmd
don't actually eval the renv() command
GH-Actions has already done this.

@@
 
-```{r message=FALSE}
+```{r message=FALSE, eval=FALSE}
 #install.packages("renv")
@@
 
-We will compare this optimal solution with results from the several different RL algorithms.
-Unlike the optimal solution, the RL methods never know the underlying growth function used by the simulation (or even if such a function actually exists).
+We will compare this optimal solution with results from a model-free RL agent.
+Unlike the optimal solution, model-free RL agents do not know the underlying growth function used by the simulation (or even if such a function actually exists).
 These methods merely seek to learn a strategy for setting harvest quotas which maximizes the cumulative reward they receive from the environment.
 At the heart of each deep RL algorithm is a neural network or a collection of neural networks. 
-Model-free RL agents use these networks to approximate a policy function and/or a value function -- see Appendix A for more detail here. 
+Model-free, deep RL agents use these networks to approximate a policy function and/or a value function -- see Appendix A for more detail here. 
 Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyperparameters_ while the parameters of a RL algorithm that are learned during training are called parameters.
@@
 
-We assume the benefit provided by the ecosystem state is assumed to be directly proportional to the state itself, $b x_t$.  
+We assume the benefit provided by the ecosystem state is directly proportional to the state itself, $b x_t$.  
 We further assume that each year the manager has the option to slow or reverse the environmental degradation by taking action $A_t$, such that under management, the resulting environment in the next time step is given by
@@
 Because the tipping point problem is non-autonomous, we cannot solve for the optimal policy even given the model and objective (utility) function using Markov Decision Process methods.
-However, a simple heuristic solution provides a reasonable starting point for comparison.
-
-As discussed in the main text, we have no existing optimal solution to the tipping point problem, and so rely on a common heuristic strategy instead:
+However, a simple heuristic solution provides a reasonable starting point for comparison:
 select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in its current state.

### 2021-08-11 manuscript/manuscript.Rmd
Adding chades reference

@@
 
-Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, RL has not been widely applied to these fields except in behavioral ecology [@yamada_evolution_2020, @wang_reinforcement_2020].
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, RL has not been widely applied across these fields except in behavioral ecology [@yamada_evolution_2020, @wang_reinforcement_2020].
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
@@
 Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
-Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013; @ferrer-mestres_k-n-momdps_2021].
+Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013; @ferrer-mestres_k-n-momdps_2021; @chades_primer_2021].
 Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.

### 2021-08-12 manuscript/manuscript.Rmd
Added a sentence on sample efficiency

@@
 An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of the environment's current *state* and an associated *reward* as output.
-In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
+In contrast to classical approaches [@Marescot2013; @chades_primer_2021], there are few restrictions on what comprises a state or action. 
 States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
@@
 Model-free algorithms have proven successful at learning effective policies even when the underlying model is difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
+Yet, model-free algorithms tend to have poor sample efficiency [@sac] which may restrain them from being insightful on simulators that are slow to run.
 

### 2021-08-12 manuscript/manuscript.Rmd
Small changes; paragraphs on behav. etc. otw to the discussion

@@
 Applications of ML in ecology have to-date illustrated the promise of two methods: *supervised learning* [@joseph_neural_2020] and *unsupervised learning* [@valletta_applications_2017]. 
-However, the fields of ecology and conservation science have largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
+However, the fields of ecology and conservation have largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
 Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing issues of global ecological change:
@@
 
-Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, RL has not been widely applied across these fields except in behavioral ecology [@yamada_evolution_2020, @wang_reinforcement_2020].
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological conservation, RL has not been widely applied across this field.
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
@@
 
-How would an RL agent be applied to empirical data?
-In principle, this is straightforward.
-After we train an agent on a simulation environment that approximates the fishery of interest, we can query the policy of the agent to find a quota for the observed stock.
+How could an RL agent be applied to empirical data?
+One solution would be to train an agent on a simulation environment that approximates the fishery of interest then query the policy of the agent to find a quota for the observed stock.
 To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ramlegacy, see Appendix D].
@@
 Model-free algorithms have proven successful at learning effective policies even when the underlying model is difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
-Yet, model-free algorithms tend to have poor sample efficiency [@sac] which may restrain them from being insightful on simulators that are slow to run.
+Yet, current model-free algorithms tend to have poor sample efficiency [@sac] which may restrain them from being insightful on simulators that are slow to run.
 

### 2021-08-16 manuscript/manuscript.Rmd
Added sentence to mention sample inefficiency

@@
 Model-free algorithms have proven successful at learning effective policies even when the underlying model is difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
-Yet, current model-free algorithms tend to have poor sample efficiency [@sac] which may restrain them from being insightful on simulators that are slow to run.
 
@@
 Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters; @berger-tal_exploration-exploitation_2014].
-As Section 5.1 and 5.2 illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
+It is also worth acknowledging that deep RL algorithms, particularly model-free algorithms, have poor sample efficiency, which could limit deep RL from being effective on environments that are slow to run [@sac].
+Thus, as Section 5.1 and 5.2 illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
 Such examples provide important benchmarks to calibrate the performance, tuning and training requirements of RL.

### 2021-08-17 manuscript/manuscript.Rmd
Added paragraph on scaling up to large problems

@@
 
+<!-- Where/how to use deep RL -->
+Yet, a more immediate barrier to deep RL being used in conservation is deep RL's hardware requirements.
+Depending on the complexity of the RL environment and agent, the equipment necessary to train an agent can vary widely.
+The examples shown above were selected so they can be replicated on a personal computer, but more realistic problems will likely require specialized computational resources.
+For instance, one of the most notable achievements in RL, Alphastar, required 33 TPUs, processors that are specialized for deep learning, for more than 40 days [@alphastar].
+Fully detailed conservation decision-making problems will likely require comparable specialized algorithms and hardware that ecologists do not generally have access to.
+For deep RL to be an effective tool for conservation, there will need to be large investments of time and money, and extensive collaboration across computer science and ecology.
+
 <!-- But never fear, hard problems have been solved before!-->

### 2021-09-28 manuscript/manuscript.Rmd
Latest pass through; some minor edits

@@
 Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
-We include an extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
 
@@
 An agent is \emph{evaluated} by recording the cumulative reward over one episode, or the mean reward over multiple episodes.
-Within evaluation, the agent does not update its neural network; instead, the agent only uses a trained neural network to select actions.}
+Within evaluation, the agent does not update its neural network; instead, the agent uses a trained neural network to select actions.}
 \end{figure}
@@
 This interface allows for the application of software agents that can interact and learn in that environment without knowing anything about the environment's internal details.
-Using the `gym` framework, we turn existing ecological models into valid environmental simulators that can be used with any RL agents.
+Using the `gym` framework, we turn existing ecological models into valid environmental simulators that can be used with any RL agent.
 In Appendix C, we give detailed instruction on how an OpenAI `gym` is constructed.
@@
 
-- $E(o_{t}| s_{t})$: an emission distribution, which accounts for an agent's observation being different from environment's state
+- $E(o_{t}| s_{t})$: an emission distribution, which accounts for an agent's observation being different from the environment's state
 
@@
 The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
-As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$, the agent will create a trajectory, $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$.
+As the agent interacts with the environment by selecting actions according to its policy, $\pi(a_t | o_t)$^[The policy can also be conditioned on a history of observations, $(o_0, ..., o_t)$.], the agent creates a trajectory, $\tau = (s_0, o_0, a_0, \dots, s_{H-1}, o_{H-1}, a_{H-1}, s_H)$.
 From these definitions, we can provide an agent's trajectory distribution for a given policy as,
@@
 Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
-Training these agents in a variety of alternative contexts can improve there ability to generalize to other scenarios.
+Training these agents in a variety of alternative contexts can improve their ability to generalize to other scenarios.
 
@@
 <!-- Where/how to use deep RL -->
-Yet, a more immediate barrier to deep RL being used in conservation is deep RL's hardware requirements.
+Yet, a more immediate barrier to the use of deep RL in conservation is deep RL's hardware requirements.
 Depending on the complexity of the RL environment and agent, the equipment necessary to train an agent can vary widely.

### 2021-09-29 manuscript/manuscript.Rmd
Added a WIP paragraph on behavioral ecology

@@
 
+This paper does not set the precedent as the first application of RL to ecology.
+In particular, there have been a number of studies applying RL to behavioral ecology, typically with multi-agent environments [@wang_reinforcement_2020, @frankenhuis_enriching_2019, @perolat_multi-agent_2017].
+Yet, it is important to distinguish the aim of these behavioral studies from the aim of applying RL to conservation management.
+In behavioral ecology RL studies, the authors employ RL agents to represent organisms.
+Assuming that animals try to behave optimally under some objective, these authors then draw conclusions about animal behavior and evolution by observing how the RL agents learn.
+When applying deep RL to conservation management, we do not make a similar logical leap to assuming that animals behave like RL agents.
+We instead propose using RL as a way to search for solutions to unsolved decision-making problems in conservation.
+
 <!-- Model free is great bc we don't know the true models -->

### 2021-09-29 manuscript/manuscript.Rmd
Touching up behav. eco paragraph; swapped order of next 2 paragraphs

--- 2021-09-28 manuscript/manuscript.Rmd
@@
 
-<!-- Model free is great bc we don't know the true models -->
-The rapidly expanding class of model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
-Rarely do we know the underlying functional forms for ecological processes.
-Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
-Frequently available ecological data is insufficient to distinguish between possible alternative models [@knape_are_2012], or the correct model may be non-identifiable with any amount of data.
-Model-free RL approaches offer a powerful solution for this thorny issue.
-Model-free algorithms have proven successful at learning effective policies even when the underlying model is difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
+<!-- RL for behavioral ecology is not the same as RL for conservation -->
+This paper does not set the precedent as the first application of RL to ecology.
+There have been a number of studies applying RL to behavioral ecology, typically with multi-agent environments [@wang_reinforcement_2020, @frankenhuis_enriching_2019, @perolat_multi-agent_2017].
+Yet, it is important to distinguish the aim of these behavioral studies from the aim of applying RL to conservation management.
+In behavioral ecology studies, RL agents are used to represent organisms.
+Assuming that animals attempt to maximize some cumulative reward like RL agents, the authors of these studies draw conclusions about animal behavior and evolution by observing how the RL agents perform.
+When applying deep RL to conservation management, we do not make the assumption that the behavior of an RL agent is a good substitute for the behavior of an animal.
+For conservation management, we instead propose using RL as a tool to search for solutions to unsolved decision-making problems.
 
@@
 This flexibility of RL could allow us to attain better decision-making insight for our most realistic ecological models such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
+
+<!-- Model free is great bc we don't know the true models -->
+The rapidly expanding class of model-free RL algorithms is particularly appealing given the ubiquitous presence of model uncertainty in ecological dynamics.
+Rarely do we know the underlying functional forms for ecological processes.
+Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
+Frequently available ecological data is insufficient to distinguish between possible alternative models [@knape_are_2012], or the correct model may be non-identifiable with any amount of data.
+Model-free RL approaches offer a powerful solution for this thorny issue.
+Model-free algorithms have proven successful at learning effective policies even when the underlying model is difficult or impossible to learn [@Pong2020], as long as simulations of possible mechanisms are available.
 

### 2021-09-29 manuscript/manuscript.Rmd
Small rewording

@@
 When applying deep RL to conservation management, we do not make the assumption that the behavior of an RL agent is a good substitute for the behavior of an animal.
-For conservation management, we instead propose using RL as a tool to search for solutions to unsolved decision-making problems.
+For conservation management, we instead propose using RL as a tool to search for solutions to decision-making problems.
 
@@
 In contrast, the process of training an RL algorithm on a more complex problem is no different than training on a simple one: we only need access to a simulation which can generate plausible future states in response to possible actions.
-This flexibility of RL could allow us to attain better decision-making insight for our most realistic ecological models such as the individual-based models used in the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
+This flexibility of RL could allow us to attain better decision-making insight for our most realistic ecological models like those used for the management of forests and wildfire [@sortie; @wildfire], disease [@covid], marine ecosystems [@ecopath], or global climate change [@dice].
 

### 2021-09-30 manuscript/manuscript.Rmd
Changing language on behav eco paragraph but more to do here

@@
 Yet, it is important to distinguish the aim of these behavioral studies from the aim of applying RL to conservation management.
-In behavioral ecology studies, RL agents are used to represent organisms.
-Assuming that animals attempt to maximize some cumulative reward like RL agents, the authors of these studies draw conclusions about animal behavior and evolution by observing how the RL agents perform.
-When applying deep RL to conservation management, we do not make the assumption that the behavior of an RL agent is a good substitute for the behavior of an animal.
-For conservation management, we instead propose using RL as a tool to search for solutions to decision-making problems.
+<!-- Compare using RL for conserv as prescriptive, behav eco as descriptive.
+We don't assume RL is the model.
+-->
+In previous behavioral ecology studies, RL algorithms as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
+When applying deep RL to conservation management, we do not make the assumption that an RL algorithm learns analogously to how an animal learns.
+We instead propose that RL be used as a tool to search for solutions to decision-making problems.
 
@@
 Methods which must first assume something about the structure or functional form of a process before estimating the corresponding parameter can only ever be as good as those structural assumptions.
-Frequently available ecological data is insufficient to distinguish between possible alternative models [@knape_are_2012], or the correct model may be non-identifiable with any amount of data.
+Frequently, available ecological data are insufficient to distinguish between possible alternative models [@knape_are_2012], or the correct model may be non-identifiable with any amount of data.
 Model-free RL approaches offer a powerful solution for this thorny issue.

### 2021-10-15 manuscript/manuscript.Rmd
Adding some sections requested by MEE

@@
 
+# Acknowledgements {-}
+
+All authors were supported by the NSF Faculty Early Career Development Program Award 1942280. We thank Benjamin Blonder, Zachary Sunberg, and Claire Tomlin for their thoughtful comments.
+
+# Conflicts of Interest {-}
+
+The authors declare there is no conflict of interest.
+
+# Authors' Contributions {-}
+
+M.L. and C.B. developed the code and wrote the manuscript with support from K.E.A.N. and M.S.C.
+
+# Data Availability {-}
+
+We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures.
+
 \pagebreak

### 2021-10-15 manuscript/manuscript.Rmd
Adding line nos and dbl spacing

@@
   - \usepackage{amssymb}
-#  - \usepackage{lineno}
+  - \usepackage{setspace}
+  - \usepackage{lineno}
 #  - \usepackage{endfloat}
-#  - \linenumbers
+  - \linenumbers
+  - \doublespacing
 
@@
 
-\pagebreak
-
 # References

### 2021-10-15 manuscript/manuscript.Rmd
Changing figure in-text refs

@@
 A step-by-step walk-through for training agents on this environment is provided in the Appendix.
-We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig 2].
+We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig. 2].
 Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
@@
 
-The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
+The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig. 2].
 Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
-Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
+Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig. 2].
 This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
@@
 To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ramlegacy, see Appendix D].
-Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
+Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig. 3].
 In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
@@
 Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
-We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 4].
+We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig. 4].
 This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
@@
 
-Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
+Examining the replicate management trajectories and corresponding rewards [Fig.
+5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
 This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady-state strategy.

### 2021-10-18 manuscript/manuscript.Rmd
Keeping up to date

@@
   most challenging conservation decision problems. RL is uniquely
-  well suited to conservation and global change challenges for three reasons: (1)
   RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
@@
 
+  well suited to conservation and global change challenges for three reasons: (1)
 keywords:
@@
   - \usepackage{lineno}
-#  - \usepackage{endfloat}
   - \linenumbers
   - \doublespacing
-
 output: 
+  word_document: default
   rticles::arxiv_article:
     keep_tex: true
-  word_document: default
 editor_options: 

### 2021-10-19 manuscript/manuscript.Rmd
Few edits

--- 2021-10-15 manuscript/manuscript.Rmd
@@
 ---
-title: Deep Reinforcement Learning for Conservation Decisions
+title: "Deep Reinforcement Learning for Conservation Decisions"
 authors:
@@
 
-keywords:
-  - Reinforcement Learning
-  - Conservation
-  - Machine Learning
-  - Artificial Intelligence
-  - Tipping points
+output:
+  rticles::arxiv_article:
+    keep_tex: yes
+  pdf_document: default
+  word_document: default
 bibliography: references.bib
@@
 - \usepackage{lineno}
-#  - \usepackage{endfloat}
 - \linenumbers
 - \doublespacing
-
-output: 
-  rticles::arxiv_article:
-    keep_tex: true
-  word_document: default
+keywords:
+- Reinforcement Learning
+- Conservation
+- Machine Learning
+- Artificial Intelligence
+- Tipping points
 editor_options:
@@
 -->
-In previous behavioral ecology studies, RL algorithms as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
+In previous behavioral ecology studies, RL algorithms are used as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
 When applying deep RL to conservation management, we do not make the assumption that an RL algorithm learns analogously to how an animal learns.
@@
 
-The authors declare there is no conflict of interest.
+The authors declare there are no conflicts of interest.
 
@@
 
-We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures.
+We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures herein.
 

### 2021-10-19 manuscript/manuscript.Rmd
Reverting to main's head section

--- 2021-10-15 manuscript/manuscript.Rmd
@@
   - \usepackage{amssymb}
+  - \usepackage{lineno}
   - \usepackage{setspace}
-  - \usepackage{lineno}
 #  - \usepackage{endfloat}

### 2021-10-21 manuscript/manuscript.Rmd
Fixing abstract and section titles

@@
 abstract: |
+  1.
 Can machine learning help us make better decisions about a changing planet?
-  In this paper, we illustrate and discuss the potential of a promising 
-  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
-  most challenging conservation decision problems. RL is uniquely
-  well suited to conservation and global change challenges for three reasons: (1)
-  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
-  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
-  utilize rather than replace existing models, simulations, and the knowledge they contain.
-  We provide a conceptual and technical introduction to RL and its relevance to ecological
-  and conservation challenges, including examples of a problem in setting
-  fisheries quotas and in managing ecological tipping points.
-  Four appendices with annotated code provide a tangible introduction to researchers 
-  looking to adopt, evaluate, or extend these approaches.
+In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as deep reinforcement learning (RL) to help tackle the most challenging conservation decision problems.
+We provide a conceptual and technical introduction to deep RL as well as annotated code so that researchers can adopt, evaluate and extend these approaches.
+  2.
+RL explicitly focuses on designing an agent who interacts with an environment which is dynamic and uncertain.
+Deep RL is the subfield of RL that incorporates deep neural networks into the agent.
+We train deep RL agents to solve sequential decision-making problems in setting fisheries quotas and managing ecological tipping points.
+  3.
+We show that a deep RL agent is able to learn a nearly optimal solution for the fisheries management problem.
+For the tipping point problem, we show that a deep RL agent can outperform a sensible rule-of-thumb strategy.
+  4.
+Our results demonstrate that deep RL has the potential to solve challenging decision problems in conservation.
+We discuss the obstacles that may restrict the successful application of deep RL to more realistic conservation problems.
+We ultimately set forth a research framework based on well-posed, public challenges so that ecologists and computer scientists can collaborate towards solving hard decision-making problems in conservation.
 
@@
 
-the number of words: **4,814**
+the number of words: **5655**
 
@@
 
-# RL overview
+# Materials and Methods
 
@@
 
-# Examples
+# Results
 

### 2021-10-21 manuscript/manuscript.Rmd
Getting keywords alphabetized

@@
 keywords:
-  - Reinforcement Learning
+  - Artificial Intelligence
   - Conservation
   - Machine Learning
-  - Artificial Intelligence
+  - Reinforcement Learning
   - Tipping points

### 2022-01-21 manuscript/manuscript.Rmd
Revise intro to address question of realistic models

--- 2021-10-19 manuscript/manuscript.Rmd
@@
   Can machine learning help us make better decisions about a changing planet?
-  In this paper, we illustrate and discuss the potential of a promising 
-  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
-  most challenging conservation decision problems. RL is uniquely
-  well suited to conservation and global change challenges for three reasons: (1)
-  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
-  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
-  utilize rather than replace existing models, simulations, and the knowledge they contain.
-  We provide a conceptual and technical introduction to RL and its relevance to ecological
-  and conservation challenges, including examples of a problem in setting
-  fisheries quotas and in managing ecological tipping points.
-  Four appendices with annotated code provide a tangible introduction to researchers 
-  looking to adopt, evaluate, or extend these approaches.
+  In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as _reinforcement learning_ (RL) to help tackle the most challenging conservation decision problems.
+  RL is uniquely well suited to conservation and global change challenges for three reasons: 
+  (1) RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
+  (2) RL approaches do not require massive amounts of data, 
+  (3) RL approaches would utilize rather than replace existing models, simulations, and the knowledge they contain.
+  While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
+  the required expertise and computational cost may place these applications may be beyond the reach of all but international technology firms.
+  Such an advent would represent a sea-change not only in what problems we can solve, but what stakeholders are involved in solving them.
+  Ecologists must establish a better understanding of how these algorithms work and fail if we are to realize this potential and avoid the pitfalls such a transition would bring.
+  To this end, we provide a conceptual and technical introduction to RL and its relevance to ecological and conservation challenges, including examples of a problem in setting fisheries quotas and in managing ecological tipping points.
+  Four appendices with annotated code provide a tangible introduction to researchers looking to adopt, evaluate, or extend these approaches.
 
@@
   - \usepackage{amssymb}
-  - \usepackage{lineno}
-  - \usepackage{setspace}
+#  - \usepackage{lineno}
 #  - \usepackage{endfloat}
-  - \linenumbers
-  - \doublespacing
+#  - \linenumbers
 
@@
 
+While deep RL may have the potential to open up such intractable problems, it also risks making those problems tractable only by stakeholders with access to extensive computational resources and expertise.
+It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.
+Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially avilable cloud computing platforms place the training of a single model at over $30 million [@alphacost],
+and many realistic ecological problems will involve greater complexity than board games of two players on a 19 x 19 grid [@alphago].
+While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.
+We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though their application will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
+
 In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
-Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
+We focus on examining the potential and limitations of deep RL through the lens of simple, classical models.
+Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2017].
+As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.
+Levins esposed simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
+More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general).
+Such simple, realistic and general models are still the bedrock of most theory and practice today, even when implementation turns to tactical models for quantitative guidance (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or $R_0$ in epidemiology, remain important concepts in management even though they do not exist in typical tactical models).
+These models provide an ideal first benchmark for evaluating the performance of emerging methods of deep RL for several reasons:
+Firstly, for some cases the optimal solution is already known, providing a clear standard-of-comparison to evaluate RL performance. 
+Prior work sometimes overlooks this essential step, assuming that whatever behavior an RL agent produces is sufficiently optimal or otherwise an accurate proxy for behavior arising from animal learning or evolution.
+As our evaluations will illustrate, such an assumption can be quickly misleading. 
+Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of threshholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
+Many readers can thus benefit from exisiting knowledge and intutition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.
+It would be unwise to consider applications of RL on more complex bespoke models without first analyzing these classic and well-understood cases.
+Third, these models include or can easily be extended to contexts for which the optimal management policy is unknown or inaccessible to classical methods.
+Our implementations of these models in two python modules published to `PyPi` code archive include many such variations which represent open problems for RL.
 We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+
+We do recognize the need for more complex, "precise" models to derive quanitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
+The examples here should provide ecologists with the greater background to engage in the collaborative development of such methods, even as they also highlight challenges which ecological problems pose to existing techniques.
+Even the simple models we consider here contain many sources of complexity -- significant stochasticity, non-linear dynamics, hysteresis -- that are likely commonplace to ecological problems but have not been well represtented in RL research to date.
+Such work is beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.
 
@@
 A step-by-step walk-through for training agents on this environment is provided in the Appendix.
-We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig. 2].
+We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig 2].
 Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.
@@
 
-The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig. 2].
+The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
 Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
-Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig. 2].
+Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
 This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
@@
 To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ramlegacy, see Appendix D].
-Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig. 3].
+Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
 In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
@@
 Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
-We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig. 4].
+We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 4].
 This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
@@
 
-Examining the replicate management trajectories and corresponding rewards [Fig.
-5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
+Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
 This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady-state strategy.
@@
 -->
-In previous behavioral ecology studies, RL algorithms are used as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
+In previous behavioral ecology studies, RL algorithms as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
 When applying deep RL to conservation management, we do not make the assumption that an RL algorithm learns analogously to how an animal learns.
@@
 
-# Acknowledgements {-}
-
-All authors were supported by the NSF Faculty Early Career Development Program Award 1942280. We thank Benjamin Blonder, Zachary Sunberg, and Claire Tomlin for their thoughtful comments.
-
-# Conflicts of Interest {-}
-
-The authors declare there are no conflicts of interest.
-
-# Authors' Contributions {-}
-
-M.L. and C.B. developed the code and wrote the manuscript with support from K.E.A.N. and M.S.C.
-
-# Data Availability {-}
-
-We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures herein.
+\pagebreak
 

### 2022-01-27 appendices/failure_plot.Rmd
Stashing here for now

---
title: "failure_plot"
output: html_document
---

```{r message=FALSE}
#install.packages("renv")
renv::restore()
```

```{r setup, cache = FALSE, message=FALSE}
# R dependencies
library(tidyverse)
library(patchwork)
library(reticulate)

## Python dependencies loaded via R
sb3      = import("stable_baselines3")
torch    = import("torch")
gym      = import("gym")
gym_fishing = import("gym_fishing")
gym_conservation = import("gym_conservation")

source("../R/plotting-helpers.R")
```

```{r knitr, include = FALSE}
# This chunk controls plot aesthetics only, and can be omitted with no material change
knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4, cache = FALSE)
ggplot2::theme_set(ggplot2::theme_bw())

scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
pal = ggthemes::solarized_pal()(8)
txtcolor = "#586e75"

colors <- set_names(c(pal[c(1,3,8,2)], txtcolor, pal[6]),
                    c("Optimal", "TD3_untuned", "TD3_tuned", "RL", "historical", "steady-state"))
scale_colour_discrete <- function(...) scale_colour_manual(..., values=colors)
scale_fill_discrete <- function(...) scale_fill_manual(..., values=colors)
```

```{r}
## reproducible settings
np = import("numpy")
seed = 24L # integer
np$random$seed(seed)

set.seed(seed)
```

```{r }
## initialize the environment
env_high = gym$make("fishing-v1", sigma=0.1, init_state=1.5)
env_low = gym$make("fishing-v1", sigma=0.1)
```

```{r}
# Simulate under the optimal solution (given the model)
opt = gym_fishing$models$escapement(env_low)
opt_sims_high = env_high$simulate(opt, reps = 100L)
opt_sims_low = env_low$simulate(opt, reps = 100L)
opt_policy = env_low$policyfn(opt)
```

```{r}
# add a column indicating the model
sims_df_high <- opt_sims_high %>% mutate(model = "Optimal")
sims_df_low <- opt_sims_low %>% mutate(model = "Optimal")
policy_df <- opt_policy  %>% mutate(model = "Optimal")
```

```{r}
agent = sb3$TD3$load("../python/cache/td3_untuned")
agent_sims_high = env_high$simulate(agent, reps = 100L)
agent_sims_low = env_low$simulate(agent, reps = 100L)
agent_policy = env_low$policyfn(agent, reps = 5L)
```

```{r}
# stack result simulations with those above and plot:
sims_df_high <- agent_sims_high %>% 
  mutate(model = "TD3_untuned") %>% 
  bind_rows(sims_df_high)

# stack result simulations with those above and plot:
sims_df_low <- agent_sims_low %>% 
  mutate(model = "TD3_untuned") %>% 
  bind_rows(sims_df_low)

policy_df <- agent_policy  %>% 
  mutate(model = "TD3_untuned")  %>% 
  bind_rows(policy_df)
```

```{r}
td3 <- sb3$TD3$load("../python/cache/td3_tuned.zip")
td3_sims_high = env_high$simulate(td3, reps = 100L)
td3_sims_low = env_low$simulate(td3, reps = 100L)
td3_policy = env_low$policyfn(td3)
```

```{r}
# stack result simulations with those above and plot:
sims_df_high <- td3_sims_high %>% 
  mutate(model = "TD3_tuned") %>% 
  bind_rows(sims_df_high)
sims_df_low <- td3_sims_low %>% 
  mutate(model = "TD3_tuned") %>% 
  bind_rows(sims_df_low)

policy_df <- td3_policy  %>% 
  mutate(model = "TD3_tuned")  %>% 
  bind_rows(policy_df)

( plot_sims(sims_df_low, "none") + plot_sims(sims_df_high, "right")) / ( plot_policy(policy_df))
```

### 2022-01-27 R/plotting-helpers.R

@@ -5,21 +5,22 @@ library(tidyverse)
 ymin = function(x) last(x[(ntile(x, 20)==1)])
 ymax = function(x) last(x[(ntile(x, 20)==19)])
 
-plot_sims <- function(sims_df)
+plot_sims <- function(sims_df, legend)
   sims_df %>%
   group_by(time, model) %>%
   summarise(ymin = ymin(state),
             ymax = ymax(state),
             state = mean(state), .groups = "drop") %>%
   ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) +
-  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
+  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model)) +
+  theme(legend.position = legend)
 
 plot_policy <- function(policy_df)
   policy_df %>% ggplot(aes(state, action,
                            group=interaction(rep, model),
                            col = model)) +
   geom_line(show.legend = FALSE) +
-  coord_cartesian(xlim = c(0, 1.2))
+  coord_cartesian(xlim = c(0, 1.5))
 
 plot_reward <- function(sims_df, gamma = 1)
   sims_df %>%

### 2022-01-28 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
committing suggestion directly in PR, how cool

Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 We do recognize the need for more complex, "precise" models to derive quanitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
-The examples here should provide ecologists with the greater background to engage in the collaborative development of such methods, even as they also highlight challenges which ecological problems pose to existing techniques.
+The examples here should provide ecologists with a greater background to engage in the collaborative development of such methods, while also highlighting challenges which ecological problems pose to existing techniques.
 Even the simple models we consider here contain many sources of complexity -- significant stochasticity, non-linear dynamics, hysteresis -- that are likely commonplace to ecological problems but have not been well represtented in RL research to date.

### 2022-01-28 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 Third, these models include or can easily be extended to contexts for which the optimal management policy is unknown or inaccessible to classical methods.
-Our implementations of these models in two python modules published to `PyPi` code archive include many such variations which represent open problems for RL.
+Our implementations of these models have been published to the python-based `PyPi` code archive and include many such variations which represent open problems for RL.
 We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.

### 2022-01-28 manuscript/manuscript.Rmd
Adding draft of a table here

--- 2022-01-28 manuscript/manuscript.Rmd
@@
   Can machine learning help us make better decisions about a changing planet?
-  In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as _reinforcement learning_ (RL) to help tackle the most challenging conservation decision problems.
-  RL is uniquely well suited to conservation and global change challenges for three reasons: 
-  (1) RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
-  (2) RL approaches do not require massive amounts of data, 
-  (3) RL approaches would utilize rather than replace existing models, simulations, and the knowledge they contain.
-  While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
-  the required expertise and computational cost may place these applications may be beyond the reach of all but international technology firms.
-  Such an advent would represent a sea-change not only in what problems we can solve, but what stakeholders are involved in solving them.
-  Ecologists must establish a better understanding of how these algorithms work and fail if we are to realize this potential and avoid the pitfalls such a transition would bring.
-  To this end, we provide a conceptual and technical introduction to RL and its relevance to ecological and conservation challenges, including examples of a problem in setting fisheries quotas and in managing ecological tipping points.
-  Four appendices with annotated code provide a tangible introduction to researchers looking to adopt, evaluate, or extend these approaches.
+  In this paper, we illustrate and discuss the potential of a promising 
+  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
+  most challenging conservation decision problems. RL is uniquely
+  well suited to conservation and global change challenges for three reasons: (1)
+  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
+  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
+  utilize rather than replace existing models, simulations, and the knowledge they contain.
+  We provide a conceptual and technical introduction to RL and its relevance to ecological
+  and conservation challenges, including examples of a problem in setting
+  fisheries quotas and in managing ecological tipping points.
+  Four appendices with annotated code provide a tangible introduction to researchers 
+  looking to adopt, evaluate, or extend these approaches.
 
@@
 
-While deep RL may have the potential to open up such intractable problems, it also risks making those problems tractable only by stakeholders with access to extensive computational resources and expertise.
-It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.
-Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially avilable cloud computing platforms place the training of a single model at over $30 million [@alphacost],
-and many realistic ecological problems will involve greater complexity than board games of two players on a 19 x 19 grid [@alphago].
-While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.
-We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though their application will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
-
 In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-We focus on examining the potential and limitations of deep RL through the lens of simple, classical models.
-Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2017].
-As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.
-Levins esposed simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
-More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general).
-Such simple, realistic and general models are still the bedrock of most theory and practice today, even when implementation turns to tactical models for quantitative guidance (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or $R_0$ in epidemiology, remain important concepts in management even though they do not exist in typical tactical models).
-These models provide an ideal first benchmark for evaluating the performance of emerging methods of deep RL for several reasons:
-Firstly, for some cases the optimal solution is already known, providing a clear standard-of-comparison to evaluate RL performance. 
-Prior work sometimes overlooks this essential step, assuming that whatever behavior an RL agent produces is sufficiently optimal or otherwise an accurate proxy for behavior arising from animal learning or evolution.
-As our evaluations will illustrate, such an assumption can be quickly misleading. 
-Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of threshholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
-Many readers can thus benefit from exisiting knowledge and intutition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.
-It would be unwise to consider applications of RL on more complex bespoke models without first analyzing these classic and well-understood cases.
-Third, these models include or can easily be extended to contexts for which the optimal management policy is unknown or inaccessible to classical methods.
-Our implementations of these models in two python modules published to `PyPi` code archive include many such variations which represent open problems for RL.
+We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
+Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
 We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
-
-We do recognize the need for more complex, "precise" models to derive quanitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
-The examples here should provide ecologists with a greater background to engage in the collaborative development of such methods, while also highlighting challenges which ecological problems pose to existing techniques.
-Even the simple models we consider here contain many sources of complexity -- significant stochasticity, non-linear dynamics, hysteresis -- that are likely commonplace to ecological problems but have not been well represtented in RL research to date.
-Such work is beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.
 
@@
 Successfully applying RL to complex ecological problems is no easy task.
-Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters; @berger-tal_exploration-exploitation_2014].
+Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons, see Table 2, like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters; @berger-tal_exploration-exploitation_2014].
 It is also worth acknowledging that deep RL algorithms, particularly model-free algorithms, have poor sample efficiency, which could limit deep RL from being effective on environments that are slow to run [@sac].
@@
 A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021, @chapman_promoting_2021].
+
+\begin{table}
+\begin{tabular}{l l }
+\hline
+Practical Challenges with Deep RL & Description   \\
+\hline
+   Generalization                   & blah blah  \\
+   Intransparency                   & blah blah \\
+   Hyperparameter Instability       & blah blah  \\
+   Reward Misspecification          & blah blah \\
+   Capital Intensive                & blah blah  \\
+   Sample Inefficiency              & blah blah \\
+
+\hline
+\end{tabular}
+\caption{Survey of common deep RL Challenges.}
+\end{table}
 

### 2022-01-29 manuscript/manuscript.Rmd
Updating this table

@@
 \begin{table}
-\begin{tabular}{l l }
+\renewcommand{\arraystretch}{1.25}
+\begin{tabular}{p{5 cm} p{10 cm}}
 \hline
-Practical Challenges with Deep RL & Description   \\
+Issue & Description   \\
 \hline
-   Generalization                   & blah blah  \\
-   Intransparency                   & blah blah \\
-   Hyperparameter Instability       & blah blah  \\
-   Reward Misspecification          & blah blah \\
-   Capital Intensive                & blah blah  \\
-   Sample Inefficiency              & blah blah \\
+   Generalization                   & Agents struggle to adapt to tasks not seen in training.   \\
+   Reproducibility                  & It can be very challenging to replicate results due to a host of reasons like differences in computational hardware. \\
+   Intransparency                   & RL users cannot interpret why agents select actions. \\
+   Hyperparameter Instability       & Agent performance can vary significantly over slight alterations in hyperparameters, like initialization seed.   \\
+   Reward Misspecification          & Agents commonly learn undesirable behavior that still maximizes the RL objective. \\
+   Capital Intensive                & Landmark successes like AlphaGo and AlphaStar have required very large teams of researchers and large amounts of computational power.  \\
+   Sample Inefficiency              & Current algorithms require large amounts of interaction with the environment to achieve reward maximization.  \\
 
@@
 \end{tabular}
-\caption{Survey of common deep RL Challenges.}
+\caption{Practical issues with deep RL.}
 \end{table}

### 2022-01-29 manuscript/manuscript.Rmd
Changing one issue name

@@
    Reproducibility                  & It can be very challenging to replicate results due to a host of reasons like differences in computational hardware. \\
-   Intransparency                   & RL users cannot interpret why agents select actions. \\
+   Lack of Transparency             & RL users cannot interpret why agents select actions. \\
    Hyperparameter Instability       & Agent performance can vary significantly over slight alterations in hyperparameters, like initialization seed.   \\
    Reward Misspecification          & Agents commonly learn undesirable behavior that still maximizes the RL objective. \\
-   Capital Intensive                & Landmark successes like AlphaGo and AlphaStar have required very large teams of researchers and large amounts of computational power.  \\
+   High Capital Demands            & Landmark successes like AlphaGo and AlphaStar have required very large teams of researchers and large amounts of computational power.  \\
    Sample Inefficiency              & Current algorithms require large amounts of interaction with the environment to achieve reward maximization.  \\

### 2022-01-31 manuscript/manuscript.Rmd

--- 2022-01-28 manuscript/manuscript.Rmd
@@
 nocite: |
-    @A2C, @trpo, @ddpg, @TD3, @impala, @hafner_learning_2019, @weber_imagination-augmented_2018
+    @A2C, @trpo, @ddpg, @TD3, @impala, @hafner_learning_2019, @weber_imagination-augmented_2018, @kirk_survey_2022
 header-includes:
@@
 \begin{table}
-\begin{tabular}{l l }
+\renewcommand{\arraystretch}{1.25}
+\begin{tabular}{p{5 cm} p{10 cm}}
 \hline
-Practical Challenges with Deep RL & Description   \\
+Issue & Description   \\
 \hline
-   Generalization                   & blah blah  \\
-   Intransparency                   & blah blah \\
-   Hyperparameter Instability       & blah blah  \\
-   Reward Misspecification          & blah blah \\
-   Capital Intensive                & blah blah  \\
-   Sample Inefficiency              & blah blah \\
+   Generalization                   & Agents struggle to adapt to tasks not seen in training (Kirk et al. 2022).   \\
+   Reproducibility                  & It can be very challenging to replicate results due to a host of reasons like differences in computational hardware (Henderson et al. 2019). \\
+   Lack of Transparency             & Deep RL users cannot interpret why agents select actions (Castelvecchi 2016). \\
+   Hyperparameter Instability       & Agent performance can vary significantly over slight alterations in hyperparameters, like initialization seed (Henderson et al. 2019).   \\
+   Reward Misspecification          & Agents commonly learn undesirable behavior that still maximizes the RL objective  (Hadfield-Menell et al. 2020). \\
+   High Capital Demands            & Landmark successes like AlphaGo and AlphaStar have required very large teams of researchers and large amounts of computational power (Silver et al.
+2017; Vinyals et al.
+2019).
+\\
+   Sample Inefficiency              & Current algorithms require large amounts of interaction with the environment to achieve reward maximization (Haarnoja et al. 2018).  \\
 

### 2022-01-31 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

--- 2022-01-29 manuscript/manuscript.Rmd
@@
   Can machine learning help us make better decisions about a changing planet?
-  In this paper, we illustrate and discuss the potential of a promising 
-  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
-  most challenging conservation decision problems. RL is uniquely
-  well suited to conservation and global change challenges for three reasons: (1)
-  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
-  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
-  utilize rather than replace existing models, simulations, and the knowledge they contain.
-  We provide a conceptual and technical introduction to RL and its relevance to ecological
-  and conservation challenges, including examples of a problem in setting
-  fisheries quotas and in managing ecological tipping points.
-  Four appendices with annotated code provide a tangible introduction to researchers 
-  looking to adopt, evaluate, or extend these approaches.
+  In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as _reinforcement learning_ (RL) to help tackle the most challenging conservation decision problems.
+  RL is uniquely well suited to conservation and global change challenges for three reasons: 
+  (1) RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
+  (2) RL approaches do not require massive amounts of data, 
+  (3) RL approaches would utilize rather than replace existing models, simulations, and the knowledge they contain.
+  While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
+  the required expertise and computational cost may place these applications may be beyond the reach of all but international technology firms.
+  Such an advent would represent a sea-change not only in what problems we can solve, but what stakeholders are involved in solving them.
+  Ecologists must establish a better understanding of these algorithms if we are to realize the potential of deep RL and avoid the pitfalls such a transition would bring.
+  To this end, we provide a conceptual and technical introduction to RL and its relevance to ecological and conservation challenges, including examples of a problem in setting fisheries quotas and in managing ecological tipping points.
+  Four appendices with annotated code provide a tangible introduction to researchers looking to adopt, evaluate, or extend these approaches.
 
@@
 
+While deep RL may have the potential to open up such intractable problems, it also risks making those problems tractable only by stakeholders with access to extensive computational resources and expertise.
+It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.
+Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially avilable cloud computing platforms place the training of a single model at over $30 million [@alphacost],
+and many realistic ecological problems will involve greater complexity than board games of two players on a 19 x 19 grid [@alphago].
+While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.
+We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though their application will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
+
 In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
-We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
-Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
+We focus on examining the potential and limitations of deep RL through the lens of simple, classical models.
+Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2017].
+As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.
+Levins esposed simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
+More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general).
+Such simple, realistic and general models are still the bedrock of most theory and practice today, even when implementation turns to tactical models for quantitative guidance (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or $R_0$ in epidemiology, remain important concepts in management even though they do not exist in typical tactical models).
+These models provide an ideal first benchmark for evaluating the performance of emerging methods of deep RL for several reasons:
+Firstly, for some cases the optimal solution is already known, providing a clear standard-of-comparison to evaluate RL performance. 
+Prior work sometimes overlooks this essential step, assuming that whatever behavior an RL agent produces is sufficiently optimal or otherwise an accurate proxy for behavior arising from animal learning or evolution.
+As our evaluations will illustrate, such an assumption can be quickly misleading. 
+Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of threshholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
+Many readers can thus benefit from exisiting knowledge and intutition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.
+It would be unwise to consider applications of RL on more complex bespoke models without first analyzing these classic and well-understood cases.
+Third, these models include or can easily be extended to contexts for which the optimal management policy is unknown or inaccessible to classical methods.
+Our implementations of these models have been published to the python-based `PyPi` code archive and include many such variations which represent open problems for RL.
 We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.
+
+We do recognize the need for more complex, "precise" models to derive quanitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
+The examples here should provide ecologists with a greater background to engage in the collaborative development of such methods, while also highlighting challenges which ecological problems pose to existing techniques.
+Even the simple models we consider here contain many sources of complexity -- significant stochasticity, non-linear dynamics, hysteresis -- that are likely commonplace to ecological problems but have not been well represtented in RL research to date.
+Such work is beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.
 
@@
 Successfully applying RL to complex ecological problems is no easy task.
-Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons, see Table 2, like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters; @berger-tal_exploration-exploitation_2014].
+Even on relatively uncomplicated environments, training an RL agent can be more challenging than expected due to an entanglement of reasons like hyperparameter instability and poor exploration that can be very difficult to resolve [@drlthatmatters; @berger-tal_exploration-exploitation_2014].
 It is also worth acknowledging that deep RL algorithms, particularly model-free algorithms, have poor sample efficiency, which could limit deep RL from being effective on environments that are slow to run [@sac].
@@
 A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021, @chapman_promoting_2021].
-
-\begin{table}
-\renewcommand{\arraystretch}{1.25}
-\begin{tabular}{p{5 cm} p{10 cm}}
-\hline
-Issue & Description   \\
-\hline
-   Generalization                   & Agents struggle to adapt to tasks not seen in training.   \\
-   Reproducibility                  & It can be very challenging to replicate results due to a host of reasons like differences in computational hardware. \\
-   Intransparency                   & RL users cannot interpret why agents select actions. \\
-   Hyperparameter Instability       & Agent performance can vary significantly over slight alterations in hyperparameters, like initialization seed.   \\
-   Reward Misspecification          & Agents commonly learn undesirable behavior that still maximizes the RL objective. \\
-   Capital Intensive                & Landmark successes like AlphaGo and AlphaStar have required very large teams of researchers and large amounts of computational power.  \\
-   Sample Inefficiency              & Current algorithms require large amounts of interaction with the environment to achieve reward maximization.  \\
-
-\hline
-\end{tabular}
-\caption{Practical issues with deep RL.}
-\end{table}
 

### 2022-01-31 manuscript/manuscript.Rmd
:pen:

@@
   While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
-  the required expertise and computational cost may place these applications may be beyond the reach of all but international technology firms.
-  Such an advent would represent a sea-change not only in what problems we can solve, but what stakeholders are involved in solving them.
-  Ecologists must establish a better understanding of these algorithms if we are to realize the potential of deep RL and avoid the pitfalls such a transition would bring.
+  the required expertise and computational cost may place these applications beyond the reach of all but international technology firms.
+  Ecologists must establish a better understanding of how these algorithms work and fail if we are to realize this potential and avoid the pitfalls such a transition would bring.
   To this end, we provide a conceptual and technical introduction to RL and its relevance to ecological and conservation challenges, including examples of a problem in setting fisheries quotas and in managing ecological tipping points.

### 2022-01-31 manuscript/manuscript.Rmd

@@
   While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
-  the required expertise and computational cost may place these applications beyond the reach of all but international technology firms.
+  the required expertise and computational cost may place these applications beyond the reach of all but large, international technology firms.
   Ecologists must establish a better understanding of how these algorithms work and fail if we are to realize this potential and avoid the pitfalls such a transition would bring.

### 2022-01-31 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.
-Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially avilable cloud computing platforms place the training of a single model at over $30 million [@alphacost],
+Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially available cloud computing platforms place the training of a single model at over $30 million [@alphacost],
 and many realistic ecological problems will involve greater complexity than board games of two players on a 19 x 19 grid [@alphago].

### 2022-01-31 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 The examples here should provide ecologists with a greater background to engage in the collaborative development of such methods, while also highlighting challenges which ecological problems pose to existing techniques.
-Even the simple models we consider here contain many sources of complexity -- significant stochasticity, non-linear dynamics, hysteresis -- that are likely commonplace to ecological problems but have not been well represtented in RL research to date.
+Even the simple models we consider here contain sources of complexity such as hysteresis that are likely commonplace to ecological problems but have not been well represented in RL research to date. 
 Such work is beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.

### 2022-01-31 manuscript/manuscript.Rmd
Update manuscript/manuscript.Rmd
Co-authored-by: mlap <marcuslapeyrolerie@me.com>

@@
 Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of threshholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
-Many readers can thus benefit from exisiting knowledge and intutition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.
+Many readers can thus benefit from existing knowledge and intuition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.
 It would be unwise to consider applications of RL on more complex bespoke models without first analyzing these classic and well-understood cases.

### 2022-02-01 manuscript/manuscript.Rmd
:pen:

@@
 As our evaluations will illustrate, such an assumption can be quickly misleading. 
-Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of threshholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
+Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of thresholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
 Many readers can thus benefit from existing knowledge and intuition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.
-It would be unwise to consider applications of RL on more complex bespoke models without first analyzing these classic and well-understood cases.
 Third, these models include or can easily be extended to contexts for which the optimal management policy is unknown or inaccessible to classical methods.
@@
 
-We do recognize the need for more complex, "precise" models to derive quanitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
+We do recognize the need for more complex, "precise" models to derive quantitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
 The examples here should provide ecologists with a greater background to engage in the collaborative development of such methods, while also highlighting challenges which ecological problems pose to existing techniques.

### 2022-02-01 manuscript/manuscript.Rmd
polite

@@
 Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially available cloud computing platforms place the training of a single model at over $30 million [@alphacost],
-and many realistic ecological problems will involve greater complexity than board games of two players on a 19 x 19 grid [@alphago].
+and many realistic ecological problems will involve even greater complexity than these landmark examples [@alphago].
 While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.

### 2022-02-01 manuscript/manuscript.Rmd

@@
 A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021, @chapman_promoting_2021].
+
+\begin{table}
+\renewcommand{\arraystretch}{1.25}
+\begin{tabular}{p{5 cm} p{10 cm}}
+\hline
+Issue & Description   \\
+\hline
+   Generalization                   & Agents struggle to adapt to tasks not seen in training (Kirk et al. 2022).   \\
+   Reproducibility                  & It can be very challenging to replicate results due to a host of reasons like differences in computational hardware (Henderson et al. 2019). \\
+   Lack of Transparency             & Deep RL users cannot interpret why agents select actions (Castelvecchi 2016). \\
+   Hyperparameter Instability       & Agent performance can vary significantly over slight alterations in hyperparameters, like initialization seed (Henderson et al. 2019).   \\
+   Reward Misspecification          & Agents commonly learn undesirable behavior that still maximizes the RL objective  (Hadfield-Menell et al. 2020). \\
+   High Capital Demands            & Landmark successes like AlphaGo and AlphaStar have required very large teams of researchers and large amounts of computational power (Silver et al.
+2017; Vinyals et al.
+2019).
+\\
+   Sample Inefficiency              & Current algorithms require large amounts of interaction with the environment to achieve reward maximization (Haarnoja et al. 2018).  \\
+
+\hline
+\end{tabular}
+\caption{Practical issues with deep RL.}
+\end{table}
 

### 2022-02-01 manuscript/manuscript.Rmd
Small edits

@@
 As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.
-Levins esposed simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
+Levins espoused simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
 More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general).
@@
 Firstly, for some cases the optimal solution is already known, providing a clear standard-of-comparison to evaluate RL performance. 
-Prior work sometimes overlooks this essential step, assuming that whatever behavior an RL agent produces is sufficiently optimal or otherwise an accurate proxy for behavior arising from animal learning or evolution.
+Prior work sometimes overlooks this essential step, assuming that whatever behavior an RL agent produces is sufficiently optimal [@DQN].
 As our evaluations will illustrate, such an assumption can be quickly misleading. 
@@
 Even the simple models we consider here contain sources of complexity such as hysteresis that are likely commonplace to ecological problems but have not been well represented in RL research to date. 
-Such work is beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.
+Applying RL to a range of "precise" models is currently beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.
 

### 2022-02-01 manuscript/manuscript.Rmd
Small edits
Deleting empty commentd
Last touches here likely

@@
 
-While deep RL may have the potential to open up such intractable problems, it also risks making those problems tractable only by stakeholders with access to extensive computational resources and expertise.
+While deep RL may have the potential to open up such intractable problems, it also risks making those problems tractable only for stakeholders with access to extensive computational resources and expertise.
 It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.

### 2022-02-01 manuscript/manuscript.Rmd
update refs

--- 2022-02-01 manuscript/manuscript.Rmd
@@
 It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.
-Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially available cloud computing platforms place the training of a single model at over $30 million [@alphacost],
-and many realistic ecological problems will involve even greater complexity than these landmark examples [@alphago].
+Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially available cloud computing platforms place the training of a single model at over $35 million [@alphacost; @openai_costs; @alphaGoZero],
+and many realistic ecological problems will involve even greater complexity than these landmark examples [@alphaGoZero; @alphazero; @openai_learning_2019].
 While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.
@@
 We focus on examining the potential and limitations of deep RL through the lens of simple, classical models.
-Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2017].
+Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2018].
 As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.
-Levins esposed simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
+Levins espoused simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
 More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general).
@@
 As our evaluations will illustrate, such an assumption can be quickly misleading. 
-Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of thresholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].
+Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of thresholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2015].
 Many readers can thus benefit from existing knowledge and intuition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model.

### 2022-02-02 manuscript/manuscript.Rmd
Another sentence edit

--- 2022-02-01 manuscript/manuscript.Rmd
@@
   - \usepackage{amssymb}
-#  - \usepackage{lineno}
+  - \usepackage{lineno}
+  - \usepackage{setspace}
 #  - \usepackage{endfloat}
 #  - \linenumbers
+#  - \doublespacing
 
@@
 While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.
-We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though their application will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
+We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though the application of RL to conservation will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
 

### 2022-02-02 manuscript/manuscript.Rmd

@@
 We focus on examining the potential and limitations of deep RL through the lens of simple, classical models.
-Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2017].
+Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2018].
 As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.

### 2022-02-03 manuscript/manuscript.Rmd
Looking like the final version here

@@
 abstract: |
+  1.
 Can machine learning help us make better decisions about a changing planet?
-  In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as _reinforcement learning_ (RL) to help tackle the most challenging conservation decision problems.
-  RL is uniquely well suited to conservation and global change challenges for three reasons: 
-  (1) RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
-  (2) RL approaches do not require massive amounts of data, 
-  (3) RL approaches would utilize rather than replace existing models, simulations, and the knowledge they contain.
-  While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
-  the required expertise and computational cost may place these applications beyond the reach of all but large, international technology firms.
+In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as deep reinforcement learning (RL) to help tackle the most challenging conservation decision problems.
+We provide a conceptual and technical introduction to deep RL as well as annotated code so that researchers can adopt, evaluate and extend these approaches.
+  2.
+RL explicitly focuses on designing an agent who interacts with an environment which is dynamic and uncertain.
+Deep RL is the subfield of RL that incorporates deep neural networks into the agent.
+We train deep RL agents to solve sequential decision-making problems in setting fisheries quotas and managing ecological tipping points.
+  3.
+We show that a deep RL agent is able to learn a nearly optimal solution for the fisheries management problem.
+For the tipping point problem, we show that a deep RL agent can outperform a sensible rule-of-thumb strategy.
+  4.
+Our results demonstrate that deep RL has the potential to solve challenging decision problems in conservation.
+While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable -- the required expertise and computational cost may place these applications beyond the reach of all but large, international technology firms.
 Ecologists must establish a better understanding of how these algorithms work and fail if we are to realize this potential and avoid the pitfalls such a transition would bring.
-  To this end, we provide a conceptual and technical introduction to RL and its relevance to ecological and conservation challenges, including examples of a problem in setting fisheries quotas and in managing ecological tipping points.
-  Four appendices with annotated code provide a tangible introduction to researchers looking to adopt, evaluate, or extend these approaches.
+We ultimately set forth a research framework based on well-posed, public challenges so that ecologists and computer scientists can collaborate towards solving hard decision-making problems in conservation.
 
 keywords:
-  - Reinforcement Learning
+  - Artificial Intelligence
   - Conservation
   - Machine Learning
-  - Artificial Intelligence
+  - Reinforcement Learning
   - Tipping points
@@
   - \usepackage{setspace}
-#  - \usepackage{endfloat}
-#  - \linenumbers
-#  - \doublespacing
+  - \linenumbers
+  - \doublespacing
 
@@
 
-the number of figures, tables, and text boxes: **6** (five figures, 1 table)
+the number of figures, tables, and text boxes: **7** (5 figures, 2 tables)
 
@@
 Even the simple models we consider here contain sources of complexity such as hysteresis that are likely commonplace to ecological problems but have not been well represented in RL research to date. 
-Applying RL to a range of "precise" models is currently beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.
+Applying RL to a range of "precise" models is currently beyond the scope of any one paper: it will require not only more computational resources but more collaborations between large teams of ecologists and computer scientists.
 
-# RL overview
+# Materials and Methods
 
@@
 
+# Acknowledgements {-}
+
+All authors were supported by the NSF Faculty Early Career Development Program Award 1942280. We thank Benjamin Blonder, Zachary Sunberg, and Claire Tomlin for their thoughtful comments.
+
+# Conflicts of Interest {-}
+
+The authors declare there are no conflicts of interest.
+
+# Authors' Contributions {-}
+
+M.L. and C.B. developed the code and wrote the manuscript with support from K.E.A.N. and M.S.C.
+
+# Data Availability {-}
+
+We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures herein.
+
 \pagebreak

### 2022-05-24 manuscript/manuscript.Rmd
Adding draft of response letter

@@
 The *agent*, which is often a computer program, continuously interacts with the environment. 
-At each time step, the agent observes the current *state* of the environment then performs an *action*. 
+At each time step, the agent *observes* the current *state* of the environment then performs an *action*^[The terms observation and state are used nearly interchangeably in describing RL, so it is worth clarifying the distinction.
+An observation is the  depiction of the environment that is given to the agent at each time step, but the state is the true underlying description of the environment.
+When the term observation is used, this usually means that the observation does not provide an accurate portrayal of the environment's state.
+Yet, in cases when the observation and state are in agreement, the term observation is typically not used at all.].
 As a result of this action, the environment transitions to a new state and transmits a numerical *reward* signal to the agent.
@@
 
-An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of the environment's current *state* and an associated *reward* as output.
+An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an *observation* of the environment's current *state* and an associated *reward* as output.
 In contrast to classical approaches [@Marescot2013; @chades_primer_2021], there are few restrictions on what comprises a state or action. 
@@
 The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
-This ability allows RL environments to model a broad range of decision making problems.
+The ability to tailor the actions, states, transition dynamics and reward function allows RL environments to model a broad range of decision making problems.
 For example, we can set the transitions to be deterministic or stochastic.
+We could map any countable set of actions to a discrete action space. 
 We can also specify the reward function to be *sparse*, whereby a positive reward can only be received after a long sequence of actions, e.g. the end point in a maze.
@@
 Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
-The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
+The RL agent's use of a more dynamic strategy out-performs the steady-state strategy.
 Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
@@
 Yet, even in the case that an agent performs well on a realistic problem, there will be a range of open questions in using deep RL to inform decision-making. 
-Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the deep RL agent will generalize its past experience to new situations?
+Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the agent will generalize its past experience to new situations -- especially when we cannot readily visualize the policy?
 Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?

### 2022-05-24 manuscript/revision_response.Rmd
Adding draft of response letter

---
output: pdf_document
---

Revision Round 1 Notes, Deep RL for Conservation Decisions

Reviewer 1:

1) “I just noticed very few typos: Line 132: “forgo”, and line 318: “out-preforms”.”

We have fixed the line 318 typo.

2) As a computer scientist, I am just curious to know more about the hardware and software used for the experiments, especially the GPU used and how much time did it take to converge? At least the order of magnitude (hours or days)?

!! CC Carl !!

Reviewer 2:

1) “It would have been nice to discuss how [the 3 ML] paradigms are not mutually exclusive. […]  Do we talk about deep RL  only when the neural networks are used in the decision process (policy, action) or even when NNs are used in the transition distribution ?”

It is possible to use techniques from supervised learning and unsupervised learning in RL. Two examples that come to mind are behavior cloning and “CURL: Contrastive Unsupervised Representations for Reinforcement Learning” by Srinivas et al. which apply supervised and unsupervised learning techniques respectively. Reinforcement learning is a sprawling discipline and since this is an introductory piece in which we don’t want to overwhelm the reader, we think it is better to focus on more commonly encountered subjects in the field.

It is possible to use neural networks to model the transition dynamics whether that is in the agent or in the simulator. The first case would be an instance of model-based RL, where the agent is modeling the environment; this approach is quite common. We do not mention that neural networks can be used to approximate the transition dynamics explicitly in the manuscript, but implicitly: “Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators.”

2) You also state that deep RL is not adopted in ecology. How about closely related fields such as evolution, genetics or biocontrol (see Treloar et al 2018) research ? Did you investigate that?

We’ve updated this statement, there are a handful applications of deep RL in ecology. !!Need to do this!!

3) I didn't get the usefulness of adding the part about precise/tactical and strategic/general in the introduction, especially since this isn't elaborated in the discussion.

**CC Carl**

4) I think the distinction state/observation merits a bit more clarification.

We added a footnote to address this problem:

“The terms observation and state are used nearly interchangeably in describing RL, so it is worth clarifying the distinction. An observation is the  depiction of the environment that is given to the agent at each time step, but the state is the true underlying description of the environment. When the term observation is used, this usually means that the observation does not provide an accurate portrayal of the environment's state. Yet, in cases when the observation and state are in agreement, the term observation is typically not used at all.”

5) How you would go about observing the policies learnt by the model in the case of large models ? For instance, if the state space is very large. Surely, a decision-maker would not blindly trust a black-box model to provide the optimal decision without any provided logic, the way you show it in your experiments in Fig[2-5].  Are current machine learning interpretability tools also useful in deep RL settings?

We have put a lot of thought into the topics of evaluating large ML models and ML. We believe the difficulty to evaluate and interpret ML models will be a significant obstacle to RL’s adoption. For large observation-action spaces, there are no available tools that will let you interpret the model, so practitioners must carry out extensive testing to gain confidence in the agent; but, extensive testing for non-tabular observation-action space is still no guarantee that the agent won’t commit undesirable actions. We provided a slight update to a sentence to address these points:

“Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the agent will generalize its past experience to new situations -- especially when we cannot readily visualize the policy?”

6) I would appreciate an explanation of the discount factor. What does it represent in practice ? Why is there a discount that depends on the state t? Where does it fit in the second experiment ?How to choose the Horizon H. Is it also an hyper-parameter just like epochs in gradient descent ?

!!Return to this!!

7) In general, do you have recommendations to identify under-explored scenarios?

 Identifying under-explored spaces of observation-action is going to be very dependent on the problem. As the observation-action spaces become large, identifying under-explored regions becomes particularly tough. When the observation-action space can easily visualized, then one can use their domain knowledge to improve exploration by say changing the exploration algorithm or tuning the hyperparameters relevant to exploration.

8) It would be great to add the optimal decision also in the Argentine Hake fishery to the figure. Was the optimal decision used at some point in that time series ?

!!CC Carl!!

9) But what if you had a good understanding of the system, how would knowing the model structure leveraging the theoretical knowledge of the ecosystems facilitate or complicate the RL problem? Specifically, in regards to the challenges in terms of computation cost you raise later.

!!We Probably want to make some edits on this point!!

There are two primary ways that one could use domain knowledge to assist the agent in the learning process. One approach is to use a model-based agent where you would instill domain knowledge into the agent’s model of the environment directly. For instance, instead of fitting a neural network to the the transition dynamics, the agent could estimate parameters for a pre-specified model. The benefit of this approach is that the agent could converge to an optimal solution more quickly as the policy search space would be much smaller — note that this benefit would only be true with on-policy algorithms. Another method is to design an exploration algorithm based off of domain knowledge so the agent can more efficiently find good solutions. For both cases, the primary benefit is the same: using domain knowledge can lead to more efficient exploration. Yet, if the domain knowledge is biased, then it is very likely that the agent will learn suboptimal solutions. For instance, with model-based exploration, the agent might spend too much time exploring a low reward region of the observation-action space.

The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. Certainly, there will be occasions when model-based approaches will outperform model-free methods, and the most significant use case of model-based algorithms is for problems that are computationally expensive.

Reviewer 3:

1) I strongly recommend that the authors rewrite portions of the manuscript to make it accessible to a broader audience unfamiliar with deep learning techniques.

We acknowledge that our paper might be difficult to grasp for the audience of MEE, but we want to underscore that deep RL is difficult to understand in general, even for people that are familiar with the other subfields of machine learning. There are a lot of concepts and terminology that are unique to RL that will likely require newcomers to do lots of re-reading and digging through the references to gain a comprehensive understanding. This being said, we believe we’ve made subject matter as accessible as possible by explaining the fundamentals in the sections “RL Environments”  and “Deep RL Agents” without any hard-to-parse equations and minimal jargon. In the appendices, we go a bit more in-depth on some theoretical and practical topics, which will hopefully guide readers that are motivated to dig deeper to find a better understanding of the subject. Lastly, on this area of concern, we have addressed some specific areas of confusion raised by other reviewers.

2) While RL is based on mathematical and probabilistic outcomes proposing actions to maximize profit, I wonder whether this shall provide feasible outcomes as the actions may only be mathematically attainable and not ecologically. Suppose this is being done to reduce human interventions at decision making, to identify and obtain viable strategies, one has to investigate the outputs of RL strategies. I suggest the authors consider discussing this briefly in the manuscript.

We interpret this comment to be partially asking the question: “What are the limitations on an action space?” We have changed some of the language in “RL Environments” to better address this question:

“The ability to tailor the actions, states, transition dynamics and reward function allows RL environments to model a broad range of decision making problems.
For example, we can set the transitions to be deterministic or stochastic.
We could map any countable set of actions to a discrete action space.”

In other words, the action space can be tailored to fit virtually any decision-making problem. The other question that we interpret this comment to be presenting is whether we can describe all ecological processes mathematically. This is a fundamental philosophy of science question, and is outside of the scope of this paper.

3) The authors should provide a more vivid explanation of the training data: especially the action space and the reward signals for both the fishing management and the ecological tipping point problem.

We provide an explanation of the training data for both problems. For the fishing problem:

Lines 216-218: “The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.”

And for the tipping point problem we give a full detailed explanation of the problem from lines 265 to 280.

4) This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes. Do the authors mean that RL provides output as lower than the actual optimal. If so, how realistic is it? I want to learn from the authors how RL can be used to obtain readily available conservation solutions.

For this question, it’s best to take a look at Figure 2 to see how closely the RL policy approximates the optimal policy for the fishing example. The agent seems to learn the general functional form of the policy but is slightly biased. We will address the re-occurring concern of whether RL is ready to be used for real problems in another comment.

5) While preventing tipping points is a major goal of modern interdisciplinary science, conservation policy using RL only in one instance is not enough to comment on its applicability. The authors should validate the method by testing on a model of different origins with a different non-linearity and available real data exhibiting a critical transition. (As saddle-node bifurcation to all such instances, the trained RL model should work well when tested on such data). This shall be instrumental in commenting on the generality of the approach for conserving tippings. 
The authors should justify the robustness of their RL method at conserving tipping in real empirical data. While such data may not be available, noisy simulated data with imperfect sampling mimicking real scenarios can be considered. Real data are often irregularly sampled with errors in reported values (quite obvious in climate data and maybe a concern while reducing emissions to prevent climate tippings ). The authors may perform an experiment and discuss the details in the discussion section as to how RL will deal with such instances.

We share this perspective with the reviewer that our study does not prove that deep RL can be immediately applied to conservation problems. This is not the intention of our paper. Instead, we intend our paper to be an introduction for ecologists to deep RL, whereby we introduce the OpenAI Gym framework, some terminology and example problems, so that researchers can investigate deep RL on their own and also better understand how it is beginning to be applied to ecology already. We are at a point in time where researchers and organizations are using deep RL in conservation contexts, so it is imperative for our community to start understanding the benefits and downfalls of using deep RL for our specialized class of problems.

!!edit intro to to voice this intention!!

6) In Appendix C, the authors mentioned that scaling the instance space to [-1 1] improves the agent performance of RL. Ecological events are highly non-linear sequences of events; such data are often non-stationary. Will scaling the state space in RL not incorporate bias and still lead to robust predictions. I think it demands clarifications.

Rescaling does not affect the dynamics. This is equivalent to measuring the process in different units.

### 2022-05-27 manuscript/manuscript.Rmd
Some edits, more to come

@@
 
-Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological conservation, RL has not been widely applied across this field.
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological conservation, RL has only been applied to this field in a few cases [@xu_robust_2021, @silvestro_improving_2022].
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
@@
 cannot be represented in a computationally tractable table.], which happens frequently with realistic decision-making problems.
-Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+Whenever there is a need for an agent to approximate some function, typically to model the policy and/or the transition dynamics, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
 Although there are other function approximators that can be used in RL, e.g.

### 2022-05-27 manuscript/revision_response.Rmd
Some edits, more to come

@@
 Two examples that come to mind are behavior cloning and “CURL: Contrastive Unsupervised Representations for Reinforcement Learning” by Srinivas et al.
-which apply supervised and unsupervised learning techniques respectively.
-Reinforcement learning is a sprawling discipline and since this is an introductory piece in which we don’t want to overwhelm the reader, we think it is better to focus on more commonly encountered subjects in the field.
+Reinforcement learning is a sprawling discipline and since this is an introductory piece where we don’t want to overwhelm the reader, we think it is better to focus on more commonly encountered subjects in the field.
 
-It is possible to use neural networks to model the transition dynamics whether that is in the agent or in the simulator.
-The first case would be an instance of model-based RL, where the agent is modeling the environment; this approach is quite common.
-We do not mention that neural networks can be used to approximate the transition dynamics explicitly in the manuscript, but implicitly: “Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators.”
+This is a good point that we have attempted to clarify in the manuscript:
+
+"Whenever there is a need for an agent to approximate some function, typically to model the policy and/or the transition dynamics, neural networks can be used in this capacity due to their property of being general function approximators."
 
@@
 
-We’ve updated this statement, there are a handful applications of deep RL in ecology. !!Need to do this!!
+We’ve updated this statement as we have found a few noteworthy applications in ecology. Cite !!line!!.
 

### 2022-05-28 manuscript/revision_response.Rmd
Small updates

@@
 Two examples that come to mind are behavior cloning and “CURL: Contrastive Unsupervised Representations for Reinforcement Learning” by Srinivas et al.
-Reinforcement learning is a sprawling discipline and since this is an introductory piece where we don’t want to overwhelm the reader, we think it is better to focus on more commonly encountered subjects in the field.
+Reinforcement learning is a sprawling discipline, and since this is an introductory piece, we don’t want to overwhelm the reader with the multitude of available approaches.
+We have tried to focus on the more commonly encountered subjects in the field.
 
-This is a good point that we have attempted to clarify in the manuscript:
+How neural networks can be used is a point that we have attempted to clarify in the manuscript:
 
@@
 
-We’ve updated this statement as we have found a few noteworthy applications in ecology. Cite !!line!!.
+We’ve updated this statement as we have found a few note-worthy applications in ecology. Cite !!line!!.
 
@@
 
-We have put a lot of thought into the topics of evaluating large ML models and ML.
-We believe the difficulty to evaluate and interpret ML models will be a significant obstacle to RL’s adoption.
-For large observation-action spaces, there are no available tools that will let you interpret the model, so practitioners must carry out extensive testing to gain confidence in the agent; but, extensive testing for non-tabular observation-action space is still no guarantee that the agent won’t commit undesirable actions.
-We provided a slight update to a sentence to address these points:
+The difficulty around evaluating and interpreting ML models will be a significant obstacle to RL’s adoption on safety-critical problems.
+For large observation-action spaces, there are no general tools that will let you interpret the model, so practitioners must carry out extensive testing to gain confidence in the agent; but, extensive testing for large (aka non-tabular) observation-action spaces will not guarantee that the agent won’t commit undesirable actions on untested situations.
+We provided a small update to address these points:
 
@@
 
-!!Return to this!!
+We have updated our description of the discount factor in the manuscript to address these questions:
+
+"a discount factor which describes how much the agent will value rewards to be received in the distant future versus the immediate future"
+
+The exponentiation based on time, t, follows from this description.
+Since the discount factor is defined on the range (0, 1), the reward will become smaller over the episode.
+In our problems, we evaluate with a discount factor of 1.
+It is common however to train the agent with a different discount factor than is used during evaluation, as it may aid exploration to use a different discount factor; thus it is treated like a hyperparameter.
+The horizon is another phrase for the length of an episode.
+It is a parameter of the POMDP and not a hyperparameter.
 
@@
 
- Identifying under-explored spaces of observation-action is going to be very dependent on the problem.
-As the observation-action spaces become large, identifying under-explored regions becomes particularly tough.
-When the observation-action space can easily visualized, then one can use their domain knowledge to improve exploration by say changing the exploration algorithm or tuning the hyperparameters relevant to exploration.
+Identifying under-explored spaces of the observation-action space is going to be very dependent on the problem.
+Usually RL practitioners identify under-explored scenarios by either doing extensive testing or visualizing the policy.
+As the observation-action spaces become large, identifying under-explored regions becomes particularly tough, since it becomes impossible to visualize the policy.
+When the observation-action space can be easily visualized, then one can rely on their domain knowledge to recognize sub-optimal performance, which is what we have done in both our examples.
+So the main recommendation is to first try to visualize the policy then second perform extensive testing to identify areas of poor explorations.
 
@@
 
-!!We Probably want to make some edits on this point!!
-
 There are two primary ways that one could use domain knowledge to assist the agent in the learning process.
-One approach is to use a model-based agent where you would instill domain knowledge into the agent’s model of the environment directly.
-For instance, instead of fitting a neural network to the the transition dynamics, the agent could estimate parameters for a pre-specified model.
-The benefit of this approach is that the agent could converge to an optimal solution more quickly as the policy search space would be much smaller — note that this benefit would only be true with on-policy algorithms.
+One approach is to use a model-based agent where the RL practitioner would instill domain knowledge into the agent’s model of the environment directly.
+For instance, instead of fitting a randomized neural network to the the transition dynamics, the agent could estimate parameters for a pre-specified model.
+The benefit of this approach is that the agent could converge to an optimal solution more quickly as the policy search space could be made much smaller for restrictive models — note that this benefit could only be realized with on-policy algorithms, algorithms that explore according to the same policy that they are updating.
 Another method is to design an exploration algorithm based off of domain knowledge so the agent can more efficiently find good solutions.
@@
 
-The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs.
-Certainly, there will be occasions when model-based approaches will outperform model-free methods, and the most significant use case of model-based algorithms is for problems that are computationally expensive.
+The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to but did not unpack,
+
+"Fully detailed conservation decision-making problems will likely require comparable specialized algorithms and hardware that ecologists do not generally have access to."
 
@@
 The other question that we interpret this comment to be presenting is whether we can describe all ecological processes mathematically.
-This is a fundamental philosophy of science question, and is outside of the scope of this paper.
+This is a fundamental philosophy of science question that is outside of the scope of this paper.
 
@@
 We share this perspective with the reviewer that our study does not prove that deep RL can be immediately applied to conservation problems.
-This is not the intention of our paper.
-Instead, we intend our paper to be an introduction for ecologists to deep RL, whereby we introduce the OpenAI Gym framework, some terminology and example problems, so that researchers can investigate deep RL on their own and also better understand how it is beginning to be applied to ecology already.
+Validating RL as method for ecological problems is not the intention of our paper.
+Instead, we have written our paper to be an introduction for ecologists to deep RL, whereby we introduce the OpenAI Gym framework, some fundamental RL concepts and example problems, so that researchers can investigate deep RL on their own and better understand how it is beginning to be applied to ecology.
 We are at a point in time where researchers and organizations are using deep RL in conservation contexts, so it is imperative for our community to start understanding the benefits and downfalls of using deep RL for our specialized class of problems.
@@
 
-Rescaling does not affect the dynamics. This is equivalent to measuring the process in different units.
+Re-scaling the state space is equivalent to measuring the process in different units, so this does not affect the transition dynamics.

### 2022-05-30 manuscript/manuscript.Rmd
Small updates
Another re-read

--- 2022-05-24 manuscript/manuscript.Rmd
@@
 
-Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological conservation, RL has not been widely applied across this field.
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological conservation, RL has only been applied to this field in a few cases [@xu_robust_2021; @silvestro_improving_2022].
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
@@
 While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research.
-We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though the application of RL to conservation will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
+We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms while keeping in mind that the application of RL to conservation will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.
 
@@
 To optimize the RL objective, agents either take a *model-free* or *model-based* approach. 
-The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
+The distinction is that *model-free* algorithms do not attempt to learn or use a predictive model of the environment; yet, *model-based* algorithms employ a predictive model of the environment to achieve the RL objective.
 A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo; @suttonbarto].
@@
 cannot be represented in a computationally tractable table.], which happens frequently with realistic decision-making problems.
-Whenever there is a need for an agent to approximate some function, say a policy function, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
+Whenever there is a need for an agent to approximate some function, typically a function to represent the policy and/or to model the transition dynamics, neural networks can be used in this capacity due to their property of being general function approximators [@universalapproximators].
 Although there are other function approximators that can be used in RL, e.g.
@@
 
-- $\gamma \in (0,1]$: a discount factor
+- $\gamma \in (0,1]$: a discount factor which describes how much the agent will value rewards to be received in the distant future versus the immediate future
 
@@
 
-For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1973] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
+For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1973 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
 The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
@@
 Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
-Training these agents in a variety of alternative contexts can improve their ability to generalize to other scenarios.
+Training these agents in a variety of alternative contexts could improve their ability to generalize to other scenarios.
 
@@
 This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
-On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
+On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of a more desirable 'high' state relative to the 'low' state.
 For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
@@
 While the precise rate of such decline may not be known in practice (and will not be known to RL algorithms before-hand either), it is easy to implement such a policy in simulation for comparative purposes.
-We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm.
+We compare this rule-of-thumb to a policy found by training an agent using the TD3 algorithm.
 
@@
 As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards.
-In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
+In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the odds of a chance transition.
 In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
@@
 <!-- Ecology has complex decisions and we lack necessary tools; RL could be it -->
-Ecological challenges facing the planet today are complex and outcomes are both uncertain and consequential.
+Ecological challenges facing the planet today are complex, and their outcomes are both uncertain and consequential.
 Even our best models and best research will never provide a crystal ball to the future, only better elucidate possible scenarios.
@@
 @Schaefer1954; @Walters1978] and remains an active area of research [@Possingham2006; @Fischer2009; @Polasky2011].
-However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure or other sources of complexity that are both pervasive and influential on ecological dynamics [@Hastings2012].
+However, the limitations of classical methods such as optimal control frequently constrain applications to relatively simplified models [@Possingham2006], ignoring elements such as spatial or temporal heterogeneity and autocorrelation, stochasticity, imperfect observations, age or state structure, and other sources of complexity that are both pervasive and influential on ecological dynamics [@Hastings2012].
 Complexity comes not only from the ecological processes but also the available actions.
@@
 This paper does not set the precedent as the first application of RL to ecology.
-There have been a number of studies applying RL to behavioral ecology, typically with multi-agent environments [@wang_reinforcement_2020, @frankenhuis_enriching_2019, @perolat_multi-agent_2017].
+There have been a number of studies applying RL to behavioral ecology, typically with multi-agent environments [@wang_reinforcement_2020; @frankenhuis_enriching_2019; @perolat_multi-agent_2017].
 Yet, it is important to distinguish the aim of these behavioral studies from the aim of applying RL to conservation management.
@@
 -->
-In previous behavioral ecology studies, RL algorithms as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
+In previous behavioral ecology studies, RL algorithms as a substitute for animal learning mechanisms [@wang_reinforcement_2020; @perolat_multi-agent_2017].
 When applying deep RL to conservation management, we do not make the assumption that an RL algorithm learns analogously to how an animal learns.
@@
 It is also worth acknowledging that deep RL algorithms, particularly model-free algorithms, have poor sample efficiency, which could limit deep RL from being effective on environments that are slow to run [@sac].
-Thus, as Section 5.1 and 5.2 illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
+Thus, as Sections 5.1 and 5.2 illustrate, it is important to begin with simple problems, including those for which an optimal strategy is already known.
 Such examples provide important benchmarks to calibrate the performance, tuning and training requirements of RL.
@@
 Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?
-A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021, @chapman_promoting_2021].
+A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021; @chapman_promoting_2021].
 

### 2022-05-30 manuscript/revision_response.Rmd
Another re-read

@@
 Reinforcement learning is a sprawling discipline, and since this is an introductory piece, we don’t want to overwhelm the reader with the multitude of available approaches.
-We have tried to focus on the more commonly encountered subjects in the field.
+We have omitted a number of methods to focus on the more commonly encountered subjects in the field.
 
-How neural networks can be used is a point that we have attempted to clarify in the manuscript:
+Regarding how neural networks can be used, we have attempted to clarify this in the manuscript:
 
@@
 When the term observation is used, this usually means that the observation does not provide an accurate portrayal of the environment's state.
-Yet, in cases when the observation and state are in agreement, the term observation is typically not used at all.”
+In cases when the observation and state are in agreement, the term observation is typically not used at all.”
 
@@
 For large observation-action spaces, there are no general tools that will let you interpret the model, so practitioners must carry out extensive testing to gain confidence in the agent; but, extensive testing for large (aka non-tabular) observation-action spaces will not guarantee that the agent won’t commit undesirable actions on untested situations.
-We provided a small update to address these points:
+We provided a small update to address these points which are open research questions:
 
@@
 Since the discount factor is defined on the range (0, 1), the reward will become smaller over the episode.
-In our problems, we evaluate with a discount factor of 1.
-It is common however to train the agent with a different discount factor than is used during evaluation, as it may aid exploration to use a different discount factor; thus it is treated like a hyperparameter.
+For evaluation, rewards are attributed a discount factor of 1.
+It is common however to train the agent with a different discount factor than is used during evaluation, as it may aid exploration to use a different discount factor; thus it can be treated like a hyperparameter.
 The horizon is another phrase for the length of an episode.
@@
 When the observation-action space can be easily visualized, then one can rely on their domain knowledge to recognize sub-optimal performance, which is what we have done in both our examples.
-So the main recommendation is to first try to visualize the policy then second perform extensive testing to identify areas of poor explorations.
+So the main recommendation is to first try to visualize the policy then perform extensive testing to identify areas of poor exploration.
 
@@
 For instance, instead of fitting a randomized neural network to the the transition dynamics, the agent could estimate parameters for a pre-specified model.
-The benefit of this approach is that the agent could converge to an optimal solution more quickly as the policy search space could be made much smaller for restrictive models — note that this benefit could only be realized with on-policy algorithms, algorithms that explore according to the same policy that they are updating.
+The benefit of this approach is that the agent could converge to an optimal solution more quickly as the policy search space could be made much smaller for restrictive models — note that this benefit can only be realized in certain scenarios like with RL algorithms that explore according to the same policy that they are updating.
 Another method is to design an exploration algorithm based off of domain knowledge so the agent can more efficiently find good solutions.
@@
 Yet, if the domain knowledge is biased, then it is very likely that the agent will learn suboptimal solutions.
-For instance, with model-based exploration, the agent might spend too much time exploring a low reward region of the observation-action space.
 
-The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to but did not unpack,
+The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to here,
 
@@
 This being said, we believe we’ve made subject matter as accessible as possible by explaining the fundamentals in the sections “RL Environments”  and “Deep RL Agents” without any hard-to-parse equations and minimal jargon.
-In the appendices, we go a bit more in-depth on some theoretical and practical topics, which will hopefully guide readers that are motivated to dig deeper to find a better understanding of the subject.
+In the appendices, we go a bit more in-depth on some theoretical and practical topics, which will hopefully guide readers that are motivated to dig deeper to reach a better understanding of the subject.
 Lastly, on this area of concern, we have addressed some specific areas of confusion raised by other reviewers.
@@
 
-!!edit intro to to voice this intention!!
+!!edit intro to to voice this intention CC Carl!!
 

### 2022-05-31 manuscript/manuscript.Rmd
Validation paragraph first go

@@
 
-We do recognize the need for more complex, "precise" models to derive quantitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts.
-The examples here should provide ecologists with a greater background to engage in the collaborative development of such methods, while also highlighting challenges which ecological problems pose to existing techniques.
-Even the simple models we consider here contain sources of complexity such as hysteresis that are likely commonplace to ecological problems but have not been well represented in RL research to date. 
-Applying RL to a range of "precise" models is currently beyond the scope of any one paper: it will require not only more computational resources but more collaborations between large teams of ecologists and computer scientists.
+This paper does not intend to validate deep RL as a method that should be used to directly inform decision-making on current conservation problems.
+Instead, we have selected example problems to provide ecologists with a greater background to engage in the collaborative development of deep RL-based methods, while also highlighting challenges which ecological problems pose to existing techniques.
+Validating deep RL for current conservation problems is beyond the scope of any one paper: this will necessitate examining a range of more "precise" models which will require more computational resources than is available to most researchers and extensive collaboration between large teams of ecologists and computer scientists.
 

### 2022-05-31 manuscript/manuscript.Rmd
clean up 'tactical'

@@
 Levins espoused simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
-More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general).
-Such simple, realistic and general models are still the bedrock of most theory and practice today, even when implementation turns to tactical models for quantitative guidance (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or $R_0$ in epidemiology, remain important concepts in management even though they do not exist in typical tactical models).
+More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_. 
+Such simple, realistic and general models are still the bedrock of most theory and practice today (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or $R_0$ in epidemiology, remain important concepts in management).
 These models provide an ideal first benchmark for evaluating the performance of emerging methods of deep RL for several reasons:

### 2022-05-31 manuscript/revision_response.Rmd
Revisions after CB's pass over

@@
 
-We have fixed the line 318 typo.
+We have fixed the line 318 typo. Use of "forgo" on line 132 was intentional.
 
@@
 
-!! CC Carl !!
+All algorithms were run on an NVIDIA Quadro RTX 8000 GPU.
+The training  budget for the fishing scenario was 300K timesteps (3K runs, taking about 1 hr).
+Tipping point training budget was 3M timesteps (6K runs, taking several hours).
+Precise run times vary with machine load and algorithm.
+Software details and hyperparameters are provided in the associated GitHub repo, https://github.com/boettiger-lab/rl-intro/.
+Hyperparameter tuning typically required 100s of training runs using both Optuna and manual adjustments.
 
@@
 
-**CC Carl**
+We agree the writing was not clear, we have sought to simplify this in the intro, sticking with Levins 1966 famous characterization.
+The essence of Levins' classic argument is merely that process-based models with only a handful of parameters are often more general than those with scores of parameters.
+We return to this point, though not this language, in the Discussion addressing how real-world simulations will be more complex then the benchmarks we have chosen for this intro.
+We avoid the language of tactical (i.e.
+'use this precise policy or algorithm on this precise problem') vs strategic (i.e.
+;evaluate the use RL as a strategy in conservation problems') as relatively unhelpful here.
+
+"Levins espoused simple mechanistic models which satisfy the goals of being both realistic and general.
+More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice generality for precision.
+Such simple, realistic and general models are still the bedrock of most theory and practice today (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or R_0 in epidemiology, remain important concepts in management)."
 
@@
 
-We acknowledge that our paper might be difficult to grasp for the audience of MEE, but we want to underscore that deep RL is difficult to understand in general, even for people that are familiar with the other subfields of machine learning.
-There are a lot of concepts and terminology that are unique to RL that will likely require newcomers to do lots of re-reading and digging through the references to gain a comprehensive understanding.
-This being said, we believe we’ve made subject matter as accessible as possible by explaining the fundamentals in the sections “RL Environments”  and “Deep RL Agents” without any hard-to-parse equations and minimal jargon.
-In the appendices, we go a bit more in-depth on some theoretical and practical topics, which will hopefully guide readers that are motivated to dig deeper to reach a better understanding of the subject.
-Lastly, on this area of concern, we have addressed some specific areas of confusion raised by other reviewers.
+The reviewer raises an excellent point about the importance of communicating these technical methods to a broad audience and we have sought to make our revision more accessible to a broad audience.
+We discuss changes to specific issues highlighted by the reviewer below, as well as those locations where other reviewers have noted that certain jargon was unclear.
+We have also gone over the manuscript and all four appendices to ensure they provide an introduction that is both accessible with hands-on executable code, while also being technically precise enough for readers to have an appropriate understanding of how the methods work.
 
@@
 
-We interpret this comment to be partially asking the question: “What are the limitations on an action space?” We have changed some of the language in “RL Environments” to better address this question:
+The reviewer raises two important issues here: "What is the appropriate reward function", and "What are the appropriate limitations on the action space".
+These questions have always been central issues to ecological management and decision-making, and are neither unique to or resolved by the use RL [@gregory_structured_2012; @conroy_decision_2013].
+We have changed some of the language in “RL Environments” to better address the action space question:
 
@@
 
-In other words, the action space can be tailored to fit virtually any decision-making problem.
-The other question that we interpret this comment to be presenting is whether we can describe all ecological processes mathematically.
-This is a fundamental philosophy of science question that is outside of the scope of this paper.
+And we address the difficulty in selecting a reward function in the discussion:
+
+"Given that there have been many examples of reward misspecification leading to undesirable behavior, what if we have selected an objective that unexpectedly causes damaging behavior?"
 
@@
 
-Re-scaling the state space is equivalent to measuring the process in different units, so this does not affect the transition dynamics.
+We have added a discussion of scaling in the appendix.

### 2022-05-31 manuscript/manuscript.Rmd

@@
 
-- $\gamma \in (0,1]$: a discount factor which describes how much the agent will value rewards to be received in the distant future versus the immediate future
+- $\gamma \in (0,1]$: a discount factor which describes how much the agent will value rewards to be received in the distant future versus the immediate future [@clark_mathematical_2010]
 
@@
 Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the agent will generalize its past experience to new situations -- especially when we cannot readily visualize the policy?
+To gain such confidence, it will be necessary to do extensive testing on previously unseen contexts [@kazak_verifying_2019], but even then, it can be difficult to verify that the agent will perform as expected.
 Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?

### 2022-05-31 manuscript/revision_response.Rmd
Missed some comments

@@
 
-Identifying under-explored spaces of the observation-action space is going to be very dependent on the problem.
-Usually RL practitioners identify under-explored scenarios by either doing extensive testing or visualizing the policy.
-As the observation-action spaces become large, identifying under-explored regions becomes particularly tough, since it becomes impossible to visualize the policy.
-When the observation-action space can be easily visualized, then one can rely on their domain knowledge to recognize sub-optimal performance, which is what we have done in both our examples.
-So the main recommendation is to first try to visualize the policy then perform extensive testing to identify areas of poor exploration.
+This is a great question that we have tried to address in the latest revision,
+
+"Because these simulations rarely reach stock sizes at or above carrying capacity, i.e.
+larger stock sizes are under-explored, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
+This observation brings up a point that is well worth discussing which is how to best identify and resolve under-explored scenarios.
+Usually, RL practitioners identify under-explored scenarios by either doing extensive testing or visualizing the policy, then tweaking the hyperparameters relevant to exploration in hopes of improving the result."
 
@@
 
-The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to here,
+The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to in line xxx,
 

### 2022-05-31 appendices/appendix-A-definitions.Rmd
typo
More edits

@@
 The aim here is not to give full treatment of all deep RL methods; instead, this section is meant to serve as an abbreviated background view on model-free deep RL^[We focus on model-free methods because the agents we use in the manuscript and Appendix B are all model-free.
-Further, model-based algorithms generalize from the methods we discuss here except model-based agents attempt to learn or use the environment's state-transition operator, which is typically used for planning or supplementary sample generation [@suttonbarto].].
+Further, model-based algorithms generalize from the methods we discuss here except model-based agents attempt to learn or use the environment's state-transition operator, which is typically used for planning or supplementary sample generation [@suttonbarto].
+The main advantage of model-based algorithms tend to be more sample efficient than model-free algorithms but can struggle to perform well when the transition dynamics are very difficult to learn.].
 We will introduce some of the important design choices that are made for model-free deep RL algorithms and will point to specific resources that give a more complete presentation.
@@
 IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)           & Off-policy & Actor-critic\\
-\hlin
+\hline
 \end{tabular}}

### 2022-05-31 manuscript/revision_response.Rmd
Updating time estimates

@@
 All algorithms were run on an NVIDIA Quadro RTX 8000 GPU.
-The training  budget for the fishing scenario was 300K timesteps (3K runs, taking about 1 hr).
-Tipping point training budget was 3M timesteps (6K runs, taking several hours).
+The training  budget for the fishing scenario was 300K timesteps (3K runs, taking about 25 minutes).
+Tipping point training budget was 3M timesteps (6K runs, taking around 3 hours).
 Precise run times vary with machine load and algorithm.

### 2022-06-01 manuscript/manuscript.Rmd

@@
 RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
-Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
-Training these agents in a variety of alternative contexts could improve their ability to generalize to other scenarios.
+Because these simulations rarely reach stock sizes at or above carrying capacity, i.e.
+larger stock sizes are under-explored, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
+This observation brings up a point that is well worth discussing which is how to best identify and resolve under-explored scenarios.
+Usually, RL practitioners identify under-explored scenarios by either doing extensive testing or visualizing the policy, then tweaking the hyperparameters relevant to exploration in hopes of improving the result.
 

### 2022-06-01 manuscript/manuscript.Rmd
Adding training details in results

@@
 Annotated code for these examples may be found in Appendix B and at <https://github.com/boettiger-lab/rl-intro>.
+All algorithms were run on an NVIDIA Quadro RTX 8000 GPU.
+The training  budget for the fishing scenario was 300K timesteps (3K runs, taking about 25 minutes).
+Tipping point training budget was 3M timesteps (6K runs, taking around 3 hours).
+Precise run times vary with machine load and algorithm.
+Software details and hyperparameters are provided in the associated GitHub repo.
+Hyperparameter tuning typically required 100s of training runs using both Optuna, a python-based hyperparameter optimization library, and manual adjustments.
 

### 2022-06-01 manuscript/revision_response.Rmd
Adding training details in results
Making sure line nos are correct

@@
 Hyperparameter tuning typically required 100s of training runs using both Optuna and manual adjustments.
+We have added these details to the manuscript in lines 199-206.
 
@@
 
-Regarding how neural networks can be used, we have attempted to clarify this in the manuscript:
+Regarding how neural networks can be used, we have attempted to clarify this in the manuscript in lines 150-153:
 
@@
 
-We’ve updated this statement as we have found a few note-worthy applications in ecology. Cite !!line!!.
+We’ve updated this statement in the introduction (lines 42-43) as we have found a few note-worthy applications in ecology from @xu_robust_2021 and @silvestro_improving_2022.
 
@@
 ;evaluate the use RL as a strategy in conservation problems') as relatively unhelpful here.
+We updated lines 79-84 as follows:
 
@@
 For large observation-action spaces, there are no general tools that will let you interpret the model, so practitioners must carry out extensive testing to gain confidence in the agent; but, extensive testing for large (aka non-tabular) observation-action spaces will not guarantee that the agent won’t commit undesirable actions on untested situations.
-We provided a small update to address these points which are open research questions:
+We provided a small update to address these points in lines 395-399:
 
-“Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the agent will generalize its past experience to new situations -- especially when we cannot readily visualize the policy?”
+“Since deep neural networks lack transparency [@castelvecchi_can_2016], can we be confident that the agent will generalize its past experience to new situations -- especially when we cannot readily visualize the policy?
+To gain such confidence, it will be necessary to do extensive testing on previously unseen contexts, but even then, it can be difficult to verify that the agent will perform as expected.”
 
@@
 
-We have updated our description of the discount factor in the manuscript to address these questions:
+We have updated our description of the discount factor in the manuscript to address these questions in lines 186-187:
 
@@
 
-This is a great question that we have tried to address in the latest revision,
+This is a great question that we have tried to address in the latest revision in lines 244-250:
 
@@
 
-The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to in line xxx,
+The reviewer here raises a very good point here about employing domain knowledge to reduce computational costs. This is something that we alluded to in lines 409-411:
 
@@
 The reviewer raises two important issues here: "What is the appropriate reward function", and "What are the appropriate limitations on the action space".
-These questions have always been central issues to ecological management and decision-making, and are neither unique to or resolved by the use RL [@gregory_structured_2012; @conroy_decision_2013].
-We have changed some of the language in “RL Environments” to better address the action space question:
+These questions have always been central issues to ecological management and decision-making, and are neither unique to or resolved by RL [@gregory_structured_2012; @conroy_decision_2013].
+We have changed some of our language in lines 126-128 to better address the action space question:
 
@@
 
-And we address the difficulty in selecting a reward function in the discussion:
+And we address the difficulty in selecting a reward function in the discussion in lines 399-401:
 
@@
 
-Lines 216-218: “The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.”
+Lines 221-223: “The manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.”
 
-And for the tipping point problem we give a full detailed explanation of the problem from lines 265 to 280.
+And for the tipping point problem we give a full detailed explanation of the problem from lines 269 to 285.
 
@@
 We are at a point in time where researchers and organizations are using deep RL in conservation contexts, so it is imperative for our community to start understanding the benefits and downfalls of using deep RL for our specialized class of problems.
-
-!!edit intro to to voice this intention CC Carl!!
+We have edited the introduction in lines 99-105 to address this comment.
 
@@
 
-We have added a discussion of scaling in the appendix.
+We have added a discussion of scaling in Appendix C.

### 2022-06-02 manuscript/manuscript.Rmd

--- 2022-06-01 manuscript/manuscript.Rmd
@@
 Annotated code for these examples may be found in Appendix B and at <https://github.com/boettiger-lab/rl-intro>.
+All algorithms were run on an NVIDIA Quadro RTX 8000 GPU.
+The training  budget for the fishing scenario was 300K timesteps (3K runs, taking about 25 minutes).
+The training budget for the tipping point example was 3M timesteps (6K runs, taking around 3 hours).
+Software details and hyperparameters are provided in the associated GitHub repo.
+Hyperparameter tuning typically required 100s of training runs using both Optuna, a python-based hyperparameter optimization module, and manual adjustments.
 

### 2022-06-03 manuscript/manuscript.Rmd
Including some refs from reviewers

@@
 
-Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological conservation, RL has only been applied to this field in a few cases [@xu_robust_2021; @silvestro_improving_2022].
+Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological control, RL has only been applied to this field in a few cases [@xu_robust_2021; @silvestro_improving_2022; @treloar_deep_2020].
 To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
@@
 Given that there have been many examples of reward misspecification leading to undesirable behavior [@hadfield-menell_inverse_2020], what if we have selected an objective that unexpectedly causes damaging behavior?
+Reward misspecification is not unique to RL and has long been a central problem in ecological management and decision-making [@gregory_structured_2012; @conroy_decision_2013], but it is important to make clear that RL does not resolve this issue.
 A greater role of algorithms in conservation decision-making also raises questions about ethics and power, particularly when those algorithms are opaque or proprietary [@scoville_algorithmic_2021; @chapman_promoting_2021].

### 2022-06-03 manuscript/manuscript.Rmd
slight tweak

--- 2022-06-02 manuscript/manuscript.Rmd
@@
 This paper does not intend to validate deep RL as a method that should be used to directly inform decision-making on current conservation problems.
-Instead, we have selected example problems to provide ecologists with a greater background to engage in the collaborative development of deep RL-based methods, while also highlighting challenges which ecological problems pose to existing techniques.
+Rather, we seek to provide ecologists with a greater understanding of both potentials and pitfalls of this emerging approach. 
+We have selected familiar example problems to provide ecologists with a greater background and intuition to understand these techniques and engage in the collaborative development of deep RL-based methods, while also highlighting challenges which ecological problems pose to existing techniques.
 Validating deep RL for current conservation problems is beyond the scope of any one paper: this will necessitate examining a range of more "precise" models which will require more computational resources than is available to most researchers and extensive collaboration between large teams of ecologists and computer scientists.

### 2022-06-03 manuscript/manuscript.Rmd
there is no 'true optimal' policy

--- 2022-06-03 manuscript/manuscript.Rmd
@@
 Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
-In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
-While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.
+In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted -- a stark contrast to the management policy evidenced in the historical catch.
+Note that we have no way of knowing for sure if the RL quotas would have led to recovery, let alone a truly optimal harvest rate, because we can never know the "true model" of the Argentine hake dynamics.
+We can confirm that fishery closures seen in the RL agent's solution is also considered optimal under the assumptions of constant escapement theory [@Reed1979] whenever the stock is below the biomass of maximum sustainable yield ($B_{MSY}$), and that most fisheries models of this stock [@ramlegacy] would suggest that the stock sizes observed in the latter two decades of the data are below that threshold.
 

### 2022-06-03 manuscript/manuscript.Rmd
Small language edits

@@
 In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted -- a stark contrast to the management policy evidenced in the historical catch.
-Note that we have no way of knowing for sure if the RL quotas would have led to recovery, let alone a truly optimal harvest rate, because we can never know the "true model" of the Argentine hake dynamics.
-We can confirm that fishery closures seen in the RL agent's solution is also considered optimal under the assumptions of constant escapement theory [@Reed1979] whenever the stock is below the biomass of maximum sustainable yield ($B_{MSY}$), and that most fisheries models of this stock [@ramlegacy] would suggest that the stock sizes observed in the latter two decades of the data are below that threshold.
+Note that we have no way of knowing for sure if the RL quotas would have led to recovery nor do we know the optimal harvest rates, because we can never know the "true model" of the Argentine hake dynamics.
+We can confirm that fishery closures seen in the RL agent's solution are considered optimal under the assumptions of constant escapement theory [@Reed1979] whenever the stock is below the biomass of maximum sustainable yield ($B_{MSY}$), and that most fisheries models of this stock [@ramlegacy] would suggest that the populations observed in the latter two decades of the data are below that threshold.
 

### 2022-06-03 manuscript/manuscript.Rmd
adding an article

@@
 Note that we have no way of knowing for sure if the RL quotas would have led to recovery nor do we know the optimal harvest rates, because we can never know the "true model" of the Argentine hake dynamics.
-We can confirm that fishery closures seen in the RL agent's solution are considered optimal under the assumptions of constant escapement theory [@Reed1979] whenever the stock is below the biomass of maximum sustainable yield ($B_{MSY}$), and that most fisheries models of this stock [@ramlegacy] would suggest that the populations observed in the latter two decades of the data are below that threshold.
+We can confirm that the fishery closures seen in the RL agent's solution are considered optimal under the assumptions of constant escapement theory [@Reed1979] whenever the stock is below the biomass of maximum sustainable yield ($B_{MSY}$), and that most fisheries models of this stock [@ramlegacy] would suggest that the populations observed in the latter two decades of the data are below that threshold.
 

### 2022-07-05 manuscript/manuscript.Rmd
Changing names in Contributions

@@
 
-All authors were supported by the NSF Faculty Early Career Development Program Award 1942280. We thank Benjamin Blonder, Zachary Sunberg, and Claire Tomlin for their thoughtful comments.
+This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280. We thank Benjamin Blonder, Zachary Sunberg, and Claire Tomlin for their thoughtful comments.
 
@@
 
-M.L. and C.B. developed the code and wrote the manuscript with support from K.E.A.N. and M.S.C.
+Marcus Lapeyrolerie and Carl Boettiger developed the code and wrote the manuscript with support from Kari Norman and Melissa Chapman.
 

### 2022-07-06 manuscript/manuscript.Rmd
Editing captions

@@
 ```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario.
-The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g.
+The ecosystem begins in the desirable 'high' state under an environmental parameter (e.g.
 global mean temperature, arbitrary units) of 0.19.
@@
 rising mean temperature) as the parameter increases.
-This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.
+This results in only a slow degradation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is annihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.
 Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
@@
 Lower left: the policy function learned using TD3 relative to the policy function under the steady state.
-Lower right: mean rewards under TD3 management evenutally exceed those expected under the steady-state strategy as a large initial investment in conservation eventually pays off."}
+Lower right: mean rewards under TD3 management eventually exceed those expected under the steady-state strategy as a large initial investment in conservation eventually pays off."}
 

### 2022-07-06 manuscript/manuscript.Rmd
Adding doi link

@@
 
-We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures herein.
+We have created a GitHub repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures herein.
+The DOI for this repo is <https://zenodo.org/badge/latestdoi/318358820>.
 
